### 2021-06-08 tex/framework.tex
[ writing ] some bits about environments

\def\prefix{../../generic-lr/src/latex}

\CatchFileBetweenTags{\exItypes}%
{\prefix/Generic/Linear/Example/PaperExamples.tex}{exItypes}
\CatchFileBetweenTags{\exIlabels}%
{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIlabels}
\CatchFileBetweenTags{\exIfunrules}%
{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIfunrules}
\CatchFileBetweenTags{\exIsumrules}%
{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIsumrules}

\CatchFileBetweenTags{\Premises}%
{\prefix/Generic/Linear/Syntax.tex}{Premises}
\CatchFileBetweenTags{\Rule}%
{\prefix/Generic/Linear/Syntax.tex}{Rule}
\CatchFileBetweenTags{\System}%
{\prefix/Generic/Linear/Syntax.tex}{System}

\CatchFileBetweenTags{\semp}%
{\prefix/Generic/Linear/Syntax/Interpretation.tex}{semp}
\CatchFileBetweenTags{\semr}%
{\prefix/Generic/Linear/Syntax/Interpretation.tex}{semr}
\CatchFileBetweenTags{\sems}%
{\prefix/Generic/Linear/Syntax/Interpretation.tex}{sems}

\CatchFileBetweenTags{\SimplePremises}%
{\prefix/Generic/Simple/Syntax.tex}{SimplePremises}

\section{Building up}

We assume that a type system comprises a set of (unparametrised) rules, each
of which has a conclusion and several premises containing subterms.
The primary investigation of this work is into what form the premises can take
while maintaining useful features of syntax.
We shall start with simple forms, allowing just for multiple subterms, and
then build on resource-sensitive bunches, variable binding, and modalities.

\System{}
\Rule{}

Given some way \AgdaFunction{⟦\AgdaUnderscore{}⟧p} of interpreting
\AgdaDatatype{Premises} in a \AgdaRecord{Ctx}, we can interpret a
\AgdaDatatype{Rule} against a conclusion and a context by checking that its
stated conclusion matches and then interpreting its premises.
Then, the entire \AgdaDatatype{System} can be interpreted by picking a rule
label (including parameters) \AgdaBound{l} and interpreting the selected rule
\AgdaBound{rs}\AgdaSpace{}\AgdaBound{l}.

\semr{}
\sems{}

\subsection{The language of Cartesian products}

\SimplePremises

\section{Bunched functions}

Suppose we have a left $R$-semimodule $M$.
Then we can define the following connectives on indexed type families.

\begin{align*}
  \top~z &:= \top \\
  (P \wedge Q)~z &:= P~z \wedge Q~z \\
  (P \to Q)~z &:= P~z \to Q~z \\
  I~z &:= z \subres 0 \\
  (P * Q)~z &:= \exists x,y.~z \subres x+y \wedge P~x \wedge Q~y \\
  (P \wand Q)~y &:= \forall x,z.~z \subres x+y \to P~x \to Q~z \\
  (r \cdot P)~z &:= \exists x.~z \subres rx \wedge P~x
\end{align*}

\paragraph{Agda definitions}
To follow.

The first six of these connectives are somewhat standard, the first three
corresponding to the usual sharing connectives and the second three
corresponding to the usual separating connectives.
The final connective, $r \cdot P$, is new.
We choose $\exists$ rather than $\forall$ because scaling will usually occur in
premises, i.e., to the left of arrows, and we want to avoid anything
higher-order.

\section{Generic syntax}

Rules $R$, premises $P$ and $Q$.

\begin{align*}
  R &::= P/A \\
  P,Q &::= \grR\Theta \vdash A \mid \top \mid P \wedge Q \mid I \mid P * Q
        \mid \gr r \cdot P
\end{align*}

Example rules:

\begin{itemize}
  \item With-introduction: ${\vdash A} \wedge {\vdash B} / A \with B$
  \item Annotated arrow introduction:
    ${\gr rA \vdash B} / \gr rA \multimap B$
  \item Annotated arrow elimination:
    ${\vdash \gr rA \multimap B} * \gr r \cdot {\vdash A} / B$
  \item Cases of annotated sum:
    ${\vdash \gr rA \oplus \gr sB}
    * ({\gr rA \vdash C} \wedge {\gr sB \vdash C}) / C$
\end{itemize}

\paragraph{Agda example}

We start by defining our types \AgdaDatatype{Ty}, which in this case will be the
conclusions of our judgements.
\AgdaFunction{Ann} is the type of usage annotations, which for now is an
arbitrary skew semiring.
We also have a type of rule labels \AgdaDatatype{`Sys}, which specifies all of
the parameters of a typing rule except for its conclusion and premises.
For example, in the \AgdaInductiveConstructor{`case} label, we need to fill in
the two summands (types \AgdaBound{A} and \AgdaBound{B}), as well as the return
type \AgdaBound{C}.
The actual form of the rule will be specified later.

\exItypes{} \exIlabels{}

We are then ready to build a system \AgdaFunction{Sys} with conclusions
\AgdaDatatype{Ty}, usage annotations \AgdaFunction{Ann}, and labels
\AgdaDatatype{`Sys}.
The rule corresponding to each label is given in the following.

The \AgdaInductiveConstructor{`lam} rule has one premise, which binds one
variable with usage annotation \AgdaBound{r} and type \AgdaBound{A}.
The \AgdaInductiveConstructor{`app} rule has two premises combined with
separating conjunction.
Furthermore, the second premise (corresponding to the argument of the function)
is subject to scaling by \AgdaBound{r}.
Intuitively, this description means that we must have enough of each variable to
use some in building a function, and then use some more \AgdaBound{r} times to
build enough arguments.

\exIfunrules{}

In the rules for sums, we use an alternative notation for rule descriptions,
with \AgdaFunction{{---}{---}} being an infix version of
\AgdaInductiveConstructor{rule}.
The introduction rules \AgdaInductiveConstructor{`inl} and
\AgdaInductiveConstructor{`inr} are straightforward --- each has one
premise which binds no new variables.
The \AgdaInductiveConstructor{`case} rule is more complicated.
The premises are first split into the eliminand and the continuations, using the
separating conjunction we saw for \AgdaInductiveConstructor{`app}.
However, the continuations are connected via the sharing conjunction, reflecting
the fact that in any given run of the program, only one of the branches will be
taken, so usages from each individually should be added to usages of the
eliminand.

\exIsumrules{}

\paragraph{Interpretation as syntax}
We can say something like ``premise connectives are interpreted as the
corresponding bunched connectives, where appropriate''.

\begin{align*}
  \sem{P/A} &:= \sem P \to {- \vdash A} \\
  \sem{\grR\Theta \vdash A} &:= {-, \grR\Theta \vdash A} \\
  \sem{\top} &:= \top \\
  \sem{P \wedge Q} &:= \sem P \wedge \sem Q \\
  \sem{I} &:= I \\
  \sem{P * Q} &:= \sem P * \sem Q \\
  &\ldots
\end{align*}

\paragraph{Agda version}
The interpretation of a system is the selection of a rule, together with the
interpretation of that rule.

\sems{}

The interpretation of a rule is a check that the rule targets the desired
conclusion, together with the interpretation of the rules premises.

\semr{}

For the premises, connectives are interpreted in the obvious way.
Premises can ask for subterms via the $\AgdaInductiveConstructor{\_`⊢\_}$
constructor, which supplies the new variables (in particular, each variable's
type and usage annotation) \AgdaBound{Γ} and the desired conclusion
\AgdaBound{A} of the subterm.

\semp{}

\section{Values and computations}

Following \cite{AACMM20}, we call any semantic realisation of a variable a
\emph{value}, and any semantic realisation of a term a \emph{computation}.
Scoped families of values are named $\mathcal V$, while scoped families of
computations are named $\mathcal C$.
\todo{This wording is difficult.}
Values and computations are concepts with attitudes.

\section{Environments}

An environment is a linear map from variables in one context to values in
another.

\missingfigure{Agda definition of \AgdaRecord{\_{---}Env}}

An environment in \AgdaBound{PΓ} for \AgdaBound{$\mathcal V$}-values in
\AgdaBound{QΔ} is a linear map \AgdaField{M} such that $Q \le PM$ and for any
$P'$ and $Q'$ such that $Q' \le P'M$, a function from variables in $P'Γ$ to
\AgdaBound{$\mathcal V$}-values in $Q'Δ$.

\section{Thinnings}

\todo[inline]{I think thinnings should be called \emph{renamings}.}

In a usage-annotated sequent calculus, their are 4 basic structural rules:
exchange, weakening of $0$-annotated formulae, contraction of $+$-annotated
formulae, and subusaging.
Their typical forms are given below.

\begin{mathpar}
  \inferrule*[right=exch]
  {\grP\Gamma, \gr sB, \gr rA, \grQ\Delta \vdash Z}
  {\grP\Gamma, \gr rA, \gr sB, \grQ\Delta \vdash Z}
  \and
  \inferrule*[right=weak]
  {\grP\Gamma, \grQ\Delta \vdash Z}
  {\grP\Gamma, 0A, \grQ\Delta \vdash Z}
  \and
  \inferrule*[right=cont]
  {\grP\Gamma, \gr rA, \gr sA, \grQ\Delta \vdash Z}
  {\grP\Gamma, (\gr r + \gr s)A, \grQ\Delta \vdash Z}
  \and
  \inferrule*[right=subuse]
  {\grP \le \grQ \\\\ \grQ\Gamma \vdash Z}
  {\grP\Gamma \vdash Z}
\end{mathpar}

Generalising away from $\vdash$, it can be helpful to know when an arbitrary
scoped family respects these structural rules.
First, we note that the structural rules of sequent calculus correspond
directly to \emph{renaming} in natural deduction.
A renaming is an environment of variables, i.e., a \emph{linear} map from
variables to variables.
The linearity is what gives us the restrictions on annotations in the weakening
and contraction rules.

\missingfigure{Example renaming}

%A thinning is an environment of syntactic variables.
%In effect, thinning allows for permutation and slackening of existing
%variables, and introduction of new discardable variables at any position.
%These actions are in direct correspondence with the usual sequent calculus
%structural rules of exchange, contraction, and weakening, as well as subusaging.

Given this definition, $\square$ and \texttt{Thinnable} are as in AACMM.
\texttt{Kripke} is modified to use separating implication.

\section{Substitution}

\todo[inline]{This seems like a good place to introduce substitution, straight
after renaming, though we don't even have terms yet.}
Whereas a renaming is a linear map from variables to variables, a substitution
is a linear map from variables to terms.

\section{A layer of syntax is functorial}

### 2021-06-08 tex/lnl.tex
[ writing ] some bits about environments

We can express Benton's linear/non-linear logic~\cite{Benton94} in the
framework.
There are two apparent discrepancies between L/nL and what we have seen so far
to be expressible in the framework: the existence of multiple judgement modes
and the restriction on the kinds of assumptions based on the mode.
We will start by addressing these discrepancies, before presenting the encoding
in full and proving the resulting system logically equivalent to
$\lambda\gr{\mathcal R}$.

\subsection{Encoding L/nL}

In L/nL, we have a \emph{linear} and an \emph{intuitionistic} mode, with
separate types (respectively $A$ and $X$) and typing judgements (respectively
$\vdash_{\mathcal L}$ and $\vdash_{\mathcal C}$) connected by modalities $F$ and
$G$.
The mode of the conclusion type is the same as the mode of the judgement, so
only a distinction in the conclusion mode is necessary.
To achieve this distinction in the types, we have an indexed type family
$\mathrm{Ty} : \mathrm{Frag} \to \mathrm{Set}$, and set the framework types to
be the type $(f : \mathrm{Frag}) \times \mathrm{Ty}\,f$.
% \AgdaDatatype{Ty}\AgdaSymbol{:}\AgdaDatatype{Frag}\AgdaSymbol{$\to$}
% \AgdaDatatype{Set}

In L/nL, while linear judgements can have both linear and intuitionistic
assumptions, intuitionistic judgements can only have intuitionistic assumptions.
We encode this invariant on intuitionistic judgements using the $\Box$
premise combinator on all intuitionistic conclusions and wherever
intuitionistic premises occur for linear conclusions.

\subsection{Translating between L/nL and $\lambda\gr{\mathcal R}$}

The translations largely follow Benton's originals.

### 2021-06-18 tex/ndsc.tex
[ writing ] start on ND/SC background

In a seminal paper~\cite{Gentzen64}, Gerhard Gentzen introduces two syntactic
paradigms which remain among the most studied to this day.
These paradigms are \emph{natural deduction} and \emph{sequent calculus}, as
exemplified by natural deduction calculi NJ and NK, and sequent calculi LJ and
LK.
Restricting attention to just the intuitionistic systems NJ and LJ, these
actually differ in two orthogonal ways, which I shall prise apart in this
section.
The simpler distinction is that the logical rules in NJ are introduction and
elimination rules, whereas the logical rules in LJ are left and right rules.
But the more important distinction for this thesis is that where NJ has
assumptions, LJ has sequents explictly manipulated by structural rules.
I take the latter distinction to define natural deduction and sequent calculus,
and wherever I need to make the former distinction I shall speak of
\emph{intro-elim systems} and \emph{left-right systems}.

I will use the rest of this section as follows.
First, I introduce NJ and LJ, and some of their basic metatheory.
Then, I will give examples of systems intermediate between Gentzen's natural
deduction and sequent calculi: a left-right natural deduction calculus
$\mu\tilde\mu$, and an intro-elim sequent calculus DILL\@.
Both of these examples will reappear in later chapters.

\subsection{Intro-elim natural deduction: NJ}

\subsection{Left-right sequent calculus: LJ}

\subsection{Left-right natural deduction: $\mu\tilde\mu$}
The $\mu\tilde\mu$-calculus~\cite{CH00} (also known as
$\overline\lambda\mu\tilde\mu$ or system L, and closely related to Wadler's
dual calculus~\cite{Wadler03}) can be seen as an adaptation of natural deduction
to classical logic.
Though originally presented as a sequent calculus, the underlying natural
deduction calculus was later given by Herbelin~\cite[p.\ 12]{Herbelin-hab}, and
I will follow the latter.
While Gentzen gave a natural deduction calculus NK for classical logic, NK
relies on adding the \emph{axiom} of excluded middle.
As axioms are not systematic parts of the calculus, they can hinder or break
metatheoretic properties like normalisation.
In contrast, the $\mu\tilde\mu$-calculus allows us to \emph{derive} excluded
middle from entirely systematic components.

In NJ, a derivation of $A$ from assumptions $\Gamma$ tells us that if each
formula in $\Gamma$ is \emph{true}, then $A$ is also \emph{true}.
The $\mu\tilde\mu$-calculus generalises the picture by allowing us to have
both \emph{true} and \emph{false} assumptions, and allowing us to conclude that
some $A$ is \emph{true}, that some $A$ is \emph{false}, or that we have reached
a contradiction.
% A similar judgement of contradiction appears in Prawitz' classical natural
% deduction calculus~\cite{Prawitz65}.
Following Herbelin, we notate the judgement that $A$ is true as ${}\vdash A$,
that $A$ is false as $A \vdash{}$, and of contradiction as $\vdash$.
The only way to derive a contradiction is to find some $A$ such that
${}\vdash A$ and $A \vdash{}$.
Meanwhile, we can derive ${}\vdash A$ by assuming $A \vdash{}$ and deriving
$\vdash$, i.e., we can prove $A$ by assuming that $A$ is false and deriving a
contradiction.
Dually, we can derive $A \vdash{}$ by assuming ${}\vdash A$ and deriving
$\vdash$.
These three methods of derivation are encoded in the following rules.

\begin{mathpar}
  \inferrule*[right=Cut]
  {{}\vdash A \\ A \vdash{}}
  {\vdash}

  \and

  \inferrule*[right=$\mu$]
  {
    [A \vdash{}] \\\\ \vdots \\\\ \vdash
    %\inferrule*[fraction={~~~}]
    %{[A \vdash{}] \\\\ \vdots}
    %{\vdash}
  }
  {{}\vdash A}

  \and

  \inferrule*[right=$\tilde\mu$]
  {[{}\vdash A] \\\\ \vdots \\\\ \vdash}
  {A \vdash{}}
\end{mathpar}

The rules for logical connectives describe how to \emph{prove} and how to
\emph{refute} a formula whose principal connective is that connective.
These correspond strongly with the right and left rules, respectively, of LJ,
and for this reason, $\mu\tilde\mu$ is usually described elsewhere as a
sequent calculus.
For example, we could choose the following rules for disjunction.
To prove $A \vee B$, we can assume that both $A$ and $B$ are false, and derive
a contradiction.
To refute $A \vee B$, we can refute $A$ and $B$ separately.

\begin{mathpar}
  \inferrule*[right=$\vee$-r]
  {[A \vdash{}][B \vdash{}] \\\\ \vdots \\\\ \vdash}
  {{}\vdash A \vee B}

  \and

  \inferrule*[right=$\vee$-l]
  {A \vdash{} \\ B \vdash{}}
  {A \vee B \vdash{}}
\end{mathpar}

\subsection{Intro-elim sequent calculus: DILL}

### 2021-06-18 tex/ill.tex
[ writing ] background on ILL sequent calculus

\subsection{Motivation of linearity}

In LJ, two distinct, but equiderivable, canonical definitions of $\wedge$ can
be given.

In the first approach, $\wedge^-$ is conceived to be a
\emph{categorial product}.
Such a product is constructed by providing two maps in from the same antecedents
$\Gamma$.
It is used by projecting out one of the sides.

\begin{mathpar}
  \inferrule*[right=$\wedge^-$-r]
  {\Gamma \vdash A \\ \Gamma \vdash B}
  {\Gamma \vdash A \wedge^- B}

  \and

  \inferrule*[right=$\wedge^-$-l$_0$]
  {\Gamma, A \vdash C}
  {\Gamma, A \wedge^- B \vdash C}

  \and

  \inferrule*[right=$\wedge^-$-l$_1$]
  {\Gamma, B \vdash C}
  {\Gamma, A \wedge^- B \vdash C}
\end{mathpar}

In the second approach, $\wedge^+$ is conceived to be a \emph{tensor product}.
Intuitively, $\wedge^+$ internalises the comma of context concatenation.
Such a product is used by giving access to both sides simultaneously.
It is constructed by constructing both sides and combining the antecedents
required by each side.

\begin{mathpar}
  \inferrule*[right=$\wedge^+$-r]
  {\Gamma \vdash A \\ \Delta \vdash B}
  {\Gamma, \Delta \vdash A \wedge^+ B}

  \and

  \inferrule*[right=$\wedge^+$-l]
  {\Gamma, A, B \vdash C}
  {\Gamma, A \wedge^+ B \vdash C}
\end{mathpar}

When we prove the logical equivalence of these two formulations, we notice that
the structural rules of weakening and contraction are essential.
When we remove weakening and contraction, $\wedge^-$ and $\wedge^+$ become
logically distinct connectives, which we notate $\&$ and $\otimes$,
respectively.

\begin{mathpar}
  \inferrule*[right=$\wedge^-$-r]
  {%
    \inferrule*[right=Weak$^*$,fraction={\cdot\cdots\cdot}]
    {\Gamma \vdash A}
    {\Gamma, \Delta \vdash A}
    \\
    \inferrule*[Right=Weak$^*$,fraction={\cdot\cdots\cdot}]
    {\Delta \vdash B}
    {\Gamma, \Delta \vdash B}
  }
  {\Gamma, \Delta \vdash A \wedge^- B}

  \and

  \inferrule*[right=Contr]
  {%
    \inferrule*[Right=$\wedge^-$-l$_0$]
    {%
      \inferrule*[Right=$\wedge^-$-l$_1$]
      {\Gamma, A, B \vdash C}
      {\Gamma, A, A \wedge^- B \vdash C}
    }
    {\Gamma, A \wedge^- B, A \wedge^- B \vdash C}
  }
  {\Gamma, A \wedge^- B \vdash C}
\end{mathpar}

\begin{mathpar}
  \inferrule*[right=Contr$^*$,fraction={\cdot\cdots\cdot}]
  {%
    \mprset{defaultfraction}
    \inferrule*[Right=$\wedge^+$-r]
    {\Gamma \vdash A \\ \Gamma \vdash B}
    {\Gamma, \Gamma \vdash A \wedge^+ B}
  }
  {\Gamma \vdash A \wedge^+ B}

  \and

  \inferrule*[right=$\wedge^+$-l]
  {%
    \inferrule*[Right=Weak]
    {\Gamma, A \vdash C}
    {\Gamma, A, B \vdash C}
  }
  {\Gamma, A \wedge^+ B \vdash C}

  \and

  \inferrule*[right=$\wedge^+$-l]
  {%
    \inferrule*[Right=Weak]
    {\Gamma, B \vdash C}
    {\Gamma, A, B \vdash C}
  }
  {\Gamma, A \wedge^+ B \vdash C}
\end{mathpar}

\subsection{The $\oc$-modality}

\begin{mathpar}
  \inferrule*[right=Promotion]
  {\oc\Gamma \vdash A}
  {\oc\Gamma \vdash \oc A}

  \and

  \inferrule*[right=Dereliction]
  {\Gamma, A \vdash B}
  {\Gamma, \oc A \vdash B}

  \and

  \inferrule*[right=Weakening]
  {\Gamma \vdash B}
  {\Gamma, \oc A \vdash B}

  \and

  \inferrule*[right=Contraction]
  {\Gamma, \oc A, \oc A \vdash B}
  {\Gamma, \oc A \vdash B}
\end{mathpar}

In the intuitionistic linear logic sequent calculus ILL, $\oc A$ is defined
to be a proposition whose occurrences as antecedents can be deleted
(\TirName{Weakening}) and duplicated (\TirName{Contraction}), from which we can
extract $A$ (\TirName{Dereliction}), and which we can form from a conclusion
$A$ only when all antecedents are of the form $\oc X$ for some proposition $X$
(\TirName{Promotion}).
In short, $\oc A$ can be seen as an intuitionistic version of $A$, supporting
all of the structural rules of LJ, and only being able to be formed when it
does not depend on anything linear.

While this definition of $\oc$ works, in the sense that it gives the intended
class of models and cut elimination is maintained, it has some disadvantages.
Firstly, while the multiplicative and additive connectives are all characterised
by universal properties, $\oc$ is not.
This can be seen by the fact that taking the rules for $\oc$ and replacing each
occurrence of $\oc$ by a fresh connective $\oc'$ produces a logically distinct
connective.
One cannot produce any derivation of $\oc' A \vdash \oc A$ because
\TirName{Promotion} does not apply when there are antecedents not of the form
$\oc X$.
Finally, while $\oc$ is supposed to be a positive connective, it sometimes
behaves like a negative connective.
For example, for a positive connective like $\otimes$, the normal form proof
of the identity sequent $P \otimes Q \vdash P \otimes Q$ starts (from the
bottom) with a left rule and then, with the left in a more amenable form,
applies the right rule.
In contrast, the normal form proof of $\oc P \vdash \oc P$ starts with the
right rule, as it needs to have everything on the left be of the form $\oc X$.

\begin{mathpar}
  \inferrule*[right=$\otimes$-l]
  {%
    \inferrule*[Right=$\otimes$-r]
    {%
      \inferrule*[right=Id]{ }{P \vdash P} \\
      \inferrule*[Right=Id]{ }{Q \vdash Q}
    }
    {P, Q \vdash P \otimes Q}
  }
  {P \otimes Q \vdash P \otimes Q}

  \and

  \inferrule*[right=Pro]
  {%
    \inferrule*[Right=Der]
    {%
      \inferrule*[Right=Id]{ }{P \vdash P}
    }
    {\oc P \vdash P}
  }
  {\oc P \vdash \oc P}
\end{mathpar}

### 2021-06-18 tex/ndsc.tex
[ edit ] change example DILL -> BBdPH

@@
 deduction and sequent calculi: a left-right natural deduction calculus
-$\mu\tilde\mu$, and an intro-elim sequent calculus DILL\@.
+$\mu\tilde\mu$, and an intro-elim sequent calculus BBdPH\@.
 Both of these examples will reappear in later chapters.
@@
 
-\subsection{Intro-elim sequent calculus: DILL}
+\subsection{Intro-elim sequent calculus: BBdPH}
+
+Term assignment system introduced in~\cite{BBdPH93}.

### 2021-07-12 tex/kits.tex
[ writing ] lots

\def\prefix{../agda/latex}
\CatchFileBetweenTags{\Var}{\prefix/SimpleKits.tex}{Var}
\CatchFileBetweenTags{\Term}{\prefix/SimpleKits.tex}{Term}
\CatchFileBetweenTags{\Ren}{\prefix/SimpleKits.tex}{Ren}
\CatchFileBetweenTags{\bindRen}{\prefix/SimpleKits.tex}{bindRen}
\CatchFileBetweenTags{\rename}{\prefix/SimpleKits.tex}{rename}
\CatchFileBetweenTags{\leftTerm}{\prefix/SimpleKits.tex}{leftTerm}
\CatchFileBetweenTags{\Sub}{\prefix/SimpleKits.tex}{Sub}
\CatchFileBetweenTags{\bindSub}{\prefix/SimpleKits.tex}{bindSub}
\CatchFileBetweenTags{\substitute}{\prefix/SimpleKits.tex}{substitute}
\CatchFileBetweenTags{\Env}{\prefix/SimpleKits.tex}{Env}
\CatchFileBetweenTags{\RenSub}{\prefix/SimpleKits.tex}{RenSub}
\CatchFileBetweenTags{\Kit}{\prefix/SimpleKits.tex}{Kit}
\CatchFileBetweenTags{\trav}{\prefix/SimpleKits.tex}{trav}
\CatchFileBetweenTags{\renKit}{\prefix/SimpleKits.tex}{renKit}
\CatchFileBetweenTags{\subKit}{\prefix/SimpleKits.tex}{subKit}

\CatchFileBetweenTags{\RenGD}{\prefix/SimpleKits.tex}{RenGD}
\CatchFileBetweenTags{\RenGADA}{\prefix/SimpleKits.tex}{RenGADA}
\CatchFileBetweenTags{\GTh}{\prefix/SimpleKits.tex}{GTh}
\CatchFileBetweenTags{\DTh}{\prefix/SimpleKits.tex}{DTh}
\CatchFileBetweenTags{\GD}{\prefix/SimpleKits.tex}{GD}

\subsection{Simultaneous renaming and simultaneous substitution}

A simultaneous renaming from $\Gamma$ to $\Delta$ is a type-preserving map from
variables in $\Delta$ to \emph{variables} in $\Gamma$, while a simultaneous
substitution is a map into \emph{terms} in $\Gamma$.
While simultaneous substitution gives us a notion of one context being
\emph{derivable} from another, simultaneous renaming gives a similar notion
of derivability restricted to structural rules.

\begin{mathpar}
  \inferrule*[right=Subst]
  {%
    \inferrule*[right=$\to$-E]
    {%
      \inferrule*[right=Var]{~}{A \to B, A \vdash A \to B}
      \\
      \inferrule*[Right=Var]{~}{A \to B, A \vdash A}
    }
    {A \to B, A \vdash B}
    \\
    B \vdash C
  }
  {A \to B, A \vdash C}
\end{mathpar}

\begin{displaymath}
  \begin{prooftree}
    \infer0[Var]{A \to B, A \vdash A \to B}
    \infer0[Var]{A \to B, A \vdash A}
    \infer2[$\to$-E]{A \to B, A \vdash B}
    \hypo{B \vdash C}
    \infer2[Subst]{A \to B, A \vdash C}
  \end{prooftree}
\end{displaymath}

\subsection{Proofs of renaming and substitution}

We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
terms.
Beside base types, the only type former we have is the function type constructor
\AgdaInductiveConstructor{\_`→\_}.
Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
(\AgdaDatatype{Ty}).
Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
embeds types into contexts.
Typed variables in a context are given by \AgdaRecord{\_∋\_}.
A variable in
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.

\Var{}
\Term{}

For this syntax, a renaming from \AgdaBound{$\Gamma$} to \AgdaBound{$\Delta$}
is a map from variables in \AgdaBound{$\Delta$} to variables in
\AgdaBound{$\Gamma$}.
Substitutions instead map into terms in \AgdaBound{$\Gamma$}.

\Ren{}
\Sub{}

In the following, \AgdaFunction{ren} gives the action of a renaming on terms.
We replace variables by new variables given by the renaming \AgdaBound{$\rho$}.
For applications, we rename each subterm using the same renaming.
For $\lambda$-abstractions, we want to rename the body, but the renaming we
have has type \RenGD{}, whereas the renaming we need is of type \RenGADA{}.
To make up the difference, we introduce the \AgdaFunction{bindRen} lemma,
saying that any renaming can be extended to the right by a context
\AgdaBound{$\Theta$}~\footnote{%
  We only require extension to the right because our syntax only has binding
  on the right.
  We could also extend on the left.
}.
To produce such an extended renaming, we receive a variable in \DTh{}, with the
two cases being that this variable is either in \AgdaBound{$\Delta$} or
\AgdaBound{$\Theta$}.
In the first case, we can use the original renaming \AgdaBound{$\rho$} to get
a variable in \AgdaBound{$\Gamma$}, which can be extended in the obvious way
to a variable in \GTh{}.
In the second case, we have a variable in \AgdaBound{$\Theta$}, and just need
to extend it to be a variable in \GTh{}.

\bindRen{}
\rename{}

We use renaming to show that, like variables, we can extend the context of terms
to the right.
The operation below renames a term by replacing all variables with variables
pointing to the left of the term's new context \GD{}.

\leftTerm{}

The action of a substitution on a term is given below.
The structure is very similar to that of renaming.
The differences are the following:

\begin{itemize}
  \item A substitution gives us terms, rather than variables.
        This means that when we use the substitution on a variable, we already
        have a term, and don't need to do any embedding into terms.
  \item When dealing with newly bound variables in \AgdaFunction{bindSub}, we
        need to produce terms, so we embed the newly bound variables into terms
        using \AgdaInductiveConstructor{var}.
  \item We use \AgdaFunction{$\swarrow^t$} instead of
        \AgdaFunction{$\swarrow^v$}, because we are dealing with terms rather
        than variables.
\end{itemize}

\bindSub{}
\substitute{}

\subsection{Kits}

To abstract over the similarities between renaming and substitution, we can use
\emph{kits} as introduced by McBride~\cite{McBride05,BHKM12}.

### 2021-07-12 tex/lrkits.tex
[ writing ] lots

As we discussed in \autoref{sec:kits}, simultaneous substitution gives a
notion of derivability of one context from another, while simultaneous renaming
gives a similar notion of derivability restricted to structural rules.
To adapt these notions from an intuitionistic setting to our substructural
setting, we must examine what it means to derive one context from another
substructually.

\begin{displaymath}
  \begin{tikzpicture}[baseline]
    \path
    (-1,1) node (Gtop) {}
    (-1,0) node (G) {$\Gamma$}
    (-1,-1) node (Gbot) {}
    ;
    \node[draw,dotted,fit=(Gtop) (G) (Gbot)] (GG) {};

    \path
    (1,1) node (Dtop) {}
    (1,0) node (D) {$\Delta$}
    (1,-1) node (Dbot) {}
    ;
    \node[draw,dotted,fit=(Dtop) (D) (Dbot)] (DD) {};

    \draw[->,double] (GG) -- (DD);
  \end{tikzpicture}
  \coloneqq
  \begin{tikzpicture}[baseline]
    \path
    (-1,1) node (Gtop) {}
    (-1,0) node (G) {$\Gamma$}
    (-1,-1) node (Gbot) {}
    ;
    \node[draw,dotted,fit=(Gtop) (G) (Gbot)] (GG) {};

    \path
    (1,1) node[draw] (Dtop) {$\Delta_1$}
    (1,0) node (D) {$\vdots$}
    (1,-1) node[draw] (Dbot) {$\Delta_n$}
    ;

    \fill[green!20!white,opacity=1] (GG.north east)
    parabola[bend at end] (Dtop.west)
    parabola[bend at start] (GG.south east)
    -- cycle;
    \fill[blue!40!white,opacity=.5] (GG.north east)
    parabola[bend at end] (Dbot.west)
    parabola[bend at start] (GG.south east)
    -- cycle;

    \draw[->] (GG.north east) parabola[bend at end] (Dtop.west);
    \draw (GG.south east) parabola[bend at end] (Dtop.west);
    \draw[->] (GG.north east) parabola[bend at end] (D.west);
    \draw (GG.south east) parabola[bend at end] (D.west);
    \draw[->] (GG.north east) parabola[bend at end] (Dbot.west);
    \draw (GG.south east) parabola[bend at end] (Dbot.west);
  \end{tikzpicture}
\end{displaymath}

Semantically, we interpret a simple substitution from $\Gamma$ to $\Delta$ as
a map of type $\sem\Gamma \to \sem\Delta$.
With semantics of contexts given by a Cartesian product of their elements,
this type is
$\left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
\left(\prod_{(y : B) \in \Delta}\sem B\right)$.
We can then do the following rearrangement.
\todo{Internal vs external homs are not clear.}

\begin{align*}
  &\phantom{{}={}} \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
    \left(\prod_{(y : B) \in \Delta}\sem B\right) \\
  &\cong \prod_{(y : B) \in \Delta}\left(
         \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to \sem B
         \right) \\
  &= \prod_{(y : B) \in \Delta}\sem{\Gamma \vdash B} \\
  &\cong \prod_B\left(\Delta \ni B \to \sem{\Gamma \vdash B}\right)
\end{align*}

Taking the semantics to be the term model, we get the standard definition of a
simultaneous substitution.
To see how this definition works, let us construct an example substitution:
$A, B \to C, B \Longrightarrow B, C$.
Intuitionistically, it suffices to give two separate substitutions,
$A, B \to C, B \Longrightarrow B$ and $A, B \to C, B \Longrightarrow C$,
which are both just terms.
We indeed have terms $x : A, y : B \to C, z : B \vdash y : B$ and
$x : A, y : B \to C, z : B \vdash y\,z : C$.

When working with our semiring-annotated calculus \name{}, instead of Cartesian
products we have (weighted) monoidal products, giving the type
$\left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
\left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right)$.
As the monoidal product is not a limit, we cannot do the rearrangement we did
above.
Instead, we must divide the domain up into pieces, with each piece supporting
one factor from the codomain.

To see this division, let us adapt the previous example to have type
$\gr0A, \gr2(B \multimap C), \gr3B \Longrightarrow \gr1B, \gr2C$.
To divide the goal up, we note that $
\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix} +
\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}
$, so it suffices to give substitutions of types
$\gr0A, \gr0(B \multimap C), \gr1B \Longrightarrow \gr1B$ and
$\gr0A, \gr2(B \multimap C), \gr2B \Longrightarrow \gr2C$.
Furthermore, our term calculus only supports $\gr1$-annotated conclusions,
so we divide the second substitution type through by $\gr2$.
Finally, we give the terms largely as before:
$\gr0x : A, \gr0y : B \to C, \gr1z : B \vdash y : B$ and
$\gr0x : A, \gr1y : B \to C, \gr1z : B \vdash z\,y : C$.

\begin{displaymath}
  \begin{tikzpicture}[baseline]
    \path
    (-1,1) node (Gtop) {}
    (-1,0) node (G) {$\grP\gamma$}
    (-1,-1) node (Gbot) {}
    ;
    \node[draw,dotted,fit=(Gtop) (G) (Gbot)] (GG) {};

    \path
    (1,1) node (Dtop) {}
    (1,0) node (D) {$\grQ\delta$}
    (1,-1) node (Dbot) {}
    ;
    \node[draw,dotted,fit=(Dtop) (D) (Dbot)] (DD) {};

    \draw[->,double] (GG) -- (DD);
  \end{tikzpicture}
  \coloneqq
  \begin{tikzpicture}[baseline]
    \path
    (-1,1) node (Gtop) {}
    (-1,0) node (G) {$\grP\gamma$}
    (-1,-1) node (Gbot) {}
    ;
    \node[draw,dotted,fit=(Gtop) (G) (Gbot)] (GG) {};

    \path
    (1,3) node (G1top) {}
    (1,2) node (G1) {$\gr\Psi_1\gamma$}
    (1,1) node (G1bot) {}
    ;
    \node[draw,dotted,fit=(G1top) (G1) (G1bot)] (GG1) {};
    \draw[->] (GG) -- (GG1);

    \path (1,0) node {$\vdots$};

    \path
    (1,-1) node (Gntop) {}
    (1,-2) node (Gn) {$\gr\Psi_n\gamma$}
    (1,-3) node (Gnbot) {}
    ;
    \node[draw,dotted,fit=(Gntop) (Gn) (Gnbot)] (GGn) {};
    \draw[->] (GG) -- (GGn);

    \path
    (3,1) node[draw] (Dtop) {$\delta_1$}
    (3,0) node (D) {$\vdots$}
    (3,-1) node[draw] (Dbot) {$\delta_n$}
    ;

    \fill[green!20!white] (GG1.north east)
    parabola[bend at end] (Dtop.west)
    parabola[bend at start] (GG1.south east)
    -- cycle;
    \draw[->] (GG1.north east) parabola[bend at end] (Dtop.west);
    \draw (GG1.south east) parabola[bend at end] (Dtop.west);

    \fill[blue!20!white] (GGn.north east)
    parabola[bend at end] (Dbot.west)
    parabola[bend at start] (GGn.south east)
    -- cycle;
    \draw[->] (GGn.north east) parabola[bend at end] (Dbot.west);
    \draw (GGn.south east) parabola[bend at end] (Dbot.west);
  \end{tikzpicture}
  \quad\textrm{where }\grP = \grQ\gr\Psi
\end{displaymath}

Generalising this division process, we derive the following (letting
$\Gamma = \grP\gamma$ and $\Delta = \grQ\delta$).
The various fragments of the original usage vector are collected up in the
existentially quantified $\gr\Psi$.

\begin{align*}
  &\phantom{{}={}}
    \left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
    \left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right) \\
  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
    \left(\grP = \sum\gr\Psi\right) \times
    \prod_{(\gr sy : B) \in \Delta}\left(
    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
    \to \oc \gr s\sem B
    \right) \\
  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
    \left(\grP = \sum_{(\gr sy : B) \in \Delta}\gr s\gr\Psi_y\right) \times
    \prod_{(\gr sy : B) \in \Delta}\left(
    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
    \to \sem B
    \right) \\
  &= \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
    \left(\grP = \grQ\gr\Psi\right) \times
    \prod_{(\gr sy : B) \in \Delta} \sem{\gr\Psi_y\gamma \vdash B}
\end{align*}

In other words, we interpret a linear substitution as a \emph{matrix}
$\gr\Psi$ relating the two usage contexts, together with a term for each
variable from $\Delta$ typed in a fragmented usage context dictated by
$\gr\Psi$.

While we naturally derive a matrix as a fragmentation of a usage vector, we can
get a slightly cleaner presentation by instead using an abstract linear map.
Let $\gr\Psi$ now be a linear map of type
$\Ann^{\size\Delta} \to \Ann^{\size\Gamma}$, with application written postfix.
The equation $\grP = \grQ\gr\Psi$ remains unchanged.
As for $\prod_{(\gr sy : B) \in \Delta}\sem{\gr\Psi_y\gamma \vdash B}$, the most
direct fix is to replace $\gr\Psi_y$ by $\langle y \rvert\gr\Psi$.
But then we notice that $\langle y \rvert$ is exactly the $\grQprime$ satisfying
$\grQprime\delta \sqni y : B$.
Thus, let us derive the following.

\begin{align*}
  &\phantom{{}={}} \prod_{(\gr sy : B) \in \Delta}
    \sem{\langle y \rvert \gr\Psi\gamma \vdash B} \\
  &\cong \prod_{B,\grQprime} \left(\grQprime\delta \sqni B \to
    \sem{\grQprime\gr\Psi\gamma \vdash B}\right) \\
  &\cong \prod_{B,\grQprime,\grPprime} \left(
    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
    \sem{\grPprime\gamma \vdash B}\right)
\end{align*}

Putting everything back together gives us the final expression.

\begin{align*}
  &\sum_{\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}}
    \left(\grP = \grQ\gr\Psi\right) \times
    \prod_{B,\grQprime,\grPprime} \left(
    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
    \sem{\grPprime\gamma \vdash B}\right)
\end{align*}

We now have a new reading for the interpretation of a linear substitution:
a linear map $\gr\Psi$ relating the two usage vectors $\grP$ and $\grQ$, and
for any two similarly related usage vectors $\grPprime$ and $\grQprime$, we
have a type-preserving function from variables in $\grQprime\delta$ to terms in
$\grPprime\gamma$.
Even though we don't use $\gr\Psi$ as a matrix containing fragmented usage
vectors, we can still justify why it should be a \emph{linear} map.
We need $\gr\Phi$ to respect all fragmentation of the usage context in a typing
rule, and we know that all such fragmentation is done by linear operations
zero, addition, and scaling by a constant.
\todo{Expand. Substitutions need to preserve everything done to the context,
and linear things are all we do to the context.}

### 2021-07-12 tex/multi.tex
[ writing ] lots

In type theory and categorical logic, the idea of multicategories is to give
an algebraic structure that is close to the syntax of type theory.
In simple type theory, sequents take the form $\Gamma \vdash A$, where $A$ is
the type of the conclusion and $\Gamma$ is a list of types, one type for each
assumption.
Intuitively, a term with assumptions $\Gamma$ and type $A$ constitutes a
morphism from $\Gamma$ to $A$.
Multicategories make this structure --- domains being lists of objects and
codomains being a single object --- part of the definition of morphisms.

\begin{definition}[multicategory]
  A \emph{multicategory} comprises a collection of objects $\obj$, for each
  list of objects $\Gamma$ and object $A$ a set of (multi)morphisms
  $\hom(\Gamma, A)$, and the following morphisms, satisfying the following
  axioms.

  \begin{tikzpicture}
    \path
    (0,1) node(s) {$A$}
    (0,0) node[draw](id) {$\id$}
    (0,-1) node(t) {$A$}
    ;

    \draw[->] (s) -- (id);
    \draw[->] (id) -- (t);
  \end{tikzpicture}
  \begin{tikzpicture}
    \path
    (-2,2) node(A11) {$A_{11}$}
    (-1.5,1.5) node(A1dots) {$\cdots$}
    (-1,2) node(A1n) {$A_{1n}$}
    (0,2) node(Adots) {$\cdots$}
    (1,2) node(Am1) {$A_{m1}$}
    (1.5,1.5) node(Amdots) {$\cdots$}
    (2,2) node(Amn) {$A_{mn}$}

    (-1.5,1) node[draw](f1) {$f_1$}
    (0,1) node(fdots) {$\cdots$}
    (1.5,1) node[draw](fm) {$f_m$}

    (0,0) node[draw](g) {$g$}

    (0,-1) node(C) {$C$}
    ;

    \node[draw,dotted,fit=(f1) (fm) (g),
    label=below right:{\footnotesize$g \circ (f_1, \ldots, f_m)$}] (box) {};

    \draw[->] (A11) -- (f1);
    \draw[->] (A1n) -- (f1);
    \draw[->] (Am1) -- (fm);
    \draw[->] (Amn) -- (fm);
    \draw[->] (f1) -- (g);
    \draw[->] (fm) -- (g);
    \draw[->] (g) -- (C);
  \end{tikzpicture}

  \begin{displaymath}
    \begin{matrix}
      \begin{tikzpicture}[baseline]
        \path
        (-1,2) node(A1) {$A_1$}
        (0,2) node(Adots) {$\cdots$}
        (1,2) node(An) {$A_n$}
        (0,1) node[draw](f) {$f$}
        (0,0) node[draw](id) {$\id$}
        (0,-1) node(t) {$A$}
        ;

        \node[draw,dotted,fit=(f) (id)] {};

        \draw[->] (A1) -- (f);
        \draw[->] (An) -- (f);
        \draw[->] (f) -- (id);
        \draw[->] (id) -- (t);
      \end{tikzpicture}
      =
      \begin{tikzpicture}[baseline]
        \path
        (-1,1) node(A1) {$A_1$}
        (0,1) node(Adots) {$\cdots$}
        (1,1) node(An) {$A_n$}
        (0,0) node[draw](f) {$f$}
        (0,-1) node(t) {$A$}
        ;

        \draw[->] (A1) -- (f);
        \draw[->] (An) -- (f);
        \draw[->] (f) -- (t);
      \end{tikzpicture}
      &\phantom{mmmm}&
      \begin{tikzpicture}[baseline]
        \path
        (-1,2) node(A1) {$A_1$}
        (0,2) node(Adots) {$\cdots$}
        (1,2) node(Am) {$A_m$}
        (-1,1) node[draw](id1) {$\id$}
        (0,1) node(iddots) {$\cdots$}
        (1,1) node[draw](idm) {$\id$}
        (0,0) node[draw](f) {$f$}
        (0,-1) node(t) {$A$}
        ;

        \node[draw,dotted,fit=(id1) (idm) (f)] {};

        \draw[->] (A1) -- (id1);
        \draw[->] (Am) -- (idm);
        \draw[->] (id1) -- (f);
        \draw[->] (idm) -- (f);
        \draw[->] (f) -- (t);
      \end{tikzpicture}
      =
      \begin{tikzpicture}[baseline]
        \path
        (-1,1) node(A1) {$A_1$}
        (0,1) node(Adots) {$\cdots$}
        (1,1) node(An) {$A_n$}
        (0,0) node[draw](f) {$f$}
        (0,-1) node(t) {$A$}
        ;

        \draw[->] (A1) -- (f);
        \draw[->] (An) -- (f);
        \draw[->] (f) -- (t);
      \end{tikzpicture}
    \end{matrix}
  \end{displaymath}

  \begin{displaymath}
    \begin{matrix}
      \begin{tikzpicture}
        \path
        % left
        (-5,2) node(A111) {$A_{111}$}
        (-4.5,1.5) node(A1dots) {$\cdots$}
        (-4,2) node(A11n) {$A_{11n}$}
        (-3,2) node(Adots) {$\cdots$}
        (-2,2) node(A1m1) {$A_{1m1}$}
        (-1.5,1.5) node(Amdots) {$\cdots$}
        (-1,2) node(A1mn) {$A_{1mn}$}

        (-4.5,1) node[draw](f11) {$f_{11}$}
        (-3,1) node(fdots) {$\cdots$}
        (-1.5,1) node[draw](f1m) {$f_{1m}$}

        (-3,0) node[draw](g1) {$g_1$}

        % right
        (1,2) node(Al11) {$A_{l11}$}
        (1.5,1.5) node(A1dots) {$\cdots$}
        (2,2) node(Al1n) {$A_{l1n}$}
        (3,2) node(Adots) {$\cdots$}
        (4,2) node(Alm1) {$A_{lm1}$}
        (4.5,1.5) node(Amdots) {$\cdots$}
        (5,2) node(Almn) {$A_{lmn}$}

        (1.5,1) node[draw](fl1) {$f_{l1}$}
        (3,1) node(fdots) {$\cdots$}
        (4.5,1) node[draw](flm) {$f_{lm}$}

        (3,0) node[draw](gl) {$g_l$}

        % middle
        (0,2) node {$\cdots$}
        (0,0.5) node {$\cdots$}
        (0,-1) node[draw](h) {$h$}
        (0,-2) node(D) {$D$}
        ;

        \node[draw,dotted,fit=(f11) (f1m) (g1)] (box1) {};
        \node[draw,dotted,fit=(fl1) (flm) (gl)] (boxl) {};
        \node[draw,dotted,fit=(box1) (boxl) (h)] (box) {};

        \draw[->] (A111) -- (f11);
        \draw[->] (A11n) -- (f11);
        \draw[->] (A1m1) -- (f1m);
        \draw[->] (A1mn) -- (f1m);
        \draw[->] (f11) -- (g1);
        \draw[->] (f1m) -- (g1);
        \draw[->] (g1) -- (h);

        \draw[->] (Al11) -- (fl1);
        \draw[->] (Al1n) -- (fl1);
        \draw[->] (Alm1) -- (flm);
        \draw[->] (Almn) -- (flm);
        \draw[->] (fl1) -- (gl);
        \draw[->] (flm) -- (gl);
        \draw[->] (gl) -- (h);

        \draw[->] (h) -- (D);
      \end{tikzpicture}
      \\=\\
      \begin{tikzpicture}
        \path
        % left
        (-5,2) node(A111) {$A_{111}$}
        (-4.5,1.5) node(A1dots) {$\cdots$}
        (-4,2) node(A11n) {$A_{11n}$}
        (-3,2) node(Adots) {$\cdots$}
        (-2,2) node(A1m1) {$A_{1m1}$}
        (-1.5,1.5) node(Amdots) {$\cdots$}
        (-1,2) node(A1mn) {$A_{1mn}$}

        (-4.5,1) node[draw](f11) {$f_{11}$}
        (-3,1) node(fdots) {$\cdots$}
        (-1.5,1) node[draw](f1m) {$f_{1m}$}

        (-3,0) node[draw](g1) {$g_1$}

        % right
        (1,2) node(Al11) {$A_{l11}$}
        (1.5,1.5) node(A1dots) {$\cdots$}
        (2,2) node(Al1n) {$A_{l1n}$}
        (3,2) node(Adots) {$\cdots$}
        (4,2) node(Alm1) {$A_{lm1}$}
        (4.5,1.5) node(Amdots) {$\cdots$}
        (5,2) node(Almn) {$A_{lmn}$}

        (1.5,1) node[draw](fl1) {$f_{l1}$}
        (3,1) node(fdots) {$\cdots$}
        (4.5,1) node[draw](flm) {$f_{lm}$}

        (3,0) node[draw](gl) {$g_l$}

        % middle
        (0,2) node {$\cdots$}
        (0,1) node {$\cdots$}
        (0,0) node {$\cdots$}
        (0,-1) node[draw](h) {$h$}
        (0,-2) node(D) {$D$}
        ;

        \node[draw,dotted,fit=(g1) (gl) (h)] (boxi) {};
        \node[draw,dotted,fit=(f11) (f1m) (fl1) (flm) (boxi)] (boxo) {};

        \draw[->] (A111) -- (f11);
        \draw[->] (A11n) -- (f11);
        \draw[->] (A1m1) -- (f1m);
        \draw[->] (A1mn) -- (f1m);
        \draw[->] (f11) -- (g1);
        \draw[->] (f1m) -- (g1);
        \draw[->] (g1) -- (h);

        \draw[->] (Al11) -- (fl1);
        \draw[->] (Al1n) -- (fl1);
        \draw[->] (Alm1) -- (flm);
        \draw[->] (Almn) -- (flm);
        \draw[->] (fl1) -- (gl);
        \draw[->] (flm) -- (gl);
        \draw[->] (gl) -- (h);

        \draw[->] (h) -- (D);
      \end{tikzpicture}
    \end{matrix}
  \end{displaymath}
\end{definition}

Multicategories have been used as a framework for reasoning with multilinear
maps in linear algebra. \todo{Reference?}
We can produce a multicategory where the objects are vector spaces, and the
multimorphisms are multilinear maps.
In this setting, we can give a universal property to the tensor product of
vector spaces.

\begin{definition}[tensor product \& tensor unit]
  Given a multicategory $\C$, a \emph{tensor product} in $\C$ is a function
  ${\otimes} : \obj\C \times \obj\C \to \obj\C$ and for each pair of objects
  $A$ and $B$, a multimorphism ${\otimes} : A, B \to A \otimes B$ such that,
  for any multimorphism $f : A, B \to C$, there is a unique way to factor $f$
  through $\otimes$, as shown below.
  Similarly, a \emph{tensor unit} is an object $I$ of $\C$ and a multimorphism
  $I : \varepsilon \to I$ supporting a nullary unique factorisation.

  \begin{tikzcd}
    A, B \arrow[rd,"f"'] \arrow[r,"\otimes"] & A \otimes B \arrow[d, dashed] \\
    & C
  \end{tikzcd}
  \begin{tikzcd}
    \varepsilon \arrow[rd,"f"'] \arrow[r,"I"] & I \arrow[d, dashed] \\
    & C
  \end{tikzcd}
\end{definition}

As the proliferation of ellipses suggests, the definition of
\emph{multicategory} I gave above is not entirely rigorous.
Indeed, mechanising the multicategory definition in a reasonably usable way is
considered an open problem. \todo{Check that no-one has done it}
However, we \emph{can} achieve a simple and usable definition in the special
case of \emph{Cartesian} multicategories.

Ordinary multicategories are to monoidal categories as Cartesian multicategories
are to Cartesian categories (i.e., categories with all finite products).
Cartesian multicategories can be defined in terms of ordinary multicategories
--- they are multicategories satisfying the usual ``structural rules'' below
and satisfying various coherence conditions on them.

\begin{align*}
  e &: \hom(\Gamma, A, B, \Delta; C) \to \hom(\Gamma, B, A, \Delta; C) \\
  w &: \hom(\Gamma; B) \to \hom(\Gamma, A; B) \\
  c &: \hom(\Gamma, A, A; B) \to \hom(\Gamma, A; B)
\end{align*}

However, we can bypass ordinary multicategories entirely, and give the following
definition inspired by our earlier formulation of intuitionistic logic.
I define Cartesian multicategories in tandem with the category of contexts and
substitutions that a Cartesian multicategory yields.

\begin{definition}
  A \emph{Cartesian multicategory} comprises the following.

  \begin{itemize}
    \item We have a collection of objects $\obj$.
          We call a list of objects a \emph{context}.
    \item For each context $\Gamma$ and object $A$, we have a set of
          (multi)morphisms $\hom(\Gamma; A)$.
          From this, we derive, for any contexts $\Gamma$ and $\Delta$, a set
          of \emph{substitutions}
          $\sub(\Gamma; \Delta) \coloneqq \prod X \in \Delta.~\hom(\Gamma; X)$.
    \item For every $i : A \in \Gamma$, we have an identity morphism
          $\id(i) : \hom(\Gamma; A)$.
          The collection of identity morphisms over $\Gamma$ serves as the
          identity substitution $\id^s : \sub(\Gamma; \Gamma)$.
    \item For substitution $\sigma : \sub(\Gamma; \Delta)$ and morphism
          $f : \hom(\Delta; A)$, we get a composite morphism
          $\sigma ; f : \hom(\Gamma; A)$.
          This allows us to compose substitutions; for
          $\sigma : \sub(\Gamma; \Delta)$ and $\tau : \sub(\Delta; \Theta)$,
          let $\sigma;^s \tau \coloneqq \lambda k.~\sigma; \tau(k)$.
    \item Identity and composition satisfy the following laws.
          \begin{itemize}
            \item $\sigma; \id(i) = \sigma(i)$
            \item $\id^s; f = f$
            \item $(\sigma;^s \tau); f = \sigma; (\tau; f)$
          \end{itemize}
          It is a simple exercise to see that these laws exactly give us the
          category laws of identity and associativity for substitutions.
  \end{itemize}
\end{definition}

The category of contexts and substitutions is furthermore Cartesian, with the
product being given by context concatenation.

### 2021-07-12 tex/semirings.tex
[ writing ] lots

Addition gives the structural rules.
Multiplication gives modalities.

Compare with other work.
About here, I can talk about Granule-style $\otimes$-pattern matching.

\subsection{Partiality (or not)} \todo{Appendix?}

As we have seen \todo{Have we?}, the way additive and multiplicative rules are
realised algebraically is related to models of separation logic.
Models of separation logic typically use \emph{partial} commutative monoids to
model a heap, so it is tempting to generalise the commutative monoid of
addition in our semirings to a \emph{partial} commutative monoid.
However, we find that the most natural notion of \emph{partial semiring} is
degenerate, in that all partial semirings are actually (total) semirings.

Recall that a commutative monoid (or commutative monoid object) can be
defined in any symmetric monoidal category.
A partial commutative monoid is exactly a commutative monoid object in the
category of sets and partial functions with the usual monoidal product
\todo{Define this monoidal product somewhere.}.
However, semirings need a Cartesian category in order to state the interaction
equations between addition and multiplication.
While the category of sets and partial functions is not Cartesian, the
standard way to manufacture a Cartesian category out of a symmetric monoidal
category $\mathcal C$ is to take the category of cocommutative comonoids
$\mathrm{CComon}(\mathcal C)$.
Intuitively, the cocommutative comonoid structure equips the underlying
object $M$ with a \emph{delete} map $\eta : M \to I$ and a \emph{duplicate}
map $\delta : M \to M \otimes M$ which are coherent with respect to each other.
All morphisms in $\mathrm{CComon}(\mathcal C)$ must respect $\eta$ and
$\delta$; in particular, both addition and multiplication must separately
form bimonoids in $\mathcal C$ together with the cocommutative comonoid.

\[
  \begin{tikzpicture}[baseline]
    \path
    (-1,1) node(0) {0}
    (1,2) node(x) {}
    (0,0) node(*) {*}
    (0,-1) node(res) {}
    ;

    \draw (0) -- (*);
    \draw (x) to[out=270,in=45] (*);
    \draw (*) -- (res);
  \end{tikzpicture}
  =\quad
  \begin{tikzpicture}[baseline]
    \path
    (0,0) node(0) {0}
    (0,2) node(x) {}
    (0,-1) node(res) {}
    (0,1) node[circle,draw](del) {}
    ;

    \draw (0) -- (res);
    \draw (x) -- (del);
  \end{tikzpicture}
  \quad=
  \begin{tikzpicture}[baseline]
    \path
    (1,1) node(0) {0}
    (-1,2) node(x) {}
    (0,0) node(*) {*}
    (0,-1) node(res) {}
    ;

    \draw (0) -- (*);
    \draw (x) to[out=270,in=135] (*);
    \draw (*) -- (res);
  \end{tikzpicture}
\]
\begin{displaymath}
  \begin{matrix}
    \begin{tikzpicture}[baseline]
      \path
      (-1,2) node(x) {}
      (0,2) node(y) {}
      (-0.5,1) node(+) {+}
      (1,2) node(z) {}
      (0,0) node(*) {*}
      (0,-1) node(res) {}
      ;

      \draw (x) to[out=270,in=135] (+);
      \draw (y) to[out=270,in=45] (+);
      \draw (+) to[out=270,in=135] (*);
      \draw (z) to[out=270,in=45] (*);
      \draw (*) -- (res);
    \end{tikzpicture}
    =
    \begin{tikzpicture}[baseline]
      \path
      (-2,3) node(x) {}
      (-1,3) node(y) {}
      (0,3) node(z) {}
      (0,2) node[circle,draw](dup) {}
      (-1,1) node(x*) {*}
      (0,1) node(y*) {*}
      (-0.5,0) node(+) {+}
      (-0.5,-1) node(res) {}
      ;

      \draw (z) -- (dup);
      \draw (x) to[out=270,in=135] (x*);
      \draw (y) to[out=270,in=135] (y*);
      \draw (dup) to[out=180,in=45] (x*);
      \draw (dup) -- (y*);
      \draw (x*) to[out=270,in=135] (+);
      \draw (y*) to[out=270,in=45] (+);
      \draw (+) -- (res);
    \end{tikzpicture}
    &\phantom{mmmm}&
    \begin{tikzpicture}[baseline]
      \path
      (1,2) node(x) {}
      (0,2) node(y) {}
      (0.5,1) node(+) {+}
      (-1,2) node(z) {}
      (0,0) node(*) {*}
      (0,-1) node(res) {}
      ;

      \draw (x) to[out=270,in=45] (+);
      \draw (y) to[out=270,in=135] (+);
      \draw (+) to[out=270,in=45] (*);
      \draw (z) to[out=270,in=135] (*);
      \draw (*) -- (res);
    \end{tikzpicture}
    =
    \begin{tikzpicture}[baseline]
      \path
      (2,3) node(x) {}
      (1,3) node(y) {}
      (0,3) node(z) {}
      (0,2) node[circle,draw](dup) {}
      (1,1) node(x*) {*}
      (0,1) node(y*) {*}
      (0.5,0) node(+) {+}
      (0.5,-1) node(res) {}
      ;

      \draw (z) -- (dup);
      \draw (x) to[out=270,in=45] (x*);
      \draw (y) to[out=270,in=45] (y*);
      \draw (dup) to[out=0,in=135] (x*);
      \draw (dup) -- (y*);
      \draw (x*) to[out=270,in=45] (+);
      \draw (y*) to[out=270,in=135] (+);
      \draw (+) -- (res);
    \end{tikzpicture}
  \end{matrix}
\end{displaymath}

\begin{lemma}
  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, there is
  a cocommutative comonoid over $X$.
\end{lemma}
\begin{proof}
  Let $\eta(x) \coloneq ()$ and $\delta(x) \coloneq (x, x)$, with both
  being defined for all $x$.
  Checking that these satisfy the laws is routine.
  Alternatively, we can see that both $\eta$ and $\delta$, being total, are
  morphisms in $\mathrm{Set}$, where it is well known that they form a
  cocommutative comonoid.
  The equations in $\mathrm{Set}$ carry over to $\mathrm{Set}_{\mathrm{part}}$.
\end{proof}

\begin{lemma}
  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, any
  comonoid over $X$ is the one described in the previous lemma.
\end{lemma}
\begin{proof}
  The left unit law says that, for all $x$ and $x'$, we have
  $\exists y.~\delta(x) = (y, x') \land \eta(y) = ()$ if and only if $x = x'$.
  Letting $x'$ be $x$ and reading from right to left, we get that there is
  some $y$ such that $\delta(x) = (y, x)$ and $\eta(y) = ()$.
  Symmetrically, from the right unit law, there is some $z$ such that
  $\delta(x) = (x, z)$ and $\eta(z) = ()$.
  But because $\delta$, as a partial function, is deterministic, we have
  $(y, x) = (x, z)$, giving us that $y = z = x$, and $\delta(x) = (x, x)$.
  Moreover, $\eta(x) = \eta(y) = ()$.
\end{proof}

### 2021-07-13 tex/kits.tex
[ writing ] finish kits section

@@
         using \AgdaInductiveConstructor{var}.
-  \item We use \AgdaFunction{$\swarrow^t$} instead of
-        \AgdaFunction{$\swarrow^v$}, because we are dealing with terms rather
-        than variables.
+  \item We use \AgdaFunction{$↙ᵗ$} instead of \AgdaFunction{$↙ᵛ$}, because we
+        are dealing with terms rather than variables.
 \end{itemize}
@@
 \emph{kits} as introduced by McBride~\cite{McBride05,BHKM12}.
+Each of the three differences above is turned into a parameter of
+\AgdaFunction{trav} (generalising \AgdaFunction{ren} and \AgdaFunction{sub})
+and \AgdaFunction{bindEnv} (generalising \AgdaFunction{bindRen} and
+\AgdaFunction{bindSub}).
+In the types, \AgdaFunction{Ren} and \AgdaFunction{Sub} are generalised by
+\AgdaFunction{Env}\AgdaSpace{}\AgdaBound{K} --- a function from variables in
+\AgdaBound{$\Delta$} to \AgdaBound{K}-things in \AgdaBound{$\Gamma$}.
+
+\Env{}
+
+The function \AgdaFunction{trav} produces a term-to-term mapping based on an
+environment \AgdaBound{$\rho$}.
+Like renaming and substitution, the traversal \AgdaFunction{trav} replaces
+variables according to \AgdaBound{$\rho$}, while keeping the rest of the
+syntactic forms intact.
+The three differences between renaming and substitution present themselves as
+requirements of the notion of \emph{kit} we can choose:
+
+\begin{itemize}
+  \item In the \AgdaInductiveConstructor{var} case of \AgdaFunction{trav}, we
+        apply the environment \AgdaBound{$\rho$} to get a \AgdaBound{K}-thing,
+        and need something to turn this \AgdaBound{K}-thing into a term.
+        We let \AgdaField{tm} be this context- and type-preserving map from
+        \AgdaBound{K}-things to terms.
+  \item When dealing with newly bound variables in \AgdaFunction{bindEnv}, we
+        need to convert the new variable into a \AgdaBound{K}-thing in order to
+        put it into the environment.
+        We let \AgdaField{vr} be this context- and type-preserving map from
+        variables to \AgdaBound{K}-things.
+  \item When working out where to map old variables in an extended context, we
+        need \AgdaBound{K}-things to be stable under context extensions.
+        We let \AgdaField{↙ᵏ} be the function embedding a \AgdaBound{K}-thing
+        into an extended context.
+\end{itemize}
+
+\trav{}
+
+The three parameters can be given to these functions by filling the fields of
+the \AgdaRecord{Kit} record.
+
+\Kit{}
+
+We may now redefine the types \AgdaFunction{Ren} and \AgdaFunction{Sub} via
+\AgdaFunction{Env}, and derive the actions of renaming and substitution from
+\AgdaFunction{trav}.
+Notice that \AgdaFunction{↙ᵗ} (written out as
+\AgdaFunction{ren}\AgdaSpace{}\AgdaFunction{↙ᵛ}) still relies on renaming, but
+because \AgdaFunction{ren} is only being used to fill a parameter of
+\AgdaFunction{trav}, \AgdaFunction{trav} itself can be used to define
+\AgdaFunction{ren} in a non-circular way.
+Thus, we have succeeded in avoiding the duplication of code between
+\AgdaFunction{ren} and \AgdaFunction{sub}.
+%\todo{Distribute code horizontally}
+
+\begin{multicols}{3}
+  \noindent\RenSub{} \columnbreak
+
+  \noindent\renKit{} \columnbreak
+
+  \noindent\subKit{}
+\end{multicols}
+
+%\noindent
+%\begin{tabular}{l|l|l}
+%  \begin{minipage}{.25\textwidth}
+%    \RenSub{}
+%  \end{minipage}
+%  &
+%  \begin{minipage}{.25\textwidth}
+%    \renKit{}
+%  \end{minipage}
+%  &
+%  \begin{minipage}{.25\textwidth}
+%    \subKit{}
+%  \end{minipage}
+%\end{tabular}

### 2021-07-27 tex/framework.tex
[ fix ] path to submodule generic-lr
[ refactor ] lagda to separate file

@@
-\def\prefix{../../generic-lr/src/latex}
-
-\CatchFileBetweenTags{\exItypes}%
-{\prefix/Generic/Linear/Example/PaperExamples.tex}{exItypes}
-\CatchFileBetweenTags{\exIlabels}%
-{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIlabels}
-\CatchFileBetweenTags{\exIfunrules}%
-{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIfunrules}
-\CatchFileBetweenTags{\exIsumrules}%
-{\prefix/Generic/Linear/Example/PaperExamples.tex}{exIsumrules}
-
-\CatchFileBetweenTags{\Premises}%
-{\prefix/Generic/Linear/Syntax.tex}{Premises}
-\CatchFileBetweenTags{\Rule}%
-{\prefix/Generic/Linear/Syntax.tex}{Rule}
-\CatchFileBetweenTags{\System}%
-{\prefix/Generic/Linear/Syntax.tex}{System}
-
-\CatchFileBetweenTags{\semp}%
-{\prefix/Generic/Linear/Syntax/Interpretation.tex}{semp}
-\CatchFileBetweenTags{\semr}%
-{\prefix/Generic/Linear/Syntax/Interpretation.tex}{semr}
-\CatchFileBetweenTags{\sems}%
-{\prefix/Generic/Linear/Syntax/Interpretation.tex}{sems}
-
-\CatchFileBetweenTags{\SimplePremises}%
-{\prefix/Generic/Simple/Syntax.tex}{SimplePremises}
-
 \section{Building up}

### 2021-07-27 tex/kits.tex
[ refactor ] lagda to separate file
[ redraft ] kits introduction

--- 2021-07-12 tex/kits.tex
@@
-\def\prefix{../agda/latex}
-\CatchFileBetweenTags{\Var}{\prefix/SimpleKits.tex}{Var}
-\CatchFileBetweenTags{\Term}{\prefix/SimpleKits.tex}{Term}
-\CatchFileBetweenTags{\Ren}{\prefix/SimpleKits.tex}{Ren}
-\CatchFileBetweenTags{\bindRen}{\prefix/SimpleKits.tex}{bindRen}
-\CatchFileBetweenTags{\rename}{\prefix/SimpleKits.tex}{rename}
-\CatchFileBetweenTags{\leftTerm}{\prefix/SimpleKits.tex}{leftTerm}
-\CatchFileBetweenTags{\Sub}{\prefix/SimpleKits.tex}{Sub}
-\CatchFileBetweenTags{\bindSub}{\prefix/SimpleKits.tex}{bindSub}
-\CatchFileBetweenTags{\substitute}{\prefix/SimpleKits.tex}{substitute}
-\CatchFileBetweenTags{\Env}{\prefix/SimpleKits.tex}{Env}
-\CatchFileBetweenTags{\RenSub}{\prefix/SimpleKits.tex}{RenSub}
-\CatchFileBetweenTags{\Kit}{\prefix/SimpleKits.tex}{Kit}
-\CatchFileBetweenTags{\trav}{\prefix/SimpleKits.tex}{trav}
-\CatchFileBetweenTags{\renKit}{\prefix/SimpleKits.tex}{renKit}
-\CatchFileBetweenTags{\subKit}{\prefix/SimpleKits.tex}{subKit}
-
-\CatchFileBetweenTags{\RenGD}{\prefix/SimpleKits.tex}{RenGD}
-\CatchFileBetweenTags{\RenGADA}{\prefix/SimpleKits.tex}{RenGADA}
-\CatchFileBetweenTags{\GTh}{\prefix/SimpleKits.tex}{GTh}
-\CatchFileBetweenTags{\DTh}{\prefix/SimpleKits.tex}{DTh}
-\CatchFileBetweenTags{\GD}{\prefix/SimpleKits.tex}{GD}
-
 \subsection{Simultaneous renaming and simultaneous substitution}
@@
 
-\begin{mathpar}
-  \inferrule*[right=Subst]
-  {%
-    \inferrule*[right=$\to$-E]
-    {%
-      \inferrule*[right=Var]{~}{A \to B, A \vdash A \to B}
-      \\
-      \inferrule*[Right=Var]{~}{A \to B, A \vdash A}
-    }
-    {A \to B, A \vdash B}
-    \\
-    B \vdash C
-  }
-  {A \to B, A \vdash C}
-\end{mathpar}
+In the derivation below, we assume the existence of a derivation of
+$B, C \to C \vdash C$, and by the admissibility of substitution we thus have a
+derivation of $A \to B, A \vdash C$.
+Intuitively, the context $A \to B, A$ derives the context $B, C \to C$, so
+anything derived from $B, C \to C$ can also be derived from $A \to B, A$.
+We see formally that $A \to B, A$ derives $B, C \to C$ by deriving each element
+of the latter from the former --- hence the first two premises of the Subst
+rule below, deriving $B$ and $C \to C$ from $A \to B, A$.
 
-\begin{displaymath}
+\begin{align*}
+  &\begin{prooftree}
+    \hypo{\Pi}
+    \infer[no rule]1{A \to B, A \vdash B}
+    \infer0[Var]{A \to B, A, C \vdash C}
+    \infer1[$\to$-I]{A \to B, A \vdash C \to C}
+    \hypo{B, C \to C \vdash C}
+    \infer3[Subst]{A \to B, A \vdash C}
+  \end{prooftree}
+  \\\\
+  &\textrm{where }\Pi \coloneqq
   \begin{prooftree}
@@
     \infer2[$\to$-E]{A \to B, A \vdash B}
-    \hypo{B \vdash C}
-    \infer2[Subst]{A \to B, A \vdash C}
   \end{prooftree}
-\end{displaymath}
+\end{align*}
 
-\subsection{Proofs of renaming and substitution}
-
-We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
-terms.
-Beside base types, the only type former we have is the function type constructor
-\AgdaInductiveConstructor{\_`→\_}.
-Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
-(\AgdaDatatype{Ty}).
-Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
-embeds types into contexts.
-Typed variables in a context are given by \AgdaRecord{\_∋\_}.
-A variable in
-\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
-is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
-with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
-Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.
-
-\Var{}
-\Term{}
+\subsection{Proofs of admissibility of renaming and substitution}
 
@@
 
-In the following, \AgdaFunction{ren} gives the action of a renaming on terms.
-We replace variables by new variables given by the renaming \AgdaBound{$\rho$}.
-For applications, we rename each subterm using the same renaming.
-For $\lambda$-abstractions, we want to rename the body, but the renaming we
-have has type \RenGD{}, whereas the renaming we need is of type \RenGADA{}.
-To make up the difference, we introduce the \AgdaFunction{bindRen} lemma,
-saying that any renaming can be extended to the right by a context
-\AgdaBound{$\Theta$}~\footnote{%
-  We only require extension to the right because our syntax only has binding
-  on the right.
-  We could also extend on the left.
-}.
-To produce such an extended renaming, we receive a variable in \DTh{}, with the
-two cases being that this variable is either in \AgdaBound{$\Delta$} or
-\AgdaBound{$\Theta$}.
-In the first case, we can use the original renaming \AgdaBound{$\rho$} to get
-a variable in \AgdaBound{$\Gamma$}, which can be extended in the obvious way
-to a variable in \GTh{}.
-In the second case, we have a variable in \AgdaBound{$\Theta$}, and just need
-to extend it to be a variable in \GTh{}.
+We define \AgdaFunction{ren} as the action of a renaming on terms.
+The idea of simultaneous renaming is to preserve the structure of the term, but
+replace all of the variables from \AgdaBound{$\Delta$} by variables from
+\AgdaBound{$\Gamma$}, with the mapping given by the renaming \AgdaBound{$\rho$}.
+
+\rename{}
+
+The \AgdaInductiveConstructor{var} case is where the action of the renaming
+happens: the variable \AgdaBound{x} from \AgdaBound{$\Delta$} is mapped to the
+variable \AgdaBound{$\rho$}\AgdaSpace{}\AgdaBound{x} from \AgdaBound{$\Gamma$}.
+In the \AgdaInductiveConstructor{app} case, we have terms \AgdaBound{M}
+\AgdaSymbol{:} \AgdaBound{$\Delta$} \AgdaDatatype{$\vdash$}
+\AgdaBound{A} \AgdaInductiveConstructor{`$\to$} \AgdaBound{B} and \AgdaBound{N}
+\AgdaSymbol{:} \AgdaBound{$\Delta$} \AgdaDatatype{$\vdash$} \AgdaBound{A}.
+We may apply \AgdaFunction{ren} \AgdaBound{$\rho$} recursively to both
+\AgdaBound{M} and \AgdaBound{N} to change their contexts from
+\AgdaBound{$\Delta$} to \AgdaBound{$\Gamma$}, and the
+\AgdaInductiveConstructor{app} constructor then produces the desired term in
+\AgdaBound{$\Gamma$}.
+Finally, in the \AgdaInductiveConstructor{lam} case, we get a term
+\AgdaBound{M} \AgdaSymbol{:} \DA{} \AgdaDatatype{$\vdash$} \AgdaBound{B} and,
+after introducing a \AgdaInductiveConstructor{lam} on the right, are in need
+of a term of type \GA{} \AgdaDatatype{$\vdash$} \AgdaBound{B}.
+To recursively apply \AgdaFunction{ren} to \AgdaBound{M}, we must thus extend
+the renaming \AgdaBound{$\rho$} \AgdaSymbol{:} \RenGD{} with the newly bound
+variable.
+For this, we need an auxiliary function \AgdaFunction{bindRen} such that
+\AgdaFunction{bindRen} \AgdaBound{$\rho$} \AgdaSymbol{:} \RenGADA{}.
+This new renaming will act like \AgdaBound{$\rho$} for variables in
+\AgdaBound{$\Delta$}, and map the new variable of type \AgdaBound{A} to the
+corresponding new variable in \GA{}.
 
 \bindRen{}
-\rename{}
 
-We use renaming to show that, like variables, we can extend the context of terms
-to the right.
-The operation below renames a term by replacing all variables with variables
-pointing to the left of the term's new context \GD{}.
+The \AgdaFunction{bindRen} given here has a slightly generalised type, where
+instead of binding just a single variable of type \AgdaBound{A}, we could
+bind a whole context \AgdaBound{$\Theta$} of new variables.
+The first case of \AgdaFunction{bindRen} is for old variables from
+\AgdaBound{$\Delta$}, where we apply \AgdaBound{$\rho$} to get a variable in
+\AgdaBound{$\Gamma$}, and then use \AgdaFunction{$↙ᵛ$} to embed that variable
+into \GTh{}.
+The second case is for new variables from \AgdaBound{$\Theta$}, which embed
+straight into \GTh{}.
+
+The \AgdaFunction{sub} function below gives the action of a substitution.
+Similarly to renaming, we want to preserve the structure of the term, except
+now variables in the original term are replaced by \emph{terms} in the new
+context.
+
+\substitute{}
+
+Given that this time, \AgdaBound{$\rho$} is a substitution rather than a
+renaming, \AgdaBound{$\rho$} \AgdaBound{x} is a term, and is sufficient in the
+\AgdaInductiveConstructor{var} case.
+The \AgdaInductiveConstructor{app} case again deals with the subterms
+recursively and then recombines them with \AgdaInductiveConstructor{app}.
+In the \AgdaInductiveConstructor{lam} case, we again have a mismatch if we
+want to apply \AgdaFunction{sub} recursively to the subterm \AgdaBound{M} with
+an extra free variable.
+We have \AgdaBound{$\rho$} \AgdaSymbol{:} \SubGD{} but need a substitution of
+type \SubGADA{}, so we introduce the auxiliary definition
+\AgdaFunction{bindSub}.
+
+\bindSub{}
+
+For the old variables in the first case, we have \AgdaBound{$\rho$} to turn
+them into terms in \AgdaBound{$\Gamma$}.
+Turning a term in \AgdaBound{$\Gamma$} into a term in \GTh{} requires a form
+of weakening we have not yet proved, so I write \AgdaFunction{$↙ᵗ$} in analogy
+with \AgdaFunction{$↙ᵛ$}, and prove it below.
+In the second case, we want to substitute the new variable by the \emph{term}
+referring to this new variable in \GTh{}.
+
+The final piece to define substitution is to define the function that weakens
+a term by some newly bound variables \AgdaBound{$\Delta$}.
+For this, we use the action of renaming, which we have fully defined already,
+and in particular rename each variable in the term from a variable in
+\AgdaBound{$\Gamma$} to a variable in \GD{}.
 
@@
 
-The action of a substitution on a term is given below.
-The structure is very similar to that of renaming.
-The differences are the following:
+With this, the action of substitution is defined, and depends on the action
+of renaming.
+
+\subsection{Syntactic kits}\label{sec:syntactic-kits}
+
+The statements of simultaneous renaming and simultaneous substitution are
+very similar, with substitution being the generalisation that allows
+replacement by terms rather than just variables.
+I will introduce a type family \AgdaFunction{Env} of \emph{environments}, and
+redefine \AgdaFunction{Ren} and \AgdaFunction{Sub} as environments of
+variables and terms, respectively.
+
+\Env{}
+\RenSub{}
+
+The processes I described for constructing proofs of the admissibility of
+renaming and substitution were also similar.
+Indeed, when we line up the resulting functions, \AgdaFunction{ren} and
+\AgdaFunction{sub}, and their auxiliaries, \AgdaFunction{bindRen} and
+\AgdaFunction{bindSub}, we notice only three key
+differences:~\cite{McBride05,BHKM12}
 
 \begin{itemize}
-  \item A substitution gives us terms, rather than variables.
-        This means that when we use the substitution on a variable, we already
-        have a term, and don't need to do any embedding into terms.
-  \item When dealing with newly bound variables in \AgdaFunction{bindSub}, we
-        need to produce terms, so we embed the newly bound variables into terms
-        using \AgdaInductiveConstructor{var}.
-  \item We use \AgdaFunction{$\swarrow^t$} instead of
-        \AgdaFunction{$\swarrow^v$}, because we are dealing with terms rather
-        than variables.
+  \item In the first cases of \AgdaFunction{bindRen} and \AgdaFunction{bindSub},
+    we do \AgdaFunction{$↙ᵛ$} and \AgdaFunction{$↙ᵗ$}, respectively, based on
+    whether we are weakening a variable or a term.
+  \item In the second case of \AgdaFunction{bindSub}, we do an extra wrapping of
+    the new variable by \AgdaInductiveConstructor{var}, so as to make it a term
+    to go in the substitution.
+  \item In the \AgdaInductiveConstructor{var} case of \AgdaFunction{ren}, we
+    have \AgdaInductiveConstructor{var} \AgdaSymbol(\AgdaBound{$\rho$}
+    \AgdaBound{x}\AgdaSymbol) rather than just \AgdaBound{$\rho$} \AgdaBound{x},
+    because the renaming \AgdaBound{$\rho$} gives us a variable rather than a
+    term.
 \end{itemize}
 
-\bindSub{}
-\substitute{}
+We may abstract over these three differences using the record \AgdaRecord{Kit}.
+As in \AgdaFunction{Env}, we think of \AgdaBound{K} as being either
+\AgdaDatatype{\_$\ni$\_} or \AgdaDatatype{\_$\vdash$\_}.
+The fields of \AgdaRecord{Kit} are given in the same order as the points of
+difference above.
+Wherever the difference was presence or absence of
+\AgdaInductiveConstructor{var}, we will be able to fill that field with either
+\AgdaInductiveConstructor{var} or the identity function \AgdaFunction{id}.
 
-\subsection{Kits}
+\Kit{}
 
-To abstract over the similarities between renaming and substitution, we can use
-\emph{kits} as introduced by McBride~\cite{McBride05,BHKM12}.
+The field \AgdaField{$\swarrow^k$} can be seen as a property of the
+judgement form \AgdaBound{K}, saying that it supports a form of weakening.
+We use \AgdaField{vr} when adding a newly bound variable to an environment, and
+use \AgdaField{tm} when we do a lookup from an environment and want to get a
+term out.
+Given a \AgdaRecord{Kit} \AgdaBound{K}, we can write the syntactic traversal
+function \AgdaFunction{trav} and its auxiliary \AgdaFunction{bindEnv}, in the
+model of \AgdaFunction{ren}, \AgdaFunction{sub}, and their auxiliaries.
+
+\trav{}
+\bindEnv{}
+
+Concrete kits can be given for variables and terms either by inspecting
+\AgdaFunction{ren} and \AgdaFunction{sub} or by following the types.
+Notice that the kit for terms requires the admissibility of renaming so as to
+achieve weakening of a substitution by newly bound variables.
+Fortunately, this can be the \AgdaFunction{ren} defined below in terms of
+\AgdaFunction{trav}, so we can keep \AgdaFunction{trav} as the only syntactic
+traversal we have to write.
+
+\begin{multicols}{2}
+  \noindent\renKit{} \columnbreak
+
+  \noindent\subKit{}
+\end{multicols}

### 2021-07-27 tex/lrkits.tex
[ redraft ] linear environments

@@
 substructually.
+
+In the intuitionistic case, we say that to derive a context $\Delta$ from a
+context $\Gamma$ is to derive each element $\Delta_i$ from $\Gamma$.
+We may justify this by an intermediate step --- noting that contexts are
+understood to be Cartesian products of their elements, and giving a map into
+a Cartesian product is the same as giving a map into each factor.
+I picture this definition as the diagram below.
 
@@
 
-Semantically, we interpret a simple substitution from $\Gamma$ to $\Delta$ as
-a map of type $\sem\Gamma \to \sem\Delta$.
-With semantics of contexts given by a Cartesian product of their elements,
-this type is
-$\left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
-\left(\prod_{(y : B) \in \Delta}\sem B\right)$.
-We can then do the following rearrangement.
-\todo{Internal vs external homs are not clear.}
-
-\begin{align*}
-  &\phantom{{}={}} \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
-    \left(\prod_{(y : B) \in \Delta}\sem B\right) \\
-  &\cong \prod_{(y : B) \in \Delta}\left(
-         \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to \sem B
-         \right) \\
-  &= \prod_{(y : B) \in \Delta}\sem{\Gamma \vdash B} \\
-  &\cong \prod_B\left(\Delta \ni B \to \sem{\Gamma \vdash B}\right)
-\end{align*}
-
-Taking the semantics to be the term model, we get the standard definition of a
-simultaneous substitution.
 To see how this definition works, let us construct an example substitution:
 $A, B \to C, B \Longrightarrow B, C$.
-Intuitionistically, it suffices to give two separate substitutions,
-$A, B \to C, B \Longrightarrow B$ and $A, B \to C, B \Longrightarrow C$,
-which are both just terms.
+Because the codomain is a Cartesian product, it suffices to give two separate
+substitutions, $A, B \to C, B \Longrightarrow B$ and
+$A, B \to C, B \Longrightarrow C$, with a substitution into a singleton
+context being just a term.
 We indeed have terms $x : A, y : B \to C, z : B \vdash y : B$ and
 $x : A, y : B \to C, z : B \vdash y\,z : C$.
+It is also instructive to look at an identity substitution (which is also a
+renaming), $A, B \Longrightarrow A, B$, witnessed by terms
+$x : A, y : B \vdash x : A$ and $x : A, y : B \vdash y : B$.
 
-When working with our semiring-annotated calculus \name{}, instead of Cartesian
-products we have (weighted) monoidal products, giving the type
-$\left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
-\left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right)$.
-As the monoidal product is not a limit, we cannot do the rearrangement we did
-above.
-Instead, we must divide the domain up into pieces, with each piece supporting
-one factor from the codomain.
+When working with our semiring-annotated calculus \name{}, contexts are no
+longer understood as Cartesian products.
+This means that substitutions of type $\Gamma \Longrightarrow \Delta$ are no
+longer equivalent to collections of substitutions
+$\Gamma \Longrightarrow \Delta_i$.
+Indeed, notice that we should still have an identity substitution of type
+$\gr1A, \gr1B \Longrightarrow \gr1A, \gr1B$, but we do not have terms proving
+either $\gr1A, \gr1B \vdash A$ or $\gr1A, \gr1B \vdash B$.
+What we do have are terms $x : \gr1A, y : \gr0B \vdash x : A$ and
+$x : \gr0A, y : \gr1B \vdash y : B$, and if we pointwise add together the
+annotations of the two terms, we get back the original context
+$x : \gr1A, y : \gr1B$.
+Furthermore, adding up the annotations is not just a random operation;
+linear contexts are understood to be tensor products of their elements, and
+introduction rule for the tensor product involves summing the annotations of
+the two sides.
 
-To see this division, let us adapt the previous example to have type
-$\gr0A, \gr2(B \multimap C), \gr3B \Longrightarrow \gr1B, \gr2C$.
-To divide the goal up, we note that $
-\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
-\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix} +
-\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}
-$, so it suffices to give substitutions of types
-$\gr0A, \gr0(B \multimap C), \gr1B \Longrightarrow \gr1B$ and
-$\gr0A, \gr2(B \multimap C), \gr2B \Longrightarrow \gr2C$.
-Furthermore, our term calculus only supports $\gr1$-annotated conclusions,
-so we divide the second substitution type through by $\gr2$.
-Finally, we give the terms largely as before:
-$\gr0x : A, \gr0y : B \to C, \gr1z : B \vdash y : B$ and
-$\gr0x : A, \gr1y : B \to C, \gr1z : B \vdash z\,y : C$.
+For any annotated context $\Delta$, we have
+$\Delta \vdash \bigotimes_{(\gr rx : A) \in \Delta}\oc\gr rA$ by iterated
+application of $\otimes$-I with $\oc$-I and Var at the leaves.
+Let $\Gamma = \grP\gamma$ and $\Delta = \grQ\delta$.
+If we are to produce substitutions from $\Gamma$ to $\Delta$ in this
+pattern, we simulate the applications of $\otimes$-I by producing, for each
+element in $\Delta$, a usage context for $\gamma$ such that the whole collection
+sums to $\grP$, then simulate the applications of $\oc$-I by dividing each of
+the new usage contexts by the corresponding annotation in $\Delta$, calling
+the divided usage contexts $\gr\Psi_x$, and finally, instead of a variable
+from $\delta$, we give a term of type $\gr\Psi_x\gamma \vdash \delta_x$.
+In summary, the constraint on the collection of usage contexts $\gr\Psi$ is
+that $\grP = \sum_{(\gr rx : A) \in \Delta}\gr r\gr\Psi_x$.
+Moreover, if we take $\grP$ and $\grQ$ to be row vectors and $\gr\Psi$ to be a
+matrix, the latter expression is equal to the vector-matrix multiplication
+$\grQ\gr\Psi$.
+The resulting definition of simultaneous substitution is depicted below.
 
@@
 
-Generalising this division process, we derive the following (letting
-$\Gamma = \grP\gamma$ and $\Delta = \grQ\delta$).
-The various fragments of the original usage vector are collected up in the
-existentially quantified $\gr\Psi$.
+In type theory, we write out the definition as follows.
 
-\begin{align*}
-  &\phantom{{}={}}
-    \left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
-    \left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right) \\
-  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
-    \left(\grP = \sum\gr\Psi\right) \times
-    \prod_{(\gr sy : B) \in \Delta}\left(
-    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
-    \to \oc \gr s\sem B
-    \right) \\
-  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
-    \left(\grP = \sum_{(\gr sy : B) \in \Delta}\gr s\gr\Psi_y\right) \times
-    \prod_{(\gr sy : B) \in \Delta}\left(
-    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
-    \to \sem B
-    \right) \\
-  &= \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
+\begin{displaymath}
+  \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
     \left(\grP = \grQ\gr\Psi\right) \times
-    \prod_{(\gr sy : B) \in \Delta} \sem{\gr\Psi_y\gamma \vdash B}
-\end{align*}
+    \prod_{(x : A) \in \delta}\gr\Psi_x\gamma \vdash A
+\end{displaymath}
 
-In other words, we interpret a linear substitution as a \emph{matrix}
-$\gr\Psi$ relating the two usage contexts, together with a term for each
-variable from $\Delta$ typed in a fragmented usage context dictated by
-$\gr\Psi$.
+We can see the step-by-step construction of a substitution play out by adapting
+the previous example to have type
+$\gr0A, \gr2(B \multimap C), \gr3B \Longrightarrow \gr1B, \gr2C$.
+To split the goal up, we note that $
+\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
+\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix} +
+\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}
+$, so it suffices to give substitutions of types
+$\gr0A, \gr0(B \multimap C), \gr1B \Longrightarrow \gr1B$ and
+$\gr0A, \gr2(B \multimap C), \gr2B \Longrightarrow \gr2C$.
+Furthermore, our term calculus only supports $\gr1$-annotated conclusions,
+so we divide the second substitution type through by $\gr2$.
+Finally, we give the terms largely as before:
+$\gr0x : A, \gr0y : B \to C, \gr1z : B \vdash y : B$ and
+$\gr0x : A, \gr1y : B \to C, \gr1z : B \vdash z\,y : C$.
 
@@
 The equation $\grP = \grQ\gr\Psi$ remains unchanged.
-As for $\prod_{(\gr sy : B) \in \Delta}\sem{\gr\Psi_y\gamma \vdash B}$, the most
-direct fix is to replace $\gr\Psi_y$ by $\langle y \rvert\gr\Psi$.
-But then we notice that $\langle y \rvert$ is exactly the $\grQprime$ satisfying
-$\grQprime\delta \sqni y : B$.
-Thus, let us derive the following.
+Where we previously wrote $\gr\Psi_x$, the most direct replacement would be
+$\langle x \rvert\gr\Psi$, with $\langle x \rvert$ being the $x$th basis row
+vector.
+But then we notice that $\langle x \rvert$ is exactly the $\grQprime$ satisfying
+$\grQprime\delta \sqni x : B$.
+This gives us the following definition, which can be verified by equationally
+substituting $\grPprime$ and expanding the definition of $\sqni$.
 
-\begin{align*}
-  &\phantom{{}={}} \prod_{(\gr sy : B) \in \Delta}
-    \sem{\langle y \rvert \gr\Psi\gamma \vdash B} \\
-  &\cong \prod_{B,\grQprime} \left(\grQprime\delta \sqni B \to
-    \sem{\grQprime\gr\Psi\gamma \vdash B}\right) \\
-  &\cong \prod_{B,\grQprime,\grPprime} \left(
-    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
-    \sem{\grPprime\gamma \vdash B}\right)
-\end{align*}
-
-Putting everything back together gives us the final expression.
-
-\begin{align*}
-  &\sum_{\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}}
+\begin{displaymath}
+  \sum_{\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}}
     \left(\grP = \grQ\gr\Psi\right) \times
-    \prod_{B,\grQprime,\grPprime} \left(
-    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
-    \sem{\grPprime\gamma \vdash B}\right)
-\end{align*}
+    \prod_{A,\grQprime,\grPprime} \left(
+    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni A \to
+    \grPprime\gamma \vdash A\right)
+\end{displaymath}
 
@@
 and linear things are all we do to the context.}
+
+Taking a lead from \autoref{sec:kits}, we deduce a definition of
+\emph{environment} by replacing the $\vdash$ in the definition of simultaneous
+substitution by an arbitrary type family $\mathcal V$.
+Letting $\mathcal V$ be $\sqni$ gives us a notion of simultaneous renaming,
+allowing for renamings with types such as
+$\gr6A, \gr0B, \gr1C \stackrel\sqni\Longrightarrow \gr1C, \gr2A, \gr4A$.
+
+It is worth noting that, when contexts are Cartesian products, passing from
+``a map into $\Delta$'' to ``for each $A \in \Delta$, a map into $A$'' does not
+lose any generality because the universal property of Cartesian products
+states that every map into a Cartesian product can be given factor-wise.
+Hence, $\Delta$ and the one-element context $\prod_{A \in \Delta}A$ are
+isomorphic in the category of contexts and simultaneous substitution.
+However, tensor products are not limits, so don't have the same universal
+property.
+Indeed, many annotated contexts are not isomorphic to the weighted product of
+their elements.
+For example, we do not have a substitution of type
+$\gr1(A \otimes B) \Longrightarrow \gr1A, \gr1B$ because we would first need
+to pattern-match on the tensor product \emph{before} trying to derive the
+target context.
+This loss of generality is however justified when we consider the action of a
+substitution.
+Substitutions should only be replacing variables by terms, whereas if
+substitutions were allowed to pattern-match before introducing the target, then
+the substitution would have to replace the original term by a term that first
+pattern-matches and then continues like the original term.

### 2021-07-27 tex/lenv.tex
[ writing ] environment properties

\begin{remark}
  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
  and a $\grQprime$ such that $\grPprime = \grQprime\plr{\rho.\gr\Psi}$,
  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
  with the same linear map and action on variables.
\end{remark}
\begin{proof}
  The only part of the definition of an environment dependent on $\grP$ or
  $\grQ$ is the constraint $\grP = \grQ\gr\Psi$, which we are able to replace
  for $\grPprime$ and $\grQprime$.
\end{proof}

When constructing an environment, we can do so by cases on the shape of the
target context.
We can create an environment into the empty context when all usage annotations
on the source context are $\gr0$.
We can create an environment into a concatenated context when we can additively
split up the annotations of the source context and produce environments into
both halves from the split sources.
We can create an environment into a singleton context when there is a context
$\gr r$ times smaller than the source context in which we can produce a value
of the appropriate type.

\begin{lemma}\label{thm:construct-env}
  We can define all of the following equivalences for any values of the free
  variables.
  \begin{itemize}
    \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
    \item $\forallb{\plr{{-} \env\V \Gamma} \sep \plr{{-} \env\V \Delta}
      \dotlr \plr{{-} \env\V \Gamma, \Delta}}$
    \item
      $\forallb{\gr r \cdot \plr{\V\,(-)\,A} \dotlr \plr{{-} \env\V \gr rA}}$
  \end{itemize}
\end{lemma}
\begin{proof}
  There are 6 cases to check.
  Throughout, we write $\Gamma$ as $\grP\gamma$ and $\Delta$ as $\grQ\delta$
  when convenient.
  \begin{description}
    \item[$I(\to)$]
      Let $\gr\Psi$ be the unique linear map out of the zero space.
      By definition, $\gr0 = \grQ\gr\Psi$.
      There are no variables to act upon.
    \item[$I(\gets)$]
      $\grQ\gr\Psi$ is an empty sum, so if $\grP = \grQ\gr\Psi$ then
      $\grP = \gr0$.
    \item[$\sep(\to)$]
      Let the given environments be $\rho : \grRl\theta \env\V \Gamma$ and
      $\sigma : \grRr\theta \env\V \Delta$, with $\grR = \grRl + \grRr$.
      Define $\gr\Psi \coloneqq [\rho.\gr\Psi, \sigma.\gr\Psi]$, using the
      coproduct structure of the concatenated vector space.
      We have $\grR = \grRl + \grRr =
      \grP\plr{\rho.\gr\Psi} + \grQ\plr{\sigma.\gr\Psi} =
      \plr{\grP, \grQ}\gr\Psi$.
      To act on variables, we are given
      $\grRprime = \plr{\grPprime, \grQprime}\gr\Psi$ and
      $\grPprime\gamma, \grQprime\delta \sqni A$.
      Without loss of generality, let us have $\grPprime\gamma \sqni A$ and
      $\grQprime = \gr0$.
      Thus, $\grRprime = \grPprime\plr{\rho.\gr\Psi}$, and we can act on the
      variable using $\rho$.
    \item[$\sep(\gets)$]
      Let the unnamed context be $\Theta$, also written $\grR\theta$.
      The linear map
      $\gr\Psi : \Ann^{\size\Gamma + \size\Delta} \to \Ann^{\size\Theta}$ splits
      into
      $\gr\Psi_{\gr l} : \Ann^{\size\Gamma} \to \Ann^{\size\Theta}
      \coloneqq \langle \id, 0 \rangle; \gr\Psi$ and
      $\gr\Psi_{\gr r} : \Ann^{\size\Delta} \to \Ann^{\size\Theta}
      \coloneqq \langle 0, \id \rangle; \gr\Psi$, using the product structure of
      the concatenated vector space.
      Let $\grRl \coloneqq \grP\gr\Psi_{\gr l}$ and
      $\grRr \coloneqq \grQ\gr\Psi_{\gr r}$, by definition satisfying the
      required equations.
      For the action on variables, let us consider the left environment.
      We are given $\grRprime = \grPprime\gr\Psi_{\gr l}$ and
      $\grPprime\gamma \sqni A$.
      From these, we get
      $\grRprime = \grPprime\gr\Psi_{\gr l} = \plr{\grPprime, \gr0}\gr\Psi$ and
      $\grPprime\gamma, \gr0\delta \sqni A$.
      We can therefore act using the original environment.
    \item[$\cdot(\to)$]
      Let $\grP$ and $\grPprime$ be such that $\grP = \gr r\grPprime$ and let
      $v : \V\,\grPprime\gamma\,A$.
      Let $\gr\Psi : \Ann \to \Ann^{\size\gamma}
      \coloneqq \gr r\gr' \mapsto \gr r\gr'\grPprime$.
      By definition and the previous assumption, we have $\grP = \gr r\gr\Psi$.
      When acting on a variable, we have $\grP\gr{''} = \gr r\gr'\gr\Psi$
      and $\gr r\gr'A \sqni A'$.
      The latter tells us that $A = A'$ and $\gr r\gr' = \gr1$.
      Thus, $\grP\gr{''} = \grPprime$.
      We therefore need a value of type $\V\,\grPprime\gamma\,A$, which we can
      take to be $v$.
    \item[$\cdot(\gets)$]
      Let us have an environment of type $\grP\gamma \env\V \gr rA$.
      We want to use its action on variables to yield a value.
      To do this, we let $\grPprime \coloneqq \gr1\gr\Psi$, and use this
      equation, together with the fact that we have a variable of type
      $\gr1A \sqni A$, to get a value of type $\V\,\grPprime\gamma\,A$.
      Furthermore, we derive $\grP = \gr r\gr\Psi = \gr r\grPprime$, as
      required.
  \end{description}
\end{proof}

We could, indeed, use these three clauses to define what an environment is.
However, I find them difficult to work with, as it is often easier to do
linear algebraic proofs separately from the rest of an environment.
For identity and composition, as we are about to see, the original definition
is easier to use because we can rely on the identity and composition of linear
maps.
Concretely, an inductive proof of identity would, for example, involve
constructing an environment of type
$\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
environments of types $\grP\gamma, \gr0\delta \env\V \grP\gamma$ and
$\gr0\gamma, \grQ\delta \env\V \grQ\delta$.
These are not identity environments, so we would have to strengthen the
induction hypothesis.

The requirements for identity and composition of environments look a bit like
the unit and lift of a Kleisli triple.

\begin{lemma}[Identity environment]
  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
\end{lemma}
\begin{proof}
  Let $\gr\Psi$ be the identity map, which clearly satisfies
  $\grP = \grP\gr\Psi$.
  When acting on a variable, the equation $\grPprime = \grQprime\gr\Psi$ means
  that $\grPprime = \grQprime$, so we want, from a variable of type
  $\grPprime\gamma \sqni A$, a value of type $\V\,\grPprime\gamma\,A$, which
  we can get from $\mathrm{vr}$.
\end{proof}

\begin{lemma}\label{thm:env-comp-lemma}
  Given an environment $\rho : \Gamma \env\U \Delta$ for which we have, for any
  $\grPprime$ and $\grQprime$ such that
  $\grPprime = \grQprime\plr{\rho.\gr\Psi}$, we have a function
  $\mathrm{lift}_\rho :
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$,
  we can map environments of type $\Delta \env\V \Theta$ into environments of
  type $\Gamma \env\W \Theta$.
\end{lemma}
\begin{proof}
  Let $\rho$ be as in the statement, and let $\sigma : \Delta \env\V \Theta$.
  For the environment we are constructing, let
  $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
  $\grP = \grQ\plr{\rho.\gr\Psi} =
  \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
  For the action on variables, we are given $\grPprime = \grRprime\gr\Psi$ with
  $\grRprime\theta \sqni A$.
  We can immediately apply the action of $\sigma$, giving us a value of type
  $\V\,\plr{\grRprime\plr{\sigma.\gr\Psi}}\,A$.
  We note that
  $\grPprime = \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
  apply $\mathrm{lift}_\rho$ to get the desired value.
\end{proof}

\begin{corollary}[Composition of environments]
  Given a function
  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
  \forall \grPprime, \grQprime.~\grPprime = \grQprime\plr{\rho.\gr\Psi} \to
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
  compose environments of types $\Gamma \env\U \Delta$ and
  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
\end{corollary}

\begin{example}
  We can derive the following instances of environment composition.
  \begin{itemize}
    \item If $\U = \V = \W = {\sqni}$, then $\mathrm{lift}$ is given by the
      action of the renaming $\rho$ on variables.
      This allows us to derive composition of renamings.
    \item More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
      the action of the environment $\rho$.
      This means that renamings post-compose with any other sort of environment.
    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
      syntactic traversal.
      For example, if $\U = {\sqni}$, we need the action of renaming on terms
      to show that a renaming followed by a substitution composes to a
      substitution.
      If $\U = {\vdash}$, then the action of substitution on terms gives us that
      substitutions compose.
    \item More generally, if $\V = {\vdash}$ and we have a semantics from
      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
      of terms.
  \end{itemize}
\end{example}

% Concatenation is difficult; save to after I've talked about renamings.

% Finally for this section, we give the conditions under which the
% context-forming operations (empty, concatenation, and singleton) have a
% functorial action with respect to $\V$-environments.
%
% \begin{lemma}
%   For any $\V$, there is an environment ${\cdot} \env\V {\cdot}$.
% \end{lemma}
% \begin{proof}
%   By \autoref{thm:construct-env}, it suffices to show $I\,{\cdot}$, which is
%   trivially true.
% \end{proof}

### 2021-07-27 tex/lnd.tex
[ writing ] environment properties

%This should stay fairly general, so it could be compiled to a DILL-style
%calculus as well as the one with usage annotations.
%Our framework compiles a ND system down to a sequent calculus.
%It isolates the building blocks of typing rules ($\vdash$, $\wedge$, $*$,
%$\Box$, $\cdot$) so that we can do more generic programming (e.g., usage
%checker).

The typing rules presented in the previous section are somewhat convoluted,
and also contain a lot of repeated patterns.
For example, nearly half of the rules include the premise
$\grR \leq \grP + \grQ$.
Also, the presence of usage annotations, which are often different in different
parts of a rule, means that we keep repeating the context.
Explicit contexts go against the style of natural deduction, which is based
around being parametric in the context, so that substitution is agnostic to the
details of typing rules.

In this section, I introduce the \emph{bunched connectives}.
These are inspired by bunched logic~\cite{oHP99}, and will not only be used for
stating the syntax, but will reappear in the semantics.
The idea is to generalise the space between premises from Gentzen's natural
deduction so as to allow for any linear manipulations of usage annotations.
Similar connectives, but with different interpretations, could be used to
define other linear-like type theories, like DILL, but here I will focus on the
usage annotation style.
The interpretations I use are defined in \autoref{fig:bunched}.

\begin{figure}
  \begin{align*}
    \dot1\,\grR &\coloneqq 1 \\
    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~ \plr{\grR \leq \grP + \grQ}
                       \times T\,\grP \times U\,\grQ \\
    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~ \plr{\grR \leq \gr r\grP}
                       \times T\,\grP
  \end{align*}
  \caption{The bunched connectives}
  \label{fig:bunched}
\end{figure}

The simplest connectives are those we've already seen for intuitionistic
systems --- $\dot1$, $\dottimes$, and $\dotto$.
The absence of premises is encoded by $\dot1$, while the space between premises
sharing a context is encoded by $\dottimes$.
When we interpret a rule as a constructor of an inductive definition, $\dotto$
interprets the horizontal line, reflecting the fact that the usage annotations
we start off with in the premises are those of the conclusion.
The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
rules for $\top$ and $\with$, respectively.

\begin{align*}
  \begin{prooftree}
    \infer0{\grR\Gamma \vdash \top}
  \end{prooftree}
  &\quad\rightsquigarrow\quad
  \begin{prooftree}
    \set{separation=0.75em}
    \hypo{\dot1}
    \infer1{\vdash \top}
  \end{prooftree}
  \\\\
  \begin{prooftree}
    \hypo{\grR\Gamma \vdash A}
    \hypo{\grR\Gamma \vdash B}
    \infer2{\grR\Gamma \vdash A \with B}
  \end{prooftree}
  &\quad\rightsquigarrow\quad
  \begin{prooftree}
    \set{separation=0.75em}
    \hypo{\vdash A}
    \hypo{\dottimes}
    \hypo{\vdash B}
    \infer3{\vdash A \with B}
  \end{prooftree}
\end{align*}

The rest of the bunched connectives are for linear decompositions of the usage
annotations.
The three basic left semimodule operators --- zero, addition, and left-scaling
--- each get a bunched connective --- $I^*$, $\sep$, and $\gr r \cdot {}$,
respectively.
The prototypical typing rules for each of these three connectives are the
introduction rules for $I$, $\otimes$, and $\oc_{\gr r}$, respectively.

\begin{align*}
  \begin{prooftree}
    \hypo{\grR \leq \gr0}
    \infer1{\grR\Gamma \vdash I}
  \end{prooftree}
  &\quad\rightsquigarrow\quad
  \begin{prooftree}
    \set{separation=0.75em}
    \hypo{I^*}
    \infer1{\vdash I}
  \end{prooftree}
  \\\\
  \begin{prooftree}
    \hypo{\grP\Gamma \vdash A}
    \hypo{\grQ\Gamma \vdash B}
    \hypo{\grR \leq \grP + \grQ}
    \infer3{\grR\Gamma \vdash A \otimes B}
  \end{prooftree}
  &\quad\rightsquigarrow\quad
  \begin{prooftree}
    \set{separation=0.75em}
    \hypo{\vdash A}
    \hypo{\sep}
    \hypo{\vdash B}
    \infer3{\vdash A \otimes B}
  \end{prooftree}
  \\\\
  \begin{prooftree}
    \hypo{\grP\Gamma \vdash A}
    \hypo{\grR \leq \gr r\grP}
    \infer2{\grR\Gamma \vdash \oc_{\gr r} A}
  \end{prooftree}
  &\quad\rightsquigarrow\quad
  \begin{prooftree}
    \set{separation=0.75em}
    \hypo{\gr r \cdot (\vdash A)}
    \infer1{\vdash \oc_{\gr r} A}
  \end{prooftree}
\end{align*}

The full system \name{} is stated in terms of bunched connectives in
\autoref{fig:lr-bunched}.
%I have not included a variable rule because, like with intuitionistic
%natural deduction,

These bunched connectives should not be confused for the linear connectives
they appear in the introduction rules of.
For example, it would make sense to define a bunched connective $\dotplus$,
defined analogously to $\dottimes$.
This $\dotplus$ could be used to rephrase the introduction rules for $\oplus$.
We then have maps both ways between $T \dottimes (U \dotplus V)$ and
$(T \dottimes U) \dotplus (T \dottimes V)$, reminiscent of bunched logic,
whereas linear logic only has a map from $(A \with B) \oplus (A \with C)$ to
$A \with (B \oplus C)$, and not a map the other way.
%For example, linearly we have
%$\oc_{\gr r}(A \with B) \simeq \oc_{\gr r} A \otimes \oc_{\gr r} B$, while the
%expressions $\gr r \cdot (A \dottimes B)$ and $\gr r \cdot A \sep \gr r \cdot B$
%are generally incomparable.
We will also see later that, while $(\dottimes, \dotto)$ gives a Cartesian
closed structure, $\sep$ is monoidally closed, with an internal hom $\wand$.
Looking at the interpretations, the connection with bunched logic makes sense.
Instead of the partial commutative monoid (representing heaps) found in
standard semantics of bunched logic, we have a left semimodule of usage
contexts, which we are similarly interested in splitting and sharing between
various subterms.

\begin{figure}
  \ebproofset{separation=0.75em}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\sqni A}
      \infer1[Var]{\vdash A}
    \end{prooftree}
  \end{displaymath}

  \begin{displaymath}
    \begin{matrix}
      \begin{prooftree}
        \hypo{I^*}
        \infer1[$I$-I]{\vdash I}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash I}
        \hypo{\sep}
        \hypo{\vdash C}
        \infer3[$I$-E]{\vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\vdash A}
        \hypo{\sep}
        \hypo{\vdash B}
        \infer3[$\otimes$-I]{\vdash A \otimes B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash A \otimes B}
        \hypo{\sep}
        \hypo{\gr1A, \gr1B \vdash C}
        \infer3[$\otimes$-E]{\vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\gr1A \vdash B}
        \infer1[$\multimap$-I]{\vdash A \multimap B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash A \multimap B}
        \hypo{\sep}
        \hypo{\vdash A}
        \infer3[$\multimap$-E]{\vdash B}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\dot1}
        \infer1[$\top$-I]{\vdash \top}
      \end{prooftree}
      &&
      \textrm{(no $\top$-E)}
      \\\\
      \begin{prooftree}
        \hypo{\vdash A}
        \hypo{\dottimes}
        \hypo{\vdash B}
        \infer3[$\with$-I]{\vdash A \with B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash A_0 \with A_1}
        \infer1[$\with$-E$_i$, for $i \in \{0,1\}$]{\vdash A_i}
      \end{prooftree}
      \\\\
      \textrm{(no $0$-I)}
      &&
      \begin{prooftree}
        \hypo{\vdash 0}
        \hypo{\sep}
        \hypo{\dot1}
        \infer3[$0$-E]{\vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\vdash A_i}
        \infer1[$\oplus$-I$_i$, for $i \in \{0,1\}$]{\vdash A_0 \oplus A_1}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash A \oplus B}
        \hypo{\sep}
        \hypo{(\gr1A \vdash C}
        \hypo{\dottimes}
        \hypo{\gr1B \vdash C)}
        \infer5[$\oplus$-E]{\vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\gr r \cdot (\vdash A)}
        \infer1[$\oc$-I]{\vdash \oc_{\gr r} A}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\vdash \oc_{\gr r} A}
        \hypo{\sep}
        \hypo{\gr rA \vdash C}
        \infer3[$\oc$-E]{\vdash C}
      \end{prooftree}
    \end{matrix}
  \end{displaymath}
  \ebproofset{separation=1.5em}
  \caption{\name{} stated using bunched connectives}
  \label{fig:lr-bunched}
\end{figure}

### 2021-07-27 tex/lrsub.tex
[ writing ] environment properties

With the notion of \emph{environment} decided, the proofs of admissibility of
renaming and substitution are structured in the same way as they are in
\autoref{sec:syntactic-kits}.
We start with a slightly modified definition of \AgdaRecord{Kit}.
While intuitionistic kits require arbitrary (right-)weakening, this assumption
is unrealistic in the linear setting.
However, we can expect weakening by $\gr0$-use variables.

\missingfigure{Kit}

### 2021-07-27 tex/terms.tex
[ writing ] environment properties

We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
terms.
Beside base types, the only type former we have is the function type constructor
\AgdaInductiveConstructor{\_`→\_}.
Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
(\AgdaDatatype{Ty}).
Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
embeds types into contexts.
Typed variables in a context are given by \AgdaRecord{\_∋\_}.
A variable in
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
is given by a path \AgdaField{idx} to a type in \AgdaBound{$\Gamma$}, together
with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.

\Var{}
\Term{}

### 2021-08-03 tex/terms.tex
[ writing ] mechanising contexts

@@
+We could mechanise Gentzen's definition of a natural deduction system directly,
+but this definition is quite complicated.
+In particular, if we want to give derivations an inductive definition, the use
+of the discharge mechanism means that we actually need an inductive-inductive
+type --- derivations, particularly those using $\to$-introduction, can involve
+references to assumptions within their subderivations.
+An inductive-inductive definition of derivations would complicate our programs
+and proofs about natural deduction derivations, so I choose an alternative
+representation.
+
+Indeed, most authors since Gentzen, whether mechanising their work or not,
+have opted to replace discharge of assumptions by explicit \emph{contexts} and
+a variable rule.
+Contexts can be justified as a way to keep track of undischarged assumptions.
+In particular, we only produce derivations in the presence of a known collection
+of \emph{free variables} specified by the context.
+In other words, derivations are \emph{indexed} over their free variables and
+their types.
+When using an assumption within a derivation, we must say which free variable
+it corresponds to.
+Free variables are introduced by \emph{variable-binding} rules, like
+$\to$-introduction.
+\autoref{fig:explicit-contexts} gives an example of the same derivation written
+in Gentzen's style and in the explicit context style.
+
+\begin{sidewaysfigure}
+  \centering
+  \begin{prooftree}
+    \hypo{[A \to A \to B]^f}
+    \hypo{[A]^x}
+    \infer2[$\to$-E]{A \to B}
+    \hypo{[A]^x}
+    \infer2[$\to$-E]{B}
+    \infer1[$\to$-I$^x$]{A \to B}
+    \infer1[$\to$-I$^f$]{\plr{A \to A \to B} \to \plr{A \to B}}
+  \end{prooftree}
+
+  \vspace{2em}
+
+  \begin{prooftree}
+    \infer0[var$^f$]{{\color{red}f : A \to A \to B, x : A} \vdash A \to A \to B}
+    \infer0[var$^x$]{{\color{red}f : A \to A \to B, x : A} \vdash A}
+    \infer2[$\to$-E]{{\color{red}f : A \to A \to B, x : A} \vdash A \to B}
+    \infer0[var$^x$]{{\color{red}f : A \to A \to B, x : A} \vdash A}
+    \infer2[$\to$-E]{{\color{red}f : A \to A \to B, x : A} \vdash B}
+    \infer1[$\to$-I$^x$]{{\color{red}f : A \to A \to B} \vdash A \to B}
+    \infer1[$\to$-I$^f$]{\vdash \plr{A \to A \to B} \to \plr{A \to B}}
+  \end{prooftree}
+  \caption{A proof in Gentzen's natural deduction syntax, and a proof using
+    explicit contexts (contexts coloured {\color{red}red})}
+  \label{fig:explicit-contexts}
+\end{sidewaysfigure}
+
+Explicit contexts can be seen as a mechanism for encoding a natural deduction
+system as a sequent calculus.
+However, the natural deduction character of the system is maintained by
+ensuring that the resultant sequent calculus is really an encoding of a
+natural deduction system.
+Concretely, this means that rules can only interact with the context in
+restricted ways:
+
+\begin{itemize}
+  \item There is a designated \emph{variable rule}, stating that any variable
+    in the context can serve as a derivation of its type.
+  \item Non-variable rules may only require subterms with \emph{extended}
+    contexts, i.e., subterms in which new variables have been bound.
+    Non-variable rules are parametric in the existing free variables.
+\end{itemize}
+
+Having chosen to use explicit contexts, the mechanisation must have a chosen
+representation of contexts as a data structure.
+While the notation in \autoref{fig:explicit-contexts} uses names $f$ and $x$
+for variables, I opt for a nameless representation.
+In a nameless representation, variables are identified by their position in
+the context, rather than by a name.
+The absence of names means that $\alpha$-equivalence is just on-the-nose
+equality, and also that we never have to reason about freshness of names.
+Agda does not have support for nominal techniques~\cite{GP02}, which may have
+made names a better option.
+
+Most mechanisations choose contexts to be an inductive list of types.
+However, I instead choose a functional, tree-shaped representation, as shown
+with the type \AgdaRecord{Ctx}.
+The type \AgdaDatatype{LTree} is the inductive type generated by leaves and
+nullary \& binary nodes, and serves as a generalised ``length'' of the context.
+The contents of the context --- the types --- are then stored in the functional
+vector \AgdaField{ty-ctx}, which is a mapping from leaves in \AgdaField{shape}
+to object language types \AgdaDatatype{Ty}.
+The advantages of the functional vector representation will not become clear
+until later chapters, where I make use of the ease of look-up and $\eta$-law
+of functions.
+However, I claim for now that there is little to no disadvantage in the
+functional vector representation --- in particular, we have no need for
+function extensionality principles because we never talk about equality of
+contexts.
+For example, instead of using an equality of contexts to coerce a term, we can
+use renaming.
+As for the tree shape, this makes context concatenation definitionally
+injective, so that in cases where multiple new variables are bound in a subterm
+(for example, $\otimes$-elimination), Agda's unification-based solving will
+be more able to infer which variables have just been bound.
+
+\Ctx{}
+
 We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed

### 2021-08-10 tex/lr.tex
[ new ] lambda-R basic type system

In this section, I introduce the syntax of the type theory \name{}, which makes
use of usage annotations.
I use this syntax to write some example programs, which will motivate the
denotational semantics explored in \cref{sec:wrel}.
For the rest of this thesis, \name{} will serve as both a prototypical
usage-constrained syntax and a target of semantic analyses.

The calculus \name{} is similar in spirit to intuitionistic linear logic (ILL).
The types of \name{}, listed in \cref{fig:lr-types}, are almost identical
to those of ILL, differing only in the exponential modality $\oc$
(read ``bang'').
In particular, I include distinguished tensor- and with-product types
($\otimes$, $\with$) and their units ($I$, $\top$), function types
($\multimap$), additive sum types and their unit ($\oplus$, $0$), and the
graded modality $\oc_{\gr r}$.
The idea of $\oc_{\gr r}$ is to internalise an annotation of $\gr r$ on a
variable in the context.
In this position, an assumption of type $\oc_{\gr r} A$ acts like an assumption
of type $A$ that is to be used according to $\gr r$ rather than the standard
$\gr1$.

\begin{figure}
  \begin{displaymath}
    A, B, C \Coloneqq I \mid A \otimes B \mid A \multimap B \mid \top
    \mid A \with B \mid 0 \mid A \oplus B \mid \oc_{\gr r} A
  \end{displaymath}
  \caption{The types of \name{}}
  \label{fig:lr-types}
\end{figure}

\begin{figure}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\gamma \ni x : A}
      \hypo{\grP \le \langle x \rvert}
      \infer2[Var]{\grP\gamma \vdash A}
    \end{prooftree}
  \end{displaymath}

  \begin{displaymath}
    \begin{matrix}
      \begin{prooftree}
        \hypo{\grP \le \gr0}
        \infer1[$I$-I]{\grP\gamma \vdash I}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash I}
        \hypo{\grQ\gamma \vdash C}
        \infer3[$I$-E]{\grR\gamma \vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash A}
        \hypo{\grQ\gamma \vdash B}
        \infer3[$\otimes$-I]{\grR\gamma \vdash A \otimes B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash A \otimes B}
        \hypo{\grQ\gamma, \gr1A, \gr1B \vdash C}
        \infer3[$\otimes$-E]{\grR\gamma \vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\grR\gamma, \gr1A \vdash B}
        \infer1[$\multimap$-I]{\grR\gamma \vdash A \multimap B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash A \multimap B}
        \hypo{\grQ\gamma \vdash A}
        \infer3[$\multimap$-E]{\grR\gamma \vdash B}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \infer0[$\top$-I]{\grR\gamma \vdash \top}
      \end{prooftree}
      &&
      \textrm{(no $\top$-E)}
      \\\\
      \begin{prooftree}
        \hypo{\grR\gamma \vdash A}
        \hypo{\grR\gamma \vdash B}
        \infer2[$\with$-I]{\grR\gamma \vdash A \with B}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR\gamma \vdash A_0 \with A_1}
        \infer1[$\with$-E$_i$, for $i \in \{0,1\}$]{\grR\gamma \vdash A_i}
      \end{prooftree}
      \\\\
      \textrm{(no $0$-I)}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash 0}
        \infer2[$0$-E]{\grR\gamma \vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\grR\gamma \vdash A_i}
        \infer1[$\oplus$-I$_i$, for $i \in \{0,1\}$]%
        {\grR\gamma \vdash A_0 \oplus A_1}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash A \oplus B}
        \hypo{\grQ\gamma, \gr1A \vdash C}
        \hypo{\grQ\gamma, \gr1B \vdash C}
        \infer4[$\oplus$-E]{\grR\gamma \vdash C}
      \end{prooftree}
      \\\\
      \begin{prooftree}
        \hypo{\grR \le \gr r\grP}
        \hypo{\grP\gamma \vdash A}
        \infer2[$\oc$-I]{\grR\gamma \vdash \oc_{\gr r} A}
      \end{prooftree}
      &&
      \begin{prooftree}
        \hypo{\grR \le \grP + \grQ}
        \hypo{\grP\gamma \vdash \oc_{\gr r} A}
        \hypo{\grQ\gamma, \gr rA \vdash C}
        \infer3[$\oc$-E]{\grR\gamma \vdash C}
      \end{prooftree}
    \end{matrix}
  \end{displaymath}
  \caption{\name{}}
  \label{fig:lr}
\end{figure}

The operational semantics of \name{} are standard.
Assuming type preservation under reduction, \name{} enjoys all of the standard
operational properties via a forgetful translation to simply typed
$\lambda$-calculus.

### 2021-08-10 tex/lenv.tex
[ new ] Env preservation lemmas

@@
 
+One of the primary test cases for environments is simultaneous substitution,
+which will look like the following rule.
+The admissibility of substitution will be by induction on the derivation of
+$\Delta \vdash A$, so we will need to be able to adapt any environment we are
+given to work with any possible context of new premises.
+In the simply typed case, the only change to the context we encountered was the
+binding of new variables.
+Now, with usage annotations, we furthermore have linear decompositions of the
+context, necessitating changes to the environment whenever usage annotations
+change.
+I will deal first with linear decompositions.
+
+\begin{displaymath}
+  \begin{prooftree}
+    \hypo{\Gamma \env{\vdash} \Delta}
+    \hypo{\Delta \vdash A}
+    \infer2[sub]{\Gamma \vdash A}
+  \end{prooftree}
+\end{displaymath}
+
+There are three kinds of linear decompositions we have to deal with: zero,
+addition, and scaling; corresponding to bunched connectives $I^*$, $\sep$, and
+$\gr r \cdot {}$, respectively.
+In each case, we have a simple preservation lemma, transforming an environment
+of type $\Gamma \env\V \Delta$ and a decomposition of $\Delta$ into a
+decomposition of $\Gamma$ and environments for all of the decomposed fragments
+of $\Gamma$ and $\Delta$.
+
+\begin{lemma}[environments preserve zero]\label{thm:lr-env-zero}
+  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  $\grQ \leq \gr 0$, we also have that $\grP \leq \gr 0$.
+\end{lemma}
+\begin{proof}
+  $\grP \leq \grQ\gr\Psi \leq \gr0\gr\Psi = \gr0$, by environment
+  compatibility and monotonicity and linearity of $\gr\Psi$.
+\end{proof}
+
+\begin{lemma}[environments preserve addition]\label{thm:lr-env-add}
+  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  $\grQ \leq \grQl + \grQr$ for some $\grQl$ and $\grQr$, we also have $\grPl$
+  and $\grPr$ such that $\grP \leq \grPl + \grPr$ and there are environments
+  of types $\grPl\gamma \env\V \grQl\delta$ and
+  $\grPr\gamma \env\V \grQr\delta$.
+\end{lemma}
+\begin{proof}
+  Let $\grPl \coloneqq \grQl\gr\Psi$ and $\grPr \coloneqq \grQr\gr\Psi$.
+  Then, $\grP \leq \grQ\gr\Psi \leq \plr{\grQl + \grQr}\gr\Psi =
+  \grQl\gr\Psi + \grQr\gr\Psi = \grPl + \grPr$, satisfying the first condition.
+  Because clearly $\grPl \leq \grQl\gr\Psi$ and $\grPr \leq \grQr\gr\Psi$,
+  \cref{thm:env-resize} on the original environment gives us the required
+  pair of new environments.
+\end{proof}
+
+\begin{lemma}[environments preserve scaling]\label{thm:lr-env-scale}
+  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  $\grQ \leq \gr r\grQprime$ for some $\grQprime$, we also have a $\grPprime$
+  such that $\grP \leq \gr r\grPprime$ and there is an environment of type
+  $\grPprime\gamma \env\V \grQprime\delta$.
+\end{lemma}
+\begin{proof}
+  Let $\grPprime \coloneqq \grQprime\gr\Psi$.
+  Then, $\grP \leq \grQ\gr\Psi \leq \plr{\gr r\grQprime}\gr\Psi =
+  \gr r\plr{\grQprime\gr\Psi} = \gr r\grPprime$, satisfying the first condition.
+  Because clearly $\grPprime \leq \grQprime\gr\Psi$,
+  \cref{thm:env-resize} on the original environment gives us the required
+  new environment.
+\end{proof}
+
+Finally, I will also take the opportunity to give the bind lemma, allowing
+environments to incorporate newly bound variables.
+In the intuitionistic case, the bind lemma had two requirements on $\V$: $\V$
+admits weakening and we can map variables into $\V$-values.
+With usage annotations, the former is unreasonable, but it turns out that we
+only need weakening by variables whose usage annotation is less than or equal
+to $\gr0$.
+The latter stays as-is, with the note that ``variable'' now means a
+usage-checked variable.
+
+\begin{lemma}[bind]\label{thm:lr-bind}
+  Given functions
+  ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
+  \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
+  $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
+  type $\Gamma \env\V \Delta$ into an environment of type
+  $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
+\end{lemma}
+\begin{proof}
+  Let $\grP\gamma \coloneqq \Gamma$, $\grQ\delta \coloneqq \Delta$, and
+  $\grR\theta \coloneqq \Theta$.
+  Let the new linear map $\gr\Psi\gr' : \Ann^{\size\Delta + \size\Theta} \to
+  \Ann^{\size\Gamma + \size\Theta}$ be $\gr\Psi \oplus \gr I$.
+  That is, in block matrix notation,
+  $\begin{pmatrix} \gr\Psi & \gr0 \\ \gr0 & \gr I \end{pmatrix}$.
+  Checking that this linear map fits, we have
+  $\begin{pmatrix}\grP & \grR\end{pmatrix}
+  \leq \begin{pmatrix}\grQ\gr\Psi & \grR\gr I\end{pmatrix}
+  = \begin{pmatrix}\grQ & \grR\end{pmatrix}\plr{\gr\Psi \oplus \gr I}$.
+  For the action on variables, we are given vectors $\grPprime$,
+  $\grR\gr'_\grP$, $\grQprime$, and $\grR\gr'_\grQ$ such that
+  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix} \leq
+  \begin{pmatrix} \grQprime & \grR\gr'_\grQ \end{pmatrix}
+  \plr{\gr\Psi \oplus \gr I}$ and we have a variable of type
+  $\grQprime\delta, \grR\gr'_\grQ\theta \sqni A$ for some type $A$.
+  The constraint on the new vectors reduces to $\grPprime \leq \grQprime\gr\Psi$
+  and $\grR\gr'_\grP \leq \grR\gr'_\grQ$.
+  From the variable we either have a variable $x$ in $\delta$ with
+  $\grQprime \leq \langle x \rvert$ and $\grR\gr'_\grQ \leq \gr0$, or a
+  variable $y$ in $\theta$ with $\grQprime \leq \gr0$ and
+  $\grR\gr'_\grQ \leq \langle y \rvert$.
+  In the former case, the action of the original environment on $x$ gives us a
+  $\V$-value in $\grPprime\gamma$, and the $\gr0$-weakening principle
+  $\swarrow^k$, noting that $\grR\gr'_\grP \leq \grR\gr'_\grQ \leq \gr0$, gives
+  us a $\V$-value in $\grPprime\gamma, \grR\gr'_\grP\theta$.
+  In the latter case, we have that
+  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix}
+  \leq \begin{pmatrix} \grQprime\gr\Psi & \grR\gr'_\grQ \end{pmatrix}
+  \leq \begin{pmatrix} \gr0\gr\Psi & \langle y \rvert \end{pmatrix}
+  = \begin{pmatrix} \gr0 & \langle y \rvert \end{pmatrix}
+  = \left\langle {\searrow}y \right\rvert$, so $y$ also serves as a
+  usage-checked variable in $\grPprime\gamma, \grR\gr'_\grP\theta$.
+  From this usage-checked variable, we get a $\V$-value in the same context
+  using $\mathrm{vr}$.
+\end{proof}
+
 The requirements for identity and composition of environments look a bit like

### 2021-08-10 tex/lenv.tex
[ edit ] lR environment introduction

@@
-\begin{remark}
+\begin{lemma}\label{thm:env-resize}
   Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
@@
   with the same linear map and action on variables.
-\end{remark}
+\end{lemma}
 \begin{proof}
@@
 % \begin{proof}
-%   By \autoref{thm:construct-env}, it suffices to show $I\,{\cdot}$, which is
+%   By \cref{thm:construct-env}, it suffices to show $I\,{\cdot}$, which is
 %   trivially true.

### 2021-08-10 tex/lnd.tex
[ edit ] lR environment introduction

@@
 
-The typing rules presented in the previous section are somewhat convoluted,
-and also contain a lot of repeated patterns.
+The typing rules presented in the previous section contain a lot of detail and
+repeated patterns.
 For example, nearly half of the rules include the premise
@@
 
-In this section, I introduce the \emph{bunched connectives}.
+To encapsulate the repeated patterns and facilitate an implicit context style,
+I introduce the \emph{bunched connectives}.
 These are inspired by bunched logic~\cite{oHP99}, and will not only be used for
@@
 deduction so as to allow for any linear manipulations of usage annotations.
+Among other things, this generalisation will allow us to distinguish between
+$\with$-introduction and $\otimes$-introduction by a choice of connective:
+either \emph{sharing} or \emph{separating} conjunction.
 Similar connectives, but with different interpretations, could be used to
@@
 usage annotation style.
-The interpretations I use are defined in \autoref{fig:bunched}.
+The interpretations I use are defined in \cref{fig:bunched}.
 
@@
 The full system \name{} is stated in terms of bunched connectives in
-\autoref{fig:lr-bunched}.
+\cref{fig:lr-bunched}.
 %I have not included a variable rule because, like with intuitionistic
@@
 %are generally incomparable.
-We will also see later that, while $(\dottimes, \dotto)$ gives a Cartesian
+We will also see later\todo{Forward reference} that, while
+$(\dottimes, \dotto)$ gives a Cartesian
 closed structure, $\sep$ is monoidally closed, with an internal hom $\wand$.
+Indeed, the Cartesian closed structure is sufficient to give us the interaction
+between $\dotplus$ and $\dottimes$, by the fact that $A \dottimes {-}$ is a
+left adjoint, and therefore preserves colimits.
 Looking at the interpretations, the connection with bunched logic makes sense.

### 2021-08-10 tex/lrkits.tex
[ edit ] lR environment introduction

@@
-As we discussed in \autoref{sec:kits}, simultaneous substitution gives a
+\paragraph{New explanation}
+Recalling from \cref{sec:kits}, we have the following definition of
+\emph{environments} for simple types.
+
+\begin{definition}[Simple environment]
+  A $\V$-\emph{environment} between simply typed contexts $\Gamma$ and $\Delta$
+  is a function, polymorphic in type $A$, from variables of type $A$ in
+  $\Delta$ to inhabitants of $\V\,\Gamma\,A$.
+  We write the type of such environments as $\Gamma \env\V \Delta$.
+\end{definition}
+
+\begin{definition}[Simple recursive environment]
+  A \emph{recursive $\V$-environment} between simply typed contexts $\Gamma$ and
+  $\Delta$ is defined by cases on the shape of $\Delta$ (where
+  $\Gamma \env\V_R \Delta$ is the notation for the type of recursive
+  environments for given $\V$, $\Gamma$, and $\Delta$):
+  \begin{itemize}
+    \item If $\Delta$ is empty, then there is one environment.
+    \item If $\Delta$ is a concatenation $\Delta_l, \Delta_r$, then an
+      environment is a pair of environments of types
+      $\Gamma \env\V_R \Delta_l$ and $\Gamma \env\V_R \Delta_r$.
+    \item If $\Delta$ is a singleton $A$, then an environment is a value of
+      type $\V\,\Gamma\,A$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Usage-annotated recursive environment]
+  A \emph{recursive $\V$-environment} between annotated contexts $\Gamma$ and
+  $\Delta$ is defined by cases on the shape of $\Delta$ (where
+  $\Gamma = \grP\gamma$ and $\Gamma \env\V_R \Delta$ is the notation for the
+  type of recursive environments for given $\V$, $\Gamma$, and $\Delta$):
+  \begin{itemize}
+    \item If $\Delta$ is empty, then an environment exists if $\grP = \gr0$.
+    \item If $\Delta$ is a concatenation $\Delta_l, \Delta_r$, then an
+      environment is a choice of usage vectors $\grPl$ and $\grPr$ such that
+      $\grP = \grPl + \grPr$ and we have a pair of environments of types
+      $\grPl\gamma \env\V_R \Delta_l$ and $\grPr\gamma \env\V_R \Delta_r$.
+    \item If $\Delta$ is a singleton $\gr rA$, then an environment is a choice
+      of a usage vector $\grPprime$ such that $\grP = \gr r\grPprime$ and we
+      have a value of type $\V\,\grPprime\gamma\,A$.
+  \end{itemize}
+\end{definition}
+
+From this definition, we can recover a functional-style definition by
+separating choices of usage vectors from the provision of $\V$-values.
+In particular, the only choices of usage vectors that are essential are the
+$\grPprime$s in the singleton case, with the choices in the concatenation case
+being determined as scalings and sums of these $\grPprime$s.
+I let $\gr\Psi$ collect up these $\size\Delta$-many choices of
+$\size\Gamma$-length usage vectors and note that the constraint on $\gr\Psi$
+generated by all the scaling and summing is
+$\grP = \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
+
+\begin{definition}[Usage-annotated environment (tentative)]
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a matrix $\gr\Psi : \Ann^{\size\Delta \times \size\Gamma}$ such that
+  $\grP = \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$ and for each
+  $\plr{x : A} \in \delta$ we have a value of type $\V\,\gr\Psi_x\gamma\,A$.
+\end{definition}
+
+We may note, further, that the constraint
+$\grP = \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$ can be stated as the
+vector-matrix multiplication $\grP = \grQ\gr\Psi$.
+Using the same operation, we have that $\gr\Psi_x = \langle x \rvert\gr\Psi$.
+Because $\langle x \rvert$ is exactly the $\grQprime$ such that
+$\plr{x : A} \sqin \grQprime\delta$, we can rephrase the function producing
+$\V$-values as: for each $A$, $\grPprime$, and $\grQprime$ such that
+$\grPprime = \grQprime\gr\Psi$, a function from $\grQprime\delta \sqni A$ to
+$\V\,\grPprime\gamma\,A$.
+Finally, I choose to switch from matrices and matrix multiplication to
+linear maps and their actions, which are easier to work with.
+All of these changes yield my primary definition of an environment for
+usage-annotated calculi.
+
+\begin{definition}[Usage-annotated environment]
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
+  postfix) such that $\grP = \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
+  $\grQprime$ such that $\grPprime = \grQprime\gr\Psi$, a function from
+  $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
+\end{definition}
+
+\paragraph{Old explanation}
+As we discussed in \cref{sec:kits}, simultaneous substitution gives a
 notion of derivability of one context from another, while simultaneous renaming
@@
 
-Taking a lead from \autoref{sec:kits}, we deduce a definition of
+Taking a lead from \cref{sec:kits}, we deduce a definition of
 \emph{environment} by replacing the $\vdash$ in the definition of simultaneous

### 2021-08-10 agda/LRKits.lagda.tex
[ new ] start on lR substitution

\begin{code}
{-# OPTIONS --safe --without-K --postfix-projections --prop #-}
open import Algebra.Po
open import Level

module LRKits (poSemiring : PoSemiring 0ℓ 0ℓ 0ℓ) where

  open PoSemiring poSemiring renaming (Carrier to Ann)

  open import Algebra.Relational renaming (_↘_↙_ to _↘-_-↙_)
  open import Data.LTree
  open import Data.LTree.Vector
  open import Data.Product
  open import Data.Wrap
  open import Function using (id; _∘_; Equivalence)
  open import Relation.Binary.PropositionalEquality as ≡ using (_≡_)
  open import Relation.Nary
  open import Relation.Unary.Bunched

  data Ty : Set where
    ι : Ty
    `! : (r : Ann) (A : Ty) → Ty

  open import Generic.Linear.Operations rawPoSemiring
  open import Generic.Linear.Algebra poSemiring
  open import Generic.Linear.Syntax Ty Ann
  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
  open import Generic.Linear.Variable Ty rawPoSemiring
  open import Generic.Linear.Environment Ty poSemiring
  open import Generic.Linear.Environment.Properties Ty poSemiring
  open import Generic.Linear.Environment.Categorical Ty poSemiring
\end{code}

%<*Tm>
\begin{code}
  Bind : Ctx → (Ctx → Set) → (Ctx → Set)
  Bind Δ T Γ = T (Γ ++ᶜ Δ)

  data _⊢_ : Ctx → Ty → Set where
    var : ∀[ _∋_ ⇒ _⊢_ ]
    !-I : ∀ {r A} → ∀[ r ·ᶜ (_⊢ A) ⇒ _⊢ `! r A ]
    !-E : ∀ {r A C} → ∀[ (_⊢ `! r A) ✴ᶜ Bind [ r · A ]ᶜ (_⊢ C) ⇒ _⊢ C ]
\end{code}
%</Tm>

\begin{code}
  module _ {𝓥 : Ctx → Ty → Set} where

    env-+ : ∀ {s t γ δ P Qₗ Qᵣ} →
      let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ Qₗ +* Qᵣ ] →
      (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
        ↘- P ≤[_+*_] -↙
      (λ Pᵣ → [ 𝓥 ] ctx Pᵣ γ ⇒ᵉ Δᵣ)
    env-+ (ρ , sp) =
      let sumsₗ ↘, sp′ ,↙ sumsᵣ = ρ .asLinRel .linRel .rel-+ₘ (sp , ρ .sums) in
      record { [_]_⇒ᵉ_ ρ; sums = sumsₗ }
        ↘, sp′ ,↙
      record { [_]_⇒ᵉ_ ρ; sums = sumsᵣ }

    env-* : ∀ {s t γ δ P Q′ r} →
      let Γ = ctx {s} P γ; Δ′ = ctx {t} Q′ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ r *ₗ Q′ ] →
      P ≤[ r *ₗ_] ◇ (λ P′ → [ 𝓥 ] ctx P′ γ ⇒ᵉ Δ′)
    env-* (ρ , sp) =
      let sp′ , sums′ = ρ .asLinRel .linRel .rel-*ₘ (sp , ρ .sums) in
      sp′ , record { [_]_⇒ᵉ_ ρ; sums = sums′ }
\end{code}

%<*Kit>
\begin{code}
  Weakening : ∀ {v} {A : Set} → (Ctx → A → Set v) → Set v
  Weakening 𝓥 =
    ∀ {s t P} {γ : Vector Ty (s <+> t)} → P ∘ ↘ ≤* 0* →
    ∀[ 𝓥 (ctx (P ∘ ↙) (γ ∘ ↙)) ⇒ 𝓥 (ctx P γ) ]

  record Kit (𝓥 : Ctx → Ty → Set) : Set where
    constructor kit
    field
      ↙ᵏ : Weakening 𝓥
      vr : ∀[ _∋_ ⇒ 𝓥 ]
      tm : ∀[ 𝓥 ⇒ _⊢_ ]
\end{code}
%</Kit>

\begin{code}
    open Equivalence
\end{code}

%<*bindEnv>
\begin{code}
    bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
    bindEnv ρ .M = [ [ ρ .M │ 0ᴹ ] ─ [ 0ᴹ │ 1ᴹ ] ]
    bindEnv ρ .asLinRel = ρ .asLinRel ⊕AsLinRel idAsLinRel
    bindEnv ρ .sums = ρ .sums , ≤*-refl
    bindEnv ρ .lookup (ll , rr) (lvar (↙ i) q b) =
      let bl , br = un++ₙ b in
      ↙ᵏ (≤*-trans rr br) (ρ .lookup ll (lvar i q bl))
    bindEnv ρ .lookup (ll , rr) (lvar (↘ i) q b) =
      let bl , br = un++ₙ b in
      let bl′ = 0*→≤* (ρ .asLinRel .linRel .rel-0ₘ (≤*→0* bl , ll)) in
      vr (lvar (↘ i) q (bl′ ++ₙ ≤*-trans rr br))
\end{code}
%</bindEnv>

%<*trav>
\begin{code}
    trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (ρ .lookup (ρ .sums) x)
    trav ρ (!-I (⟨ sp* ⟩· M)) =
      let sp*′ , ρ′ = env-* (ρ , sp*) in
      !-I (⟨ sp*′ ⟩· (trav ρ′ M))
    trav ρ (!-E (M ✴⟨ sp+ ⟩ N)) =
      let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
      !-E (trav ρₗ M ✴⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
\end{code}
%</trav>

%<*dlv>
\begin{code}
  ↙ᵛ : Weakening _∋_
  ↙ᵛ Q0 v .idx = ↙ (v .idx)
  ↙ᵛ Q0 v .tyq = v .tyq
  ↙ᵛ Q0 v .basis = v .basis ++ₙ Q0
\end{code}
%</dlv>

\begin{code}
  open Kit
\end{code}

%<*var-kit>
\begin{code}
  ∋-kit : Kit _∋_
  ∋-kit = kit ↙ᵛ id var

  ren = trav ∋-kit
\end{code}
%</var-kit>

\begin{code}
  instance
    identityEnv-⊢ : IdentityEnv _⊢_
    identityEnv-⊢ .pure = var
\end{code}

%<*dlv-env>
\begin{code}
  ↙^Env : {𝓥 : Ctx → Ty → Set} → Weakening 𝓥 → Weakening [ 𝓥 ]_⇒ᵉ_
  ↙^Env ↙^𝓥 Q0 ρ .M = [ ρ .M │ 0ᴹ ]
  ↙^Env ↙^𝓥 Q0 ρ .asLinRel = [ ρ .asLinRel │ 0AsLinRel ]AsLinRel
  ↙^Env ↙^𝓥 Q0 ρ .sums = ρ .sums , ≤*→0* Q0
  ↙^Env ↙^𝓥 Q0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)

  ↙ᵛ-env :
    ∀ {s t P} {γ : Vector Ty (s <+> t)} → P ∘ ↘ ≤* 0* →
    [ _∋_ ] ctx P γ ⇒ᵉ ctx (P ∘ ↙) (γ ∘ ↙)
  ↙ᵛ-env Q0 = ↙^Env ↙ᵛ Q0 id^Env
\end{code}
%</dlv-env>

%<*sub>
\begin{code}
  ⊢-kit : Kit _⊢_
  ⊢-kit = kit (λ Q0 → ren (↙ᵛ-env Q0)) var id

  sub = trav ⊢-kit
\end{code}
%</sub>

\begin{code}
  idˢ : ∀ {Γ} → [ _⊢_ ] Γ ⇒ᵉ Γ
  idˢ = id^Env

  psh-⊢ : IsPresheaf _⊢_
  psh-⊢ QP = ren (subuse^Env QP)

  mutual
    open With-psh^𝓥 psh-⊢
\end{code}

%<*subSingle>
\begin{code}
    sub[-] : ∀ {s P Q R γ r A B} → R ≤[ Q +*_] ◇ _≤[ r *ₗ P ] →
      ctx {s} P γ ⊢ A → ctx Q γ ++ᶜ [ r · A ]ᶜ ⊢ B → ctx R γ ⊢ B
    sub[-] {Q = Q} {R} {γ} {r} {A} (sp+ , sp*) t = sub σ
      where
      σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r · A ]ᶜ
      σ = ++ᵉ (idˢ ✴⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩· t))
\end{code}
%</subSingle>

### 2021-08-10 tex/lrsub.tex
[ new ] start on lR substitution

@@
-With the notion of \emph{environment} decided, the proofs of admissibility of
-renaming and substitution are structured in the same way as they are in
-\autoref{sec:syntactic-kits}.
+\def\LRKits{../agda/latex/LRKits.tex}
+
+From \cref{sec:lrkits,sec:lenv}, we have fixed a notion of \emph{environment}
+and proved that it is preserved under all syntax-forming operations.
+In this section, I put these properties together in the style of
+\cref{sec:syntactic-kits} to prove the admissibility of simultaneous renaming,
+simultaneous substitution, and then single substitution.
+
 We start with a slightly modified definition of \AgdaRecord{Kit}.
-While intuitionistic kits require arbitrary (right-)weakening, this assumption
-is unrealistic in the linear setting.
-However, we can expect weakening by $\gr0$-use variables.
+We saw in \cref{thm:lr-bind} that in the usage-annotated context, we replace
+general weakening of $\V$-values by weakening by $\gr0$-use variables.
+Meanwhile, the function $\mathrm{vr}$, also seen in \cref{thm:lr-bind}, maps
+usage-checked variables to $\V$-values, and the function $\mathrm{tm}$, used
+to coerce $V$-values yielded by the environment into terms, stays the same.
+I state weakening in a slightly different way than previously, so as to help
+unification against a known result type.
+The type \AgdaFunction{Weakening}\AgdaSpace{}\AgdaBound{$\V$} can be read as
+saying that, for any context $\grP\gamma$ of shape $s + t$, if the right of
+$\grP$ is below $\gr0$, then a value in the left part of $\grP\gamma$ weakens
+to a value in the whole of $\grP\gamma$.
 
-\missingfigure{Kit}
+\ExecuteMetaData[\LRKits]{Kit}
+
+To demonstrate the important points succinctly, I cut \name{} down to just the
+$\oc\gr r$-fragment.
+The introduction rule and pattern-matching eliminator feature scaling, addition,
+and variable binding, missing out only on sharing (which is trivial) and zero
+(which is simpler than, and analogous to, addition).
+The resulting type of well typed terms is below.
+
+\ExecuteMetaData[\LRKits]{Tm}
+
+Given a \AgdaRecord{Kit}, \cref{thm:lr-bind} looks like the following.
+The \AgdaField{lookup} clauses still contain essentially the same structure as
+in the intuitionistic case: discriminating on whether the variable is old or
+new, using the given environment \AgdaBound{$\rho$} and weakening on the old
+variables, and using \AgdaField{vr} to repackage new variables.
+I will not explain any of the algebraic manipulations here; see
+\cref{thm:lr-bind}.
+
+\ExecuteMetaData[\LRKits]{bindEnv}
+
+Given \AgdaFunction{bindEnv}, along with \AgdaFunction{env-+} and
+\AgdaFunction{env-*}, standing for \cref{thm:lr-env-add} and
+\cref{thm:lr-env-scale}, respectively, we can reproduce the syntactic traversal
+\AgdaFunction{trav}.
+With all these lemmas in place, we can see how writing \AgdaFunction{trav}
+becomes routine --- when processing a rule, we work our way up through the
+premise connectives, applying \AgdaFunction{env-*} wherever we see a
+\AgdaFunction{$\cdot^c$}, \AgdaFunction{env-+} wherever we see a
+\AgdaFunction{$*^c$}, and \AgdaFunction{bindEnv} wherever we see a
+\AgdaFunction{Bind}.
+We when use whatever environments (with names beginning with
+\AgdaBound{$\rho$}) and whatever usage vector splitting facts (with names
+beginning with \AgdaBound{sp}) come out of this process to recursively
+traverse the subterms and recombine the results.
+
+\ExecuteMetaData[\LRKits]{trav}

### 2021-08-10 tex/terms.tex
[ minor ] references

@@
 $\to$-introduction.
-\autoref{fig:explicit-contexts} gives an example of the same derivation written
+\cref{fig:explicit-contexts} gives an example of the same derivation written
 in Gentzen's style and in the explicit context style.
@@
 representation of contexts as a data structure.
-While the notation in \autoref{fig:explicit-contexts} uses names $f$ and $x$
+While the notation in \cref{fig:explicit-contexts} uses names $f$ and $x$
 for variables, I opt for a nameless representation.
@@
 The advantages of the functional vector representation will not become clear
-until later chapters, where I make use of the ease of look-up and $\eta$-law
-of functions.
+until later chapters\todo{forward reference}, where I make use of the ease of
+look-up and $\eta$-law of functions.
 However, I claim for now that there is little to no disadvantage in the
@@
 
+\missingfigure{LTree}
 \Ctx{}

### 2021-08-17 agda/LRKits.lagda.tex
[ write ] section on λR substitution

@@
   ↙ᵛ : Weakening _∋_
-  ↙ᵛ Q0 v .idx = ↙ (v .idx)
-  ↙ᵛ Q0 v .tyq = v .tyq
-  ↙ᵛ Q0 v .basis = v .basis ++ₙ Q0
+  ↙ᵛ R0 v .idx = ↙ (v .idx)
+  ↙ᵛ R0 v .tyq = v .tyq
+  ↙ᵛ R0 v .basis = v .basis ++ₙ R0
 \end{code}
@@
   ↙^Env : {𝓥 : Ctx → Ty → Set} → Weakening 𝓥 → Weakening [ 𝓥 ]_⇒ᵉ_
-  ↙^Env ↙^𝓥 Q0 ρ .M = [ ρ .M │ 0ᴹ ]
-  ↙^Env ↙^𝓥 Q0 ρ .asLinRel = [ ρ .asLinRel │ 0AsLinRel ]AsLinRel
-  ↙^Env ↙^𝓥 Q0 ρ .sums = ρ .sums , ≤*→0* Q0
-  ↙^Env ↙^𝓥 Q0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)
+  ↙^Env ↙^𝓥 R0 ρ .M = [ ρ .M │ 0ᴹ ]
+  ↙^Env ↙^𝓥 R0 ρ .asLinRel = [ ρ .asLinRel │ 0AsLinRel ]AsLinRel
+  ↙^Env ↙^𝓥 R0 ρ .sums = ρ .sums , ≤*→0* R0
+  ↙^Env ↙^𝓥 R0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)
 
@@
     [ _∋_ ] ctx P γ ⇒ᵉ ctx (P ∘ ↙) (γ ∘ ↙)
-  ↙ᵛ-env Q0 = ↙^Env ↙ᵛ Q0 id^Env
+  ↙ᵛ-env R0 = ↙^Env ↙ᵛ R0 id^Env
 \end{code}
@@
   ⊢-kit : Kit _⊢_
-  ⊢-kit = kit (λ Q0 → ren (↙ᵛ-env Q0)) var id
+  ⊢-kit = kit (λ R0 → ren (↙ᵛ-env R0)) var id
 
@@
       σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r · A ]ᶜ
-      σ = ++ᵉ (idˢ ✴⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩· t))
+      σ = ++ᵉ (id^Env ✴⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩· t))
 \end{code}

### 2021-08-17 tex/lrsub.tex
[ write ] section on λR substitution

@@
 
-From \cref{sec:lrkits,sec:lenv}, we have fixed a notion of \emph{environment}
-and proved that it is preserved under all syntax-forming operations.
-In this section, I put these properties together in the style of
-\cref{sec:syntactic-kits} to prove the admissibility of simultaneous renaming,
-simultaneous substitution, and then single substitution.
+In this section, I show that, using the notion of \emph{environment} derived in
+\cref{sec:lrkits}, we can replicate the Agda proofs from
+\cref{sec:syntactic-kits} in the usage-aware setting.
+From \cref{sec:lenv}, we know that environments are preserved under all
+syntax-forming operations: zero, addition, scaling, and binding.
+What is left is to show how these properties are deployed, and also how to
+go on and prove the admissibility of simultaneous renaming, simultaneous
+substitution, and then single substitution.
+
+There are a few notational changes necessary in the Agda code.
+Usage vectors, elsewhere called $\grP$, $\grQ$, and $\grR$ are rendered as
+\AgdaBound{P}, \AgdaBound{Q}, and \AgdaBound{R}, respectively.
+Environments, elsewhere notated $\Gamma \env\V \Delta$, are rendered as
+\AgdaRecord{[}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaRecord{]}%
+\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\Rightarrow^e$}%
+\AgdaSpace{}\AgdaBound{$\Delta$}.
 
@@
 \ExecuteMetaData[\LRKits]{trav}
+
+Instantiating the generic syntactic traversal \AgdaFunction{trav} to renaming
+looks just like it did in the intuitionistic case.
+I have consistently replaced intuitionistic variables by linear variables, so
+\AgdaFunction{id} and \AgdaInductiveConstructor{var} still work to embed
+variables into variables and terms, respectively.
+Weakening for variables \AgdaFunction{$\swarrow^v$} (not pictured) has been
+updated to note that, for $\grP \leq \bra x$ and $\grR \leq \gr0$, we also have
+$\begin{pmatrix} \grP & \grR \end{pmatrix} \leq \bra{{\swarrow}x}$.
+
+\ExecuteMetaData[\LRKits]{var-kit}
+
+In the intuitionistic case, environments were just functions, so we passed the
+variable weakening function \AgdaFunction{$\swarrow^v$} to the function
+\AgdaFunction{ren} to yield a term weakening function.
+However, a usage-aware environment is a function packed together with some
+extra data.
+As such, we must make an environment version of \AgdaFunction{$\swarrow^v$}.
+I start with a general lemma \AgdaFunction{$\swarrow$\^{}Env}, stating that if
+$\V$ supports weakening, then so do $\V$-environments (in their domain
+context).
+This lemma then specialises to variables, with the identity renaming
+\AgdaFunction{id\^{}Env} on the left part of the context and the proof
+\AgdaBound{R0} that the right part of the context is below $\gr0$ combining
+to give the desired weakening environment.
+
+\ExecuteMetaData[\LRKits]{dlv-env}
+
+This is what we need to instantiate \AgdaFunction{trav} for substitution.
+
+\ExecuteMetaData[\LRKits]{sub}
+
+Finally, the simultaneous substitution \AgdaFunction{sub} specialises to
+single substitution \AgdaFunction{sub[-]}.
+Single substitution is stated as an admissible rule below.
+To substitute in for $\gr r$-many $A$ in the second term, we need to derive
+one $A$ with usages $\grP$, and then assert that the result can handle the
+usages of the original term $\grQ$, plus $\gr r$-many copies of $\grP$.
+
+\[
+  \begin{prooftree}
+    \hypo{\grR \leq \grQ + \gr r\grP}
+    \hypo{\grP\gamma \vdash A}
+    \hypo{\grQ\gamma, \gr rA \vdash B}
+    \infer3[singleSub]{\grR\gamma \vdash B}
+  \end{prooftree}
+\]
+
+The proof strategy for producing the substitution \AgdaFunction{$\sigma$} is
+to proceed structurally on the codomain context $\grQ\gamma, \gr rA$ using
+\cref{thm:construct-env}, applying the identity substitution
+\AgdaFunction{id\^{}Env} on the $\gamma$ half, and dropping the term
+\AgdaBound{t} in place of the variable we are substituting for.
+
+\ExecuteMetaData[\LRKits]{subSingle}

### 2021-08-17 tex/lenv.tex
[ write ] redraft bits on environments

@@
   Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
-  and a $\grQprime$ such that $\grPprime = \grQprime\plr{\rho.\gr\Psi}$,
+  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
   there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
@@
   The only part of the definition of an environment dependent on $\grP$ or
-  $\grQ$ is the constraint $\grP = \grQ\gr\Psi$, which we are able to replace
-  for $\grPprime$ and $\grQprime$.
+  $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
+  replace for $\grPprime$ and $\grQprime$.
 \end{proof}
@@
   We can define all of the following equivalences for any values of the free
-  variables.
+  variables, assuming that $\V$ respects subusaging (i.e.,
+  $\grPprime \leq \grP \to
+  \forallb{\V\,\grP\gamma \dotto \V\,\grPprime\gamma}$).
   \begin{itemize}
     \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
-    \item $\forallb{\plr{{-} \env\V \Gamma} \sep \plr{{-} \env\V \Delta}
-      \dotlr \plr{{-} \env\V \Gamma, \Delta}}$
+    \item $\forallb{\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
+      \dotlr \plr{{-} \env\V \Delta_l, \Delta_r}}$
     \item
@@
       Let $\gr\Psi$ be the unique linear map out of the zero space.
-      By definition, $\gr0 = \grQ\gr\Psi$.
+      By assumption and definition, $\grP \leq \gr0 = \grQ\gr\Psi$.
       There are no variables to act upon.
     \item[$I(\gets)$]
-      $\grQ\gr\Psi$ is an empty sum, so if $\grP = \grQ\gr\Psi$ then
-      $\grP = \gr0$.
+      $\grQ\gr\Psi$ is an empty sum, so if $\grP \leq \grQ\gr\Psi$ then
+      $\grP \leq \gr0$.
     \item[$\sep(\to)$]
-      Let the given environments be $\rho : \grRl\theta \env\V \Gamma$ and
-      $\sigma : \grRr\theta \env\V \Delta$, with $\grR = \grRl + \grRr$.
-      Define $\gr\Psi \coloneqq [\rho.\gr\Psi, \sigma.\gr\Psi]$, using the
+      Let the given environments be $\rho_l : \grPl\gamma \env\V \grQl\delta$
+      and $\rho_r : \grPr\gamma \env\V \grQr\delta$, with
+      $\grP \leq \grPl + \grPr$.
+      Define $\gr\Psi \coloneqq [\rho_l.\gr\Psi, \rho_r.\gr\Psi]$, using the
       coproduct structure of the concatenated vector space.
-      We have $\grR = \grRl + \grRr =
-      \grP\plr{\rho.\gr\Psi} + \grQ\plr{\sigma.\gr\Psi} =
-      \plr{\grP, \grQ}\gr\Psi$.
-      To act on variables, we are given
-      $\grRprime = \plr{\grPprime, \grQprime}\gr\Psi$ and
-      $\grPprime\gamma, \grQprime\delta \sqni A$.
-      Without loss of generality, let us have $\grPprime\gamma \sqni A$ and
-      $\grQprime = \gr0$.
-      Thus, $\grRprime = \grPprime\plr{\rho.\gr\Psi}$, and we can act on the
-      variable using $\rho$.
+      We have $\grP \leq \grPl + \grPr \leq
+      \grQl\plr{\rho_l.\gr\Psi} + \grQr\plr{\rho_r.\gr\Psi} =
+      \begin{pmatrix} \grQl & \grQr \end{pmatrix}\gr\Psi$.
+      To act on variables, we are given $\grPprime \leq
+      \begin{pmatrix} \gr{\grQ'_l} & \gr{\grQ'_r} \end{pmatrix}\gr\Psi$ and
+      $\gr{\grQ'_l}\delta_l, \gr{\grQ'_r}\delta_r \sqni A$.
+      Without loss of generality, let us have $\gr{\grQ'_l}\delta_l \sqni A$
+      and $\gr{\grQ'_r} \leq \gr0$.
+      Thus, $\grPprime \leq
+      \gr{\grQ'_l}\plr{\rho_l.\gr\Psi} + \gr{\grQ'_r}\plr{\rho_r.\gr\Psi} \leq
+      \gr{\grQ'_l}\plr{\rho_l.\gr\Psi}$,
+      and we can act on the variable using $\rho_l$.
     \item[$\sep(\gets)$]
-      Let the unnamed context be $\Theta$, also written $\grR\theta$.
+      Let the unnamed context be $\Gamma$, also written $\grP\gamma$.
       The linear map
-      $\gr\Psi : \Ann^{\size\Gamma + \size\Delta} \to \Ann^{\size\Theta}$ splits
-      into
-      $\gr\Psi_{\gr l} : \Ann^{\size\Gamma} \to \Ann^{\size\Theta}
-      \coloneqq \langle \id, 0 \rangle; \gr\Psi$ and
-      $\gr\Psi_{\gr r} : \Ann^{\size\Delta} \to \Ann^{\size\Theta}
-      \coloneqq \langle 0, \id \rangle; \gr\Psi$, using the product structure of
+      $\gr\Psi : \Ann^{\size{\Delta_l} + \size{\Delta_r}} \to \Ann^{\size\Gamma}$
+      splits into
+      $\gr\Psi_{\gr l} : \Ann^{\size{\Delta_l}} \to \Ann^{\size\Gamma}
+      \coloneqq \alr{\id, 0}; \gr\Psi$ and
+      $\gr\Psi_{\gr r} : \Ann^{\size{\Delta_r}} \to \Ann^{\size\Gamma}
+      \coloneqq \alr{0, \id}; \gr\Psi$, using the product structure of
       the concatenated vector space.
-      Let $\grRl \coloneqq \grP\gr\Psi_{\gr l}$ and
-      $\grRr \coloneqq \grQ\gr\Psi_{\gr r}$, by definition satisfying the
-      required equations.
-      For the action on variables, let us consider the left environment.
-      We are given $\grRprime = \grPprime\gr\Psi_{\gr l}$ and
-      $\grPprime\gamma \sqni A$.
+      Let $\grPl \coloneqq \grQl\gr\Psi_{\gr l}$ and
+      $\grPr \coloneqq \grQr\gr\Psi_{\gr r}$, by definition satisfying the
+      required constraints.
+      For the action on variables, let us consider the left environment (with
+      the right environment following symmetrically).
+      We are given $\gr{\grP'_l} \leq \gr{\grQ'_l}\gr\Psi_{\gr l}$ and
+      $\gr{\grQ'_l}\delta_l \sqni A$.
       From these, we get
-      $\grRprime = \grPprime\gr\Psi_{\gr l} = \plr{\grPprime, \gr0}\gr\Psi$ and
-      $\grPprime\gamma, \gr0\delta \sqni A$.
+      $\gr{\grP'_l} \leq \gr{\grQ'_l}\gr\Psi_{\gr l} =
+      \begin{pmatrix} \gr{\grQ'_l} & \gr0 \end{pmatrix}\gr\Psi$ and
+      $\gr{\grQ'_l}\delta_l, \gr0\delta_r \sqni A$.
       We can therefore act using the original environment.
     \item[$\cdot(\to)$]
-      Let $\grP$ and $\grPprime$ be such that $\grP = \gr r\grPprime$ and let
+      Let $\grP$ and $\grPprime$ be such that $\grP \leq \gr r\grPprime$ and let
       $v : \V\,\grPprime\gamma\,A$.
@@
       \coloneqq \gr r\gr' \mapsto \gr r\gr'\grPprime$.
-      By definition and the previous assumption, we have $\grP = \gr r\gr\Psi$.
-      When acting on a variable, we have $\grP\gr{''} = \gr r\gr'\gr\Psi$
+      By definition and the previous assumption, we have
+      $\grP \leq \gr r\gr\Psi$.
+      When acting on a variable, we have $\grP\gr{''} \leq \gr r\gr'\gr\Psi$
       and $\gr r\gr'A \sqni A'$.
-      The latter tells us that $A = A'$ and $\gr r\gr' = \gr1$.
-      Thus, $\grP\gr{''} = \grPprime$.
-      We therefore need a value of type $\V\,\grPprime\gamma\,A$, which we can
-      take to be $v$.
+      The latter tells us that $A = A'$ and $\gr r\gr' \leq \gr1$.
+      Thus, $\grP\gr{''} \leq \grPprime$.
+      Therefore, by subusaging, we may produce a value of type
+      $\V\,\grPprime\gamma\,A$, which we can take to be $v$.
     \item[$\cdot(\gets)$]
@@
       $\gr1A \sqni A$, to get a value of type $\V\,\grPprime\gamma\,A$.
-      Furthermore, we derive $\grP = \gr r\gr\Psi = \gr r\grPprime$, as
+      Furthermore, we derive $\grP \leq \gr r\gr\Psi = \gr r\grPprime$, as
       required.
@@
 the unit and lift of a Kleisli triple.
+\todo{Change from equality to inequality}
 
@@
 
+\todo{What do these choices mean?}
 \begin{example}

### 2021-08-17 tex/lnd.tex
[ write ] redraft bits on environments

@@
 The idea is to generalise the space between premises from Gentzen's natural
-deduction so as to allow for any linear manipulations of usage annotations.
+deduction so as to allow for any linear combination of usage annotations.
 Among other things, this generalisation will allow us to distinguish between

### 2021-08-17 tex/lrkits.tex
[ write ] redraft bits on environments

--- 2021-07-12 tex/lrkits.tex
@@
-As we discussed in \autoref{sec:kits}, simultaneous substitution gives a
-notion of derivability of one context from another, while simultaneous renaming
-gives a similar notion of derivability restricted to structural rules.
-To adapt these notions from an intuitionistic setting to our substructural
-setting, we must examine what it means to derive one context from another
-substructually.
+In an effort to stick to the syntactic kits and traversals approach of
+\cref{sec:syntactic-kits}, I will derive the types of simultaneous renaming
+and simultaneous substitution from a generic type of \emph{environments}.
+To get a type of environments suitable for the usage-aware setting, I first
+analyse intuitionistic environments, distilling the slick functional definition
+(\cref{def:simple-env}) into a more basic recursive definition
+(\cref{def:simple-rec-env}).
+This recursive definition is easy to make usage-aware (\cref{def:lr-rec-env}),
+which gives a basis from which to derive the function-based definition I will
+take as primary (\cref{def:lr-env}).
+The resulting definition makes explicit the role of algebraic linearity in the
+metatheory of semiring-annotated calculi.
+
+Recalling from \cref{sec:kits}, we have the following definition of
+environments for simple types.
+
+\begin{definition}[Simple environment]\label{def:simple-env}
+  A $\V$-\emph{environment} between simply typed contexts $\Gamma$ and $\Delta$
+  is a function, polymorphic in type $A$, from variables of type $A$ in
+  $\Delta$ to inhabitants of $\V\,\Gamma\,A$.
+  We write the type of such environments as $\Gamma \env\V \Delta$.
+\end{definition}
+
+The first thing to note is that this definition is inadequate for \name{}.
+For example, suppose we have a term
+$\plr{M{_\otimes}N} : \grR\gamma \vdash A \otimes B$ and a substitution
+$\sigma : \grR\gamma \env\vdash \grR\gr'\delta$.
+From the $\otimes$-I rule, we have $M : \grP\gamma \vdash A$ and
+$N : \grQ\gamma \vdash B$ for some $\grP$ and $\grQ$ such that
+$\grR \leq \grP + \grQ$.
+We want to apply $\sigma$ to the subterms $M$ and $N$, but this is impossible
+because their contexts are not $\grR\gamma$, and we have no way to adapt
+$\sigma$ to these new contexts.
+Another instructive failure is the general non-existence of identity
+environments, like a renaming of type $\gr1A, \gr1B \env\sqni \gr1A, \gr1B$.
+We do not have a variable of type $\gr1A, \gr1B \sqni A$ or, symmetrically,
+$\gr1A, \gr1B \sqni B$, because, in each case, there is one variable with
+annotation $\gr1$ which we have not actually used.
+This example suggests that the values of a usage-aware environment should be
+derived in \emph{different} usage contexts, such as in $\gr1A, \gr0B \sqni A$.
+
+To see why this definition of environment works for simply typed
+$\lambda$-calculus but not \name{}, I produce an equivalent definition by
+recursion on the target context.
+This recursive definition (\cref{def:simple-rec-env}), and particularly the
+case where $\Delta$ is a concatenation, makes it clear how $\Gamma$ is being
+copied for use in each $\V$-value.
+I take the equivalence of \cref{def:simple-env} and \cref{def:simple-rec-env}
+as obvious, because any function from variables in $\Delta$ can be
+defunctionalised as a data structure with the same shape as $\Delta$.
+
+\begin{definition}[Simple recursive environment]\label{def:simple-rec-env}
+  A \emph{recursive $\V$-environment} between simply typed contexts $\Gamma$ and
+  $\Delta$ is defined by cases on the shape of $\Delta$ (where
+  $\Gamma \env\V_R \Delta$ is the notation for the type of recursive
+  environments for given $\V$, $\Gamma$, and $\Delta$):
+  \begin{itemize}
+    \item There is one environment $\alr{} : \Gamma \env\V {\cdot}$.
+    \item For $\rho_l : \Gamma \env\V_R \Delta_l$ and
+      $\rho_r : \Gamma \env\V \Delta_r$, we have an environment
+      $\alr{\rho_l, \rho_r} : \Gamma \env\V_R \Delta_l, \Delta_r$.
+    \item For any value $v : \V\,\Gamma\,A$, we have an environment
+      $\alr{v} : \Gamma \env\V_R A$.
+  \end{itemize}
+\end{definition}
+
+I picture the sharing of $\Gamma$ in \cref{def:simple-rec-env} in the diagram
+below.
+The converging arrows from $\Gamma$ to each $\Delta_i$ represent the indices of
+values appearing in a simple environment.
 
@@
 
-Semantically, we interpret a simple substitution from $\Gamma$ to $\Delta$ as
-a map of type $\sem\Gamma \to \sem\Delta$.
-With semantics of contexts given by a Cartesian product of their elements,
-this type is
-$\left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
-\left(\prod_{(y : B) \in \Delta}\sem B\right)$.
-We can then do the following rearrangement.
-\todo{Internal vs external homs are not clear.}
+To account for usage, we must replace the simple copying of $\Gamma$ by
+copying of just the types $\gamma$ and \emph{sharing up} of the usage
+annotations $\grP$.
+Fortunately, our three basic ways of sharing up usage vectors --- zero,
+addition, and scaling --- apply directly to the three possible shapes of the
+target context --- empty, concatenation, and a usage-annotated singleton.
 
-\begin{align*}
-  &\phantom{{}={}} \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to
-    \left(\prod_{(y : B) \in \Delta}\sem B\right) \\
-  &\cong \prod_{(y : B) \in \Delta}\left(
-         \left(\prod_{(x : A) \in \Gamma}\sem A\right) \to \sem B
-         \right) \\
-  &= \prod_{(y : B) \in \Delta}\sem{\Gamma \vdash B} \\
-  &\cong \prod_B\left(\Delta \ni B \to \sem{\Gamma \vdash B}\right)
-\end{align*}
+\begin{definition}[Usage-annotated recursive environment]\label{def:lr-rec-env}
+  A \emph{recursive $\V$-environment} between annotated contexts $\Gamma$ and
+  $\Delta$ is defined by cases on the shape of $\Delta$ (where
+  $\Gamma \env\V_R \Delta$ is the notation for the
+  type of recursive environments for given $\V$, $\Gamma$, and $\Delta$):
+  \begin{itemize}
+    \item There is one environment $\alr{} : \grP\gamma \env\V {\cdot}$
+      whenever $\grP \leq \gr0$.
+    \item For $\rho_l : \grPl\gamma \env\V_R \Delta_l$ and
+      $\rho_r : \grPr\gamma \env\V \Delta_r$, we have an environment
+      $\alr{\rho_l, \rho_r} : \grP\gamma \env\V_R \Delta_l, \Delta_r$ whenever
+      $\grP \leq \grPl + \grPr$.
+    \item For any value $v : \V\,\grPprime\gamma\,A$, we have an environment
+      $\alr{v} : \grP\gamma \env\V_R \gr rA$ whenever
+      $\grP \leq \gr r\grPprime$.
+  \end{itemize}
+\end{definition}
 
-Taking the semantics to be the term model, we get the standard definition of a
-simultaneous substitution.
-To see how this definition works, let us construct an example substitution:
-$A, B \to C, B \Longrightarrow B, C$.
-Intuitionistically, it suffices to give two separate substitutions,
-$A, B \to C, B \Longrightarrow B$ and $A, B \to C, B \Longrightarrow C$,
-which are both just terms.
-We indeed have terms $x : A, y : B \to C, z : B \vdash y : B$ and
-$x : A, y : B \to C, z : B \vdash y\,z : C$.
+\begin{example}
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
+  addition and multiplication.
+  There is an intuitionistic recursive environment
+  \[
+    \alr{\alr{z},\alr{y\,z}} :
+    \plr{x : A, y : B \to C, z : B} \env\vdash_R \plr{B, C}.
+  \]
+  There is also a usage-aware recursive environment
+  \[
+    \alr{\alr{z},\alr{y\,z}} :
+    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash_R
+    \plr{\gr1B, \gr2C}.
+  \]
+  The latter relies on the observations that
+  $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
+  \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}
+  + \begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ and, on the right, that
+  $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix} =
+  \gr2\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
+  Then, we have $\gr0x : A, \gr0y : B \multimap C, \gr1z : B \vdash z : B$ and
+  $\gr0x : A, \gr1y : B \multimap C, \gr1z : B \vdash y\,z : C$.
+\end{example}
 
-When working with our semiring-annotated calculus \name{}, instead of Cartesian
-products we have (weighted) monoidal products, giving the type
-$\left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
-\left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right)$.
-As the monoidal product is not a limit, we cannot do the rearrangement we did
-above.
-Instead, we must divide the domain up into pieces, with each piece supporting
-one factor from the codomain.
-
-To see this division, let us adapt the previous example to have type
-$\gr0A, \gr2(B \multimap C), \gr3B \Longrightarrow \gr1B, \gr2C$.
-To divide the goal up, we note that $
-\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
-\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix} +
-\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}
-$, so it suffices to give substitutions of types
-$\gr0A, \gr0(B \multimap C), \gr1B \Longrightarrow \gr1B$ and
-$\gr0A, \gr2(B \multimap C), \gr2B \Longrightarrow \gr2C$.
-Furthermore, our term calculus only supports $\gr1$-annotated conclusions,
-so we divide the second substitution type through by $\gr2$.
-Finally, we give the terms largely as before:
-$\gr0x : A, \gr0y : B \to C, \gr1z : B \vdash y : B$ and
-$\gr0x : A, \gr1y : B \to C, \gr1z : B \vdash z\,y : C$.
+From the example, we can see that the important usage vectors are the initial
+one $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix}$ and the usage vectors
+at which terms are derived: $\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$
+and $\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
+I will call the latter the \emph{leaf vectors}.
+The intermediate vector $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ can
+be worked out from the leaf vector
+$\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$ and the scaling factor
+$\gr2$ found in the codomain context $\gr1B, \gr2C$.
+Even when the ordering on annotations is given by a non-equivalence relation
+$\leq$, there is a canonical least choice for all of the intermediate vectors,
+together with a constraint that the entire linear combination of all the leaf
+vectors is less than or equal to the initial usage vector.
+In symbols, we may let $\gr\Psi$ be the collection of leaf vectors, and state
+the constraint as $\grP \leq \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
+Seeing $\gr\Psi$ instead as a $\size\Delta \times \size\Gamma$ matrix, this
+constraint is $\grP \leq \grQ\gr\Psi$, using vector-matrix multiplication.
+The resulting picture is below, showing $\grP$ being split up into $\gr\Psi$,
+and then each $\V$-value being constructed in a separate $\gr\Psi_i\gamma$.
 
@@
   \end{tikzpicture}
-  \quad\textrm{where }\grP = \grQ\gr\Psi
+  \quad\textrm{where }\grP \leq \grQ\gr\Psi
 \end{displaymath}
 
-Generalising this division process, we derive the following (letting
-$\Gamma = \grP\gamma$ and $\Delta = \grQ\delta$).
-The various fragments of the original usage vector are collected up in the
-existentially quantified $\gr\Psi$.
+From this point, we can recover a functional-style definition of usage-aware
+environments.
+We choose our leaf vectors $\gr\Psi$ up-front, check the inequality, and then
+produce a value at each leaf vector.
+%From this definition, we can recover a functional-style definition by
+%separating choices of usage vectors from the provision of $\V$-values.
+%In particular, the only choices of usage vectors that are essential are the
+%$\grPprime$s in the singleton case, with the choices in the concatenation case
+%being determined as scalings and sums of these $\grPprime$s.
+%I let $\gr\Psi$ collect up these $\size\Delta$-many choices of
+%$\size\Gamma$-length usage vectors and note that the constraint on $\gr\Psi$
+%generated by all the scaling and summing is
+%$\grP = \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
 
-\begin{align*}
-  &\phantom{{}={}}
-    \left(\bigotimes_{(\gr rx : A) \in \Gamma}\oc \gr r\sem A\right) \to
-    \left(\bigotimes_{(\gr sy : B) \in \Delta}\oc \gr s\sem B\right) \\
-  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
-    \left(\grP = \sum\gr\Psi\right) \times
-    \prod_{(\gr sy : B) \in \Delta}\left(
-    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
-    \to \oc \gr s\sem B
-    \right) \\
-  &\cong \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
-    \left(\grP = \sum_{(\gr sy : B) \in \Delta}\gr s\gr\Psi_y\right) \times
-    \prod_{(\gr sy : B) \in \Delta}\left(
-    \left(\bigotimes_{(\gr rx : A) \in \gr\Psi_y\gamma}\oc \gr r\sem A\right)
-    \to \sem B
-    \right) \\
-  &= \sum_{\gr\Psi : \size\Delta \to \size\Gamma \to \Ann}
-    \left(\grP = \grQ\gr\Psi\right) \times
-    \prod_{(\gr sy : B) \in \Delta} \sem{\gr\Psi_y\gamma \vdash B}
-\end{align*}
+\begin{definition}[Usage-annotated environment (tentative)]
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a matrix $\gr\Psi : \Ann^{\size\Delta \times \size\Gamma}$ such that
+  $\grP \leq \grQ\gr\Psi$ and for each
+  $\plr{x : A} \in \delta$ we have a value of type $\V\,\gr\Psi_x\gamma\,A$.
+\end{definition}
 
-In other words, we interpret a linear substitution as a \emph{matrix}
-$\gr\Psi$ relating the two usage contexts, together with a term for each
-variable from $\Delta$ typed in a fragmented usage context dictated by
-$\gr\Psi$.
+I find this definition somewhat fiddly because of its reliance on low-level
+concepts like non-usage-checked variables and rows of a matrix.
+We may note that $\gr\Psi_x = \bra x\gr\Psi$, from which point, requiring not
+just $\V\,\gr\Psi_x\gamma\,A$ but rather $\V\,\plr{\grQprime\gr\Psi}\gamma\,A$
+for any $\grQprime \leq \bra x$ is a minor change (and equivalent if $\V$
+respects subusaging, which is practically always the case).
+``An $x$ such that $(x : A) \in \delta$ and $\grQprime \leq \bra x\gr\Psi$''
+is exactly the definition of $\grQprime\delta \sqni A$.
+I further regularise this clause by asking for a
+$\grPprime \leq \grQprime\gr\Psi$ rather than $\grQprime\gr\Psi$ exactly,
+leaving us needing, for each $\grPprime$ and $\grQprime$ related in the same
+way ($\gr\Psi$) as $\grP$ and $\grQ$, a function from $\grQprime\delta \sqni A$
+to $\V\,\grPprime\gamma\,A$.
+Finally, I choose to switch from matrices and matrix multiplication to
+linear maps and their actions, which are easier to work with.
+All of these changes yield my primary definition of an environment for
+usage-annotated calculi.
 
-While we naturally derive a matrix as a fragmentation of a usage vector, we can
-get a slightly cleaner presentation by instead using an abstract linear map.
-Let $\gr\Psi$ now be a linear map of type
-$\Ann^{\size\Delta} \to \Ann^{\size\Gamma}$, with application written postfix.
-The equation $\grP = \grQ\gr\Psi$ remains unchanged.
-As for $\prod_{(\gr sy : B) \in \Delta}\sem{\gr\Psi_y\gamma \vdash B}$, the most
-direct fix is to replace $\gr\Psi_y$ by $\langle y \rvert\gr\Psi$.
-But then we notice that $\langle y \rvert$ is exactly the $\grQprime$ satisfying
-$\grQprime\delta \sqni y : B$.
-Thus, let us derive the following.
+\begin{definition}[Usage-annotated environment]\label{def:lr-env}
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
+  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
+  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
+  $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
+\end{definition}
 
-\begin{align*}
-  &\phantom{{}={}} \prod_{(\gr sy : B) \in \Delta}
-    \sem{\langle y \rvert \gr\Psi\gamma \vdash B} \\
-  &\cong \prod_{B,\grQprime} \left(\grQprime\delta \sqni B \to
-    \sem{\grQprime\gr\Psi\gamma \vdash B}\right) \\
-  &\cong \prod_{B,\grQprime,\grPprime} \left(
-    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
-    \sem{\grPprime\gamma \vdash B}\right)
-\end{align*}
+The following simple lemma shows that usage-annotated environments are, in a
+sense, as good as simple environments on usage-checked variables.
+What usage-annotated environments give us beyond simple environments is the
+ability to accommodate linear decompositions, in a way I will make precise in
+the next section.
 
-Putting everything back together gives us the final expression.
-
-\begin{align*}
-  &\sum_{\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}}
-    \left(\grP = \grQ\gr\Psi\right) \times
-    \prod_{B,\grQprime,\grPprime} \left(
-    \grPprime = \grQprime\gr\Psi \to \grQprime\delta \sqni B \to
-    \sem{\grPprime\gamma \vdash B}\right)
-\end{align*}
-
-We now have a new reading for the interpretation of a linear substitution:
-a linear map $\gr\Psi$ relating the two usage vectors $\grP$ and $\grQ$, and
-for any two similarly related usage vectors $\grPprime$ and $\grQprime$, we
-have a type-preserving function from variables in $\grQprime\delta$ to terms in
-$\grPprime\gamma$.
-Even though we don't use $\gr\Psi$ as a matrix containing fragmented usage
-vectors, we can still justify why it should be a \emph{linear} map.
-We need $\gr\Phi$ to respect all fragmentation of the usage context in a typing
-rule, and we know that all such fragmentation is done by linear operations
-zero, addition, and scaling by a constant.
-\todo{Expand. Substitutions need to preserve everything done to the context,
-and linear things are all we do to the context.}
+\begin{lemma}
+  We can use an environment $\rho : \Gamma \env\V \Delta$ to get from a
+  usage-checked variable $x : \Delta \sqni A$ to a value of type
+  $\V\,\Gamma\,A$.
+\end{lemma}
+\begin{proof}
+  Let $\Gamma = \grP\gamma$ and $\Delta = \grQ\delta$.
+  Set $\grPprime \coloneqq \grP$ and $\grQprime \coloneqq \grQ$, then
+  $\grPprime \leq \grQprime\gr\Psi$ by the constraint in $\rho$, and we can
+  use the lookup function to map $x$ to a value.
+\end{proof}

### 2021-08-17 tex/gen-sem.tex
[ new ] start writing about generic semantics

The traversal \AgdaFunction{trav} from the last section is generic in the sense
that $\V$, the type of entries in an environment, can be instantiated to many
different things.
However, in practice we only use $\ni$ and $\vdash$, giving us renaming and
substitution, respectively.
This is because \AgdaFunction{trav} only targets terms, and does so by keeping
term constructors intact and replacing only the variables by things from the
environment.
This makes substitution the most general possible traversal.

If we want to capture a broader range of traversals, including not just
syntactic but also \emph{semantic} operations, we must be able to target things
other than terms, and act in an interesting way on term constructors.
Doing a straight generalisation of the type of \AgdaFunction{trav}, this
suggests that we want a function with the following type.

\missingfigure{Semantic traversal type signature}

### 2021-08-23 tex/lr.tex
[ format ] make typing rules fit in margins
[ format ] add comb proof style
[ writing ] explain some features of λR

@@
         \hypo{\grR \le \grP + \grQ}
-        \hypo{\grP\gamma \vdash A}
-        \hypo{\grQ\gamma \vdash B}
-        \infer3[$\otimes$-I]{\grR\gamma \vdash A \otimes B}
+        \hypo{%
+          \begin{matrix}
+            \grP\gamma \vdash A \\ \grQ\gamma \vdash B
+          \end{matrix}%
+        }
+        \infer2[$\otimes$-I]{\grR\gamma \vdash A \otimes B}
       \end{prooftree}
@@
       \begin{prooftree}
-        \hypo{\grR \le \grP + \grQ}
-        \hypo{\grP\gamma \vdash A \otimes B}
+        \hypo{%
+          \begin{matrix}
+            \grR \le \grP + \grQ \\ \grP\gamma \vdash A \otimes B
+          \end{matrix}%
+        }
         \hypo{\grQ\gamma, \gr1A, \gr1B \vdash C}
-        \infer3[$\otimes$-E]{\grR\gamma \vdash C}
+        \infer2[$\otimes$-E]{\grR\gamma \vdash C}
       \end{prooftree}
@@
       \begin{prooftree}
-        \hypo{\grR \le \grP + \grQ}
-        \hypo{\grP\gamma \vdash A \oplus B}
-        \hypo{\grQ\gamma, \gr1A \vdash C}
-        \hypo{\grQ\gamma, \gr1B \vdash C}
-        \infer4[$\oplus$-E]{\grR\gamma \vdash C}
+        \hypo{%
+          \begin{matrix}
+            \grR \le \grP + \grQ \\ \grP\gamma \vdash A \oplus B
+          \end{matrix}%
+        }
+        \hypo{%
+          \begin{matrix}
+            \grQ\gamma, \gr1A \vdash C \\ \grQ\gamma, \gr1B \vdash C
+          \end{matrix}%
+        }
+        \infer2[$\oplus$-E]{\grR\gamma \vdash C}
       \end{prooftree}
@@
         \hypo{\grP\gamma \vdash A}
-        \infer2[$\oc$-I]{\grR\gamma \vdash \oc_{\gr r} A}
+        \infer2[$\oc$-I]{\grR\gamma \vdash \oc\gr rA}
       \end{prooftree}
@@
         \hypo{\grR \le \grP + \grQ}
-        \hypo{\grP\gamma \vdash \oc_{\gr r} A}
+        \hypo{\grP\gamma \vdash \oc\gr rA}
         \hypo{\grQ\gamma, \gr rA \vdash C}
@@
 
-The operational semantics of \name{} are standard.
-Assuming type preservation under reduction, \name{} enjoys all of the standard
-operational properties via a forgetful translation to simply typed
-$\lambda$-calculus.
+I will not cover the operational semantics or equational theory of \name{} in
+this thesis.
+I will discuss a denotational semantics in \cref{sec:wrel}.
+
+The following features are of note.
+
+\paragraph{Subusaging}
+Several typing rules contain constraints of the form $\grP \leq \grQ$, for
+certain usage vectors $\grP$ and $\grQ$.
+To a first approximation, $\leq$ can simply be read as $=$.
+The point of using $\leq$ rather than $=$ is to allow for \emph{subusaging},
+i.e., subsumption of usage annotations.
+For usage annotations $\gr r$ and $\gr s$, the inequality $\gr r \leq \gr s$
+states that an assumption with annotation $\gr r$ can be used wherever an
+assumption with annotation $\gr s$ is required.
+A mnemonic is that $\gr r$ is less specific than $\gr s$.
+The principle is reflected by the admissible subusaging rule.
+
+\[
+  \begin{prooftree}
+    \hypo{\grP \leq \grQ}
+    \hypo{\grQ\gamma \vdash A}
+    \infer2[Subuse]{\grP\gamma \vdash A}
+  \end{prooftree}
+\]
+
+Subusaging is essential to nearly all interesting choices of $\Ann$.
+For example, we can capture intuitionistic linear logic by choosing $\Ann$ to
+be $\{\gr0 > \gr\omega < \gr1\}$.
+This allows variables annotated $\gr\omega$ (``unrestricted'') to be both
+weakened/discarded (because $\gr\omega \leq \gr0$) and derelicted/used
+(because $\gr\omega \leq \gr1$).
+
+\paragraph{Tensor- and with-products}
+Like intuitionistic linear logic (ILL), \name{} distinguishes tensor-products
+($A \otimes B$) from with-products ($A \with B$).
+Whereas in ILL, rules like $\otimes$-introduction involve splitting the
+assumptions between the two subterms, in \name{}, this splitting is done by
+choosing usage annotations which add up to the usage annotations of the
+conclusion.
+For example, we can derive $\vdash A \otimes B \multimap B \otimes A$ as
+follows.
+Notice that the assumption $A \otimes B$ is still present in all subderivations,
+even after it has been ``used up''.
+The only thing that stops us using the assumption again is that, for a general
+choice of $\Ann$, we do not have $\gr0 \leq \gr1$ or $\gr1 \leq \gr1 + \gr1$.
+
+\begin{small}
+  \[
+    \nabla \coloneqq
+    \begin{prooftree}
+      \infer0{\plr{\gr0\;\gr1\;\gr1} \leq
+        \plr{\gr0\;\gr0\;\gr1} + \plr{\gr0\;\gr1\;\gr0}}
+      \infer0{\plr{\gr0\;\gr0\;\gr1} \leq \plr{\gr0\;\gr0\;\gr1}}
+      \infer1[Var]{\gr0\plr{A \otimes B}, \gr0A, \gr1B \vdash B}
+      \infer0{\plr{\gr0\;\gr1\;\gr0} \leq \plr{\gr0\;\gr1\;\gr0}}
+      \infer1[Var]{\gr0\plr{A \otimes B}, \gr1A, \gr0B \vdash A}
+      \infer3[$\otimes$-I]%
+      {\gr0\plr{A \otimes B}, \gr1A, \gr1B \vdash B \otimes A}
+    \end{prooftree}
+  \]
+
+  \[
+    \begin{prooftree}
+      \infer0{\plr{\gr1} \leq \plr{\gr1} + \plr{\gr0}}
+      \infer0{\plr{\gr1} \leq \plr{\gr1}}
+      \infer1[Var]{\gr1\plr{A \otimes B} \vdash A \otimes B}
+      \hypo{\nabla}
+      \infer[no rule]1{\gr0\plr{A \otimes B}, \gr1A, \gr1B \vdash B \otimes A}
+      \infer3[$\otimes$-E]{\gr1\plr{A \otimes B} \vdash B \otimes A}
+      \infer1[$\multimap$-I]{\vdash A \otimes B \multimap B \otimes A}
+    \end{prooftree}
+  \]
+\end{small}
+
+\begin{example}
+  The following judgements hold for any partially ordered semiring.
+  Derivations are left as an exercise to the reader.
+  \begin{itemize}
+    \item $\vdash A \oplus A \multimap A$
+    \item $\vdash A \multimap A \with A$
+    \item $\vdash A \oplus 0 \multimapboth A$
+    \item $\vdash A \otimes 0 \multimapboth 0$
+    \item $\vdash \oc\gr1A \multimapboth A$
+    \item If $\gr r \leq \gr s$, then $\vdash \oc\gr rA \multimap \oc\gr sA$
+  \end{itemize}
+\end{example}
+
+To get a feeling for \name{}, I will temporarily fix
+$\Ann \coloneqq (\mathbb N, =, 0, +, 1, \times)$.
+That is, the usual semiring of natural numbers with ordering given by equality.
+Under this discipline, the usage constraints enforce a form of exact usage
+counting.
+
+\begin{example}
+  The following judgements hold for the natural number semiring.
+  Derivations are left as an exercise to the reader.
+  \begin{itemize}
+    \item $\vdash \oc\gr2A \multimap A \otimes A$
+    \item $\vdash \oc\gr5A \multimap \oc\gr2A \otimes \oc\gr3A$
+  \end{itemize}
+\end{example}

### 2021-08-23 tex/rec.tex
[ writing ] a bit about recursion/lists

Based on an intuitive understanding of ``usage'', recursion introduces a new
phenomenon relative to the forms of programs we have seen so far:
terms can be used an unbounded number of times.
For example, notice the following reduction in Agda.

\missingfigure{\texttt{foldr \_+\_ 0 (1 :: 2 :: 3 :: []) = 1 + (2 + (3 + 0))}}

The function \AgdaFunction{\_+\_} has been copied into 3 different places in
the running of the program.
This copying is despite no type telling us that \AgdaFunction{\_+\_} would be
used 3 times (both \verb|[1,2,3]| and \verb|[2,3]| have type
\AgdaDatatype{List}\AgdaSpace{}\AgdaDatatype{$\mathbb N$}, despite the
corresponding folds using \AgdaFunction{\_+\_} a different number of times).
As such, when checking an application of \AgdaFunction{foldr}, we need check
that we can use its functional argument (\AgdaFunction{\_+\_} in this case) an
arbitrary number of times.
If we were to fix $\Ann$ as the $\{\gr0, \gr1, \gr\omega\}$ posemiring, then
wrapping the type of the functional argument in $\oc\gr\omega$ would suffice.
However, we want to remain generic in the choice of semiring.

I propose the following additions to \name{} to support a broad class of
inductive types.
I define strictly positive functors syntactically, with the only notable
restriction being not being allowed to use the type variable $X$ in the domain
of a function type.
I then add least fixed points of such strictly positive functors to the syntax
of types.

\begin{align*}
  U &\Coloneqq A \multimap (-) \mid \oc\gr r(-) \\
  {\odot} &\Coloneqq {\otimes} \mid {\oplus} \mid {\with} \\
  F[X], G[X] &\Coloneqq X \mid A \mid U(F[X]) \mid F[X] \odot G[X] \\
  A &\Coloneqq \cdots \mid \mu X.~F[X]
\end{align*}

\begin{example}
  We may define $\mathrm{List}_A \coloneqq \mu X.~I \oplus \plr{A \otimes X}$.
\end{example}

In the typing rules, introduction of an inductive type is standard.
For the elimination rule, we follow a similar pattern to other pattern-matching
rules --- $\oplus$-E, $\otimes$-E, and $\oc$-E --- by splitting the context
and typing the eliminand in one half ($\grP$).
We type the continuation in the other half, but because the continuation may
be used multiple times, and in a modal context, we require that $\grQ$ is
preserved by all linear operations.

\begin{displaymath}
  \begin{prooftree}
    \hypo{\grR\gamma \vdash F[\mu X.~F[X]]}
    \infer1[$\mu$-I]{\grR\gamma \vdash \mu X.~F[X]}
  \end{prooftree}
\end{displaymath}
\begin{displaymath}
  \begin{prooftree}
    \hypo{\grR \leq \grP + \grQ}
    \hypo{\grP\gamma \vdash \mu X.~F[X]}
    \hypo{%
      \begin{matrix*}[l]
        \grQ \leq \gr0 \\
        \grQ \leq \grQ + \grQ \\
        \forall \gr r.~\grQ \leq \gr r\grQ
      \end{matrix*}%
    }
    \hypo{\grQ\gamma, \gr1F[C] \vdash C}
    \infer4[$\mu$-E]{\grR\gamma \vdash C}
  \end{prooftree}
\end{displaymath}

\begin{example}\label{thm:list-rules}
  For lists, we can derive the following introduction and elimination rules
  (with usage constraints omitted for brevity when obvious).

  \begin{align*}
    \begin{prooftree}
      \hypo{\grR \leq \gr0}
      \infer1[$I$-I]{I}
      \infer1[$\oplus$-I$_0$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
    &&
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash A}
      \hypo{\grQ\gamma \vdash \mathrm{List}_A}
      \infer3[$\otimes$-I]{\grR\gamma \vdash A \otimes \mathrm{List}_A}
      \infer1[$\oplus$-I$_1$]%
      {\grR\gamma \vdash I \oplus \plr{A \otimes \mathrm{List}_A}}
      \infer1[$\mu$-I]{\grR\gamma \vdash \mathrm{List}_A}
    \end{prooftree}
  \end{align*}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\grP\gamma \vdash \mathrm{List}_A}
      \infer0[Var]{\gr0\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash I \oplus \plr{A \otimes C}}
      \hypo{\nabla^n}
      \hypo{\nabla^c}
      \infer3[$\oplus$-E]{\grQ\gamma, \gr1\plr{I \oplus \plr{A \otimes C}}
        \vdash C}
      \infer2[$\mu$-E]{\grR\gamma \vdash C}
    \end{prooftree}
  \end{displaymath}
  \begin{align*}
    \textrm{where }\nabla^n &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1I \vdash I}
      \hypo{\grQ\gamma \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0I \vdash C}
      \infer2[$I$-E]{\grQ\gamma, \gr1I \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1I
        \vdash C}
    \end{prooftree}
    \\\\
    \textrm{and }\nabla^c &\coloneqq
    \begin{prooftree}
      \infer0[Var]{\gr0\gamma, \gr1\plr{A \otimes C} \vdash A \otimes C}
      \hypo{\grQ\gamma, \gr1A, \gr1C \vdash C}
      \infer1[Wk]{\grQ\gamma, \gr0\plr{A \otimes C}, \gr1A, \gr1C \vdash C}
      \infer2[$\otimes$-E]{\grQ\gamma, \gr1\plr{A \otimes C} \vdash C}
      \infer1[Wk]%
      {\grQ\gamma, \gr0\plr{I \oplus \plr{A \otimes C}}, \gr1\plr{A \otimes C}
        \vdash C}
    \end{prooftree}
  \end{align*}
\end{example}

Following \cref{sec:lnd}, I want to turn the ad hoc constraints on $\grP$,
$\grQ$, and $\grR$ into the result of some premise combinators.
To do this, I introduce a new combinator $\Box^{0{+}{\times}}$ defined below,
along with the resulting implicit-context typing rules.

\begin{align*}
  \plr{\Box^{0{+}{\times}}\,T}\grR \coloneqq
  \plr{\grR \leq \gr0} \times \plr{\grR \leq \grR + \grR} \times
  \plr{\forall \gr r.~\grR \leq \gr r\grR} \times T\,\grR
\end{align*}

\begin{align*}
  \begin{prooftree}[comb]
    \hypo{\vdash F[\mu X.~F[X]]}
    \infer1[$\mu$-I]{\vdash \mu X.~F[X]}
  \end{prooftree}
  &&
  \begin{prooftree}[comb]
    \hypo{\vdash \mu X.~F[X]}
    \hypo{\sep}
    \hypo{\Box^{0{+}{\times}}\plr{\gr1F[C] \vdash C}}
    \infer3[$\mu$-E]{\vdash C}
  \end{prooftree}
\end{align*}

\begin{example}
  We can state the rules for lists derived in \cref{thm:list-rules} as follows.
  \begin{align*}
    \begin{prooftree}[comb]
      \hypo{I^*}
      \infer1{\vdash \mathrm{List}_A}
    \end{prooftree}
    &&
    \begin{prooftree}[comb]
      \hypo{\vdash A}
      \hypo{\sep}
      \hypo{\vdash \mathrm{List}_A}
      \infer3{\vdash \mathrm{List}_A}
    \end{prooftree}
    &&
    \begin{prooftree}[comb]
      \hypo{\vdash \mathrm{List}_A}
      \hypo{\sep}
      \hypo{\Box^{0{+}{\times}}
        \plr{\vdash C\hskip0.75em\dottimes\hskip0.75em\gr1A, \gr1C \vdash C}}
      \infer3{\vdash C}
    \end{prooftree}
  \end{align*}
\end{example}

### 2021-08-23 tex/semirings.tex
[ refactor ] put sec 5 in an \include

@@
+\chapter{Usage restriction via semirings}
+
 Addition gives the structural rules.
@@
 
-\subsection{Partiality (or not)} \todo{Appendix?}
-
-As we have seen \todo{Have we?}, the way additive and multiplicative rules are
-realised algebraically is related to models of separation logic.
-Models of separation logic typically use \emph{partial} commutative monoids to
-model a heap, so it is tempting to generalise the commutative monoid of
-addition in our semirings to a \emph{partial} commutative monoid.
-However, we find that the most natural notion of \emph{partial semiring} is
-degenerate, in that all partial semirings are actually (total) semirings.
-
-Recall that a commutative monoid (or commutative monoid object) can be
-defined in any symmetric monoidal category.
-A partial commutative monoid is exactly a commutative monoid object in the
-category of sets and partial functions with the usual monoidal product
-\todo{Define this monoidal product somewhere.}.
-However, semirings need a Cartesian category in order to state the interaction
-equations between addition and multiplication.
-While the category of sets and partial functions is not Cartesian, the
-standard way to manufacture a Cartesian category out of a symmetric monoidal
-category $\mathcal C$ is to take the category of cocommutative comonoids
-$\mathrm{CComon}(\mathcal C)$.
-Intuitively, the cocommutative comonoid structure equips the underlying
-object $M$ with a \emph{delete} map $\eta : M \to I$ and a \emph{duplicate}
-map $\delta : M \to M \otimes M$ which are coherent with respect to each other.
-All morphisms in $\mathrm{CComon}(\mathcal C)$ must respect $\eta$ and
-$\delta$; in particular, both addition and multiplication must separately
-form bimonoids in $\mathcal C$ together with the cocommutative comonoid.
-
-\[
-  \begin{tikzpicture}[baseline]
-    \path
-    (-1,1) node(0) {0}
-    (1,2) node(x) {}
-    (0,0) node(*) {*}
-    (0,-1) node(res) {}
-    ;
-
-    \draw (0) -- (*);
-    \draw (x) to[out=270,in=45] (*);
-    \draw (*) -- (res);
-  \end{tikzpicture}
-  =\quad
-  \begin{tikzpicture}[baseline]
-    \path
-    (0,0) node(0) {0}
-    (0,2) node(x) {}
-    (0,-1) node(res) {}
-    (0,1) node[circle,draw](del) {}
-    ;
-
-    \draw (0) -- (res);
-    \draw (x) -- (del);
-  \end{tikzpicture}
-  \quad=
-  \begin{tikzpicture}[baseline]
-    \path
-    (1,1) node(0) {0}
-    (-1,2) node(x) {}
-    (0,0) node(*) {*}
-    (0,-1) node(res) {}
-    ;
-
-    \draw (0) -- (*);
-    \draw (x) to[out=270,in=135] (*);
-    \draw (*) -- (res);
-  \end{tikzpicture}
-\]
-\begin{displaymath}
-  \begin{matrix}
-    \begin{tikzpicture}[baseline]
-      \path
-      (-1,2) node(x) {}
-      (0,2) node(y) {}
-      (-0.5,1) node(+) {+}
-      (1,2) node(z) {}
-      (0,0) node(*) {*}
-      (0,-1) node(res) {}
-      ;
-
-      \draw (x) to[out=270,in=135] (+);
-      \draw (y) to[out=270,in=45] (+);
-      \draw (+) to[out=270,in=135] (*);
-      \draw (z) to[out=270,in=45] (*);
-      \draw (*) -- (res);
-    \end{tikzpicture}
-    =
-    \begin{tikzpicture}[baseline]
-      \path
-      (-2,3) node(x) {}
-      (-1,3) node(y) {}
-      (0,3) node(z) {}
-      (0,2) node[circle,draw](dup) {}
-      (-1,1) node(x*) {*}
-      (0,1) node(y*) {*}
-      (-0.5,0) node(+) {+}
-      (-0.5,-1) node(res) {}
-      ;
-
-      \draw (z) -- (dup);
-      \draw (x) to[out=270,in=135] (x*);
-      \draw (y) to[out=270,in=135] (y*);
-      \draw (dup) to[out=180,in=45] (x*);
-      \draw (dup) -- (y*);
-      \draw (x*) to[out=270,in=135] (+);
-      \draw (y*) to[out=270,in=45] (+);
-      \draw (+) -- (res);
-    \end{tikzpicture}
-    &\phantom{mmmm}&
-    \begin{tikzpicture}[baseline]
-      \path
-      (1,2) node(x) {}
-      (0,2) node(y) {}
-      (0.5,1) node(+) {+}
-      (-1,2) node(z) {}
-      (0,0) node(*) {*}
-      (0,-1) node(res) {}
-      ;
-
-      \draw (x) to[out=270,in=45] (+);
-      \draw (y) to[out=270,in=135] (+);
-      \draw (+) to[out=270,in=45] (*);
-      \draw (z) to[out=270,in=135] (*);
-      \draw (*) -- (res);
-    \end{tikzpicture}
-    =
-    \begin{tikzpicture}[baseline]
-      \path
-      (2,3) node(x) {}
-      (1,3) node(y) {}
-      (0,3) node(z) {}
-      (0,2) node[circle,draw](dup) {}
-      (1,1) node(x*) {*}
-      (0,1) node(y*) {*}
-      (0.5,0) node(+) {+}
-      (0.5,-1) node(res) {}
-      ;
-
-      \draw (z) -- (dup);
-      \draw (x) to[out=270,in=45] (x*);
-      \draw (y) to[out=270,in=45] (y*);
-      \draw (dup) to[out=0,in=135] (x*);
-      \draw (dup) -- (y*);
-      \draw (x*) to[out=270,in=45] (+);
-      \draw (y*) to[out=270,in=135] (+);
-      \draw (+) -- (res);
-    \end{tikzpicture}
-  \end{matrix}
-\end{displaymath}
-
-\begin{lemma}
-  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, there is
-  a cocommutative comonoid over $X$.
-\end{lemma}
-\begin{proof}
-  Let $\eta(x) \coloneq ()$ and $\delta(x) \coloneq (x, x)$, with both
-  being defined for all $x$.
-  Checking that these satisfy the laws is routine.
-  Alternatively, we can see that both $\eta$ and $\delta$, being total, are
-  morphisms in $\mathrm{Set}$, where it is well known that they form a
-  cocommutative comonoid.
-  The equations in $\mathrm{Set}$ carry over to $\mathrm{Set}_{\mathrm{part}}$.
-\end{proof}
-
-\begin{lemma}
-  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, any
-  comonoid over $X$ is the one described in the previous lemma.
-\end{lemma}
-\begin{proof}
-  The left unit law says that, for all $x$ and $x'$, we have
-  $\exists y.~\delta(x) = (y, x') \land \eta(y) = ()$ if and only if $x = x'$.
-  Letting $x'$ be $x$ and reading from right to left, we get that there is
-  some $y$ such that $\delta(x) = (y, x)$ and $\eta(y) = ()$.
-  Symmetrically, from the right unit law, there is some $z$ such that
-  $\delta(x) = (x, z)$ and $\eta(z) = ()$.
-  But because $\delta$, as a partial function, is deterministic, we have
-  $(y, x) = (x, z)$, giving us that $y = z = x$, and $\delta(x) = (x, x)$.
-  Moreover, $\eta(x) = \eta(y) = ()$.
-\end{proof}
+\section{A usage-annotated calculus}\label{sec:lr}
+\input{lr.tex}
+\section{Linear natural deduction}\label{sec:lnd}
+\input{lnd.tex}
+\section{What are linear renaming and substitution?}\label{sec:lrkits}
+\input{lrkits.tex}
+\section{Properties of linear environments}\label{sec:lenv}
+\input{lenv.tex}
+\section{Substitution is admissible in \name{}}\label{sec:lrsub}
+\input{lrsub.tex}
+\section{Recursion}\label{sec:rec}
+\input{rec.tex}
+\section{Addendum: (lack of) partiality}\label{sec:part}
+\input{part.tex}

### 2021-09-07 tex/cpp21/algebra.tex
[ new ] start CPP paper

The basic algebraic structure we concern ourselves with is \emph{partially
ordered semirings}, or \emph{posemirings} for short.
A posemiring is a (not necessarily commutative) semiring on a partially ordered
set, where both operations are monotonic.

\begin{definition}
  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that the
  following laws hold (universally quantified in all the variables).
  \begin{itemize}
    \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
    \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
    \item $0 + x = x \land x + 0 = x$
    \item $(x + y) + z = x + (y + z)$
    \item $x + y = y + x$
    \item $1 * x = x \land x * 1 = x$
    \item $(x * y) * z = x * (y * z)$
    \item $0 * x = 0 \land x * 0 = 0$
    \item $(x + y) * z = x * z + y * z$
    \item $x * (y + z) = x * y + x * z$
  \end{itemize}
\end{definition}

We also make use, particularly in the mechanisation, of an alternative
definition based on relations.

\begin{definition}
  A \emph{relational posemiring} is a tuple
  $(\Ann, \leq, {{-}\leq}0, {{-}\leq}[{-}+{-}],
  {{-}\leq}1, {{-}\leq}[{-}*{-}])$
  such that the following laws hold (with free variables universally
  quantified).
  \begin{itemize}
    \item $x' \leq x \to {x \leq}0 \to {x' \leq}0$
    \item $x' \leq x \to y \leq y' \to z \leq z' \to
      {x \leq}[y+z] \to {x' \leq}[y'+z']$
    \item $x' \leq x \to {x \leq}1 \to {x' \leq}1$
    \item $x' \leq x \to y \leq y' \to z \leq z' \to
      {x \leq}[y*z] \to {x' \leq}[y'*z']$
    \item $\plr{\exists y.~{y \leq}0 \land {x \leq}[y+z]}
      \leftrightarrow x \leq z$
    \item $\plr{\exists z.~{x \leq}[y+z] \land {z \leq}0}
      \leftrightarrow x \leq y$
    \item $\plr{\exists wx.~{wx \leq}[w+x] \land {z \leq}[wx+y]}$
      $\leftrightarrow$

      $\plr{\exists xy.~{z \leq}[w+xy] \land {xy \leq}[x+y]}$
    \item $\exists x.~{x \leq}0 \land \forall x'.~{x' \leq}0 \to x' \leq x$
    \item $\vdots$
    %\item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
    %\item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
    %\item $0 + x = x \land x + 0 = x$
    %\item $(x + y) + z = x + (y + z)$
    %\item $x + y = y + x$
    %\item $1 * x = x \land x * 1 = x$
    %\item $(x * y) * z = x * (y * z)$
    %\item $0 * x = 0 \land x * 0 = 0$
    %\item $(x + y) * z = x * z + y * z$
    %\item $x * (y + z) = x * y + x * z$
  \end{itemize}
\end{definition}

An element of a chosen posemiring $\Ann$ describes the usage restrictions on
a variable.
Therefore, a \emph{vector} of elements from $\Ann$ describes the usage
restrictions of a whole context's worth of variables.
From the posemiring operations of $\Ann$, we derive the standard vector
operations of zero, addition, and multiplication by a scalar.
We can also form the standard basis vectors at any given dimension.

\begin{figure}
  \begin{align*}
    \dot1\,\grR &\coloneqq 1 \\
    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
                       \times T\,\grP \times U\,\grQ \\
    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
                       \times T\,\grP \\
    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
                       \to T\,\grQ \to U\,\grR
  \end{align*}
  \caption{The bunched connectives}
  \label{fig:bunched}
\end{figure}

Operations like renaming and substitution are essentially translations from one
context to another.
When faced with two vector spaces arranged in this way, a natural thing to
consider is the \emph{linear maps} from one space to the other.

### 2021-09-07 tex/cpp21/env.tex
[ new ] start CPP paper

--- 2021-08-17 tex/lenv.tex
@@
+\subsection{Definition}
+
+\begin{definition}[Usage-annotated recursive environment]\label{def:lr-rec-env}
+  A \emph{recursive $\V$-environment} between annotated contexts $\Gamma$ and
+  $\Delta$ is defined by cases on the shape of $\Delta$ (where
+  $\Gamma \env\V_R \Delta$ is the notation for the
+  type of recursive environments for given $\V$, $\Gamma$, and $\Delta$):
+  \begin{itemize}
+    \item There is one environment $\alr{} : \grP\gamma \env\V {\cdot}$
+      whenever $\grP \leq \gr0$.
+    \item For $\rho_l : \grPl\gamma \env\V_R \Delta_l$ and
+      $\rho_r : \grPr\gamma \env\V \Delta_r$, we have an environment
+      $\alr{\rho_l, \rho_r} : \grP\gamma \env\V_R \Delta_l, \Delta_r$ whenever
+      $\grP \leq \grPl + \grPr$.
+    \item For any value $v : \V\,\grPprime\gamma\,A$, we have an environment
+      $\alr{v} : \grP\gamma \env\V_R \gr rA$ whenever
+      $\grP \leq \gr r\grPprime$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Usage-annotated environment]\label{def:lr-env}
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
+  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
+  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
+  $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
+\end{definition}
+
+\subsection{Properties}
+
 \begin{lemma}\label{thm:env-resize}
@@
 \end{example}
-
-% Concatenation is difficult; save to after I've talked about renamings.
-
-% Finally for this section, we give the conditions under which the
-% context-forming operations (empty, concatenation, and singleton) have a
-% functorial action with respect to $\V$-environments.
-%
-% \begin{lemma}
-%   For any $\V$, there is an environment ${\cdot} \env\V {\cdot}$.
-% \end{lemma}
-% \begin{proof}
-%   By \cref{thm:construct-env}, it suffices to show $I\,{\cdot}$, which is
-%   trivially true.
-% \end{proof}

### 2021-09-07 tex/cpp21/quant-framework.tex
[ new ] start CPP paper

\documentclass[sigplan,10pt,anonymous,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage[conor]{agda}
\usepackage{catchfilebetweentags}
\usepackage{cleveref}
\usepackage{ebproof}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{newunicodechar}
\usepackage{stmaryrd}
\usepackage{todonotes}

\input{../macros.tex}
\input{lagda.tex}

\def\Syntaxtex{../../generic-lr/src/latex/Generic/Linear/Syntax.tex}
\def\Interpretationtex%
{../../generic-lr/src/latex/Generic/Linear/Syntax/Interpretation.tex}

\begin{document}

\title{A Framework for Substructural Programming Languages}

\author{James Wood}
\orcid{0000-0002-8080-3350}
\affiliation{%
  \institution{University of Strathclyde}%
  \city{Glasgow}%
  \country{UK}%
}
\email{james.wood.100@strath.ac.uk}

\author{Robert Atkey}
\orcid{0000-0002-4414-5047}
\affiliation{%
  \institution{University of Strathclyde}%
  \city{Glasgow}%
  \country{UK}%
}
\email{robert.atkey@strath.ac.uk}

\maketitle

\section{Introduction}\label{sec:intro}

\section{Vectors over semirings}\label{sec:algebra}
\input{algebra}

\section{Environments}\label{sec:env}
\input{env}

\section{Generic syntax}\label{sec:syntax}
\input{syntax}

\section{Semantics}\label{sec:semantics}

\section{Conclusions}\label{sec:conc}

\end{document}

### 2021-09-07 tex/cpp21/syntax.tex
[ new ] start CPP paper

\subsection{Descriptions}

We capture the form of rules exemplified previously\todo{Previously?} via
\emph{descriptions} of rules.
The key to making these descriptions work is that they only allow syntactic
forms that preserve environments.
These forms are: absence and multiplicity of subterms with the same usage
annotations, absence and multiplicity of subterms with summed usage annotations,
scaling of a subterm, and variable binding.

A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and a type
of conclusion.

\ExecuteMetaData[\Syntaxtex]{Scoped}

A single premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$} constructor.
This allows binding of some variables \AgdaBound{$\Gamma$} (with specified
types and usage annotations) and the specification of a conclusion type
\AgdaBound{A}.
The rest of the constructors are descriptions for the first-order bunched
connectives, and are interpreted directly as such.
\todo{Remove Box}
\todo{Remove instance arguments}

\ExecuteMetaData[\Syntaxtex]{Premises}
\ExecuteMetaData[\Interpretationtex]{semp}

\ExecuteMetaData[\Syntaxtex]{Rule}
\ExecuteMetaData[\Interpretationtex]{semr}

\ExecuteMetaData[\Syntaxtex]{System}
\ExecuteMetaData[\Interpretationtex]{sems}

### 2021-09-14 tex/gen-sem.tex
[ writing ] various thesis additions

@@
+\def\SimpleSem{../agda/latex/SimpleSem.tex}
+
 The traversal \AgdaFunction{trav} from the last section is generic in the sense
@@
 Doing a straight generalisation of the type of \AgdaFunction{trav}, this
-suggests that we want a function with the following type.
+suggests that we want a function with the following type, where
+\AgdaBound{$\C$} is the type family we are targeting.
 
-\missingfigure{Semantic traversal type signature}
+\ExecuteMetaData[\SimpleSem]{semType}
+
+Following the implementation of \AgdaFunction{trav}, we see that
+\AgdaBound{$\C$} will need to support a semantic counterpart of each syntactic
+form (\AgdaInductiveConstructor{var}, \AgdaInductiveConstructor{app}, and
+\AgdaInductiveConstructor{lam}).
+With syntactic kits, we already asked for the field \AgdaField{tm} to interpret
+\AgdaBound{$\V$}-values as terms.
+We rename \AgdaField{tm} to \AgdaField{⟦var⟧} to reflect its role in the
+semantic traversal \AgdaFunction{sem}.
+Now, we will also ask for fields to replace the right-hand side applications of
+the other term constructors.
+For application, we can stick with the obvious thing: we should be able to
+combine a semantic function and its semantic argument to get the semantic
+result.
+
+\ExecuteMetaData[\SimpleSem]{semVarApp}
+
+However, we want to treat binding constructs specially, particularly because
+there are semanticses with no notion of binding.
+We instead provide a function from values to computations that works in any
+\emph{extension} of the current context.
+Keeping \AgdaField{$\swarrow^k$} as before, we get the following semantic
+replacement for \AgdaRecord{Kit}.
+
+\ExecuteMetaData[\SimpleSem]{SemanticsExplicit}
+
+With the aim of abstracting away from explicit contexts, bringing us closer to
+natural deduction, we can use some new notation to rephrase these requirements.
+We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
+One of the basic connectives in this setting is the \emph{pointwise} arrow
+\AgdaFunction{\_$\Rightarrow$\_}, which acts in
+\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
+like the non-dependent arrow does in \AgdaPrimitive{Set}.
+Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
+embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
+to quantify over \emph{all} contexts.
+Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
+encapsulating the pattern of considering arbitrary \emph{extensions} of a
+context.
+To facilitate working in this point-free setting, I give infix versions of
+the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
+\AgdaFunction{\_$\V\vdash$\_} and \AgdaFunction{\_$\C\vdash$\_}).
+The principal use of these aliases is to fill the right argument with a type
+(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
+a context given through the point-free machinery.
+
+\ExecuteMetaData[\SimpleSem]{Circle}
+
+\ExecuteMetaData[\SimpleSem]{SemanticsCircle}
+
+To illustrate this definition, I will discuss a syntactic traversal ---
+renaming --- and a semantic traversal --- a standard $\mathrm{Set}$ semantics.
+
+For the renaming semantics, as with the renaming kit, we specify that
+environments hold variables (\AgdaDatatype{\_$\ni$\_}) and show that variables
+satisfy the required form of weakening (\AgdaFunction{$\swarrow^v$}).
+Meanwhile, whereas all syntactic kits target terms
+(\AgdaDatatype{\_$\vdash$\_}), with a semantic traversal we must specify the
+target.
+The fields \AgdaField{$\llbracket$var$\rrbracket$} and
+\AgdaField{$\llbracket$app$\rrbracket$} follow straightforwardly, with variables
+embedding into terms and a pair of terms of the right types giving an
+application term in the same context, via the relevant constructors.
+For the \AgdaField{$\llbracket$lam$\rrbracket$} case, we are given
+\ExecuteMetaData[\SimpleSem]{bRenSemCircle}, and after producing a
+\AgdaInductiveConstructor{lam}, are left needing a term in
+\ExecuteMetaData[\SimpleSem]{resRenSemCircle}.
+That the type of \AgdaBound{b} is wrapped in \AgdaFunction{$\bigcirc$} gives
+us the ability to use \AgdaBound{b} in the extended context
+\ExecuteMetaData[\SimpleKits]{GA}.
+In particular, we point at the new variable to yield the desired term in the
+same context.
+
+\ExecuteMetaData[\SimpleSem]{RenSemCircle}
+
+To produce a $\mathrm{Set}$ semantics, we shift from targeting terms to
+targeting the interpretation of terms.
+In particular, \ExecuteMetaData[\SimpleSem]{semGA}\ is the type of functions
+from the interpretation of \AgdaBound{$\Gamma$} to the interpretation of
+\AgdaBound{A}.
+The interpretation of a type is defined as usual, by recursion on the structure
+of the type.
+The interpretation of a context is the interpretation for each of its types.
+We still have environments storing variables, which delays the interpretation of
+variables to the \AgdaField{$\llbracket$var$\rrbracket$} case and allows newly
+bound variables to be referred to directly as variables, rather than fetching
+them up-front from an environment of interpretations.
+In the \AgdaField{$\llbracket$app$\rrbracket$} case, we have
+\ExecuteMetaData[\SimpleSem]{mSetSemCircle}\ and
+\ExecuteMetaData[\SimpleSem]{nSetSemCircle}, and combine them in the usual way
+by distributing the interpretation of the context
+\ExecuteMetaData[\SimpleSem]{gammaSetSemCircle}.
+The \AgdaField{$\llbracket$lam$\rrbracket$} case involves the same placement of
+the new variable into the environment as in \AgdaFunction{RenSem}.
+Finally, we get the function \AgdaFunction{set} from terms to their
+interpretations by passing in the identity $\ni$-environment \AgdaFunction{id}.
+
+\ExecuteMetaData[\SimpleSem]{SetSemCircle}
+
+The definition of \AgdaRecord{Semantics} above essentially enforces that the
+term being traversed and the result of the traversal share the same binding
+structure.
+Concretely, \AgdaInductiveConstructor{lam} is the only case where we can bind
+new variables, and at that point we must do exactly one binding.
+This is fine for renaming and substitution, which preserve the binding
+structure, and also for a standard denotational semantics, which is sufficiently
+abstracted from binding.
+However, if we want to do other syntactic translations --- for example,
+converting from a syntax with $n$-ary functions to a syntax with only unary
+Curried functions --- it would be useful to allow more choices when going under
+a binder.
+To this end, I replace the one-step binding modality \AgdaFunction{$\bigcirc$}
+by the all-possible-renamings modality \AgdaFunction{$\Box$}.
+\AgdaFunction{$\Box$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaBound{$\Gamma$}
+states that \AgdaBound{T} holds not only at \AgdaBound{$\Gamma$}, but also at
+any context \AgdaBound{$\Gamma^+$} containing \AgdaBound{$\Gamma$} (including
+\ExecuteMetaData[\SimpleKits]{GD}\ for any \AgdaBound{$\Delta$}).
+
+\ExecuteMetaData[\SimpleSem]{Box}
+
+As well as the flexibility in binding structure, the \AgdaFunction{$\Box$}
+modality allows us to use the somewhat more well behaved and standard
+relation of renaming, rather than strict context extension.
+The resulting definition of \AgdaRecord{Semantics} is as follows, and is
+simply a version of the previous definition where \AgdaFunction{$\bigcirc$}
+has been replaced by \AgdaFunction{$\Box$}.
+It will become apparent when we implement the traversal \AgdaFunction{sem} why
+the first field also changes to include a \AgdaFunction{$\Box$}.
+
+\ExecuteMetaData[\SimpleSem]{Semantics}
+
+Writing a \AgdaFunction{$\Box$}-based semantics is very similar to writing a
+\AgdaFunction{$\bigcirc$}-based semantics, so I will only give one further
+example.
+I generalise the renaming example to derive a semantic traversal from any
+syntactic traversal.
+We need a slightly modified definition of \AgdaRecord{Kit} to provide
+\emph{renaming} of \AgdaBound{$\V$}-values, rather than just extension.
+
+\ExecuteMetaData[\SimpleSem]{Kit}
+
+The interesting feature of the corresponding \AgdaRecord{Semantics} is that
+we now pass \AgdaBound{b} the renaming \AgdaFunction{$\swarrow^v$}, projecting
+the original context \AgdaBound{$\Gamma$} out of the
+\AgdaInductiveConstructor{lam}-extended context
+\ExecuteMetaData[\SimpleKits]{GA}.
+
+\ExecuteMetaData[\SimpleSem]{kit}

### 2021-09-14 tex/linear.tex
[ writing ] various thesis additions

\chapter{Linearity}\label{sec:linearity}

\section{Intuitionistic linear logic}
\input{ill.tex}
\section{DILL}
DILL provides a syntax in which $\oc$ is characterised by a universal
property.
There is no need for any rule to depend on the entire context and rebind
variables (as done by BBdPH), so is more ND-like.
DILL makes $\oc$ behave more like a positive type former (see the deep
identity function $\oc A \vdash \oc A$).
DILL has been extended to dependent types: Vakar15.

The idea of split contexts can also be applied to left-/right-rule systems.
(Define this DILL variant.)

\section{Mechanisation survey}

### 2021-09-14 tex/lrkits.tex
[ writing ] various thesis additions

@@
     \item For $\rho_l : \grPl\gamma \env\V_R \Delta_l$ and
-      $\rho_r : \grPr\gamma \env\V \Delta_r$, we have an environment
+      $\rho_r : \grPr\gamma \env\V_R \Delta_r$, we have an environment
       $\alr{\rho_l, \rho_r} : \grP\gamma \env\V_R \Delta_l, \Delta_r$ whenever

### 2021-09-14 tex/part.tex
[ refactor ] put sec 5 in an \include
[ writing ] various thesis additions

--- 2021-07-12 tex/semirings.tex
@@
-Addition gives the structural rules.
-Multiplication gives modalities.
-
-Compare with other work.
-About here, I can talk about Granule-style $\otimes$-pattern matching.
-
-\subsection{Partiality (or not)} \todo{Appendix?}
-
-As we have seen \todo{Have we?}, the way additive and multiplicative rules are
+As we have seen, the way additive and multiplicative rules are
 realised algebraically is related to models of separation logic.

### 2021-09-14 tex/prelim.tex
[ writing ] various thesis additions

\chapter{Mathematical preliminaries}

Categories, (co)products, monoidal categories, string diagrams (in Rel),
monoid objects.
Examples: Set, Rel, partial functions, commutative monoids.

### 2021-09-14 tex/semirings.tex
[ writing ] various thesis additions

@@
-\chapter{Usage restriction via semirings}
+\chapter{Usage restriction via semirings}\label{sec:semirings}
 
-Addition gives the structural rules.
-Multiplication gives modalities.
+The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
+make crucial use of \emph{weakening} --- the fact that if we have
+$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
+We use this property to update environments as they go under binders.
+However, as we saw in \cref{sec:linearity}, there are interesting calculi in
+which general weakening does not hold.
+As such, one of the aims of this chapter will be to find a form of weakening
+applicable to variables of any type, while essentially retaining linearity
+(as opposed to affinity).
 
-Compare with other work.
-About here, I can talk about Granule-style $\otimes$-pattern matching.
+As explained by McBride~\cite{McBride16}, the first insight is to, instead of
+removing variables from the context of certain subterms, add an annotation to
+free variables saying whether they are to be used or not.
+We may use an annotation $\gr0$ on variables that are not to be used, and an
+annotation $\gr1$ on variables that are to be used.
+This convention lets us transcribe the usual $\otimes$-introduction rule
+(left) as a rule with usage annotations (right).
+I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context is
+now a list of types together with their usage annotations.
+\todo{Reword: what's capital $\Gamma$}
+If $\gamma = x_1 : A_1, \ldots, x_m : A_m$, then
+$\gr r\gamma = \gr rx_1 : A_1, \ldots, \gr rx_m : A_m$.
+
+\[
+  \begin{prooftree}
+    \hypo{\Gamma \vdash A}
+    \hypo{\Delta \vdash B}
+    \infer2{\Gamma, \Delta \vdash A \otimes B}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \hypo{\gr1\gamma, \gr0\delta \vdash A}
+    \hypo{\gr0\gamma, \gr1\delta \vdash B}
+    \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
+  \end{prooftree}
+\]
+
+What the existence of $\gr0$ has bought us is that variables never go out of
+scope in subterms; rather, we lose the ability to use certain variables.
+Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
+$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
+variables from $\delta$.
+A rigorous proof of weakening for terms will come in \cref{sec:lrsub}.
+
+If we follow the DILL style of variable management, there are not just the two
+states \emph{used} ($\gr1$) and \emph{not used} ($\gr0$), but also
+\emph{usable unrestrictedly}.
+If we assign unrestricted (or \emph{intuitionistic}) variables an annotation
+$\gr\omega$, we can make the following transcription of the DILL
+$\otimes$-introduction rule.
+
+\[
+  \begin{prooftree}
+    \hypo{\Theta; \Gamma \vdash A}
+    \hypo{\Theta; \Delta \vdash B}
+    \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
+    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
+    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
+  \end{prooftree}
+\]
+
+To conceptualise the criteria on the usage annotations involved in this rule,
+I introduce a formal $+$ operation on usage annotations.
+The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
+$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
+The capital calligraphic letters $\grP$, $\grQ$, and $\grR$ stand for all of
+the usage annotations in a given context, with $\grR\gamma$ standing for
+$\grR_1x_1 : \gamma_1, \ldots, \grR_mx_m : \gamma_m$.
+Addition lifts pointwise to these vectors of usage annotations.
+It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
+rule on the right now accepts $\gr0$-annotated variables in its conclusion,
+which is essential for weakening to be admissible.
+
+\[
+  \begin{prooftree}
+    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
+    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
+    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \hypo{\grR = \grP + \grQ}
+    \hypo{\grP\gamma \vdash A}
+    \hypo{\grQ\gamma \vdash B}
+    \infer3{\grR\gamma \vdash A \otimes B}
+  \end{prooftree}
+\]
+
+Some other transcriptions:
+
+\[
+  \begin{prooftree}
+    \infer0{\Theta; A \vdash A}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \hypo{\grR \leq \bra x}
+    \hypo{\gamma_x = A}
+    \infer2{\grR\gamma \vdash A}
+  \end{prooftree}
+\]
+
+\[
+  \begin{prooftree}
+    \infer0{\Theta, A; {\cdot} \vdash A}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
+  \end{prooftree}
+  \quad\rightsquigarrow\quad
+  \begin{prooftree}
+    \hypo{\grR \leq \bra x}
+    \hypo{\gamma_x = A}
+    \infer2{\grR\gamma \vdash A}
+  \end{prooftree}
+\]
+
+%In \cref{sec:simple}, we saw that the logical rules of simply typed
+%$\lambda$-calculus can be described in terms of three basic premise combinators:
+%$\dot1$, standing for no premises; $\dottimes$, allowing for multiple premises
+%in the same context; and $\Theta \vdash A$, requiring a subterm of type $A$
+%having bound the extra variables from context $\Theta$.
+%However, we remember from \cref{sec:linearity} that in a substructural setting,
+%we do not always want to copy assumptions for use in all subterms.
+%This motivates me to introduce the additional premise combinators $I^*$, $\sep$,
+%and $\cdot$ in \cref{sec:lnd}, allowing for the modes of usage exhibited in
+%the introduction rules for $I$, $\otimes$, and $\oc$, respectively.
+
+In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
+appeared previously in my work with Atkey~\cite{WA20}, and can be seen as a
+simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
+The idea of \name{} is to augment simply typed $\lambda$-calculus with
+annotations on free variables, and give enough types to manipulate these
+annotations.
+I use \cref{sec:lnd} to introduce some notation that allows us to restate the
+typing rules of \name{} so as to not mention contexts explicitly.
+The first problem I tackle with this new syntax is proving admissibility of
+substitution.
+I follow the methodology from \cref{sec:kits}, which
 
@@
 \input{lrsub.tex}
-\section{Recursion}\label{sec:rec}
+\section{Adding recursion to \name{}}\label{sec:rec}
 \input{rec.tex}

### 2021-09-14 tex/simple.tex
[ writing ] various thesis additions

\chapter{Mechanisation of simple types}\label{sec:simple}

\section{Term representation}
\input{terms.tex}
\section{Renaming and substitution}\label{sec:kits}
\input{kits.tex}
\section{Generic semantics}
\input{gen-sem.tex}
\section{Generic syntax}
\section{Natural deduction vs sequent calculus}
\input{ndsc.tex}

### 2021-09-14 tex/wrel.tex
[ writing ] various thesis additions

\chapter{Semantics in worldly relations}\label{sec:wrel}

### 2021-09-14 tex/cpp21/env.tex
[ writing ] CPP paper

--- 2021-08-17 tex/lenv.tex
@@
-\begin{lemma}\label{thm:env-resize}
-  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
-  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
-  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
-  with the same linear map and action on variables.
+This section is a development of our previous work~\cite{WA20}. \todo{Blind}
+
+Let us write $\sdtstile{}\V$ as the infix version of $\V$.
+That is, $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$.
+The idea of a $\V$-environment is to extend the notion of $\V$-values from
+types to contexts.
+Specifically, where the judgement $\Gamma \sdtstile{}\V A$ says that we
+have a $\V$-value of type $A$ in context $\Gamma$, the judgement
+$\Gamma \env\V \Delta$ says that we have a $\V$-value for each entry in
+$\Delta$ (at the specified multiplicity), all in context $\Gamma$.
+For our purposes, it is essential that the ``all'' in the previous sentence is
+multiplicative --- we want to split the usage annotations of $\Gamma$ up into
+parts such that each part supports one $\V$-value.
+
+We have seen two kinds of values already: $\sqni$-values are variables and
+$\vdash$-values are terms.
+The corresponding environments are also standard concepts: $\sqni$-environments
+are \emph{simultaneous renamings} and $\vdash$-environments are
+\emph{simultaneous substitutions}.
+
+\subsection{Definition}
+
+\begin{definition}\label{def:lr-env}
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
+  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
+  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
+  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
+  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
+  $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.
+\end{definition}
+
+\begin{example}
+  We can form the identity renaming on a two-variable context.
+  \[
+    \id : \plr{\gr rA, \gr sB} \env\sqni \plr{\gr rA, \gr sB}
+  \]
+  Linear map $\gr\Psi$ is the identity map, clearly satisfying
+  \(
+    \begin{pmatrix} \gr r & \gr s \end{pmatrix} \leq
+    \begin{pmatrix} \gr r & \gr s \end{pmatrix}\gr\Psi
+  \).
+  When considering values, the fact that $\grPprime \leq \grQprime\gr\Psi$
+  reduces to $\grPprime \leq \grQprime$.
+  The two cases to consider are when $\grQprime\delta \sqni A$ and when
+  $\grQprime\delta \sqni B$.
+  In the first case, $\grPprime \leq \grQprime \leq
+  \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, so we have
+  $\grPprime\plr{A, B} \sqni A$.
+  The second case follows symmetrically.
+\end{example}
+
+\begin{example}
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
+  addition and multiplication.
+  There is an environment (substitution) of type
+  \[
+    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
+    \plr{\gr1B, \gr2C}.
+  \]
+  Linear map $\gr\Psi$ is given by the matrix
+  \(
+    \begin{pmatrix}
+      \gr0 & \gr0 & \gr1 \\
+      \gr0 & \gr1 & \gr1
+    \end{pmatrix}
+  \),
+  noticing that $\gr1$ times the first row plus $\gr2$ times the second gives
+  the original $\grP$.
+  For $\grQprime = \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, we have
+  $\grPprime = \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$, and thus
+  $\grPprime\gamma \vdash z : B$.
+  For $\grQprime = \begin{pmatrix} \gr0 & \gr1 \end{pmatrix}$, we have
+  $\grPprime = \begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$, and thus
+  $\grPprime\gamma \vdash y\,z : C$.
+\end{example}
+
+As a mnemonic, one may use notation like the following to see what values are
+needed in the environment.
+\[
+  \begin{pmatrix}
+    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
+    \gr0A & \gr1\plr{B \multimap C} & \gr1B
+  \end{pmatrix}
+  \begin{matrix}
+    {} \vdash B \\
+    {} \vdash C
+  \end{matrix}
+\]
+This notation assumes that the notion of value supports subusaging, which is
+always the case when we are using environments for traversals.
+
+\begin{example}
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
+  addition and multiplication.
+  There is an environment (renaming) of type
+  \[
+    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D} \env\sqni
+    \plr{\gr1C, \gr2A, \gr4A}.
+  \]
+  Linear map $\gr\Psi$ is given by the matrix
+  \(
+    \begin{pmatrix}
+      \gr0 & \gr0 & \gr1 & \gr0 \\
+      \gr1 & \gr0 & \gr0 & \gr0 \\
+      \gr1 & \gr0 & \gr0 & \gr0
+    \end{pmatrix}
+  \),
+  which we can check satisfies the required inequality.
+  The values are given by
+  \begin{align*}
+    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
+    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
+    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A.
+  \end{align*}
+
+  We can read from the columns of the matrix what happened to each of the
+  variables in $\Gamma$.
+  The first column, corresponding to variable $\gr6a : A$, contains two $\gr1$s
+  because it has been duplicated (via contraction).
+  Meanwhile, the second and fourth columns are all $\gr0$ because variables
+  $b$ and $d$ have been discarded (via weakening).
+  The third column contains one $\gr1$ because $c$ is used once.
+  This $\gr1$ appears above the $\gr1$s to its left because $c$ has been
+  permuted (via exchange) past $a$.
+  Each of the rows in the matrix is a basis vector because variables can only
+  be formed in contexts with basis annotations or less.
+\end{example}
+
+\subsection{Properties}
+
+\begin{lemma}
+  \cref{def:lr-env} is sound and complete for the following syntax.
+  \begin{align*}
+    \begin{prooftree}[comb,center=false]
+      \hypo{I^*}
+      \infer1{\alr{} : {} \env\V {\cdot}}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb,center=false]
+      \hypo{\rho : {} \env\V \Delta_l}
+      \hypo{\sep}
+      \hypo{\sigma : {} \env\V \Delta_r}
+      \infer3{\alr{\rho,\sigma} : {} \env\V \Delta_l, \Delta_r}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb,center=false]
+      \hypo{\gr r\cdot\plr{M : {} \sdtstile{}\V A}}
+      \infer1{\alr{M} : {} \env\V \gr rA}
+    \end{prooftree}
+  \end{align*}
 \end{lemma}
 \begin{proof}
-  The only part of the definition of an environment dependent on $\grP$ or
-  $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
-  replace for $\grPprime$ and $\grQprime$.
+  This replaces \cref{thm:construct-env}.
 \end{proof}
+
+\begin{example}
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
+  addition and multiplication.
+  There is an environment (substitution) of type
+  \[
+    \alr{\alr{z},\alr{y\,z}} :
+    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
+    \plr{\gr1B, \gr2C}.
+  \]
+  We rely on the observations that
+  $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
+  \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}
+  + \begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ and, on the right, that
+  $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix} =
+  \gr2\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
+  Then, we have $\gr0x : A, \gr0y : B \multimap C, \gr1z : B \vdash z : B$ and
+  $\gr0x : A, \gr1y : B \multimap C, \gr1z : B \vdash y\,z : C$.
+\end{example}
 
@@
 
+\begin{lemma}\label{thm:env-resize}
+  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
+  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
+  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
+  with the same linear map and action on variables.
+\end{lemma}
+\begin{proof}
+  The only part of the definition of an environment dependent on $\grP$ or
+  $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
+  replace for $\grPprime$ and $\grQprime$.
+\end{proof}
+
 One of the primary test cases for environments is simultaneous substitution,
@@
 
-Finally, I will also take the opportunity to give the bind lemma, allowing
+Finally, I will also take the opportunity to give the extend lemma, allowing
 environments to incorporate newly bound variables.
-In the intuitionistic case, the bind lemma had two requirements on $\V$: $\V$
+\todo{Motivate}
+In the intuitionistic case, the extend lemma had two requirements on $\V$: $\V$
 admits weakening and we can map variables into $\V$-values.
@@
 
-\begin{lemma}[bind]\label{thm:lr-bind}
+\begin{lemma}[extend]\label{thm:lr-bind}
   Given functions
@@
 \end{proof}
+
+\subsection{Identity and composition}
 

### 2021-09-14 tex/cpp21/quant-framework.tex
[ writing ] CPP paper

@@
 \usepackage{cleveref}
+\usepackage{cmll}
 \usepackage{ebproof}
@@
 \usepackage{stmaryrd}
+\setlength{\marginparwidth}{2cm}
 \usepackage{todonotes}
+\usepackage{turnstile}
 
@@
 
-\def\Syntaxtex{../../generic-lr/src/latex/Generic/Linear/Syntax.tex}
-\def\Interpretationtex%
-{../../generic-lr/src/latex/Generic/Linear/Syntax/Interpretation.tex}
+\def\genericlr{../../generic-lr/src/processed-latex}
+\def\Syntaxtex{\genericlr/Generic/Linear/Syntax.tex}
+\def\Interpretationtex{\genericlr/Generic/Linear/Syntax/Interpretation.tex}
+\def\Maptex{\genericlr/Generic/Linear/Syntax/Interpretation/Map.tex}
+\def\Termtex{\genericlr/Generic/Linear/Syntax/Term.tex}
+\def\Semanticstex{\genericlr/Generic/Linear/Semantics.tex}
+\def\Renamingtex{\genericlr/Generic/Linear/Renaming.tex}
+\def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
 
@@
 
-\title{A Framework for Substructural Programming Languages}
+\title{A Framework for Substructural Type Systems}
 
@@
 
+\keywords{substructural, linear, metatheory, type system, mechanisation}
+
+\begin{abstract}
+  We present a framework for mechanising the metatheory of certain substructural
+  type systems.
+\end{abstract}
+
 \maketitle
@@
 \section{Introduction}\label{sec:intro}
+\input{intro}
 
@@
 
+\section{Generic syntax}\label{sec:syntax}
+\input{syntax}
+
 \section{Environments}\label{sec:env}
@@
 
-\section{Generic syntax}\label{sec:syntax}
-\input{syntax}
-
 \section{Semantics}\label{sec:semantics}
+\input{semantics}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
+\bibliography{quant-framework}
+
 \end{document}

### 2021-09-14 tex/cpp21/semantics.tex
[ writing ] CPP paper

Our aim with the following traversals is to derive a recursion scheme for
terms which abstracts over some of the details of variable binding and
changes to usage annotations.
We follow the structure given by Allais et.\ al.~\cite{AACMM20} closely, with
the main difference having come already in the syntax and environments.

\subsection{Renameability}

{\color{red}(Maybe this goes in the Environment section.)}
We saw in \cref{sec:env} that our notion of \emph{environment} specialises to
\emph{simultaneous renaming} by setting $\V$ to be $\sqni$.

\ExecuteMetaData[\Renamingtex]{Ren}
\ExecuteMetaData[\Renamingtex]{Boxr}
\ExecuteMetaData[\Renamingtex]{Renameable}

\subsection{Syntactic traversal}

\subsection{Semantic traversal}

\ExecuteMetaData[\Semanticstex]{Kripke}
\ExecuteMetaData[\Semanticstex]{Semantics}

### 2021-09-14 tex/cpp21/syntax.tex
[ writing ] CPP paper

@@
+\subsection{A prototypical type system}
+
+\begin{figure}
+  \begin{align*}
+    \begin{prooftree}
+      \hypo{\grR \leq \bra x}
+      \hypo{\gamma_x = A}
+      \infer2[Var]{\grR\gamma \vdash A}
+    \end{prooftree}
+  \end{align*}
+  \begin{align*}
+    \begin{prooftree}
+      \hypo{\grR\gamma, \gr1A \vdash B}
+      \infer1[$\multimap$-I]{\grR\gamma \vdash A \multimap B}
+    \end{prooftree}
+    &&
+    \begin{prooftree}
+      \hypo{\grR \leq \grP + \grQ}
+      \hypo{\grP\gamma &\vdash A \multimap B}
+      \infer[no rule]1{\grQ\gamma &\vdash A \phantom{{} \multimap B}}
+      \infer2[$\multimap$-E]{\grR\gamma \vdash B}
+    \end{prooftree}
+  \end{align*}
+  \begin{align*}
+    \begin{prooftree}
+      \hypo{\grR\gamma \vdash A_i}
+      \infer1[$\oplus$-I$_i$]{\grR\gamma \vdash A_0 \oplus A_1}
+    \end{prooftree}
+    &&
+    \begin{prooftree}
+      \hypo{\grR \leq \grP + \grQ}
+      \infer[no rule]1{\grP\gamma \vdash A \oplus B}
+      \hypo{\grQ\gamma, \gr1A &\vdash C}
+      \infer[no rule]1{\grQ\gamma, \gr1B &\vdash C}
+      \infer2[$\oplus$-E]{\grR\gamma \vdash C}
+    \end{prooftree}
+  \end{align*}
+  \begin{align*}
+    \begin{prooftree}
+      \hypo{\grR \leq \gr r\grP}
+      \hypo{\grP\gamma \vdash A}
+      \infer2[$\oc_{\gr r}$-I]{\grR\gamma \vdash \oc_{\gr r} A}
+    \end{prooftree}
+    &&
+    \begin{prooftree}
+      \hypo{\grR \leq \grP + \grQ}
+      \hypo{\grP\gamma \vdash \oc_{\gr r} A}
+      \infer[no rule]1{\grQ\gamma, \gr rA \vdash C}
+      \infer2[$\oc_{\gr r}$-E]{\grR\gamma \vdash C}
+    \end{prooftree}
+  \end{align*}
+  \caption{A fragment of \name{}}\label{fig:lr}
+\end{figure}
+
+\begin{figure}
+  \begin{align*}
+    \begin{prooftree}[comb]
+      \hypo{\sqni A}
+      \infer1[Var]{\vdash A}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb]
+      \hypo{\gr1A \vdash B}
+      \infer1[$\multimap$-I]{\vdash A \multimap B}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb]
+      \hypo{\vdash A \multimap B}
+      \hypo{\sep}
+      \hypo{\vdash A}
+      \infer3[$\multimap$-E]{\vdash B}
+    \end{prooftree}
+  \end{align*}
+  \begin{align*}
+    \begin{prooftree}[comb]
+      \hypo{\vdash A_i}
+      \infer1[$\oplus$-I$_i$]{\vdash A_0 \oplus A_1}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb]
+      \hypo{\vdash A \oplus B}
+      \hypo{\sep}
+      \hypo{(\gr1A \vdash C}
+      \hypo{\dottimes}
+      \hypo{\gr1B \vdash C)}
+      \infer5[$\oplus$-E]{\vdash C}
+    \end{prooftree}
+  \end{align*}
+  \begin{align*}
+    \begin{prooftree}[comb]
+      \hypo{\gr r \cdot \plr{\vdash A}}
+      \infer1[$\oc_{\gr r}$-I]{\vdash \oc_{\gr r} A}
+    \end{prooftree}
+    &&
+    \begin{prooftree}[comb]
+      \hypo{\vdash \oc_{\gr r} A}
+      \hypo{\sep}
+      \hypo{\gr rA \vdash C}
+      \infer3[$\oc_{\gr r}$-E]{\vdash C}
+    \end{prooftree}
+  \end{align*}
+  \caption{\cref{fig:lr} restated in terms of bunched combinators}
+  \label{fig:lr-comb}
+\end{figure}
+
 \subsection{Descriptions}
@@
 
-\ExecuteMetaData[\Syntaxtex]{Scoped}
-
 A single premise is introduced by the
 \AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$} constructor.
-This allows binding of some variables \AgdaBound{$\Gamma$} (with specified
+This allows binding of some variables \AgdaBound{$\Delta$} (with specified
 types and usage annotations) and the specification of a conclusion type
@@
 connectives, and are interpreted directly as such.
-\todo{Remove Box}
-\todo{Remove instance arguments}
 
 \ExecuteMetaData[\Syntaxtex]{Premises}
+\ExecuteMetaData[\Syntaxtex]{Rule}
+\ExecuteMetaData[\Syntaxtex]{System}
+
+As an example, we transcribe the system defined in \cref{fig:lr-comb} into a
+description.
+We first give the set of types in a standard way (together with base type
+\AgdaInductiveConstructor{$\iota$}).
+
+\ExecuteMetaData[\PaperExamplestex]{Ty}
+
+Then, we give the set of labels.
+There is one label for each instantiation of a logical rule, but the labels
+contain no further information about subterms or restrictions on the context.
+
+\ExecuteMetaData[\PaperExamplestex]{qlR}
+
+Finally, we specify, for each label, the rest of the information about that
+rule.
+We do not have a variable rule, because the variable rule is implicit to any
+type system.
+
+\ExecuteMetaData[\PaperExamplestex]{lR}
+
+The next thing we want to do is to build terms in the described type system.
+
+\ExecuteMetaData[\Syntaxtex]{Scoped}
+
+To specify the meaning of descriptions, we assume some
+\ExecuteMetaData[\Interpretationtex]{X}, over which we form one layer of syntax.
+The first argument to \AgdaBound{X} is the new variables bound by this layer
+of syntax, as exemplified in the first clause of
+\AgdaFunction{$\llbracket$\_$\rrbracket$p}.
+The second argument is the context containing the variables being carried over
+from the previous layer.
+Notice that this is not, in general, the same as the context from the previous
+layer, because the usage annotations may have been changed by connectives like
+\AgdaInductiveConstructor{\_`$*$\_} and \AgdaInductiveConstructor{\_`$\cdot$\_}.
+The third argument is the type of subterm required.
+
+With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
+the rest are simply interpretations into bunched combinators.
+
 \ExecuteMetaData[\Interpretationtex]{semp}
 
-\ExecuteMetaData[\Syntaxtex]{Rule}
+The interpretation of a \AgdaRecord{Rule} is to check that the rule targets
+the desired type and then interpret the rules premises \AgdaBound{ps}.
+Notice that the interpretation of the premises is independent of the conclusion
+of the rule, which accounts for the difference in type between
+\AgdaFunction{$\llbracket$\_$\rrbracket$p} and
+\AgdaFunction{$\llbracket$\_$\rrbracket$r}.
+
 \ExecuteMetaData[\Interpretationtex]{semr}
 
-\ExecuteMetaData[\Syntaxtex]{System}
+The interpretation of a \AgdaRecord{System} is to choose a rule label
+\AgdaBound{l} from \AgdaBound{L} and interpret the corresponding rule
+\AgdaBound{rs}\AgdaSpace{}\AgdaBound{l} in the same context and for the same
+conclusion.
+
 \ExecuteMetaData[\Interpretationtex]{sems}
+
+The most obvious way to make such an \AgdaBound{X} is to use some existing
+\AgdaFunction{Scoped} on an extended context.
+This is what \AgdaFunction{Scope} does --- take the new variables
+\AgdaBound{$\Delta$}, concatenate them onto the existing context
+\AgdaBound{$\Gamma$}, and pass the extended context onto the judgement
+\AgdaBound{T}.
+
+\ExecuteMetaData[\Syntaxtex]{Scope}
+
+It is this method we use to deal with new variables in syntax.
+Terms resemble the free monad over a layer-of-syntax functor, though that
+picture is complicated by variable binding.
+A term is either a variable or a use of a logical rule together with terms
+for each of the required subterms.
+
+\ExecuteMetaData[\Termtex]{Term}
+
+\missingfigure{Example term}
+
+A layer of syntax supports the following functorial action.
+{\color{red}(Maybe this has to go later, because it talks about linear maps.)}
+
+\ExecuteMetaData[\Maptex]{map-s-type}

### 2021-09-17 tex/cpp21/env.tex
[ writing ] more CPP paper

@@
-This section is a development of our previous work~\cite{WA20}. \todo{Blind}
+This section is a development of the work of \citet{WA20}.
 
@@
 
-\begin{lemma}[Identity environment]
+\begin{lemma}[Identity environment]\label{thm:env-id}
   Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
@@
 
-\begin{corollary}[Composition of environments]
+\begin{corollary}[Composition of environments]\label{thm:env-comp}
   Given a function

### 2021-09-17 tex/cpp21/examples.tex
[ writing ] more CPP paper

\subsection{Renaming and substitution}

In an unpublished note, \citet{McBride05} gives a parametrised traversal
yielding homomorphisms of syntax.
The parameters are collected in the record \AgdaRecord{Kit}.
We make a minor change to the original presentation, where instead of our
\AgdaField{ren\textasciicircum{}$\V$} field, \citeauthor{McBride05} has the
field \AgdaField{wk} allowing only context extensions.
As for the other two fields, \AgdaField{vr} allows us to map variables to
$\V$-values, so as to put newly bound variables in environments; and
\AgdaField{tm} allows us to extract terms from $\V$-values, as required when
we use the environment to evaluate a free variable.

\ExecuteMetaData[\Syntactictex]{Kit}

Where \citeauthor{McBride05} gave the traversal explicitly, we go via our
generic semantic traversal.
The first two fields of \AgdaRecord{Semantics} derive directly from fields of
\AgdaRecord{Kit}.
Meanwhile, to handle term constructors, we first \AgdaFunction{reify} to get a
collection of traversed subterms, and then use \AgdaInductiveConstructor{`con}
to assemble these subterms into a similarly shaped syntactic form as we started
with.
The \AgdaField{vr} field is used implicitly in \AgdaFunction{reify}, as it is
used to show that $\V$-identity environments exist.

\ExecuteMetaData[\Syntactictex]{kit-to-sem}

\ExecuteMetaData[\Syntactictex]{Ren-Kit}
\ExecuteMetaData[\Syntactictex]{Sub-Kit}

\subsection{A denotational semantics}

\subsection{A usage elaborator}

### 2021-09-17 tex/cpp21/quant-framework.tex
[ writing ] more CPP paper

@@
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
+
+\citestyle{acmauthoryear}
+\bibliographystyle{ACM-Reference-Format}
 
@@
 \def\Semanticstex{\genericlr/Generic/Linear/Semantics.tex}
+\def\Syntactictex{\genericlr/Generic/Linear/Semantics/Syntactic.tex}
 \def\Renamingtex{\genericlr/Generic/Linear/Renaming.tex}
 \def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
+\def\Snippetstex{../../agda/processed-latex/Snippets.tex}
 
@@
 \begin{abstract}
-  We present a framework for mechanising the metatheory of certain substructural
-  type systems.
+  Mechanisation of programming language research is of growing interest, and
+  the act of mechanising type systems and their metatheory is generally becoming
+  easier as new techniques are invented.
+  However, state-of-the-art techniques mostly rely on \emph{structurality} of
+  the type system --- that weakening, contraction, and exchange are admissible
+  and variables can be used unrestrictedly once assumed.
+  Linear logic, and many related subsequent systems, provide motivations for
+  breaking some of these assumptions.
+
+  We present a framework for mechanising the metatheory of certain
+  substructural type systems, in a style resembling mechanised metatheory of
+  structural type systems.
+  The framework covers a wide range of simply typed syntaxes with semiring
+  usage annotations, via a metasyntax of typing rules.
+  The metasyntax for the premises of a typing rule is related to bunched logic,
+  featuring both sharing and separating conjunction, roughly corresponding to
+  the additive and multiplicative features of linear logic.
+  The bunched flavour is carried over into the semantics, together with the use
+  of basic linear algebra constructs.
+  For example, \emph{environments} are presented equivalently as values
+  accumulated via separating conjunction, and as functions from variables to
+  values supported by linear maps.
+  Producing a generic semantic traversal has us combine environments with a
+  separating implication, producing a Kripke function space of the form
+  $\Box(A \wand B)$.
+  From the generic semantic traversal, we derive totally generic renaming and
+  substitution operations, a specific denotational semantics, and a
+  syntax-generic \emph{usage elaborator} which greatly facilitates writing
+  concrete terms.
 \end{abstract}
@@
 
+\section{Example traversals}\label{sec:examples}
+\input{examples}
+
 \section{Conclusions}\label{sec:conc}
 
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
 \bibliography{quant-framework}

### 2021-09-17 tex/cpp21/semantics.tex
[ writing ] more CPP paper

@@
 
-\subsection{Syntactic traversal}
+\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
+  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
+  $\plr{-} \env\V \Delta$ for all $\Delta$.
+\end{lemma}
+\begin{proof}
+  For renameable $\V$, we can implement $\mathrm{lift}$ as in
+  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
+  $\V \coloneqq \W \coloneqq \V$.
+  Renaming the environment then becomes an instance of environment composition.
+\end{proof}
 
@@
 
+Earlier, when talking about syntax, we saw \AgdaFunction{Scope} as a simple way
+to let an open family admit context extensions.
+However, a key component of our generic semantic traversal is to make use of
+the open family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
+we store in an environment.
+The definition \AgdaFunction{Kripke} gives an alternative to
+\AgdaFunction{Scope} which interprets the newly bound variables via a
+requirement of $\V$-values rather than extra assumptions for the
+$\C$-computation.
+
 \ExecuteMetaData[\Semanticstex]{Kripke}
+
+\AgdaFunction{Wrap} is a device that turns any type family into an equivalent
+type family that is definitionally injective in its indices.
+It turns the type family into a parametrised record with a single field
+\AgdaField{get} whose type is the type in the body of the $\lambda$-abstraction.
+For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
+ignored.
+If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
+\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
+\ExecuteMetaData[\Snippetstex]{KripkeExpanded}.
+That is to say, the Kripke function is expecting a value for each newly bound
+variable, at the multiplicity of its annotation, together with the resources
+supporting each of those values.
+The need for \AgdaFunction{$\Box^r$} is exemplified by the following construct
+\AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
+version of \AgdaBound{$\Delta$}.
+The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
+
+For typical choices of \AgdaBound{$\V$} (specifically, those which support
+identity environments as per \cref{thm:env-id}),
+\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}
+is stronger than \AgdaFunction{Scope}\AgdaSpace{}\AgdaBound{$\C$}.
+To show this, we instantiate the Kripke function with the renaming
+$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
+pass it an argument environment
+$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
+extended part.
+Post-composition of a renaming comes from \cref{thm:env-comp}.
+{\color{red}(TODO: refer to specific example of composition.)}
+
+\ExecuteMetaData[\Syntactictex]{reify}
+
+We tend to use \AgdaFunction{reify} in syntactic traversals, where, for example,
+\ExecuteMetaData[\Snippetstex]{term-with-bound}\ could be passed to a syntax
+constructor and make sense.
+
 \ExecuteMetaData[\Semanticstex]{Semantics}
+
+\ExecuteMetaData[\Semanticstex]{Comp}
+
+\ExecuteMetaData[\Semanticstex]{semantics-type}
+\ExecuteMetaData[\Semanticstex]{body-type}
+
+\ExecuteMetaData[\Semanticstex]{semantics}
+\ExecuteMetaData[\Semanticstex]{body}

### 2021-09-17 tex/cpp21/syntax.tex
[ writing ] more CPP paper

@@
 
-Then, we give the set of labels.
+%Then, we give the set of labels.
 There is one label for each instantiation of a logical rule, but the labels
@@
 Finally, we specify, for each label, the rest of the information about that
-rule.
+rule {\color{red}(premises and conclusion)}.
 We do not have a variable rule, because the variable rule is implicit to any
 type system.
+{\color{red}(Looks like \cref{fig:lr-comb})}
 
@@
 
-\ExecuteMetaData[\Syntaxtex]{Scoped}
+\ExecuteMetaData[\Syntaxtex]{OpenFam}
 
@@
 The most obvious way to make such an \AgdaBound{X} is to use some existing
-\AgdaFunction{Scoped} on an extended context.
+\AgdaFunction{OpenFam} on an extended context.
 This is what \AgdaFunction{Scope} does --- take the new variables
@@
 \ExecuteMetaData[\Syntaxtex]{Scope}
+
+{\color{red}(Forward ref: for now, we could have inlined \texttt{Scope}.)}
 

### 2021-09-20 tex/cpp21/env.tex
[ cpp21 ] compilation and bibliography fixes

@@
   \cref{def:lr-env} is sound and complete for the following syntax.
-  \begin{align*}
+  \begin{displaymath}
     \begin{prooftree}[comb,center=false]
@@
     \end{prooftree}
-    &&
+    \qquad
     \begin{prooftree}[comb,center=false]
@@
     \end{prooftree}
-  \end{align*}
+  \end{displaymath}
 \end{lemma}
@@
 the unit and lift of a Kleisli triple.
-\todo{Change from equality to inequality}
 
@@
   Let $\gr\Psi$ be the identity map, which clearly satisfies
-  $\grP = \grP\gr\Psi$.
-  When acting on a variable, the equation $\grPprime = \grQprime\gr\Psi$ means
-  that $\grPprime = \grQprime$, so we want, from a variable of type
-  $\grPprime\gamma \sqni A$, a value of type $\V\,\grPprime\gamma\,A$, which
-  we can get from $\mathrm{vr}$.
+  $\grP \leq \grP\gr\Psi$.
+  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
+  means that $\grPprime \leq \grQprime$, so the variable of type
+  $\grQprime\gamma \sqni A$ can be coerced to have type
+  $\grPprime\gamma \sqni A$.
+  From this point, $\mathrm{vr}$ gives us a value of type
+  $\V\,\grPprime\gamma\,A$, as required.
 \end{proof}
@@
   $\grPprime$ and $\grQprime$ such that
-  $\grPprime = \grQprime\plr{\rho.\gr\Psi}$, we have a function
+  $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$, we have a function
   $\mathrm{lift}_\rho :
@@
   $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
-  $\grP = \grQ\plr{\rho.\gr\Psi} =
+  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
   \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
-  For the action on variables, we are given $\grPprime = \grRprime\gr\Psi$ with
-  $\grRprime\theta \sqni A$.
+  For the action on variables, we are given $\grPprime \leq \grRprime\gr\Psi$
+  with $\grRprime\theta \sqni A$.
   We can immediately apply the action of $\sigma$, giving us a value of type
@@
   We note that
-  $\grPprime = \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
+  $\grPprime \leq \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
   apply $\mathrm{lift}_\rho$ to get the desired value.
@@
 
-\begin{corollary}[Composition of environments]\label{thm:env-comp}
+\begin{lemma}[Composition of environments]\label{thm:env-comp}
   Given a function
   $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
-  \forall \grPprime, \grQprime.~\grPprime = \grQprime\plr{\rho.\gr\Psi} \to
+  \forall \grPprime, \grQprime.~\grPprime \leq \grQprime\plr{\rho.\gr\Psi} \to
   \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
@@
   $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
+\end{lemma}
+
+\begin{corollary}\label{thm:env-postren}
+  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
+  $\sigma : \Delta \env\sqni \Theta$, we can get an environment of type
+  $\Gamma \env\U \Theta$.
+  When $\U = {\sqni}$, this gives composition of renamings.
 \end{corollary}
+\begin{proof}
+  By \cref{thm:env-comp}, with $\mathrm{lift}$ being given exactly by the
+  action of $\rho$ on variables.
+\end{proof}
 
@@
   \begin{itemize}
-    \item If $\U = \V = \W = {\sqni}$, then $\mathrm{lift}$ is given by the
-      action of the renaming $\rho$ on variables.
-      This allows us to derive composition of renamings.
-    \item More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
-      the action of the environment $\rho$.
-      This means that renamings post-compose with any other sort of environment.
     \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a

### 2021-09-21 tex/cpp21/examples.tex
[ writing ] semantics and examples

@@
 
+The action of a syntactic traversal on logical rules is basically fixed: we
+preserve the logical rule and extend the environment with any newly bound
+variables according to \AgdaField{vr}.
+Meanwhile, the action on variables is relatively unconstrained: we look up the
+variable in the environment to get a $\V$-value, then transform that $\V$-value
+into a term using \AgdaField{tm}.
+
+The idea of renaming is that variables replace variables, whereas with
+substitution, terms replace variables.
+This translates to environments for renaming containing $\sqni$-values
+(variables), and environments for substitution containing $\vdash$-values
+(terms).
+
+%To implement renaming and substitution for terms, we now just implement
+%syntactic kits for variables and terms, respectively.
+
 \ExecuteMetaData[\Syntactictex]{Ren-Kit}
+
+Notice that \AgdaFunction{ren\textasciicircum$\vdash$}, witnessing the fact
+that terms are renameable, is a corollary of \AgdaFunction{Ren-Kit}.
+
 \ExecuteMetaData[\Syntactictex]{Sub-Kit}
@@
 \subsection{A usage elaborator}
+
+Using the constructs we have seen so far, producing example terms soon becomes
+extremely tedious.
+We achieved some abbreviation by using pattern synonyms, but we still have to
+produce essentially bespoke proofs whenever we use a usage-sensitive part of the
+syntax.
+The size of each of these proofs is roughly proportional to the number of free
+variables, so the amount of proof we have to write grows roughly quadratically
+with the size of terms.
+An additional factor, which we can't see on paper, is that type checking time
+for these proofs soon becomes prohibitive to interactive development.
+
+Our aim in this subsection is to automate usage constraint proofs, making terms
+both easier to write and more performant to check.
+We invoke the automation by writing terms in a syntax where usage constraints
+have been trivialised, and then use a semantic traversal over the simplified
+syntax to try to produce a fully elaborated term in the original syntax.
+We write the automation in a way that is generic in the syntax description, thus
+avoiding repetition and facilitating the prototyping of new type systems.
+
+The type of syntax descriptions depends on the type of usage annotations because
+of variable binding.
+For example, in the $\oc_{\gr r}$-E rule of \cref{fig:lr-comb}, the right
+premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
+from the ambient posemiring.
+The scaling combinator also makes direct reference to the posemiring.
+To produce a simplified syntax description, where usage constraints are
+trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
+posemiring.
+In contrast to syntax descriptions, even though types can contain usage
+annotations, the type of types does not depend on the type of usage annotations.
+This means that, in our simplified syntax, terms have types from the original
+system even though variables have trivial usage annotations.
+We define the $\mathbf0$ posemiring as follows, being careful to use the
+0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
+solved by Agda's $\eta$-laws.
+Indeed, in this very definition, all of the semiring operations and laws are
+canonically inferred.
+
+\ExecuteMetaData[\UsageChecktex]{0-poSemiring}
+
+The elaboration process is monadic.
+In particular, we use the \AgdaDatatype{List}/non-determinism monad to give
+\emph{all} of the possible annotation choices on the free variables of a term.
+We believe the commitment to multiple solutions is inherent when the syntax
+contains \AgdaInductiveConstructor{`$\dot1$}.
+For example, in the intermediate stages of elaborating
+$\plr{\vdash \lambda x.~\plr{*,*}} : A \multimap \top \otimes \top$ with a
+usage counting posemiring (assuming reasonable rules for $\top$ and $\otimes$),
+it is unclear whether to use the variable $x$ in the left $*$ or the right $*$.
+This uncertainty should be reflected in the final result.
+
+The non-deterministic choices we make during elaboration are enumerated by
+the fields of \AgdaRecord{NonDetInverses}.
+These choices are driven by the typing rules and a candidate usage vector for
+the conclusion.
+For example, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is needed when we
+encounter a \AgdaInductiveConstructor{`$\sep$} in the syntax and the candidate
+usage annotation we are considering is \AgdaBound{r}.
+Then, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is a list of pairs of
+annotations \AgdaBound{p} and \AgdaBound{q} that \AgdaBound{r} can split into,
+together with a proof of the splitting.
+For \AgdaField{0\#$^{-1}$} and \AgdaField{1\#$^{-1}$}, inverses to constants,
+we are given the candidate \AgdaBound{r} and typically return an empty list if
+the constraint cannot be satisfied, or a singleton list containing a proof.
+\AgdaField{*$^{-1}$} is used when we encounter scaling, in which case we know
+both the scaling factor \AgdaBound{r} (from the syntax description) and the
+candidate \AgdaBound{q}.
+These inverse operations combine monadically (in fact, applicatively) to give
+inverses to the vector operations of zero, addition, scaling, and basis.
+
+\ExecuteMetaData[\UsageChecktex]{NonDetInverses}
+
+We choose the \AgdaBound{$\V$} of our semantics to be (unannotated) variables.
+For the \AgdaBound{$\C$}, we consider \emph{functions} from candidate usage
+vectors \AgdaBound{R} to the list of elaborated derivations with usage
+annotations given by \AgdaBound{R}.
+The module name \AgdaModule{U} refers to the fact that we are taking the
+ambient posemiring to be $\mathbf0$ in \AgdaFunction{OpenFam}.
+The effect on \AgdaFunction{OpenFam} is that the usage annotations of any
+contexts we consider are uninformative (hence the \AgdaSymbol{\_} on the left).
+
+\ExecuteMetaData[\UsageChecktex]{C}
+
+To traverse the unannotated terms, we produce a \AgdaRecord{Semantics} over the
+unannotated system \AgdaFunction{uSystem}\AgdaSpace{}\AgdaBound{sys}.
+We already know that variables are renameable.
+To interpret a variable, we consider all the possible proofs that such a
+variable could be well annotated, and package them up as a variable term.
+
+\ExecuteMetaData[\UsageChecktex]{elab-sem}
+
+\ExecuteMetaData[\UsageChecktex]{lemma-type}
+
+To actually use \AgdaFunction{elab-sem} on terms, we take the associated
+\AgdaFunction{semantics} and pass it the identity environment (an identity
+\emph{renaming} in this case, because $\V$ is a family of variables).
+The candidate usage vector \AgdaBound{R} will be empty for closed terms, and
+otherwise we have to supply the intended usage annotations.

### 2021-09-28 tex/cpp21/quant-framework.tex
[ cpp21 ] compilation and bibliography fixes
[ writing ] semantics and examples
[ writing ] remove duplicate bibliography directives
[ esop ] A draft table of contents

@@
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
-
-\citestyle{acmauthoryear}
-\bibliographystyle{ACM-Reference-Format}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
+
 \def\genericlr{../../generic-lr/src/processed-latex}
@@
 \def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
+\def\UsageChecktex{\genericlr/Generic/Linear/Example/UsageCheck.tex}
 \def\Snippetstex{../../agda/processed-latex/Snippets.tex}
@@
 
+\begin{enumerate}
+\item (Section 1)
+  \begin{enumerate}
+  \item   Motivate intrinsically typed syntax (in Agda)
+    \begin{enumerate}
+    \item Show the ``GADT'' for well scoped syntax
+    \item Show the well typed version
+    \end{enumerate}
+  \item Motivate adding posemiring annotations
+    \begin{enumerate}
+    \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
+    \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
+    \end{enumerate}
+  \end{enumerate}
+\item (Section 2) Background / notation fixing for:
+  \begin{enumerate}
+  \item Partially ordered semirings
+  \item Vectors
+  \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
+  \end{enumerate}
+\item (Section 3) Bunched Typing Rules
+  \begin{enumerate}
+  \item Introduces $\lambda\mathcal{R}$
+  \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
+  \end{enumerate}
+\item (Section 4) Universe of Generic Usage-annotated syntax
+  \begin{enumerate}
+  \item Premises
+  \item Rules
+  \item Systems
+  \item Example(s)
+  \item Intepretation as Terms
+  \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
+  \end{enumerate}
+\item (Section 5) Example Syntaxes
+  \begin{enumerate}
+  \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
+  \item Inductive types with a well resourced fold
+  \item Bernardy-Abel style
+  \item First-order (circuitry)
+  \item LNL
+  \item $\lambda\mu\tilde{\mu}$
+  \end{enumerate}
+\item (Section 6) Environments
+  \begin{enumerate}
+  \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
+  \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
+    \begin{enumerate}
+    \item Key idea / slogan : environment = change of basis
+    \end{enumerate}
+  \item Definition of environments
+  \item Examples
+  \item Properties, incl. preservation of linear operations
+  \end{enumerate}
+\item (Section 7) Semantics, or How to apply an Environment to a Term
+  \begin{enumerate}
+  \item Renameability
+  \item $\mathsf{semantics}$
+  \end{enumerate}
+\item (Section 8) Examples of Semantics
+  \begin{enumerate}
+  \item Renaming and Substitution (term-valued)
+  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/AnnotatedArrow.agda})
+  \item Usage Elaborator
+  \end{enumerate}
+\item (Section 9) Related Work
+\item (Section 10) Conclusion
+\end{enumerate}
+
 \section{Introduction}\label{sec:intro}

### 2021-10-04 tex/esop22/quant-framework.tex
[ new ] switch to working on ESOP submission

--- 2021-09-28 tex/cpp21/quant-framework.tex
@@
-\documentclass[sigplan,10pt,anonymous,review]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.20 of 2017/10/04
+%
+\documentclass[runningheads]{llncs}
 
 \usepackage[conor]{agda}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{catchfilebetweentags}
@@
 \usepackage{mathtools}
+\usepackage[numbers,sort&compress]{natbib}
 \usepackage{newunicodechar}
 \usepackage{stmaryrd}
-\setlength{\marginparwidth}{2cm}
 \usepackage{todonotes}
@@
 \input{lagda.tex}
-
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
 
@@
 \begin{document}
-
-\title{A Framework for Substructural Type Systems}
-
-\author{James Wood}
-\orcid{0000-0002-8080-3350}
-\affiliation{%
-  \institution{University of Strathclyde}%
-  \city{Glasgow}%
-  \country{UK}%
-}
-\email{james.wood.100@strath.ac.uk}
-
-\author{Robert Atkey}
-\orcid{0000-0002-4414-5047}
-\affiliation{%
-  \institution{University of Strathclyde}%
-  \city{Glasgow}%
-  \country{UK}%
-}
-\email{robert.atkey@strath.ac.uk}
-
-\keywords{substructural, linear, metatheory, type system, mechanisation}
-
+%
+\title{A Framework for Substructural Type Systems
+  \thanks{Supported by organization x.}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{James Wood\inst{1}\orcidID{0000-0002-8080-3350} \and
+Robert Atkey\inst{1}\orcidID{0000-0002-4414-5047}}
+%
+\authorrunning{J. Wood, R. Atkey}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{University of Strathclyde, Glasgow, UK
+\email{\{james.wood.100,robert.atkey\}@strath.ac.uk}}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
@@
   concrete terms.
+
+\keywords{First keyword  \and Second keyword \and Another keyword.}
 \end{abstract}
-
-\maketitle
-
+%
+%
+%
 \begin{enumerate}
@@
   \item Renaming and Substitution (term-valued)
-  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/AnnotatedArrow.agda})
+  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda})
   \item Usage Elaborator
@@
 \section{Conclusions}\label{sec:conc}
-
+%
+% ---- Bibliography ----
+%
+% BibTeX users should specify bibliography style 'splncs04'.
+% References will then be sorted and formatted in the correct style.
+%
+\bibliographystyle{splncs04nat}
 \bibliography{quant-framework}
-
 \end{document}

### 2021-10-05 tex/esop22/env.tex
[ writing ] explaining semantics

--- 2021-09-20 tex/cpp21/env.tex
@@
 
-\begin{lemma}
+\begin{lemma}\label{thm:env-rules}
   \cref{def:lr-env} is sound and complete for the following syntax.
@@
 
-\begin{lemma}\label{thm:env-resize}
+\begin{lemma}[\AgdaFunction{relocate}]\label{thm:env-resize}
   Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$

### 2021-10-05 tex/esop22/semantics.tex
[ new ] switch to working on ESOP submission
[ writing ] explaining semantics

--- 2021-09-17 tex/cpp21/semantics.tex
@@
 changes to usage annotations.
-We follow the structure given by Allais et.\ al.~\cite{AACMM20} closely, with
+We follow the structure given by \citet{AACMM20} closely, with
 the main difference having come already in the syntax and environments.
@@
 extended part.
-Post-composition of a renaming comes from \cref{thm:env-comp}.
-{\color{red}(TODO: refer to specific example of composition.)}
+Post-composition of a renaming comes from \cref{thm:env-postren}.
 
@@
 constructor and make sense.
+\todo{It's also used in the denotational semantics, so we need a better
+  explanation.}
 
@@
 
-\ExecuteMetaData[\Semanticstex]{Comp}
+%\ExecuteMetaData[\Semanticstex]{Comp}
+
+We mutually define the action \AgdaFunction{semantics} and its lemma
+\AgdaFunction{body}.
+The purpose of \AgdaFunction{semantics} is to turn a term into a
+\AgdaBound{$\C$}-value using a \AgdaBound{$\V$}-environment and the fields of
+\AgdaRecord{Semantics}.
+Meanwhile, \AgdaFunction{body} does a similar job, but also deals with
+newly bound variables.
+In particular, \AgdaFunction{body} takes a term in a context extended by
+\AgdaBound{$\Theta$}, and produces a Kripke function from
+\AgdaBound{$\V$}-values for \AgdaBound{$\Theta$} to \AgdaBound{$\C$}-values.
 
 \ExecuteMetaData[\Semanticstex]{semantics-type}
-\ExecuteMetaData[\Semanticstex]{body-type}
+
+In \AgdaFunction{semantics}, to deal with a variable \AgdaBound{v}, we look it
+up in the environment \AgdaBound{$\rho$}, then use the
+\AgdaField{$\sem{\text{var}}$} field to map the resulting
+\AgdaBound{$\V$}-value to a \AgdaBound{$\C$}-value.
+To deal with a syntactic construct, we use the functoriality of a layer of
+syntax, together with recursive calls to \AgdaFunction{semantics} via
+\AgdaFunction{body}.
+Recall that \AgdaFunction{relocate} (\cref{thm:env-resize}) adjusts the
+environment \AgdaBound{$\rho$} to work in the usage contexts of the subterms.
 
 \ExecuteMetaData[\Semanticstex]{semantics}
+
+For \AgdaFunction{body}, we are given
+$\rho : \grP\gamma \env\V \Delta$,
+$\mathit{th} : \grP\gr{^+}\gamma^+ \env\sqni \grP\gamma$,
+$\sigma : \grQ\gr{^+}\gamma^+ \env\V \Theta$,
+with $r : \grR\gr{^+} \leq \grP\gr{^+} + \grQ\gr{^+}$.
+\todo{This needs some explanation.}
+We are aiming for an environment of type
+$\grR\gr{^+}\gamma \env\V \Delta, \Theta$, which we get by renaming $\rho$ by
+$\mathit{th}$ to give $\rho'$, then using \cref{thm:env-rules} to stick
+together the environment $\rho'$ for old variables $\Delta$ and the
+environment $\sigma$ for the new variables $\Theta$.
+
 \ExecuteMetaData[\Semanticstex]{body}

### 2021-10-07 tex/esop22/semantics.tex
[ writing ] explain Kripke things, semantics

@@
 
-\subsection{Semantic traversal}
+\subsection{The Kripke function space}
 
@@
 \AgdaFunction{Wrap} is a device that turns any type family into an equivalent
-type family that is definitionally injective in its indices.
+type family that is judgementally injective in its indices.
 It turns the type family into a parametrised record with a single field
@@
 
-For typical choices of \AgdaBound{$\V$} (specifically, those which support
-identity environments as per \cref{thm:env-id}),
-\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}
-is stronger than \AgdaFunction{Scope}\AgdaSpace{}\AgdaBound{$\C$}.
+When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
+we can coerce a \AgdaFunction{Kripke} function into just a
+\AgdaBound{$\C$}-value in an extended context.
 To show this, we instantiate the Kripke function with the renaming
@@
 extended part.
-Post-composition of a renaming comes from \cref{thm:env-postren}.
+Post-composition of a renaming onto an environment comes from
+\cref{thm:env-postren}.
 
@@
 
-We tend to use \AgdaFunction{reify} in syntactic traversals, where, for example,
-\ExecuteMetaData[\Snippetstex]{term-with-bound}\ could be passed to a syntax
-constructor and make sense.
-\todo{It's also used in the denotational semantics, so we need a better
-  explanation.}
+We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
+identity environments.
+In this paper, all examples support identity environments.
+However, \citet[p.~27]{AACMM20} give some important examples that do not
+support identity environments, and thus cannot use \AgdaFunction{reify}.
+The feature that causes the lack of support for identity environments is that
+a semantics can make use of the fact that only variables of particular kinds
+are bound by the syntax.
+In the examples of \citeauthor{AACMM20}, a bidirectionally typed language only
+binds variables that synthesise their type, as opposed to those whose type is
+checked.
+The semantics of type-checking and elaboration rely on variables synthesising
+their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
+Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
+only binds such synthesising variables.
+Similar phenomena would appear in, say, a call-by-value language where
+variables are values (not computations), or a polarised language where
+variables always have a polarity matching their type.
+
+\subsection{Semantic traversal}
+
+The aim of this subsection is to give an alternative recursion principle for
+terms that incorporates some of the environment-handling seen in the
+implementations of renaming and substitution.
+The rest of this section assumes the following: a renameable open family
+\AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
+algebra for a layer of syntax at \AgdaBound{$\C$}.
 
@@
 %\ExecuteMetaData[\Semanticstex]{Comp}
+
+Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
+\cref{thm:lr-bind} as
+\AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
+\AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
+We can think of \AgdaFunction{extend} as our best effort to extend an
+environment by \AgdaBound{$\Theta$} without access to an identity environment
+at \AgdaBound{$\Theta$}.
+We instead take \AgdaBound{$\rho$}, which is an environment covering
+\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an environment
+covering \AgdaBound{$\Theta$}, and glue them together using
+\cref{thm:env-rules}, after having renamed \AgdaBound{$\rho$} via
+\cref{thm:env-ren} to make it fit the new context \AgdaBound{$\Gamma^+$}
+(intended to be \ExecuteMetaData[\Snippetstex]{GT}).
+
+\ExecuteMetaData[\Semanticstex]{extend}
 
@@
 
-In \AgdaFunction{semantics}, to deal with a variable \AgdaBound{v}, we look it
+To implement the new recursor \AgdaFunction{semantics}, we use the standard
+recursor, which in one case gives us a variable \AgdaBound{v}, and in the other
+gives us a structure of subterms \AgdaBound{M}, each of which is in an extended
+context.
+To deal with a variable \AgdaBound{v}, we look it
 up in the environment \AgdaBound{$\rho$}, then use the
@@
 \AgdaBound{$\V$}-value to a \AgdaBound{$\C$}-value.
-To deal with a syntactic construct, we use the functoriality of a layer of
-syntax, together with recursive calls to \AgdaFunction{semantics} via
-\AgdaFunction{body}.
+To deal with a structure of subtems \AgdaBound{M}, we use the functoriality of
+the syntactic structure to consider each subterm separately.
+On a subterm, we apply \AgdaFunction{body}, which amounts to a recursive call
+to \AgdaFunction{semantics} with an extended environment.
 Recall that \AgdaFunction{relocate} (\cref{thm:env-resize}) adjusts the
@@
 
-For \AgdaFunction{body}, we are given
-$\rho : \grP\gamma \env\V \Delta$,
-$\mathit{th} : \grP\gr{^+}\gamma^+ \env\sqni \grP\gamma$,
-$\sigma : \grQ\gr{^+}\gamma^+ \env\V \Theta$,
-with $r : \grR\gr{^+} \leq \grP\gr{^+} + \grQ\gr{^+}$.
-\todo{This needs some explanation.}
-We are aiming for an environment of type
-$\grR\gr{^+}\gamma \env\V \Delta, \Theta$, which we get by renaming $\rho$ by
-$\mathit{th}$ to give $\rho'$, then using \cref{thm:env-rules} to stick
-together the environment $\rho'$ for old variables $\Delta$ and the
-environment $\sigma$ for the new variables $\Theta$.
+For \AgdaFunction{body}, we are given a subterm \AgdaFunction{M}, to which we
+want to apply \AgdaFunction{semantics}.
+To do so, we need an extended version of the initial environment
+\AgdaBound{$\rho$}.
+The best we can achieve without identity environments for \AgdaBound{$\V$} is
+a Kripke function returning an extended environment.
+We use \AgdaFunction{mapK$\C$} to post-compose this function by the
+\AgdaSymbol{$\lambda$}-function taking an extended environment and acting with
+it on \AgdaBound{M}.
 

### 2021-10-08 tex/esop22/syntax.tex
[ new ] switch to working on ESOP submission
[ writing ] add example term

--- 2021-09-17 tex/cpp21/syntax.tex
@@
 \begin{figure}
-  \begin{align*}
+  \begin{displaymath}
     \begin{prooftree}
@@
     \end{prooftree}
-  \end{align*}
-  \begin{align*}
+  \end{displaymath}
+  \begin{displaymath}
     \begin{prooftree}
@@
     \end{prooftree}
-    &&
+    \qquad
     \begin{prooftree}
@@
     \end{prooftree}
-  \end{align*}
+  \end{displaymath}
   \caption{A fragment of \name{}}\label{fig:lr}
@@
 \begin{figure}
-  \begin{align*}
+  \begin{displaymath}
     \begin{prooftree}[comb]
@@
     \end{prooftree}
-    &&
+    \qquad
     \begin{prooftree}[comb]
@@
     \end{prooftree}
-  \end{align*}
-  \begin{align*}
+  \end{displaymath}
+  \begin{displaymath}
     \begin{prooftree}[comb]
@@
     \end{prooftree}
-  \end{align*}
+  \end{displaymath}
   \caption{\cref{fig:lr} restated in terms of bunched combinators}
@@
 
-\missingfigure{Example term}
+Here is an example term, using the \AgdaFunction{$\lambda$R} system.
+First, for ease of writing, we introduce pattern synonyms for each of the
+typing rules we use.
+
+\ExecuteMetaData[\PaperExamplestex]{patterns}
+
+Our example term is a function that flips a tagged union wrapped in an
+arbitrarily annotated \emph{bang}.
+Much of the effort in writing such a term goes into writing the various
+relatedness proofs between usage contexts --- observing, for example, that two
+usage contexts sum together to make a third, or that a usage context used for
+a variable is a basis vector.
+We give a method of automating these proofs in \cref{sec:usage-elaborator}.
+
+\ExecuteMetaData[\PaperExamplestex]{lR-term}
 

### 2021-10-12 tex/esop22/examples.tex
[ writing ] add example term
[ new ] WRel example first draft

--- 2021-09-21 tex/cpp21/examples.tex
@@
 
-\subsection{A usage elaborator}
+To abbreviate this section, we use a simplified syntax compared to \name{}.
+We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
+type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
+\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.
+
+\ExecuteMetaData[\WReltex]{Ty}
+
+In the term syntax, $\lambda$-abstraction now binds a variable with annotation
+\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
+(both in accordance with the function type they are acting on).
+
+\ExecuteMetaData[\WReltex]{AnnArr}
+
+In this subsection, we take the usage annotations to be the 4-element variance
+posemiring.
+We establish the property that all terms are monotonic in their free variables.
+This monotonicity can be covariant or contravariant (or neither or both)
+depending on the annotation of each free variable.
+This provides an additional example to those of \citet{AbelBernardy2020}.
+\todo{Cite before here.}
+
+We will take semantics of this system into
+\emph{worldly relations}~\cite{AbelBernardy2020}.
+A worldly relation over a poset of worlds \AgdaBound{W} is a set over which
+we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
+property with respect to the order on \AgdaBound{W}.
+
+\ExecuteMetaData[\WReltex]{WRel}
+
+\begin{example}
+  When \AgdaBound{W} is the 1-element set, a worldly relation is just a set
+  equipped with a binary relation.
+\end{example}
+
+Morphisms between worldly relations \AgdaBound{R} and \AgdaBound{S} consist of
+a mapping between the underlying sets such that that mapping preserves
+relatedness from \AgdaBound{R} to \AgdaBound{S}.
+
+\ExecuteMetaData[\WReltex]{WRelMor}
+
+When the poset of worlds forms a (relational) commutative monoid, such worldly
+relations support a symmetric monoidal closed structure.
+We reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
+\AgdaRecord{$\wand$}, now over worlds rather than contexts.
+
+\ExecuteMetaData[\WReltex]{IR}
+\ExecuteMetaData[\WReltex]{tensorR}
+\ExecuteMetaData[\WReltex]{lollyR}
+
+The final piece of sematics we need is a \emph{bang} operator.
+Instead of requiring extra algebraic structure on the worlds, we allow the
+semantic \emph{bang} to be an arbitrary annotation-indexed functor on worldly
+relations.
+This functor must respect all of the structure on the indices, making it a
+graded comonad over multiplication, as well as being lax monoidal at any
+particular index \AgdaBound{r}.
+
+\ExecuteMetaData[\WReltex]{Bang}
+
+\begin{example}
+  With \AgdaBound{W} being the 1-element set and annotations coming from the
+  variance semiring, we can define the following \emph{bang}.
+  It is always the identity on the set component, while the relation component
+  consists of flipping the relation for contravariance and taking conjunctions
+  to achieve both covariance and contravariance.
+  When we want neither covariance nor contravariance, we use the always true
+  predicate on worlds \AgdaFunction{$\dot1$}.
+
+  \ExecuteMetaData[\Monotonicitytex]{BangR}
+\end{example}
+
+To associate semantics to syntax, we start as standard by associating worldly
+relations to types.
+We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
+\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
+context, concatenation, and usage annotations on singletons, respectively.
+
+\ExecuteMetaData[\WReltex]{sem}
+
+The semantics of a term is then to be a morphism from the interpretation of the
+context to the interpretation of the term's type.
+
+\ExecuteMetaData[\WReltex]{sem-vdash}
+
+Variables are given semantics by \AgdaFunction{lookup$^R$} (definition omitted).
+
+\ExecuteMetaData[\WReltex]{lookupR-type}
+
+Now, we give a \AgdaRecord{Semantics}.
+The choice of \AgdaBound{$\V$} as
+\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
+given that a standard denotational semantics would not use intermediate
+environments in the same sense as renaming and substitution, but allows us to
+reuse the standard facts that variables support renaming and identity
+environments.
+With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
+environment entries by \AgdaFunction{lookup$^R$}.
+Meanwhile, for the logical rules, we ignore environments by using
+\AgdaFunction{reify} to just deal with morphisms in an extended context.
+As such, $\lambda$-abstractions are easy to interpret, while applications
+require some massaging to remove the extension by an empty context, followed by
+some plumbing to split the interpretation of the context according to the usage
+constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
+the interpretation of the function \AgdaBound{m$'$}.
+
+\ExecuteMetaData[\WReltex]{Wrel}
+
+In order to map open terms to interpretations, we take the action of the
+semantics and give the identity renaming as the starting environment.
+
+\ExecuteMetaData[\WReltex]{wrel}
+
+\begin{example}
+  We can make a subtraction function from primitive addition and negation on
+  integers.
+  Subtraction is covariant in its first argument and contravariant in its
+  second argument.
+  We give the definition in pseudocode, as we have not yet seen how to
+  conveniently write terms (\cref{sec:usage-elaborator}).
+
+  \begin{align*}
+    &{\sim\sim}p :
+      {\uparrow\uparrow}\mathbb Z \multimap
+      {\uparrow\uparrow}\mathbb Z \multimap \mathbb Z,
+      {\sim\sim}n : {\downarrow\downarrow}\mathbb Z \multimap \mathbb Z
+      \vdash \mathnormal{minus} :
+      {\uparrow\uparrow}\mathbb Z \multimap
+      {\downarrow\downarrow}\mathbb Z \multimap
+      \mathbb Z
+    \\
+    &\mathnormal{minus} \coloneqq \lambda x.~\lambda y.~p\,x\,(n\,y)
+  \end{align*}
+
+  We observe that the set component of this term's semantics is just the
+  expected Agda function when the two free variables are given appropriate
+  interpretations.
+
+  \ExecuteMetaData[\Monotonicitytex]{minus-set}
+
+  Furthermore, the relational component of the semantics yields the free
+  theorem that the Agda subtraction so defined is monotonic in the expected way.
+  This relies on library proofs that addition and negation are suitably
+  monotonic.
+
+  \ExecuteMetaData[\Monotonicitytex]{thm}
+\end{example}
+
+\subsection{A usage elaborator}\label{sec:usage-elaborator}
 

### 2021-10-12 tex/esop22/quant-framework.tex
[ writing ] explaining semantics
[ refactor ] example term in separate file
[ new ] WRel example first draft

@@
 \def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
+\def\HeavyItex{\genericlr/Generic/Linear/Example/PaperExamples/HeavyI.tex}
 \def\UsageChecktex{\genericlr/Generic/Linear/Example/UsageCheck.tex}
+\def\WReltex{\genericlr/Generic/Linear/Example/WRel.tex}
+\def\Monotonicitytex{\genericlr/Generic/Linear/Example/WRel/Monotonicity.tex}
 \def\Snippetstex{../../agda/processed-latex/Snippets.tex}
@@
   \item Renaming and Substitution (term-valued)
-  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda})
+  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
   \item Usage Elaborator

### 2021-10-13 tex/esop22/algebra.tex
[ writing ] more algebra (modules, maps)

--- 2021-09-07 tex/cpp21/algebra.tex
@@
 
-We also make use, particularly in the mechanisation, of an alternative
-definition based on relations.
-
-\begin{definition}
-  A \emph{relational posemiring} is a tuple
-  $(\Ann, \leq, {{-}\leq}0, {{-}\leq}[{-}+{-}],
-  {{-}\leq}1, {{-}\leq}[{-}*{-}])$
-  such that the following laws hold (with free variables universally
-  quantified).
-  \begin{itemize}
-    \item $x' \leq x \to {x \leq}0 \to {x' \leq}0$
-    \item $x' \leq x \to y \leq y' \to z \leq z' \to
-      {x \leq}[y+z] \to {x' \leq}[y'+z']$
-    \item $x' \leq x \to {x \leq}1 \to {x' \leq}1$
-    \item $x' \leq x \to y \leq y' \to z \leq z' \to
-      {x \leq}[y*z] \to {x' \leq}[y'*z']$
-    \item $\plr{\exists y.~{y \leq}0 \land {x \leq}[y+z]}
-      \leftrightarrow x \leq z$
-    \item $\plr{\exists z.~{x \leq}[y+z] \land {z \leq}0}
-      \leftrightarrow x \leq y$
-    \item $\plr{\exists wx.~{wx \leq}[w+x] \land {z \leq}[wx+y]}$
-      $\leftrightarrow$
-
-      $\plr{\exists xy.~{z \leq}[w+xy] \land {xy \leq}[x+y]}$
-    \item $\exists x.~{x \leq}0 \land \forall x'.~{x' \leq}0 \to x' \leq x$
-    \item $\vdots$
-    %\item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
-    %\item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
-    %\item $0 + x = x \land x + 0 = x$
-    %\item $(x + y) + z = x + (y + z)$
-    %\item $x + y = y + x$
-    %\item $1 * x = x \land x * 1 = x$
-    %\item $(x * y) * z = x * (y * z)$
-    %\item $0 * x = 0 \land x * 0 = 0$
-    %\item $(x + y) * z = x * z + y * z$
-    %\item $x * (y + z) = x * y + x * z$
-  \end{itemize}
-\end{definition}
-
 An element of a chosen posemiring $\Ann$ describes the usage restrictions on
@@
 We can also form the standard basis vectors at any given dimension.
+
+Vectors of a given length form a \emph{module} over the posemiring $\Ann$,
+analogously to how vectors over a field form a vector space.
+
+\begin{definition}
+  A \emph{(left) module over a posemiring}, given a posemiring $\Ann$, is a
+  partially ordered commutative monoid $(M, 0_M, +_M)$ with, for each
+  $r \in \Ann$, a pomonoid morphism $r \cdot \plr{-} : M \to M$, such that the
+  collection of these respects the posemiring structure on $r$.
+  Specifically:
+  \begin{itemize}
+    \item $r \leq r' \to u \leq u' \to r \cdot u \leq r' \cdot u'$
+    \item $r \cdot 0_M = 0_M$
+    \item $r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$
+    \item $0 \cdot u = 0_M$
+    \item $\plr{r + s} \cdot u = r \cdot u +_M s \cdot u$
+    \item $1 \cdot u = u$
+    \item $\plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$
+  \end{itemize}
+\end{definition}
+
+We care to define modules so as to define \emph{module morphisms}, also known
+as \emph{linear maps}, which we use extensively when relating two contexts (as
+we do, for example, in simultaneous substitution).
+For the sake of mechanisation, we choose to define module morphisms
+\emph{relationally} rather than \emph{functionally}, giving a somewhat
+unfamiliar-looking definition that is equivalent to the usual functional
+definition.
+The main advantage of this relational approach is that proofs of relatedness
+for typical linear maps compose and decompose via data constructors and
+pattern matching.
+% I'm not sure this is true:
+%An auxiliary advantage is that with relations rather than functions, we can
+%much more often take advantage of judgemental injectivity, thus making
+%unification-based solving of implicits more effective.
+%For example, if \AgdaBound{R} is a free variable of relation type, then
+%\AgdaInductiveConstructor{refl} serves as a proof of
+%\ExecuteMetaData[\Snippetstex]{Rxy-R}{}, solving the underscores as
+%\AgdaBound{x} and \AgdaBound{y}, respectively.
 

### 2021-10-13 tex/esop22/bunchedrules.tex
[ writing ] New section: bunched typing rules

Need to introduce:
\begin{enumerate}
\item Intrinsically typed syntax
\item Intrinsically typed semiring annotated syntax
\item Context modification
\end{enumerate}

To build our universe of semiring annotated syntaxes with binding, we
need to closely examine the structures underlying the kinds of typing
rules present in these systems. We take a fragment of Wood and Atkey's
\name{} calculus, enough to show the key details, presented in
Figure \ref{fig:lr}.

\newcommand{\constr}[1]{\mathsf{#1}}

\begin{figure}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\grR \leq \bra x}
      \hypo{\gamma_x = A}
      \infer2{\constr{var} : \grR\gamma \vdash A}
    \end{prooftree}
  \end{displaymath}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{t : \grR\gamma, \gr1A \vdash B}
      \infer1{\constr{lolliI} : \grR\gamma \vdash A \multimap B}
    \end{prooftree}
    \qquad
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \hypo{s : \grP\gamma &\vdash A \multimap B}
      \infer[no rule]1{t : \grQ\gamma &\vdash A \phantom{{} \multimap B}}
      \infer2{\constr{lolliE} : \grR\gamma \vdash B}
    \end{prooftree}
  \end{displaymath}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\grR\gamma \vdash A_i}
      \infer1[$\oplus$-I$_i$]{\grR\gamma \vdash A_0 \oplus A_1}
    \end{prooftree}
    \qquad
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \infer[no rule]1{\grP\gamma \vdash A \oplus B}
      \hypo{\grQ\gamma, \gr1A &\vdash C}
      \infer[no rule]1{\grQ\gamma, \gr1B &\vdash C}
      \infer2[$\oplus$-E]{\grR\gamma \vdash C}
    \end{prooftree}
  \end{displaymath}
  \begin{displaymath}
    \begin{prooftree}
      \hypo{\grR \leq \gr r\grP}
      \hypo{\grP\gamma \vdash A}
      \infer2[$\oc_{\gr r}$-I]{\grR\gamma \vdash \oc_{\gr r} A}
    \end{prooftree}
    \qquad
    \begin{prooftree}
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash \oc_{\gr r} A}
      \infer[no rule]1{\grQ\gamma, \gr rA \vdash C}
      \infer2[$\oc_{\gr r}$-E]{\grR\gamma \vdash C}
    \end{prooftree}
  \end{displaymath}
  \caption{A fragment of \name{}}\label{fig:lr}
\end{figure}

In common with other mechanisations of syntax using dependent types,
we use a

These rules are the constructors of a datatype
$(\vdash) : \mathrm{Ctx} \to \mathrm{Ty} \to \mathrm{Set}$.

In keeping with our Agda formalisation, we have used an
\emph{intrinsically typed} syntax. Each typing rule is presented as a
constructor whose arguments are the premises above the line. Typed
terms are represented as nothing more than their typing derivations,
constructed from the constructors. To account for the semiring
annotations, the rules are

The key features we wish to describe from this

The $\constr{var}$ rule constructs variables,

Nevertheless, there are several features of this system that are non
standard. In the $\constr{var}$ rule,

There are two key features

first make a number of key observations on an exemplar syntax that
will guide us.

In systems with Linear types, a crucial distinction is made between
multiplicative and additive typing rules that does not arise in
non-substructural systems. The distinction is exemplified by the
Intuitionistic Linear Logic introduction rules for \emph{tensor}
products $A \otimes B$ and \emph{with} products $A \with B$:
\begin{mathpar}
  \inferrule* [right=$\otimes$-Intro]
  {\Gamma \vdash s : A \\ \Delta \vdash t : B}
  {\Gamma \vdash s \otimes t : A \otimes B}

  \inferrule* [right=$\with$-Intro]
  {\Gamma \vdash s : A \\ \Gamma \vdash t : B}
  {\Gamma \vdash s \with t : A \with B}
\end{mathpar}
In the \TirName{$\otimes$-Intro} rule, the two premises have disjoint
contexts $\Gamma$ and $\Delta$ that are combined in the conclusion,
indicating that the terms $s$ and $t$ refer to separate
resources. This form of rule is referred to as
\emph{multiplicative}. In the \TirName{$\with$-Intro} rule, the two
premises share the context $\Gamma$, indicating that the terms $s$ and
$t$ may refer to the same resources.

In systems that use semiring annotations to track resource usage, the
distinction between multiplicative and additive still remains but now
contexts in multiplicative rules are combined by adding their semiring
annotations together.

### 2021-10-13 tex/esop22/examples.tex
[ writing ] start on μμ~ account

@@
-\subsection{Renaming and substitution}
+\subsection{Renaming and substitution}\label{sec:kits}
 
@@
 
+\todo{Introduction}
 To abbreviate this section, we use a simplified syntax compared to \name{}.
@@
 posemiring.
+\todo{This works for any semiring.}
 We establish the property that all terms are monotonic in their free variables.
@@
 
+\todo{Define big intersection.}
 When the poset of worlds forms a (relational) commutative monoid, such worldly
@@
 The final piece of sematics we need is a \emph{bang} operator.
+\todo{No instead}
 Instead of requiring extra algebraic structure on the worlds, we allow the

### 2021-10-13 tex/esop22/syntax.tex
[ refactor ] example term in separate file
[ writing ] New section: bunched typing rules
[ writing ] start on μμ~ account

@@
 \subsection{A prototypical type system}
-
-\begin{figure}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR \leq \bra x}
-      \hypo{\gamma_x = A}
-      \infer2[Var]{\grR\gamma \vdash A}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR\gamma, \gr1A \vdash B}
-      \infer1[$\multimap$-I]{\grR\gamma \vdash A \multimap B}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \hypo{\grP\gamma &\vdash A \multimap B}
-      \infer[no rule]1{\grQ\gamma &\vdash A \phantom{{} \multimap B}}
-      \infer2[$\multimap$-E]{\grR\gamma \vdash B}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR\gamma \vdash A_i}
-      \infer1[$\oplus$-I$_i$]{\grR\gamma \vdash A_0 \oplus A_1}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \infer[no rule]1{\grP\gamma \vdash A \oplus B}
-      \hypo{\grQ\gamma, \gr1A &\vdash C}
-      \infer[no rule]1{\grQ\gamma, \gr1B &\vdash C}
-      \infer2[$\oplus$-E]{\grR\gamma \vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR \leq \gr r\grP}
-      \hypo{\grP\gamma \vdash A}
-      \infer2[$\oc_{\gr r}$-I]{\grR\gamma \vdash \oc_{\gr r} A}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \hypo{\grP\gamma \vdash \oc_{\gr r} A}
-      \infer[no rule]1{\grQ\gamma, \gr rA \vdash C}
-      \infer2[$\oc_{\gr r}$-E]{\grR\gamma \vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \caption{A fragment of \name{}}\label{fig:lr}
-\end{figure}
 
@@
 We give a method of automating these proofs in \cref{sec:usage-elaborator}.
+\todo{To be clear, we don't actually write this.}
 
-\ExecuteMetaData[\PaperExamplestex]{lR-term}
+\ExecuteMetaData[\HeavyItex]{lR-term}
 
@@
 \ExecuteMetaData[\Maptex]{map-s-type}
+
+\subsection{Other syntaxes}
+
+We can encode a usage-annotated version of System $L$/the
+$\mu\tilde\mu$-calculus~\cite{CH00} --- a syntax for classical logic --- in
+such a way that contexts capture the undistinguished parts of the sequent.
+As such, the generic substitution lemma we get in \cref{sec:kits} is the form
+of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
+Though the $\mu\tilde\mu$-calculus is originally described as a sequent
+calculus~\cite{CH00}, we use the techniques of
+\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it as a natural
+deduction system, thus giving a notion of \emph{variable} to the system.
+
+Unlike the single judgement form of \name{} and standard simply typed
+$\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
+terms, coterms, and commands.
+Read logically, terms and coterms are seen to, respectively, prove and refute
+propositions (types), while commands exhibit contradictions.
+This means that the abstract \AgdaBound{Ty} in the generic framework is
+instantiated to \AgdaDatatype{Conc} as below, with \AgdaDatatype{Ty} not being
+exposed directly to the generic framework.
+
+\begin{minipage}[t]{0.5\textwidth}
+  \centering
+  \ExecuteMetaData[\MuMuTildetex]{Ty}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+  \centering
+  \ExecuteMetaData[\MuMuTildetex]{Conc}
+\end{minipage}
+
+\ExecuteMetaData[\MuMuTildetex]{MMT}
+
+\ExecuteMetaData[\MuMuTildeTermtex]{patterns}
+\ExecuteMetaData[\MuMuTildeTermtex]{myComm}

### 2021-10-13 tex/esop22/bunchedrules.tex
[ writing ] a miserably small amount

@@
+Let $\Ann$ be a posemiring. Our framework represents well typed and
+$\Ann$-resourced terms \emph{intrinsically}. Intrinsic typing means
+that we represent well typed and $\Ann$-resourced terms (and
+\emph{only} those) as inhabitants of an inductive family
+$\grR\gamma \vdash A$ indexed by resource contexts $\grR$, type
+contexts $\gamma$, and types $A$. We represent typing contexts as
+trees with types at the leaves, and resource contexts as functions
+that assign elements of $\Ann$ to members of the context. Using trees
+instead of lists for typing contexts has the advantage that extension
+of a context by multiple variables does not lead to complex counting
+argument to access the pre-existing variables.\todo{left and right
+  constructors for paths}
+
+\begin{figure}
+  \begin{displaymath}
+    \begin{prooftree}
+      \hypo{x : \grR\gamma \sqni A}
+      \infer1{\AgdaInductiveConstructor{var}~x : \grR\gamma \vdash A}
+    \end{prooftree}
+  \end{displaymath}
+  \begin{displaymath}
+    \begin{prooftree}
+      \hypo{t : \grR\gamma, \gr1A \vdash B}
+      \infer1{\AgdaInductiveConstructor{⊸I}~t : \grR\gamma \vdash A \multimap B}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}
+      \hypo{p : \grR \leq \grP + \grQ}
+      \hypo{s : \grP\gamma &\vdash A \multimap B}
+      \infer[no rule]1{t : \grQ\gamma &\vdash A \phantom{{} \multimap B}}
+      \infer2{\AgdaInductiveConstructor{⊸E}~p~s~t : \grR\gamma \vdash B}
+    \end{prooftree}
+  \end{displaymath}
+  \begin{displaymath}
+    \begin{prooftree}
+      \hypo{t : \grR\gamma \vdash A_i}
+      \infer1{\AgdaInductiveConstructor{⊕I}_i~t : \grR\gamma \vdash A_0 \oplus A_1}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}
+      \hypo{p : \grR \leq \grP + \grQ}
+      \infer[no rule]1{s : \grP\gamma \vdash A \oplus B}
+      \hypo{t : \grQ\gamma, \gr1A &\vdash C}
+      \infer[no rule]1{u : \grQ\gamma, \gr1B &\vdash C}
+      \infer2{\AgdaInductiveConstructor{⊕E}~p~s~t~u : \grR\gamma \vdash C}
+    \end{prooftree}
+  \end{displaymath}
+  \begin{displaymath}
+    \begin{prooftree}
+      \hypo{p : \grR \leq \gr r\grP}
+      \hypo{t : \grP\gamma \vdash A}
+      \infer2{\AgdaInductiveConstructor{!I}~p~t : \grR\gamma \vdash \oc_{\gr r} A}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}
+      \hypo{p : \grR \leq \grP + \grQ}
+      \hypo{s : \grP\gamma \vdash \oc_{\gr r} A}
+      \infer[no rule]1{t : \grQ\gamma, \gr rA \vdash C}
+      \infer2{\AgdaInductiveConstructor{!E}~p~s~t : \grR\gamma \vdash C}
+    \end{prooftree}
+  \end{displaymath}
+  \caption{A fragment of \name{}}\label{fig:lr}
+\end{figure}
+
+\figref{fig:lr} presents a prototypical example of a system that our
+framework can represent.
+
+As a prototypical example of the kind of system that our framework can
+handle, we
+
+To motivate the way our framework represents
+
 Need to introduce:
@@
 Figure \ref{fig:lr}.
-
-\newcommand{\constr}[1]{\mathsf{#1}}
-
-\begin{figure}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR \leq \bra x}
-      \hypo{\gamma_x = A}
-      \infer2{\constr{var} : \grR\gamma \vdash A}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{t : \grR\gamma, \gr1A \vdash B}
-      \infer1{\constr{lolliI} : \grR\gamma \vdash A \multimap B}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \hypo{s : \grP\gamma &\vdash A \multimap B}
-      \infer[no rule]1{t : \grQ\gamma &\vdash A \phantom{{} \multimap B}}
-      \infer2{\constr{lolliE} : \grR\gamma \vdash B}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR\gamma \vdash A_i}
-      \infer1[$\oplus$-I$_i$]{\grR\gamma \vdash A_0 \oplus A_1}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \infer[no rule]1{\grP\gamma \vdash A \oplus B}
-      \hypo{\grQ\gamma, \gr1A &\vdash C}
-      \infer[no rule]1{\grQ\gamma, \gr1B &\vdash C}
-      \infer2[$\oplus$-E]{\grR\gamma \vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}
-      \hypo{\grR \leq \gr r\grP}
-      \hypo{\grP\gamma \vdash A}
-      \infer2[$\oc_{\gr r}$-I]{\grR\gamma \vdash \oc_{\gr r} A}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}
-      \hypo{\grR \leq \grP + \grQ}
-      \hypo{\grP\gamma \vdash \oc_{\gr r} A}
-      \infer[no rule]1{\grQ\gamma, \gr rA \vdash C}
-      \infer2[$\oc_{\gr r}$-E]{\grR\gamma \vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \caption{A fragment of \name{}}\label{fig:lr}
-\end{figure}
 
@@
 
-The $\constr{var}$ rule constructs variables,
-
-Nevertheless, there are several features of this system that are non
-standard. In the $\constr{var}$ rule,
-
-There are two key features
-
-first make a number of key observations on an exemplar syntax that
-will guide us.
+The $\AgdaInductiveConstructor{var}$ rule constructs variables,
 

### 2021-10-13 tex/esop22/syntax.tex
[ writing ] a bit more on the μμ~ syntax

@@
 This means that the abstract \AgdaBound{Ty} in the generic framework is
-instantiated to \AgdaDatatype{Conc} as below, with \AgdaDatatype{Ty} not being
-exposed directly to the generic framework.
+instantiated to \AgdaDatatype{Conc} (for \emph{conclusion}) as below, with
+\AgdaDatatype{Ty} not being exposed directly to the generic framework.
+For now, we just consider multiplicative disjunction $\parr$ (\emph{par}) and
+negation/duality, beside an uninterpreted base type.
+These are enough to exhibit classical behaviour.
 
@@
 
+With \AgdaBound{Ty} instantiated as \AgdaDatatype{Conc}, all terms are assigned
+\AgdaDatatype{Conc} type, as are all the variables.
+No variables are given \AgdaInductiveConstructor{com} type, similar to how in
+the bidirectional typing syntax of \citet[p.~25]{AACMM20}, no variables are
+given \AgdaInductiveConstructor{Check} type.
+How to observe this invariant is covered in the latter paper, so we will not
+repeat it here (having not yet seen how to write traverals on terms).
+
+The syntax comprises a \emph{cut} between a term and a coterm of the same type,
+the eponymous $\mu$ and $\tilde\mu$ constructs for proof by contradiction, and
+then term and coterm (introduction and elimination) forms for negation and
+\emph{par}.
+
 \ExecuteMetaData[\MuMuTildetex]{MMT}
+
+With a collection of pattern synonyms and the machinery from
+\cref{sec:usage-elaborator}, we can write an example term: a function which
+flips the disjuncts of a \emph{par}.
 

### 2021-10-14 tex/esop22/algebra.tex
[ writing ] start on μμ~ account
[ writing ] lots of stuff; intro and conclusion

@@
     \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
-    \item $0 + x = x \land x + 0 = x$
-    \item $(x + y) + z = x + (y + z)$
-    \item $x + y = y + x$
-    \item $1 * x = x \land x * 1 = x$
-    \item $(x * y) * z = x * (y * z)$
-    \item $0 * x = 0 \land x * 0 = 0$
-    \item $(x + y) * z = x * z + y * z$
-    \item $x * (y + z) = x * y + x * z$
+    \item $0 + x = x
+      \land x + 0 = x
+      \land (x + y) + z = x + (y + z)
+      \land x + y = y + x$
+    \item $1 * x = x
+      \land x * 1 = x
+      \land (x * y) * z = x * (y * z)$
+    \item $0 * x = 0
+      \land (x + y) * z = x * z + y * z$
+    \item $x * 0 = 0
+      \land x * (y + z) = x * y + x * z$
   \end{itemize}
@@
     \item $r \leq r' \to u \leq u' \to r \cdot u \leq r' \cdot u'$
-    \item $r \cdot 0_M = 0_M$
-    \item $r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$
-    \item $0 \cdot u = 0_M$
-    \item $\plr{r + s} \cdot u = r \cdot u +_M s \cdot u$
-    \item $1 \cdot u = u$
-    \item $\plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$
+    \item $r \cdot 0_M = 0_M
+      \land r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$
+    \item $0 \cdot u = 0_M
+      \land \plr{r + s} \cdot u = r \cdot u +_M s \cdot u$
+    \item $1 \cdot u = u
+      \land \plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$
   \end{itemize}
@@
 pattern matching.
-% I'm not sure this is true:
+% I'm not sure this is a real difference:
 %An auxiliary advantage is that with relations rather than functions, we can
@@
 %\AgdaBound{x} and \AgdaBound{y}, respectively.
+
+\begin{definition}
+  A \emph{(relational) linear map} $\Psi$ between modules $M$ and $N$ over a
+  posemiring $\Ann$ is a relation $\sim$ on the underlying sets of $M$ and $N$
+  satisfying the following laws.
+  \begin{itemize}
+    \item $u \leq u' \to v' \leq v \to u \sim v \to u' \sim v'$
+    \item $\forall v.~\plr{\exists u.~u \leq 0 \land u \sim v} \to v \leq 0$
+    \item $\forall u_0,u_1,v.~\plr{\exists u.~u \leq u_0 + u_1 \land u \sim v}
+      \to {}$\\$\plr{\exists v_0,v_1.~u_0 \sim v_0
+      \land u_1 \sim v_1 \land v \leq v_0 + v_1}$
+    \item $\forall r,u',v.~\plr{\exists u.~u \leq ru' \land u \sim v} \to
+      \plr{\exists v'.~u' \sim v' \land v \leq rv'}$
+    \item
+      $\forall u.~\exists v.~u \sim v \land \forall v'.~u \sim v' \to v' \leq v$
+  \end{itemize}
+\end{definition}
+
+Intuitively, $Q \sim P$, where $P$ and $Q$ are row vectors, is equivalent to
+$P \leq Q\Psi$, where $\Psi$ is the matrix representing the linear map and on
+the right is a vector-matrix multiplication.
+It is important that we think of \emph{row} vectors and
+\emph{right}-multiplication by a matrix because, without commutativity of the
+underlying posemiring, we can only expect $\plr{rQ}\Psi = r\plr{Q\Psi}$ and
+not $\Psi\plr{rQ} = r\plr{\Psi Q}$.
+In \cref{sec:env}, we use the matrix notation for convenience, while in the
+Agda code we see \ExecuteMetaData[\Snippetstex]{Psi-rel-P-Q}.
 
@@
 
-Operations like renaming and substitution are essentially translations from one
-context to another.
-When faced with two vector spaces arranged in this way, a natural thing to
-consider is the \emph{linear maps} from one space to the other.
+%Operations like renaming and substitution are essentially translations from one
+%context to another.
+%When faced with two vector spaces arranged in this way, a natural thing to
+%consider is the \emph{linear maps} from one space to the other.

### 2021-10-14 tex/esop22/appendix.tex
[ writing ] lots of stuff; intro and conclusion

--- 2021-09-20 tex/cpp21/env.tex
@@
-This section is a development of the work of \citet{WA20}.
-
-Let us write $\sdtstile{}\V$ as the infix version of $\V$.
-That is, $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$.
-The idea of a $\V$-environment is to extend the notion of $\V$-values from
-types to contexts.
-Specifically, where the judgement $\Gamma \sdtstile{}\V A$ says that we
-have a $\V$-value of type $A$ in context $\Gamma$, the judgement
-$\Gamma \env\V \Delta$ says that we have a $\V$-value for each entry in
-$\Delta$ (at the specified multiplicity), all in context $\Gamma$.
-For our purposes, it is essential that the ``all'' in the previous sentence is
-multiplicative --- we want to split the usage annotations of $\Gamma$ up into
-parts such that each part supports one $\V$-value.
-
-We have seen two kinds of values already: $\sqni$-values are variables and
-$\vdash$-values are terms.
-The corresponding environments are also standard concepts: $\sqni$-environments
-are \emph{simultaneous renamings} and $\vdash$-environments are
-\emph{simultaneous substitutions}.
-
-\subsection{Definition}
-
-\begin{definition}\label{def:lr-env}
-  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
-  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
-  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
-  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
-  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
-  $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.
-\end{definition}
-
-\begin{example}
-  We can form the identity renaming on a two-variable context.
-  \[
-    \id : \plr{\gr rA, \gr sB} \env\sqni \plr{\gr rA, \gr sB}
-  \]
-  Linear map $\gr\Psi$ is the identity map, clearly satisfying
-  \(
-    \begin{pmatrix} \gr r & \gr s \end{pmatrix} \leq
-    \begin{pmatrix} \gr r & \gr s \end{pmatrix}\gr\Psi
-  \).
-  When considering values, the fact that $\grPprime \leq \grQprime\gr\Psi$
-  reduces to $\grPprime \leq \grQprime$.
-  The two cases to consider are when $\grQprime\delta \sqni A$ and when
-  $\grQprime\delta \sqni B$.
-  In the first case, $\grPprime \leq \grQprime \leq
-  \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, so we have
-  $\grPprime\plr{A, B} \sqni A$.
-  The second case follows symmetrically.
-\end{example}
-
-\begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an environment (substitution) of type
-  \[
-    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
-    \plr{\gr1B, \gr2C}.
-  \]
-  Linear map $\gr\Psi$ is given by the matrix
-  \(
-    \begin{pmatrix}
-      \gr0 & \gr0 & \gr1 \\
-      \gr0 & \gr1 & \gr1
-    \end{pmatrix}
-  \),
-  noticing that $\gr1$ times the first row plus $\gr2$ times the second gives
-  the original $\grP$.
-  For $\grQprime = \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, we have
-  $\grPprime = \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$, and thus
-  $\grPprime\gamma \vdash z : B$.
-  For $\grQprime = \begin{pmatrix} \gr0 & \gr1 \end{pmatrix}$, we have
-  $\grPprime = \begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$, and thus
-  $\grPprime\gamma \vdash y\,z : C$.
-\end{example}
-
-As a mnemonic, one may use notation like the following to see what values are
-needed in the environment.
-\[
-  \begin{pmatrix}
-    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
-    \gr0A & \gr1\plr{B \multimap C} & \gr1B
-  \end{pmatrix}
-  \begin{matrix}
-    {} \vdash B \\
-    {} \vdash C
-  \end{matrix}
-\]
-This notation assumes that the notion of value supports subusaging, which is
-always the case when we are using environments for traversals.
-
-\begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an environment (renaming) of type
-  \[
-    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D} \env\sqni
-    \plr{\gr1C, \gr2A, \gr4A}.
-  \]
-  Linear map $\gr\Psi$ is given by the matrix
-  \(
-    \begin{pmatrix}
-      \gr0 & \gr0 & \gr1 & \gr0 \\
-      \gr1 & \gr0 & \gr0 & \gr0 \\
-      \gr1 & \gr0 & \gr0 & \gr0
-    \end{pmatrix}
-  \),
-  which we can check satisfies the required inequality.
-  The values are given by
-  \begin{align*}
-    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
-    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
-    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A.
-  \end{align*}
-
-  We can read from the columns of the matrix what happened to each of the
-  variables in $\Gamma$.
-  The first column, corresponding to variable $\gr6a : A$, contains two $\gr1$s
-  because it has been duplicated (via contraction).
-  Meanwhile, the second and fourth columns are all $\gr0$ because variables
-  $b$ and $d$ have been discarded (via weakening).
-  The third column contains one $\gr1$ because $c$ is used once.
-  This $\gr1$ appears above the $\gr1$s to its left because $c$ has been
-  permuted (via exchange) past $a$.
-  Each of the rows in the matrix is a basis vector because variables can only
-  be formed in contexts with basis annotations or less.
-\end{example}
-
-\subsection{Properties}
-
-\begin{lemma}
-  \cref{def:lr-env} is sound and complete for the following syntax.
-  \begin{displaymath}
-    \begin{prooftree}[comb,center=false]
-      \hypo{I^*}
-      \infer1{\alr{} : {} \env\V {\cdot}}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb,center=false]
-      \hypo{\rho : {} \env\V \Delta_l}
-      \hypo{\sep}
-      \hypo{\sigma : {} \env\V \Delta_r}
-      \infer3{\alr{\rho,\sigma} : {} \env\V \Delta_l, \Delta_r}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb,center=false]
-      \hypo{\gr r\cdot\plr{M : {} \sdtstile{}\V A}}
-      \infer1{\alr{M} : {} \env\V \gr rA}
-    \end{prooftree}
-  \end{displaymath}
-\end{lemma}
-\begin{proof}
-  This replaces \cref{thm:construct-env}.
-\end{proof}
-
-\begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an environment (substitution) of type
-  \[
-    \alr{\alr{z},\alr{y\,z}} :
-    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
-    \plr{\gr1B, \gr2C}.
-  \]
-  We rely on the observations that
-  $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix} =
-  \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}
-  + \begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix}$ and, on the right, that
-  $\begin{pmatrix} \gr0 & \gr2 & \gr2 \end{pmatrix} =
-  \gr2\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
-  Then, we have $\gr0x : A, \gr0y : B \multimap C, \gr1z : B \vdash z : B$ and
-  $\gr0x : A, \gr1y : B \multimap C, \gr1z : B \vdash y\,z : C$.
-\end{example}
-
-When constructing an environment, we can do so by cases on the shape of the
-target context.
-We can create an environment into the empty context when all usage annotations
-on the source context are $\gr0$.
-We can create an environment into a concatenated context when we can additively
-split up the annotations of the source context and produce environments into
-both halves from the split sources.
-We can create an environment into a singleton context when there is a context
-$\gr r$ times smaller than the source context in which we can produce a value
-of the appropriate type.
-
-\begin{lemma}\label{thm:construct-env}
-  We can define all of the following equivalences for any values of the free
-  variables, assuming that $\V$ respects subusaging (i.e.,
-  $\grPprime \leq \grP \to
-  \forallb{\V\,\grP\gamma \dotto \V\,\grPprime\gamma}$).
-  \begin{itemize}
-    \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
-    \item $\forallb{\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
-      \dotlr \plr{{-} \env\V \Delta_l, \Delta_r}}$
-    \item
-      $\forallb{\gr r \cdot \plr{\V\,(-)\,A} \dotlr \plr{{-} \env\V \gr rA}}$
-  \end{itemize}
-\end{lemma}
+%thm:env-rules
 \begin{proof}
@@
 
-We could, indeed, use these three clauses to define what an environment is.
-However, I find them difficult to work with, as it is often easier to do
-linear algebraic proofs separately from the rest of an environment.
-For identity and composition, as we are about to see, the original definition
-is easier to use because we can rely on the identity and composition of linear
-maps.
-Concretely, an inductive proof of identity would, for example, involve
-constructing an environment of type
-$\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
-environments of types $\grP\gamma, \gr0\delta \env\V \grP\gamma$ and
-$\gr0\gamma, \grQ\delta \env\V \grQ\delta$.
-These are not identity environments, so we would have to strengthen the
-induction hypothesis.
-
-\begin{lemma}\label{thm:env-resize}
-  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
-  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
-  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
-  with the same linear map and action on variables.
-\end{lemma}
-\begin{proof}
-  The only part of the definition of an environment dependent on $\grP$ or
-  $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
-  replace for $\grPprime$ and $\grQprime$.
-\end{proof}
-
-One of the primary test cases for environments is simultaneous substitution,
-which will look like the following rule.
-The admissibility of substitution will be by induction on the derivation of
-$\Delta \vdash A$, so we will need to be able to adapt any environment we are
-given to work with any possible context of new premises.
-In the simply typed case, the only change to the context we encountered was the
-binding of new variables.
-Now, with usage annotations, we furthermore have linear decompositions of the
-context, necessitating changes to the environment whenever usage annotations
-change.
-I will deal first with linear decompositions.
-
-\begin{displaymath}
-  \begin{prooftree}
-    \hypo{\Gamma \env{\vdash} \Delta}
-    \hypo{\Delta \vdash A}
-    \infer2[sub]{\Gamma \vdash A}
-  \end{prooftree}
-\end{displaymath}
-
-There are three kinds of linear decompositions we have to deal with: zero,
-addition, and scaling; corresponding to bunched connectives $I^*$, $\sep$, and
-$\gr r \cdot {}$, respectively.
-In each case, we have a simple preservation lemma, transforming an environment
-of type $\Gamma \env\V \Delta$ and a decomposition of $\Delta$ into a
-decomposition of $\Gamma$ and environments for all of the decomposed fragments
-of $\Gamma$ and $\Delta$.
-
-\begin{lemma}[environments preserve zero]\label{thm:lr-env-zero}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \gr 0$, we also have that $\grP \leq \gr 0$.
-\end{lemma}
+%thm:lr-env-zero
 \begin{proof}
@@
 
-\begin{lemma}[environments preserve addition]\label{thm:lr-env-add}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \grQl + \grQr$ for some $\grQl$ and $\grQr$, we also have $\grPl$
-  and $\grPr$ such that $\grP \leq \grPl + \grPr$ and there are environments
-  of types $\grPl\gamma \env\V \grQl\delta$ and
-  $\grPr\gamma \env\V \grQr\delta$.
-\end{lemma}
+%thm:lr-env-add
 \begin{proof}
@@
 
-\begin{lemma}[environments preserve scaling]\label{thm:lr-env-scale}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \gr r\grQprime$ for some $\grQprime$, we also have a $\grPprime$
-  such that $\grP \leq \gr r\grPprime$ and there is an environment of type
-  $\grPprime\gamma \env\V \grQprime\delta$.
-\end{lemma}
+%thm:lr-env-scale
 \begin{proof}
@@
 
-Finally, I will also take the opportunity to give the extend lemma, allowing
-environments to incorporate newly bound variables.
-\todo{Motivate}
-In the intuitionistic case, the extend lemma had two requirements on $\V$: $\V$
-admits weakening and we can map variables into $\V$-values.
-With usage annotations, the former is unreasonable, but it turns out that we
-only need weakening by variables whose usage annotation is less than or equal
-to $\gr0$.
-The latter stays as-is, with the note that ``variable'' now means a
-usage-checked variable.
-
-\begin{lemma}[extend]\label{thm:lr-bind}
-  Given functions
-  ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
-  \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
-  $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
-  type $\Gamma \env\V \Delta$ into an environment of type
-  $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
-\end{lemma}
+%thm:lr-bind
 \begin{proof}
@@
 
-\subsection{Identity and composition}
-
-The requirements for identity and composition of environments look a bit like
-the unit and lift of a Kleisli triple.
-
-\begin{lemma}[Identity environment]\label{thm:env-id}
-  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
-  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
-\end{lemma}
+%thm:env-id
 \begin{proof}
@@
 \end{proof}
-
-\begin{lemma}[Composition of environments]\label{thm:env-comp}
-  Given a function
-  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
-  \forall \grPprime, \grQprime.~\grPprime \leq \grQprime\plr{\rho.\gr\Psi} \to
-  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
-  compose environments of types $\Gamma \env\U \Delta$ and
-  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
-\end{lemma}
-
-\begin{corollary}\label{thm:env-postren}
-  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
-  $\sigma : \Delta \env\sqni \Theta$, we can get an environment of type
-  $\Gamma \env\U \Theta$.
-  When $\U = {\sqni}$, this gives composition of renamings.
-\end{corollary}
-\begin{proof}
-  By \cref{thm:env-comp}, with $\mathrm{lift}$ being given exactly by the
-  action of $\rho$ on variables.
-\end{proof}
-
-\todo{What do these choices mean?}
-\begin{example}
-  We can derive the following instances of environment composition.
-  \begin{itemize}
-    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
-      syntactic traversal.
-      For example, if $\U = {\sqni}$, we need the action of renaming on terms
-      to show that a renaming followed by a substitution composes to a
-      substitution.
-      If $\U = {\vdash}$, then the action of substitution on terms gives us that
-      substitutions compose.
-    \item More generally, if $\V = {\vdash}$ and we have a semantics from
-      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
-      of terms.
-  \end{itemize}
-\end{example}

### 2021-10-14 tex/esop22/conc.tex
[ writing ] lots of stuff; intro and conclusion

We have presented a framework for doing metatheory for a class of substructural
type systems in Agda.
The framework gives us renaming, substitution, and a usage elaborator for new
syntaxes for free, which we hope can facilitate prototyping and the
mechanisation of more interesting semantic results.
Beside the mechanised framework itself, we believe its methodology --- the use
of bunched premise combinators --- can guide and simplify the development of
(potentially unmechanised) substructural type systems.

Our account of substructurality is based on the linear algebraic
principles described by \citet{WA20}.
However, these details only really affect the definition of environment,
in which the use of linear maps is motivated by them being the standard notion
of morphism between vectors.
We could imagine that a similar notion of morphism is found for the kind of
annotations found in \citet{LicataSR17}, allowing a framework to consider
finer substructural systems.

### 2021-10-14 tex/esop22/env.tex
[ writing ] lots of stuff; intro and conclusion

@@
 That is, $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$.
+In this section, we introduce \emph{$\V$-environments}, which will form a key
+part of the generic traversal of \cref{sec:semantics}.
 The idea of a $\V$-environment is to extend the notion of $\V$-values from
@@
 
+When constructing an environment, we can do so by cases on the shape of the
+target context.
+We can create an environment into the empty context when all usage annotations
+on the source context are $\gr0$.
+We can create an environment into a concatenated context when we can additively
+split up the annotations of the source context and produce environments into
+both halves from the split sources.
+We can create an environment into a singleton context when there is a context
+$\gr r$ times smaller than the source context in which we can produce a value
+of the appropriate type.
+
 \begin{lemma}\label{thm:env-rules}
-  \cref{def:lr-env} is sound and complete for the following syntax.
+  \Cref{def:lr-env} is sound and complete for the following syntax.
   \begin{displaymath}
@@
 \end{lemma}
-\begin{proof}
-  This replaces \cref{thm:construct-env}.
-\end{proof}
 
@@
 
-When constructing an environment, we can do so by cases on the shape of the
-target context.
-We can create an environment into the empty context when all usage annotations
-on the source context are $\gr0$.
-We can create an environment into a concatenated context when we can additively
-split up the annotations of the source context and produce environments into
-both halves from the split sources.
-We can create an environment into a singleton context when there is a context
-$\gr r$ times smaller than the source context in which we can produce a value
-of the appropriate type.
-
-\begin{lemma}\label{thm:construct-env}
-  We can define all of the following equivalences for any values of the free
-  variables, assuming that $\V$ respects subusaging (i.e.,
-  $\grPprime \leq \grP \to
-  \forallb{\V\,\grP\gamma \dotto \V\,\grPprime\gamma}$).
-  \begin{itemize}
-    \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
-    \item $\forallb{\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
-      \dotlr \plr{{-} \env\V \Delta_l, \Delta_r}}$
-    \item
-      $\forallb{\gr r \cdot \plr{\V\,(-)\,A} \dotlr \plr{{-} \env\V \gr rA}}$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  There are 6 cases to check.
-  Throughout, we write $\Gamma$ as $\grP\gamma$ and $\Delta$ as $\grQ\delta$
-  when convenient.
-  \begin{description}
-    \item[$I(\to)$]
-      Let $\gr\Psi$ be the unique linear map out of the zero space.
-      By assumption and definition, $\grP \leq \gr0 = \grQ\gr\Psi$.
-      There are no variables to act upon.
-    \item[$I(\gets)$]
-      $\grQ\gr\Psi$ is an empty sum, so if $\grP \leq \grQ\gr\Psi$ then
-      $\grP \leq \gr0$.
-    \item[$\sep(\to)$]
-      Let the given environments be $\rho_l : \grPl\gamma \env\V \grQl\delta$
-      and $\rho_r : \grPr\gamma \env\V \grQr\delta$, with
-      $\grP \leq \grPl + \grPr$.
-      Define $\gr\Psi \coloneqq [\rho_l.\gr\Psi, \rho_r.\gr\Psi]$, using the
-      coproduct structure of the concatenated vector space.
-      We have $\grP \leq \grPl + \grPr \leq
-      \grQl\plr{\rho_l.\gr\Psi} + \grQr\plr{\rho_r.\gr\Psi} =
-      \begin{pmatrix} \grQl & \grQr \end{pmatrix}\gr\Psi$.
-      To act on variables, we are given $\grPprime \leq
-      \begin{pmatrix} \gr{\grQ'_l} & \gr{\grQ'_r} \end{pmatrix}\gr\Psi$ and
-      $\gr{\grQ'_l}\delta_l, \gr{\grQ'_r}\delta_r \sqni A$.
-      Without loss of generality, let us have $\gr{\grQ'_l}\delta_l \sqni A$
-      and $\gr{\grQ'_r} \leq \gr0$.
-      Thus, $\grPprime \leq
-      \gr{\grQ'_l}\plr{\rho_l.\gr\Psi} + \gr{\grQ'_r}\plr{\rho_r.\gr\Psi} \leq
-      \gr{\grQ'_l}\plr{\rho_l.\gr\Psi}$,
-      and we can act on the variable using $\rho_l$.
-    \item[$\sep(\gets)$]
-      Let the unnamed context be $\Gamma$, also written $\grP\gamma$.
-      The linear map
-      $\gr\Psi : \Ann^{\size{\Delta_l} + \size{\Delta_r}} \to \Ann^{\size\Gamma}$
-      splits into
-      $\gr\Psi_{\gr l} : \Ann^{\size{\Delta_l}} \to \Ann^{\size\Gamma}
-      \coloneqq \alr{\id, 0}; \gr\Psi$ and
-      $\gr\Psi_{\gr r} : \Ann^{\size{\Delta_r}} \to \Ann^{\size\Gamma}
-      \coloneqq \alr{0, \id}; \gr\Psi$, using the product structure of
-      the concatenated vector space.
-      Let $\grPl \coloneqq \grQl\gr\Psi_{\gr l}$ and
-      $\grPr \coloneqq \grQr\gr\Psi_{\gr r}$, by definition satisfying the
-      required constraints.
-      For the action on variables, let us consider the left environment (with
-      the right environment following symmetrically).
-      We are given $\gr{\grP'_l} \leq \gr{\grQ'_l}\gr\Psi_{\gr l}$ and
-      $\gr{\grQ'_l}\delta_l \sqni A$.
-      From these, we get
-      $\gr{\grP'_l} \leq \gr{\grQ'_l}\gr\Psi_{\gr l} =
-      \begin{pmatrix} \gr{\grQ'_l} & \gr0 \end{pmatrix}\gr\Psi$ and
-      $\gr{\grQ'_l}\delta_l, \gr0\delta_r \sqni A$.
-      We can therefore act using the original environment.
-    \item[$\cdot(\to)$]
-      Let $\grP$ and $\grPprime$ be such that $\grP \leq \gr r\grPprime$ and let
-      $v : \V\,\grPprime\gamma\,A$.
-      Let $\gr\Psi : \Ann \to \Ann^{\size\gamma}
-      \coloneqq \gr r\gr' \mapsto \gr r\gr'\grPprime$.
-      By definition and the previous assumption, we have
-      $\grP \leq \gr r\gr\Psi$.
-      When acting on a variable, we have $\grP\gr{''} \leq \gr r\gr'\gr\Psi$
-      and $\gr r\gr'A \sqni A'$.
-      The latter tells us that $A = A'$ and $\gr r\gr' \leq \gr1$.
-      Thus, $\grP\gr{''} \leq \grPprime$.
-      Therefore, by subusaging, we may produce a value of type
-      $\V\,\grPprime\gamma\,A$, which we can take to be $v$.
-    \item[$\cdot(\gets)$]
-      Let us have an environment of type $\grP\gamma \env\V \gr rA$.
-      We want to use its action on variables to yield a value.
-      To do this, we let $\grPprime \coloneqq \gr1\gr\Psi$, and use this
-      equation, together with the fact that we have a variable of type
-      $\gr1A \sqni A$, to get a value of type $\V\,\grPprime\gamma\,A$.
-      Furthermore, we derive $\grP \leq \gr r\gr\Psi = \gr r\grPprime$, as
-      required.
-  \end{description}
-\end{proof}
-
-We could, indeed, use these three clauses to define what an environment is.
-However, I find them difficult to work with, as it is often easier to do
+We could, indeed, use this syntax to define what an environment is.
+However, we find them difficult to work with, as it is often easier to do
 linear algebraic proofs separately from the rest of an environment.
-For identity and composition, as we are about to see, the original definition
+For example, for identity and composition of environments, the original
+definition
 is easier to use because we can rely on the identity and composition of linear
@@
 change.
-I will deal first with linear decompositions.
+We deal first with linear decompositions.
 
@@
 \end{lemma}
-\begin{proof}
-  $\grP \leq \grQ\gr\Psi \leq \gr0\gr\Psi = \gr0$, by environment
-  compatibility and monotonicity and linearity of $\gr\Psi$.
-\end{proof}
 
@@
 \end{lemma}
-\begin{proof}
-  Let $\grPl \coloneqq \grQl\gr\Psi$ and $\grPr \coloneqq \grQr\gr\Psi$.
-  Then, $\grP \leq \grQ\gr\Psi \leq \plr{\grQl + \grQr}\gr\Psi =
-  \grQl\gr\Psi + \grQr\gr\Psi = \grPl + \grPr$, satisfying the first condition.
-  Because clearly $\grPl \leq \grQl\gr\Psi$ and $\grPr \leq \grQr\gr\Psi$,
-  \cref{thm:env-resize} on the original environment gives us the required
-  pair of new environments.
-\end{proof}
 
@@
 \end{lemma}
-\begin{proof}
-  Let $\grPprime \coloneqq \grQprime\gr\Psi$.
-  Then, $\grP \leq \grQ\gr\Psi \leq \plr{\gr r\grQprime}\gr\Psi =
-  \gr r\plr{\grQprime\gr\Psi} = \gr r\grPprime$, satisfying the first condition.
-  Because clearly $\grPprime \leq \grQprime\gr\Psi$,
-  \cref{thm:env-resize} on the original environment gives us the required
-  new environment.
-\end{proof}
 
-Finally, I will also take the opportunity to give the extend lemma, allowing
+Finally, we also take the opportunity to give the extend lemma, allowing
 environments to incorporate newly bound variables.
@@
 \end{lemma}
-\begin{proof}
-  Let $\grP\gamma \coloneqq \Gamma$, $\grQ\delta \coloneqq \Delta$, and
-  $\grR\theta \coloneqq \Theta$.
-  Let the new linear map $\gr\Psi\gr' : \Ann^{\size\Delta + \size\Theta} \to
-  \Ann^{\size\Gamma + \size\Theta}$ be $\gr\Psi \oplus \gr I$.
-  That is, in block matrix notation,
-  $\begin{pmatrix} \gr\Psi & \gr0 \\ \gr0 & \gr I \end{pmatrix}$.
-  Checking that this linear map fits, we have
-  $\begin{pmatrix}\grP & \grR\end{pmatrix}
-  \leq \begin{pmatrix}\grQ\gr\Psi & \grR\gr I\end{pmatrix}
-  = \begin{pmatrix}\grQ & \grR\end{pmatrix}\plr{\gr\Psi \oplus \gr I}$.
-  For the action on variables, we are given vectors $\grPprime$,
-  $\grR\gr'_\grP$, $\grQprime$, and $\grR\gr'_\grQ$ such that
-  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix} \leq
-  \begin{pmatrix} \grQprime & \grR\gr'_\grQ \end{pmatrix}
-  \plr{\gr\Psi \oplus \gr I}$ and we have a variable of type
-  $\grQprime\delta, \grR\gr'_\grQ\theta \sqni A$ for some type $A$.
-  The constraint on the new vectors reduces to $\grPprime \leq \grQprime\gr\Psi$
-  and $\grR\gr'_\grP \leq \grR\gr'_\grQ$.
-  From the variable we either have a variable $x$ in $\delta$ with
-  $\grQprime \leq \langle x \rvert$ and $\grR\gr'_\grQ \leq \gr0$, or a
-  variable $y$ in $\theta$ with $\grQprime \leq \gr0$ and
-  $\grR\gr'_\grQ \leq \langle y \rvert$.
-  In the former case, the action of the original environment on $x$ gives us a
-  $\V$-value in $\grPprime\gamma$, and the $\gr0$-weakening principle
-  $\swarrow^k$, noting that $\grR\gr'_\grP \leq \grR\gr'_\grQ \leq \gr0$, gives
-  us a $\V$-value in $\grPprime\gamma, \grR\gr'_\grP\theta$.
-  In the latter case, we have that
-  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix}
-  \leq \begin{pmatrix} \grQprime\gr\Psi & \grR\gr'_\grQ \end{pmatrix}
-  \leq \begin{pmatrix} \gr0\gr\Psi & \langle y \rvert \end{pmatrix}
-  = \begin{pmatrix} \gr0 & \langle y \rvert \end{pmatrix}
-  = \left\langle {\searrow}y \right\rvert$, so $y$ also serves as a
-  usage-checked variable in $\grPprime\gamma, \grR\gr'_\grP\theta$.
-  From this usage-checked variable, we get a $\V$-value in the same context
-  using $\mathrm{vr}$.
-\end{proof}
 
@@
 \end{lemma}
-\begin{proof}
-  Let $\gr\Psi$ be the identity map, which clearly satisfies
-  $\grP \leq \grP\gr\Psi$.
-  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
-  means that $\grPprime \leq \grQprime$, so the variable of type
-  $\grQprime\gamma \sqni A$ can be coerced to have type
-  $\grPprime\gamma \sqni A$.
-  From this point, $\mathrm{vr}$ gives us a value of type
-  $\V\,\grPprime\gamma\,A$, as required.
-\end{proof}
-
-\begin{lemma}\label{thm:env-comp-lemma}
-  Given an environment $\rho : \Gamma \env\U \Delta$ for which we have, for any
-  $\grPprime$ and $\grQprime$ such that
-  $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$, we have a function
-  $\mathrm{lift}_\rho :
-  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$,
-  we can map environments of type $\Delta \env\V \Theta$ into environments of
-  type $\Gamma \env\W \Theta$.
-\end{lemma}
-\begin{proof}
-  Let $\rho$ be as in the statement, and let $\sigma : \Delta \env\V \Theta$.
-  For the environment we are constructing, let
-  $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
-  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
-  \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
-  For the action on variables, we are given $\grPprime \leq \grRprime\gr\Psi$
-  with $\grRprime\theta \sqni A$.
-  We can immediately apply the action of $\sigma$, giving us a value of type
-  $\V\,\plr{\grRprime\plr{\sigma.\gr\Psi}}\,A$.
-  We note that
-  $\grPprime \leq \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
-  apply $\mathrm{lift}_\rho$ to get the desired value.
-\end{proof}
 

### 2021-10-14 tex/esop22/examples.tex
[ writing ] lots of stuff; intro and conclusion

@@
 
-In this subsection, we take the usage annotations to be the 4-element variance
-posemiring.
-\todo{This works for any semiring.}
+As a running example, we take the usage annotations to be the 4-element
+variance posemiring.
+\todo{Define earlier}
 We establish the property that all terms are monotonic in their free variables.
@@
 We will take semantics of this system into
-\emph{worldly relations}~\cite{AbelBernardy2020}.
-A worldly relation over a poset of worlds \AgdaBound{W} is a set over which
+\emph{world-indexed relations}~\cite{AbelBernardy2020}.
+A world-indexed relation over a poset of worlds \AgdaBound{W} is a set over
+which
 we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
@@
 \begin{example}
-  When \AgdaBound{W} is the 1-element set, a worldly relation is just a set
-  equipped with a binary relation.
+  When \AgdaBound{W} is the 1-element set, a world-indexed relation is just a
+  set equipped with a binary relation.
 \end{example}
 
-Morphisms between worldly relations \AgdaBound{R} and \AgdaBound{S} consist of
-a mapping between the underlying sets such that that mapping preserves
-relatedness from \AgdaBound{R} to \AgdaBound{S}.
+Morphisms between world-indexed relations \AgdaBound{R} and \AgdaBound{S}
+consist of a mapping between the underlying sets such that that mapping
+preserves relatedness from \AgdaBound{R} to \AgdaBound{S}.
 
@@
 \todo{Define big intersection.}
-When the poset of worlds forms a (relational) commutative monoid, such worldly
-relations support a symmetric monoidal closed structure.
-We reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
+When the poset of worlds forms a (relational) commutative monoid, such
+world-indexed relations support a symmetric monoidal closed structure, with
+objects denoted \AgdaFunction{I$^R$},
+\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
+\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
+These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
 \AgdaRecord{$\wand$}, now over worlds rather than contexts.
 
-\ExecuteMetaData[\WReltex]{IR}
-\ExecuteMetaData[\WReltex]{tensorR}
-\ExecuteMetaData[\WReltex]{lollyR}
+%\ExecuteMetaData[\WReltex]{IR}
+%\ExecuteMetaData[\WReltex]{tensorR}
+%\ExecuteMetaData[\WReltex]{lollyR}
 
 The final piece of sematics we need is a \emph{bang} operator.
-\todo{No instead}
-Instead of requiring extra algebraic structure on the worlds, we allow the
-semantic \emph{bang} to be an arbitrary annotation-indexed functor on worldly
-relations.
+We allow the
+semantic \emph{bang} to be an arbitrary annotation-indexed functor on
+world-indexed relations.
 This functor must respect all of the structure on the indices, making it a
@@
 
-To associate semantics to syntax, we start as standard by associating worldly
-relations to types.
+To associate semantics to syntax, we start as standard by associating
+world-indexed relations to types.
 We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
@@
 annotations given by \AgdaBound{R}.
+The protocol this encodes is that the user will provide an unannotated term
+together with a candidate usage context \AgdaBound{R}, and usage elaboration
+returns a list of possible ways the term could be annotated such that the
+conclusion has usage context \AgdaBound{R}.
 The module name \AgdaModule{U} refers to the fact that we are taking the
@@
 unannotated system \AgdaFunction{uSystem}\AgdaSpace{}\AgdaBound{sys}.
-We already know that variables are renameable.
+To write it, we make use of idiom brackets
+\AgdaSymbol{(|}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaSymbol{|)},
+which have the effect of replacing top-level spines of applications by
+(\AgdaDatatype{List}-)applicative applications.
+Field by field, we already know that variables are renameable.
 To interpret a variable, we consider all the possible proofs that such a
-variable could be well annotated, and package them up as a variable term.
+variable could be well annotated, and package them up as a variable term via
+the applicative machinery.
+Finally, for compound terms, we first reify the unannotated subterms, and then
+combine the subterms via a \AgdaFunction{lemma}.
 
 \ExecuteMetaData[\UsageChecktex]{elab-sem}
+
+The \AgdaFunction{lemma} essentially goes through the shape of the premises,
+combining the collections of subterms in the natural way.
+For example, at each
+\AgdaInductiveConstructor{\AgdaUnderscore{}$\dottimes$\AgdaUnderscore{}},
+we take the Cartesian product of the possibilities of each half, and at each
+\AgdaInductiveConstructor{\AgdaUnderscore{}$\sep$\AgdaUnderscore{}},
+we non-deterministically split the usage annotations coming in, and then take
+the Cartesian product.
+When it comes to newly bound variables, the \emph{syntax description} tells us
+their annotations, so there is no further non-determinism introduced here.
 
@@
 \emph{renaming} in this case, because $\V$ is a family of variables).
+In the previous examples, we have used \AgdaFunction{elab-unique}, which further
+checks statically that exactly one derivation is returned.
 The candidate usage vector \AgdaBound{R} will be empty for closed terms, and

### 2021-10-14 tex/esop22/intro.tex
[ writing ] lots of stuff; intro and conclusion

\begin{itemize}
  \item Review standard linear algebra constructions in the posemiring setting.
  \item Give a general definition of simple type systems with usage annotations
    based on connectives from bunched logic and constructions from linear
    algebra.
  \item Give some novel examples of such systems.
  \item Derive admissibility of renaming and substitution for all such systems.
  \item Give a usage checker generic in syntax.
\end{itemize}

In this paper, we concern the metatheory of a class of substructural type
systems related to linear logic~\cite{girard87linear}.
This class is variously known as
\emph{coeffectful}~\cite{PetricekOM14,Granule18},
\emph{quantitative}~\cite{BrunelGMZ14,Atkey18}, or
\emph{resource-aware}~\cite{GhicaS14},
or is given no particular name~\cite{reed10distance,abadi99core},
and generalises bounded linear logic to track variable usage with semiring
annotations.
In all of these systems, we have some ambient semiring $\Ann$, and in the
judgements of the type system, variables are annotated by elements of $\Ann$
describing \emph{how} that variable can be used.
The additive structure of $\Ann$ gives the ability to count, or otherwise
accumulate, usages of variables in multiple subterms.
The multiplicative structure gives rise to a form of modality, for example
allowing multiple or unlimited reuse, or movement between security levels, in
the type system.

The aspect of such systems we tackle here is their basic metatheory and
mechanisation thereof.

We build upon both the general structural framework of
\citet{AACMM20} and the substructural techniques of \citet{WA20}.
The way \citeauthor{AACMM20} consolidate and codify mechanisation techniques for
propositional natural deduction systems based on intrinsically typed syntax and
de Bruijn indices, we aim to replicate for linear-like systems based on
semiring usage annotations.
By picking a trivial semiring, our work can subsume that of
\citeauthor{AACMM20}, except for the many pieces of machinery we have not yet
ported to this new framework.

Our work complements that of \citet{Granule18} on the Granule programming
language.
Where Granule focuses on writing programs \emph{in} the language and running
them, we focus on metatheoretic reasoning \emph{about} type systems.
%As such, whereas Granule has a convenient syntax, performant type-checker,
%interpreter

Our work is similar in scope to that of \citet{LicataSR17}, though we work in
a natural deduction style rather than a sequent calculus style.
Where \citeauthor{LicataSR17} are much more agnostic in terms of
substructurality --- allowing for non-commutative and bunched logics ---
we are much more agnostic in terms of syntax.
The system of \citeauthor{LicataSR17} is essentially a single calculus,
supporting ``product'' ($\mathrm F$) types and ``function'' ($\mathrm U$)
types, parametrised on a \emph{mode theory} describing its structural rules.
For this system, they derive the strong result of cut elimination.
Meanwhile, we leave syntax design to the user, and consequently can only
guarantee substitution (which we can only get because of our commitment to
natural deduction).

### 2021-10-14 tex/esop22/quant-framework.tex
[ writing ] New section: bunched typing rules
[ writing ] start on μμ~ account
[ writing ] a miserably small amount
[ writing ] lots of stuff; intro and conclusion

@@
 \usepackage{turnstile}
+\usepackage{mathpartir}
 
@@
 \def\Monotonicitytex{\genericlr/Generic/Linear/Example/WRel/Monotonicity.tex}
+\def\MuMuTildetex{\genericlr/Generic/Linear/Example/MuMuTilde.tex}
+\def\MuMuTildeTermtex{\genericlr/Generic/Linear/Example/MuMuTilde/Term.tex}
 \def\Snippetstex{../../agda/processed-latex/Snippets.tex}
+
+\newcommand{\figref}[1]{Figure \ref{#1}}
 
@@
 %
-\author{James Wood\inst{1}\orcidID{0000-0002-8080-3350} \and
-Robert Atkey\inst{1}\orcidID{0000-0002-4414-5047}}
+\author{Anonymous authors}
+%\author{James Wood\inst{1}\orcidID{0000-0002-8080-3350} \and
+%  Robert Atkey\inst{1}\orcidID{0000-0002-4414-5047}}
 %
-\authorrunning{J. Wood, R. Atkey}
+\authorrunning{Anon.}
+%\authorrunning{J. Wood, R. Atkey}
 % First names are abbreviated in the running head.
@@
 %
-\institute{University of Strathclyde, Glasgow, UK
-\email{\{james.wood.100,robert.atkey\}@strath.ac.uk}}
+%\institute{University of Strathclyde, Glasgow, UK
+%\email{\{james.wood.100,robert.atkey\}@strath.ac.uk}}
 %
@@
   the additive and multiplicative features of linear logic.
-  The bunched flavour is carried over into the semantics, together with the use
-  of basic linear algebra constructs.
-  For example, \emph{environments} are presented equivalently as values
-  accumulated via separating conjunction, and as functions from variables to
-  values supported by linear maps.
-  Producing a generic semantic traversal has us combine environments with a
-  separating implication, producing a Kripke function space of the form
-  $\Box(A \wand B)$.
-  From the generic semantic traversal, we derive totally generic renaming and
-  substitution operations, a specific denotational semantics, and a
-  syntax-generic \emph{usage elaborator} which greatly facilitates writing
-  concrete terms.
+  We use the uniformity of syntaxes to derive type system-generic renaming,
+  substitution, and a form of linearity checking.
+  %The bunched flavour is carried over into the semantics, together with the use
+  %of basic linear algebra constructs.
+  %For example, \emph{environments} are presented equivalently as values
+  %accumulated via separating conjunction, and as functions from variables to
+  %values supported by linear maps.
+  %Producing a generic semantic traversal has us combine environments with a
+  %separating implication, producing a Kripke function space of the form
+  %$\Box(A \wand B)$.
+  %From the generic semantic traversal, we derive totally generic renaming and
+  %substitution operations, a specific denotational semantics, and a
+  %syntax-generic \emph{usage elaborator} which greatly facilitates writing
+  %concrete terms.
 
@@
 
+\section{Bunched Typing Rules}\label{sec:bunched-rules}
+\input{bunchedrules}
+
 \section{Generic syntax}\label{sec:syntax}
@@
 \section{Conclusions}\label{sec:conc}
+\input{conc}
 %

### 2021-10-14 tex/esop22/syntax.tex
[ writing ] lots of stuff; intro and conclusion

@@
 
-\subsection{Other syntaxes}
+\subsection{Other syntaxes and syntactic forms}
 
+\paragraph{$\mu\tilde\mu$}
 We can encode a usage-annotated version of System $L$/the
@@
 
-With a collection of pattern synonyms and the machinery from
-\cref{sec:usage-elaborator}, we can write an example term: a function which
-flips the disjuncts of a \emph{par}.
+%With a collection of pattern synonyms and the machinery from
+%\cref{sec:usage-elaborator}, we can write an example term: a function which
+%flips the disjuncts of a \emph{par}.
 
-\ExecuteMetaData[\MuMuTildeTermtex]{patterns}
-\ExecuteMetaData[\MuMuTildeTermtex]{myComm}
+%\ExecuteMetaData[\MuMuTildeTermtex]{patterns}
+%\ExecuteMetaData[\MuMuTildeTermtex]{myComm}
+
+\paragraph{Duplicability}
+There is one more bunched combinator we have experimented with adding to the
+framework:
+
+\[
+  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
+  \times \plr{\grRprime \leq \gr0}
+  \times \plr{\grRprime \leq \grRprime + \grRprime}
+  \times T\,\grRprime
+\]
+
+The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
+of it, can be both discarded and duplicated indefinitely, and in the
+refinement we have a $T$.
+We use this combinator to introduce subterms that are used an unknown number of
+times, for example the continuations of the eliminator of an inductive type,
+or other fixed points.
+We can also use it in linear/non-linear style systems~\cite{Benton94} to make
+sure linear variables are not available in the intuitionistic fragment.
+
+Adding the $\Box$ combinator is the only thing we have found that requires our
+linear maps be functional rather than merely relational.

### 2021-10-15 tex/esop22/bunchedrules.tex
[ writing ] Bunched Typing Rules section

@@
-Let $\Ann$ be a posemiring. Our framework represents well typed and
-$\Ann$-resourced terms \emph{intrinsically}. Intrinsic typing means
-that we represent well typed and $\Ann$-resourced terms (and
-\emph{only} those) as inhabitants of an inductive family
+We now let $\Ann$ be an arbitrary posemiring. Our framework represents
+well typed and $\Ann$-resourced terms \emph{intrinsically}. Intrinsic
+typing means that we represent well typed and $\Ann$-resourced terms
+(and \emph{only} those) as inhabitants of an inductive family
 $\grR\gamma \vdash A$ indexed by resource contexts $\grR$, type
@@
   \end{displaymath}
-  \caption{A fragment of \name{}}\label{fig:lr}
+  \caption{A prototypical posemiring-resourced system}\label{fig:lr}
 \end{figure}
@@
 \figref{fig:lr} presents a prototypical example of a system that our
-framework can represent.
+framework can represent, which is a subsystem of the \name{} system of
+\cite{WA20}. Each rule is given as a constructor: the premises are
+named $p$, $s$, $t$, etc., and the conclusion is a constructor applied
+to those variables. Variables are represented intrinsically as members
+of the type $\grR\gamma \sqni A$, which is a proof that the type $A$
+appears in the typing context, $i : \gamma \ni A$, together with a
+proof that $\grR \leq \langle i \rvert$.  Expanding the vector
+notation, the latter condition says that the selected variable $i$
+must have a usage annotation $\leq \gr1$ in $\grR$, while all other
+variables must have a usage annotation $\leq \gr0$. The
+\AgdaInductiveConstructor{var} rule imports variables into terms.
 
-As a prototypical example of the kind of system that our framework can
-handle, we
+The remaining rules are the introduction and elimination rules for
+three type constructors: \AgdaInductiveConstructor{⊸I} and
+\AgdaInductiveConstructor{⊸E} for function types $A \multimap B$ where
+the bound variable is annotated with $\gr 1$ for ``use once'';
+\AgdaInductiveConstructor{⊕I} and \AgdaInductiveConstructor{⊕E} for
+sum types $A \oplus B$; and \AgdaInductiveConstructor{!I} and
+\AgdaInductiveConstructor{!E} for a $\Ann$-annotated exponential
+modality $\oc_{\gr r} A$.
 
-To motivate the way our framework represents
-
-Need to introduce:
+There are two key observations to make about this system, which will
+guide the way we build our generic framework for $\Ann$-annotated
+substructural systems:
 \begin{enumerate}
-\item Intrinsically typed syntax
-\item Intrinsically typed semiring annotated syntax
-\item Context modification
+\item Every rule repeats the typing context $\gamma$ throughout its
+  premises and conclusion. The only time the typing context is
+  modified is to add additional variables in the rules that bind fresh
+  variables (\AgdaInductiveConstructor{⊸I},
+  \AgdaInductiveConstructor{⊕E}, \AgdaInductiveConstructor{!E}).
+\item Rules with multiple typing premises must describe how the
+  resources of the conclusion (always denoted $\grR$) are distributed
+  across the premises. In the \AgdaInductiveConstructor{⊸E} rule, the
+  resources are separated into two parts $\grP$ and $\grQ$ for the
+  premises. This is an example of a \emph{multiplicative} rule in the
+  terminology of Linear Logic \cite{girard87linear}. In the
+  \AgdaInductiveConstructor{⊕E} we see an example of an
+  \emph{additive} rule, where the resource context $\grQ$ is shared
+  between the premises $t$ and $u$\footnote{There is an unfortunate
+    clash of terminology here: multiplicative rules \emph{add} their
+    resources, while additive rules \emph{share} their
+    resources.}. The \AgdaInductiveConstructor{!I} rule uses scaling
+  by $\gr r$ of the resources of the premise.
 \end{enumerate}
 
-To build our universe of semiring annotated syntaxes with binding, we
-need to closely examine the structures underlying the kinds of typing
-rules present in these systems. We take a fragment of Wood and Atkey's
-\name{} calculus, enough to show the key details, presented in
-Figure \ref{fig:lr}.
+\begin{figure}
+  \begin{align*}
+    \dot1\,\grR &\coloneqq 1 \\
+    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
+    % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
+    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
+    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
+                       \times T\,\grP \times U\,\grQ \\
+    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
+                       \times T\,\grP \\
+    % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
+    %                    \to T\,\grQ \to U\,\grR
+  \end{align*}
+  \caption{The bunched connectives for combining premises.}
+  \label{fig:bunched}
+\end{figure}
 
-In common with other mechanisations of syntax using dependent types,
-we use a
+These observations indicate a way to regularise and streamline the
+presentation of this system. Instead of treating each premise and the
+conclusion as having potentially unrelated typing and resource
+conflicts, we make use of combinators for combining premises that will
+relate their resource and typing contexts to the conclusion by
+construction. To handle binders, which introduce variables, we make
+use of a combinator that adds a variable with a given $Ann$-annotation
+to an ambient context, without having to explicitly mention the parts
+of the context that have not changed. This technique is already
+present in some paper presentations of type systems, and is used as a
+simplifying technique by Allais et al.~\cite{AACMM20}. To manage how
+resources are distributed between premises, we make use the separating
+($*$) and sharing ($\dot{\times}$) conjunction connectives from
+Bunched Implications \cite{oHP99}, instantiated for typing rules. To
+handle the \AgdaInductiveConstructor{!I} rule, we will need a
+\emph{scaling} modality, $\gr r \cdot -$. The semantics of the bunched
+connectives we will use are given in \figref{fig:bunched}. An
+important point to note is that bunched combinators induce
+\emph{linear} combinations of substructures, in the sense of the
+linear algebra of posemirings described in the previous section.
 
-These rules are the constructors of a datatype
-$(\vdash) : \mathrm{Ctx} \to \mathrm{Ty} \to \mathrm{Set}$.
+\begin{figure}
+  \begin{displaymath}
+    \begin{prooftree}[comb]
+      \hypo{x :~\sqni A}
+      \infer1{\AgdaInductiveConstructor{var}~x :~\vdash A}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}[comb]
+      \hypo{t : \gr1A \vdash B}
+      \infer1{\AgdaInductiveConstructor{⊸I}~t~ \mathord: \mathord\vdash A \multimap B}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}[comb]
+      \hypo{(t :~\vdash A \multimap B)}
+      \hypo{\sep}
+      \hypo{(s :~\vdash A)}
+      \infer3{\AgdaInductiveConstructor{⊸E}~t~s :~\vdash B}
+    \end{prooftree}
+  \end{displaymath}
+  \begin{displaymath}
+    \begin{prooftree}[comb]
+      \hypo{(t :~\vdash A_i)}
+      \infer1{\AgdaInductiveConstructor{⊕I}_i~t : \vdash A_0 \oplus A_1}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}[comb]
+      \hypo{(s : \vdash A \oplus B)}
+      \hypo{\sep}
+      \hypo{((t : \gr1A \vdash C)}
+      \hypo{\dottimes}
+      \hypo{(u : \gr1B \vdash C))}
+      \infer5{\AgdaInductiveConstructor{⊕E}~s~t~u : \vdash C}
+    \end{prooftree}
+  \end{displaymath}
+  \begin{displaymath}
+    \begin{prooftree}[comb]
+      \hypo{t : \gr r \cdot \plr{\vdash A}}
+      \infer1{\AgdaInductiveConstructor{!I}~t : \vdash \oc_{\gr r} A}
+    \end{prooftree}
+    \qquad
+    \begin{prooftree}[comb]
+      \hypo{(s : \vdash \oc_{\gr r} A)}
+      \hypo{\sep}
+      \hypo{(t : \gr rA \vdash C)}
+      \infer3{\AgdaInductiveConstructor{!E}~s~t : \vdash C}
+    \end{prooftree}
+  \end{displaymath}
+  \caption{The prototypical system of \figref{fig:lr} restated in terms of bunched combinators.}
+  \label{fig:lr-comb}
+\end{figure}
 
-In keeping with our Agda formalisation, we have used an
-\emph{intrinsically typed} syntax. Each typing rule is presented as a
-constructor whose arguments are the premises above the line. Typed
-terms are represented as nothing more than their typing derivations,
-constructed from the constructors. To account for the semiring
-annotations, the rules are
+\figref{fig:lr-comb} shows our prototypical system restated with
+implicit contexts and the bunched combinators. The inductive family is
+now denoted $\vdash A$, only mentioning context extensions, as we do
+in the rules \AgdaInductiveConstructor{⊸I},
+\AgdaInductiveConstructor{⊕E} and \AgdaInductiveConstructor{!E}. Thus,
+in the \AgdaInductiveConstructor{var} rule, the context is completely
+supressed. The \AgdaInductiveConstructor{⊸I} just has to state that a
+new variable with resource annotation $\gr1$ is added to the
+context. The \AgdaInductiveConstructor{⊸E} rule uses the separating
+conjunction ($*$) to combine the premises, indicating that the
+resources required by the two premises are added together for the
+conclusion. The \AgdaInductiveConstructor{⊕E} rule demonstrates the
+sharing conjunction $\dot\times$: the scrutinee term $s$ and the
+clause terms $t, u$ are combined by separating conjunction, because
+their resources must be combined, but the clause terms are combined by
+the sharing conjunction, because they must use the same resources.
 
-The key features we wish to describe from this
-
-The $\AgdaInductiveConstructor{var}$ rule constructs variables,
-
-In systems with Linear types, a crucial distinction is made between
-multiplicative and additive typing rules that does not arise in
-non-substructural systems. The distinction is exemplified by the
-Intuitionistic Linear Logic introduction rules for \emph{tensor}
-products $A \otimes B$ and \emph{with} products $A \with B$:
-\begin{mathpar}
-  \inferrule* [right=$\otimes$-Intro]
-  {\Gamma \vdash s : A \\ \Delta \vdash t : B}
-  {\Gamma \vdash s \otimes t : A \otimes B}
-
-  \inferrule* [right=$\with$-Intro]
-  {\Gamma \vdash s : A \\ \Gamma \vdash t : B}
-  {\Gamma \vdash s \with t : A \with B}
-\end{mathpar}
-In the \TirName{$\otimes$-Intro} rule, the two premises have disjoint
-contexts $\Gamma$ and $\Delta$ that are combined in the conclusion,
-indicating that the terms $s$ and $t$ refer to separate
-resources. This form of rule is referred to as
-\emph{multiplicative}. In the \TirName{$\with$-Intro} rule, the two
-premises share the context $\Gamma$, indicating that the terms $s$ and
-$t$ may refer to the same resources.
-
-In systems that use semiring annotations to track resource usage, the
-distinction between multiplicative and additive still remains but now
-contexts in multiplicative rules are combined by adding their semiring
-annotations together.
+Bunched combinators, along with supression of unchanged typing
+contexts, leads to a more streamlined presentation of the system
+without the clutter of explicit resource context
+inequalities. However, the larger advantage for us is that systems are
+constructed using these combinators \emph{automatically} admit
+renaming, substitution, and other scope-, type-, and resource-safe
+traversals. If we were to allow arbitrary modification of the context
+in premises, this would not always be possible, since there would be
+no guarantee that a substitution (for instance) could be ``pushed'' up
+from a conclusion to the premises. As we will see in \secref{sec:env},
+our generic notion of environment (e.g., a simultaneous substitution)
+is essentially a linear transformation, and so automatically commutes
+with the linear combinations of premises induced by the bunched
+connectives. This is the key to our generic results for all of the
+systems describable in our framework.

### 2021-10-15 tex/esop22/quant-framework.tex
[ writing ] Bunched Typing Rules section

@@
 \newcommand{\figref}[1]{Figure \ref{#1}}
+\newcommand{\secref}[1]{Section \ref{#1}}
 
@@
 %
-\begin{enumerate}
-\item (Section 1)
-  \begin{enumerate}
-  \item   Motivate intrinsically typed syntax (in Agda)
-    \begin{enumerate}
-    \item Show the ``GADT'' for well scoped syntax
-    \item Show the well typed version
-    \end{enumerate}
-  \item Motivate adding posemiring annotations
-    \begin{enumerate}
-    \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
-    \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
-    \end{enumerate}
-  \end{enumerate}
-\item (Section 2) Background / notation fixing for:
-  \begin{enumerate}
-  \item Partially ordered semirings
-  \item Vectors
-  \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
-  \end{enumerate}
-\item (Section 3) Bunched Typing Rules
-  \begin{enumerate}
-  \item Introduces $\lambda\mathcal{R}$
-  \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
-  \end{enumerate}
-\item (Section 4) Universe of Generic Usage-annotated syntax
-  \begin{enumerate}
-  \item Premises
-  \item Rules
-  \item Systems
-  \item Example(s)
-  \item Intepretation as Terms
-  \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
-  \end{enumerate}
-\item (Section 5) Example Syntaxes
-  \begin{enumerate}
-  \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
-  \item Inductive types with a well resourced fold
-  \item Bernardy-Abel style
-  \item First-order (circuitry)
-  \item LNL
-  \item $\lambda\mu\tilde{\mu}$
-  \end{enumerate}
-\item (Section 6) Environments
-  \begin{enumerate}
-  \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
-  \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
-    \begin{enumerate}
-    \item Key idea / slogan : environment = change of basis
-    \end{enumerate}
-  \item Definition of environments
-  \item Examples
-  \item Properties, incl. preservation of linear operations
-  \end{enumerate}
-\item (Section 7) Semantics, or How to apply an Environment to a Term
-  \begin{enumerate}
-  \item Renameability
-  \item $\mathsf{semantics}$
-  \end{enumerate}
-\item (Section 8) Examples of Semantics
-  \begin{enumerate}
-  \item Renaming and Substitution (term-valued)
-  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
-  \item Usage Elaborator
-  \end{enumerate}
-\item (Section 9) Related Work
-\item (Section 10) Conclusion
-\end{enumerate}
+% \begin{enumerate}
+% \item (Section 1)
+%   \begin{enumerate}
+%   \item   Motivate intrinsically typed syntax (in Agda)
+%     \begin{enumerate}
+%     \item Show the ``GADT'' for well scoped syntax
+%     \item Show the well typed version
+%     \end{enumerate}
+%   \item Motivate adding posemiring annotations
+%     \begin{enumerate}
+%     \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
+%     \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
+%     \end{enumerate}
+%   \end{enumerate}
+% \item (Section 2) Background / notation fixing for:
+%   \begin{enumerate}
+%   \item Partially ordered semirings
+%   \item Vectors
+%   \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
+%   \end{enumerate}
+% \item (Section 3) Bunched Typing Rules
+%   \begin{enumerate}
+%   \item Introduces $\lambda\mathcal{R}$
+%   \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
+%   \end{enumerate}
+% \item (Section 4) Universe of Generic Usage-annotated syntax
+%   \begin{enumerate}
+%   \item Premises
+%   \item Rules
+%   \item Systems
+%   \item Example(s)
+%   \item Intepretation as Terms
+%   \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
+%   \end{enumerate}
+% \item (Section 5) Example Syntaxes
+%   \begin{enumerate}
+%   \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
+%   \item Inductive types with a well resourced fold
+%   \item Bernardy-Abel style
+%   \item First-order (circuitry)
+%   \item LNL
+%   \item $\lambda\mu\tilde{\mu}$
+%   \end{enumerate}
+% \item (Section 6) Environments
+%   \begin{enumerate}
+%   \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
+%   \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
+%     \begin{enumerate}
+%     \item Key idea / slogan : environment = change of basis
+%     \end{enumerate}
+%   \item Definition of environments
+%   \item Examples
+%   \item Properties, incl. preservation of linear operations
+%   \end{enumerate}
+% \item (Section 7) Semantics, or How to apply an Environment to a Term
+%   \begin{enumerate}
+%   \item Renameability
+%   \item $\mathsf{semantics}$
+%   \end{enumerate}
+% \item (Section 8) Examples of Semantics
+%   \begin{enumerate}
+%   \item Renaming and Substitution (term-valued)
+%   \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
+%   \item Usage Elaborator
+%   \end{enumerate}
+% \item (Section 9) Related Work
+% \item (Section 10) Conclusion
+% \end{enumerate}
 

### 2021-10-15 tex/esop22/algebra.tex
[ writing ] Bunched Typing Rules section
[ writing ] more redrafting near deadline

@@
 \begin{definition}
-  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that the
-  following laws hold (universally quantified in all the variables).
-  \begin{itemize}
-    \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
-    \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
-    \item $0 + x = x
-      \land x + 0 = x
-      \land (x + y) + z = x + (y + z)
-      \land x + y = y + x$
-    \item $1 * x = x
-      \land x * 1 = x
-      \land (x * y) * z = x * (y * z)$
-    \item $0 * x = 0
-      \land (x + y) * z = x * z + y * z$
-    \item $x * 0 = 0
-      \land x * (y + z) = x * y + x * z$
-  \end{itemize}
+  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that
+  $(\Ann, \leq)$ is a partially ordered set, $(\Ann, 0, +)$ is a commutative
+  monoid, $(\Ann, 1, *)$ is a monoid, $+$ and $*$ are monotonic, and $*$
+  distributes over $0$ and $+$ in the usual way on both sides.
+  %the
+  %following laws hold (universally quantified in all the variables).
+  %\begin{itemize}
+  %  \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
+  %  \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
+  %  \item $0 + x = x
+  %    \land x + 0 = x
+  %    \land (x + y) + z = x + (y + z)
+  %    \land x + y = y + x$
+  %  \item $1 * x = x
+  %    \land x * 1 = x
+  %    \land (x * y) * z = x * (y * z)$
+  %  \item $0 * x = 0
+  %    \land (x + y) * z = x * z + y * z$
+  %  \item $x * 0 = 0
+  %    \land x * (y + z) = x * y + x * z$
+  %\end{itemize}
 \end{definition}
+
+\begin{example}[Zero-one-many]\label{thm:linearity}
+  The poset $\{0 > \omega < 1\}$ forms a posemiring under normal numeric
+  addition (with $1 + 1 = 1 + \omega = \omega + \omega = \omega$) and
+  multiplication (with $\omega * \omega = \omega$).
+  This gives us a way to mark whether variables are unused ($0$), used linearly
+  ($1$), or used unrestrictedly ($\omega$) in the current (sub)term.
+  The ordering says that unrestricted-use variables can also remain unused or
+  be used linearly.
+\end{example}
+
+\begin{example}[Variance]\label{thm:variance}
+  The set
+  $\{{\sim\sim}, {\uparrow\uparrow}, {\downarrow\downarrow}, {\wn\wn}\}$,
+  with ${\sim\sim}$ at the bottom and ${\wn\wn}$ at the top of the order, forms
+  a posemiring with addition being \emph{meet}, $0$ being \emph{top}
+  (${\wn\wn}$), $1$ being ${\uparrow\uparrow}$, and multiplication being
+  commutative and determined by
+  ${\downarrow\downarrow} * {\downarrow\downarrow} = {\uparrow\uparrow}$ and
+  ${\sim\sim} * {\downarrow\downarrow} = {\sim\sim} * {\sim\sim} = {\sim\sim}$.
+  This gives us a way to track the variance with which variables are used, in
+  the aim of all terms being monotonic in their free variables.
+  ${\uparrow\uparrow}$ stands for covariance, ${\downarrow\downarrow}$ for
+  contravariance, ${\sim\sim}$ for invariance, and ${\wn\wn}$ for a variable
+  with no guarantees, in which we must be constant.
+\end{example}
 
@@
 
-\begin{figure}
-  \begin{align*}
-    \dot1\,\grR &\coloneqq 1 \\
-    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
-    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
-    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
-    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
-                       \times T\,\grP \times U\,\grQ \\
-    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
-                       \times T\,\grP \\
-    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
-                       \to T\,\grQ \to U\,\grR
-  \end{align*}
-  \caption{The bunched connectives}
-  \label{fig:bunched}
-\end{figure}
-
 %Operations like renaming and substitution are essentially translations from one

### 2021-10-15 tex/esop22/intro.tex
[ writing ] more redrafting near deadline

@@
-\begin{itemize}
-  \item Review standard linear algebra constructions in the posemiring setting.
-  \item Give a general definition of simple type systems with usage annotations
-    based on connectives from bunched logic and constructions from linear
-    algebra.
-  \item Give some novel examples of such systems.
-  \item Derive admissibility of renaming and substitution for all such systems.
-  \item Give a usage checker generic in syntax.
-\end{itemize}
+%\begin{itemize}
+%  \item Review standard linear algebra constructions in the posemiring setting.
+%  \item Give a general definition of simple type systems with usage annotations
+%    based on connectives from bunched logic and constructions from linear
+%    algebra.
+%  \item Give some novel examples of such systems.
+%  \item Derive admissibility of renaming and substitution for all such systems.
+%  \item Give a usage checker generic in syntax.
+%\end{itemize}
 
@@
 natural deduction).
+
+This paper proceeds as follows.
+In \cref{sec:algebra}, we review and fix conventions pertaining to partially
+ordered semirings and vectors over them.
+In \cref{sec:bunched-rules}, we introduce an informal meta-syntax allowing us
+to state substructural typing rules succinctly and without explicit reference
+to contexts.
+In \cref{sec:syntax}, we mechanise that meta-syntax, giving a type of
+\emph{descriptions} of type systems, and interpreting those descriptions as
+types of intrinsically typed terms.
+In \cref{sec:env}, we discuss usage-aware environments: a generalisation of
+the structures used in simultaneous renaming and substitution proofs.
+We use environments in \cref{sec:semantics} to state an alternative
+elimination principle for terms, and give examples of such eliminations in
+\cref{sec:examples}.
+The examples are syntax-generic renaming and substitution, a specific
+denotational semantics, and a syntax-generic usage elaborator.
+Finally, we conclude and discuss future work in \cref{sec:conc}

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] Bunched Typing Rules section
[ writing ] more redrafting near deadline

@@
 \subsection{A prototypical type system}
-
-\begin{figure}
-  \begin{displaymath}
-    \begin{prooftree}[comb]
-      \hypo{\sqni A}
-      \infer1[Var]{\vdash A}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb]
-      \hypo{\gr1A \vdash B}
-      \infer1[$\multimap$-I]{\vdash A \multimap B}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb]
-      \hypo{\vdash A \multimap B}
-      \hypo{\sep}
-      \hypo{\vdash A}
-      \infer3[$\multimap$-E]{\vdash B}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}[comb]
-      \hypo{\vdash A_i}
-      \infer1[$\oplus$-I$_i$]{\vdash A_0 \oplus A_1}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb]
-      \hypo{\vdash A \oplus B}
-      \hypo{\sep}
-      \hypo{(\gr1A \vdash C}
-      \hypo{\dottimes}
-      \hypo{\gr1B \vdash C)}
-      \infer5[$\oplus$-E]{\vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \begin{displaymath}
-    \begin{prooftree}[comb]
-      \hypo{\gr r \cdot \plr{\vdash A}}
-      \infer1[$\oc_{\gr r}$-I]{\vdash \oc_{\gr r} A}
-    \end{prooftree}
-    \qquad
-    \begin{prooftree}[comb]
-      \hypo{\vdash \oc_{\gr r} A}
-      \hypo{\sep}
-      \hypo{\gr rA \vdash C}
-      \infer3[$\oc_{\gr r}$-E]{\vdash C}
-    \end{prooftree}
-  \end{displaymath}
-  \caption{\cref{fig:lr} restated in terms of bunched combinators}
-  \label{fig:lr-comb}
-\end{figure}
 
@@
 The next thing we want to do is to build terms in the described type system.
+The following definitions are useful for talking about types indexed over
+contexts, judgement forms, and judgement forms admitting newly bound variables,
+respectively.
 
@@
 
-Here is an example term, using the \AgdaFunction{$\lambda$R} system.
-First, for ease of writing, we introduce pattern synonyms for each of the
-typing rules we use.
+Terms defined like this are still quite difficult to write, mainly because of
+frequently changing usage contexts and the need for proofs that they all match
+up.
+We will see how to automate these proofs in \cref{sec:usage-elaborator}.
 
-\ExecuteMetaData[\PaperExamplestex]{patterns}
+%Here is an example term, using the \AgdaFunction{$\lambda$R} system.
+%First, for ease of writing, we introduce pattern synonyms for each of the
+%typing rules we use.
 
-Our example term is a function that flips a tagged union wrapped in an
-arbitrarily annotated \emph{bang}.
-Much of the effort in writing such a term goes into writing the various
-relatedness proofs between usage contexts --- observing, for example, that two
-usage contexts sum together to make a third, or that a usage context used for
-a variable is a basis vector.
-We give a method of automating these proofs in \cref{sec:usage-elaborator}.
-\todo{To be clear, we don't actually write this.}
+%\ExecuteMetaData[\PaperExamplestex]{patterns}
 
-\ExecuteMetaData[\HeavyItex]{lR-term}
+%Our example term is a function that flips a tagged union wrapped in an
+%arbitrarily annotated \emph{bang}.
+%Much of the effort in writing such a term goes into writing the various
+%relatedness proofs between usage contexts --- observing, for example, that two
+%usage contexts sum together to make a third, or that a usage context used for
+%a variable is a basis vector.
+%We give a method of automating these proofs in \cref{sec:usage-elaborator}.
+%\todo{To be clear, we don't actually write this.}
+
+%\ExecuteMetaData[\HeavyItex]{lR-term}
 
 A layer of syntax supports the following functorial action.
-{\color{red}(Maybe this has to go later, because it talks about linear maps.)}
 

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] cut Section 4 into paragraphs

@@
-\subsection{A prototypical type system}
-
-\subsection{Descriptions}
-
 We capture the form of rules exemplified previously\todo{Previously?} via
@@
 annotations, absence and multiplicity of subterms with summed usage annotations,
-scaling of a subterm, and variable binding.
+scaling of a subterm, and variable binding.\todo{Switching to Agda}
+
+\paragraph{Premises, Rules, and Systems.}
 
@@
 \ExecuteMetaData[\Syntaxtex]{System}
+
+\paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}
 
@@
 \ExecuteMetaData[\PaperExamplestex]{lR}
+
+\paragraph{Terms of a \AgdaRecord{System}.}
 
@@
 
-A layer of syntax supports the following functorial action.
+% A layer of syntax supports the following functorial action.
 
-\ExecuteMetaData[\Maptex]{map-s-type}
+% \ExecuteMetaData[\Maptex]{map-s-type}
 

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] fix up some of the 2-column code
[ writing ] tweaks to Section 4

@@
-We capture the form of rules exemplified previously\todo{Previously?} via
-\emph{descriptions} of rules.
-The key to making these descriptions work is that they only allow syntactic
-forms that preserve environments.
-These forms are: absence and multiplicity of subterms with the same usage
-annotations, absence and multiplicity of subterms with summed usage annotations,
-scaling of a subterm, and variable binding.\todo{Switching to Agda}
+We take the insights of the previous section and use them to build a
+generic framework for posemiring-annotated substructural systems in
+Agda. We will first show \emph{descriptions} of systems, which are
+comprised of rules that have premises combined using the bunched
+combinators. We then show how to construct the Agda data type of
+intrinsically well scoped, typed, and resourced terms for any given
+system of our framework. We use the prototypical system from
+\figref{fig:comb-lr} as a running example. \secref{sec:other-syntaxes}
+presents further examples that our framework can express.
 
-\paragraph{Premises, Rules, and Systems.}
+We now start to use Agda notation for record and data type
+declarations, to emphasise that our framework has been implemented.
+
+\subsection{Descriptions of Systems}
+
+% We capture the form of rules exemplified previously\todo{Previously?} via
+% \emph{descriptions} of rules.
+% The key to making these descriptions work is that they only allow syntactic
+% forms that preserve environments.
+% These forms are: absence and multiplicity of subterms with the same usage
+% annotations, absence and multiplicity of subterms with summed usage annotations,
+% scaling of a subterm, and variable binding.\todo{Switching to Agda}
+
+\paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}
 
 A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
-Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and a type
-of conclusion.
+Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and
+a type of conclusion. We assume that there is a
+$\AgdaBound{Ty} : \AgdaPrimitiveType{Set}$ of types for the system in
+scope.
 
-A single premise is introduced by the
-\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$} constructor.
-This allows binding of some variables \AgdaBound{$\Delta$} (with specified
-types and usage annotations) and the specification of a conclusion type
-\AgdaBound{A}.
-The rest of the constructors are descriptions for the first-order bunched
-connectives, and are interpreted directly as such.
+The \AgdaDatatype{Premise} data type describes premises of rules,
+using the bunched combinators from the previous section. A single
+premise is introduced by the
+\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
+constructor.  This allows binding of additional variables
+\AgdaBound{$\Delta$} (with specified types and usage annotations) and
+the specification of a conclusion type \AgdaBound{A} for this premise.
+The remaining constructors are descriptions for the first-order
+bunched connectives, and will be interpreted directly as such, below.
 
 \ExecuteMetaData[\Syntaxtex]{Premises}
+
+A \AgdaRecord{Rule} is a pair of some \AgdaDatatype{Premises} and a
+conclusion. We use an infix arrow as a suggestive notation for rules.
+
 \ExecuteMetaData[\Syntaxtex]{Rule}
+
+Finally, a \AgdaRecord{System} consists of a set of rule labels (i.e.,
+constructor names), and for each label a decsription of the
+corresponding rule. We use $\rhd$ as infix notation for systems to
+associate the label set with the rules.
+
 \ExecuteMetaData[\Syntaxtex]{System}
@@
 
-As an example, we transcribe the system defined in \cref{fig:lr-comb} into a
-description.
-We first give the set of types in a standard way (together with base type
-\AgdaInductiveConstructor{$\iota$}).
+As an example, we transcribe the system defined in
+\figref{fig:lr-comb} into a description.  We first give the set of
+types as a data type (together with a base type
+\AgdaInductiveConstructor{$\iota$}). We assume that there is a
+posemiring \AgdaInductiveConstructor{Ann} in scope for the
+annotations.
 
@@
 %Then, we give the set of labels.
-There is one label for each instantiation of a logical rule, but the labels
-contain no further information about subterms or restrictions on the context.
+There is one label for each instantiation of a logical rule, but the
+labels contain no further information about subterms or restrictions
+on the context. This will be provided when we associate labels with
+\AgdaRecord{Rule}s in a \AgdaRecord{System}.
 
+\noindent
+\begin{minipage}[t]{0.5\textwidth}
+  \ExecuteMetaData[\PaperExamplestex]{Ty}
+  \ExecuteMetaData[\PaperExamplestex]{Side}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
   \ExecuteMetaData[\PaperExamplestex]{qlR}
+\end{minipage}
 
-Finally, we specify, for each label, the rest of the information about that
-rule {\color{red}(premises and conclusion)}.
-We do not have a variable rule, because the variable rule is implicit to any
-type system.
-{\color{red}(Looks like \cref{fig:lr-comb})}
+To build a system, we associate with each label a rule:
 
 \ExecuteMetaData[\PaperExamplestex]{lR}
+
+Compared to \figref{fig:lr-comb}, modulo the Agda notation, we can see
+that the fundamental structure has been preserved: the rules match
+one-to-one, and the bunched premises are the same. A major difference
+is that we do not include a counterpart to the
+\AgdaInductiveConstructor{var} rule in a
+\AgdaRecord{System}. Variables are common to all the systems
+representable in our framework.
 
@@
 
-To specify the meaning of descriptions, we assume some
-\ExecuteMetaData[\Interpretationtex]{X}, over which we form one layer of syntax.
-The first argument to \AgdaBound{X} is the new variables bound by this layer
-of syntax, as exemplified in the first clause of
-\AgdaFunction{$\llbracket$\_$\rrbracket$p}.
-The second argument is the context containing the variables being carried over
-from the previous layer.
-Notice that this is not, in general, the same as the context from the previous
-layer, because the usage annotations may have been changed by connectives like
-\AgdaInductiveConstructor{\_`$*$\_} and \AgdaInductiveConstructor{\_`$\cdot$\_}.
-The third argument is the type of subterm required.
+To specify the meaning of descriptions, we assume some \AgdaBound{X} : \AgdaFunction{ExtOpenFam},
+% \ExecuteMetaData[\Interpretationtex]{X},
+over which we form one layer of syntax, using the function
+\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
+\AgdaDatatype{Premises} defined below.  The first argument to
+\AgdaBound{X} is the new variables bound by this layer of syntax, as
+exemplified in the first clause of
+\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
+the context containing the variables being carried over from the
+previous layer.  Notice that this is not, in general, the same as the
+context from the previous layer, because the usage annotations may
+have been changed by connectives like
+\AgdaInductiveConstructor{\_`$*$\_} and
+\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
+type of subterm required.
 
 With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
-the rest are simply interpretations into bunched combinators.
+the rest are simply interpretations of premises into bunched combinators.
 
@@
 
-The interpretation of a \AgdaRecord{Rule} is to check that the rule targets
-the desired type and then interpret the rules premises \AgdaBound{ps}.
+The interpretation of a \AgdaRecord{Rule} checks that the rule targets
+the desired type and then interprets the rule's premises \AgdaBound{ps}.
 Notice that the interpretation of the premises is independent of the conclusion
@@
 \AgdaFunction{OpenFam} on an extended context.
-This is what \AgdaFunction{Scope} does --- take the new variables
+We defined \AgdaFunction{Scope} to do this: take the new variables
 \AgdaBound{$\Delta$}, concatenate them onto the existing context
@@
 
-{\color{red}(Forward ref: for now, we could have inlined \texttt{Scope}.)}
+%{\color{red}(Forward ref: for now, we could have inlined \texttt{Scope}.)}
 
-It is this method we use to deal with new variables in syntax.
-Terms resemble the free monad over a layer-of-syntax functor, though that
-picture is complicated by variable binding.
-A term is either a variable or a use of a logical rule together with terms
-for each of the required subterms.
+We use \AgdaFunction{Scope} to deal with new variables in syntax.
+Terms resemble the free monad over a layer-of-syntax functor, though
+that picture is complicated by variable binding.  A term is either a
+variable or a use of a logical rule together with terms for each of
+the required subterms. The \AgdaFunction{Size} argument is where we
+use sized types to convince Agda that this type is strictly positive.
 
@@
 
-\subsection{Other syntaxes and syntactic forms}
+\subsection{Other syntaxes and syntactic forms}\label{sec:other-syntaxes}
 
-\paragraph{$\mu\tilde\mu$}
+\paragraph{The system $\mu\tilde\mu$.}
 We can encode a usage-annotated version of System $L$/the
@@
 
+\noindent
 \begin{minipage}[t]{0.5\textwidth}
-  \centering
   \ExecuteMetaData[\MuMuTildetex]{Ty}
@@
 \begin{minipage}[t]{0.5\textwidth}
-  \centering
   \ExecuteMetaData[\MuMuTildetex]{Conc}

### 2021-10-15 tex/esop22/bunchedrules.tex
[ writing ] more redrafting near deadline
[ writing ] panic edits

@@
       \hypo{t : \grP\gamma \vdash A}
-      \infer2{\AgdaInductiveConstructor{!I}~p~t : \grR\gamma \vdash \oc_{\gr r} A}
+      \infer2{\AgdaInductiveConstructor{!I}~p~t : \grR\gamma \vdash \oc{\gr r} A}
     \end{prooftree}
@@
       \hypo{p : \grR \leq \grP + \grQ}
-      \hypo{s : \grP\gamma \vdash \oc_{\gr r} A}
+      \hypo{s : \grP\gamma \vdash \oc{\gr r} A}
       \infer[no rule]1{t : \grQ\gamma, \gr rA \vdash C}
@@
 
-\begin{figure}
-  \begin{align*}
-    \dot1\,\grR &\coloneqq 1 \\
-    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
-    % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
-    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
-    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
-                       \times T\,\grP \times U\,\grQ \\
-    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
-                       \times T\,\grP \\
-    % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
-    %                    \to T\,\grQ \to U\,\grR
-  \end{align*}
-  \caption{The bunched connectives for combining premises.}
-  \label{fig:bunched}
-\end{figure}
+% \begin{figure}
+%   \begin{align*}
+%     \dot1\,\grR &\coloneqq 1 \\
+%     (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
+%     % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
+%     I^*\,\grR &\coloneqq \grR \leq \gr0 \\
+%     (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
+%                        \times T\,\grP \times U\,\grQ \\
+%     (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
+%                        \times T\,\grP \\
+%     % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
+%     %                    \to T\,\grQ \to U\,\grR
+%   \end{align*}
+%   \caption{The bunched connectives for combining premises.}
+%   \label{fig:bunched}
+% \end{figure}
 
@@
 \emph{scaling} modality, $\gr r \cdot -$. The semantics of the bunched
-connectives we will use are given in \figref{fig:bunched}. An
-important point to note is that bunched combinators induce
+connectives we will use are:
+  \begin{align*}
+    \dot1\,\grR &\coloneqq 1 \\
+    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
+    % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
+    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
+    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
+                       \times T\,\grP \times U\,\grQ \\
+    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
+                       \times T\,\grP \\
+    % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
+    %                    \to T\,\grQ \to U\,\grR
+  \end{align*}
+An important point to note is that bunched combinators induce
 \emph{linear} combinations of substructures, in the sense of the

### 2021-10-15 tex/esop22/syntax.tex
[ writing ] panic edits

@@
 As an example, we transcribe the system defined in
-\figref{fig:lr-comb} into a description.  We first give the set of
-types as a data type (together with a base type
-\AgdaInductiveConstructor{$\iota$}). We assume that there is a
+\figref{fig:lr-comb} into a description.  We give the set of types of
+this system as a data type \AgdaDatatype{Ty} (together with a base
+type \AgdaInductiveConstructor{$\iota$}). We assume that there is a
 posemiring \AgdaInductiveConstructor{Ann} in scope for the
-annotations.
-
-\ExecuteMetaData[\PaperExamplestex]{Ty}
-
-%Then, we give the set of labels.
-There is one label for each instantiation of a logical rule, but the
-labels contain no further information about subterms or restrictions
-on the context. This will be provided when we associate labels with
-\AgdaRecord{Rule}s in a \AgdaRecord{System}.
+annotations.There is one label for each instantiation of a logical
+rule, but the labels contain no further information about subterms or
+restrictions on the context. This will be provided when we associate
+labels with \AgdaRecord{Rule}s in a \AgdaRecord{System}.
 

### 2021-10-26 tex/framework.tex
[ writing ] various thesis additions
[ dump ] framework work from the paper

@@
-\section{Building up}
+\chapter{A framework for usage-restricted calculi}
 
-We assume that a type system comprises a set of (unparametrised) rules, each
-of which has a conclusion and several premises containing subterms.
-The primary investigation of this work is into what form the premises can take
-while maintaining useful features of syntax.
-We shall start with simple forms, allowing just for multiple subterms, and
-then build on resource-sensitive bunches, variable binding, and modalities.
+\section{Syntax}
 
-\System{}
-\Rule{}
+\input{syntax}
 
-Given some way \AgdaFunction{⟦\AgdaUnderscore{}⟧p} of interpreting
-\AgdaDatatype{Premises} in a \AgdaRecord{Ctx}, we can interpret a
-\AgdaDatatype{Rule} against a conclusion and a context by checking that its
-stated conclusion matches and then interpreting its premises.
-Then, the entire \AgdaDatatype{System} can be interpreted by picking a rule
-label (including parameters) \AgdaBound{l} and interpreting the selected rule
-\AgdaBound{rs}\AgdaSpace{}\AgdaBound{l}.
+\section{Semantics}
 
-\semr{}
-\sems{}
-
-\subsection{The language of Cartesian products}
-
-\SimplePremises
-
-\section{Bunched functions}
-
-Suppose we have a left $R$-semimodule $M$.
-Then we can define the following connectives on indexed type families.
-
-\begin{align*}
-  \top~z &:= \top \\
-  (P \wedge Q)~z &:= P~z \wedge Q~z \\
-  (P \to Q)~z &:= P~z \to Q~z \\
-  I~z &:= z \subres 0 \\
-  (P * Q)~z &:= \exists x,y.~z \subres x+y \wedge P~x \wedge Q~y \\
-  (P \wand Q)~y &:= \forall x,z.~z \subres x+y \to P~x \to Q~z \\
-  (r \cdot P)~z &:= \exists x.~z \subres rx \wedge P~x
-\end{align*}
-
-\paragraph{Agda definitions}
-To follow.
-
-The first six of these connectives are somewhat standard, the first three
-corresponding to the usual sharing connectives and the second three
-corresponding to the usual separating connectives.
-The final connective, $r \cdot P$, is new.
-We choose $\exists$ rather than $\forall$ because scaling will usually occur in
-premises, i.e., to the left of arrows, and we want to avoid anything
-higher-order.
-
-\section{Generic syntax}
-
-Rules $R$, premises $P$ and $Q$.
-
-\begin{align*}
-  R &::= P/A \\
-  P,Q &::= \grR\Theta \vdash A \mid \top \mid P \wedge Q \mid I \mid P * Q
-        \mid \gr r \cdot P
-\end{align*}
-
-Example rules:
-
-\begin{itemize}
-  \item With-introduction: ${\vdash A} \wedge {\vdash B} / A \with B$
-  \item Annotated arrow introduction:
-    ${\gr rA \vdash B} / \gr rA \multimap B$
-  \item Annotated arrow elimination:
-    ${\vdash \gr rA \multimap B} * \gr r \cdot {\vdash A} / B$
-  \item Cases of annotated sum:
-    ${\vdash \gr rA \oplus \gr sB}
-    * ({\gr rA \vdash C} \wedge {\gr sB \vdash C}) / C$
-\end{itemize}
-
-\paragraph{Agda example}
-
-We start by defining our types \AgdaDatatype{Ty}, which in this case will be the
-conclusions of our judgements.
-\AgdaFunction{Ann} is the type of usage annotations, which for now is an
-arbitrary skew semiring.
-We also have a type of rule labels \AgdaDatatype{`Sys}, which specifies all of
-the parameters of a typing rule except for its conclusion and premises.
-For example, in the \AgdaInductiveConstructor{`case} label, we need to fill in
-the two summands (types \AgdaBound{A} and \AgdaBound{B}), as well as the return
-type \AgdaBound{C}.
-The actual form of the rule will be specified later.
-
-\exItypes{} \exIlabels{}
-
-We are then ready to build a system \AgdaFunction{Sys} with conclusions
-\AgdaDatatype{Ty}, usage annotations \AgdaFunction{Ann}, and labels
-\AgdaDatatype{`Sys}.
-The rule corresponding to each label is given in the following.
-
-The \AgdaInductiveConstructor{`lam} rule has one premise, which binds one
-variable with usage annotation \AgdaBound{r} and type \AgdaBound{A}.
-The \AgdaInductiveConstructor{`app} rule has two premises combined with
-separating conjunction.
-Furthermore, the second premise (corresponding to the argument of the function)
-is subject to scaling by \AgdaBound{r}.
-Intuitively, this description means that we must have enough of each variable to
-use some in building a function, and then use some more \AgdaBound{r} times to
-build enough arguments.
-
-\exIfunrules{}
-
-In the rules for sums, we use an alternative notation for rule descriptions,
-with \AgdaFunction{{---}{---}} being an infix version of
-\AgdaInductiveConstructor{rule}.
-The introduction rules \AgdaInductiveConstructor{`inl} and
-\AgdaInductiveConstructor{`inr} are straightforward --- each has one
-premise which binds no new variables.
-The \AgdaInductiveConstructor{`case} rule is more complicated.
-The premises are first split into the eliminand and the continuations, using the
-separating conjunction we saw for \AgdaInductiveConstructor{`app}.
-However, the continuations are connected via the sharing conjunction, reflecting
-the fact that in any given run of the program, only one of the branches will be
-taken, so usages from each individually should be added to usages of the
-eliminand.
-
-\exIsumrules{}
-
-\paragraph{Interpretation as syntax}
-We can say something like ``premise connectives are interpreted as the
-corresponding bunched connectives, where appropriate''.
-
-\begin{align*}
-  \sem{P/A} &:= \sem P \to {- \vdash A} \\
-  \sem{\grR\Theta \vdash A} &:= {-, \grR\Theta \vdash A} \\
-  \sem{\top} &:= \top \\
-  \sem{P \wedge Q} &:= \sem P \wedge \sem Q \\
-  \sem{I} &:= I \\
-  \sem{P * Q} &:= \sem P * \sem Q \\
-  &\ldots
-\end{align*}
-
-\paragraph{Agda version}
-The interpretation of a system is the selection of a rule, together with the
-interpretation of that rule.
-
-\sems{}
-
-The interpretation of a rule is a check that the rule targets the desired
-conclusion, together with the interpretation of the rules premises.
-
-\semr{}
-
-For the premises, connectives are interpreted in the obvious way.
-Premises can ask for subterms via the $\AgdaInductiveConstructor{\_`⊢\_}$
-constructor, which supplies the new variables (in particular, each variable's
-type and usage annotation) \AgdaBound{Γ} and the desired conclusion
-\AgdaBound{A} of the subterm.
-
-\semp{}
-
-\section{Values and computations}
-
-Following \cite{AACMM20}, we call any semantic realisation of a variable a
-\emph{value}, and any semantic realisation of a term a \emph{computation}.
-Scoped families of values are named $\mathcal V$, while scoped families of
-computations are named $\mathcal C$.
-\todo{This wording is difficult.}
-Values and computations are concepts with attitudes.
-
-\section{Environments}
-
-An environment is a linear map from variables in one context to values in
-another.
-
-\missingfigure{Agda definition of \AgdaRecord{\_{---}Env}}
-
-An environment in \AgdaBound{PΓ} for \AgdaBound{$\mathcal V$}-values in
-\AgdaBound{QΔ} is a linear map \AgdaField{M} such that $Q \le PM$ and for any
-$P'$ and $Q'$ such that $Q' \le P'M$, a function from variables in $P'Γ$ to
-\AgdaBound{$\mathcal V$}-values in $Q'Δ$.
-
-\section{Thinnings}
-
-\todo[inline]{I think thinnings should be called \emph{renamings}.}
-
-In a usage-annotated sequent calculus, their are 4 basic structural rules:
-exchange, weakening of $0$-annotated formulae, contraction of $+$-annotated
-formulae, and subusaging.
-Their typical forms are given below.
-
-\begin{mathpar}
-  \inferrule*[right=exch]
-  {\grP\Gamma, \gr sB, \gr rA, \grQ\Delta \vdash Z}
-  {\grP\Gamma, \gr rA, \gr sB, \grQ\Delta \vdash Z}
-  \and
-  \inferrule*[right=weak]
-  {\grP\Gamma, \grQ\Delta \vdash Z}
-  {\grP\Gamma, 0A, \grQ\Delta \vdash Z}
-  \and
-  \inferrule*[right=cont]
-  {\grP\Gamma, \gr rA, \gr sA, \grQ\Delta \vdash Z}
-  {\grP\Gamma, (\gr r + \gr s)A, \grQ\Delta \vdash Z}
-  \and
-  \inferrule*[right=subuse]
-  {\grP \le \grQ \\\\ \grQ\Gamma \vdash Z}
-  {\grP\Gamma \vdash Z}
-\end{mathpar}
-
-Generalising away from $\vdash$, it can be helpful to know when an arbitrary
-scoped family respects these structural rules.
-First, we note that the structural rules of sequent calculus correspond
-directly to \emph{renaming} in natural deduction.
-A renaming is an environment of variables, i.e., a \emph{linear} map from
-variables to variables.
-The linearity is what gives us the restrictions on annotations in the weakening
-and contraction rules.
-
-\missingfigure{Example renaming}
-
-%A thinning is an environment of syntactic variables.
-%In effect, thinning allows for permutation and slackening of existing
-%variables, and introduction of new discardable variables at any position.
-%These actions are in direct correspondence with the usual sequent calculus
-%structural rules of exchange, contraction, and weakening, as well as subusaging.
-
-Given this definition, $\square$ and \texttt{Thinnable} are as in AACMM.
-\texttt{Kripke} is modified to use separating implication.
-
-\section{Substitution}
-
-\todo[inline]{This seems like a good place to introduce substitution, straight
-after renaming, though we don't even have terms yet.}
-Whereas a renaming is a linear map from variables to variables, a substitution
-is a linear map from variables to terms.
-
-\section{A layer of syntax is functorial}
+\input{semantics}

### 2021-10-26 tex/semantics.tex
[ dump ] framework work from the paper

--- 2021-10-07 tex/esop22/semantics.tex
@@
 
-{\color{red}(Maybe this goes in the Environment section.)}
-We saw in \cref{sec:env} that our notion of \emph{environment} specialises to
-\emph{simultaneous renaming} by setting $\V$ to be $\sqni$.
+\todo{Maybe this goes in the Environment section.}
+We saw in \cref{sec:env} that our notion of environment specialises to
+simultaneous renaming by setting $\V$ to be $\sqni$.
+This is encoded below as
+\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
 
 \ExecuteMetaData[\Renamingtex]{Ren}
+
+In a setting where new variables can be bound in the middle of a derivation,
+it is important that the values we carry around while traversing a term can
+handle the existence of variables that appear but they do not use.
+We call any such notion of value \emph{renameable}.
+The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
+\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
+\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
+\AgdaFunction{$\Box^r$}-coalgebra.
+
 \ExecuteMetaData[\Renamingtex]{Boxr}
 \ExecuteMetaData[\Renamingtex]{Renameable}
+
+A renameable notion of value gives rise to a renameable notion of environment,
+essentially by renaming each contained value in an appropriate way.
 

### 2021-10-26 tex/syntax.tex
[ dump ] framework work from the paper

--- 2021-10-15 tex/esop22/syntax.tex
@@
 system of our framework. We use the prototypical system from
-\figref{fig:comb-lr} as a running example. \secref{sec:other-syntaxes}
+\cref{fig:lr-comb} as a running example. \cref{sec:other-syntaxes}
 presents further examples that our framework can express.
@@
 
-\paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}
+\paragraph{\cref{fig:lr-comb} as a \AgdaRecord{System}.}
 
 As an example, we transcribe the system defined in
-\figref{fig:lr-comb} into a description.  We give the set of types of
+\cref{fig:lr-comb} into a description.  We give the set of types of
 this system as a data type \AgdaDatatype{Ty} (together with a base
@@
 
-Compared to \figref{fig:lr-comb}, modulo the Agda notation, we can see
+Compared to \cref{fig:lr-comb}, modulo the Agda notation, we can see
 that the fundamental structure has been preserved: the rules match

### 2021-12-03 tex/examples.tex
[ fix ] some citations and missing files

--- 2021-10-14 tex/esop22/examples.tex
@@
 
-\todo{Introduction}
+To justify the name \emph{semantics}, we give an example traversal that is a
+denotational semantics in the usual sense.
+The semantics we take is a refinement of that of \citet{AbelBernardy2020},
+which gives a way to extract parametricity theorems from substructurally typed
+programs.
+Example theorems are that all linear terms act as permutations on some fixed
+set of resources, and all monotonically typed terms are really monotonic in the
+way the typing suggests they are.
+
 To abbreviate this section, we use a simplified syntax compared to \name{}.
@@
 As a running example, we take the usage annotations to be the 4-element
-variance posemiring.
-\todo{Define earlier}
+variance posemiring (\cref{thm:variance}).
 We establish the property that all terms are monotonic in their free variables.
@@
 depending on the annotation of each free variable.
-This provides an additional example to those of \citet{AbelBernardy2020}.
-\todo{Cite before here.}
+This provides an additional example to those of \citeauthor{AbelBernardy2020}.
 
 We will take semantics of this system into
-\emph{world-indexed relations}~\cite{AbelBernardy2020}.
+\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
 A world-indexed relation over a poset of worlds \AgdaBound{W} is a set over
@@
 
-  We observe that the set component of this term's semantics is just the
-  expected Agda function when the two free variables are given appropriate
-  interpretations.
+  After feeding in Agda's addition and negation functions as the
+  interpretations of the free variables (noting that they are both monotonic
+  in the required way), we get the following free theorem.
 
-  \ExecuteMetaData[\Monotonicitytex]{minus-set}
+  \ExecuteMetaData[\Monotonicitytex]{thm-type}
 
-  Furthermore, the relational component of the semantics yields the free
-  theorem that the Agda subtraction so defined is monotonic in the expected way.
-  This relies on library proofs that addition and negation are suitably
-  monotonic.
+  %We observe that the set component of this term's semantics is just the
+  %expected Agda function when the two free variables are given appropriate
+  %interpretations.
 
-  \ExecuteMetaData[\Monotonicitytex]{thm}
+  %\ExecuteMetaData[\Monotonicitytex]{minus-set}
+
+  %Furthermore, the relational component of the semantics yields the free
+  %theorem that the Agda subtraction so defined is monotonic in the expected way.
+  %This relies on library proofs that addition and negation are suitably
+  %monotonic.
+
+  %\ExecuteMetaData[\Monotonicitytex]{thm}
 \end{example}
@@
 of variable binding.
-For example, in the $\oc_{\gr r}$-E rule of \cref{fig:lr-comb}, the right
+For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-comb}, the right
 premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
@@
 \emph{renaming} in this case, because $\V$ is a family of variables).
+\todo{Previous examples are now deleted.}
 In the previous examples, we have used \AgdaFunction{elab-unique}, which further

### 2021-12-03 tex/ndsc.tex
[ fix ] some citations and missing files

@@
 Though originally presented as a sequent calculus, the underlying natural
-deduction calculus was later given by Herbelin~\cite[p.\ 12]{Herbelin-hab}, and
+deduction calculus was later given by Herbelin~\cite[p.\ 12]{herbelin-hab}, and
 I will follow the latter.

### 2021-12-03 tex/dill.tex
[ writing ] WIP about DILL

Dual intuitionistic linear logic is a syntax for intuitionistic linear logic
introduced by \citep{Barber1997}.
Its key feature is sequents with \emph{dual contexts} --- splitting assumptions
into \emph{linear} assumptions and \emph{intuitionistic} assumptions.
The idea of the intuitionistic assumptions is that they will behave like the
variables of simply typed $\lambda$-calculus, in contrast to the linear
assumptions, which behave as in the linear calculus we saw in \autoref{sec:ill}.
For example, an intuitionistic assumption of $A$ in an instance of the
$\otimes$-introduction rule is automatically copied to both premises.
In contrast, an assumption of $\oc A$ in the purely linear calculus must first
be contracted into two assumptions, with one going to each premise.

The high-level purpose of intuitionistic assumptions is to universally
characterise the connective $\oc$.
That is to say, the existence of intuitionistic assumptions, differentiated from
linear assumptions, gives rise to $\oc$ in the same way that the existence of
multiple (linear) assumptions gives rise to $\otimes$.
We informally express the latter as ``$\otimes$ is the internalisation of
context concatenation (comma)'', and we can similarly express the former as
``$\oc$ is the internalisation of intuitionistic assumptions''.

%\begin{align*}
%  \begin{proof}
%    \hypo{\Gamma_0 \vdash \oc A}
%    \hypo{\Gamma_1, \mathbf{int}~A \vdash C}
%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
%  \end{proof}
%  &
%  \begin{proof}
%    \hypo{\Gamma_0 \vdash A \otimes B}
%    \hypo{\Gamma_1, \mathbf{lin}~A, \mathbf{lin}~B \vdash C}
%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
%  \end{proof}
%\end{align*}

### 2021-12-03 tex/linear.tex
[ writing ] WIP about DILL

@@
 
-\section{Intuitionistic linear logic}
+\section{Intuitionistic linear logic}\label{sec:ill}
 \input{ill.tex}
+
 \section{DILL}
-DILL provides a syntax in which $\oc$ is characterised by a universal
-property.
-There is no need for any rule to depend on the entire context and rebind
-variables (as done by BBdPH), so is more ND-like.
-DILL makes $\oc$ behave more like a positive type former (see the deep
-identity function $\oc A \vdash \oc A$).
-DILL has been extended to dependent types: Vakar15.
-
-The idea of split contexts can also be applied to left-/right-rule systems.
-(Define this DILL variant.)
+\input{dill.tex}
 

### 2022-01-04 tex/dill.tex
[ new ] start writing about DILL

@@
 Dual intuitionistic linear logic is a syntax for intuitionistic linear logic
-introduced by \citep{Barber1997}.
+introduced by \citet{Barber1996}.
 Its key feature is sequents with \emph{dual contexts} --- splitting assumptions
@@
 
-%\begin{align*}
-%  \begin{proof}
-%    \hypo{\Gamma_0 \vdash \oc A}
-%    \hypo{\Gamma_1, \mathbf{int}~A \vdash C}
-%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
-%  \end{proof}
-%  &
-%  \begin{proof}
-%    \hypo{\Gamma_0 \vdash A \otimes B}
-%    \hypo{\Gamma_1, \mathbf{lin}~A, \mathbf{lin}~B \vdash C}
-%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
-%  \end{proof}
-%\end{align*}
+The rules witnessing the characterisations of $\otimes$ and $\oc$ are listed
+below.
+In particular, these are elimination rules, as found in the intro-elim system
+given by \citeauthor{Barber1996}.
+I take $\uplus$ to be a partial function, with $\Gamma_0 \uplus \Gamma_1$
+defined only when $\Gamma_0$ and $\Gamma_1$ have the same intuitionistic
+entries (these being the same intuitionistic entries as in
+$\Gamma_0 \uplus \Gamma_1$).
+Linear assumptions are accumulated by $\uplus$.
+The use of $\uplus$ in these two rules separates the eliminand from the
+eliminator/continuation, rather than having anything particular to do with
+the connectives $\otimes$ and $\oc$.
+
+\[
+  \begin{matrix}
+    \begin{prooftree}
+      \hypo{\Gamma_0 \vdash A \otimes B}
+      \hypo{\Gamma_1, \mathbf{lin}\,A, \mathbf{lin}\,B \vdash C}
+      \infer2[$\otimes$-E]{\Gamma_0 \uplus \Gamma_1 \vdash C}
+    \end{prooftree}
+    &&
+    \begin{prooftree}
+      \hypo{\Gamma_0 \vdash \oc A}
+      \hypo{\Gamma_1, \mathbf{int}\,A \vdash C}
+      \infer2[$\oc$-E]{\Gamma_0 \uplus \Gamma_1 \vdash C}
+    \end{prooftree}
+  \end{matrix}
+\]
+
+The introduction rule for $\oc$ follows the pattern of Promotion, allowing us
+to conclude $\oc A$ from $A$ only when that proof of $A$ does not use linear
+assumptions.
+Now that the distinction between linear and intuitionistic assumptions is part
+of the structure of the \emph{sequent}, we do not check the \emph{types} of any
+assumptions in this rule.
+I write $\mathbf{int}\,\Gamma$, where $\Gamma = A_1, \ldots, A_n$, for the
+context $\mathbf{int}\,A_1, \ldots, \mathbf{int}\,A_n$.
+
+\[
+  \begin{prooftree}
+    \hypo{\mathbf{int}\,\Gamma \vdash A}
+    \infer1[$\oc$-I]{\mathbf{int}\,\Gamma \vdash \oc A}
+  \end{prooftree}
+\]

### 2022-01-04 tex/esop22/algebra.tex
[ edit ] various changes after reviews

@@
 set, where both operations are monotonic.
+As in many similar formalisms, posemiring elements represent usage restrictions,
+with addition collecting restrictions from multiple uses, multiplication
+handling usage under a modality, and the order giving subsumption of
+restrictions, comparable to subtyping.
 
@@
   distributes over $0$ and $+$ in the usual way on both sides.
-  %the
-  %following laws hold (universally quantified in all the variables).
-  %\begin{itemize}
-  %  \item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
-  %  \item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
-  %  \item $0 + x = x
-  %    \land x + 0 = x
-  %    \land (x + y) + z = x + (y + z)
-  %    \land x + y = y + x$
-  %  \item $1 * x = x
-  %    \land x * 1 = x
-  %    \land (x * y) * z = x * (y * z)$
-  %  \item $0 * x = 0
-  %    \land (x + y) * z = x * z + y * z$
-  %  \item $x * 0 = 0
-  %    \land x * (y + z) = x * y + x * z$
-  %\end{itemize}
 \end{definition}
@@
 We can also form the standard basis vectors at any given dimension.
+From the order on $\Ann$, we get a pointwise order on vectors.
 
@@
   collection of these respects the posemiring structure on $r$.
-  Specifically:
+  Specifically, for all instantiations of the variables:
   \begin{itemize}
-    \item $r \leq r' \to u \leq u' \to r \cdot u \leq r' \cdot u'$
-    \item $r \cdot 0_M = 0_M
-      \land r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$
-    \item $0 \cdot u = 0_M
-      \land \plr{r + s} \cdot u = r \cdot u +_M s \cdot u$
-    \item $1 \cdot u = u
-      \land \plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$
+      \item If $r \leq r'$ and $u \leq u'$, then $r \cdot u \leq r' \cdot u'$.
+    \item $r \cdot 0_M = 0_M$ and
+      $r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$.
+    \item $0 \cdot u = 0_M$ and
+      $\plr{r + s} \cdot u = r \cdot u +_M s \cdot u$.
+    \item $1 \cdot u = u$ and
+      $\plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$.
   \end{itemize}
@@
   posemiring $\Ann$ is a relation $\sim$ on the underlying sets of $M$ and $N$
-  satisfying the following laws.
+  satisfying the following laws (with $\to$ standing for implication and
+  quantifiers binding most loosely).
   \begin{itemize}
-    \item $u \leq u' \to v' \leq v \to u \sim v \to u' \sim v'$
+    \item $\forall u,u',v,v'.~u \leq u' \to v' \leq v \to u \sim v \to u' \sim v'$
     \item $\forall v.~\plr{\exists u.~u \leq 0 \land u \sim v} \to v \leq 0$

### 2022-01-04 tex/esop22/bunchedrules.tex
[ edit ] various changes after reviews

@@
 We now let $\Ann$ be an arbitrary posemiring. Our framework represents
-well typed and $\Ann$-resourced terms \emph{intrinsically}. Intrinsic
-typing means that we represent well typed and $\Ann$-resourced terms
+well typed and $\Ann$-usaged terms \emph{intrinsically}. Intrinsic
+typing means that we represent well typed and $\Ann$-usaged terms
 (and \emph{only} those) as inhabitants of an inductive family
-$\grR\gamma \vdash A$ indexed by resource contexts $\grR$, type
-contexts $\gamma$, and types $A$. We represent typing contexts as
-trees with types at the leaves, and resource contexts as functions
+$\grR\gamma \vdash A$ indexed by usage context $\grR$, type
+context $\gamma$, and type $A$. We represent typing contexts as
+trees with types at the leaves, and usage contexts as functions
 that assign elements of $\Ann$ to members of the context. Using trees
@@
 of a context by multiple variables does not lead to complex counting
-argument to access the pre-existing variables.\todo{left and right
+arguments to access the pre-existing variables.\todo{left and right
   constructors for paths}
@@
   \end{displaymath}
-  \caption{A prototypical posemiring-resourced system}\label{fig:lr}
+  \caption{A prototypical posemiring-usaged system}\label{fig:lr}
 \end{figure}
@@
 presentation of this system. Instead of treating each premise and the
-conclusion as having potentially unrelated typing and resource
-conflicts, we make use of combinators for combining premises that will
-relate their resource and typing contexts to the conclusion by
-construction. To handle binders, which introduce variables, we make
+conclusion as having potentially unrelated typing and usage
+constraints, we make use of combinators for combining premises that will
+relate their usage and typing contexts to the conclusion by
+construction.
+Similar combinators have appeared for the same purpose in the work of
+\citet{RPKV20}, including the internal homs we introduce later.
+To handle binders, which introduce variables, we make
 use of a combinator that adds a variable with a given $Ann$-annotation
@@
 present in some paper presentations of type systems, and is used as a
-simplifying technique by Allais et al.~\cite{AACMM20}. To manage how
-resources are distributed between premises, we make use the separating
-($*$) and sharing ($\dot{\times}$) conjunction connectives from
-Bunched Implications \cite{oHP99}, instantiated for typing rules. To
-handle the \AgdaInductiveConstructor{!I} rule, we will need a
-\emph{scaling} modality, $\gr r \cdot -$. The semantics of the bunched
+simplifying technique by Allais et al.~\cite{AACMM20}.
+To manage how usage annotations are distributed between premises, we use the
+separating ($*$) and sharing ($\dot{\times}$) conjunction connectives from
+Bunched Implications~\cite{oHP99}.
+To handle the \AgdaInductiveConstructor{!I} rule, we will need a
+\emph{scaling} modality, $\gr r \cdot {-}$. The semantics of the bunched
 connectives we will use are:
@@
 supressed. The \AgdaInductiveConstructor{⊸I} just has to state that a
-new variable with resource annotation $\gr1$ is added to the
+new variable with usage annotation $\gr1$ is added to the
 context. The \AgdaInductiveConstructor{⊸E} rule uses the separating
@@
 contexts, leads to a more streamlined presentation of the system
-without the clutter of explicit resource context
+without the clutter of explicit usage context
 inequalities. However, the larger advantage for us is that systems are
 constructed using these combinators \emph{automatically} admit
-renaming, substitution, and other scope-, type-, and resource-safe
+renaming, substitution, and other scope-, type-, and usage-safe
 traversals. If we were to allow arbitrary modification of the context

### 2022-01-04 tex/esop22/examples.tex
[ edit ] various changes after reviews

@@
 
-\begin{example}
+\begin{example}\label{thm:minus}
   We can make a subtraction function from primitive addition and negation on
@@
 \emph{renaming} in this case, because $\V$ is a family of variables).
-In the previous examples, we have used \AgdaFunction{elab-unique}, which further
+We use \AgdaFunction{elab-unique}, which further
 checks statically that exactly one derivation is returned.
@@
 otherwise we have to supply the intended usage annotations.
+
+The example below is the term from \cref{thm:minus}.
+We have instantiated the usage elaborator so that \AgdaField{0\#$^{-1}$},
+\AgdaField{+$^{-1}$}, and \AgdaField{*$^{-1}$} are always singletons (i.e.,
+deterministic), and \AgdaField{1\#$^{-1}$} is a singleton whenever it can be
+(and empty for $\downarrow\downarrow$ and $\wn\wn$).
+The notation \AgdaFunction{\#}\AgdaSpace{}\AgdaBound{i} turns statically known
+natural number \AgdaBound{i} into the corresponding variable pointer, with
+\AgdaNumber{0} being the leftmost.
+In other words, we can write variables using de Bruijn levels.
+The definition \AgdaFunction{\#0} is a workaround for an Agda bug.
+
+\ExecuteMetaData[\Monotonicitytex]{minus}

### 2022-01-04 tex/esop22/semantics.tex
[ edit ] various changes after reviews

@@
 
+Just as $\dotto$ is the internal hom for $\dottimes$, we introduce $\wand$ as
+the internal hom for $\sep$, with \AgdaFunction{$\wand^c$} acting on open types
+analogously to \AgdaFunction{$\sep^c$}.
+It is given by $\plr{T \wand U}\,\grP \coloneqq
+\Pi\grQ,\grR.~\plr{\grR \leq \grP + \grQ} \to T\,\grQ \to U\,\grR$.
+
+For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
+ignored.
 \AgdaFunction{Wrap} is a device that turns any type family into an equivalent
@@
 \AgdaField{get} whose type is the type in the body of the $\lambda$-abstraction.
-For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
-ignored.
 If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
 \ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
-\ExecuteMetaData[\Snippetstex]{KripkeExpanded}.
+\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.
 That is to say, the Kripke function is expecting a value for each newly bound
@@
 The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
+
+At this point we introduce a minor generalisation to \AgdaFunction{OpenFam}
+and \AgdaFunction{ExtOpenFam}:
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.
+We obtain the definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam}
+by replacing the textual occurrence of \AgdaBound{Ty} by the parameter
+\AgdaBound{I}.
+This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
+type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
+We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
+to \AgdaDatatype{Ctx}.
 

### 2022-01-04 tex/esop22/syntax.tex
[ edit ] various changes after reviews

@@
 the rest are simply interpretations of premises into bunched combinators.
+The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
+they have been lifted from predicates on usage vectors to predicates on
+contexts, with the type component of the context shared throughout.
+Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
+already polymorphic (not relying on anything specific about usage vectors), so
+do not need a \AgdaFunction{$^c$} variant.
 
@@
 the required subterms. The \AgdaFunction{Size} argument is where we
-use sized types to convince Agda that this type is strictly positive.
+use sized types to record that subterms are smaller than the
+surrounding term.
 
 \ExecuteMetaData[\Termtex]{Term}
+
+This definition uses \AgdaFunction{$\dotto$}, which, analogously to
+\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
+space.
+We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
+case two (the context and the type).
+Informally,
+\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
+stands for
+\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
+\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
+\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
+\AgdaSpace{}%
+\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
+where \AgdaBound{T} is a type family with $n$ many indices.
 

### 2022-01-21 tex/esop22/syntax.tex
[ edit ] TODO list

@@
 Finally, a \AgdaRecord{System} consists of a set of rule labels (i.e.,
-constructor names), and for each label a decsription of the
+constructor names), and for each label a description of the
 corresponding rule. We use $\rhd$ as infix notation for systems to

### 2022-01-21 tex/esop22/env.tex
[ writing ] more redrafting near deadline
[ edit ] changes and edits to Section 5 (Environments)

@@
-This section is a development of the work of \citet{WA20}.
+We have now seen how to generically build data types of terms for a
+given \AgdaRecord{System}, giving us a setting in which to construct
+intrinsically well typed and well annotated terms. Before defining a
+generic assignment of semantics for terms, we need to develop the
+notion of \emph{environment}, a mapping of variables to their semantic
+interpretations that is preserved by structural decompositions and
+variable bindings.
 
-Let us write $\sdtstile{}\V$ as the infix version of $\V$.
-That is, $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$.
-In this section, we introduce \emph{$\V$-environments}, which will form a key
-part of the generic traversal of \cref{sec:semantics}.
-The idea of a $\V$-environment is to extend the notion of $\V$-values from
-types to contexts.
-Specifically, where the judgement $\Gamma \sdtstile{}\V A$ says that we
-have a $\V$-value of type $A$ in context $\Gamma$, the judgement
-$\Gamma \env\V \Delta$ says that we have a $\V$-value for each entry in
-$\Delta$ (at the specified multiplicity), all in context $\Gamma$.
-For our purposes, it is essential that the ``all'' in the previous sentence is
-multiplicative --- we want to split the usage annotations of $\Gamma$ up into
-parts such that each part supports one $\V$-value.
+Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
+use the notation $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$ for
+the type of inhabitants of $\V$ in the context $\Gamma$ at type
+$A$. In the non-substructural systems of Allais et al. \cite{AACMM21},
+a $\V$-environment $\Gamma \env\V \Delta$ is nothing more than a
+function $\forall A \to \Delta \sqni A \to \Gamma \sdtstile{}\V A$,
+mapping variables to $\V$-things. In our usage annotated setting
+though, we must correctly distribute resources tracked by the
+annotations; making sure that we have enough resources in $\Gamma$ to
+cover all the demands in $\Delta$. Following the work of \citet{WA20},
+this accounting is expressed via the presence of a linear
+transformation:
 
-We have seen two kinds of values already: $\sqni$-values are variables and
-$\vdash$-values are terms.
-The corresponding environments are also standard concepts: $\sqni$-environments
-are \emph{simultaneous renamings} and $\vdash$-environments are
-\emph{simultaneous substitutions}.
+% the
+% abstract notion of variable in different ways, several familiar
+% definitions can be expressed as environments: if the abstract
+% variables
 
-\subsection{Definition}
+% If we take the latter to be
 
-\begin{definition}\label{def:lr-env}
-  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and $\Delta$
-  (written $\grP\gamma$ and $\grQ\delta$, respectively, when convenient)
-  is a linear map $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
-  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$, $\grPprime$, and
-  $\grQprime$ such that $\grPprime \leq \grQprime\gr\Psi$, a function from
+% This section is a development of the work of \citet{WA20}.
+
+% Let us write $\sdtstile{}\V$ as the infix version of $\V$.
+% That is, $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$.
+% In this section, we introduce \emph{$\V$-environments}, which will form a key
+% part of the generic traversal of \cref{sec:semantics}.
+% The idea of a $\V$-environment is to extend the notion of $\V$-values from
+% types to contexts.
+% Specifically, where the judgement $\Gamma \sdtstile{}\V A$ says that we
+% have a $\V$-value of type $A$ in context $\Gamma$, the judgement
+% $\Gamma \env\V \Delta$ says that we have a $\V$-value for each entry in
+% $\Delta$ (at the specified multiplicity), all in context $\Gamma$.
+% For our purposes, it is essential that the ``all'' in the previous sentence is
+% multiplicative --- we want to split the usage annotations of $\Gamma$ up into
+% parts such that each part supports one $\V$-value.
+
+% We have seen two kinds of values already: $\sqni$-values are variables and
+% $\vdash$-values are terms.
+% The corresponding environments are also standard concepts: $\sqni$-environments
+% are \emph{simultaneous renamings} and $\vdash$-environments are
+% \emph{simultaneous substitutions}.
+
+% \subsection{Definition}
+
+\begin{definition}[Environment]\label{def:lr-env}
+  A \emph{$\V$-environment} between annotated contexts $\Gamma$ and
+  $\Delta$ (decomposed as $\grP\gamma$ and $\grQ\delta$, respectively,
+  when convenient) is a linear map
+  $\gr\Psi : \Ann^{\size\Delta} \to \Ann^{\size\Gamma}$ (written
+  postfix) such that $\grP \leq \grQ\gr\Psi$ and for each $A$,
+  $\grPprime$, and $\grQprime$ such that
+  $\grPprime \leq \grQprime\gr\Psi$, a function from
   $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.
+
+  \bob{Write out the Agda notation for environments}
 \end{definition}
+
+By instantiating $\V$, we obtain resource correct versions of familiar
+notions: letting $\V$ be $\sqni$ yields resource correct renamings;
+and letting $\V$ be $\vdash$ (i.e., terms) yields resource correct
+substitutions.
 
@@
   \]
-  Linear map $\gr\Psi$ is the identity map, clearly satisfying
+  We take $\gr\Psi$ to be the identity map, clearly satisfying
   \(
@@
 \begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an environment (substitution) of type
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and usual
+  addition and multiplication.\bob{and $\vdash$ is the terms from a suitable system}
+  There is a $\vdash$-environment (substitution)
   \[
@@
   \]
-  Linear map $\gr\Psi$ is given by the matrix
+  The linear map $\gr\Psi$ is given by the matrix
   \(
@@
 \]
-This notation assumes that the notion of value supports subusaging, which is
-always the case when we are using environments for traversals.
+This notation assumes that $\V$ supports subusaging, which is
+always the case when we are using environments for traversals.\bob{Forward ref to where this is explicitly assumed?}
 
@@
   addition and multiplication.
-  There is an environment (renaming) of type
+  There is a $\sqni$-environment (renaming)
   \[
@@
 
-\subsection{Properties}
+\paragraph{Relocation} An environment
+$\rho : \grP\gamma \env\V \grQ\delta$ does not determine $\grP$ and
+$\grQ$, we can replace them with any $\grPprime$ and $\grQprime$ that
+are related by the linear map $\rho.\gr\Psi$:
 
-When constructing an environment, we can do so by cases on the shape of the
-target context.
-We can create an environment into the empty context when all usage annotations
-on the source context are $\gr0$.
-We can create an environment into a concatenated context when we can additively
-split up the annotations of the source context and produce environments into
-both halves from the split sources.
-We can create an environment into a singleton context when there is a context
-$\gr r$ times smaller than the source context in which we can produce a value
-of the appropriate type.
+\begin{lemma}[\AgdaFunction{relocate}]\label{thm:env-resize}
+  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
+  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
+  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
+  with the same linear map and action on variables.
+\end{lemma}
+% \begin{proof}
+%   The only part of the definition of an environment dependent on $\grP$ or
+%   $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
+%   replace for $\grPprime$ and $\grQprime$.
+% \end{proof}
+Relocation will be used when pushing environments into subterms in
+\secref{XX}.
 
-\begin{lemma}\label{thm:env-rules}
-  \Cref{def:lr-env} is sound and complete for the following syntax.
+\paragraph{Inductive Construction}
+
+When constructing an environment, we can do so by cases on the shape
+of the target context by the following rules, which use the bunched
+connectives from \secref{sec:bunched-rules}:
 \begin{displaymath}
@@
 \end{displaymath}
-\end{lemma}
+Left to right, we can create an environment into the empty context
+when all usage annotations on the source context are $\gr0$; we can
+create an environment into a concatenated context when we can
+additively split up the annotations of the source context and produce
+environments into both halves from the split sources; and we can
+create an environment into a singleton context $\gr rA$ when there is
+a context $\gr r$ times smaller than the source context in which we
+can produce a $\V$-value of the appropriate type.
+
+% \begin{lemma}\label{thm:env-rules}
+%   \Cref{def:lr-env} is sound and complete for the following syntax.
+% \end{lemma}
 
 \begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an environment (substitution) of type
+  Assume $\Ann$ is the natural numbers with ordering given by $=$ and
+  the usual addition and multiplication, and $\vdash$ is the type of
+  terms for a \AgdaRecord{System} with function application.  There is
+  an environment (substitution)
   \[
@@
 
-We could, indeed, use this syntax to define what an environment is.
-However, we find them difficult to work with, as it is often easier to do
+We could have used these rules to inductively define what environments are.
+However, we found that this was difficult to work with. It is often easier to do
 linear algebraic proofs separately from the rest of an environment.
-For example, for identity and composition of environments, the original
-definition
+For example, for identity and composition of environments (below), \defref{def:lr-env}
 is easier to use because we can rely on the identity and composition of linear
@@
 These are not identity environments, so we would have to strengthen the
-induction hypothesis.
+induction hypothesis.\bob{FIXME: make this paragraph more neatly flow into the next ones}
 
-\begin{lemma}[\AgdaFunction{relocate}]\label{thm:env-resize}
-  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
-  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
-  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
-  with the same linear map and action on variables.
-\end{lemma}
-\begin{proof}
-  The only part of the definition of an environment dependent on $\grP$ or
-  $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
-  replace for $\grPprime$ and $\grQprime$.
-\end{proof}
+% One of the primary test cases for environments is simultaneous substitution,
+% which will look like the following rule.
+% The admissibility of substitution will be by induction on the derivation of
+% $\Delta \vdash A$, so we will need to be able to adapt any environment we are
+% given to work with any possible context of new premises.
+% In the simply typed case, the only change to the context we encountered was the
+% binding of new variables.
+% Now, with usage annotations, we furthermore have linear decompositions of the
+% context, necessitating changes to the environment whenever usage annotations
+% change.
+% We deal first with linear decompositions.
 
-One of the primary test cases for environments is simultaneous substitution,
-which will look like the following rule.
-The admissibility of substitution will be by induction on the derivation of
-$\Delta \vdash A$, so we will need to be able to adapt any environment we are
-given to work with any possible context of new premises.
-In the simply typed case, the only change to the context we encountered was the
-binding of new variables.
-Now, with usage annotations, we furthermore have linear decompositions of the
-context, necessitating changes to the environment whenever usage annotations
-change.
-We deal first with linear decompositions.
+% \begin{displaymath}
+%   \begin{prooftree}
+%     \hypo{\Gamma \env{\vdash} \Delta}
+%     \hypo{\Delta \vdash A}
+%     \infer2[sub]{\Gamma \vdash A}
+%   \end{prooftree}
+% \end{displaymath}
 
-\begin{displaymath}
-  \begin{prooftree}
-    \hypo{\Gamma \env{\vdash} \Delta}
-    \hypo{\Delta \vdash A}
-    \infer2[sub]{\Gamma \vdash A}
-  \end{prooftree}
-\end{displaymath}
+% There are three kinds of linear decompositions we have to deal with: zero,
+% addition, and scaling; corresponding to bunched connectives $I^*$, $\sep$, and
+% $\gr r \cdot {}$, respectively.
+% In each case, we have a simple preservation lemma, transforming an environment
+% of type $\Gamma \env\V \Delta$ and a decomposition of $\Delta$ into a
+% decomposition of $\Gamma$ and environments for all of the decomposed fragments
+% of $\Gamma$ and $\Delta$.
 
-There are three kinds of linear decompositions we have to deal with: zero,
-addition, and scaling; corresponding to bunched connectives $I^*$, $\sep$, and
-$\gr r \cdot {}$, respectively.
-In each case, we have a simple preservation lemma, transforming an environment
-of type $\Gamma \env\V \Delta$ and a decomposition of $\Delta$ into a
-decomposition of $\Gamma$ and environments for all of the decomposed fragments
-of $\Gamma$ and $\Delta$.
+% \begin{lemma}[environments preserve zero]\label{thm:lr-env-zero}
+%   Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+%   $\grQ \leq \gr 0$, we also have that $\grP \leq \gr 0$.
+% \end{lemma}
 
-\begin{lemma}[environments preserve zero]\label{thm:lr-env-zero}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \gr 0$, we also have that $\grP \leq \gr 0$.
-\end{lemma}
+% \begin{lemma}[environments preserve addition]\label{thm:lr-env-add}
+%   Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+%   $\grQ \leq \grQl + \grQr$ for some $\grQl$ and $\grQr$, we also have $\grPl$
+%   and $\grPr$ such that $\grP \leq \grPl + \grPr$ and there are environments
+%   of types $\grPl\gamma \env\V \grQl\delta$ and
+%   $\grPr\gamma \env\V \grQr\delta$.
+% \end{lemma}
 
-\begin{lemma}[environments preserve addition]\label{thm:lr-env-add}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \grQl + \grQr$ for some $\grQl$ and $\grQr$, we also have $\grPl$
-  and $\grPr$ such that $\grP \leq \grPl + \grPr$ and there are environments
-  of types $\grPl\gamma \env\V \grQl\delta$ and
-  $\grPr\gamma \env\V \grQr\delta$.
-\end{lemma}
+% \begin{lemma}[environments preserve scaling]\label{thm:lr-env-scale}
+%   Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+%   $\grQ \leq \gr r\grQprime$ for some $\grQprime$, we also have a $\grPprime$
+%   such that $\grP \leq \gr r\grPprime$ and there is an environment of type
+%   $\grPprime\gamma \env\V \grQprime\delta$.
+% \end{lemma}
 
-\begin{lemma}[environments preserve scaling]\label{thm:lr-env-scale}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
-  $\grQ \leq \gr r\grQprime$ for some $\grQprime$, we also have a $\grPprime$
-  such that $\grP \leq \gr r\grPprime$ and there is an environment of type
-  $\grPprime\gamma \env\V \grQprime\delta$.
-\end{lemma}
+% Finally, we also take the opportunity to give the extend lemma, allowing
+% environments to incorporate newly bound variables.
+% \todo{Motivate}
+% In the intuitionistic case, the extend lemma had two requirements on $\V$: $\V$
+% admits weakening and we can map variables into $\V$-values.
+% With usage annotations, the former is unreasonable, but it turns out that we
+% only need weakening by variables whose usage annotation is less than or equal
+% to $\gr0$.
+% The latter stays as-is, with the note that ``variable'' now means a
+% usage-checked variable.
 
-Finally, we also take the opportunity to give the extend lemma, allowing
-environments to incorporate newly bound variables.
-\todo{Motivate}
-In the intuitionistic case, the extend lemma had two requirements on $\V$: $\V$
-admits weakening and we can map variables into $\V$-values.
-With usage annotations, the former is unreasonable, but it turns out that we
-only need weakening by variables whose usage annotation is less than or equal
-to $\gr0$.
-The latter stays as-is, with the note that ``variable'' now means a
-usage-checked variable.
+% \begin{lemma}[extend]\label{thm:lr-bind}
+%   Given functions
+%   ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
+%   \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
+%   $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
+%   type $\Gamma \env\V \Delta$ into an environment of type
+%   $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
+% \end{lemma}
 
-\begin{lemma}[extend]\label{thm:lr-bind}
-  Given functions
-  ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
-  \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
-  $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
-  type $\Gamma \env\V \Delta$ into an environment of type
-  $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
-\end{lemma}
-
-\subsection{Identity and composition}
-
-The requirements for identity and composition of environments look a bit like
-the unit and lift of a Kleisli triple.
+\paragraph{Identity and Composition} $\V$-environments map variables
+to $\V$-things, so they do not {\it a priori} support identities and
+composition. However, by assuming extra structure we can obtain
+identities and composition for environments:\bob{Need to define the
+  $\forallb{\dotto}$ notation and mention when these are real id/comp}
 
@@
   Given a function
-  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
-  \forall \grPprime, \grQprime.~\grPprime \leq \grQprime\plr{\rho.\gr\Psi} \to
-  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
-  compose environments of types $\Gamma \env\U \Delta$ and
-  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
+  \begin{displaymath}
+    \mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to \forallb{\V\,\grQ\delta \dotto \W\,\grP\gamma}
+  \end{displaymath}
+  then we can compose environments $\rho : \Gamma \env\U \Delta$ and
+  $\rho' : \Delta \env\V \Theta$ into an environment
+  $\rho; \rho' : \Gamma \env\W \Theta$.
 \end{lemma}
@@
 \end{corollary}
-\begin{proof}
-  By \cref{thm:env-comp}, with $\mathrm{lift}$ being given exactly by the
-  action of $\rho$ on variables.
-\end{proof}
+% \begin{proof}
+%   By \cref{thm:env-comp}, with $\mathrm{lift}$ being given exactly by the
+%   action of $\rho$ on variables.
+% \end{proof}
 
-\todo{What do these choices mean?}
-\begin{example}
-  We can derive the following instances of environment composition.
-  \begin{itemize}
-    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
-      syntactic traversal.
-      For example, if $\U = {\sqni}$, we need the action of renaming on terms
-      to show that a renaming followed by a substitution composes to a
-      substitution.
-      If $\U = {\vdash}$, then the action of substitution on terms gives us that
-      substitutions compose.
-    \item More generally, if $\V = {\vdash}$ and we have a semantics from
-      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
-      of terms.
-  \end{itemize}
-\end{example}
+\bob{Forward reference to where we use identity and composition}
+
+% \begin{example}
+%   We can derive the following instances of environment composition.
+%   \begin{itemize}
+%     \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
+%       syntactic traversal.
+%       For example, if $\U = {\sqni}$, we need the action of renaming on terms
+%       to show that a renaming followed by a substitution composes to a
+%       substitution.
+%       If $\U = {\vdash}$, then the action of substitution on terms gives us that
+%       substitutions compose.
+%     \item More generally, if $\V = {\vdash}$ and we have a semantics from
+%       $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
+%       of terms.
+%       This shows that a substitution and a semantics can turn a
+%       $\U$-environment into a $\W$-environment.
+%   \end{itemize}
+% \end{example}

### 2022-01-21 tex/esop22/examples.tex
[ minor ] changes made during meeting
[ rework ] usage elab above den sem, and 01ω

@@
+TODO: introductory paragraph
+
 \subsection{Renaming and substitution}\label{sec:kits}
@@
 \ExecuteMetaData[\Syntactictex]{Sub-Kit}
+
+\subsection{A usage elaborator}\label{sec:usage-elaborator}
+
+Using the constructs we have seen so far, producing example terms soon becomes
+extremely tedious.
+We can achieve some abbreviation by using pattern synonyms to wrap around
+\AgdaInductiveConstructor{`con} expressions, but we still have to
+produce essentially bespoke proofs whenever we use a usage-sensitive part of the
+syntax.
+The size of each of these proofs is roughly proportional to the number of free
+variables, so the amount of proof we have to write grows roughly quadratically
+with the size of terms.
+An additional factor, which we can't see on paper, is that type checking time
+for these proofs soon becomes prohibitive to interactive development.
+
+Our aim in this subsection is to automate usage constraint proofs, making terms
+both easier to write and more performant to check.
+We invoke the automation by writing terms in a syntax where usage constraints
+have been trivialised, and then use a semantic traversal over the simplified
+syntax to try to produce a fully elaborated term in the original syntax.
+We write the automation in a way that is generic in the syntax description, thus
+avoiding repetition and facilitating the prototyping of new type systems.
+
+The type of syntax descriptions depends on the type of usage annotations because
+of variable binding.
+For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-comb}, the right
+premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
+from the ambient posemiring.
+The scaling combinator also makes direct reference to the posemiring.
+To produce a simplified syntax description, where usage constraints are
+trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
+posemiring.
+In contrast to syntax descriptions, even though types can contain usage
+annotations, the type of types does not depend on the type of usage annotations.
+This means that, in our simplified syntax, terms have types from the original
+system even though variables have trivial usage annotations.
+We define the $\mathbf0$ posemiring as follows, being careful to use the
+0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
+solved by Agda's $\eta$-laws.
+Indeed, in this very definition, all of the semiring operations and laws are
+canonically inferred.
+
+\ExecuteMetaData[\UsageChecktex]{0-poSemiring}
+
+The elaboration process is monadic.
+In particular, we use the \AgdaDatatype{List}/non-determinism monad to give
+\emph{all} of the possible annotation choices on the free variables of a term.
+We believe the commitment to multiple solutions is inherent when the syntax
+contains \AgdaInductiveConstructor{`$\dot1$}.
+For example, in the intermediate stages of elaborating
+$\plr{\vdash \lambda x.~\plr{*,*}} : A \multimap \top \otimes \top$ with a
+usage counting posemiring (assuming reasonable rules for $\top$ and $\otimes$),
+it is unclear whether to use the variable $x$ in the left $*$ or the right $*$.
+This uncertainty should be reflected in the final result.
+
+The non-deterministic choices we make during elaboration are enumerated by
+the fields of \AgdaRecord{NonDetInverses}.
+These choices are driven by the typing rules and a candidate usage vector for
+the conclusion.
+For example, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is needed when we
+encounter a \AgdaInductiveConstructor{`$\sep$} in the syntax and the candidate
+usage annotation we are considering is \AgdaBound{r}.
+Then, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is a list of pairs of
+annotations \AgdaBound{p} and \AgdaBound{q} that \AgdaBound{r} can split into,
+together with a proof of the splitting.
+For \AgdaField{0\#$^{-1}$} and \AgdaField{1\#$^{-1}$}, inverses to constants,
+we are given the candidate \AgdaBound{r} and typically return an empty list if
+the constraint cannot be satisfied, or a singleton list containing a proof.
+\AgdaField{*$^{-1}$} is used when we encounter scaling, in which case we know
+both the scaling factor \AgdaBound{r} (from the syntax description) and the
+candidate \AgdaBound{q}.
+These inverse operations combine monadically (in fact, applicatively) to give
+inverses to the vector operations of zero, addition, scaling, and basis.
+
+\ExecuteMetaData[\UsageChecktex]{NonDetInverses}
+
+We choose the \AgdaBound{$\V$} of our semantics to be (unannotated) variables.
+For the \AgdaBound{$\C$}, we consider \emph{functions} from candidate usage
+vectors \AgdaBound{R} to the list of elaborated derivations with usage
+annotations given by \AgdaBound{R}.
+The protocol this encodes is that the user will provide an unannotated term
+together with a candidate usage context \AgdaBound{R}, and usage elaboration
+returns a list of possible ways the term could be annotated such that the
+conclusion has usage context \AgdaBound{R}.
+The module name \AgdaModule{U} refers to the fact that we are taking the
+ambient posemiring to be $\mathbf0$ in \AgdaFunction{OpenFam}.
+The effect on \AgdaFunction{OpenFam} is that the usage annotations of any
+contexts we consider are uninformative (hence the \AgdaSymbol{\_} on the left).
+
+\ExecuteMetaData[\UsageChecktex]{C}
+
+To traverse the unannotated terms, we produce a \AgdaRecord{Semantics} over the
+unannotated system \AgdaFunction{uSystem}\AgdaSpace{}\AgdaBound{sys}.
+To write it, we make use of idiom brackets
+\AgdaSymbol{(|}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaSymbol{|)},
+which have the effect of replacing top-level spines of applications by
+(\AgdaDatatype{List}-)applicative applications.
+Field by field, we already know that variables are renameable.
+To interpret a variable, we consider all the possible proofs that such a
+variable could be well annotated, and package them up as a variable term via
+the applicative machinery.
+Finally, for compound terms, we first reify the unannotated subterms, and then
+combine the subterms via a \AgdaFunction{lemma}.
+
+\ExecuteMetaData[\UsageChecktex]{elab-sem}
+
+The \AgdaFunction{lemma} essentially goes through the shape of the premises,
+combining the collections of subterms in the natural way.
+For example, at each
+\AgdaInductiveConstructor{\AgdaUnderscore{}$\dottimes$\AgdaUnderscore{}},
+we take the Cartesian product of the possibilities of each half, and at each
+\AgdaInductiveConstructor{\AgdaUnderscore{}$\sep$\AgdaUnderscore{}},
+we non-deterministically split the usage annotations coming in, and then take
+the Cartesian product.
+When it comes to newly bound variables, the \emph{syntax description} tells us
+their annotations, so there is no further non-determinism introduced here.
+
+\ExecuteMetaData[\UsageChecktex]{lemma-type}
+
+To actually use \AgdaFunction{elab-sem} on terms, we take the associated
+\AgdaFunction{semantics} and pass it the identity environment (an identity
+\emph{renaming} in this case, because $\V$ is a family of variables).
+We use \AgdaFunction{elab-unique}, which further
+checks statically that exactly one derivation is returned.
+The candidate usage vector \AgdaBound{R} will be \AgdaFunction{[]} for closed
+terms, and otherwise we have to supply the intended usage annotations.
+
+The example below is of a term needed to show that, in the
+$\{\gr0, \gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
+$\oc\gr\omega$ forms a comonad.
+We have instantiated the usage elaborator so that:
+\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and empty on
+$\gr1$;
+\AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and $\gr\omega$, and empty on
+$\gr0$;
+\AgdaField{+$^{-1}$} gives $\gr0 \mapsto [(\gr0,\gr0)]$,
+$\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
+$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and
+\AgdaField{*$^{-1}$} gives $(\gr\omega, \gr0) \mapsto [\gr0]$,
+$(\gr\omega, \gr1) \mapsto []$, and
+$(\gr\omega, \gr\omega) \mapsto [\gr\omega]$
+(omitting $(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity).
+Note that we do not consider splitting $\gr\omega$ up as, say,
+$\gr1 + \gr\omega$, because this splitting would introduce more
+non-determinism but not allow any more terms to be typed.
+As such, the only non-determinism comes when we have variables annotated
+$\gr1$ and need to do an additive split, like when we apply the
+\AgdaInductiveConstructor{!E} rule below.
+At this point, the variable can become either $\gr0$-annotated in the left
+subterm and $\gr1$-annotated on the right, or vice-versa.
+We will find that, because the left subterm wants to use that variable, the
+former choice will be rejected.
+The function \AgdaFunction{var\#} is a convenience for converting statically
+known natural numbers, representing de Bruijn \emph{levels}, into variable
+terms.
+
+\ExecuteMetaData[\PaperExamplestex]{cojoin}
 
@@
   second argument.
-  We give the definition in pseudocode, as we have not yet seen how to
-  conveniently write terms (\cref{sec:usage-elaborator}).
+  We give the definition in pseudocode, though it is also amenable to the
+  usage elaborator of \cref{sec:usage-elaborator}, suitably instantiated.
 
@@
 \end{example}
-
-\subsection{A usage elaborator}\label{sec:usage-elaborator}
-
-Using the constructs we have seen so far, producing example terms soon becomes
-extremely tedious.
-We achieved some abbreviation by using pattern synonyms, but we still have to
-produce essentially bespoke proofs whenever we use a usage-sensitive part of the
-syntax.
-The size of each of these proofs is roughly proportional to the number of free
-variables, so the amount of proof we have to write grows roughly quadratically
-with the size of terms.
-An additional factor, which we can't see on paper, is that type checking time
-for these proofs soon becomes prohibitive to interactive development.
-
-Our aim in this subsection is to automate usage constraint proofs, making terms
-both easier to write and more performant to check.
-We invoke the automation by writing terms in a syntax where usage constraints
-have been trivialised, and then use a semantic traversal over the simplified
-syntax to try to produce a fully elaborated term in the original syntax.
-We write the automation in a way that is generic in the syntax description, thus
-avoiding repetition and facilitating the prototyping of new type systems.
-
-The type of syntax descriptions depends on the type of usage annotations because
-of variable binding.
-For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-comb}, the right
-premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
-from the ambient posemiring.
-The scaling combinator also makes direct reference to the posemiring.
-To produce a simplified syntax description, where usage constraints are
-trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
-posemiring.
-In contrast to syntax descriptions, even though types can contain usage
-annotations, the type of types does not depend on the type of usage annotations.
-This means that, in our simplified syntax, terms have types from the original
-system even though variables have trivial usage annotations.
-We define the $\mathbf0$ posemiring as follows, being careful to use the
-0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
-solved by Agda's $\eta$-laws.
-Indeed, in this very definition, all of the semiring operations and laws are
-canonically inferred.
-
-\ExecuteMetaData[\UsageChecktex]{0-poSemiring}
-
-The elaboration process is monadic.
-In particular, we use the \AgdaDatatype{List}/non-determinism monad to give
-\emph{all} of the possible annotation choices on the free variables of a term.
-We believe the commitment to multiple solutions is inherent when the syntax
-contains \AgdaInductiveConstructor{`$\dot1$}.
-For example, in the intermediate stages of elaborating
-$\plr{\vdash \lambda x.~\plr{*,*}} : A \multimap \top \otimes \top$ with a
-usage counting posemiring (assuming reasonable rules for $\top$ and $\otimes$),
-it is unclear whether to use the variable $x$ in the left $*$ or the right $*$.
-This uncertainty should be reflected in the final result.
-
-The non-deterministic choices we make during elaboration are enumerated by
-the fields of \AgdaRecord{NonDetInverses}.
-These choices are driven by the typing rules and a candidate usage vector for
-the conclusion.
-For example, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is needed when we
-encounter a \AgdaInductiveConstructor{`$\sep$} in the syntax and the candidate
-usage annotation we are considering is \AgdaBound{r}.
-Then, \AgdaField{+$^{-1}$}\AgdaSpace{}\AgdaBound{r} is a list of pairs of
-annotations \AgdaBound{p} and \AgdaBound{q} that \AgdaBound{r} can split into,
-together with a proof of the splitting.
-For \AgdaField{0\#$^{-1}$} and \AgdaField{1\#$^{-1}$}, inverses to constants,
-we are given the candidate \AgdaBound{r} and typically return an empty list if
-the constraint cannot be satisfied, or a singleton list containing a proof.
-\AgdaField{*$^{-1}$} is used when we encounter scaling, in which case we know
-both the scaling factor \AgdaBound{r} (from the syntax description) and the
-candidate \AgdaBound{q}.
-These inverse operations combine monadically (in fact, applicatively) to give
-inverses to the vector operations of zero, addition, scaling, and basis.
-
-\ExecuteMetaData[\UsageChecktex]{NonDetInverses}
-
-We choose the \AgdaBound{$\V$} of our semantics to be (unannotated) variables.
-For the \AgdaBound{$\C$}, we consider \emph{functions} from candidate usage
-vectors \AgdaBound{R} to the list of elaborated derivations with usage
-annotations given by \AgdaBound{R}.
-The protocol this encodes is that the user will provide an unannotated term
-together with a candidate usage context \AgdaBound{R}, and usage elaboration
-returns a list of possible ways the term could be annotated such that the
-conclusion has usage context \AgdaBound{R}.
-The module name \AgdaModule{U} refers to the fact that we are taking the
-ambient posemiring to be $\mathbf0$ in \AgdaFunction{OpenFam}.
-The effect on \AgdaFunction{OpenFam} is that the usage annotations of any
-contexts we consider are uninformative (hence the \AgdaSymbol{\_} on the left).
-
-\ExecuteMetaData[\UsageChecktex]{C}
-
-To traverse the unannotated terms, we produce a \AgdaRecord{Semantics} over the
-unannotated system \AgdaFunction{uSystem}\AgdaSpace{}\AgdaBound{sys}.
-To write it, we make use of idiom brackets
-\AgdaSymbol{(|}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaSymbol{|)},
-which have the effect of replacing top-level spines of applications by
-(\AgdaDatatype{List}-)applicative applications.
-Field by field, we already know that variables are renameable.
-To interpret a variable, we consider all the possible proofs that such a
-variable could be well annotated, and package them up as a variable term via
-the applicative machinery.
-Finally, for compound terms, we first reify the unannotated subterms, and then
-combine the subterms via a \AgdaFunction{lemma}.
-
-\ExecuteMetaData[\UsageChecktex]{elab-sem}
-
-The \AgdaFunction{lemma} essentially goes through the shape of the premises,
-combining the collections of subterms in the natural way.
-For example, at each
-\AgdaInductiveConstructor{\AgdaUnderscore{}$\dottimes$\AgdaUnderscore{}},
-we take the Cartesian product of the possibilities of each half, and at each
-\AgdaInductiveConstructor{\AgdaUnderscore{}$\sep$\AgdaUnderscore{}},
-we non-deterministically split the usage annotations coming in, and then take
-the Cartesian product.
-When it comes to newly bound variables, the \emph{syntax description} tells us
-their annotations, so there is no further non-determinism introduced here.
-
-\ExecuteMetaData[\UsageChecktex]{lemma-type}
-
-To actually use \AgdaFunction{elab-sem} on terms, we take the associated
-\AgdaFunction{semantics} and pass it the identity environment (an identity
-\emph{renaming} in this case, because $\V$ is a family of variables).
-We use \AgdaFunction{elab-unique}, which further
-checks statically that exactly one derivation is returned.
-The candidate usage vector \AgdaBound{R} will be empty for closed terms, and
-otherwise we have to supply the intended usage annotations.
-
-The example below is the term from \cref{thm:minus}.
-We have instantiated the usage elaborator so that \AgdaField{0\#$^{-1}$},
-\AgdaField{+$^{-1}$}, and \AgdaField{*$^{-1}$} are always singletons (i.e.,
-deterministic), and \AgdaField{1\#$^{-1}$} is a singleton whenever it can be
-(and empty for $\downarrow\downarrow$ and $\wn\wn$).
-The notation \AgdaFunction{\#}\AgdaSpace{}\AgdaBound{i} turns statically known
-natural number \AgdaBound{i} into the corresponding variable pointer, with
-\AgdaNumber{0} being the leftmost.
-In other words, we can write variables using de Bruijn levels.
-The definition \AgdaFunction{\#0} is a workaround for an Agda bug.
-
-\ExecuteMetaData[\Monotonicitytex]{minus}

### 2022-01-21 tex/esop22/semantics.tex
[ writing ] explain map-s/map-p

@@
 
+\subsection{A layer of syntax is functorial}
+
+A basic property of the universe of syntaxes we described in \cref{sec:syntax}
+is that every syntax supports a functorial action on subterms.
+The type of \AgdaFunction{map-s} says that to map a function \AgdaBound{f}
+over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
+domain and codomain usage contexts, and \AgdaBound{f} should be usable
+wherever the domain and codomain usage contexts are similarly related by
+\AgdaBound{F}.
+This generality is needed because usage contexts typically will change between
+a term and its immediate subterms, but there will be a linear combination of
+usage contexts for the subterms that make the original usage contexts.
+\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
+\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
+variables newly bound in that subterm).
+Unlike usage annotations, types are preserved throughout a term (including
+\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).
+
+\ExecuteMetaData[\Maptex]{map-s-type}
+
+The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
+works through the structure \AgdaBound{ps} of premises of the rule applied,
+acting on each subterm it finds.
+Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
+\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
+function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
+--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
+and \AgdaBound{Q$'$}.
+In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
+$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
+collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
+Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
+$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
+$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
+The cases for \AgdaInductiveConstructor{`$\cdot$} and
+\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
+of linearity.
+In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
+\AgdaInductiveConstructor{`$\dot\times$}, which exist in fully structural
+systems, do not involve any changes in the usage contexts.
+
+\ExecuteMetaData[\Maptex]{map-p}
+
 \subsection{Semantic traversal}

### 2022-01-24 agda/LRKits.lagda.tex
[ edit ] minor updates

@@
   open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
-  open import Generic.Linear.Variable Ty rawPoSemiring
+  open import Generic.Linear.Variable Ty rawPoSemiring hiding (_↙ᵛ_; _↘ᵛ_)
   open import Generic.Linear.Environment Ty poSemiring
@@
     env-+ (ρ , sp) =
-      let sumsₗ ↘, sp′ ,↙ sumsᵣ = ρ .asLinRel .linRel .rel-+ₘ (sp , ρ .sums) in
-      record { [_]_⇒ᵉ_ ρ; sums = sumsₗ }
+      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (sp , ρ .fit-here) in
+      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereₗ }
         ↘, sp′ ,↙
-      record { [_]_⇒ᵉ_ ρ; sums = sumsᵣ }
+      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereᵣ }
 
@@
     env-* (ρ , sp) =
-      let sp′ , sums′ = ρ .asLinRel .linRel .rel-*ₘ (sp , ρ .sums) in
-      sp′ , record { [_]_⇒ᵉ_ ρ; sums = sums′ }
+      let fit-here′ , sp′ = ρ .Ψ .rel-*ₘ (sp , ρ .fit-here) in
+      sp′ , record { [_]_⇒ᵉ_ ρ; fit-here = fit-here′ }
 \end{code}
@@
     open Equivalence
+    private module Vᶠᴿ {s} = FRelLeftSemimodule (Vᶠᴿ s)
+    open Vᶠᴿ
 \end{code}
@@
     bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
-    bindEnv ρ .M = [ [ ρ .M │ 0ᴹ ] ─ [ 0ᴹ │ 1ᴹ ] ]
-    bindEnv ρ .asLinRel = ρ .asLinRel ⊕AsLinRel idAsLinRel
-    bindEnv ρ .sums = ρ .sums , ≤*-refl
-    bindEnv ρ .lookup (ll , rr) (lvar (↙ i) q b) =
-      let bl , br = un++ₙ b in
-      ↙ᵏ (≤*-trans rr br) (ρ .lookup ll (lvar i q bl))
-    bindEnv ρ .lookup (ll , rr) (lvar (↘ i) q b) =
-      let bl , br = un++ₙ b in
-      let bl′ = 0*→≤* (ρ .asLinRel .linRel .rel-0ₘ (≤*→0* bl , ll)) in
-      vr (lvar (↘ i) q (bl′ ++ₙ ≤*-trans rr br))
+    bindEnv ρ .Ψ = [ [ ρ .Ψ │ 0ᴿ ]ᴿ ─ [ 0ᴿ │ 1ᴿ ]ᴿ ]ᴿ
+    bindEnv ρ .fit-here = _↘,_,↙_
+      {left = _ ++ _} {right = _ ++ _}
+      (ρ .fit-here , 0*-triv)
+      (+*-identity↘ _ ++ₙ +*-identity↙ _)
+      (0*-triv , ≤*-refl)
+    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar i q b)
+      with un++ₙ sp+ | un++ₙ b
+    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↙ i) q b)
+      | sp+l , sp+r | bl , br = ↙ᵏ
+      (≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br))
+      (ρ .lookup (ρ .Ψ .rel-≤ₘ ≤*-refl (+ₘ-identityʳ→ (sp+l , dl)) ul)
+        (lvar i q bl))
+    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↘ i) q b)
+      | sp+l , sp+r | bl , br = vr (lvar (↘ i) q
+      (≤*-trans (+ₘ-identityʳ→ (sp+l , dl))
+        (0*→≤* (ρ .Ψ .rel-0ₘ (≤*→0* bl , ul)))
+      ++ₙ
+      ≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br)))
 \end{code}
@@
     trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
-    trav ρ (var x) = tm (ρ .lookup (ρ .sums) x)
+    trav ρ (var x) = tm (ρ .lookup (ρ .fit-here) x)
     trav ρ (!-I (⟨ sp* ⟩· M)) =
@@
   ↙^Env : {𝓥 : Ctx → Ty → Set} → Weakening 𝓥 → Weakening [ 𝓥 ]_⇒ᵉ_
-  ↙^Env ↙^𝓥 R0 ρ .M = [ ρ .M │ 0ᴹ ]
-  ↙^Env ↙^𝓥 R0 ρ .asLinRel = [ ρ .asLinRel │ 0AsLinRel ]AsLinRel
-  ↙^Env ↙^𝓥 R0 ρ .sums = ρ .sums , ≤*→0* R0
+  ↙^Env ↙^𝓥 R0 ρ .Ψ = [ ρ .Ψ │ 0ᴿ ]ᴿ
+  ↙^Env ↙^𝓥 R0 ρ .fit-here = ρ .fit-here , ≤*→0* R0
   ↙^Env ↙^𝓥 R0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)

### 2022-01-24 tex/esop22/algebra.tex
[ edit ] minor updates

@@
-The basic algebraic structure we concern ourselves with is \emph{partially
+The basic algebraic structure we deal with is \emph{partially
 ordered semirings}, or \emph{posemirings} for short.
@@
   monoid, $(\Ann, 1, *)$ is a monoid, $+$ and $*$ are monotonic, and $*$
-  distributes over $0$ and $+$ in the usual way on both sides.
+  distributes over $0$ and $+$ on both sides.
 \end{definition}
@@
 analogously to how vectors over a field form a vector space.
+The partial order on such vectors is pointwise.
 

### 2022-01-24 tex/esop22/examples.tex
[ edit ] minor updates

@@
 In an unpublished note, \citet{McBride05} gives a parametrised traversal
-yielding homomorphisms of syntax.
+yielding homomorphisms of syntax, the canonical examples of which are
+simultaneous renaming and simultaneous substitution.
 The parameters are collected in the record \AgdaRecord{Kit}.
@@
 \AgdaField{tm} allows us to extract terms from $\V$-values, as required when
-we use the environment to evaluate a free variable.
+we use the environment to handle a free variable.
 
@@
 
-The idea of renaming is that variables replace variables, whereas with
-substitution, terms replace variables.
+The idea of simultaneous renaming is that variables replace variables, whereas
+with simultaneous substitution, terms replace variables.
 This translates to environments for renaming containing $\sqni$-values

### 2022-01-25 tex/esop22/bunchedrules.tex
[ minor ] changes made during meeting
[ esop22 ] various minor fixes to formatting

@@
 construction.
-Similar combinators have appeared for the same purpose in the work of
-\citet{RPKV20}, including the internal homs we introduce later.
+This idea comes from the work of \citet{RPKV20}, including the $\dotto$ and
+$\wand$ connectives we introduce later.
 To handle binders, which introduce variables, we make
@@
 present in some paper presentations of type systems, and is used as a
-simplifying technique by Allais et al.~\cite{AACMM20}.
+simplifying technique by Allais et al.~\cite{AACMM21}.
 To manage how usage annotations are distributed between premises, we use the
@@
     \begin{prooftree}[comb]
-      \hypo{x :~\sqni A}
-      \infer1{\AgdaInductiveConstructor{var}~x :~\vdash A}
+      \hypo{x : {} \sqni A}
+      \infer1{\AgdaInductiveConstructor{var}~x : {} \vdash A}
     \end{prooftree}
@@
       \hypo{t : \gr1A \vdash B}
-      \infer1{\AgdaInductiveConstructor{⊸I}~t~ \mathord: \mathord\vdash A \multimap B}
+      \infer1{\AgdaInductiveConstructor{⊸I}~t : {} \vdash A \multimap B}
     \end{prooftree}
@@
     \begin{prooftree}[comb]
-      \hypo{(t :~\vdash A \multimap B)}
+      \hypo{(t : {} \vdash A \multimap B)}
       \hypo{\sep}
-      \hypo{(s :~\vdash A)}
-      \infer3{\AgdaInductiveConstructor{⊸E}~t~s :~\vdash B}
+      \hypo{(s : {} \vdash A)}
+      \infer3{\AgdaInductiveConstructor{⊸E}~t~s : {} \vdash B}
     \end{prooftree}
@@
     \begin{prooftree}[comb]
-      \hypo{(t :~\vdash A_i)}
-      \infer1{\AgdaInductiveConstructor{⊕I}_i~t : \vdash A_0 \oplus A_1}
+      \hypo{t : {} \vdash A_i}
+      \infer1{\AgdaInductiveConstructor{⊕I}_i~t : {} \vdash A_0 \oplus A_1}
     \end{prooftree}
@@
     \begin{prooftree}[comb]
-      \hypo{(s : \vdash A \oplus B)}
+      \hypo{(s : {} \vdash A \oplus B)}
       \hypo{\sep}
@@
       \hypo{(u : \gr1B \vdash C))}
-      \infer5{\AgdaInductiveConstructor{⊕E}~s~t~u : \vdash C}
+      \infer5{\AgdaInductiveConstructor{⊕E}~s~t~u : {} \vdash C}
     \end{prooftree}
@@
       \hypo{t : \gr r \cdot \plr{\vdash A}}
-      \infer1{\AgdaInductiveConstructor{!I}~t : \vdash \oc_{\gr r} A}
+      \infer1{\AgdaInductiveConstructor{!I}~t : {} \vdash \oc_{\gr r} A}
     \end{prooftree}
@@
     \begin{prooftree}[comb]
-      \hypo{(s : \vdash \oc_{\gr r} A)}
+      \hypo{(s : {} \vdash \oc_{\gr r} A)}
       \hypo{\sep}
       \hypo{(t : \gr rA \vdash C)}
-      \infer3{\AgdaInductiveConstructor{!E}~s~t : \vdash C}
+      \infer3{\AgdaInductiveConstructor{!E}~s~t : {} \vdash C}
     \end{prooftree}

### 2022-01-25 tex/esop22/intro.tex
[ minor ] changes made during meeting
[ esop22 ] various minor fixes to formatting

@@
 
-In this paper, we concern the metatheory of a class of substructural type
+In this paper, we treat the metatheory of a class of substructural type
 systems related to linear logic~\cite{girard87linear}.
@@
 We build upon both the general structural framework of
-\citet{AACMM20} and the substructural techniques of \citet{WA20}.
-The way \citeauthor{AACMM20} consolidate and codify mechanisation techniques for
+\citet{AACMM21} and the substructural techniques of \citet{WA20}.
+The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
 propositional natural deduction systems based on intrinsically typed syntax and
@@
 By picking a trivial semiring, our work can subsume that of
-\citeauthor{AACMM20}, except for the many pieces of machinery we have not yet
+\citeauthor{AACMM21}, except for the many pieces of machinery we have not yet
 ported to this new framework.

### 2022-01-25 tex/esop22/quant-framework.tex
[ writing ] panic edits
[ writing ] keywords
[ minor ] changes made during meeting
[ edit ] changes and edits to Section 5 (Environments)
[ esop22 ] various minor fixes to formatting

@@
 
+\usepackage[paperheight=235mm, paperwidth=155mm,textwidth=12.2cm,textheight=19.3cm]{geometry}
 \usepackage[conor]{agda}
@@
 \usepackage{stmaryrd}
-\usepackage{todonotes}
+%\usepackage{todonotes}
 \usepackage{turnstile}
@@
 
+\usepackage{bbding}
+\usepackage{graphicx}
+
 \input{../macros.tex}
 \input{lagda.tex}
+
+\newcommand{\todo}[1]{}
 
@@
 
+% \makeatletter
+% \RequirePackage[bookmarks,dvips,unicode,colorlinks=true]{hyperref}%
+%   \def\@citecolor{blue}%
+%   \def\@urlcolor{blue}%
+%   \def\@linkcolor{blue}%
+% \def\UrlFont{\rmfamily}
+% \def\orcidID#1{\smash{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid_color.eps}}}}}
+% \makeatother
+
 \newcommand{\figref}[1]{Figure \ref{#1}}
 \newcommand{\secref}[1]{Section \ref{#1}}
+\newcommand{\defref}[1]{Definition \ref{#1}}
+
+\newcommand{\bob}[1]{\textbf{#1}}
 
@@
 \title{A Framework for Substructural Type Systems
-  \thanks{Supported by organization x.}}
+  \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}}
 %
@@
 %
-\author{Anonymous authors}
-%\author{James Wood\inst{1}\orcidID{0000-0002-8080-3350} \and
-%  Robert Atkey\inst{1}\orcidID{0000-0002-4414-5047}}
+\author{James Wood\inst{1}(\Envelope)\orcidID{0000-0002-8080-3350} \and \\
+  Robert Atkey\inst{1}(\Envelope)\orcidID{0000-0002-4414-5047}}
 %
-\authorrunning{Anon.}
-%\authorrunning{J. Wood, R. Atkey}
+\authorrunning{J. Wood, R. Atkey}
 % First names are abbreviated in the running head.
@@
 %
-%\institute{University of Strathclyde, Glasgow, UK
-%\email{\{james.wood.100,robert.atkey\}@strath.ac.uk}}
+\institute{University of Strathclyde, Glasgow, UK
+\email{\{james.wood.100,robert.atkey\}@strath.ac.uk}}
 %
@@
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+  \keywords{Formalised syntax \and substructural types \and mechanised metatheory
+  \and quantitative typing}
 \end{abstract}
@@
 \input{conc}
+
 %
@@
 \bibliography{quant-framework}
+
+\vfill
+
+{\small\medskip\noindent{\bf Open Access} This chapter is licensed under the terms of the Creative Commons\break Attribution 4.0 International License (\url{http://creativecommons.org/licenses/by/4.0/}), which permits use, sharing, adaptation, distribution and reproduction in any medium or format, as long as you give appropriate credit to the original author(s) and the source, provide a link to the Creative Commons license and indicate if changes were made.}
+
+{\small \spaceskip .28em plus .1em minus .1em The images or other third party material in this chapter are included in the chapter's Creative Commons license, unless indicated otherwise in a credit line to the material.~If material is not included in the chapter's Creative Commons license and your intended\break use is not permitted by statutory regulation or exceeds the permitted use, you will need to obtain permission directly from the copyright holder.}
+
+\medskip\noindent\includegraphics{cc_by_4-0.eps}
+
 \end{document}

### 2022-01-25 tex/esop22/semantics.tex
[ esop22 ] various minor fixes to formatting

@@
 changes to usage annotations.
-We follow the structure given by \citet{AACMM20} closely, with
+We follow the structure given by \citet{AACMM21} closely, with
 the main difference having come already in the syntax and environments.
@@
 
-At this point we introduce a minor generalisation to \AgdaFunction{OpenFam}
+\bob{Move this para} At this point we introduce a minor generalisation to \AgdaFunction{OpenFam}
 and \AgdaFunction{ExtOpenFam}:
@@
 In this paper, all examples support identity environments.
-However, \citet[p.~27]{AACMM20} give some important examples that do not
+However, \citet[p.~27]{AACMM21} give some important examples that do not
 support identity environments, and thus cannot use \AgdaFunction{reify}.
@@
 are bound by the syntax.
-In the examples of \citeauthor{AACMM20}, a bidirectionally typed language only
+In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
 binds variables that synthesise their type, as opposed to those whose type is

### 2022-01-25 tex/esop22/syntax.tex
[ minor ] changes made during meeting
[ esop22 ] various minor fixes to formatting

@@
 up.
-We will see how to automate these proofs in \cref{sec:usage-elaborator}.
+We will see a way of automating these proofs in \cref{sec:usage-elaborator}.
 
@@
 No variables are given \AgdaInductiveConstructor{com} type, similar to how in
-the bidirectional typing syntax of \citet[p.~25]{AACMM20}, no variables are
+the bidirectional typing syntax of \citet[p.~25]{AACMM21}, no variables are
 given \AgdaInductiveConstructor{Check} type.

### 2022-01-25 tex/esop22/examples.tex
[ esop22 ] Reorder Section 6, try to smooth narrative

@@
 TODO: introductory paragraph
+
+\bob{Reification dumped here for now} When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
+we can coerce a \AgdaFunction{Kripke} function into just a
+\AgdaBound{$\C$}-value in an extended context.
+To show this, we instantiate the Kripke function with the renaming
+$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
+pass it an argument environment
+$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
+extended part.
+Post-composition of a renaming onto an environment comes from
+\cref{thm:env-postren}.
+
+\ExecuteMetaData[\Syntactictex]{reify}
+
+We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
+identity environments.
+In this paper, all examples support identity environments.
+However, \citet[p.~27]{AACMM21} give some important examples that do not
+support identity environments, and thus cannot use \AgdaFunction{reify}.
+The feature that causes the lack of support for identity environments is that
+a semantics can make use of the fact that only variables of particular kinds
+are bound by the syntax.
+In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
+binds variables that synthesise their type, as opposed to those whose type is
+checked.
+The semantics of type-checking and elaboration rely on variables synthesising
+their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
+Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
+only binds such synthesising variables.
+Similar phenomena would appear in, say, a call-by-value language where
+variables are values (not computations), or a polarised language where
+variables always have a polarity matching their type.
 

### 2022-01-25 tex/esop22/semantics.tex
[ esop22 ] Reorder Section 6, try to smooth narrative

@@
-Our aim with the following traversals is to derive a recursion scheme for
-terms which abstracts over some of the details of variable binding and
-changes to usage annotations.
-We follow the structure given by \citet{AACMM21} closely, with
-the main difference having come already in the syntax and environments.
+Our goal in this section is to define \AgdaFunction{semantics}, a
+recursor that turns a term into a \AgdaBound{$\C$}-value using a
+\AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
+  ``body'' here}
 
-\subsection{Renameability}
+\ExecuteMetaData[\Semanticstex]{semantics-type}
 
-\todo{Maybe this goes in the Environment section.}
-We saw in \cref{sec:env} that our notion of environment specialises to
-simultaneous renaming by setting $\V$ to be $\sqni$.
-This is encoded below as
-\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
+The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
+representing the interpretations of variables and terms
+respectively. In \cref{sec:traversal} we will see the data that must
+be provided to make a \AgdaFunction{semantics} for a given
+system. Before that, we must see how to deal with the two complicated
+features of our syntax: the usage annotations (\cref{sec:functorial})
+and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}
 
-\ExecuteMetaData[\Renamingtex]{Ren}
+\subsection{A layer of syntax is functorial}\label{sec:functorial}
 
-In a setting where new variables can be bound in the middle of a derivation,
-it is important that the values we carry around while traversing a term can
-handle the existence of variables that appear but they do not use.
-We call any such notion of value \emph{renameable}.
-The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
-\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
-\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
-\AgdaFunction{$\Box^r$}-coalgebra.
+A basic property of the universe of syntaxes we described in \cref{sec:syntax}
+is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
+Its says that to map a function \AgdaBound{f}
+over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
+domain and codomain usage contexts, and \AgdaBound{f} should be usable
+wherever the domain and codomain usage contexts are similarly related by
+\AgdaBound{F}.
 
-\ExecuteMetaData[\Renamingtex]{Boxr}
-\ExecuteMetaData[\Renamingtex]{Renameable}
+This generality is needed because usage contexts change between
+a term and its immediate subterms, but there will be a linear combination of
+the subterms' usage contexts for the subterms that make the original usage contexts.
+\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
+\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
+variables newly bound in that subterm).
+Unlike usage annotations, types are preserved throughout a term (including
+\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).
 
-A renameable notion of value gives rise to a renameable notion of environment,
-essentially by renaming each contained value in an appropriate way.
+\ExecuteMetaData[\Maptex]{map-s-type}
 
-\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
-  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
-  $\plr{-} \env\V \Delta$ for all $\Delta$.
-\end{lemma}
-\begin{proof}
-  For renameable $\V$, we can implement $\mathrm{lift}$ as in
-  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
-  $\V \coloneqq \W \coloneqq \V$.
-  Renaming the environment then becomes an instance of environment composition.
-\end{proof}
+The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
+works through the structure \AgdaBound{ps} of premises of the rule applied,
+acting on each subterm it finds.
+Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
+\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
+function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
+--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
+and \AgdaBound{Q$'$}.
+In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
+$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
+collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
+Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
+$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
+$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
+The cases for \AgdaInductiveConstructor{`$\cdot$} and
+\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
+of linearity.
+In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
+\AgdaInductiveConstructor{`$\dot\times$}, which exist in fully structural
+systems, do not involve any changes in the usage contexts.
 
-\subsection{The Kripke function space}
+\ExecuteMetaData[\Maptex]{map-p}
 
-Earlier, when talking about syntax, we saw \AgdaFunction{Scope} as a simple way
-to let an open family admit context extensions.
-However, a key component of our generic semantic traversal is to make use of
-the open family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
-we store in an environment.
-The definition \AgdaFunction{Kripke} gives an alternative to
-\AgdaFunction{Scope} which interprets the newly bound variables via a
-requirement of $\V$-values rather than extra assumptions for the
-$\C$-computation.
+\subsection{The Kripke function space}\label{sec:kripke}
+
+In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
+judgement-indexed family admit context extensions. However, a key
+component of our generic semantic traversal is to make use of the open
+family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
+we store in an environment.  The definition \AgdaFunction{Kripke}
+gives an alternative to \AgdaFunction{Scope} which interprets the
+newly bound variables via a requirement of $\V$-values rather than
+extra assumptions for the $\C$-computation.
 
@@
 
-When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
-we can coerce a \AgdaFunction{Kripke} function into just a
-\AgdaBound{$\C$}-value in an extended context.
-To show this, we instantiate the Kripke function with the renaming
-$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
-pass it an argument environment
-$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
-extended part.
-Post-composition of a renaming onto an environment comes from
-\cref{thm:env-postren}.
+\subsection{Renameability}
 
-\ExecuteMetaData[\Syntactictex]{reify}
+\todo{Maybe this goes in the Environment section.}
+We saw in \cref{sec:env} that our notion of environment specialises to
+simultaneous renaming by setting $\V$ to be $\sqni$.
+This is encoded below as
+\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
 
-We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
-identity environments.
-In this paper, all examples support identity environments.
-However, \citet[p.~27]{AACMM21} give some important examples that do not
-support identity environments, and thus cannot use \AgdaFunction{reify}.
-The feature that causes the lack of support for identity environments is that
-a semantics can make use of the fact that only variables of particular kinds
-are bound by the syntax.
-In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
-binds variables that synthesise their type, as opposed to those whose type is
-checked.
-The semantics of type-checking and elaboration rely on variables synthesising
-their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
-Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
-only binds such synthesising variables.
-Similar phenomena would appear in, say, a call-by-value language where
-variables are values (not computations), or a polarised language where
-variables always have a polarity matching their type.
+\ExecuteMetaData[\Renamingtex]{Ren}
 
-\subsection{A layer of syntax is functorial}
+In a setting where new variables can be bound in the middle of a derivation,
+it is important that the values we carry around while traversing a term can
+handle the existence of variables that appear but they do not use.
+We call any such notion of value \emph{renameable}.
+The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
+\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
+\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
+\AgdaFunction{$\Box^r$}-coalgebra.
 
-A basic property of the universe of syntaxes we described in \cref{sec:syntax}
-is that every syntax supports a functorial action on subterms.
-The type of \AgdaFunction{map-s} says that to map a function \AgdaBound{f}
-over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
-domain and codomain usage contexts, and \AgdaBound{f} should be usable
-wherever the domain and codomain usage contexts are similarly related by
-\AgdaBound{F}.
-This generality is needed because usage contexts typically will change between
-a term and its immediate subterms, but there will be a linear combination of
-usage contexts for the subterms that make the original usage contexts.
-\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
-\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
-variables newly bound in that subterm).
-Unlike usage annotations, types are preserved throughout a term (including
-\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).
+\ExecuteMetaData[\Renamingtex]{Boxr}
+\ExecuteMetaData[\Renamingtex]{Renameable}
 
-\ExecuteMetaData[\Maptex]{map-s-type}
+A renameable notion of value gives rise to a renameable notion of environment,
+essentially by renaming each contained value in an appropriate way.
 
-The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
-works through the structure \AgdaBound{ps} of premises of the rule applied,
-acting on each subterm it finds.
-Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
-\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
-function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
---- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
-and \AgdaBound{Q$'$}.
-In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
-$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
-collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
-Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
-$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
-$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
-The cases for \AgdaInductiveConstructor{`$\cdot$} and
-\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
-of linearity.
-In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
-\AgdaInductiveConstructor{`$\dot\times$}, which exist in fully structural
-systems, do not involve any changes in the usage contexts.
+\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
+  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
+  $\plr{-} \env\V \Delta$ for all $\Delta$.
+\end{lemma}
+\bob{Proof probably isn't needed}
+\begin{proof}
+  For renameable $\V$, we can implement $\mathrm{lift}$ as in
+  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
+  $\V \coloneqq \W \coloneqq \V$.
+  Renaming the environment then becomes an instance of environment composition.
+\end{proof}
 
-\ExecuteMetaData[\Maptex]{map-p}
-
-\subsection{Semantic traversal}
+\subsection{Semantic traversal}\label{sec:traversal}
 

### 2022-01-25 tex/esop22/examples.tex
[ edit ] reify becomes lemma

--- 2022-01-24 tex/esop22/examples.tex
@@
 TODO: introductory paragraph
+
+\bob{Reification dumped here for now}
+When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
+we can simplify our obligations in giving a \AgdaRecord{Semanitics} by coercing
+\AgdaFunction{Kripke} functions into just
+\AgdaBound{$\C$}-values in an extended context.
+
+%To show this, we instantiate the Kripke function with the renaming
+%$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
+%pass it an argument environment
+%$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
+%extended part.
+%Post-composition of a renaming onto an environment comes from
+%\cref{thm:env-postren}.
+
+%\ExecuteMetaData[\Syntactictex]{reify}
+
+\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
+  If $\V$ is an open family such that there is a function
+  $v : \forallb{{\sqni} \dotto \V}$, we get a function of type
+  $\forallb{\mathrm{Kripke}\,\V\,\C \dotto \mathrm{Scope}\,\C}$ for any $\C$.
+\end{lemma}
+\begin{proof}
+  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
+  That is, $b$ is a Kripke function yielding $\C$-computations, and we want to
+  apply it so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
+  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
+  The $\Box^r$ in the type of $b$ allows us to rename $\Gamma$ to
+  $\Gamma, \gr0\delta$.
+  Then we give the $\wand$-function an argument in context
+  $\gr0\gamma, \Delta$, noting that
+  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
+  as we wanted from the result.
+  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
+  We produce this through the composition
+  $\gr0\gamma, \Delta \env\V \gr0\gamma, \Delta \env{\sqni} \Delta$,
+  using the instance of composition mentioned in \cref{thm:env-postren}.
+  The $\V$-environment follows from $v$ and \cref{thm:env-id}, while the
+  renaming is the complement to that used on $\Box^r$.
+\end{proof}
+
+All of the $\V$s used in examples in this paper support identity environments.
+However, \citet[p.~27]{AACMM21} give some important examples that do not
+support identity environments, and thus cannot use \AgdaFunction{reify}
+(\cref{thm:reify}).
+The feature that causes the lack of support for identity environments is that
+a semantics can make use of the fact that only variables of particular kinds
+are bound by the syntax.
+In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
+binds variables that synthesise their type, as opposed to those whose type is
+checked.
+The semantics of type-checking and elaboration rely on variables synthesising
+their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
+Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
+only binds such synthesising variables.
+Similar phenomena would appear in, say, a call-by-value language where
+variables are values (not computations), or a polarised language where
+variables always have a polarity matching their type.
 

### 2022-01-26 tex/esop22/syntax.tex
[ esop22 ] Reorder Section 6, try to smooth narrative
[ fix ] references
[ edit ] Some minor tidy ups in Section 4

@@
 system of our framework. We use the prototypical system from
-\figref{fig:comb-lr} as a running example. \secref{sec:other-syntaxes}
+\figref{fig:lr-comb} as a running example. \secref{sec:other-syntaxes}
 presents further examples that our framework can express.
@@
 
-\paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}
+% \paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}
 
 A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
-Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and
-a type of conclusion. We assume that there is a
+Each \AgdaRecord{Rule} comprises a \AgdaDatatype{Premises} and
+a conclusion type. We assume that there is a
 $\AgdaBound{Ty} : \AgdaPrimitiveType{Set}$ of types for the system in
@@
 The \AgdaDatatype{Premise} data type describes premises of rules,
-using the bunched combinators from the previous section. A single
+using the bunched combinators from \cref{sec:bunched-rules}. A single
 premise is introduced by the
@@
 the specification of a conclusion type \AgdaBound{A} for this premise.
-The remaining constructors are descriptions for the first-order
-bunched connectives, and will be interpreted directly as such, below.
+The remaining constructors are descriptions for the
+bunched connectives. %, and will be interpreted directly as such, below.
 
@@
 
-\paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}
+% \paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}
 
@@
 
-\paragraph{Terms of a \AgdaRecord{System}.}
+\subsection{Terms of a \AgdaRecord{System}.}\label{sec:terms}
 
@@
 
-With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
-the rest are simply interpretations of premises into bunched combinators.
+The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
+are the interpretation into bunched combinators.
 The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
@@
 The interpretation of a \AgdaRecord{Rule} checks that the rule targets
-the desired type and then interprets the rule's premises \AgdaBound{ps}.
-Notice that the interpretation of the premises is independent of the conclusion
-of the rule, which accounts for the difference in type between
-\AgdaFunction{$\llbracket$\_$\rrbracket$p} and
-\AgdaFunction{$\llbracket$\_$\rrbracket$r}.
+the desired type and then interprets the rule's premises
+\AgdaBound{ps}.  Notice that the interpretation of the premises is
+independent of the conclusion of the rule, which accounts for the use
+of \AgdaFunction{OpenType} in
+\AgdaFunction{$\llbracket$\_$\rrbracket$p} versus
+\AgdaFunction{OpenFam} in \AgdaFunction{$\llbracket$\_$\rrbracket$r}.
 
@@
 variable or a use of a logical rule together with terms for each of
-the required subterms. The \AgdaFunction{Size} argument is where we
-use sized types to record that subterms are smaller than the
-surrounding term.
+the required subterms. The \AgdaFunction{Size} argument is a use of
+Agda's sized types to record that subterms are smaller than the
+surrounding term for the termination checker.
 
@@
 case two (the context and the type).
-Informally,
+The notation
 \AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
@@
 
-Terms defined like this are still quite difficult to write, mainly because of
-frequently changing usage contexts and the need for proofs that they all match
-up.
+Terms defined like this are difficult to write by hand, due to the
+need for proofs that the usage contexts are handled correctly for the rules used.
 We will see a way of automating these proofs in \cref{sec:usage-elaborator}.

### 2022-01-26 tex/esop22/bunchedrules.tex
[ edit ] Major edit to the Semantics section

@@
 \emph{scaling} modality, $\gr r \cdot {-}$. The semantics of the bunched
-connectives we will use are:
+connectives we will use in this paper are:
   \begin{align*}
@@
     (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
-    % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
+    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
     I^*\,\grR &\coloneqq \grR \leq \gr0 \\
@@
                        \times T\,\grP \\
-    % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
-    %                    \to T\,\grQ \to U\,\grR
+    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
+                        \to T\,\grQ \to U\,\grR
   \end{align*}
+where the function connectives $\dotto$ and $\wand$ are not used in typing rules, but are used in the rest of the framework.
 An important point to note is that bunched combinators induce

### 2022-01-26 tex/esop22/env.tex
[ fix ] references
[ edit ] Major edit to the Semantics section

@@
-We have now seen how to generically build data types of terms for a
-given \AgdaRecord{System}, giving us a setting in which to construct
-intrinsically well typed and well annotated terms. Before defining a
-generic assignment of semantics for terms, we need to develop the
-notion of \emph{environment}, a mapping of variables to their semantic
-interpretations that is preserved by structural decompositions and
-variable bindings.
+We have now seen how to build data types of intrinsically well typed
+and well uasged terms for a given \AgdaRecord{System}. In the next
+section, we will define a generic traversal function that assigns a
+``semantics'' to each term. Traversals operate on open terms, so they
+need a way to assign semantics to variables in a typed and usage
+respecting manner. This is the function fulfiled by
+\emph{environments}.
 
 Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
-use the notation $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$ for
+use the notation $\Gamma \sdtstile{}\V A$ meaning $\V\,\Gamma\,A$ for
 the type of inhabitants of $\V$ in the context $\Gamma$ at type
@@
   addition and multiplication.
-  There is a $\sqni$-environment (renaming)
+  There is a $\sqni$-environment (a renaming)
   \[
@@
   \]
-  Linear map $\gr\Psi$ is given by the matrix
-  \(
+  The mapping of variables to variables and matrix giving the linear
+  map $\gr\Phi$ are:
+  \begin{displaymath}
+    \begin{array}{ll}
+    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
+    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
+    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A
+    \end{array}
+    \qquad\qquad
     \begin{pmatrix}
@@
     \end{pmatrix}
-  \),
-  which we can check satisfies the required inequality.
-  The values are given by
-  \begin{align*}
-    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
-    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
-    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A.
-  \end{align*}
-
-  We can read from the columns of the matrix what happened to each of the
-  variables in $\Gamma$.
-  The first column, corresponding to variable $\gr6a : A$, contains two $\gr1$s
+  \end{displaymath}
+  Note that $(\gr 6\,\gr 0\,\gr 1\,\gr 0) = (\gr 1\,\gr 2\,\gr 4)\gr \Phi$.
+  % We can read from the columns of the matrix what happened to each of the
+  % variables in $\Gamma$.
+  The first column of $\gr\Phi$, corresponding to variable $\gr6a : A$, contains two $\gr1$s
   because it has been duplicated (via contraction).
-  Meanwhile, the second and fourth columns are all $\gr0$ because variables
+  The second and fourth columns are all $\gr0$ because variables
   $b$ and $d$ have been discarded (via weakening).
@@
 Relocation will be used when pushing environments into subterms in
-\secref{XX}.
+\secref{sec:traversal}.
 
@@
 % \end{example}
+
+\subsection{Renameability}
+
+\todo{Sort this out: we get a category of renamings, and the comonad a la AACMM}
+We saw in \cref{sec:env} that our notion of environment specialises to
+simultaneous renaming by setting $\V$ to be $\sqni$.
+This is encoded below as
+\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
+
+\ExecuteMetaData[\Renamingtex]{Ren}
+
+In a setting where new variables can be bound in the middle of a derivation,
+it is important that the values we carry around while traversing a term can
+handle the existence of variables that appear but they do not use.
+We call any such notion of value \emph{renameable}.
+The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
+\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
+\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
+\AgdaFunction{$\Box^r$}-coalgebra.
+
+\ExecuteMetaData[\Renamingtex]{Boxr}
+\ExecuteMetaData[\Renamingtex]{Renameable}
+
+A renameable notion of value gives rise to a renameable notion of environment,
+essentially by renaming each contained value in an appropriate way.
+
+\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
+  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
+  $\plr{-} \env\V \Delta$ for all $\Delta$.
+\end{lemma}
+\bob{Proof probably isn't needed}
+\begin{proof}
+  For renameable $\V$, we can implement $\mathrm{lift}$ as in
+  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
+  $\V \coloneqq \W \coloneqq \V$.
+  Renaming the environment then becomes an instance of environment composition.
+\end{proof}

### 2022-01-26 tex/esop22/semantics.tex
[ edit ] Major edit to the Semantics section

@@
-Our goal in this section is to define \AgdaFunction{semantics}, a
-recursor that turns a term into a \AgdaBound{$\C$}-value using a
-\AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
-  ``body'' here}
+Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
+\AgdaFunction{semantics} we define in this section assigns a
+$\C$-value in context $\Gamma$ to every term in context $\Delta$,
+where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
+semantic interpretation of terms ($\V$ being the semantic
+interpretation of variables). Before we can define
+\AgdaFunction{semantics}, we need to treat recursion through rules'
+premises (\cref{sec:functorial}) and extension of environments when
+going under variable binders (\cref{sec:kripke}).
 
-\ExecuteMetaData[\Semanticstex]{semantics-type}
+% Our goal in this section is to define \AgdaFunction{semantics}, a
+% recursor that turns a term into a \AgdaBound{$\C$}-value using a
+% \AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
+%   ``body'' here}
 
-The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
-representing the interpretations of variables and terms
-respectively. In \cref{sec:traversal} we will see the data that must
-be provided to make a \AgdaFunction{semantics} for a given
-system. Before that, we must see how to deal with the two complicated
-features of our syntax: the usage annotations (\cref{sec:functorial})
-and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}
+% \ExecuteMetaData[\Semanticstex]{semantics-type}
+
+% The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
+% representing the interpretations of variables and terms
+% respectively. In \cref{sec:traversal} we will see the data that must
+% be provided to make a \AgdaFunction{semantics} for a given
+% system. Before that, we must see how to deal with the two complicated
+% features of our syntax: the usage annotations (\cref{sec:functorial})
+% and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}
 
@@
 is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
-Its says that to map a function \AgdaBound{f}
+Its type says that to map a function \AgdaBound{f}
 over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
@@
 
+\ExecuteMetaData[\Maptex]{map-s-type}
+
 This generality is needed because usage contexts change between
-a term and its immediate subterms, but there will be a linear combination of
-the subterms' usage contexts for the subterms that make the original usage contexts.
+a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
 \AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
@@
 variables newly bound in that subterm).
-Unlike usage annotations, types are preserved throughout a term (including
-\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).
-
-\ExecuteMetaData[\Maptex]{map-s-type}
+Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
+This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.
 
@@
 In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
-\AgdaInductiveConstructor{`$\dot\times$}, which exist in fully structural
+\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
 systems, do not involve any changes in the usage contexts.
@@
 
-In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
-judgement-indexed family admit context extensions. However, a key
-component of our generic semantic traversal is to make use of the open
-family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
-we store in an environment.  The definition \AgdaFunction{Kripke}
-gives an alternative to \AgdaFunction{Scope} which interprets the
-newly bound variables via a requirement of $\V$-values rather than
-extra assumptions for the $\C$-computation.
+At this point we introduce a minor generalisation to
+\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam}:
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.  We obtain the
+definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} by
+replacing the textual occurrence of \AgdaBound{Ty} by the parameter
+\AgdaBound{I}.
+
+The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
+of function space that describes a $\C$ value parameterised by
+$\Delta$-many additional $\V$s (all correctly typed and usage
+annotated). It is used to describe how to go under binders in a
+Higher-Order Abstract Syntax style---to go under a binder we must
+provide semantic interpretations for all the additional variables:
+
+% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environemnt must be extended with $\Theta$s-worth of $\V$s
+
+% we need the ability to say that
+
+% Kripke V C is given the extension \Theta
+
+% In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
+% judgement-indexed family admit context extensions. However, a key
+% component of our generic semantic traversal is to make use of the open
+% family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
+% we store in an environment.  The definition \AgdaFunction{Kripke}
+% gives an alternative to \AgdaFunction{Scope} which interprets the
+% newly bound variables via a requirement of $\V$-values rather than
+% extra assumptions for the $\C$-computation.
 
@@
 
-Just as $\dotto$ is the internal hom for $\dottimes$, we introduce $\wand$ as
-the internal hom for $\sep$, with \AgdaFunction{$\wand^c$} acting on open types
-analogously to \AgdaFunction{$\sep^c$}.
-It is given by $\plr{T \wand U}\,\grP \coloneqq
-\Pi\grQ,\grR.~\plr{\grR \leq \grP + \grQ} \to T\,\grQ \to U\,\grR$.
+\AgdaFunction{Wrap}
+is a device that turns any type family into an equivalent type family
+that is judgementally injective in its indices, which helps with
+Agda's type inference.
+It turns the type family into a parametrised
+record with a single field \AgdaField{get} whose type is the type in
+the body of the $\lambda$-abstraction.
+For understanding the meaning of
+\AgdaFunction{Kripke}, \AgdaFunction{Wrap} can be ignored.
 
-For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
-ignored.
-\AgdaFunction{Wrap} is a device that turns any type family into an equivalent
-type family that is judgementally injective in its indices.
-It turns the type family into a parametrised record with a single field
-\AgdaField{get} whose type is the type in the body of the $\lambda$-abstraction.
 If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
 \ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
-\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.
-That is to say, the Kripke function is expecting a value for each newly bound
-variable, at the multiplicity of its annotation, together with the resources
-supporting each of those values.
-The need for \AgdaFunction{$\Box^r$} is exemplified by the following construct
-\AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
-version of \AgdaBound{$\Delta$}.
-The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
+\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
+to say, the Kripke function is expecting a value for each newly bound
+variable, at the multiplicity of its annotation, together with the
+resources supporting each of those values. We use the ``magic wand''
+function space here to enforce the invariant that the freshly bound
+variables have usage annotations that are added to the existing
+variables, not shared with them. The use of the
+\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
+the presence of additional variables introduced by weakening.
 
-\bob{Move this para} At this point we introduce a minor generalisation to \AgdaFunction{OpenFam}
-and \AgdaFunction{ExtOpenFam}:
-\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
-\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.
-We obtain the definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam}
-by replacing the textual occurrence of \AgdaBound{Ty} by the parameter
-\AgdaBound{I}.
-This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
-type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
-We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
-to \AgdaDatatype{Ctx}.
+\AgdaFunction{Kripke} is functorial in the $\C$ argument, as witnessed
+by the \AgdaFunction{mapKC} function, which is essentially
+post-composition:
 
-\subsection{Renameability}
+\ExecuteMetaData[\Semanticstex]{mapKC}
 
-\todo{Maybe this goes in the Environment section.}
-We saw in \cref{sec:env} that our notion of environment specialises to
-simultaneous renaming by setting $\V$ to be $\sqni$.
-This is encoded below as
-\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
+% is exemplified by the following construct
+% \AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
+% version of \AgdaBound{$\Delta$}.
+% The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
 
-\ExecuteMetaData[\Renamingtex]{Ren}
-
-In a setting where new variables can be bound in the middle of a derivation,
-it is important that the values we carry around while traversing a term can
-handle the existence of variables that appear but they do not use.
-We call any such notion of value \emph{renameable}.
-The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
-\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
-\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
-\AgdaFunction{$\Box^r$}-coalgebra.
-
-\ExecuteMetaData[\Renamingtex]{Boxr}
-\ExecuteMetaData[\Renamingtex]{Renameable}
-
-A renameable notion of value gives rise to a renameable notion of environment,
-essentially by renaming each contained value in an appropriate way.
-
-\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
-  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
-  $\plr{-} \env\V \Delta$ for all $\Delta$.
-\end{lemma}
-\bob{Proof probably isn't needed}
-\begin{proof}
-  For renameable $\V$, we can implement $\mathrm{lift}$ as in
-  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
-  $\V \coloneqq \W \coloneqq \V$.
-  Renaming the environment then becomes an instance of environment composition.
-\end{proof}
+% \bob{Move this para}
+% This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
+% type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
+% We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
+% to \AgdaDatatype{Ctx}.
 
@@
 
-The aim of this subsection is to give an alternative recursion principle for
-terms that incorporates some of the environment-handling seen in the
-implementations of renaming and substitution.
-The rest of this section assumes the following: a renameable open family
-\AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
-algebra for a layer of syntax at \AgdaBound{$\C$}.
+We can now state the data required to implement a traversal assigning
+semantics to terms. For open families $\V$ and $\C$, interpreting
+variables and terms respectively, we assume that $\V$ is renamable,
+that $\V$ is embeddable in $\C$, and that we have an algebra for a
+layer of syntax, where bound variables are handled using the Kripke
+function space:
+
+% The aim of this subsection is to give an alternative recursion principle for
+% terms that incorporates some of the environment-handling seen in the
+% implementations of renaming and substitution.
+% The rest of this section assumes the following: a renameable open family
+% \AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
+% algebra for a layer of syntax at \AgdaBound{$\C$}.
 
@@
 %\ExecuteMetaData[\Semanticstex]{Comp}
-
-Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
-\cref{thm:lr-bind} as
-\AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
-\AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
-We can think of \AgdaFunction{extend} as our best effort to extend an
-environment by \AgdaBound{$\Theta$} without access to an identity environment
-at \AgdaBound{$\Theta$}.
-We instead take \AgdaBound{$\rho$}, which is an environment covering
-\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an environment
-covering \AgdaBound{$\Theta$}, and glue them together using
-\cref{thm:env-rules}, after having renamed \AgdaBound{$\rho$} via
-\cref{thm:env-ren} to make it fit the new context \AgdaBound{$\Gamma^+$}
-(intended to be \ExecuteMetaData[\Snippetstex]{GT}).
-
-\ExecuteMetaData[\Semanticstex]{extend}
 
@@
 
-For \AgdaFunction{body}, we are given a subterm \AgdaFunction{M}, to which we
-want to apply \AgdaFunction{semantics}.
-To do so, we need an extended version of the initial environment
-\AgdaBound{$\rho$}.
-The best we can achieve without identity environments for \AgdaBound{$\V$} is
-a Kripke function returning an extended environment.
-We use \AgdaFunction{mapK$\C$} to post-compose this function by the
-\AgdaSymbol{$\lambda$}-function taking an extended environment and acting with
-it on \AgdaBound{M}.
+For \AgdaFunction{body}, we are given a subterm \AgdaBound{M}, to
+which we want to apply \AgdaFunction{semantics}.  To do so, we need an
+extended version of the initial environment \AgdaBound{$\rho$}. We
+express this as the generation of a Kripke function that produces the
+extended enviroment given interpretations of the fresh variables. We
+take \AgdaBound{$\rho$}, which is an environment covering
+\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an
+environment covering \AgdaBound{$\Theta$}, and glue them together
+using the inductive rules for generating environments, after having
+renamed \AgdaBound{$\rho$} via \cref{thm:env-ren} to make it fit the
+new context \AgdaBound{$\Gamma^+$} (intended to be
+\ExecuteMetaData[\Snippetstex]{GT}):
+
+\ExecuteMetaData[\Semanticstex]{extend}
+
+% The best we can achieve without identity environments for \AgdaBound{$\V$} is
+% a Kripke function returning an extended environment.
+To define \AgdaFunction{body}, we use \AgdaFunction{mapK$\C$} to
+post-compose the environment extension by the
+\AgdaSymbol{$\lambda$}-function taking an extended environment and
+acting with it on \AgdaBound{M}.
 
 \ExecuteMetaData[\Semanticstex]{body}
+
+% \todo{FIX} Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
+% \cref{thm:lr-bind} as
+% \AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
+% \AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
+% We can think of \AgdaFunction{extend} as our best effort to extend an
+% environment by \AgdaBound{$\Theta$} without access to an identity environment
+% at \AgdaBound{$\Theta$}.

### 2022-01-26 tex/esop22/env.tex
[ edit ] various additions and changes

@@
 annotations; making sure that we have enough resources in $\Gamma$ to
-cover all the demands in $\Delta$. Following the work of \citet{WA20},
+cover all the demands in $\Delta$. Following our previous work~\cite{WA20},
 this accounting is expressed via the presence of a linear
@@
   $\grPprime$, and $\grQprime$ such that
-  $\grPprime \leq \grQprime\gr\Psi$, a function from
+  $\grPprime \leq \grQprime\gr\Psi$, a ``lookup'' function from
   $\grQprime\delta \sqni A$ to $\grPprime\gamma \sdtstile{}\V A$.
-
-  \bob{Write out the Agda notation for environments}
 \end{definition}
 
-By instantiating $\V$, we obtain resource correct versions of familiar
-notions: letting $\V$ be $\sqni$ yields resource correct renamings;
-and letting $\V$ be $\vdash$ (i.e., terms) yields resource correct
+In Agda code, we use \ExecuteMetaData[\Snippetstex]{GvDashA}\ instead of
+$\Gamma \sdtstile{}\V A$ and \ExecuteMetaData[\Snippetstex]{GenvD}\ instead of
+$\Gamma \env\V \Delta$.
+
+The specification of the lookup function has some redundancy.
+Notice that, for $\grQprime\Delta \sqni A$ to hold, we must have
+$\grQprime \leq \langle i \rvert$ for some $i$.
+Instead of $\grPprime \leq \grQprime\gr\Psi$, asking for
+$\grPprime \leq \langle i \rvert\gr\Psi$ would be just as general.
+Additionally, all of the $\V$s we consider satisfy the \emph{subusaging}
+property (that $\grPprime \leq \grP$ yields a coercion
+$\grP\Gamma \sdtstile{}\V A \to \grPprime\Gamma \sdtstile{}\V A$), in which case
+we could just ask for an inhabitant of
+$\plr{\langle i \rvert\gr\Psi}\gamma \sdtstile{}\V A$.
+However, we find the stated definition technically expedient because, by this
+point, basis vectors and raw indices (instead of usage-checked variables) are
+below our level of abstraction.
+We prefer to work with linear relatedness and $\sqni$-variables.
+
+By instantiating $\V$ in \cref{def:lr-env}, we obtain resource-correct versions
+of familiar
+notions: letting $\V$ be $\sqni$ yields resource-correct renamings;
+and letting $\V$ be $\vdash$ (i.e., terms) yields resource-correct
 substitutions.
@@
 \]
-This notation assumes that $\V$ supports subusaging, which is
-always the case when we are using environments for traversals.\bob{Forward ref to where this is explicitly assumed?}
+This notation assumes that $\V$ supports subusaging.
 

### 2022-01-26 tex/esop22/quant-framework.tex
[ esop22 ] Reorder Section 6, try to smooth narrative
[ edit ] various additions and changes

@@
 
-\newcommand{\todo}[1]{}
+\newcommand{\todo}[1]{\textbf{#1}}
 
@@
 
+\appendix
+
+{\bf TODO: explain and format properly.}
+\AgdaInductiveConstructor{!E$'$} and \AgdaInductiveConstructor{!I$'$} are
+variants on \AgdaInductiveConstructor{!E} and \AgdaInductiveConstructor{!I}
+with some of the usage contexts given explicitly (as well as the proofs that the
+usage contexts are correctly related).
+On concrete posemirings, unification is particularly poor at inferring these
+usage contexts from the proofs because addition and multiplication are no longer
+(judgementally) injective.
+\ExecuteMetaData[\HeavyItex]{cojoin-explicit}
+
 \end{document}

### 2022-01-26 tex/esop22/semantics.tex
[ edit ] various additions and changes

@@
 Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
-$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
+$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
 $(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.

### 2022-01-28 tex/esop22/env.tex
[ edit ] another pass over environments

@@
 
-\begin{example}
-  We can form the identity renaming on a two-variable context.
-  \[
-    \id : \plr{\gr rA, \gr sB} \env\sqni \plr{\gr rA, \gr sB}
-  \]
-  We take $\gr\Psi$ to be the identity map, clearly satisfying
-  \(
-    \begin{pmatrix} \gr r & \gr s \end{pmatrix} \leq
-    \begin{pmatrix} \gr r & \gr s \end{pmatrix}\gr\Psi
-  \).
-  When considering values, the fact that $\grPprime \leq \grQprime\gr\Psi$
-  reduces to $\grPprime \leq \grQprime$.
-  The two cases to consider are when $\grQprime\delta \sqni A$ and when
-  $\grQprime\delta \sqni B$.
-  In the first case, $\grPprime \leq \grQprime \leq
-  \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, so we have
-  $\grPprime\plr{A, B} \sqni A$.
-  The second case follows symmetrically.
-\end{example}
+%\begin{example}
+%  We can form the identity renaming on a two-variable context.
+%  \[
+%    \id : \plr{\gr rA, \gr sB} \env\sqni \plr{\gr rA, \gr sB}
+%  \]
+%  We take $\gr\Psi$ to be the identity map, clearly satisfying
+%  \(
+%    \begin{pmatrix} \gr r & \gr s \end{pmatrix} \leq
+%    \begin{pmatrix} \gr r & \gr s \end{pmatrix}\gr\Psi
+%  \).
+%  When considering values, the fact that $\grPprime \leq \grQprime\gr\Psi$
+%  reduces to $\grPprime \leq \grQprime$.
+%  The two cases to consider are when $\grQprime\delta \sqni A$ and when
+%  $\grQprime\delta \sqni B$.
+%  In the first case, $\grPprime \leq \grQprime \leq
+%  \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, so we have
+%  $\grPprime\plr{A, B} \sqni A$.
+%  The second case follows symmetrically.
+%\end{example}
+%
+%\begin{example}
+%  Assume $\Ann$ is the natural numbers with ordering given by $=$ and usual
+%  addition and multiplication.\bob{and $\vdash$ is the terms from a suitable system}
+%  There is a $\vdash$-environment (substitution)
+%  \[
+%    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
+%    \plr{\gr1B, \gr2C}.
+%  \]
+%  The linear map $\gr\Psi$ is given by the matrix
+%  \(
+%    \begin{pmatrix}
+%      \gr0 & \gr0 & \gr1 \\
+%      \gr0 & \gr1 & \gr1
+%    \end{pmatrix}
+%  \),
+%  noticing that $\gr1$ times the first row plus $\gr2$ times the second gives
+%  the original $\grP$.
+%  For $\grQprime = \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, we have
+%  $\grPprime = \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$, and thus
+%  $\grPprime\gamma \vdash z : B$.
+%  For $\grQprime = \begin{pmatrix} \gr0 & \gr1 \end{pmatrix}$, we have
+%  $\grPprime = \begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$, and thus
+%  $\grPprime\gamma \vdash y\,z : C$.
+%\end{example}
 
-\begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and usual
-  addition and multiplication.\bob{and $\vdash$ is the terms from a suitable system}
-  There is a $\vdash$-environment (substitution)
-  \[
-    \plr{\gr0x : A, \gr2y : B \multimap C, \gr3z : B} \env\vdash
-    \plr{\gr1B, \gr2C}.
-  \]
-  The linear map $\gr\Psi$ is given by the matrix
-  \(
-    \begin{pmatrix}
-      \gr0 & \gr0 & \gr1 \\
-      \gr0 & \gr1 & \gr1
-    \end{pmatrix}
-  \),
-  noticing that $\gr1$ times the first row plus $\gr2$ times the second gives
-  the original $\grP$.
-  For $\grQprime = \begin{pmatrix} \gr1 & \gr0 \end{pmatrix}$, we have
-  $\grPprime = \begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$, and thus
-  $\grPprime\gamma \vdash z : B$.
-  For $\grQprime = \begin{pmatrix} \gr0 & \gr1 \end{pmatrix}$, we have
-  $\grPprime = \begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$, and thus
-  $\grPprime\gamma \vdash y\,z : C$.
-\end{example}
+%As a mnemonic, one may use notation like the following to see what values are
+%needed in the environment.
+%This notation assumes that $\V$ supports subusaging.
+%\[
+%  \begin{pmatrix}
+%    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
+%    \gr0A & \gr1\plr{B \multimap C} & \gr1B
+%  \end{pmatrix}
+%  \begin{matrix}
+%    {} \vdash B \\
+%    {} \vdash C
+%  \end{matrix}
+%\]
 
-As a mnemonic, one may use notation like the following to see what values are
-needed in the environment.
-\[
-  \begin{pmatrix}
-    \gr0A & \gr0\plr{B \multimap C} & \gr1B \\
-    \gr0A & \gr1\plr{B \multimap C} & \gr1B
-  \end{pmatrix}
-  \begin{matrix}
-    {} \vdash B \\
-    {} \vdash C
-  \end{matrix}
-\]
-This notation assumes that $\V$ supports subusaging.
+We may informally assign variable names to the entries in the domain
+context.
 
@@
   The mapping of variables to variables and matrix giving the linear
-  map $\gr\Phi$ are:
+  map $\gr\Psi$ are:
   \begin{displaymath}
@@
   \end{displaymath}
-  Note that $(\gr 6\,\gr 0\,\gr 1\,\gr 0) = (\gr 1\,\gr 2\,\gr 4)\gr \Phi$.
+  Note that $(\gr 6\,\gr 0\,\gr 1\,\gr 0) = (\gr 1\,\gr 2\,\gr 4)\gr \Psi$.
   % We can read from the columns of the matrix what happened to each of the
   % variables in $\Gamma$.
-  The first column of $\gr\Phi$, corresponding to variable $\gr6a : A$, contains two $\gr1$s
+  The first column of $\gr\Psi$, corresponding to variable $\gr6a : A$, contains two $\gr1$s
   because it has been duplicated (via contraction).
@@
   Each of the rows in the matrix is a basis vector because variables can only
-  be formed in contexts with basis annotations or less.
+  be formed in contexts with basis-compatible annotations.
 \end{example}
@@
 $\grQ$, we can replace them with any $\grPprime$ and $\grQprime$ that
-are related by the linear map $\rho.\gr\Psi$:
+are related by the linear map $\rho.\gr\Psi$ (that is, the linear map of
+environment $\rho$):
 
@@
 
-When constructing an environment, we can do so by cases on the shape
-of the target context by the following rules, which use the bunched
+When $\V$ supports subusaging, we can construct a $\V$-environment by cases on
+the shape of the target context by the following rules, which use the bunched
 connectives from \secref{sec:bunched-rules}:
@@
 environments into both halves from the split sources; and we can
-create an environment into a singleton context $\gr rA$ when there is
-a context $\gr r$ times smaller than the source context in which we
-can produce a $\V$-value of the appropriate type.
+create an environment into a singleton context $\gr rA$ when we can divide the
+source context by $\gr r$ and produce a $\V$-value in the divided context of the
+appropriate type.
 
@@
 These are not identity environments, so we would have to strengthen the
-induction hypothesis.\bob{FIXME: make this paragraph more neatly flow into the next ones}
+induction hypothesis.
 
@@
 composition. However, by assuming extra structure we can obtain
-identities and composition for environments:\bob{Need to define the
-  $\forallb{\dotto}$ notation and mention when these are real id/comp}
+identities and composition for environments:
+\bob{Need to mention when these are real id/comp}
 
@@
 
-\begin{corollary}\label{thm:env-postren}
-  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
-  $\sigma : \Delta \env\sqni \Theta$, we can get an environment of type
-  $\Gamma \env\U \Theta$.
-  When $\U = {\sqni}$, this gives composition of renamings.
-\end{corollary}
 % \begin{proof}
@@
 
-\subsection{Renameability}
+\paragraph{Renameability}
+Renamings, i.e. $\sqni$-environments, are a particularly important case of
+environments.
+As in the work of \citet{FPT99}, presheaves over renamings are an important
+class of open families.
+In Agda code, we abbreviate $\env\sqni$ (which would usually be
+\AgdaOperator{\AgdaRecord{[}}\AgdaSpace{}%
+\AgdaOperator{\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}}}\AgdaSpace{}%
+\AgdaOperator{\AgdaRecord{]\AgdaUnderscore{}$\Rightarrow^e$\AgdaUnderscore{}}}%
+) as \AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
 
 \todo{Sort this out: we get a category of renamings, and the comonad a la AACMM}
-We saw in \cref{sec:env} that our notion of environment specialises to
-simultaneous renaming by setting $\V$ to be $\sqni$.
-This is encoded below as
-\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
-
-\ExecuteMetaData[\Renamingtex]{Ren}
 
@@
 We call any such notion of value \emph{renameable}.
-The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
-\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
-\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
-\AgdaFunction{$\Box^r$}-coalgebra.
+The cofree renameable open type on an open type $T$ is $\Box^r\,T$ (unrelated to
+the $\Box$ combinator mentioned at the end of \cref{sec:other-syntaxes}), with
+$T$ then being renameable if it forms a $\Box^r$-coalgebra.
 
-\ExecuteMetaData[\Renamingtex]{Boxr}
-\ExecuteMetaData[\Renamingtex]{Renameable}
+\begin{definition}\label{def:box-r}
+  For $T$ an open type,
+  $\plr{\Box^r\,T}\,\Gamma \coloneqq
+    \forallb{\plr{\plr{-} \env{\sqni} \Gamma} \dotto T}$.
+  That is, $\Box^r\,T$ holds at $\Gamma$ when $T$ holds not only at $\Gamma$,
+  but also at any other $\Gamma^+$ which renames to $\Gamma$.
+\end{definition}
+\begin{definition}\label{def:renameable}
+  We say that $T$ is \emph{renameable} whenever there is a function
+  $\mathit{ren}^T : \forallb{T \dotto \Box^r\,T}$.
+  That is, whenever $T$ holds at $\Gamma$, it also holds at any $\Gamma^+$ which
+  renames to $\Gamma$.
+\end{definition}
 
@@
 essentially by renaming each contained value in an appropriate way.
+On the other side, environments are always renameable in their codomains.
 
@@
 \end{proof}
+\begin{lemma}\label{thm:env-postren}
+  $\Gamma \env\V \plr{-}$ is renameable for all $\V$ and $\Gamma$.
+\end{lemma}
+\begin{proof}
+  Corollary of \cref{thm:env-comp}.
+  \todo{Or we may be removing environment composition.}
+\end{proof}

### 2022-01-28 tex/esop22/examples.tex
[ edit ] another pass over environments

@@
   Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
-  That is, $b$ is a Kripke function yielding $\C$-computations, and we want to
-  apply it so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
+  That is, $b$ is a Kripke function yielding $\C$-computations
+  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
   Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
-  The $\Box^r$ in the type of $b$ allows us to rename $\Gamma$ to
+  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
   $\Gamma, \gr0\delta$.

### 2022-01-28 tex/esop22/bunchedrules.tex
[ edit ] pass over bunched typing rules

@@
 $\grR\gamma \vdash A$ indexed by usage context $\grR$, type
-context $\gamma$, and type $A$. We represent typing contexts as
-trees with types at the leaves, and usage contexts as functions
-that assign elements of $\Ann$ to members of the context. Using trees
+context $\gamma$, and type $A$.
+We represent the shape of a context as a nullary-binary tree, with typing and
+usage contexts being functions that assign types and elements of $\Ann$,
+respectively, to leaves of the tree.
+Using trees
 instead of lists for typing contexts has the advantage that extension
 of a context by multiple variables does not lead to complex counting
-arguments to access the pre-existing variables.\todo{left and right
-  constructors for paths}
+arguments to access the pre-existing variables, because context extension is
+(judgementally) injective.
+However, these precise details will eventually become irrelevant, as we will be
+able to use simultaneous renaming to smooth over any structural differences
+between contexts.
 
@@
 
-\figref{fig:lr} presents a prototypical example of a system that our
+\Cref{fig:lr} presents a prototypical example of a system that our
 framework can represent, which is a subsystem of the \name{} system of
-\cite{WA20}. Each rule is given as a constructor: the premises are
+\citet{WA20}. Each rule is given as a constructor: the premises are
 named $p$, $s$, $t$, etc., and the conclusion is a constructor applied
-to those variables. Variables are represented intrinsically as members
+to those metalanguage variables.
+Object language variables are represented intrinsically as members
 of the type $\grR\gamma \sqni A$, which is a proof that the type $A$
@@
 must have a usage annotation $\leq \gr1$ in $\grR$, while all other
-variables must have a usage annotation $\leq \gr0$. The
-\AgdaInductiveConstructor{var} rule imports variables into terms.
+variables must have a usage annotation $\leq \gr0$.
+We use the constructors \AgdaInductiveConstructor{$\swarrow$} and
+\AgdaInductiveConstructor{$\searrow$} to describe a path down the
+nullary-binary tree, terminated by the word
+\AgdaInductiveConstructor{here}.
+The \AgdaInductiveConstructor{var} rule imports variables into terms.
 
@@
 \AgdaInductiveConstructor{!E} for a $\Ann$-annotated exponential
-modality $\oc_{\gr r} A$.
+modality $\oc{\gr r} A$.
 
@@
 \item Rules with multiple typing premises must describe how the
-  resources of the conclusion (always denoted $\grR$) are distributed
+  usages of the conclusion (always denoted $\grR$) are distributed
   across the premises. In the \AgdaInductiveConstructor{⊸E} rule, the
-  resources are separated into two parts $\grP$ and $\grQ$ for the
+  usages are separated into two parts $\grP$ and $\grQ$ for the
   premises. This is an example of a \emph{multiplicative} rule in the
@@
   \AgdaInductiveConstructor{⊕E} we see an example of an
-  \emph{additive} rule, where the resource context $\grQ$ is shared
+  \emph{additive} rule, where the usage context $\grQ$ is shared
   between the premises $t$ and $u$\footnote{There is an unfortunate
     clash of terminology here: multiplicative rules \emph{add} their
-    resources, while additive rules \emph{share} their
-    resources.}. The \AgdaInductiveConstructor{!I} rule uses scaling
-  by $\gr r$ of the resources of the premise.
+    usage contexts, while additive rules \emph{share} their
+    usage contexts.}. The \AgdaInductiveConstructor{!I} rule uses scaling
+  by $\gr r$ of the usages of the premise.
 \end{enumerate}
@@
 This idea comes from the work of \citet{RPKV20}, including the $\dotto$ and
-$\wand$ connectives we introduce later.
+$\wand$ connectives we use later.
 To handle binders, which introduce variables, we make
-use of a combinator that adds a variable with a given $Ann$-annotation
+use of a combinator that adds a variable with a given $\Ann$-annotation
 to an ambient context, without having to explicitly mention the parts
 of the context that have not changed. This technique is already
-present in some paper presentations of type systems, and is used as a
-simplifying technique by Allais et al.~\cite{AACMM21}.
+present in some paper presentations of type systems, and is formalised by
+\citet{AACMM21}.
 To manage how usage annotations are distributed between premises, we use the
@@
                        \times T\,\grP \times U\,\grQ \\
+    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
+                        \to T\,\grQ \to U\,\grR \\
     (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
-                       \times T\,\grP \\
-    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
-                        \to T\,\grQ \to U\,\grR
+                       \times T\,\grP.
   \end{align*}
-where the function connectives $\dotto$ and $\wand$ are not used in typing rules, but are used in the rest of the framework.
+The function connectives $\dotto$ and $\wand$ are not used in typing rules, but
+are used in the rest of the framework (though one can interpret the horizontal
+line in a typing rule as $\dotto$ plus universal quantification).
 An important point to note is that bunched combinators induce
@@
   \end{displaymath}
-  \caption{The prototypical system of \figref{fig:lr} restated in terms of bunched combinators.}
+  \caption{The prototypical system of \cref{fig:lr} restated in terms of bunched combinators.}
   \label{fig:lr-comb}
@@
 
-\figref{fig:lr-comb} shows our prototypical system restated with
+\Cref{fig:lr-comb} shows our prototypical system restated with
 implicit contexts and the bunched combinators. The inductive family is
@@
 in the \AgdaInductiveConstructor{var} rule, the context is completely
-supressed. The \AgdaInductiveConstructor{⊸I} just has to state that a
-new variable with usage annotation $\gr1$ is added to the
+supressed. The \AgdaInductiveConstructor{⊸I} rule just has to state that a
+new variable with usage annotation $\gr1$ and type $A$ is added to the
 context. The \AgdaInductiveConstructor{⊸E} rule uses the separating
 conjunction ($*$) to combine the premises, indicating that the
-resources required by the two premises are added together for the
+usages of the two premises are added together for the
 conclusion. The \AgdaInductiveConstructor{⊕E} rule demonstrates the
@@
 clause terms $t, u$ are combined by separating conjunction, because
-their resources must be combined, but the clause terms are combined by
-the sharing conjunction, because they must use the same resources.
+their usages must be combined, but the clause terms are combined by
+the sharing conjunction, because they have the same usage context.
 
@@
 traversals. If we were to allow arbitrary modification of the context
-in premises, this would not always be possible, since there would be
+in premises, these results would not be possible, since there would be
 no guarantee that a substitution (for instance) could be ``pushed'' up
-from a conclusion to the premises. As we will see in \secref{sec:env},
+from a conclusion to the premises. As we will see in \cref{sec:env},
 our generic notion of environment (e.g., a simultaneous substitution)
-is essentially a linear transformation, and so automatically commutes
+is based around linear transformations, and so automatically commutes
 with the linear combinations of premises induced by the bunched

### 2022-01-29 tex/esop22/env.tex
[ edit ] squash and address todos

@@
 Relocation will be used when pushing environments into subterms in
-\secref{sec:traversal}.
+\cref{sec:traversal}.
 
@@
 the shape of the target context by the following rules, which use the bunched
-connectives from \secref{sec:bunched-rules}:
+connectives from \cref{sec:bunched-rules}:
 \begin{displaymath}
@@
 linear algebraic proofs separately from the rest of an environment.
-For example, for identity and composition of environments (below), \defref{def:lr-env}
+For example, for identity and composition of environments (below), \cref{def:lr-env}
 is easier to use because we can rely on the identity and composition of linear
@@
 
-\paragraph{Identity and Composition} $\V$-environments map variables
-to $\V$-things, so they do not {\it a priori} support identities and
-composition. However, by assuming extra structure we can obtain
-identities and composition for environments:
-\bob{Need to mention when these are real id/comp}
-
-\begin{lemma}[Identity environment]\label{thm:env-id}
-  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
-  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
-\end{lemma}
-
-\begin{lemma}[Composition of environments]\label{thm:env-comp}
-  Given a function
-  \begin{displaymath}
-    \mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to \forallb{\V\,\grQ\delta \dotto \W\,\grP\gamma}
-  \end{displaymath}
-  then we can compose environments $\rho : \Gamma \env\U \Delta$ and
-  $\rho' : \Delta \env\V \Theta$ into an environment
-  $\rho; \rho' : \Gamma \env\W \Theta$.
-\end{lemma}
+%\paragraph{Identity and Composition} $\V$-environments map variables
+%to $\V$-things, so they do not {\it a priori} support identities and
+%composition. However, by assuming extra structure we can obtain
+%identities and composition for environments:
+%\bob{Need to mention when these are real id/comp}
+%
+%\begin{lemma}[Identity environment]\label{thm:env-id}
+%  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
+%  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
+%\end{lemma}
+%
+%\begin{lemma}[Composition of environments]\label{thm:env-comp}
+%  Given a function
+%  \begin{displaymath}
+%    \mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to \forallb{\V\,\grQ\delta \dotto \W\,\grP\gamma}
+%  \end{displaymath}
+%  then we can compose environments $\rho : \Gamma \env\U \Delta$ and
+%  $\rho' : \Delta \env\V \Theta$ into an environment
+%  $\rho; \rho' : \Gamma \env\W \Theta$.
+%\end{lemma}
 
@@
 
-\bob{Forward reference to where we use identity and composition}
+%\bob{Forward reference to where we use identity and composition}
 
@@
 environments.
+Renamings form a category, with identity and composition following from the
+identity and composition of linear maps.
 As in the work of \citet{FPT99}, presheaves over renamings are an important
@@
 ) as \AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.
-
-\todo{Sort this out: we get a category of renamings, and the comonad a la AACMM}
 
@@
 essentially by renaming each contained value in an appropriate way.
-On the other side, environments are always renameable in their codomains.
+On the other side, environments admit renamings of their codomains in the
+opposite direction to that given by renameability.
 
@@
 \end{lemma}
-\bob{Proof probably isn't needed}
-\begin{proof}
-  For renameable $\V$, we can implement $\mathrm{lift}$ as in
-  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
-  $\V \coloneqq \W \coloneqq \V$.
-  Renaming the environment then becomes an instance of environment composition.
-\end{proof}
+%\bob{Proof probably isn't needed}
+%\begin{proof}
+%  For renameable $\V$, we can implement $\mathrm{lift}$ as in
+%  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
+%  $\V \coloneqq \W \coloneqq \V$.
+%  Renaming the environment then becomes an instance of environment composition.
+%\end{proof}
 \begin{lemma}\label{thm:env-postren}
-  $\Gamma \env\V \plr{-}$ is renameable for all $\V$ and $\Gamma$.
+  From $\Gamma \env\V \Delta$ and $\Delta \env{\sqni} \Theta$, we get
+  $\Gamma \env\V \Theta$.
 \end{lemma}
-\begin{proof}
-  Corollary of \cref{thm:env-comp}.
-  \todo{Or we may be removing environment composition.}
-\end{proof}
+\begin{proofsketch}
+  Notice that the lookup component of an environment maps variables in the
+  codomain to values in the domain.
+  We can apply the renaming to these variables.
+\end{proofsketch}
+%\begin{proof}
+%  Corollary of \cref{thm:env-comp}.
+%  \todo{Or we may be removing environment composition.}
+%\end{proof}

### 2022-01-29 tex/esop22/examples.tex
[ edit ] squash and address todos

@@
-TODO: introductory paragraph
+In this section, we provide three example uses of semantic traversals: generic
+renaming and substitution, a usage elaborator, and a denotational semantics.
+The reader is also encouraged to see the far greater range of examples in the
+work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
+Renaming and substitution are essential results, while the latter two examples
+focus on usage annotations.
 
-\bob{Reification dumped here for now}
-When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
-we can simplify our obligations in giving a \AgdaRecord{Semanitics} by coercing
+A result we will use throughout this section is \emph{reification}.
+When we have an index-preserving mapping from usage-checked variables to
+$\V$-environments, we can construct environments of the form
+$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
+This lets us write the \AgdaFunction{reify} function, which  simplifies our
+obligations in giving a \AgdaRecord{Semanitics} by coercing
 \AgdaFunction{Kripke} functions into just
@@
   The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
-  We produce this through the composition
-  $\gr0\gamma, \Delta \env\V \gr0\gamma, \Delta \env{\sqni} \Delta$,
-  using the instance of composition mentioned in \cref{thm:env-postren}.
-  The $\V$-environment follows from $v$ and \cref{thm:env-id}, while the
-  renaming is the complement to that used on $\Box^r$.
+  We produce this via \cref{thm:env-postren} from an environment
+  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
+  and a renaming which is the complement to that used on $\Box^r$.
 \end{proof}
@@
 \emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
-A world-indexed relation over a poset of worlds \AgdaBound{W} is a set over
-which
+A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
+\AgdaBound{W} is a set over which
 we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
 property with respect to the order on \AgdaBound{W}.
-
-\ExecuteMetaData[\WReltex]{WRel}
+The rendered Agda code for world-indexed relations and constructions on them
+can be found in \cref{sec:den-sem-extras}.
 
@@
 
-Morphisms between world-indexed relations \AgdaBound{R} and \AgdaBound{S}
-consist of a mapping between the underlying sets such that that mapping
-preserves relatedness from \AgdaBound{R} to \AgdaBound{S}.
+Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
+and \AgdaBound{S} consist of a mapping between the underlying sets such that, at
+each fixed world \AgdaBound{w}, the mapping preserves relatedness from
+\AgdaBound{R} to \AgdaBound{S}.
 
-\ExecuteMetaData[\WReltex]{WRelMor}
-
-\todo{Define big intersection.}
 When the poset of worlds forms a (relational) commutative monoid, such
@@
 
-%\ExecuteMetaData[\WReltex]{IR}
-%\ExecuteMetaData[\WReltex]{tensorR}
-%\ExecuteMetaData[\WReltex]{lollyR}
-
 The final piece of sematics we need is a \emph{bang} operator.
@@
 particular index \AgdaBound{r}.
-
-\ExecuteMetaData[\WReltex]{Bang}
+These laws are listed in \cref{sec:den-sem-extras}.
 
@@
 
-To associate semantics to syntax, we start as standard by associating
-world-indexed relations to types.
-We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
-\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
-context, concatenation, and usage annotations on singletons, respectively.
+%To associate semantics to syntax, we start as standard by associating
+%world-indexed relations to types.
+%We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
+%\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
+%context, concatenation, and usage annotations on singletons, respectively.
+%
+%\ExecuteMetaData[\WReltex]{sem}
+%
+%The semantics of a term is then to be a morphism from the interpretation of the
+%context to the interpretation of the term's type.
+%
+%\ExecuteMetaData[\WReltex]{sem-vdash}
 
-\ExecuteMetaData[\WReltex]{sem}
-
-The semantics of a term is then to be a morphism from the interpretation of the
-context to the interpretation of the term's type.
-
-\ExecuteMetaData[\WReltex]{sem-vdash}
-
-Variables are given semantics by \AgdaFunction{lookup$^R$} (definition omitted).
+The semantics of a type is given by \AgdaFunction{$\llbracket$\_$\rrbracket$},
+which maps into world-indexed relations.
+The function type is interpreted using \AgdaFunction{$\multimap^R$} and
+\AgdaField{!$^R$}.
+Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
+\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}.
+Terms are interpreted as morphisms by the open family
+\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
+Variables are interpreted by \AgdaFunction{lookup$^R$} (definition omitted).
 
@@
 
-Now, we give a \AgdaRecord{Semantics}.
+Now we give a \AgdaRecord{Semantics}.
 The choice of \AgdaBound{$\V$} as
@@
 given that a standard denotational semantics would not use intermediate
-environments in the same sense as renaming and substitution, but allows us to
+environments in the same sense as renaming and substitution, but it allows us to
 reuse the standard facts that variables support renaming and identity
@@
 
-In order to map open terms to interpretations, we take the action of the
-semantics and give the identity renaming as the starting environment.
-
-\ExecuteMetaData[\WReltex]{wrel}
+Then, the semantics of terms is given by the function
+\AgdaFunction{semantics}\AgdaSpace{}\AgdaFunction{Wrel}\AgdaSpace{}%
+\AgdaFunction{1$^r$}, where \AgdaFunction{1$^r$} is the identity renaming.
+%In order to map open terms to interpretations, we take the action of the
+%semantics and give the identity renaming as the starting environment.
+%
+%\ExecuteMetaData[\WReltex]{wrel}
 

### 2022-01-29 tex/esop22/semantics.tex
[ edit ] squash and address todos

@@
 
-% In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
+% In \cref{sec:terms}, we defined \AgdaFunction{Scope} to let a
 % judgement-indexed family admit context extensions. However, a key
@@
 
-\AgdaFunction{Kripke} is functorial in the $\C$ argument, as witnessed
-by the \AgdaFunction{mapKC} function, which is essentially
+\AgdaFunction{Kripke} is functorial in the \AgdaBound{$\C$} argument,
+as witnessed by the \AgdaFunction{mapKC} function, which is essentially
 post-composition:

### 2022-01-29 tex/esop22/syntax.tex
[ edit ] squash and address todos

@@
 system of our framework. We use the prototypical system from
-\figref{fig:lr-comb} as a running example. \secref{sec:other-syntaxes}
+\cref{fig:lr-comb} as a running example. \Cref{sec:other-syntaxes}
 presents further examples that our framework can express.
@@
 
-% \paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}
+% \paragraph{\Cref{fig:lr-comb} as a \AgdaRecord{System}.}
 
 As an example, we transcribe the system defined in
-\figref{fig:lr-comb} into a description.  We give the set of types of
+\cref{fig:lr-comb} into a description.  We give the set of types of
 this system as a data type \AgdaDatatype{Ty} (together with a base
@@
 
-\subsection{Terms of a \AgdaRecord{System}.}\label{sec:terms}
+\subsection{Terms of a \AgdaRecord{System}}\label{sec:terms}
 
@@
 need for proofs that the usage contexts are handled correctly for the rules used.
-We will see a way of automating these proofs in \cref{sec:usage-elaborator}.
+We give an example of a hand-written term in \cref{sec:explicit-term}, and will
+see a way of automating the necessary proofs in \cref{sec:usage-elaborator}.
 

### 2022-01-30 tex/esop22/bunchedrules.tex
[ edit ] prepare for submission

@@
 in the \AgdaInductiveConstructor{var} rule, the context is completely
-supressed. The \AgdaInductiveConstructor{⊸I} rule just has to state that a
+suppressed. The \AgdaInductiveConstructor{⊸I} rule just has to state that a
 new variable with usage annotation $\gr1$ and type $A$ is added to the
@@
 
-Bunched combinators, along with supression of unchanged typing
+Bunched combinators, along with suppression of unchanged typing
 contexts, leads to a more streamlined presentation of the system

### 2022-01-30 tex/esop22/env.tex
[ edit ] prepare for submission

@@
 We have now seen how to build data types of intrinsically well typed
-and well uasged terms for a given \AgdaRecord{System}. In the next
+and well usaged terms for a given \AgdaRecord{System}. In the next
 section, we will define a generic traversal function that assigns a
@@
 need a way to assign semantics to variables in a typed and usage
-respecting manner. This is the function fulfiled by
+respecting manner. This is the function fulfilled by
 \emph{environments}.

### 2022-01-30 tex/esop22/examples.tex
[ edit ] prepare for submission

@@
 This lets us write the \AgdaFunction{reify} function, which  simplifies our
-obligations in giving a \AgdaRecord{Semanitics} by coercing
+obligations in giving a \AgdaRecord{Semantics} by coercing
 \AgdaFunction{Kripke} functions into just
@@
 
-The final piece of sematics we need is a \emph{bang} operator.
+The final piece of semantics we need is a \emph{bang} operator.
 We allow the

### 2022-01-30 tex/esop22/quant-framework.tex
[ edit ] squash and address todos
[ edit ] prepare for submission

@@
 \usepackage{catchfilebetweentags}
-\usepackage{cleveref}
+\usepackage{hyperref}
+\usepackage[noabbrev,nameinlink]{cleveref}
 \usepackage{cmll}
@@
 
-\newcommand{\figref}[1]{Figure \ref{#1}}
-\newcommand{\secref}[1]{Section \ref{#1}}
-\newcommand{\defref}[1]{Definition \ref{#1}}
+\newcommand{\bob}[1]{\textbf{#1}}
 
-\newcommand{\bob}[1]{\textbf{#1}}
+\newenvironment{proofsketch}{%
+  \renewcommand{\proofname}{Proof sketch}\proof}{\endproof}
+
+\def\orcidID#1{\smash{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid_color.eps}}}}}
 
@@
 %   \item Example(s)
-%   \item Intepretation as Terms
+%   \item Interpretation as Terms
 %   \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
@@
 
-{\bf TODO: explain and format properly.}
+\section{A term written without the usage elaborator}\label{sec:explicit-term}
+
+Below is a fully witten-out version of the example term at the end of
+\cref{sec:usage-elaborator}.
 \AgdaInductiveConstructor{!E$'$} and \AgdaInductiveConstructor{!I$'$} are
@@
 usage contexts are correctly related).
-On concrete posemirings, unification is particularly poor at inferring these
-usage contexts from the proofs because addition and multiplication are no longer
-(judgementally) injective.
+On concrete posemirings (as in this example), unification is particularly poor
+at inferring these usage contexts from the proofs because addition and
+multiplication are no longer (judgementally) injective.
 \ExecuteMetaData[\HeavyItex]{cojoin-explicit}
 
+\section{Omitted pieces of the denotational semantics}\label{sec:den-sem-extras}
+
+\ExecuteMetaData[\WReltex]{WRel}
+
+\ExecuteMetaData[\WReltex]{IR}
+\ExecuteMetaData[\WReltex]{tensorR}
+\ExecuteMetaData[\WReltex]{lollyR}
+
+\todo{Define big intersection.}
+\ExecuteMetaData[\WReltex]{WRelMor}
+
+\ExecuteMetaData[\WReltex]{Bang}
+
 \end{document}

### 2022-01-30 tex/esop22/semantics.tex
[ edit ] prepare for submission

@@
 The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
-of function space that describes a $\C$ value parameterised by
+of function space that describes a $\C$ value parametrised by
 $\Delta$-many additional $\V$s (all correctly typed and usage
@@
 
-% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environemnt must be extended with $\Theta$s-worth of $\V$s
+% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environment must be extended with $\Theta$s-worth of $\V$s
 
@@
 semantics to terms. For open families $\V$ and $\C$, interpreting
-variables and terms respectively, we assume that $\V$ is renamable,
+variables and terms respectively, we assume that $\V$ is renameable,
 that $\V$ is embeddable in $\C$, and that we have an algebra for a
@@
 \AgdaBound{$\V$}-value to a \AgdaBound{$\C$}-value.
-To deal with a structure of subtems \AgdaBound{M}, we use the functoriality of
+To deal with a structure of subterms \AgdaBound{M}, we use the functoriality of
 the syntactic structure to consider each subterm separately.
@@
 express this as the generation of a Kripke function that produces the
-extended enviroment given interpretations of the fresh variables. We
+extended environment given interpretations of the fresh variables. We
 take \AgdaBound{$\rho$}, which is an environment covering

### 2022-02-19 tex/esop22/examples.tex
[ esop ] Inline the appendix

@@
 
-The example below is of a term needed to show that, in the
-$\{\gr0, \gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
-$\oc\gr\omega$ forms a comonad.
+We can now use the elaborator to automatically infer the usage
+annotations for the example at the end of \cref{sec:terms}. This
+allows us to write:
+\ExecuteMetaData[\PaperExamplestex]{cojoin}
 We have instantiated the usage elaborator so that:
-\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and empty on
-$\gr1$;
-\AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and $\gr\omega$, and empty on
-$\gr0$;
-\AgdaField{+$^{-1}$} gives $\gr0 \mapsto [(\gr0,\gr0)]$,
-$\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
-$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and
-\AgdaField{*$^{-1}$} gives $(\gr\omega, \gr0) \mapsto [\gr0]$,
-$(\gr\omega, \gr1) \mapsto []$, and
-$(\gr\omega, \gr\omega) \mapsto [\gr\omega]$
-(omitting $(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity).
-Note that we do not consider splitting $\gr\omega$ up as, say,
-$\gr1 + \gr\omega$, because this splitting would introduce more
-non-determinism but not allow any more terms to be typed.
-As such, the only non-determinism comes when we have variables annotated
-$\gr1$ and need to do an additive split, like when we apply the
-\AgdaInductiveConstructor{!E} rule below.
-At this point, the variable can become either $\gr0$-annotated in the left
-subterm and $\gr1$-annotated on the right, or vice-versa.
-We will find that, because the left subterm wants to use that variable, the
-former choice will be rejected.
-The function \AgdaFunction{var\#} is a convenience for converting statically
-known natural numbers, representing de Bruijn \emph{levels}, into variable
-terms.
-
-\ExecuteMetaData[\PaperExamplestex]{cojoin}
+\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
+empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
+$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
+\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
+$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
+gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
+[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
+$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
+consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
+this splitting would introduce more non-determinism but not allow any
+more terms to be typed. As such, the only non-determinism comes when
+we have variables annotated $\gr1$ and need to do an additive split,
+like when we apply the \AgdaInductiveConstructor{!E} rule below. At
+this point, the variable can become either $\gr0$-annotated in the
+left subterm and $\gr1$-annotated on the right, or vice-versa. We will
+find that, because the left subterm wants to use that variable, the
+former choice will be rejected. The function \AgdaFunction{var\#} is a
+convenience for converting statically known natural numbers,
+representing de Bruijn \emph{levels}, into variable terms.
 
@@
 property with respect to the order on \AgdaBound{W}.
-The rendered Agda code for world-indexed relations and constructions on them
-can be found in \cref{sec:den-sem-extras}.
+The Agda code for world-indexed relations and constructions on them
+can be found in \cite{generic-lr}. %\cref{sec:den-sem-extras}.
 
@@
 particular index \AgdaBound{r}.
-These laws are listed in \cref{sec:den-sem-extras}.
+These laws are listed in the \texttt{Generic.Linear.Example.WRel} module in \cite{generic-lr}. %\cref{sec:den-sem-extras}.
 

### 2022-02-19 tex/esop22/syntax.tex
[ edit ] prepare for submission
[ esop ] Inline the appendix

@@
 
-Terms defined like this are difficult to write by hand, due to the
-need for proofs that the usage contexts are handled correctly for the rules used.
-We give an example of a hand-written term in \cref{sec:explicit-term}, and will
-see a way of automating the necessary proofs in \cref{sec:usage-elaborator}.
+Terms in this data type are difficult to write by hand, due to the
+need for proofs that the usage contexts are handled correctly. For
+example, the following term is needed to show that, in the $\{\gr0,
+\gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
+$\oc\gr\omega$ forms a comonad. \AgdaInductiveConstructor{!E$'$} and
+\AgdaInductiveConstructor{!I$'$} are variants on
+\AgdaInductiveConstructor{!E} and \AgdaInductiveConstructor{!I} with
+some of the usage contexts given explicitly (as well as the proofs
+that the usage contexts are correctly related). On concrete
+posemirings (as in this example), unification is particularly poor at
+inferring these usage contexts from the proofs because addition and
+multiplication are no longer (judgementally) injective.
+\ExecuteMetaData[\HeavyItex]{cojoin-explicit} Writing terms like this
+is clearly unsustainable. We will see a way of automating the
+necessary proofs via a \AgdaRecord{System}-generic elaborator in
+\cref{sec:usage-elaborator}.
 
@@
 How to observe this invariant is covered in the latter paper, so we will not
-repeat it here (having not yet seen how to write traverals on terms).
+repeat it here (having not yet seen how to write traversals on terms).
 

### 2022-02-20 tex/esop22/examples.tex
[ edit ] citation as noun
[ esop ] idiom brackets in text

@@
 To write it, we make use of idiom brackets
-\AgdaSymbol{(|}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaSymbol{|)},
+\AgdaSymbol{\ensuremath{\llparenthesis}}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaSymbol{\ensuremath{\rrparenthesis}},
 which have the effect of replacing top-level spines of applications by
@@
 The Agda code for world-indexed relations and constructions on them
-can be found in \cite{generic-lr}. %\cref{sec:den-sem-extras}.
+can be found in \citet{generic-lr}. %\cref{sec:den-sem-extras}.
 

### 2022-02-21 tex/esop22/rebuttal.md
[ esop ] reviews and rebuttal

Rebuttal
========

Thank you to the reviewers for their careful reviews and kind comments
and suggestions. We have commented below on the major questions and
concerns brought up by the reviews, and given our suggestions for how
we will amend in the paper for any further revision.

Reviewer A
----------

> Please clarify the "Page 9: As far as I can tell, the *c and I*c and
> r.c used in the definition of [[_]]p are not defined defined
> anywhere. " to help me understand this definition.

These are liftings of the corresponding bunched connectives from page
5 to be indexed by vector/context pairs. We will make this clear in
any revised version. The $\dot\times$ product works because it is a
polymorphic pointwise use of $\times$ for any indexing set. Our
notation on page 5 implies that it is only defined for vectors, but it
is actually polymorphic. We will state this explicitly.

> It might be worth giving some general guidance about how to
> think about the orientation of <=

We think of the ordering `x <= y` as convertibility; so that something
labelled with `x` is sufficient to supply a demand of `y`. This gives
the upside-down ordering when compared the standard natural
numbers. Abel and Bernardy (ICFP 2020) make the same choice.

> In the "An element of a chosen posemiring" paragraph, it would help
> to define <= on vectors (the green scripty P's and Q's that occur in
> the typing rules) --- I assume this means pointwise <= of resources?

Yes, this is the pointwise ordering.

> It would have helped me to have an explicit comment about what
> isomorphisms you're eliding when representing resource vectors (the
> script P and script Q) in different ways.

Conceptually, contexts with usage annotations are pairs of vectors,
one of semiring values and one of types. We have found it useful in
the formalisation to store these vectors in an "unnormalised" fashion
as trees that record their history of appends used to construct
them. This non-canonicity does not cause a problem because we can
always apply renaming to shift to a different association
structure. The isomorphisms come from renamings. We never compare
contexts for equality directly, we only make demands on what variables
appear in them and what their usage annotations and types are. This is
a minor formalisation "trick" that is difficult to get across in a
paper intended for a general audience.

> Page 5, bottom: As far as I understand, the resource 0 doesn't have
> any special status with respect to the preorder <=. So why does R <=
> 0 get a "connective" in the framework

The vector 0 is special with respect to the vector space structure,
being the unit for the addition. There is no multiplication structure
between vectors, and so no use for the vector of all 1s. I* is also
defined with `R <= 0` to ensure it is the unit of the * connective.

> In Definition 5.1, I found the role of the Q' confusing at
> first

You are correct about what it is used for -- it is a convenience. We
will add a comment to clarify.

> [Agda notation]

We will track these down and add the necessary definitions.

> It would be good to give some empirical sense for how well this
> resource elaborator works --- what examples is it able to handle,
> what times out, etc.?

We will elaborate on the elaborator, as also suggested by the other
reviewers.

> The type systems for Amortized Automated Resource Analysis by
> Hofman, Jost, Hoffmann, etc. have the flavor of these
> resource-annotated type systems -- can you encode those?

We can't definitively say, but we think the answer is that we cannot
directly encode these systems because the contraction rule they use
does not fully preserve the types: a variable of type A can be
duplicated to two variables of types A1 and A2 whose integrated
potentials add up to the ones in the type A. In our system,
duplicating a variable affects its top-level usage annotation, but
preserves its type. It may be possible to encode a system for
amortised reasoning however.

> It would also be nice to show that you can encode a system with
> "non-logical" rules, e.g. some of the primitives for differential
> privacy in [18].

This is certainly possible. For the Reed and Pierce system, we would
need to add rules for the graded differentially private distribution
monad, which in our system is just some more rules.

> What are the prospects for extending the framework beyond simple
> types, e.g. to handle resource polymorphism (quantification over
> resources) or dependent types?

We think that extension to polymorphism (over types or usage
annotations) would be relatively straightforward conceptually given
that the kinding or resourcing contexts would not interact with the
substructural parts. It would become harder to make a generic
elaborator, however, due to the need to reason about open terms in the
semiring (something that the Granule systems needs to use external
solvers for). Extension to dependent types is much harder; though we
think that our insights will help extend any formalisation of non
substructural dependent types to having semiring resource annotations.

Reviewer B
----------

> Can only really be followed by an expert in this kind of extremely
> general formalization in Agda.

We have tried to mitigate this by using examples throughout. For a
revised final version we will guide our examples by trying to get
across how the framework is used in practice. For example, an example
of using the elaborator defined in Section 7.3, as suggested.

> Extremely dense in notation and fairly sparse in giving intuitions
> of what's going on.

In addressing reviewer C's points, we will add more intuition to help
readers used to extrinsic presentations of type systems in
Section 3. We will simplify Section 4, as explained below. We will
also more high-level discussion in the later sections on what is
possible with the syntactic traversals for generic metatheory, again
see below. To make space will we move some of the technical content of
Section 5 to an appendix to concentrate more on the high level ideas
of using linear algebra.

> Explanation after page 8 (construction of terms) is too opaque

We can simplify this section by removing the I--OpenFam and I--OpenExt
definitions, since I is always Ty. We will then add proper
explanations for the remaining three defnitions (OpenType, OpenFam and
OpenExtFam) and point to the places where they are required in our
development. The more general definitions are not actually used until
the `extend` function on page 18, where we will explicitly say that we
need the extra generality in a revised version.

> How does one use this for *metatheory* as opposed just creating
> syntaxes.
and
> What kind of metatheory can you do?

We can define functions such as generic renaming and substitution, and
generic denotational semantics. But we cannot yet (easily) reason
about them beyond their types. We expect that the `Simulation`
construct from the Allais et al. paper [3] will smoothly transfer to
this setting, which allow us to prove equational properties of these
traversals. For example, the substitution of a composition of two
simultaneous substitutions being the composition of two separate
subsitutions. In this paper, we wanted to show that the core of Allais
et al. transfers to the semiring-annotated setting, where even
defining the syntax and the operations of renaming and substitution
are nontrivial and require a linearly algebraic approach. We leave
transfer of the equational reasoning principles to future work. We
expect that given our foundational work, equational reasoning will
follow easily.

> what kind of "things" can I expect to reasonably define as
> traversals?
and
> The applicability of the framework in terms of metareasoning beyond
> the use cases in the paper is not really discussed and the
> automation/elaborator is not really exercised in the presentation.

Traversals are a "nicer" fold / recursion principle over the syntax,
that automatically handles dealing with updating of parameters when
going under binders. This is particularly useful in the renaming and
substitution cases, where the parameter is the renaming or
substitution. Using a traversal saves on repeated work that would be
required if one were to do this directly with the "native" recursion
principle for the type of Terms. The key point is that defining
renaming and substitution for substructural systems is not trivial,
and our framework yields a generic definition for a wide range of
such systems.

> An elaborator for this is great, but how can I use it?

We will give an actual use of the elaborator to show how it is used in
the context of Examples 8-10 (the monotone subtraction function),
which will also be compressed to give more space to discussion of the
framework.

> Concretely, I think some of the examples in the paper (e.g. Examples
> 8-10) could be cut for extra space so that definitions, roadmaps and
> actual explanations can be given in a way that makes the paper
> understandable to a wider audience.

We take the reviewer's point that examples 8-10 are more on an example
application of one part of the metatheory (a denotational semantics)
than the overall picture. We will do as the reviewer suggests to
compress these examples to give more space for high level discussion
of the features of the framework. In particular, we will add a short
discussion of how to do denotational semantics for other systems
within the framework. The main point of this section is that our
framework provides a way to formalise much of Abel and Bernardy's ICFP
2020 paper, including the free theorems arising from the usage
annotations.

Reviewer C
----------

> Basically: Expand section 3, starting with a non-intrinsic version
> of Figure 1; explain how to get from the non-intrinsic system to
> Figure 1; explain how to get to Figure 2.

Good idea. We agree that Section 3 is the key section to understanding
our approach. Our plan is to expand the discussion of the intrinsic
presentation in Figure 1 by comparing them to the extrinsic
presentations directly. Hopefully this will lead the reader more
gently to the final presentation in Fig 2. As suggested, comparing our
approach to HOAS's use of ambient contexts is also a useful way to
help readers' intuitions.

> missing some work that may be related: ambient contexts are used in
> higher-order abstract syntax (as in LF/Twelf and its successors)

We will add a discussion of the use of ambient contexts in HOAS
presentations. The interesting point there, we think, is that to
easily encode substructural systems, the common approach is to
replicate the substructural discipline in the logical framework (e.g.,
Linear LF). Here, we are exploiting the same idea internally in Agda
by using the bunched connectives, which means that we don't need to
switch to a completely new logical framework. Thank you for the prompt
to discuss this related work.

> In Definition 2, in the first line, I take it that → is implication,

Yes, this is implication. We will write this out using English words
instead.

> Is the "Rγ" notation chosen for a reason, compared to something like
> R; γ?

We felt that this notation was more concise. Figure 1 with $(R; \gamma)$
everywhere would be much messier, in our opinion.

### 2022-03-18 tex/examples.tex
[ backport ] examples from ESOP22 to thesis

@@
+In this section, we provide three example uses of semantic traversals: generic
+renaming and substitution, a usage elaborator, and a denotational semantics.
+The reader is also encouraged to see the far greater range of examples in the
+work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
+Renaming and substitution are essential results, while the latter two examples
+focus on usage annotations.
+
+A result we will use throughout this section is \emph{reification}.
+When we have an index-preserving mapping from usage-checked variables to
+$\V$-environments, we can construct environments of the form
+$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
+This lets us write the \AgdaFunction{reify} function, which  simplifies our
+obligations in giving a \AgdaRecord{Semantics} by coercing
+\AgdaFunction{Kripke} functions into just
+\AgdaBound{$\C$}-values in an extended context.
+
+%To show this, we instantiate the Kripke function with the renaming
+%$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
+%pass it an argument environment
+%$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
+%extended part.
+%Post-composition of a renaming onto an environment comes from
+%\cref{thm:env-postren}.
+
+%\ExecuteMetaData[\Syntactictex]{reify}
+
+\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
+  If $\V$ is an open family such that there is a function
+  $v : \forallb{{\sqni} \dotto \V}$, we get a function of type
+  $\forallb{\mathrm{Kripke}\,\V\,\C \dotto \mathrm{Scope}\,\C}$ for any $\C$.
+\end{lemma}
+\begin{proof}
+  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
+  That is, $b$ is a Kripke function yielding $\C$-computations
+  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
+  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
+  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
+  $\Gamma, \gr0\delta$.
+  Then we give the $\wand$-function an argument in context
+  $\gr0\gamma, \Delta$, noting that
+  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
+  as we wanted from the result.
+  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
+  We produce this via \cref{thm:env-postren} from an environment
+  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
+  and a renaming which is the complement to that used on $\Box^r$.
+\end{proof}
+
+All of the $\V$s used in examples in this paper support identity environments.
+However, \citet[p.~27]{AACMM21} give some important examples that do not
+support identity environments, and thus cannot use \AgdaFunction{reify}
+(\cref{thm:reify}).
+The feature that causes the lack of support for identity environments is that
+a semantics can make use of the fact that only variables of particular kinds
+are bound by the syntax.
+In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
+binds variables that synthesise their type, as opposed to those whose type is
+checked.
+The semantics of type-checking and elaboration rely on variables synthesising
+their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
+Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
+only binds such synthesising variables.
+Similar phenomena would appear in, say, a call-by-value language where
+variables are values (not computations), or a polarised language where
+variables always have a polarity matching their type.
+
 \subsection{Renaming and substitution}\label{sec:kits}
@@
 
-\subsection{A denotational semantics}
-
-To justify the name \emph{semantics}, we give an example traversal that is a
-denotational semantics in the usual sense.
-The semantics we take is a refinement of that of \citet{AbelBernardy2020},
-which gives a way to extract parametricity theorems from substructurally typed
-programs.
-Example theorems are that all linear terms act as permutations on some fixed
-set of resources, and all monotonically typed terms are really monotonic in the
-way the typing suggests they are.
-
-To abbreviate this section, we use a simplified syntax compared to \name{}.
-We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
-type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
-\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.
-
-\ExecuteMetaData[\WReltex]{Ty}
-
-In the term syntax, $\lambda$-abstraction now binds a variable with annotation
-\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
-(both in accordance with the function type they are acting on).
-
-\ExecuteMetaData[\WReltex]{AnnArr}
-
-As a running example, we take the usage annotations to be the 4-element
-variance posemiring (\cref{thm:variance}).
-We establish the property that all terms are monotonic in their free variables.
-This monotonicity can be covariant or contravariant (or neither or both)
-depending on the annotation of each free variable.
-This provides an additional example to those of \citeauthor{AbelBernardy2020}.
-
-We will take semantics of this system into
-\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
-A world-indexed relation over a poset of worlds \AgdaBound{W} is a set over
-which
-we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
-property with respect to the order on \AgdaBound{W}.
-
-\ExecuteMetaData[\WReltex]{WRel}
-
-\begin{example}
-  When \AgdaBound{W} is the 1-element set, a world-indexed relation is just a
-  set equipped with a binary relation.
-\end{example}
-
-Morphisms between world-indexed relations \AgdaBound{R} and \AgdaBound{S}
-consist of a mapping between the underlying sets such that that mapping
-preserves relatedness from \AgdaBound{R} to \AgdaBound{S}.
-
-\ExecuteMetaData[\WReltex]{WRelMor}
-
-\todo{Define big intersection.}
-When the poset of worlds forms a (relational) commutative monoid, such
-world-indexed relations support a symmetric monoidal closed structure, with
-objects denoted \AgdaFunction{I$^R$},
-\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
-\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
-These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
-\AgdaRecord{$\wand$}, now over worlds rather than contexts.
-
-%\ExecuteMetaData[\WReltex]{IR}
-%\ExecuteMetaData[\WReltex]{tensorR}
-%\ExecuteMetaData[\WReltex]{lollyR}
-
-The final piece of sematics we need is a \emph{bang} operator.
-We allow the
-semantic \emph{bang} to be an arbitrary annotation-indexed functor on
-world-indexed relations.
-This functor must respect all of the structure on the indices, making it a
-graded comonad over multiplication, as well as being lax monoidal at any
-particular index \AgdaBound{r}.
-
-\ExecuteMetaData[\WReltex]{Bang}
-
-\begin{example}
-  With \AgdaBound{W} being the 1-element set and annotations coming from the
-  variance semiring, we can define the following \emph{bang}.
-  It is always the identity on the set component, while the relation component
-  consists of flipping the relation for contravariance and taking conjunctions
-  to achieve both covariance and contravariance.
-  When we want neither covariance nor contravariance, we use the always true
-  predicate on worlds \AgdaFunction{$\dot1$}.
-
-  \ExecuteMetaData[\Monotonicitytex]{BangR}
-\end{example}
-
-To associate semantics to syntax, we start as standard by associating
-world-indexed relations to types.
-We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
-\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
-context, concatenation, and usage annotations on singletons, respectively.
-
-\ExecuteMetaData[\WReltex]{sem}
-
-The semantics of a term is then to be a morphism from the interpretation of the
-context to the interpretation of the term's type.
-
-\ExecuteMetaData[\WReltex]{sem-vdash}
-
-Variables are given semantics by \AgdaFunction{lookup$^R$} (definition omitted).
-
-\ExecuteMetaData[\WReltex]{lookupR-type}
-
-Now, we give a \AgdaRecord{Semantics}.
-The choice of \AgdaBound{$\V$} as
-\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
-given that a standard denotational semantics would not use intermediate
-environments in the same sense as renaming and substitution, but allows us to
-reuse the standard facts that variables support renaming and identity
-environments.
-With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
-environment entries by \AgdaFunction{lookup$^R$}.
-Meanwhile, for the logical rules, we ignore environments by using
-\AgdaFunction{reify} to just deal with morphisms in an extended context.
-As such, $\lambda$-abstractions are easy to interpret, while applications
-require some massaging to remove the extension by an empty context, followed by
-some plumbing to split the interpretation of the context according to the usage
-constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
-the interpretation of the function \AgdaBound{m$'$}.
-
-\ExecuteMetaData[\WReltex]{Wrel}
-
-In order to map open terms to interpretations, we take the action of the
-semantics and give the identity renaming as the starting environment.
-
-\ExecuteMetaData[\WReltex]{wrel}
-
-\begin{example}
-  We can make a subtraction function from primitive addition and negation on
-  integers.
-  Subtraction is covariant in its first argument and contravariant in its
-  second argument.
-  We give the definition in pseudocode, as we have not yet seen how to
-  conveniently write terms (\cref{sec:usage-elaborator}).
-
-  \begin{align*}
-    &{\sim\sim}p :
-      {\uparrow\uparrow}\mathbb Z \multimap
-      {\uparrow\uparrow}\mathbb Z \multimap \mathbb Z,
-      {\sim\sim}n : {\downarrow\downarrow}\mathbb Z \multimap \mathbb Z
-      \vdash \mathnormal{minus} :
-      {\uparrow\uparrow}\mathbb Z \multimap
-      {\downarrow\downarrow}\mathbb Z \multimap
-      \mathbb Z
-    \\
-    &\mathnormal{minus} \coloneqq \lambda x.~\lambda y.~p\,x\,(n\,y)
-  \end{align*}
-
-  After feeding in Agda's addition and negation functions as the
-  interpretations of the free variables (noting that they are both monotonic
-  in the required way), we get the following free theorem.
-
-  \ExecuteMetaData[\Monotonicitytex]{thm-type}
-
-  %We observe that the set component of this term's semantics is just the
-  %expected Agda function when the two free variables are given appropriate
-  %interpretations.
-
-  %\ExecuteMetaData[\Monotonicitytex]{minus-set}
-
-  %Furthermore, the relational component of the semantics yields the free
-  %theorem that the Agda subtraction so defined is monotonic in the expected way.
-  %This relies on library proofs that addition and negation are suitably
-  %monotonic.
-
-  %\ExecuteMetaData[\Monotonicitytex]{thm}
-\end{example}
-
 \subsection{A usage elaborator}\label{sec:usage-elaborator}
@@
 extremely tedious.
-We achieved some abbreviation by using pattern synonyms, but we still have to
+We can achieve some abbreviation by using pattern synonyms to wrap around
+\AgdaInductiveConstructor{`con} expressions, but we still have to
 produce essentially bespoke proofs whenever we use a usage-sensitive part of the
@@
 \emph{renaming} in this case, because $\V$ is a family of variables).
-\todo{Previous examples are now deleted.}
-In the previous examples, we have used \AgdaFunction{elab-unique}, which further
+We use \AgdaFunction{elab-unique}, which further
 checks statically that exactly one derivation is returned.
-The candidate usage vector \AgdaBound{R} will be empty for closed terms, and
-otherwise we have to supply the intended usage annotations.
+The candidate usage vector \AgdaBound{R} will be \AgdaFunction{[]} for closed
+terms, and otherwise we have to supply the intended usage annotations.
+
+We can now use the elaborator to automatically infer the usage
+annotations for the example at the end of \cref{sec:terms}. This
+allows us to write:
+\ExecuteMetaData[\PaperExamplestex]{cojoin}
+We have instantiated the usage elaborator so that:
+\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
+empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
+$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
+\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
+$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
+gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
+[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
+$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
+consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
+this splitting would introduce more non-determinism but not allow any
+more terms to be typed. As such, the only non-determinism comes when
+we have variables annotated $\gr1$ and need to do an additive split,
+like when we apply the \AgdaInductiveConstructor{!E} rule below. At
+this point, the variable can become either $\gr0$-annotated in the
+left subterm and $\gr1$-annotated on the right, or vice-versa. We will
+find that, because the left subterm wants to use that variable, the
+former choice will be rejected. The function \AgdaFunction{var\#} is a
+convenience for converting statically known natural numbers,
+representing de Bruijn \emph{levels}, into variable terms.
+
+\subsection{A denotational semantics}
+
+To justify the name \emph{semantics}, we give an example traversal that is a
+denotational semantics in the usual sense.
+The semantics we take is a refinement of that of \citet{AbelBernardy2020},
+which gives a way to extract parametricity theorems from substructurally typed
+programs.
+Example theorems are that all linear terms act as permutations on some fixed
+set of resources, and all monotonically typed terms are really monotonic in the
+way the typing suggests they are.
+
+To abbreviate this section, we use a simplified syntax compared to \name{}.
+We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
+type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
+\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.
+
+\ExecuteMetaData[\WReltex]{Ty}
+
+In the term syntax, $\lambda$-abstraction now binds a variable with annotation
+\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
+(both in accordance with the function type they are acting on).
+
+\ExecuteMetaData[\WReltex]{AnnArr}
+
+As a running example, we take the usage annotations to be the 4-element
+variance posemiring (\cref{thm:variance}).
+We establish the property that all terms are monotonic in their free variables.
+This monotonicity can be covariant or contravariant (or neither or both)
+depending on the annotation of each free variable.
+This provides an additional example to those of \citeauthor{AbelBernardy2020}.
+
+We will take semantics of this system into
+\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
+A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
+\AgdaBound{W} is a set over which
+we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
+property with respect to the order on \AgdaBound{W}.
+
+\ExecuteMetaData[\WReltex]{WRel}
+
+\begin{example}
+  When \AgdaBound{W} is the 1-element set, a world-indexed relation is just a
+  set equipped with a binary relation.
+\end{example}
+
+Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
+and \AgdaBound{S} consist of a mapping between the underlying sets such that, at
+each fixed world \AgdaBound{w}, the mapping preserves relatedness from
+\AgdaBound{R} to \AgdaBound{S}.
+
+\ExecuteMetaData[\WReltex]{WRelMor}
+
+\todo{Define big intersection.}
+When the poset of worlds forms a (relational) commutative monoid, such
+world-indexed relations support a symmetric monoidal closed structure, with
+objects denoted \AgdaFunction{I$^R$},
+\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
+\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
+These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
+\AgdaRecord{$\wand$}, now over worlds rather than contexts.
+
+The final piece of semantics we need is a \emph{bang} operator.
+We allow the
+semantic \emph{bang} to be an arbitrary annotation-indexed functor on
+world-indexed relations.
+This functor must respect all of the structure on the indices, making it a
+graded comonad over multiplication, as well as being lax monoidal at any
+particular index \AgdaBound{r}.
+
+\ExecuteMetaData[\WReltex]{Bang}
+
+\begin{example}
+  With \AgdaBound{W} being the 1-element set and annotations coming from the
+  variance semiring, we can define the following \emph{bang}.
+  It is always the identity on the set component, while the relation component
+  consists of flipping the relation for contravariance and taking conjunctions
+  to achieve both covariance and contravariance.
+  When we want neither covariance nor contravariance, we use the always true
+  predicate on worlds \AgdaFunction{$\dot1$}.
+
+  \ExecuteMetaData[\Monotonicitytex]{BangR}
+\end{example}
+
+%To associate semantics to syntax, we start as standard by associating
+%world-indexed relations to types.
+%We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
+%\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
+%context, concatenation, and usage annotations on singletons, respectively.
+%
+%\ExecuteMetaData[\WReltex]{sem}
+%
+%The semantics of a term is then to be a morphism from the interpretation of the
+%context to the interpretation of the term's type.
+%
+%\ExecuteMetaData[\WReltex]{sem-vdash}
+
+The semantics of a type is given by \AgdaFunction{$\llbracket$\_$\rrbracket$},
+which maps into world-indexed relations.
+The function type is interpreted using \AgdaFunction{$\multimap^R$} and
+\AgdaField{!$^R$}.
+Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
+\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}.
+Terms are interpreted as morphisms by the open family
+\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
+Variables are interpreted by \AgdaFunction{lookup$^R$} (definition omitted).
+
+\ExecuteMetaData[\WReltex]{lookupR-type}
+
+Now we give a \AgdaRecord{Semantics}.
+The choice of \AgdaBound{$\V$} as
+\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
+given that a standard denotational semantics would not use intermediate
+environments in the same sense as renaming and substitution, but it allows us to
+reuse the standard facts that variables support renaming and identity
+environments.
+With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
+environment entries by \AgdaFunction{lookup$^R$}.
+Meanwhile, for the logical rules, we ignore environments by using
+\AgdaFunction{reify} to just deal with morphisms in an extended context.
+As such, $\lambda$-abstractions are easy to interpret, while applications
+require some massaging to remove the extension by an empty context, followed by
+some plumbing to split the interpretation of the context according to the usage
+constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
+the interpretation of the function \AgdaBound{m$'$}.
+
+\ExecuteMetaData[\WReltex]{Wrel}
+
+Then, the semantics of terms is given by the function
+\AgdaFunction{semantics}\AgdaSpace{}\AgdaFunction{Wrel}\AgdaSpace{}%
+\AgdaFunction{1$^r$}, where \AgdaFunction{1$^r$} is the identity renaming.
+%In order to map open terms to interpretations, we take the action of the
+%semantics and give the identity renaming as the starting environment.
+
+\ExecuteMetaData[\WReltex]{wrel}
+
+\begin{example}\label{thm:minus}
+  We can make a subtraction function from primitive addition and negation on
+  integers.
+  Subtraction is covariant in its first argument and contravariant in its
+  second argument.
+  We give the definition in pseudocode, though it is also amenable to the
+  usage elaborator of \cref{sec:usage-elaborator}, suitably instantiated.
+
+  \begin{align*}
+    &{\sim\sim}p :
+      {\uparrow\uparrow}\mathbb Z \multimap
+      {\uparrow\uparrow}\mathbb Z \multimap \mathbb Z,
+      {\sim\sim}n : {\downarrow\downarrow}\mathbb Z \multimap \mathbb Z
+      \vdash \mathnormal{minus} :
+      {\uparrow\uparrow}\mathbb Z \multimap
+      {\downarrow\downarrow}\mathbb Z \multimap
+      \mathbb Z
+    \\
+    &\mathnormal{minus} \coloneqq \lambda x.~\lambda y.~p\,x\,(n\,y)
+  \end{align*}
+
+  After feeding in Agda's addition and negation functions as the
+  interpretations of the free variables (noting that they are both monotonic
+  in the required way), we get the following free theorem.
+
+  \ExecuteMetaData[\Monotonicitytex]{thm-type}
+
+  %We observe that the set component of this term's semantics is just the
+  %expected Agda function when the two free variables are given appropriate
+  %interpretations.
+
+  %\ExecuteMetaData[\Monotonicitytex]{minus-set}
+
+  %Furthermore, the relational component of the semantics yields the free
+  %theorem that the Agda subtraction so defined is monotonic in the expected way.
+  %This relies on library proofs that addition and negation are suitably
+  %monotonic.
+
+  %\ExecuteMetaData[\Monotonicitytex]{thm}
+\end{example}

### 2022-03-18 tex/syntax.tex
[ backport ] syntax from ESOP22 to thesis

@@
 system of our framework. We use the prototypical system from
-\cref{fig:lr-comb} as a running example. \cref{sec:other-syntaxes}
+\cref{fig:lr-comb} as a running example. \Cref{sec:other-syntaxes}
 presents further examples that our framework can express.
@@
 
-\paragraph{\cref{fig:lr-comb} as a \AgdaRecord{System}.}
+% \paragraph{\Cref{fig:lr-comb} as a \AgdaRecord{System}.}
 
@@
 
-\paragraph{Terms of a \AgdaRecord{System}.}
+\subsection{Terms of a \AgdaRecord{System}}\label{sec:terms}
 
@@
 
-With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
-the rest are simply interpretations of premises into bunched combinators.
+The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
+are the interpretation into bunched combinators.
+The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
+they have been lifted from predicates on usage vectors to predicates on
+contexts, with the type component of the context shared throughout.
+Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
+already polymorphic (not relying on anything specific about usage vectors), so
+do not need a \AgdaFunction{$^c$} variant.
 
@@
 variable or a use of a logical rule together with terms for each of
-the required subterms. The \AgdaFunction{Size} argument is where we
-use sized types to convince Agda that this type is strictly positive.
+the required subterms. The \AgdaFunction{Size} argument is a use of
+Agda's sized types to record that subterms are smaller than the
+surrounding term for the termination checker.
 
@@
 
-Terms defined like this are still quite difficult to write, mainly because of
-frequently changing usage contexts and the need for proofs that they all match
-up.
-We will see how to automate these proofs in \cref{sec:usage-elaborator}.
+This definition uses \AgdaFunction{$\dotto$}, which, analogously to
+\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
+space.
+We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
+case two (the context and the type).
+The notation
+\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
+stands for
+\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
+\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
+\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
+\AgdaSpace{}%
+\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
+where \AgdaBound{T} is a type family with $n$ many indices.
+
+Terms in this data type are difficult to write by hand, due to the
+need for proofs that the usage contexts are handled correctly. For
+example, the following term is needed to show that, in the $\{\gr0,
+\gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
+$\oc\gr\omega$ forms a comonad.
+Pattern synonyms \AgdaInductiveConstructor{$\multimap$I},
+\AgdaInductiveConstructor{!E$'$}, and
+\AgdaInductiveConstructor{!I$'$} stand for applications of
+\AgdaInductiveConstructor{`con}, with the latter two taking explicit usage
+contexts and proofs.
+On concrete
+posemirings (as in this example), unification is particularly poor at
+inferring the usage contexts from the proofs because addition and
+multiplication are no longer (judgementally) injective.
+The function \AgdaFunction{var\#} is a way of turning a statically known de
+Bruijn level and a usage proof into an application of \AgdaInductiveConstructor{`var}.
+
+\ExecuteMetaData[\HeavyItex]{cojoin-explicit}
+
+ Writing terms like this
+is clearly unsustainable. We will see a way of automating the
+necessary proofs via a \AgdaRecord{System}-generic elaborator in
+\cref{sec:usage-elaborator}.
 

### 2022-05-27 tex/ill.tex
[ more ] thesis writing

@@
+Linear logic is a good starting point for our investigations into usage
+restrictions because it is well known, well understood, and exhibits most of the
+difficulties we will come across.
+Linear logic's lack of general weakening and contraction will be a typical
+feature of the calculi we will see, and the exponential modality $\oc$ (bang)
+will be the prototype for a range of modalities I will consider.
+
 \subsection{Motivation of linearity}
 
-In LJ, two distinct, but equiderivable, canonical definitions of $\wedge$ can
-be given.
-
-In the first approach, $\wedge^-$ is conceived to be a
-\emph{categorial product}.
-Such a product is constructed by providing two maps in from the same antecedents
-$\Gamma$.
-It is used by projecting out one of the sides.
-
-\begin{mathpar}
-  \inferrule*[right=$\wedge^-$-r]
-  {\Gamma \vdash A \\ \Gamma \vdash B}
-  {\Gamma \vdash A \wedge^- B}
-
-  \and
-
-  \inferrule*[right=$\wedge^-$-l$_0$]
-  {\Gamma, A \vdash C}
-  {\Gamma, A \wedge^- B \vdash C}
-
-  \and
-
-  \inferrule*[right=$\wedge^-$-l$_1$]
-  {\Gamma, B \vdash C}
-  {\Gamma, A \wedge^- B \vdash C}
-\end{mathpar}
-
-In the second approach, $\wedge^+$ is conceived to be a \emph{tensor product}.
-Intuitively, $\wedge^+$ internalises the comma of context concatenation.
-Such a product is used by giving access to both sides simultaneously.
-It is constructed by constructing both sides and combining the antecedents
-required by each side.
-
-\begin{mathpar}
-  \inferrule*[right=$\wedge^+$-r]
-  {\Gamma \vdash A \\ \Delta \vdash B}
-  {\Gamma, \Delta \vdash A \wedge^+ B}
-
-  \and
-
-  \inferrule*[right=$\wedge^+$-l]
-  {\Gamma, A, B \vdash C}
-  {\Gamma, A \wedge^+ B \vdash C}
-\end{mathpar}
-
-When we prove the logical equivalence of these two formulations, we notice that
-the structural rules of weakening and contraction are essential.
-When we remove weakening and contraction, $\wedge^-$ and $\wedge^+$ become
-logically distinct connectives, which we notate $\&$ and $\otimes$,
-respectively.
-
-\begin{mathpar}
-  \inferrule*[right=$\wedge^-$-r]
-  {%
-    \inferrule*[right=Weak$^*$,fraction={\cdot\cdots\cdot}]
-    {\Gamma \vdash A}
-    {\Gamma, \Delta \vdash A}
-    \\
-    \inferrule*[Right=Weak$^*$,fraction={\cdot\cdots\cdot}]
-    {\Delta \vdash B}
-    {\Gamma, \Delta \vdash B}
-  }
-  {\Gamma, \Delta \vdash A \wedge^- B}
-
-  \and
-
-  \inferrule*[right=Contr]
-  {%
-    \inferrule*[Right=$\wedge^-$-l$_0$]
-    {%
-      \inferrule*[Right=$\wedge^-$-l$_1$]
-      {\Gamma, A, B \vdash C}
-      {\Gamma, A, A \wedge^- B \vdash C}
-    }
-    {\Gamma, A \wedge^- B, A \wedge^- B \vdash C}
-  }
-  {\Gamma, A \wedge^- B \vdash C}
-\end{mathpar}
-
-\begin{mathpar}
-  \inferrule*[right=Contr$^*$,fraction={\cdot\cdots\cdot}]
-  {%
-    \mprset{defaultfraction}
-    \inferrule*[Right=$\wedge^+$-r]
-    {\Gamma \vdash A \\ \Gamma \vdash B}
-    {\Gamma, \Gamma \vdash A \wedge^+ B}
-  }
-  {\Gamma \vdash A \wedge^+ B}
-
-  \and
-
-  \inferrule*[right=$\wedge^+$-l]
-  {%
-    \inferrule*[Right=Weak]
-    {\Gamma, A \vdash C}
-    {\Gamma, A, B \vdash C}
-  }
-  {\Gamma, A \wedge^+ B \vdash C}
-
-  \and
-
-  \inferrule*[right=$\wedge^+$-l]
-  {%
-    \inferrule*[Right=Weak]
-    {\Gamma, B \vdash C}
-    {\Gamma, A, B \vdash C}
-  }
-  {\Gamma, A \wedge^+ B \vdash C}
-\end{mathpar}
+%In LJ, two distinct, but equiderivable, canonical definitions of $\wedge$ can
+%be given.
+%
+%In the first approach, $\wedge^-$ is conceived to be a
+%\emph{categorial product}.
+%Such a product is constructed by providing two maps in from the same antecedents
+%$\Gamma$.
+%It is used by projecting out one of the sides.
+%
+%\begin{mathpar}
+%  \inferrule*[right=$\wedge^-$-r]
+%  {\Gamma \vdash A \\ \Gamma \vdash B}
+%  {\Gamma \vdash A \wedge^- B}
+%
+%  \and
+%
+%  \inferrule*[right=$\wedge^-$-l$_0$]
+%  {\Gamma, A \vdash C}
+%  {\Gamma, A \wedge^- B \vdash C}
+%
+%  \and
+%
+%  \inferrule*[right=$\wedge^-$-l$_1$]
+%  {\Gamma, B \vdash C}
+%  {\Gamma, A \wedge^- B \vdash C}
+%\end{mathpar}
+%
+%In the second approach, $\wedge^+$ is conceived to be a \emph{tensor product}.
+%Intuitively, $\wedge^+$ internalises the comma of context concatenation.
+%Such a product is used by giving access to both sides simultaneously.
+%It is constructed by constructing both sides and combining the antecedents
+%required by each side.
+%
+%\begin{mathpar}
+%  \inferrule*[right=$\wedge^+$-r]
+%  {\Gamma \vdash A \\ \Delta \vdash B}
+%  {\Gamma, \Delta \vdash A \wedge^+ B}
+%
+%  \and
+%
+%  \inferrule*[right=$\wedge^+$-l]
+%  {\Gamma, A, B \vdash C}
+%  {\Gamma, A \wedge^+ B \vdash C}
+%\end{mathpar}
+%
+%When we prove the logical equivalence of these two formulations, we notice that
+%the structural rules of weakening and contraction are essential.
+%When we remove weakening and contraction, $\wedge^-$ and $\wedge^+$ become
+%logically distinct connectives, which we notate $\&$ and $\otimes$,
+%respectively.
+%
+%\begin{mathpar}
+%  \inferrule*[right=$\wedge^-$-r]
+%  {%
+%    \inferrule*[right=Weak$^*$,fraction={\cdot\cdots\cdot}]
+%    {\Gamma \vdash A}
+%    {\Gamma, \Delta \vdash A}
+%    \\
+%    \inferrule*[Right=Weak$^*$,fraction={\cdot\cdots\cdot}]
+%    {\Delta \vdash B}
+%    {\Gamma, \Delta \vdash B}
+%  }
+%  {\Gamma, \Delta \vdash A \wedge^- B}
+%
+%  \and
+%
+%  \inferrule*[right=Contr]
+%  {%
+%    \inferrule*[Right=$\wedge^-$-l$_0$]
+%    {%
+%      \inferrule*[Right=$\wedge^-$-l$_1$]
+%      {\Gamma, A, B \vdash C}
+%      {\Gamma, A, A \wedge^- B \vdash C}
+%    }
+%    {\Gamma, A \wedge^- B, A \wedge^- B \vdash C}
+%  }
+%  {\Gamma, A \wedge^- B \vdash C}
+%\end{mathpar}
+%
+%\begin{mathpar}
+%  \inferrule*[right=Contr$^*$,fraction={\cdot\cdots\cdot}]
+%  {%
+%    \mprset{defaultfraction}
+%    \inferrule*[Right=$\wedge^+$-r]
+%    {\Gamma \vdash A \\ \Gamma \vdash B}
+%    {\Gamma, \Gamma \vdash A \wedge^+ B}
+%  }
+%  {\Gamma \vdash A \wedge^+ B}
+%
+%  \and
+%
+%  \inferrule*[right=$\wedge^+$-l]
+%  {%
+%    \inferrule*[Right=Weak]
+%    {\Gamma, A \vdash C}
+%    {\Gamma, A, B \vdash C}
+%  }
+%  {\Gamma, A \wedge^+ B \vdash C}
+%
+%  \and
+%
+%  \inferrule*[right=$\wedge^+$-l]
+%  {%
+%    \inferrule*[Right=Weak]
+%    {\Gamma, B \vdash C}
+%    {\Gamma, A, B \vdash C}
+%  }
+%  {\Gamma, A \wedge^+ B \vdash C}
+%\end{mathpar}
 
@@
 \begin{mathpar}
-  \inferrule*[right=Promotion]
-  {\oc\Gamma \vdash A}
-  {\oc\Gamma \vdash \oc A}
+  \ebrule{%
+    \hypo{\oc\Gamma \vdash A}
+    \infer1[Promotion]{\oc\Gamma \vdash \oc A}
+  }
 
@@
 
-  \inferrule*[right=Dereliction]
-  {\Gamma, A \vdash B}
-  {\Gamma, \oc A \vdash B}
+  \ebrule{%
+    \hypo{\Gamma, A \vdash B}
+    \infer1[Dereliction]{\Gamma, \oc A \vdash B}
+  }
 
@@
 
-  \inferrule*[right=Weakening]
-  {\Gamma \vdash B}
-  {\Gamma, \oc A \vdash B}
+  \ebrule{%
+    \hypo{\Gamma \vdash B}
+    \infer1[Weakening]{\Gamma, \oc A \vdash B}
+  }
 
@@
 
-  \inferrule*[right=Contraction]
-  {\Gamma, \oc A, \oc A \vdash B}
-  {\Gamma, \oc A \vdash B}
+  \ebrule{%
+    \hypo{\Gamma, \oc A, \oc A \vdash B}
+    \infer1[Contraction]{\Gamma, \oc A \vdash B}
+  }
 \end{mathpar}
@@
 \begin{mathpar}
-  \inferrule*[right=$\otimes$-l]
-  {%
-    \inferrule*[Right=$\otimes$-r]
-    {%
-      \inferrule*[right=Id]{ }{P \vdash P} \\
-      \inferrule*[Right=Id]{ }{Q \vdash Q}
+  \ebrule{%
+    \infer0[Id]{P \vdash P}
+    \infer0[Id]{Q \vdash Q}
+    \infer2[$\otimes$-r]{P, Q \vdash P \otimes Q}
+    \infer1[$\otimes$-l]{P \otimes Q \vdash P \otimes Q}
   }
-    {P, Q \vdash P \otimes Q}
-  }
-  {P \otimes Q \vdash P \otimes Q}
 
@@
 
-  \inferrule*[right=Pro]
-  {%
-    \inferrule*[Right=Der]
-    {%
-      \inferrule*[Right=Id]{ }{P \vdash P}
+  \ebrule{%
+    \infer0[Id]{P \vdash P}
+    \infer1[Der]{\oc P \vdash P}
+    \infer1[Pro]{\oc P \vdash \oc P}
   }
-    {\oc P \vdash P}
-  }
-  {\oc P \vdash \oc P}
 \end{mathpar}

### 2022-05-27 tex/linmech.tex
[ more ] thesis writing

In this section, I give an overview of techniques which have been used in
previous work to encode linear logic in proof assistants.
To my knowledge, there have been no other significant attempts to mechanise
semiring-annotated logics.

\subsection{Typing with leftovers}

Typing with leftovers~\cite{allais17} is a technique developed to capture a
straightforward approach to linearity checking.
The idea is to, at any point, consider an input context, a term, and an output
context.
The input context contains all of the variables in scope, while the output
context is the same minus any variables used by the term.
Checking of adjacent terms, for example in a $\otimes$-introduction, is done by
threading the context through from the output of the first term to the input of
the second.
Bound variables are introduced in the input context of the term in which they
are bound, and are expected to be absent in the output context of that term.
\Cref{fig:twl,fig:twl-mult} give rules in this style for the multiplicative
fragment of intuitionistic linear logic.

\begin{figure}
  \begin{align*}
    \Gamma, \Delta, \Theta
    &\Coloneqq {\cdot} \mid \Gamma, \gr1x : A \mid \Gamma, \gr0x : A \\
    \mathcal{S} &\Coloneqq \Gamma \vdash M : A \boxtimes \Delta
  \end{align*}
  \caption{Typing with leftovers, context and sequent syntax}
  \label{fig:twl}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \infer0[Var]{\Gamma, \gr1x : A \vdash x : A \boxtimes \Gamma, \gr0x : A}
    }
    \and
    \ebrule{%
      \infer0[$I$-I]{\Gamma \vdash {*} : I \boxtimes \Gamma}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : I \boxtimes \Delta}
      \hypo{\Delta \vdash N : C \boxtimes \Theta}
      \infer2[$I$-E]{\Gamma \vdash \text{let }{*} = M\text{ in }N : C
        \boxtimes \Theta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : A \boxtimes \Delta}
      \hypo{\Delta \vdash N : B \boxtimes \Theta}
      \infer2[$\otimes$-I]{\Gamma \vdash (M, N) : A \otimes B \boxtimes \Theta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : A \otimes B \boxtimes \Delta}
      \hypo{\Delta, \gr1x : A, \gr1y : B \vdash N : C \boxtimes
        \Theta, \gr0x : A, \gr0y : B}
      \infer2[$\otimes$-E]{\Gamma \vdash \text{let }(x, y) = M\text{ in }N : C
        \boxtimes \Theta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma, \gr1x : A \vdash M : B \boxtimes \Delta, \gr0x : A}
      \infer1[$\multimap$-I]{\Gamma \vdash \lambda x.~M : A \multimap B
        \boxtimes \Delta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : A \multimap B \boxtimes \Delta}
      \hypo{\Delta \vdash N : A \boxtimes \Theta}
      \infer2[$\multimap$-E]{\Gamma \vdash M\,N : B \boxtimes \Theta}
    }
  \end{mathpar}
  \caption{Typing with leftovers, multiplicative fragment}
  \label{fig:twl-mult}
\end{figure}

The original paper extends the logic covered to binary additives --- $\with$ and
$\oplus$ --- with rules that check that terms agree on output contexts, as seen
in \cref{fig:twl-add}.
However, it is less clear how to handle nullary additives --- $\top$ and $0$ ---
as we would have 0 (rather than 2) candidates for the output context.
At some level, this problem is unavoidable in a system modelling linearity
checking because any checking strategy will expose the ambiguity in sequents
like $\gr1x : A \vdash (\{\}, \{\}) : \top \otimes \top$ of whether the variable
$x$ was used in the left half or the right half.
As such, we may expect uses of $\top$-introduction (and similarly
$0$-elimination) to be annotated with which variables they use, in which case we
get a viable typing with leftovers rule.

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma \vdash M : A \boxtimes \Delta}
      \hypo{\Gamma \vdash N : B \boxtimes \Delta}
      \infer2[$\with$-I]{\Gamma \vdash \{M,N\} : A \with B \boxtimes \Delta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : A \oplus B \boxtimes \Delta}
      \hypo{\Delta, \gr1x : A \vdash N : C \boxtimes \Theta, \gr0x : A}
      \hypo{\Delta, \gr1y : B \vdash O : C \boxtimes \Theta, \gr0y : B}
      \infer3[$\oplus$-E]{\Gamma \vdash
        \text{case }M\text{ of }\{x.~N; y.~O\} : C \boxtimes \Theta}
    }
  \end{mathpar}
  \caption{Typing with leftovers, a selection of the additive rules}
  \label{fig:twl-add}
\end{figure}

The original paper also does not show how to capture the exponential modality
$\oc$.
We could perhaps, inspired by DILL, distinguish between linear variables and
intuitionistic variables.
This gives rules like those of \cref{fig:twl-exp}.
The important invariant is that linear and intuitionistic variables stay
distinct, so any intuitionistic variable in the input context (annotated by
$\gr\omega$) must be intuitionistic in the output context.
In the $\oc$-introduction rule, there are several choices we could make, but I
have chosen to keep all the linear variables in scope but used so as to match
the general style of variables staying in scope with the $\gr0$ annotation..

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \infer0[IVar]{\Gamma, \gr\omega x : A \vdash x : A
        \boxtimes \Gamma, \gr\omega x : A}
    }
    \and
    \ebrule{%
      \hypo{\gr0\gamma, \gr0\delta, \gr\omega\theta \vdash
        M : A \boxtimes \gr0\gamma, \gr0\delta, \gr\omega\theta}
      \infer1[$\oc$-I]{\gr0\gamma, \gr1\delta, \gr\omega\theta \vdash
        [M] : \oc A \boxtimes \gr0\gamma, \gr1\delta, \gr\omega\theta}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash M : \oc A \boxtimes \Delta}
      \hypo{\Delta, \gr\omega x : A \vdash N : C
        \boxtimes \Theta, \gr\omega x : A}
      \infer2[$\oc$-E]{\Gamma \vdash
        \text{let }[x] = M\text{ in }N : C \boxtimes \Theta}
    }
  \end{mathpar}
  \caption{Typing with leftovers, a possible way to capture $\oc$}
  \label{fig:twl-exp}
\end{figure}

However, this adaptation of the DILL style does not obviously generalise to
semiring annotations.
Even for the multiplicative fragment, we seem to be working against the
direction of addition, instead using a subtraction operation whenever we use a
variable.
For exponentials, though, and particularly the $\oc$-introduction rule, what I
have done seems ad-hoc, not based on any pointwise algebraic operation.

Also, the unusual form of sequents can cause some problems when working with a
typing with leftovers system.
For example, a traditional intuitionistic linear logic sequent
$\Gamma \vdash M : A$ corresponds to many different typing with leftovers
sequents:
\begin{itemize}
  \item $\gr1\Gamma \vdash M : A \boxtimes \gr0\Gamma$
  \item $\gr1\Gamma, \gr1x : B \vdash M : A \boxtimes \gr0\Gamma, \gr1x : B$
  \item $\gr1\Gamma, \gr0x : B \vdash M : A \boxtimes \gr0\Gamma, \gr0x : B$
  \item \&c.
\end{itemize}

Generally, any variable not used in the term can be added to both the input and
the output context with the same annotation.
Many of these variations are likely to appear in various typing derivations,
depending on what terms surround a given subterm.
This means that if we want to implement substitution, which involves putting a
term into an unknown surrounding, we have to navigate these different forms via
the \emph{framing property}.

The unusal form of sequents also somewhat obscures the denotational semantics of
a typing with leftovers system.
Though the $\boxtimes$ notation suggests a semantics into symmetric monoidal
closed categories where terms are morphisms from one iterated tensor product
(the input context) to another (the type and output context), the syntax is
incomplete for this semantics because we cannot produce anything interesting in
the output context.

\subsection{yalla}
\begin{itemize}
  \item Sequent calculus
  \item Specialised to linear logic
\end{itemize}

### 2022-05-27 tex/lnl.tex
[ writing ] lots
[ more ] state the other LnL translation lemma
[ more ] thesis writing

@@
 only a distinction in the conclusion mode is necessary.
-To achieve this distinction in the types, we have an indexed type family
+To achieve this distinction in the types, I have an indexed type family
 $\mathrm{Ty} : \mathrm{Frag} \to \mathrm{Set}$, and set the framework types to
@@
 % \AgdaDatatype{Set}
+I present these judgements in a standard paper style in \cref{fig:lnl-types}.
 
@@
 assumptions, intuitionistic judgements can only have intuitionistic assumptions.
-We encode this invariant on intuitionistic judgements using the $\Box$
+I encode this invariant on intuitionistic judgements using the $\Box$
 premise combinator on all intuitionistic conclusions and wherever
-intuitionistic premises occur for linear conclusions.
+intuitionistic premises occur for linear conclusions, as seen in
+\cref{fig:lnl-desc}.
+This scheme means that the rules contain the maximal number of boxes, which is
+convenient when consuming terms.
+For ease of \emph{producing} terms, we would want a ``garbage in; garbage out''
+style with a minimal number of boxes, which could be achieved in two different
+ways:
+\begin{itemize}
+  \item Ensure that $\Gamma, \gr1\,\lin A \vdash \intu X$ is never derivable.
+    This leaves boxes only on rules $1$-I and $G$-I.
+  \item Ensure that, when working bottom-up, we never transition from a linear
+    to an intuitionistic conclusion with linear assumptions present.
+    This leaves boxes only on rules $F$-I and $G$-E.
+\end{itemize}
+It would be interesting to show the equivalence of these three approaches, but
+that is left to future work.
+
+\begin{figure}
+  \begin{align*}
+    A, B, C &\Coloneqq I \mid A \otimes B \mid A \multimap B \mid FX \\
+    X, Y, Z &\Coloneqq 1 \mid X \times Y \mid X \to Y \mid GA \\
+    \mathcal J &\Coloneqq \lin A \mid \intu X
+  \end{align*}
+  \caption{Types and judgement forms of L/nL}
+  \label{fig:lnl-types}
+\end{figure}
+
+\begin{figure}
+  \begin{mathpar}
+    % lin rules
+    \ebrule[comb]{%
+      \hypo{I^*}
+      \infer1[$I$-I]{\vdash \lin I}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdash \lin I}
+      \hypo{*}
+      \hypo{\vdash \lin C}
+      \infer3[$I$-E]{\vdash \lin C}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdash \lin A}
+      \hypo{*}
+      \hypo{\vdash \lin B}
+      \infer3[$\otimes$-I]{\vdash \lin A \otimes B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdash \lin A \otimes B}
+      \hypo{*}
+      \hypo{\gr1\,\lin A, \gr1\,\lin B \vdash \lin C}
+      \infer3[$\otimes$-E]{\vdash \lin C}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\gr1\,\lin A \vdash \lin B}
+      \infer1[$\multimap$-I]{\vdash \lin A \multimap B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdash \lin A \multimap B}
+      \hypo{*}
+      \hypo{\vdash \lin A}
+      \infer3[$\multimap$-E]{\vdash \lin B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdash \intu X}}
+      \infer1[$F$-I]{\vdash \lin FX}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdash \lin FX}
+      \hypo{*}
+      \hypo{\gr\omega\,\intu X \vdash \lin C}
+      \infer3[$F$-E]{\vdash \lin C}
+    }
+    \and
+    % int rules
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\dot1}}
+      \infer1[$1$-I]{\vdash \intu 1}
+    }
+    \and
+    \text{(no $1$-E)}
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box(\vdash \intu X}
+      \hypo{\dottimes}
+      \hypo{\vdash \intu Y)}
+      \infer3[$\times$-I]{\vdash \intu X \times Y}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdash \intu X \times Y}}
+      \infer1[$\times$-El]{\vdash \intu X}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdash \intu X \times Y}}
+      \infer1[$\times$-Er]{\vdash \intu Y}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\gr\omega\,\intu A \vdash \intu B}}
+      \infer1[$\to$-I]{\vdash \intu A \to B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box(\vdash \intu A \to B}
+      \hypo{\dottimes}
+      \hypo{\vdash \intu A)}
+      \infer3[$\to$-E]{\vdash \intu B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdash \lin A}}
+      \infer1[$G$-I]{\vdash \intu GA}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdash \intu GA}}
+      \infer1[$G$-E]{\vdash \lin A}
+    }
+  \end{mathpar}
+  \caption{Description of the logical rules of L/nL}
+  \label{fig:lnl-desc}
+\end{figure}
+
+\begin{proposition}
+  We can construct the following translations.
+  \begin{align}
+    (\Theta \vdash_{\mathcal C} X) &\to (\gr\omega\Theta \vdash \intu X) \\
+    (\Theta; \Gamma \vdash_{\mathcal L} A) &\to
+    (\gr\omega\Theta, \gr1\Gamma \vdash \lin A)
+  \end{align}
+\end{proposition}
+
+\begin{proposition}
+  We can construct the following translations, where $\Gamma_{\gr r}$ is the
+  list of types in $\Gamma$ that are annotated $\gr r$.%
+  \todo{I'm not sure the resultant contexts are well formed.}
+  \begin{align}
+    (\Gamma \vdash \intu X) &\to (\Gamma_{\gr\omega} \vdash_{\mathcal C} X) \\
+    (\Gamma \vdash \lin A) &\to
+    (\Gamma_{\gr\omega}; \Gamma_{\gr1} \vdash_{\mathcal L} A)
+  \end{align}
+\end{proposition}
 
@@
 The translations largely follow Benton's originals.
+In this section, I take $\name$ to be the system described in
+\cref{fig:lr-bunched} instantiated to the $\{\gr0 > \gr\omega < \gr1\}$
+posemiring and restricted to the fragment with only connectives $I$, $\otimes$,
+$\multimap$, and $\oc\gr\omega$.
+Notably, this excludes $\oc\gr0$ and $\oc\gr1$.
+I write $\oc\gr\omega$ as just $\oc$, as in traditional linear logic.
+Under these restrictions, Benton's translations of types can be used verbatim
+(\cref{fig:lnl-lr-types}).
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{.49\linewidth}
+    \centering
+    \begin{align*}
+      (-)^\circ : \mathrm{Ty}_{\name} \to \mathrm{Ty}_{\lin} \\
+      \begin{aligned}
+        I^\circ &= I \\
+        \plr{A \otimes B}^\circ &= A^\circ \otimes B^\circ \\
+        \plr{A \multimap B}^\circ &= A^\circ \multimap B^\circ \\
+        \plr{\oc A}^\circ &= GFA^\circ
+      \end{aligned}
+    \end{align*}
+  \end{subfigure}
+  \begin{subfigure}{.49\linewidth}
+    \centering
+    \begin{align*}
+      (-)^* : \mathrm{Ty}_\lnl \to \mathrm{Ty}_{\name} \\
+      \begin{aligned}
+        I^* &= I \\
+        \plr{A \otimes B}^* &= A^* \otimes B^* \\
+        \plr{A \multimap B}^* &= A^* \multimap B^* \\
+        \plr{FX}^* &= \oc X^* \\
+        1^* &= I \\
+        \plr{X \times Y}^* &= \oc X^* \otimes \oc Y^* \\
+        \plr{X \to Y}^* &= \oc X^* \multimap Y^* \\
+        \plr{GA}^* &= A^*
+      \end{aligned}
+    \end{align*}
+  \end{subfigure}
+  \caption{Translation of types between L/nL and $\name$}
+  \label{fig:lnl-lr-types}
+\end{figure}
+
+We extend both $(-)^\circ$ and $(-)^*$ to contexts pointwise on the types.
+For $(-)^\circ$, this means that the translation lands outside what is usually
+expressible in L/nL whenever the context contains annotation $\gr\omega$.
+For example, $\plr{\gr\omega I}^\circ = \plr{\gr\omega\,\lin I}$, and we do not
+see ``$\gr\omega\,\lin$'' in translations from Benton's L/nL.
+This could be fixed by inserting $G$ in such cases.
+As for $(-)^*$, note that we do not insert a $\oc$ on intuitionistic assumptions
+as Benton does.
+Instead, the annotation $\gr\omega$ achieves the same effect for us.
+This discrepancy essentially amounts to the fact that $\name$ is more like DILL
+than Girard's ILL, and Benton's translations act upon the latter.
+
+\begin{theorem}
+  We can translate from $\name$ to the linear fragment of L/nL.
+  \begin{align}
+    (\Gamma \vdash_{\name} A) \to (\Gamma^\circ \vdash_\lnl \lin A^\circ)
+  \end{align}
+\end{theorem}
+
+\begin{theorem}
+  We can translate any L/nL term to a $\name$ term as follows.
+  \begin{align}
+    (\Gamma \vdash_\lnl \intu X) &\to (\Gamma^* \vdash_{\name} X^*) \\
+    (\Gamma \vdash_\lnl \lin A) &\to (\Gamma^* \vdash_{\name} A^*)
+  \end{align}
+\end{theorem}

### 2022-05-27 tex/lrkits.tex
[ more ] thesis writing

@@
   \begin{itemize}
-    \item There is one environment $\alr{} : \Gamma \env\V {\cdot}$.
+    \item There is one environment $\alr{} : \Gamma \env\V_R {\cdot}$.
     \item For $\rho_l : \Gamma \env\V_R \Delta_l$ and
-      $\rho_r : \Gamma \env\V \Delta_r$, we have an environment
+      $\rho_r : \Gamma \env\V_R \Delta_r$, we have an environment
       $\alr{\rho_l, \rho_r} : \Gamma \env\V_R \Delta_l, \Delta_r$.
@@
   \begin{itemize}
-    \item There is one environment $\alr{} : \grP\gamma \env\V {\cdot}$
+    \item There is one environment $\alr{} : \grP\gamma \env\V_R {\cdot}$
       whenever $\grP \leq \gr0$.

### 2022-05-27 tex/prelim.tex
[ more ] thesis writing

@@
 Examples: Set, Rel, partial functions, commutative monoids.
+
+\begin{definition}
+  A \emph{category} is a collection $\obj$ of objects, a family
+  $\hom : \obj \times \obj \to \Set$, for each object $A$ an element
+  $\id : \hom(A,A)$, and for each triple of objects $A, B, C$ a function
+  ${\circ} : \hom(B,C) \times \hom(A,B) \to \hom(A,C)$ such that the following
+  laws hold for any $f : \hom(A,B), g : \hom(B,C), h : \hom(C,D)$.
+  \begin{itemize}
+    \item $\id \circ f = f$
+    \item $f \circ \id = f$
+    \item $(h \circ g) \circ f = h \circ (g \circ f)$
+  \end{itemize}
+  I sometimes write $f; g$ for $g \circ f$.
+\end{definition}
+
+I give a somewhat unusual definition of \emph{monoidal categories} via
+\emph{multicategories}.
+I use this approach to capture the purpose of monoidal categories in many
+examples, particularly in algebra and programming languages: to provide a way to
+encode morphisms of arbitrary arity as unary morphisms by representing $n$-ary
+domains as internal $n$-ary products.
+
+I choose to present multicategories in a style where composition models
+simultaneous substitution rather than single substitution.
+This choice is consistent with the rest of this thesis, which concerns
+simultaneous substitution.
+
+\begin{definition}
+  A \emph{multicategory} is a collection $\obj$ of objects, a family
+  $\hom : \List \obj \times \obj \to \Set$ (with $\hom([A_1, \ldots, A_n], B)$
+  written $\hom(A_1, \ldots, A_n; B)$), for each object $A$ an element
+  $\id : \hom(A; A)$, and for each list of lists $\Psi$, each list $\Gamma$, and
+  each object $A$, a function
+  ${\circ} : \hom(\Gamma; A) \times \plr{\prod_i \hom(\Psi_i; \Gamma_i)} \to
+  \hom\plr{\sum_i \Psi_i; A}$ such that the following laws hold.
+  \begin{itemize}
+    \item $\id \circ f = f$
+    \item $f \circ \plr{\lambda i.~\id} = f$
+    \item $h \circ \plr{\lambda i.~g_i \circ \lambda j.~f_{ij}} =
+      (h \circ \lambda i.~g_i) \circ \lambda ij.~f_{ij}$
+  \end{itemize}
+\end{definition}
+
+\begin{example}
+  The following give rise to multicategories:
+  \begin{itemize}
+    \item Sets and multi-ary functions (functions of the form
+      $A_1 \times \cdots \times A_n \to B$)
+    \item $K$-vector spaces and multilinear maps
+    \item For any simple type theory with sequents of the form
+      $\Gamma \vdash A$, where $\Gamma$ is a list of types and $A$ is a type,
+      types and open terms (up to $\beta\eta$-equivalence) form a multicategory.
+  \end{itemize}
+\end{example}
+
+Related to the multicategory of $K$-vector spaces and multilinear maps is the
+multicategory of commutative monoids and multilinear maps.
+I use the latter later in this chapter\todo{forward reference}, so I will
+develop it fully as an example.
+
+\begin{example}
+  The multicategory of \emph{commutative monoids and multilinear maps} has as
+  objects commutative monoids.
+  A morphism of type $A_1, \ldots, A_n \to B$ is a function
+  $f : A_1 \times \cdots \times A_n \to B$ such that the following laws hold:
+  \begin{itemize}
+    \item $f(x_1, \ldots, 0_{A_i}, \ldots, x_n) = 0_B$
+    \item $f(x_1, \ldots, y +_{A_i} z, \ldots, x_n) =
+      f(x_1, \ldots, y, \ldots, x_n) +_B f(x_1, \ldots, z, \ldots, x_n)$
+  \end{itemize}
+  The identity function of type $A \to A$ serves as the identity
+  (multi)morphism.
+  Composition is given by multi-place composition of functions, which satisfies
+  the above laws because in a composite $g \circ [f_1, \ldots, f_n]$, a
+  commutative monoid operation in the domain of the composite bubbles through
+  one of the $f_i$ to become one operation in position $i$ of the domain of $g$,
+  and then bubbles through $g$.
+  That this identity and composition follow the equational laws follows
+  straightforwardly by considering the underlying functions.
+\end{example}
+
+\begin{definition}
+  A multicategory $\C$ gives rise to its \emph{category of contexts} $\C^*$ as
+  follows:
+  \begin{itemize}
+    \item $\obj(\C^*) \coloneqq \List(\obj\C)$
+    \item $\C^*\plr{\sum_i \Psi_i, \Delta} \coloneqq \prod_i \C(\Psi_i; \Delta_i)$
+      for any $\Psi : \List(\List(\obj\C))$
+    \item $\id^* : \Gamma \to^* \Gamma \coloneqq \overline\id$
+    \item $(\sigma : \Delta \to^* \Theta) \circ^* (\rho : \Gamma \to^* \Delta) :
+      (\Gamma \to^* \Theta) \coloneqq \lambda i.~\sigma_i \circ \lambda j.~\rho_{ij}$
+      \todo{Still wrong}
+  \end{itemize}
+\end{definition}
+
+\begin{definition}
+  Each multicategory $\C$ \emph{restricts} to a category $\C^1$, where
+  $\C^1(A, B) \coloneqq \C(A; B)$ and identity and composition are as in $\C$.
+\end{definition}
+
+\begin{definition}\label{def:tensor-product}
+  A \emph{tensor product} in a multicategory $\C$ of objects $A$ and $B$ is an
+  object $A \otimes B$ and a map ${\otimes} : A, B \to A \otimes B$ such that
+  any morphism $f : \Gamma, A, B, \Delta \to C$ uniquely factors as:
+
+  \begin{tikzcd}
+    \Gamma, A, B, \Delta \arrow[d,"f"']
+    \arrow[r,"{\overline\id, \otimes, \overline\id}"]
+    & \Gamma, A \otimes B, \Delta
+    \arrow[dl,dashed,"f^{\Gamma, A \otimes B, \Delta}"] \\
+    C
+  \end{tikzcd}
+
+  When it is clear from context which two objects are being multiplied, I will
+  write just $f^{\otimes}$ for the unique morphism.
+\end{definition}
+
+\begin{definition}\label{def:tensor-unit}
+  A \emph{tensor unit} in a multicategory $\C$ is an object $I$ and a map
+  $I : {\cdot} \to I$ such that any morphism
+  $f : \Gamma, \Delta \to C$ uniquely factors as:
+
+  \begin{tikzcd}
+    \Gamma, \Delta \arrow[d,"f"']
+    \arrow[r,"{\overline\id, I, \overline\id}"]
+    & \Gamma, I, \Delta \arrow[dl,dashed,"f^{\Gamma, I, \Delta}"] \\
+    C
+  \end{tikzcd}
+
+  When it is clear from context where the $I$ is being introduced, I will
+  write just $f^I$ for the unique morphism.
+\end{definition}
+
+\begin{lemma}\label{thm:tensor-product-terms}
+  For each morphism $f : \Gamma \to A$ and $g : \Delta \to B$, we have a
+  morphism $(f, g) : \Gamma, \Delta \to A \otimes B$ such that we have, for any
+  $h : \Gamma', A, B, \Delta' \to C$, that
+  $h^{\otimes} \circ [\overline\id, (f, g), \overline\id] =
+  h \circ [\overline\id, f, g, \overline\id]$, and that
+  $\id = (\id, \id)^{\otimes} : A \otimes B \to A \otimes B$.
+\end{lemma}
+\begin{proof}
+  The composite $(\otimes) \circ [f, g]$ has the required type.
+
+  To prove the first equation, note that
+  \begin{align*}
+    h^{\otimes} \circ [\overline\id, (f, g), \overline\id]
+    &= h^{\otimes} \circ [\overline\id, {\otimes} \circ [f, g], \overline\id] \\
+    &= h^{\otimes} \circ [\overline\id, {\otimes}, \overline\id]
+    \circ [\overline\id, f, g, \overline\id].
+  \end{align*}
+  Then, we have that
+  $h^{\otimes} \circ [\overline\id, {\otimes}, \overline\id] = h$
+  by the defining diagram of the tensor product (\cref{def:tensor-product}).
+
+  To prove the second equation, note that the diagram below commutes.
+  Because this is an instance of the diagram in \cref{def:tensor-product},
+  uniqueness gives us the desired equation.
+
+  \begin{tikzcd}
+    A, B \arrow[d,"{\otimes}"'] \arrow[r,"{\otimes}"]
+    & A \otimes B \arrow[dl,"\id"] \\
+    A \otimes B
+  \end{tikzcd}
+\end{proof}
+
+\begin{lemma}\label{thm:tensor-unit-terms}
+  For any $h : \Gamma', \Delta' \to C$, we have that
+  $h^I \circ [\overline\id, I, \overline\id] =
+  h \circ [\overline\id, \overline\id]$
+  and $\id = I^I : I \to I$.
+\end{lemma}
+\begin{proof}
+  The first equation is just the commuting diagram of \cref{def:tensor-unit},
+  modulo composition with identity morphisms.
+  For the second equation, note that the diagram below commutes.
+  Because this is an instance of the diagram in \cref{def:tensor-unit},
+  uniqueness gives us the desired equation.
+
+  \begin{tikzcd}
+    {\cdot} \arrow[d,"I"'] \arrow[r,"I"]
+    & I \arrow[dl,"\id"] \\
+    I
+  \end{tikzcd}
+\end{proof}
+
+\begin{lemma}
+  In a multicategory $\C$ with all tensor products, $\otimes$ induces a
+  bifunctor ${\otimes} : \C, \C \to \C$.
+\end{lemma}
+\begin{proof}
+  The action on objects is given by $\otimes$.
+  The action on morphisms $f : A \to A'$ and $g : B \to B'$ is given by
+  $(f, g)^{\otimes}$.
+  Preservation of identity is already proved in \cref{thm:tensor-product-terms},
+  and preservation of composition is left as an exercise.
+\end{proof}
+
+\begin{lemma}
+  For any tensor product and tensor unit, we have the following natural
+  isomorphisms:
+  \begin{itemize}
+    \item $I \otimes A \cong A$
+    \item $A \otimes I \cong A$
+    \item $(A \otimes B) \otimes C \cong A \otimes (B \otimes C)$
+  \end{itemize}
+\end{lemma}
+
+\begin{definition}\label{def:monoidal-category}
+  A \emph{monoidal category} is a multicategory equipped with a tensor product
+  and tensor unit.
+  %A \emph{monoidal category} is a category $\C = (\obj, \hom, \id, \circ)$ with
+  %functors $I : 1 \to \C$ and ${\otimes} : \C \times \C \to \C$ together with
+  %the following natural isomorphisms:
+  %\begin{itemize}
+  %  \item $\lambda : (I \otimes {-}) \simeq {-}$
+  %  \item $\rho : ({-} \otimes I) \simeq {-}$
+  %  \item $\alpha :
+  %    (({-} \otimes {-}) \otimes {-}) \simeq ({-} \otimes ({-} \otimes {-}))$
+  %\end{itemize}
+  %such that the following coherence conditions hold:
+  %\begin{itemize}
+  %  \item
+  %    \begin{tikzcd}
+  %      (A \otimes I) \otimes B \arrow[dr,"\rho_A \otimes B"'] \arrow[rr,"\alpha_{A,I,B}"]
+  %      && A \otimes (I \otimes B) \arrow[dl,"A \otimes \lambda_B"] \\
+  %      & A \otimes B
+  %    \end{tikzcd}
+  %  \item
+  %    \begin{tikzcd}
+  %      & (A \otimes B) \otimes (C \otimes D)
+  %      \arrow[dr,"\alpha_{A,B,C \otimes D}"] &
+  %      \\
+  %      ((A \otimes B) \otimes C) \otimes D
+  %      \arrow[ur,"\alpha_{A \otimes B,C,D}"]
+  %      \arrow[d,"\alpha_{A,B,C} \otimes D"]
+  %      &&
+  %      A \otimes (B \otimes (C \otimes D))
+  %      \\
+  %      (A \otimes (B \otimes C)) \otimes D
+  %      \arrow[rr,"\alpha_{A,B \otimes C,D}"]
+  %      &&
+  %      A \otimes ((B \otimes C) \otimes D)
+  %      \arrow[u,"A \otimes \alpha_{B,C,D}"]
+  %    \end{tikzcd}
+  %\end{itemize}
+\end{definition}
+
+\begin{remark}
+  We can talk about monoidal products on a category $\C$ by finding a
+  multicategory which restricts to $\C$ and constructing tensor products in it.
+\end{remark}
+
+\begin{definition}
+  A \emph{monoid} in a multicategory is an object $M$ and multimorphisms
+  $\eta : {\cdot} \to M$ and $\mu : M, M \to M$ such that the following
+  equations hold:
+  \begin{itemize}
+    \item $[\eta, \id]; \mu = \id : M \to M$
+    \item $[\id, \eta]; \mu = \id : M \to M$
+    \item $[\mu, \id]; \mu = [\id, \mu]; \mu : M, M, M \to M$
+  \end{itemize}
+\end{definition}
+
+\begin{definition}
+  TODO: \emph{symmetric monoidal category}
+\end{definition}
+
+\begin{definition}
+  TODO: \emph{commutative monoid in a symmetric monoidal category}
+\end{definition}
+
+Algebra
+
+\begin{definition}
+  A \emph{semiring} is a monoid in the multicategory of commutative monoids and
+  multilinear maps.
+  Unpacked, this means that we have a set $\mathscr R$ together with elements
+  $0$ and $1$ and binary operators $+$ and $\cdot$ (with $\cdot$ usually written
+  as juxtaposition) such that the following hold for all
+  $x, y, z \in \mathscr R$.
+  \begin{itemize}
+    \item $0 + x = x$; $x + 0 = x$; $(x + y) + z = x + (y + z)$; $x + y = y + x$
+    \item $1x = x$; $x1 = x$; $(xy)z = x(yz)$
+    \item $0x = 0$; $x0 = 0$; $(x + y)z = xz + yz$; $x(y + z) = xy + xz$
+  \end{itemize}
+\end{definition}

### 2022-05-27 tex/semantics.tex
[ more ] thesis writing

--- 2021-10-07 tex/esop22/semantics.tex
@@
-Our aim with the following traversals is to derive a recursion scheme for
-terms which abstracts over some of the details of variable binding and
-changes to usage annotations.
-We follow the structure given by \citet{AACMM20} closely, with
-the main difference having come already in the syntax and environments.
+Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
+\AgdaFunction{semantics} we define in this section assigns a
+$\C$-value in context $\Gamma$ to every term in context $\Delta$,
+where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
+semantic interpretation of terms ($\V$ being the semantic
+interpretation of variables). Before we can define
+\AgdaFunction{semantics}, we need to treat recursion through rules'
+premises (\cref{sec:functorial}) and extension of environments when
+going under variable binders (\cref{sec:kripke}).
 
-\subsection{Renameability}
+% Our goal in this section is to define \AgdaFunction{semantics}, a
+% recursor that turns a term into a \AgdaBound{$\C$}-value using a
+% \AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
+%   ``body'' here}
 
-{\color{red}(Maybe this goes in the Environment section.)}
-We saw in \cref{sec:env} that our notion of \emph{environment} specialises to
-\emph{simultaneous renaming} by setting $\V$ to be $\sqni$.
+% \ExecuteMetaData[\Semanticstex]{semantics-type}
 
-\ExecuteMetaData[\Renamingtex]{Ren}
-\ExecuteMetaData[\Renamingtex]{Boxr}
-\ExecuteMetaData[\Renamingtex]{Renameable}
+% The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
+% representing the interpretations of variables and terms
+% respectively. In \cref{sec:traversal} we will see the data that must
+% be provided to make a \AgdaFunction{semantics} for a given
+% system. Before that, we must see how to deal with the two complicated
+% features of our syntax: the usage annotations (\cref{sec:functorial})
+% and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}
 
-\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
-  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
-  $\plr{-} \env\V \Delta$ for all $\Delta$.
-\end{lemma}
-\begin{proof}
-  For renameable $\V$, we can implement $\mathrm{lift}$ as in
-  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
-  $\V \coloneqq \W \coloneqq \V$.
-  Renaming the environment then becomes an instance of environment composition.
-\end{proof}
+\subsection{A layer of syntax is functorial}\label{sec:functorial}
 
-\subsection{The Kripke function space}
+A basic property of the universe of syntaxes we described in \cref{sec:syntax}
+is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
+Its type says that to map a function \AgdaBound{f}
+over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
+domain and codomain usage contexts, and \AgdaBound{f} should be usable
+wherever the domain and codomain usage contexts are similarly related by
+\AgdaBound{F}.
 
-Earlier, when talking about syntax, we saw \AgdaFunction{Scope} as a simple way
-to let an open family admit context extensions.
-However, a key component of our generic semantic traversal is to make use of
-the open family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
-we store in an environment.
-The definition \AgdaFunction{Kripke} gives an alternative to
-\AgdaFunction{Scope} which interprets the newly bound variables via a
-requirement of $\V$-values rather than extra assumptions for the
-$\C$-computation.
+\ExecuteMetaData[\Maptex]{map-s-type}
+
+This generality is needed because usage contexts change between
+a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
+\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
+\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
+variables newly bound in that subterm).
+Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
+This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.
+
+The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
+works through the structure \AgdaBound{ps} of premises of the rule applied,
+acting on each subterm it finds.
+Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
+\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
+function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
+--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
+and \AgdaBound{Q$'$}.
+In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
+$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
+collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
+Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
+$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
+$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
+The cases for \AgdaInductiveConstructor{`$\cdot$} and
+\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
+of linearity.
+In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
+\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
+systems, do not involve any changes in the usage contexts.
+
+\ExecuteMetaData[\Maptex]{map-p}
+
+\subsection{The Kripke function space}\label{sec:kripke}
+
+At this point we introduce a minor generalisation to
+\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam}:
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
+\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.  We obtain the
+definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} by
+replacing the textual occurrence of \AgdaBound{Ty} by the parameter
+\AgdaBound{I}.
+
+The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
+of function space that describes a $\C$ value parametrised by
+$\Delta$-many additional $\V$s (all correctly typed and usage
+annotated). It is used to describe how to go under binders in a
+Higher-Order Abstract Syntax style---to go under a binder we must
+provide semantic interpretations for all the additional variables:
+
+% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environment must be extended with $\Theta$s-worth of $\V$s
+
+% we need the ability to say that
+
+% Kripke V C is given the extension \Theta
+
+% In \cref{sec:terms}, we defined \AgdaFunction{Scope} to let a
+% judgement-indexed family admit context extensions. However, a key
+% component of our generic semantic traversal is to make use of the open
+% family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
+% we store in an environment.  The definition \AgdaFunction{Kripke}
+% gives an alternative to \AgdaFunction{Scope} which interprets the
+% newly bound variables via a requirement of $\V$-values rather than
+% extra assumptions for the $\C$-computation.
 
@@
 
-\AgdaFunction{Wrap} is a device that turns any type family into an equivalent
-type family that is judgementally injective in its indices.
-It turns the type family into a parametrised record with a single field
-\AgdaField{get} whose type is the type in the body of the $\lambda$-abstraction.
-For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
-ignored.
+\AgdaFunction{Wrap}
+is a device that turns any type family into an equivalent type family
+that is judgementally injective in its indices, which helps with
+Agda's type inference.
+It turns the type family into a parametrised
+record with a single field \AgdaField{get} whose type is the type in
+the body of the $\lambda$-abstraction.
+For understanding the meaning of
+\AgdaFunction{Kripke}, \AgdaFunction{Wrap} can be ignored.
+
 If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
 \ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
-\ExecuteMetaData[\Snippetstex]{KripkeExpanded}.
-That is to say, the Kripke function is expecting a value for each newly bound
-variable, at the multiplicity of its annotation, together with the resources
-supporting each of those values.
-The need for \AgdaFunction{$\Box^r$} is exemplified by the following construct
-\AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
-version of \AgdaBound{$\Delta$}.
-The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
+\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
+to say, the Kripke function is expecting a value for each newly bound
+variable, at the multiplicity of its annotation, together with the
+resources supporting each of those values. We use the ``magic wand''
+function space here to enforce the invariant that the freshly bound
+variables have usage annotations that are added to the existing
+variables, not shared with them. The use of the
+\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
+the presence of additional variables introduced by weakening.
 
-When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
-we can coerce a \AgdaFunction{Kripke} function into just a
-\AgdaBound{$\C$}-value in an extended context.
-To show this, we instantiate the Kripke function with the renaming
-$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
-pass it an argument environment
-$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
-extended part.
-Post-composition of a renaming onto an environment comes from
-\cref{thm:env-postren}.
+\AgdaFunction{Kripke} is functorial in the \AgdaBound{$\C$} argument,
+as witnessed by the \AgdaFunction{mapK$\C$} function, which is essentially
+post-composition:
 
-\ExecuteMetaData[\Syntactictex]{reify}
+\ExecuteMetaData[\Semanticstex]{mapKC}
 
-We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
-identity environments.
-In this paper, all examples support identity environments.
-However, \citet[p.~27]{AACMM20} give some important examples that do not
-support identity environments, and thus cannot use \AgdaFunction{reify}.
-The feature that causes the lack of support for identity environments is that
-a semantics can make use of the fact that only variables of particular kinds
-are bound by the syntax.
-In the examples of \citeauthor{AACMM20}, a bidirectionally typed language only
-binds variables that synthesise their type, as opposed to those whose type is
-checked.
-The semantics of type-checking and elaboration rely on variables synthesising
-their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
-Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
-only binds such synthesising variables.
-Similar phenomena would appear in, say, a call-by-value language where
-variables are values (not computations), or a polarised language where
-variables always have a polarity matching their type.
+% is exemplified by the following construct
+% \AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
+% version of \AgdaBound{$\Delta$}.
+% The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.
 
-\subsection{Semantic traversal}
+% \bob{Move this para}
+% This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
+% type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
+% We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
+% to \AgdaDatatype{Ctx}.
 
-The aim of this subsection is to give an alternative recursion principle for
-terms that incorporates some of the environment-handling seen in the
-implementations of renaming and substitution.
-The rest of this section assumes the following: a renameable open family
-\AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
-algebra for a layer of syntax at \AgdaBound{$\C$}.
+\subsection{Semantic traversal}\label{sec:traversal}
+
+We can now state the data required to implement a traversal assigning
+semantics to terms. For open families $\V$ and $\C$, interpreting
+variables and terms respectively, we assume that $\V$ is renameable,
+that $\V$ is embeddable in $\C$, and that we have an algebra for a
+layer of syntax, where bound variables are handled using the Kripke
+function space:
+
+% The aim of this subsection is to give an alternative recursion principle for
+% terms that incorporates some of the environment-handling seen in the
+% implementations of renaming and substitution.
+% The rest of this section assumes the following: a renameable open family
+% \AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
+% algebra for a layer of syntax at \AgdaBound{$\C$}.
 
@@
 
-For \AgdaFunction{body}, we are given a subterm \AgdaFunction{M}, to which we
-want to apply \AgdaFunction{semantics}.
-To do so, we need an extended version of the initial environment
-\AgdaBound{$\rho$}.
-The best we can achieve without identity environments for \AgdaBound{$\V$} is
-a Kripke function returning an extended environment.
-We use \AgdaFunction{mapK$\C$} to post-compose this function by the
-\AgdaSymbol{$\lambda$}-function taking an extended environment and acting with
-it on \AgdaBound{M}.
+For \AgdaFunction{body}, we are given a subterm \AgdaBound{M}, to
+which we want to apply \AgdaFunction{semantics}.  To do so, we need an
+extended version of the initial environment \AgdaBound{$\rho$}. We
+express this as the generation of a Kripke function that produces the
+extended environment given interpretations of the fresh variables. We
+take \AgdaBound{$\rho$}, which is an environment covering
+\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an
+environment covering \AgdaBound{$\Theta$}, and glue them together
+using the inductive rules for generating environments, after having
+renamed \AgdaBound{$\rho$} via \cref{thm:env-ren} to make it fit the
+new context \AgdaBound{$\Gamma^+$} (intended to be
+\ExecuteMetaData[\Snippetstex]{GT}):
+
+\ExecuteMetaData[\Semanticstex]{extend}
+
+% The best we can achieve without identity environments for \AgdaBound{$\V$} is
+% a Kripke function returning an extended environment.
+To define \AgdaFunction{body}, we use \AgdaFunction{mapK$\C$} to
+post-compose the environment extension by the
+\AgdaSymbol{$\lambda$}-function taking an extended environment and
+acting with it on \AgdaBound{M}.
 
 \ExecuteMetaData[\Semanticstex]{body}
+
+% \todo{FIX} Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
+% \cref{thm:lr-bind} as
+% \AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
+% \AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
+% We can think of \AgdaFunction{extend} as our best effort to extend an
+% environment by \AgdaBound{$\Theta$} without access to an identity environment
+% at \AgdaBound{$\Theta$}.

### 2022-05-27 tex/semirings.tex
[ more ] thesis writing

@@
 In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
-appeared previously in my work with Atkey~\cite{WA20}, and can be seen as a
+appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
 simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
@@
 \input{lr.tex}
-\section{Linear natural deduction}\label{sec:lnd}
+\section{Bunched connectives}\label{sec:lnd}
 \input{lnd.tex}

### 2022-06-06 tex/ill.tex
[ writing ] introduction to MALL

@@
 \subsection{Motivation of linearity}
+
+While classical linear logic is perhaps better understood, for the sake of this
+thesis I focus on intuitionistic linear logic.
+The syntactic difference between the two is analogous to the difference between
+classical and intuitionistic logic: The latter restricts the former by only
+allowing one formula on the right of a sequent.
+Classical linear logic is not irrelevant to this thesis, and appears as an
+instance of the framework presented in \cref{sec:framework}.
+However, the connection of this work to intuitionistic linear logic is more
+direct, and reflects the basis of existing semiring-annotated calculi on
+(intuitionistic) typed $\lambda$-calculi.
+
+Intuitionistic linear logic is often motivated by its sensitivity to resources.
+The fact that, in intuitionistic (and classical) logic, \emph{any} hypothesis
+can be discarded or duplicated means that propositions cannot be used to model
+things that may be ``possessed'' or ``spent'', despite hypotheses often being
+described as things we ``have''.
+In contrast, linear logic allows propositions to behave like physical objects,
+being moved around but not being created or destroyed (a priori).
+
+Programming applications of linearity include stateful protocols
+\citep{Wadler12}, mutually exclusive capabilities (TODO: cite),
+and mutation (TODO: cite).
+In each of these cases, the act of using up a hypothesis allows us to divide
+time into ``before'' and ``after'', with lack of duplication avoiding confusion
+over which half we are in, and lack of deletion allowing us to know whether the
+corresponding action actually happened.
+For example, when modelling mutation, a variable refers to a single state of a
+mutable value.
+The \texttt{write} operation uses up such a variable, making the old state
+inaccessible, and returns a new linear value to represent the new state of the
+mutable value.
+Such a protocol naturally also supports a \emph{freezing} operation, by which we
+relinquish the ability to mutate the value in return for an immutable reference
+to the value produced.
 
@@
 
-\subsection{The $\oc$-modality}
+\subsection{The multiplicative-additive fragment}
+
+The multiplicative-additive fragment of linear logic (MALL) is the fragment
+where all hypotheses are linear (must be used exactly once).
+We will extend MALL with the \emph{exponential} modality in
+\cref{sec:bang-modality}.
+MALL is unable to embed intuitionistic or classical logic (collectively
+known as \emph{structural logics}), as MALL is unable to reflect any of the
+discarding or duplication that can be done in proofs in structural logics.
+
+In short, the syntax of intuitionistic MALL can be described as intuitionistic
+logic with the structural rules of \emph{weakening} and \emph{contraction}
+removed.
+However, without the presence of weakening and contraction, we have to be more
+careful about the rules we state, so as not to accidentally introduce weakening
+and contraction admissibly.
+The lack of these structural rules also allows us to observe a new phenomenon:
+the distinction (at the level of provability) between \emph{additive} and
+\emph{multiplicative} formulations of existing connectives (in particular, in
+the intuitionistic case, the conjunction connective).
+
+I present MALL in \cref{fig:mall} in a sequent calculus style, as it was
+presented by \citet{girard87linear}.
+
+To encode what it means to use a hypothesis \emph{exactly once}, we first need
+to decide what counts as a use.
+The simplest case is that the identity sequent counts as a single use of its
+sole hypothesis, and conversely does not count as a use of any other hypotheses.
+For sequential proofs, created by the \TirName{Cut} rule, if we have a proof of
+$A$ using $\Gamma$, and a proof of $B$ using $\Delta$ and $A$, then we have a
+proof of $B$ transitively using $\Delta$ and $\Gamma$.
+The exchange rule Exch says that use is invariant under permutation.
+
+For the logical connectives, we have genuine choices as to what it means to use
+them.
+Two cases --- disjunction ($\oplus$) and (linear) implication ($\multimap$) ---
+are somewhat intuitive from intuitionistic logic.
+A canonical proof of a disjunction is a tag and a proof of one of the two
+disjuncts.
+This suggests that a proof of a disjunction only uses the same hypotheses as
+the proof of the disjunct we actually choose, with the other disjunct being
+irrelevant.
+Correspondingly, when we use a disjunction hypothesis, we will only actually use
+one of the cases, so each branch should use the same hypotheses.
+For implication, use is sequential like with the \TirName{Cut} rule, and its
+left rule is more or less the only choice that allows use of the surrounding
+hypotheses.
+
+For conjunction, there are two choices: Either the conjuncts \emph{together} use
+all of the hypotheses, or each of the conjuncts \emph{individually} uses all of
+the hypotheses.
+The former choice gives us the tensor-product ($\otimes$), and the latter choice
+gives us the with-product ($\with$).
+These products are equivalent up to provability in structural logics but
+distinct in linear logic.
+
+\begin{figure}
+  \begin{align*}
+    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
+              \mid 0 \mid A \oplus B \mid \top \mid A \with B \\
+    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A
+  \end{align*}
+  \begin{mathpar}
+    \ebrule{%
+      \infer0[Id]{A \vdash A}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash A}
+      \hypo{\Delta, A \vdash B}
+      \infer2[Cut]{\Gamma, \Delta \vdash B}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma, B, A, \Delta \vdash C}
+      \infer1[Exch]{\Gamma, A, B, \Delta \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash C}
+      \infer1[$I$-L]{\Gamma, I \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \infer0[$I$-R]{{\cdot} \vdash I}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma, A, B \vdash C}
+      \infer1[$\otimes$-L]{\Gamma, A \otimes B \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash A}
+      \hypo{\Delta \vdash B}
+      \infer2[$\otimes$-R]{\Gamma, \Delta \vdash A \otimes B}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash A}
+      \hypo{\Delta, B \vdash C}
+      \infer2[$\multimap$-L]{\Gamma, \Delta, A \multimap B \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma, A \vdash B}
+      \infer1[$\multimap$-R]{\Gamma \vdash A \multimap B}
+    }
+
+    \and
+
+    \ebrule{%
+      \infer0[$0$-L]{\Gamma, 0 \vdash C}
+    }
+
+    \and
+
+    \text{(no $0$-R)}
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma, A \vdash C}
+      \hypo{\Gamma, B \vdash C}
+      \infer2[$\oplus$-L]{\Gamma, A \oplus B \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash A_i}
+      \infer1[$\oplus$-R$_i$]{\Gamma \vdash A_0 \oplus A_1}
+    }
+
+    \and
+
+    \text{(no $\top$-L)}
+
+    \and
+
+    \ebrule{%
+      \infer0[$\top$-R]{\Gamma \vdash \top}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma, A_i \vdash C}
+      \infer1[$\with$-L$_i$]{\Gamma, A_0 \with A_1 \vdash C}
+    }
+
+    \and
+
+    \ebrule{%
+      \hypo{\Gamma \vdash A}
+      \hypo{\Gamma \vdash B}
+      \infer2[$\with$-R]{\Gamma \vdash A \with B}
+    }
+  \end{mathpar}
+  \caption{Multiplicative-additive fragment of linear logic}
+  \label{fig:mall}
+\end{figure}
+
+Implication ($\multimap$) and the tensor-product ($\otimes$, $I$) comprise the
+\emph{multiplicative} fragment, while disjunction ($\oplus$, $0$) and the
+with-product ($\with$, $\top$) comprise the \emph{additive} fragment.
+Categorically, the additive fragment corresponds to products and coproducts,
+while the multiplicative fragment corresponds to multicategorical tensor
+products and closure.
+
+\subsection{The $\oc$-modality}\label{sec:bang-modality}
 

### 2022-06-06 tex/linear.tex
[ more ] thesis writing
[ writing ] introduction to MALL

@@
 \chapter{Linearity}\label{sec:linearity}
+
+Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
+Its distinguishing feature is that hypotheses, by default, must be used exactly
+once.
+I devote a chapter to introducing linear logic because it is better understood
+than any individual or general semiring-annotated calculi, and is thus an
+important point of comparison.
+Linear logic will be an instance of the semiring-annotated systems we study
+later.
 
@@
 \section{Mechanisation survey}
+\input{linmech.tex}

### 2022-06-06 tex/simple.tex
[ writing ] introduction to MALL

@@
 \chapter{Mechanisation of simple types}\label{sec:simple}
+
+In this chapter, I review and justify the family of approaches usually used to
+represent simple type systems inside dependently typed proof assistants.
+These approaches were first presented by \citet{AR99}, who showed a way of
+representing well \emph{scoped} terms in a language with polymorphic recursion,
+and extended the representation to well \emph{typed} terms in a language with
+dependent types.
+The representation of terms relies on indexing on both a context --- giving the
+types of all the free variables --- and a type for the term itself.
+A basic operation on terms is \emph{simultaneous substitution}, which replaces
+each variable in the context by a term in another context.
+This simultaneous substitution operation fits the form of composition in a
+multicategory.
+Specifically, in the case of a simple intuitionistic type system, this is the
+composition of a Cartesian multicategory, which allows all of the terms being
+substituted in to share a context.
 

### 2022-08-01 tex/framework.tex
[ fix ] some citations and missing files
[ writing ] introduction to MALL
[ writing ] content sections to full draft

@@
-\chapter{A framework for usage-restricted calculi}
+\chapter{A framework for usage-restricted calculi}\label{sec:framework}
+
+In \cref{sec:semirings}, we saw how to use parametrisation over a partially
+ordered semiring to recreate a range of usage-aware calculi.
+However, $\name$, with its fixed set of type formers and syntactic forms, is a
+long way from capturing the full range of linear-like programming languages
+studied in the literature and required in practice.
+
+In this chapter, I take the framework for typed syntaxes with binding developed
+by \citet{AACMM21} and apply the principles we discovered in
+\cref{sec:semirings} to yield a framework allowing for semiring-based usage
+restrictions on variables.
+Syntactically, I claim that this framework ranges over all finitiary
+variable-based simply typed semiring-annotated calculi, with justification by
+comparison to the framework of \citet{AACMM21} and some novel examples in
+\cref{sec:other-syntaxes}.
+I also derive analogues to some of the semantic results of \citet{AACMM21},
+strengthening them to take advantage of usage restrictions.
+
+The work in this chapter is fully mechanised in Agda, which allows me to be
+precise about the various levels of domain-specific languages which appear.
+I assume that the reader is familiar with the bunched connectives introduced in
+\cref{fig:bunched} and the usage-aware environments of \cref{def:lr-env}.
 
@@
 \input{semantics}
+
+\section{Example semantics}\label{sec:example-semantics}
+
+\input{examples}

### 2022-08-01 tex/lr.tex
[ more ] thesis writing
[ writing ] content sections to full draft

@@
 In this section, I introduce the syntax of the type theory \name{}, which makes
-use of usage annotations.
+use of posemiring usage annotations.
 I use this syntax to write some example programs, which will motivate the
@@
 \begin{example}
-  The following judgements hold for any partially ordered semiring.
+  Let $A \multimapboth B$ abbreviate
+  $\plr{A \multimap B} \with \plr{B \multimap A}$.
+  Then the following judgements hold for any partially ordered semiring.
   Derivations are left as an exercise to the reader.
@@
 \end{example}
+
+\subsection{Other posemirings}\label{sec:example-posemirings}
+
+Now that we have seen the role of usage annotations in $\name$, I will give more
+examples of posemirings for tracking interesting usage patterns.
+
+\begin{example}\label{def:trivial-posemiring}
+  The singleton set gives rise to a posemiring in a unique way.
+  When the usage annotations of $\name$ are taken from this trivial posemiring,
+  we recover a version of intuitionistic simply typed $\lambda$-calculus.
+\end{example}
+
+{\color{red}TODO: cite rntz}
+\begin{example}\label{def:monotonicity-posemiring}
+  The \emph{monotonicity} posemiring is defined over the set of symbols
+  $\{\gr{\wn\wn}, \gr{\uparrow\uparrow}, \gr{\downarrow\downarrow},
+  \gr{\sim\sim}\}$, with $0 \coloneqq \gr{\wn\wn}$,
+  $1 \coloneqq \gr{\uparrow\uparrow}$, and the following operations:
+
+  \makebox[\textwidth][s]{
+    \begin{tabular}{c|cccc}
+      $+$ & $\gr{\wn\wn}$ & $\gr{\uparrow\uparrow}$ & $\gr{\downarrow\downarrow}$ & $\gr{\sim\sim}$ \\ \hline
+      $\gr{\wn\wn}$ & $\gr{\wn\wn}$ & $\gr{\uparrow\uparrow}$ & $\gr{\downarrow\downarrow}$  & $\gr{\sim\sim}$ \\
+      $\gr{\uparrow\uparrow}$ & $\gr{\uparrow\uparrow}$ & $\gr{\uparrow\uparrow}$ & $\gr{\sim\sim}$ & $\gr{\sim\sim}$ \\
+      $\gr{\downarrow\downarrow}$ & $\gr{\downarrow\downarrow}$ & $\gr{\sim\sim}$ & $\gr{\downarrow\downarrow}$ & $\gr{\sim\sim}$ \\
+      $\gr{\sim\sim}$ & $\gr{\sim\sim}$  & $\gr{\sim\sim}$ & $\gr{\sim\sim}$ & $\gr{\sim\sim}$ \\
+    \end{tabular}
+    \begin{tabular}{c|cccc}
+      $*$ & $\gr{\wn\wn}$ & $\gr{\uparrow\uparrow}$ & $\gr{\downarrow\downarrow}$ & $\gr{\sim\sim}$ \\ \hline
+      $\gr{\wn\wn}$ & $\gr{\wn\wn}$ & $\gr{\wn\wn}$ & $\gr{\wn\wn}$  & $\gr{\wn\wn}$ \\
+      $\gr{\uparrow\uparrow}$ & $\gr{\wn\wn}$ & $\gr{\uparrow\uparrow}$ & $\gr{\downarrow\downarrow}$ & $\gr{\sim\sim}$ \\
+      $\gr{\downarrow\downarrow}$ & $\gr{\wn\wn}$ & $\gr{\downarrow\downarrow}$ & $\gr{\uparrow\uparrow}$ & $\gr{\sim\sim}$ \\
+      $\gr{\sim\sim}$ & $\gr{\wn\wn}$  & $\gr{\sim\sim}$ & $\gr{\sim\sim}$ & $\gr{\sim\sim}$ \\
+    \end{tabular}
+    \begin{tikzpicture}[baseline]
+      \node(omega) at (0,-1) {$\gr{\sim\sim}$};
+      \node(0) [above left of=omega] {$\gr{\uparrow\uparrow}$};
+      \node(1) [above right of=omega] {$\gr{\downarrow\downarrow}$};
+      \node(qq) [above right of=0] {$\gr{\wn\wn}$};
+
+      \draw (omega) -- (0);
+      \draw (omega) -- (1);
+      \draw (0) -- (qq);
+      \draw (1) -- (qq);
+    \end{tikzpicture}
+  }
+
+  The idea is that each symbol represents the possible \emph{variance} of an
+  input (free variable) with respect to some partial ordering on a semantic
+  domain of elements.
+  $\gr{\uparrow\uparrow}$ represents covariance (if that input goes up, the
+  output goes up), $\gr{\downarrow\downarrow}$ represents contravariance
+  (if that input goes \emph{down}, the output goes up), $\gr{\sim\sim}$ gives no
+  guarantees (if that input remains constant, the output (trivially) goes up),
+  and $\gr{\wn\wn}$ says that that input is irrelevant (whatever changes are
+  made to that input, the output (trivially) goes up).
+
+  Addition represents an intersection of guarantees.
+  For example, if a variable is used covariantly in one subterm and
+  contravariantly in another, we can only make the trivial guaratee represented
+  by $\gr{\sim\sim}$.
+  Multiplication is mainly interesting for multiplication by
+  $\gr{\downarrow\downarrow}$, which flips the variance on any other annotation.
+  As such, $\oc\gr{\downarrow\downarrow}A$ represents a contravariant $A$.
+\end{example}
+
+\begin{example}\label{def:sensitivity-posemiring}
+  The \emph{sensitivity} posemiring~\citep{reed10distance} is given by
+  $(\mathbb R^+, \geq, \gr0, +, \gr1, \times)$, where $\mathbb R^+$ is the
+  non-negative real numbers extended with $\gr\infty$ (distances), and the rest
+  of the structure comes from the standard operations on real numbers (except
+  that $\gr0 \times \gr\infty = \gr\infty \times \gr0 = \gr0$).
+  Note that the order is reversed, making $\gr\infty$ coercible to any other
+  annotation and anything coercible to $\gr0$.
+
+  This posemiring can be used for sensitivity analysis, where we want to bound
+  the effect of a perturbation on inputs in terms of some semantic notion of
+  distance between values.
+  An annotation $\gr r$ says that if that input is perturbed by at most $r$,
+  then the output will change by at most $1$.
+  An $\gr\infty$-annotated variable gives the very strong guarantee that any
+  change in the input will make a minimal change to the output, while a
+  $\gr0$-annotated variable provides no guarantee at all.
+  Addition forbids general contraction, which would otherwise allow arbitrary
+  finite blow-up of the effect of any non-$\gr0$-annotated variable.
+  However, the ordering, with $\gr0$ at the top, means that we have general
+  weakening, so the resulting sensitivity calculus has an affine flavour.
+\end{example}

### 2022-08-01 tex/semirings.tex
[ writing ] content sections to full draft

@@
 
-As explained by McBride~\cite{McBride16}, the first insight is to, instead of
+As explained by \citet{McBride16}, the first insight is to, instead of
 removing variables from the context of certain subterms, add an annotation to
@@
 (left) as a rule with usage annotations (right).
-I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context is
-now a list of types together with their usage annotations.
-\todo{Reword: what's capital $\Gamma$}
-If $\gamma = x_1 : A_1, \ldots, x_m : A_m$, then
-$\gr r\gamma = \gr rx_1 : A_1, \ldots, \gr rx_m : A_m$.
+I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context now
+looks like $\grP\gamma$, where $\grP$ is a list of usage annotations
+$\gr{r_1}, \ldots, \gr{r_m}$ and $\gamma$ is a list of types $A_1, \ldots, A_m$
+of the same length.
+Explicit contexts will usually be written with usage annotations and types
+interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
+I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.
 
 \[
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\Gamma \vdash A}
@@
     \infer2{\Gamma, \Delta \vdash A \otimes B}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\gr1\gamma, \gr0\delta \vdash A}
@@
     \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
-  \end{prooftree}
+  }
 \]
@@
 What the existence of $\gr0$ has bought us is that variables never go out of
-scope in subterms; rather, we lose the ability to use certain variables.
+scope in subterms; rather, we lose the ability to use certain variables that
+remain in scope.
 Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
@@
 \[
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\Theta; \Gamma \vdash A}
@@
     \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
@@
     \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  \end{prooftree}
+  }
 \]
@@
 $\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
-The capital calligraphic letters $\grP$, $\grQ$, and $\grR$ stand for all of
-the usage annotations in a given context, with $\grR\gamma$ standing for
-$\grR_1x_1 : \gamma_1, \ldots, \grR_mx_m : \gamma_m$.
-Addition lifts pointwise to these vectors of usage annotations.
+Addition lifts pointwise to vectors of usage annotations (the green capital
+calligraphic $\grP$, $\grQ$, and $\grR$).
 It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
-rule on the right now accepts $\gr0$-annotated variables in its conclusion,
-which is essential for weakening to be admissible.
+rule on the right below now accepts $\gr0$-annotated variables in its
+conclusion, which is essential for weakening to be admissible.
 
 \[
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
@@
     \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\grR = \grP + \grQ}
@@
     \infer3{\grR\gamma \vdash A \otimes B}
-  \end{prooftree}
+  }
 \]
 
-Some other transcriptions:
+Some other transcriptions are as follows.
+I unify the variable rules (the one for linear variables and the one for
+intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
+annotations.
+We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
+demand of a linear variable by dereliction.
+We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
+weakened away like $\gr0$-annotated variables.
+All together, this means that at the unified variable rule, the variable being
+used must have annotation less than or equal to $\gr1$, and every other variable
+must have annotation less than or equal to $\gr0$.
+Thus the basis vector $\langle x \rvert$ describes the upper bound on the
+current usage context $\grR$.
 
 \[
-  \begin{prooftree}
+  \ebrule{%
     \infer0{\Theta; A \vdash A}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\grR \leq \bra x}
@@
     \infer2{\grR\gamma \vdash A}
-  \end{prooftree}
+  }
 \]
@@
 \[
-  \begin{prooftree}
+  \ebrule{%
     \infer0{\Theta, A; {\cdot} \vdash A}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
-  \end{prooftree}
+  }
   \quad\rightsquigarrow\quad
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\grR \leq \bra x}
@@
     \infer2{\grR\gamma \vdash A}
-  \end{prooftree}
+  }
 \]
+
+The final interesting rule form to cover is that found in DILL's
+$\oc$-introduction rule.
+I choose to see DILL's $\oc$-introduction as an $\gr\omega$-ary counterpart to
+$\otimes$-introduction, though with the same premise each time rather than
+$\gr\omega$-many premises.
+This explanation recovers the behaviour that only $\gr\omega$- and
+$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
+and also justifies the designation of multiplication (vector scaling) as the
+algebraic operation controlling the $\oc$-modality.
+
+\[
+  \ebrule{%
+    \hypo{\Theta; {\cdot} \vdash A}
+    \infer1{\Theta; {\cdot} \vdash \oc A}
+  }
+  \quad\rightsquigarrow\quad
+  \ebrule{%
+    \hypo{\gr\omega\theta, \gr0\delta \vdash A}
+    \infer1{\gr\omega\theta, \gr0\delta \vdash \oc A}
+  }
+  \quad\rightsquigarrow\quad
+  \ebrule{%
+    \hypo{\grR \leq \gr\omega\grP}
+    \hypo{\grP\gamma \vdash A}
+    \infer2{\grR\gamma \vdash \oc_{\gr\omega} A}
+  }
+\]
+
+In summary, the structure we have required of the set of usage annotations is
+that they have addition (for $\otimes$-introduction and similar rules),
+multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
+$0$ (for a variable being discarded), and an ordering (allowing for subsumption
+of usage restrictions).
+Together, these form a \emph{partially ordered semiring} (posemiring), the laws
+of which are both supported by examples and necessary for the syntax to be well
+behaved.
+\todo{Refer back to preliminaries for definition of posemiring}
+For concreteness, I collect together the definition of the
+$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.
+
+\begin{example}\label{def:lin-semiring}
+  The \emph{$\{\gr0, \gr1, \gr\omega\}$ semiring}, also known as the
+  \emph{linearity semiring}, has the operations given as follows, with
+  $0 \coloneqq \gr0$ and $1 \coloneqq \gr1$:
+
+  \makebox[\textwidth][s]{
+    \begin{tabular}{c|ccc}
+      $+$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
+      $\gr0$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
+      $\gr1$ & $\gr1$ & $\gr\omega$ & $\gr\omega$ \\
+      $\gr\omega$ & $\gr\omega$ & $\gr\omega$ & $\gr\omega$ \\
+    \end{tabular}
+    \begin{tabular}{c|ccc}
+      $*$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
+      $\gr0$ & $\gr0$ & $\gr0$ & $\gr0$ \\
+      $\gr1$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
+      $\gr\omega$ & $\gr0$ & $\gr\omega$ & $\gr\omega$ \\
+    \end{tabular}
+    \begin{tikzpicture}[baseline]
+      \node(omega) at (0,0) {$\gr\omega$};
+      \node(0) [above left of=omega] {$\gr0$};
+      \node(1) [above right of=omega] {$\gr1$};
+
+      \draw (omega) -- (0);
+      \draw (omega) -- (1);
+    \end{tikzpicture}
+  }
+\end{example}
 
@@
 
+The rest of this chapter proceeds as follows.
 In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has

### 2022-08-01 tex/syntax.tex
[ writing ] content sections to full draft

@@
-We take the insights of the previous section and use them to build a
-generic framework for posemiring-annotated substructural systems in
-Agda. We will first show \emph{descriptions} of systems, which are
-comprised of rules that have premises combined using the bunched
-combinators. We then show how to construct the Agda data type of
-intrinsically well scoped, typed, and resourced terms for any given
-system of our framework. We use the prototypical system from
-\cref{fig:lr-comb} as a running example. \Cref{sec:other-syntaxes}
-presents further examples that our framework can express.
+The aim of this section is to produce a domain-specific language of
+\emph{syntax descriptions}.
+One may also use the terminology ``universe'' in place of ``domain-specific
+language'', alluding to the \emph{universe pattern} common in dependently typed
+programming~\citep{BDJ03}.
+A syntax description is essentially a concise, high-level representation of a
+type system's syntactic rules.
+The information contained in a syntax description is comparable to what is
+written (informally) in \cref{fig:lr-bunched}.
+The key features allowing these descriptions to capture semiring-annotated
+calculi are the distinction between sharing ($\dottimes$) and separating ($*$)
+conjunction of premises, modal scaling by a semiring element ($\cdot$), and the
+inclusion of semiring annotations on newly bound variables.
 
-We now start to use Agda notation for record and data type
-declarations, to emphasise that our framework has been implemented.
+%We take the insights of the previous section and use them to build a
+%generic framework for posemiring-annotated substructural systems in
+%Agda. We will first show \emph{descriptions} of systems, which are
+%comprised of rules that have premises combined using the bunched
+%combinators. We then show how to construct the Agda data type of
+%intrinsically well scoped, typed, and resourced terms for any given
+%system of our framework. We use the prototypical system from
+%\cref{fig:lr-comb} as a running example. \Cref{sec:other-syntaxes}
+%presents further examples that our framework can express.
+
+%We now start to use Agda notation for record and data type
+%declarations, to emphasise that our framework has been implemented.
 
@@
 
+I introduce syntax descriptions in three layers: \AgdaRecord{System},
+\AgdaRecord{Rule}, and \AgdaRecord{Premises}.
 A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
@@
 The \AgdaDatatype{Premise} data type describes premises of rules,
-using the bunched combinators from \cref{sec:bunched-rules}. A single
+using the bunched combinators from \cref{fig:bunched}. A single
 premise is introduced by the
@@
 
-As an example, we transcribe the system defined in
-\cref{fig:lr-comb} into a description.  We give the set of types of
+As an example, we transcribe a fragment of $\name$ (as defined in
+\cref{fig:lr} and \cref{fig:lr-bunched}) into a description.
+We give the set of types of
 this system as a data type \AgdaDatatype{Ty} (together with a base
@@
 posemiring \AgdaInductiveConstructor{Ann} in scope for the
-annotations.There is one label for each instantiation of a logical
+annotations.
+There is one label for each instantiation of a logical
 rule, but the labels contain no further information about subterms or
@@
 
-Compared to \cref{fig:lr-comb}, modulo the Agda notation, we can see
-that the fundamental structure has been preserved: the rules match
+Compared to \cref{fig:lr-bunched}, modulo the Agda notation, we can see
+that the fundamental structure has been preserved: The rules match
 one-to-one, and the bunched premises are the same. A major difference
@@
 example, the following term is needed to show that, in the $\{\gr0,
-\gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
+\gr1, \gr\omega\}$ (linearity) posemiring of \cref{def:lin-semiring},
 $\oc\gr\omega$ forms a comonad.
@@
 such a way that contexts capture the undistinguished parts of the sequent.
-As such, the generic substitution lemma we get in \cref{sec:kits} is the form
-of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
+As such, the generic substitution lemma we get in \cref{sec:kit-to-sem} is the
+form of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
 Though the $\mu\tilde\mu$-calculus is originally described as a sequent

### 2022-08-01 tex/wrel.tex
[ writing ] content sections to full draft

@@
-\chapter{Semantics in worldly relations}\label{sec:wrel}
+\chapter{Semantics in world-indexed relations}\label{sec:wrel}
+
+While most of the results in this thesis are syntactic in nature, it is
+important to make sure that our syntaxes can be interpreted in semantically
+interesting ways.
+The family of denotational semantics I chose to focus on is those based around
+\emph{world-indexed relations}.
+Such a semantics allows us to compare runs of a program on two different inputs
+(for example allowing us to prove monotonicity with respect to inputs) in a
+modal or resource-sensitive way (for example allowing us to prove that a program
+is a permutation of its inputs).
+The general semantics is intantiated in ways that make use of the particular
+semiring used in the syntax for usage restriction.
+
+This chapter can be seen as a refinement and development of the work of
+\citet{AbelBernardy2020}.
+The subtly different syntax I use with regards to pattern-matching
+eliminators~\citep{HMWO21} means that I require fewer and arguably more natural
+laws to hold of instantiations of the semantics.
+I also provide additional examples to those given by
+\citeauthor{AbelBernardy2020}.
+
+\section{Semantics in sets}\label{sec:set-sem}
+
+The eventual semantics based on world-indexed relations will offer theorems
+about the behaviour of programs.
+The behaviour we care about is just the extension of the program --- what result
+we get when semantic values are substituted in for the free variables.
+As such, a simple set-based semantics suffices.
+
+The set semantics is the same as one we could give to simply typed
+$\lambda$-calculus.
+Types are interpreted as sets, with function types interpreted as function sets
+and with sum types interpreted as disjoint unions.
+Both kinds of products --- with-products and tensor-products --- are interpreted
+as Cartesian products, while all $\oc\gr r$ modalities are ignored.
+These latter features show us that the set semantics is unaware of usage
+restrictions.
+
+% TODO: maybe figure
+\begin{displaymath}
+  \begin{array}{c@{\hspace{0.5in}}c}
+    \begin{eqns}
+      \sem{\iota} &=& X_\iota \\
+      \sem{\oc\gr rA} &=& \sem{A} \\
+      \sem{I} &=& \{*\} \\
+      \sem{\top} &=& \{*\} \\
+      \sem{0} &=& \{\} \\
+    \end{eqns}
+    &
+    \begin{eqns}
+      \\
+      \sem{A \multimap B} &=& \sem{A} \rightarrow \sem{B} \\
+      \sem{A \otimes B} &=& \sem{A} \times \sem{B} \\
+      \sem{A \with B} &=& \sem{A} \times \sem{B} \\
+      \sem{A \oplus B} &=& \sem{A} \uplus \sem{B} \\
+    \end{eqns}
+  \end{array}
+\end{displaymath}
+
+Contexts are always interpreted as iterated tensor products of their $\oc$-ed
+entries.
+In the case of this set semantics, this makes the interpretation of a context
+just the Cartesian product of its types, ignoring any usage annotations.
+Open terms are interpreted as functions from the interpretation of their context
+to the interpretation of their type.
+
+\section{Example results}
+
+The interesting part of the semantics is to prove properties about the set
+semantics based on the type of the program.
+These ``free theorems'' (in the terminology of \citet{Wadler89}) resemble the
+abstraction theorem for parametric polymorphism of \citet{Reynolds83}, but
+capture properties the abstraction theorem cannot due to the specificity of
+usage-aware typing.
+Before discussing technical details, I give some example results below to give a
+flavour of what we will be proving.
+
+\paragraph{Linearity}
+We should be able to prove, when we instantiate $\name$ with the
+$\{\gr0, \gr1, \gr\omega\}$ linearity semiring, that all programs are
+permutations, given that they cannot delete or duplicate values freely.
+Technically, all programs will be permutations in a designated base type,
+because deletion and duplication is derivable for any types made purely of
+logical type formers.
+To achieve the result, I will assign a ``weight'' to both the context and the
+result type of an open term, corresponding to how many values of base type that
+term or context needs to support it.
+I will then show that all term formers preserve this weight, yielding a
+permutation.
+
+\paragraph{Monotonicity}
+When we instantiate $\name$ with the $\{\gr{\sim\sim}, \gr{\uparrow\uparrow},
+\gr{\downarrow\downarrow}, \gr{\wn\wn}\}$ monotonicity semiring, we expect that
+all programs really are monotonic.
+Precisely, that means that we impose an ordering on the set denoted by a
+designated base type, extend this ordering to all types, then show that passing
+(the denotation of) a term bigger inputs makes it produce bigger outputs.
+To achieve this, we need to be able to relate two runs of a program.
+
+\paragraph{Sensitivity}
+When we instantiate $\name$ with the $\mathbb R^+$ sensitivity semiring, we want
+to show that all programs are non-expansive.
+Similarly to the case for monotonicity, the notion of distance is semantic,
+based on a distinguished base type whose denotation has a notion of distance.
+
+\section{World-indexed relations}
+
+Inspired by modal logic \todo{ref}, our relational semantics assumes a partially
+ordered set of worlds $\mathcal W$.
+The world keeps track of what resources we have allocated to which parts of the
+program.
+Our syntax tracks division of usages between subterms, and correspondingly we
+have a commutative monoid on $\mathcal W$ to model division of resources.
+The monoid structure respects the ordering, i.e., addition of worlds is
+monotonic.
+
+{\color{red}intuitively: the elements of $\mathcal W$ determine the
+  resources that we are talking about, the pre-order determines a
+  notion of ``swappability'', and the monoidal product determines what
+  it means for two resources to be combined.}
+
+\begin{example}[Trivial]\label{def:trivial-worlds}
+  The one element pre-ordered set $\{*\}$, with ${*} \leq {*}$ and
+  ${*} \otimes {*} = {*}$ gives a ``track nothing'' semantics of
+  resources. \todo{fwd ref the where this is used.}
+\end{example}
+
+\newcommand{\append}{\mathop{++}}
+
+\begin{example}[Lists and Permutations]\label{def:permutation-worlds}
+  Let $K$ be some set of values that will be used as keys for sorting.
+  I consider multisets over $K$ with finite support, denoted $\mathcal MK$,
+  where the multiplicity of each element is either a natural number or $\infty$.
+  These multisets represent the keys in scope at any particular point.
+  The order on $\mathcal MK$ is multiset equality, plus the fact that
+  $\lbag k^\infty \rbag \leq \lbag k^n \rbag$ for any finite $n$ (freely
+  extended to respect bag-addition).
+  The monoid structure is given by bag-addition.
+  The elements with multiplicity $\infty$ exist to have a sound interpretation
+  of $\oc\gr\omega$, with $\gr\omega$ coming from the
+  $\{\gr0, \gr1, \gr\omega\}$ semiring.
+  I am more interested in the elements with finite multiplicity, because in that
+  case the ordering corresponds to preservation of keys.
+
+  %Define $\mathcal{P}_K$ to be the pre-ordered set consisting
+  %of lists of elements of $K$, with $l_1 \leq l_2$ exactly when $l_1$
+  %is a permutation of $l_2$. The monoidal product is defined to be
+  %list concatenation: $l_1 \otimes l_2 = l_1 \append l_2$, with the
+  %empty list as the monoidal unit.
+
+  %\todo{explain the interpretation of the base type of keys}
+
+  %\todo{explain how this will be used in the $\{0,1,\omega\}$ example.}
+\end{example}
+
+\begin{example}[Security Principles]\label{def:privacy-worlds}
+  Lattice of sets of names of principles. In terms of resources, this
+  can be thought of as the collection of principles that are ``happy''
+  with this computation. \todo{still need to fully work out this
+    example}
+\end{example}
+
+\begin{example}[Distances]\label{def:sensitivity-worlds}
+  For sensitivity analysis, I take the partially ordered monoid
+  $(\mathbb R^+, \geq, 0, {+})$ of non-negative real numbers under the opposite
+  of the usual order.
+  We want to establish non-expansiveness of terms, which is why the order is
+  reversed.
+  The world represents the maximum permissible amount of deviation allowed in
+  the result when a unit change is made to the inputs.
+\end{example}
+
+\section{Semantics in world-indexed relations}
+
+I give each type former $F$ a semantic counterpart $F^R$.
+Where $F$ is an $n$-ary type former, $F^R$ is an operator taking $n$
+world-indexed relations and producing another world-indexed relation.
+The interpretation of any type is given recursively in the obvious way, with
+each base type interpreted as an arbitrary provided world-indexed relation.
+
+\subsubsection{Interpretation of $\multimap$ and $\otimes$}
+
+Given a preorder of semantic resources
+$(\mathcal{W}, \leq, \otimes, I)$, we have enough structure to
+interpret the function types ($A \multimap B$) and tensor product types
+($A \oplus B$) of \name{}.
+I define the following notations for connectives of world-indexed relations,
+and give interpretations of the type formers in terms of the notations.
+
+\begin{displaymath}
+  \arraycolsep=1.5pt
+  \begin{array}{rllcl}
+    I^R & w & (*,*) &\iff& w \leq I
+    \\
+    \plr{R \otimes^R S} & w & ((a,b),(a',b'))
+    &\iff& \exists x,y.~w = x \otimes y
+           \wedge R~x~(a,a')
+           \wedge S~y~(b,b')
+    \\
+    \plr{R \multimap^R S} & w & (f,f')
+    &\iff& \forall x.~\forall a,a'.~R~x~(a,a')
+           \Rightarrow S~(w \otimes x)~(f~a,f'~a')
+  \end{array}
+\end{displaymath}
+
+%If we assume we are given semantic interpretations
+%$\sem{A}^R : \mathcal{W}^{op} \to \operatorname{Rel}{\sem{A}}$ and
+%respectively for $B$, then we define the interpretations
+%$\sem{A \multimap B}^R$, $\sem{A \otimes B}^R$, and
+%$\sem{I}^R$ as follows:
+
+%\begin{align*}
+%  \sem{I}^R &\coloneqq I^R \\
+%  \sem{A \otimes B}^R &\coloneqq \sem A^R \otimes^R \sem B^R \\
+%  \sem{A \multimap B}^R &\coloneqq \sem A^R \multimap^R \sem B^R
+%\end{align*}
+
+The connective $\otimes^R$ is a form of Day convolution~\citep{Day70}.
+Using the multiset worlds of \cref{def:permutation-worlds}, $R \otimes^R S$ can
+be understood as a pointwise relation on pairs, but where the world (collection
+of keys) at which it is true has to be split up into the $R$ world and the $S$
+world.
+In the implication $R \multimap^R S$, $w$ is the multiset of keys provided by
+the function, $x$ is the multiset of keys supporting the argument, and thus all
+of the keys in $w \uplus x$ are in the result.
+Using the distance worlds of \cref{def:sensitivity-worlds}, we see that the
+sensitivity of a tensor product is the sum of the sensitivities of its two
+halves, and the sensitivity of the result of an application is similarly the sum
+of the sensitivity of the function and the sensitivity of the argument.
+
+\todo{Point out what happens when the monoidal product
+  is actually a join}
+
+\subsubsection{Interpretation of lists}
+
+As mentioned in section XXX, we effectively treat lists as tensored
+together. \todo{spell out what this entails, and relate it to the
+  permutations example}
+
+\subsubsection{Interpretation of $\oc\gr r$}
+
+The choice of interpretation of $\oc\gr r$ links the usage
+annotations present in \name{}'s typing contexts to the semantics.
+
+\begin{example}[Trivial]
+  The identity
+\end{example}
+
+\begin{example}[Linear Types]
+  $0$ is the identifies everything and forces the resource to be $I$;
+  $1$ is the identity; and $\omega$ forces the resource to be
+  duplicable.
+\end{example}
+
+\begin{example}[Monotonicity Annotations]
+  The worlds we use for monotonicity tracking are trivial, so $\oc^R$ is just
+  a transformer of homogeneous binary relations.
+  $\oc^R~\gr{\uparrow\uparrow}$ is the identity transformation, while
+  $\oc^R~\gr{\downarrow\downarrow}~R$ is $R^{\mathrm{op}}$, the flipped version of
+  $R$.
+  $\oc^R~\gr{\wn\wn}~R$ is the total relation, while
+  $\oc^R~\gr{\sim\sim}~R \coloneqq R \cap R^{\mathrm{op}}$.
+\end{example}
+
+\begin{example}[Distances]
+  Merp
+\end{example}
+
+\subsubsection{Interpretation of the additives: $\with$ and $\oplus$}
+
+Copy the clauses from below.
+
+\begin{displaymath}
+  \arraycolsep=1.5pt
+  \begin{array}{rllcl}
+    \sem{\top}^R & w & (*,*) &\iff& \top \\
+    \sem{A \with B}^R & w & ((a,b),(a',b')) &\iff&
+    \sem{A}^R~w~(a,a') \wedge \sem{B}^R~w~(b,b') \\
+    \sem{A \oplus B}^R & w & (\operatorname{inl} a,\operatorname{inl} a')
+                               &\iff& \sem{A}^R~w~(a,a') \\
+    \sem{A \oplus B}^R & w & (\operatorname{inr} b,\operatorname{inr} b')
+                               &\iff& \sem{B}^R~w~(b,b') \\
+  \end{array}
+\end{displaymath}
+
+\subsubsection{Assignment of Relations to Types}
+
+Let $\left| X \right|$ be the proposition that the set $X$ is
+inhabited.\todo{don't need this if we're just using a preorder.}
+
+\todo{put these in a figure to summarise them, and merge this
+  section with the one on the fundamental lemma}
+
+\begin{displaymath}
+  \arraycolsep=1.5pt
+  \begin{array}{rllcl}
+    \sem{\iota}^R & w & (c,c') &\iff& R_\iota~w~(c,c') \\
+    \sem{A \multimap B}^R & w & (f,f')
+                               &\iff& \forall x,y.~
+                                 \left| \mathcal W(x, y \otimes w) \right|
+                                 \Rightarrow \\
+    &&&& \forall a,a'.~\sem{A}^R~y~(a,a')
+                                 \Rightarrow \sem{B}^R~x~(f~a,f'~a') \\
+    \sem{\oc\gr rA}^R & w & (a,a') &\iff& \oc\gr r\sem{A}^R~w~(a,a') \\
+    \sem{I}^R & w & (*,*) &\iff& \left|\mathcal W(w, I)\right| \\
+    \sem{A \otimes B}^R & w & ((a,b),(a',b'))
+                               &\iff& \exists x,y.~
+                                 \left| \mathcal W(w, x \otimes y) \right|
+                                 \wedge \sem{A}^R~x~(a,a')
+                                 \wedge \sem{B}^R~y~(b,b') \\
+    \sem{\top}^R & w & (*,*) &\iff& \top \\
+    \sem{A \with B}^R & w & ((a,b),(a',b')) &\iff&
+    \sem{A}^R~w~(a,a') \wedge \sem{B}^R~w~(b,b') \\
+    \sem{A \oplus B}^R & w & (\operatorname{inl} a,\operatorname{inl} a')
+                               &\iff& \sem{A}^R~w~(a,a') \\
+    \sem{A \oplus B}^R & w & (\operatorname{inr} b,\operatorname{inr} b')
+                               &\iff& \sem{B}^R~w~(b,b') \\
+  \end{array}
+\end{displaymath}
+
+A context $\grR\Gamma$ is interpreted as a tensor product of all the
+$\oc^R\gr rA$ for each $\gr rA$ in $\grR\Gamma$.
+\todo{I use $\grR\gamma$ in other places, but this clashes with
+  $\gamma : \sem\Gamma$.}
+
+\begin{displaymath}
+ \begin{eqns}
+   \sem{\cdot}^R~w~(*,*) &=& \mathcal W(w, I) \\
+   \sem{\grR\Gamma, \gr rA}^R~w
+   ~((\gamma,a), (\gamma',a')) &=&
+   \sem{\grR\Gamma}^R \otimes \oc\gr r\sem A^R
+ \end{eqns}
+\end{displaymath}
+
+\subsubsection{Semantic Soundness}
+
+We now give the fundamental semantic soundness result for \name{},
+that states that the set-theoretic interpretation of every term
+(\autoref{sec:set-interp}) preserves the Kripke relational semantics
+of contexts and types described above. The proof of this theorem
+proceeds by induction on the typing derivation, and has been
+formalised in Agda \autoref{sec:agda-formalisation}.
+
+\begin{theorem}
+  If $\grR\Gamma \vdash t : T$, then for all worlds $w$ and
+  context interpretations $\gamma, \gamma' \in \sem{\Gamma}$, we have:
+  \begin{displaymath}
+    \sem{\grR\Gamma}^R w~(\gamma, \gamma') \implies \sem{T}^R w~(\sem{t}\gamma, \sem{t}\gamma')
+  \end{displaymath}
+\end{theorem}
+
+In the very general setting of this section, it is difficult to see
+the significance of this result. In the next section, we will
+specialise to instantiations of \name{} with specific usage
+semirings and specific semantic settings, which will allow us to
+derive concrete semantic results of interest.
+
+\section{Examples}

### 2022-11-06 tex/gen-syn.tex
[ writing ] lots of background, before feedback on later chapters

We have seen in previous sections a method for defining well typed terms,
providing them with the basic operations of renaming and substitution, and
defining type-preserving semantic traversals over those terms.
However, the Agda code we have seen only deals with one specific kind of terms
--- simply typed $\lambda$-calculus with a base type and function types.
The aim of this section is to write some code to which we can pass a
\emph{description} or \emph{signature} of a syntax and have it produce all of
the same machinery.

The description of a syntax will closely resemble the logical rules Gentzen gave
for natural deduction systems NJ and NK, but we give them a revised
interpretation.
Where Gentzen intended his rules to be applied schematically, and hypothetical
proofs to be handled via \emph{discharge} of hypotheses, we will take the rules
formally to produce a system with explicit contexts and a variable rule.
However, knowing that this resulting system came from such a description means
that we can derive variable-handling features, such as substitution, in a
generic way.

I will present a scheme such that \cref{fig:app-lam} is interpreted as the type
system we studied in the previous sections (simply typed $\lambda$-caluclus with
a base type and function types).
Remember that, while these look like inference rules, I am treating them
entirely formally, collected together into a \emph{syntax description}.
The information presented in \cref{fig:app-lam} is essentially all of the
information needed for the type system sans any details about variables.
In particular, notice:
\begin{itemize}
  \item Contexts, in particular the context of a rule's conclusion, which is
    shared in all premises in the resulting type system, are elided.
    The only part of any context I record is the newly bound variables in
    premises, such as the variable bound by a $\lambda$-abstraction.
  \item There is no explicit variable rule.
    It is understood that any $x : A$ in the context of the resulting type
    system can be used to yield a term with type $A$.
\end{itemize}

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\vdash A \to B}
      \hypo{\vdash A}
      \infer2[app]{\vdash B}
    }
    \and
    \ebrule{%
      \hypo{A \vdash B}
      \infer1[lam]{\vdash A \to B}
    }
  \end{mathpar}
  \caption{An example syntax description}
  \label{fig:app-lam}
\end{figure}

Such a scheme commits us to a certain approach to variable binding and context
management, but does not commit us to anything about the meaning of types.
For example, we do not declare that \TirName{app} and \TirName{lam} are
``elimination'' and ``introduction'' forms for the function type former.
This limits our generic results to matters of syntax and variables, but provides
a platform upon which a future semantic scheme could rest.

\begin{figure}
  \begin{align*}
    \text{Premises} && \mathit{ps}, \mathit{qs} &\Coloneqq
      \Delta \vdash A \mid {} \mid \mathit{ps} \quad \mathit{qs}
    \\
    \text{Rule} && r, s &\Coloneqq {\mathit{ps} \over \vdash A}
  \end{align*}
  \caption{The grammar of typing rules}
  \label{fig:simple-syntax}
\end{figure}

A syntax description is a set of fully instantiated rules.
In our running example, this set contains a \TirName{app}-rule and a
\TirName{lam}-rule for each pair of types $A$ and $B$.

To construct the syntax given by a description, we keep
\AgdaInductiveConstructor{var} as before, and have another constructor
\AgdaInductiveConstructor{con} for all of the logical rules.
\AgdaInductiveConstructor{con} takes a rule \AgdaBound{r} with premises
$\Delta_1 \vdash A_1; \ldots; \Delta_n \vdash A_n$ and conclusion $A$, and the
remainder of its type is as follows.
\[
  \AgdaInductiveConstructor{con}~\AgdaBound{r}~\AgdaSymbol{:}~%
  \forall\Gamma.~(\Gamma, \Delta_1 \vdash A_1) \times \cdots
  \times (\Gamma, \Delta_n \vdash A_n) \to \Gamma \vdash A
\]
Note that, in this type, $\vdash$ is the type family of terms we are
inductively constructing, as opposed to the description syntax found in the
premises.

In our generic version of \AgdaRecord{Semantics}, we keep the
\AgdaField{ren\textasciicaron$\V$} and \AgdaField{$\llbracket$var$\rrbracket$}
fields as before, and replace \AgdaField{$\llbracket$app$\rrbracket$} and
\AgdaField{$\llbracket$lam$\rrbracket$} by a
\AgdaField{$\llbracket$con$\rrbracket$} field as follows.
\[
  \AgdaField{$\llbracket$con$\rrbracket$}~\AgdaBound{r}~\AgdaSymbol{:}~%
  \forallb{%
    \Box\plr{{} \env\V \Delta_1 \dotto {} \sdtstile{}\C A_1} \dottimes
    \cdots \dottimes
    \Box\plr{{} \env\V \Delta_n \dotto {} \sdtstile{}\C A_n} \dotto
    {} \sdtstile{}\C A}
\]
I use ${} \sdtstile{}\C A$ for the Agda notation
\AgdaFunction{\_$\C\vDash$}\AgdaSpace{}\AgdaBound{A}, while ${} \env\V \Delta$
stands for the type family of environments
\AgdaSymbol{$\lambda$}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaFunction{Env}\AgdaSpace{}\AgdaBound{$\V$}%
\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\Delta$}.
Environments appear in this definition simply as a way to write a product of
$\V$-values --- one value for each element of $\Delta$.
We could make a special case of premises which do not bind any variables, as did
\citet{AACMM21}, eliding the $\Box$ and empty environment, but I choose not to
for uniformity and simplicity of presentation.

### 2022-11-06 tex/kits.tex
[ writing ] various thesis additions
[ build ] get thesis into building state
[ writing ] lots of background, before feedback on later chapters

@@
+\def\SimpleKits{../agda/processed-latex/SimpleKits.tex}
+
+%Explain:
+%
+%\begin{itemize}
+%  \item Specific uses of renaming/substitution in $\lambda$-calculus semantics.
+%  \item General role of renaming/substitution in abstract algebra/syntax with
+%    binding.
+%\end{itemize}
+
+A basic operation on any syntax with variables is \emph{substitution} --- the
+replacement of variables in a term by terms with the same type as the variables.
+In a sense, this is the defining operation of variables --- a variable is a
+placeholder for a term, or equivalently in logic, a hypothesis is a placeholder
+for an arbitrary proof.
+In a type theory or logic, terms can bind variables, and we will typically have
+operational semantics rules combining a term binding a variable with a term that
+is to be substituted into the place of that variable, like the $\beta$-rule for
+$\lambda$-calculus functions.
+
+While substitution has this extra role in a lot of the syntaxes with binding we
+care about, variable-binding also significantly complicates the substitution
+operation.
+Substitution acts on the free variables of a term, replacing them by terms, but
+binders mean that some subterms have \emph{more} free variables than our
+original term.
+This causes different challenges for different representations of terms.
+For example, with named variables and shadowing, na\"{i}vely defined
+substitution could fall foul of variable capture.
+In our approach, based on de Bruijn indices, the difficulty is that an index $i$
+outside a binder of $n$ variables corresponds to an index $n + i$ inside the
+binder.
+Therefore, when substituting under a binder, we must first increment any free
+variables contained in terms we are substituting in, which is a form of
+\emph{renaming}.
+Renaming replaces each free variable by another free variable, and is a special
+case of substitution.
+We must, however, define renaming before substitution, so as to avoid the
+definition of substitution being circular.
+Renaming avoids a similar circularity because when renaming goes under a binder,
+we only have to increment each variable being renamed in, rather than each
+variable \emph{in each term} being substituted in.
+
+In this section, I formally implement simultaneous renaming and substitution for
+the terms defined in the previous section.
+Simultaneous substitution turns out to have a simple definition, which
+generalises into other algorithms over terms with binders.
+The section concludes with a unified implementation of renaming and
+substitution, leaving further generalisation to the next section.
+
 \subsection{Simultaneous renaming and simultaneous substitution}

### 2022-11-06 tex/linmech.tex
[ writing ] lots of background, before feedback on later chapters

@@
 
-\subsection{yalla}
+\subsection{Yalla}
+
+\Citet{laurent18} presents a collection of linear logics in a uniform style, and
+various proofs relating them.
+The logics share varying amounts of definitions and theorems --- for example,
+the main linear logic is parametrised on whether to include mix rules and
+whether to restrict exchange to cyclic permutations, whereas systems like the
+Lambek calculus (with no exchange) and polarised linear logic are defined
+separately from scratch.
+
+The style used in Yalla is to realise sequents as lists of formulas.
+The active formula tends to be forced to be the first formula in such a list,
+with an explicit exchange rule being used to move such formulas into place.
+\Citet{laurent18} explains why explicit permutations are necessary in such a
+system, and why using multisets instead of lists would conflate distinct proofs
+when there are repeated formulas in a sequent.
+The system I will present in the next chapter avoids permutations essentially by
+being a refinemet of the system presented in the previous chapter, which has an
+ordered context and specific references (de Bruijn indices) into it.
+
+Despite making sure to distinguish distinct proofs, the Yalla library does not
+define any equational theories of proofs, so does not prove any results relying
+on these distinctions.
+While the complication of defining an equational theory in the presence of an
+exchange rule is probably largely inevitable, having the exchange rule
+introduces redundancy such that many equivalent proofs are not equal as data
+structures.
+The mechanisations presented in \cref{sec:simple} all sought to reduce this kind
+of redundancy for intuitionistic logic, so that the only non-trivial equations
+in the equational theory are $\beta$- and $\eta$-rules (i.e.\ computationally
+interesting rules).
+I will restore this property of the representation in \cref{sec:semirings}.
+
 \begin{itemize}

### 2022-11-06 tex/mech-related.tex
[ writing ] lots of background, before feedback on later chapters

There is a vast literature on formalisations of syntaxes with binding, which I
cannot possibly do justice to in a reasonably sized thesis chapter.
Instead, I limit myself to comparisons of the \citet{AACMM21} method I follow in
this thesis to just its closest related work.

\subsection{Autosubst}

\citet{Autosubst15} present the system \emph{Autosubst}, which provides various
tools for working with syntaxes with binding in the Coq proof assistant.
Autosubst is based on similar ideas to those \citeauthor{AACMM21} use:
de Bruijn-indexed terms with a distinguished variable rule and notion of
binding, acted upon by simultaneous renaming and substitution.

The simplest differences are essentially matters of choosing the encoding that
best fits the proof assistant being used.
Coq users tend to prefer using unindexed types and propositions indexed over
them --- in this case, a type of unscoped and untyped terms plus a
``well typed'' predicate --- whereas Agda users prefer to work with only well
formed data (well scoped and well typed terms).
The latter approach more readily allows us to show generically that substitution
preserves scoping and typing, but the former approach, conversely, allows for
bespoke proofs of such facts.
For example, one theorem of \citet{Autosubst15} is type preservation for
$\mathrm{CC}_\omega$, a dependent type system we cannot express using the
machinery of \citet{AACMM21}.
In principle, one could use \citeauthor{AACMM21}'s machinery as the basis of a
similar bespoke proof, but as far as I am aware, this has not been tried.

Another main difference is that Autosubst is presented to the user largely as
a black-box implementation of substitution and related lemmas, in contrast to
\citeauthor{AACMM21}'s work exposing the \AgdaRecord{Semantics} bundle to the
user, and having substitution be just one instance.
\citet{ACMM17} and \citet{AACMM21} provide many examples of traversals over
syntax using the same generic environment management as used by substitution.
However, the focus on substitution in Autosubst has meant that reasoning about
substitutions has been given more developed support.
For example, the library provides a tactic \texttt{autosubst} which automates
many equational proofs involving substitutions based on the $\sigma$-calculus of
\citet{ACCL91}.

An interesting feature of Autosubst is \emph{heterogeneous substitution}.
The motivation for heterogeneous substitution is to handle systems like system
F, where types and terms are syntactically distinct, but both feature binding
and require a substitution operation.
Furthermore, binding and substitution of types also affects the syntax of terms,
thanks to $\Lambda$ terms.
\citeauthor{AACMM21} provide no direct equivalent to heterogeneous substitution,
and it is unclear how well their work can handle polymorphic calculi.

\citet{Autosubst18} propose some modifications to Autosubst which, as far as I
can tell, have not yet been incorporated, but are presented in mechanised form
for the paper.
Some of these modifications aim to bring Autosubst into line with
\citeauthor{AACMM21}'s work, in particular taking semantic traversals as a
theoretical basis.
TODO: more.

\subsection{Fiore, Plotkin, Turi}

TODO: cite \citet{FPT99} and \citet{FS22}.

### 2022-11-06 tex/prelim.tex
[ writing ] content sections to full draft
[ writing ] lots of background, before feedback on later chapters

@@
 \chapter{Mathematical preliminaries}
+
+In this section, I recall some standard categorical and algebraic notions that
+are used in the rest of this thesis.
+I present most of these notions in terms of \emph{multicategories}, whose main
+purpose I see as giving universal properties to monoidal products --- for
+example, vector spaces and multilinear maps form a multicategory, and in this
+multicategory we can state the universal property of the tensor product of
+vector spaces.
+However, a reader already familiar with monoidal categories presented some other
+way may wish to ignore the parts about multicategories and just skim the other
+definitions to make sure they seem familiar.
 
@@
 Examples: Set, Rel, partial functions, commutative monoids.
+
+\section{Categories and multicategories}
 
@@
       $A_1 \times \cdots \times A_n \to B$)
-    \item $K$-vector spaces and multilinear maps
+    \item $K$-vector spaces and multilinear maps, for any field $K$
     \item For any simple type theory with sequents of the form
@@
   \begin{itemize}
-    \item $\obj(\C^*) \coloneqq \List(\obj\C)$
+    \item $\obj(\C^*) \coloneqq \List(\obj\C)$.
     \item $\C^*\plr{\sum_i \Psi_i, \Delta} \coloneqq \prod_i \C(\Psi_i; \Delta_i)$
-      for any $\Psi : \List(\List(\obj\C))$
-    \item $\id^* : \Gamma \to^* \Gamma \coloneqq \overline\id$
+      for any $\Psi : \List(\List(\obj\C))$.
+    \item $\id^* : \Gamma \to^* \Gamma \coloneqq \overline\id$.
     \item $(\sigma : \Delta \to^* \Theta) \circ^* (\rho : \Gamma \to^* \Delta) :
-      (\Gamma \to^* \Theta) \coloneqq \lambda i.~\sigma_i \circ \lambda j.~\rho_{ij}$
-      \todo{Still wrong}
+      (\Gamma \to^* \Theta) \coloneqq \lambda i.~\sigma_i \circ \rho|_{\sigma,i}$,
+      where $\rho|_{\sigma,i}$ is the restriction of $\rho$ to those objects
+      in the domain of $\sigma_i$.
   \end{itemize}
@@
 \begin{definition}
-  TODO: \emph{symmetric monoidal category}
+  A \emph{symmetric multicategory} is a multicategory $\C$ together with, for
+  each permutation $\sigma : \Gamma \tilde\to \Delta$, an action
+  $\sigma(-) : \C(\Delta; A) \to \C(\Gamma; A)$ respecting identity and
+  composition of permutations.
+\end{definition}
+
+\begin{remark}
+  It is tempting to define symmetric multicategories by replacing the lists of
+  objects in the definition of multicategories by finite multisets.
+  However, doing so entirely forgets the order of inputs to an operation, for
+  example forcing every morphism with a type of the form $A, A \to B$ to be
+  commutative.
+  There is a way to refine this idea to get a correct definition, at the cost of
+  requiring more sophisticated higher categorical notions.
+\end{remark}
+
+\begin{definition}
+  A \emph{symmetric monoidal category} is a representable symmetric
+  multicategory, analogously to how I defined a monoidal category.
 \end{definition}
@@
 \begin{definition}
-  TODO: \emph{commutative monoid in a symmetric monoidal category}
+  A \emph{commutative monoid} in a symmetric multicategory is a monoid in the
+  underlying multicategory such that $\mathrm{swap}(\mu) = \mu$, where
+  $\mathrm{swap} : M, M \tilde\to M, M$ is the permutation swapping two
+  elements.
 \end{definition}
 
-Algebra
+\section{Algebra}
 

### 2022-11-06 tex/terms.tex
[ writing ] lots of background, before feedback on later chapters

@@
 
-\missingfigure{LTree}
+\ExecuteMetaData[\LTreetex]{LTree}
 \Ctx{}
 
-We start with a data type \AgdaDatatype{\_⊢\_} of intrinsically simply typed
-terms.
-Beside base types, the only type former we have is the function type constructor
-\AgdaInductiveConstructor{\_`→\_}.
-Contexts (of type \AgdaRecord{Ctx}) are implemented as the free magma on types
-(\AgdaDatatype{Ty}).
-Context concatenation is \AgdaFunction{\_++ᶜ\_}, and \AgdaFunction{[\_]ᶜ}
-embeds types into contexts.
-Typed variables in a context are given by \AgdaRecord{\_∋\_}.
-A variable in
+Our first data structure involving contexts is that of intrinsically typed
+variables.
+A variable of type
 \AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaRecord{$\ni$}\AgdaSpace{}\AgdaBound{A}
@@
 with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
-Variables embed into terms via the \AgdaInductiveConstructor{var} constructor.
+Variables embed into terms via the \AgdaInductiveConstructor{var} constructor of
+the family \AgdaDatatype{\_⊢\_} of intrinsically simply typed terms.
+The only other syntactic forms we consider for now are the eliminator and
+constructor of function types \AgdaInductiveConstructor{\_`→\_} ---
+\AgdaInductiveConstructor{app} and \AgdaInductiveConstructor{lam}.
+Application \AgdaInductiveConstructor{app} takes two subterms of the appropriate
+types, while the subterm of $\lambda$-abstraction \AgdaInductiveConstructor{lam}
+is in an extended context \GA{} --- \AgdaBound{$\Gamma$} concatenated with a
+singleton context containing the type \AgdaBound{A}.
 

### 2022-11-18 tex/lnd.tex
[ format ] add comb proof style
[ writing ] various thesis additions
[ more ] thesis writing
[ edit ] some corrections and notes after meeting

@@
 
-The typing rules presented in the previous section contain a lot of detail and
-repeated patterns.
+The typing rules of $\name$ presented in \cref{fig:lr} contain a lot of detail
+and repeated patterns.
 For example, nearly half of the rules include the premise
@@
 To encapsulate the repeated patterns and facilitate an implicit context style,
-I introduce the \emph{bunched connectives}.
+I introduce the \emph{bunched connectives} for premises.
 These are inspired by bunched logic~\cite{oHP99}, and will not only be used for
@@
     I^*\,\grR &\coloneqq \grR \leq \gr0 \\
-    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~ \plr{\grR \leq \grP + \grQ}
+    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
                        \times T\,\grP \times U\,\grQ \\
-    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~ \plr{\grR \leq \gr r\grP}
+    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
                        \times T\,\grP
@@
 
-The simplest connectives are those we've already seen for intuitionistic
+The first three connectives are those we've already seen for intuitionistic
 systems --- $\dot1$, $\dottimes$, and $\dotto$.
@@
   &\quad\rightsquigarrow\quad
-  \begin{prooftree}
-    \set{separation=0.75em}
+  \begin{prooftree}[comb]
     \hypo{\dot1}
@@
   &\quad\rightsquigarrow\quad
-  \begin{prooftree}
-    \set{separation=0.75em}
+  \begin{prooftree}[comb]
     \hypo{\vdash A}
@@
 The prototypical typing rules for each of these three connectives are the
-introduction rules for $I$, $\otimes$, and $\oc_{\gr r}$, respectively.
+introduction rules for $I$, $\otimes$, and $\oc{\gr r}$, respectively.
 
@@
   &\quad\rightsquigarrow\quad
-  \begin{prooftree}
-    \set{separation=0.75em}
+  \begin{prooftree}[comb]
     \hypo{I^*}
@@
     \hypo{\grR \leq \gr r\grP}
-    \infer2{\grR\Gamma \vdash \oc_{\gr r} A}
+    \infer2{\grR\Gamma \vdash \oc{\gr r} A}
   \end{prooftree}
   &\quad\rightsquigarrow\quad
-  \begin{prooftree}
-    \set{separation=0.75em}
+  \begin{prooftree}[comb]
     \hypo{\gr r \cdot (\vdash A)}
-    \infer1{\vdash \oc_{\gr r} A}
+    \infer1{\vdash \oc{\gr r} A}
   \end{prooftree}
@@
 %For example, linearly we have
-%$\oc_{\gr r}(A \with B) \simeq \oc_{\gr r} A \otimes \oc_{\gr r} B$, while the
+%$\oc{\gr r}(A \with B) \simeq \oc{\gr r} A \otimes \oc{\gr r} B$, while the
 %expressions $\gr r \cdot (A \dottimes B)$ and $\gr r \cdot A \sep \gr r \cdot B$
@@
 We will also see later\todo{Forward reference} that, while
-$(\dottimes, \dotto)$ gives a Cartesian
+$(\dottimes, \dotto)$ gives a Cartesian \todo{wrong}
 closed structure, $\sep$ is monoidally closed, with an internal hom $\wand$.
@@
 various subterms.
+\todo{Bunched type system to Agda}
+\todo{Split chapter here. Play around with other systems using bunched
+  connectives. Lists/recursion, annotated arrow, Granule elim, sharing elim.}
 
@@
         \hypo{\gr r \cdot (\vdash A)}
-        \infer1[$\oc$-I]{\vdash \oc_{\gr r} A}
+        \infer1[$\oc$-I]{\vdash \oc{\gr r} A}
       \end{prooftree}
@@
       \begin{prooftree}
-        \hypo{\vdash \oc_{\gr r} A}
+        \hypo{\vdash \oc{\gr r} A}
         \hypo{\sep}

### 2022-11-18 tex/lr.tex
[ edit ] some corrections and notes after meeting

@@
 assumptions between the two subterms, in \name{}, this splitting is done by
-choosing usage annotations which add up to the usage annotations of the
-conclusion.
+choosing usage annotations for the premises which add up to the usage
+annotations of the conclusion.
 For example, we can derive $\vdash A \otimes B \multimap B \otimes A$ as
@@
   When the usage annotations of $\name$ are taken from this trivial posemiring,
-  we recover a version of intuitionistic simply typed $\lambda$-calculus.
+  we recover a version of intuitionistic simply typed $\lambda$-calculus
+  featuring redundant connectives $\otimes$ (equivalent to $\with$ in the pure
+  setting) and $\oc{\gr*}$ (where $\oc{\gr*} A \simeq A$).
 \end{example}

### 2022-11-18 tex/lrkits.tex
[ edit ] some corrections and notes after meeting

@@
 \begin{definition}[Simple environment]\label{def:simple-env}
-  A $\V$-\emph{environment} between simply typed contexts $\Gamma$ and $\Delta$
+  For $\V : \mathrm{Ctx} \to \mathrm{Ty} \to \mathrm{Set}$,
+  a $\V$-\emph{environment} between simply typed contexts $\Gamma$ and $\Delta$
   is a function, polymorphic in type $A$, from variables of type $A$ in
@@
 
-The first thing to note is that this definition is inadequate for \name{}.
+This definition is inadequate for \name{}.
 For example, suppose we have a term
@@
 To see why this definition of environment works for simply typed
-$\lambda$-calculus but not \name{}, I produce an equivalent definition by
+$\lambda$-calculus but not \name{}, let us look at an equivalent definition by
 recursion on the target context.
@@
 vectors is less than or equal to the initial usage vector.
-In symbols, we may let $\gr\Psi$ be the collection of leaf vectors, and state
+In symbols, we may let $\gr\Psi$ be the collection of leaf vectors indexed by
+items in $\Delta$, and state
 the constraint as $\grP \leq \sum_{\plr{x : \gr rA} \in \Delta} \gr r\gr\Psi_x$.
@@
 concepts like non-usage-checked variables and rows of a matrix.
-We may note that $\gr\Psi_x = \bra x\gr\Psi$, from which point, requiring not
+We note that $\gr\Psi_x = \bra x\gr\Psi$, from which point, requiring not
 just $\V\,\gr\Psi_x\gamma\,A$ but rather $\V\,\plr{\grQprime\gr\Psi}\gamma\,A$
@@
 All of these changes yield my primary definition of an environment for
-usage-annotated calculi.
+usage-annotated calculi, which will be used for the rest of this chapter and in
+\cref{sec:framework}.
 

### 2022-11-18 tex/semirings.tex
[ edit ] some corrections and notes after meeting

@@
 applicable to variables of any type, while essentially retaining linearity
-(as opposed to affinity).
+(as opposed to affineness).
 
@@
 removing variables from the context of certain subterms, add an annotation to
-free variables saying whether they are to be used or not.
-We may use an annotation $\gr0$ on variables that are not to be used, and an
+free variables saying whether or not they are to be used.
+I use an annotation $\gr0$ on variables that are not to be used, and an
 annotation $\gr1$ on variables that are to be used.
 This convention lets us transcribe the usual $\otimes$-introduction rule
-(left) as a rule with usage annotations (right).
-I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context now
-looks like $\grP\gamma$, where $\grP$ is a list of usage annotations
-$\gr{r_1}, \ldots, \gr{r_m}$ and $\gamma$ is a list of types $A_1, \ldots, A_m$
-of the same length.
+(below left) as a rule with usage annotations (below right).
+In the notation on the right, I let $\Gamma = \grP\gamma$ and
+$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
+\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
+A usage context is a list of usage annotations, so
+$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
+$\gamma = A_1, \ldots, A_m$.
+When combined, the usage context and the typing context will be of the same
+length.
 Explicit contexts will usually be written with usage annotations and types
@@
 
-What the existence of $\gr0$ has bought us is that variables never go out of
-scope in subterms; rather, we lose the ability to use certain variables that
-remain in scope.
+The use of $\gr0$ gives us the property that variables never go out of
+scope in subterms; rather, we lose the ability to use certain variables, but
+retain the ability to refer to them metatheoretically.
 Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
@@
 variables from $\delta$.
-A rigorous proof of weakening for terms will come in \cref{sec:lrsub}.
+I prove the admissibility of weakening for terms will come in \cref{sec:lrsub}.
 
@@
 To conceptualise the criteria on the usage annotations involved in this rule,
-I introduce a formal $+$ operation on usage annotations.
+I introduce an additive structure over usage annotations.
 The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
@@
 must have annotation less than or equal to $\gr0$.
-Thus the basis vector $\langle x \rvert$ describes the upper bound on the
-current usage context $\grR$.
+I write this requirement as $\grR \leq \langle x \rvert$, where
+$\langle x \rvert$ is the \emph{basis vector} at position $x$.
 
@@
 
-The final interesting rule form to cover is that found in DILL's
+The final interesting rule form to cover is found in DILL's
 $\oc$-introduction rule.
-I choose to see DILL's $\oc$-introduction as an $\gr\omega$-ary counterpart to
+DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
 $\otimes$-introduction, though with the same premise each time rather than
@@
 multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
-$0$ (for a variable being discarded), and an ordering (allowing for subsumption
+$0$ (for a variable not being used), and an ordering (allowing for subsumption
 of usage restrictions).
@@
 
-\section{A usage-annotated calculus}\label{sec:lr}
+\section{A usage-annotated calculus $\name$}\label{sec:lr}
 \input{lr.tex}

### 2022-12-12 tex/lenv.tex
[ edit ] some corrections and notes after meeting
[ edit ] new draft of chapters 5 and 6

@@
+I settle on \cref{def:lr-env}, and prove various properties about it.
+
 \begin{lemma}\label{thm:env-resize}
@@
 
-We could, indeed, use these three clauses to define what an environment is.
+We could, as in \cref{def:lr-rec-env}, use these three clauses to define what an
+environment is.
 However, I find them difficult to work with, as it is often easier to do
 linear algebraic proofs separately from the rest of an environment.
-For identity and composition, as we are about to see, the original definition
-is easier to use because we can rely on the identity and composition of linear
-maps.
-Concretely, an inductive proof of identity would, for example, involve
-constructing an environment of type
+For example, for identity and composition of environments, as we are about to
+see, our working definition (\cref{def:lr-env})
+is easier to use because we can rely on the closure of linear maps under
+identity and composition.
+Concretely, an inductive construction of, for example, an identity environment
+would involve constructing an environment of type
 $\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
@@
 induction hypothesis.
+Similar strengthening of the induction hypothesis would be required for
+composition.
 
 One of the primary test cases for environments is simultaneous substitution,
-which will look like the following rule.
-The admissibility of substitution will be by induction on the derivation of
-$\Delta \vdash A$, so we will need to be able to adapt any environment we are
-given to work with any possible context of new premises.
-In the simply typed case, the only change to the context we encountered was the
-binding of new variables.
-Now, with usage annotations, we furthermore have linear decompositions of the
-context, necessitating changes to the environment whenever usage annotations
-change.
-I will deal first with linear decompositions.
+which will look like the \TirName{sub} rule below.
+Note that we have taken $\V \coloneqq {\vdash}$ --- i.e.\ that the values
+yielded by the environment are terms, namely the terms to be substituted in for
+the free variables of the derivation of $\Delta \vdash A$.
 
@@
 
+The admissibility of substitution will be by induction on the derivation of
+$\Delta \vdash A$, so we will need to be able to adapt any environment we are
+given to work with any possible context of new premises yielded by the rules of
+\cref{fig:lr-bunched}.
+In the simply typed case, the only change to the context we encountered was the
+binding of new variables.
+With usage annotations, we furthermore have linear decompositions of the
+context, necessitating changes to the environment whenever usage annotations
+change.
+
 There are three kinds of linear decompositions we have to deal with: zero,
@@
 $\gr r \cdot {}$, respectively.
-In each case, we have a simple preservation lemma, transforming an environment
+In each of these three cases, we have a simple preservation lemma, transforming
+an environment
 of type $\Gamma \env\V \Delta$ and a decomposition of $\Delta$ into a
@@
 \begin{lemma}[environments preserve zero]\label{thm:lr-env-zero}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ such that
   $\grQ \leq \gr 0$, we also have that $\grP \leq \gr 0$.
@@
   $\grP \leq \grQ\gr\Psi \leq \gr0\gr\Psi = \gr0$, by environment
-  compatibility and monotonicity and linearity of $\gr\Psi$.
+  compatibility from $\rho$ and monotonicity and linearity of $\gr\Psi$.
 \end{proof}
@@
 \begin{lemma}[environments preserve addition]\label{thm:lr-env-add}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ such that
   $\grQ \leq \grQl + \grQr$ for some $\grQl$ and $\grQr$, we also have $\grPl$
   and $\grPr$ such that $\grP \leq \grPl + \grPr$ and there are environments
-  of types $\grPl\gamma \env\V \grQl\delta$ and
-  $\grPr\gamma \env\V \grQr\delta$.
+  $\rho_l : \grPl\gamma \env\V \grQl\delta$ and
+  $\rho_r : \grPr\gamma \env\V \grQr\delta$.
 \end{lemma}
@@
   Because clearly $\grPl \leq \grQl\gr\Psi$ and $\grPr \leq \grQr\gr\Psi$,
-  \cref{thm:env-resize} on the original environment gives us the required
-  pair of new environments.
+  applying \cref{thm:env-resize} to $\rho$ gives us the required
+  new environments $\rho_l$ and $\rho_r$.
 \end{proof}
@@
 \begin{lemma}[environments preserve scaling]\label{thm:lr-env-scale}
-  Given an environment of type $\grP\gamma \env\V \grQ\delta$ such that
+  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ such that
   $\grQ \leq \gr r\grQprime$ for some $\grQprime$, we also have a $\grPprime$
-  such that $\grP \leq \gr r\grPprime$ and there is an environment of type
-  $\grPprime\gamma \env\V \grQprime\delta$.
+  such that $\grP \leq \gr r\grPprime$ and there is an environment
+  $\rho' : \grPprime\gamma \env\V \grQprime\delta$.
 \end{lemma}
@@
   Because clearly $\grPprime \leq \grQprime\gr\Psi$,
-  \cref{thm:env-resize} on the original environment gives us the required
-  new environment.
+  applying \cref{thm:env-resize} to $\rho$ gives us the required
+  new environment $\rho'$.
 \end{proof}
 
-Finally, I will also take the opportunity to give the bind lemma, allowing
-environments to incorporate newly bound variables.
-In the intuitionistic case, the bind lemma had two requirements on $\V$: $\V$
-admits weakening and we can map variables into $\V$-values.
-With usage annotations, the former is unreasonable, but it turns out that we
-only need weakening by variables whose usage annotation is less than or equal
-to $\gr0$.
-The latter stays as-is, with the note that ``variable'' now means a
-usage-checked variable.
+The final change environments need to preserve is the binding of new free
+variables.
+In \cref{sec:syntactic-kits}, we had the operation \AgdaFunction{bindEnv} for
+this purpose in the intuitionistic setting.
+There, we relied on $\V$ supporting a map from $\ni$-variables and admitting
+weakening.
+In the usage-annotated setting, the former requirement is updated to having a
+map from usage-checked $\sqni$-variables.
+As for the latter requirement, it turns out that we only need $\V$ to admit
+weakening by $\gr0$-annotated variables, which is much more reasonable than
+general weakening.
+\Cref{thm:lr-bind} adapts \AgdaFunction{bindEnv} for the usage-annotated
+setting.
 
-\begin{lemma}[bind]\label{thm:lr-bind}
+\begin{lemma}[\AgdaFunction{bindEnv}]\label{thm:lr-bind}
   Given functions
@@
 
-The requirements for identity and composition of environments look a bit like
-the unit and lift of a Kleisli triple.
-\todo{Change from equality to inequality}
+I put together the preceeding pieces to give a syntactic traversal operation over
+$\name$ in the following section.
+For the rest of this section, I observe some more constructions purely on
+environments --- in particular, composition of environments given certain
+assumptions about the families of values.
 
-\begin{lemma}[Identity environment]
-  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
-  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
+Following \citet{ACU15}, we expect (intuitionistic) ST$\lambda$C syntax to form
+a relative monad over $\ni$ seen as a functor from the category of contexts
+under renaming to the functor category $\blr{\mathrm{Ty}, \Set}$, where
+$\mathrm{Ty}$ is the discrete category of ST$\lambda$C types.
+Notice that, given $F,G : \blr{\mathrm{Ty}, \Set}$, a morphism from $F$ to $G$
+is a function of type $\forallb{F \dotto G}$ (with naturality being trivial).
+Therefore, we expect a relative monad, given as a Kleisli triple, to have a unit
+$\eta_\Gamma : \forallb{\Gamma \ni \plr{-} \dotto \Gamma \vdash \plr{-}}$
+given by the variable rule, and a Kleisli extension operator
+${^*}_{\Gamma,\Delta} :
+\forallb{\Gamma \ni \plr{-} \dotto \Delta \vdash \plr{-}} \to
+\forallb{\Gamma \vdash \plr{-} \dotto \Delta \vdash \plr{-}}$
+given by substitution.
+Composition of substitutions falls out of this framework as Kleisli composition.
+However, in the usage-aware case, substitution needs not just a mapping of
+variables $f : \forallb{\Gamma \sqni \plr{-} \dotto \Delta \vdash \plr{-}}$,
+but rather an environment $\rho : \Delta \env\vdash \Gamma$, as we have already
+discussed.
+It therefore makes sense for our replacement for the Kleisli extension operator
+to similarly take an environment rather than a simple variable mapping.
+
+\Cref{thm:env-comp-lemma} below amounts to deriving a modified notion of Kleisli
+composition from a modified Kleisli extension.
+Additionally, \cref{thm:env-id-lemma} is required to turn a monadic unit into an
+identity environment.
+Both lemmas are stated in terms of general $\U$/$\V$/$\W$-environments, with
+some specific examples (e.g.\ for renaming and substitution) below them.
+
+\begin{lemma}[Identity environment]\label{thm:env-id-lemma}
+  Given a function
+  \[
+    \mathrm{vr}_{\Gamma'} : \forallb{\Gamma' \sqni \plr{-} \dotto \V\,\Gamma'}
+  \]
+  for any
+  $\Gamma$ we have an environment $\mathrm{id} : \Gamma \env\V \Gamma$.
 \end{lemma}
 \begin{proof}
+  Let $\Gamma = \grP\gamma$.
   Let $\gr\Psi$ be the identity map, which clearly satisfies
-  $\grP = \grP\gr\Psi$.
-  When acting on a variable, the equation $\grPprime = \grQprime\gr\Psi$ means
-  that $\grPprime = \grQprime$, so we want, from a variable of type
-  $\grPprime\gamma \sqni A$, a value of type $\V\,\grPprime\gamma\,A$, which
-  we can get from $\mathrm{vr}$.
+  $\grP \leq \grP\gr\Psi$.
+  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
+  means that $\grPprime \leq \grQprime$.
+  We are given a variable of type $\grQprime\gamma \sqni A$, which we can
+  coerce to a variable of type $\grPprime\gamma \sqni A$, upon which we apply
+  $\mathrm{vr}$ to get the required value of type $\V\,\grPprime\gamma\,A$.
 \end{proof}
 
-\begin{lemma}\label{thm:env-comp-lemma}
-  Given an environment $\rho : \Gamma \env\U \Delta$ for which we have, for any
-  $\grPprime$ and $\grQprime$ such that
-  $\grPprime = \grQprime\plr{\rho.\gr\Psi}$, we have a function
-  $\mathrm{lift}_\rho :
-  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$,
-  we can map environments of type $\Delta \env\V \Theta$ into environments of
-  type $\Gamma \env\W \Theta$.
+\begin{lemma}[Composition of environments]\label{thm:env-comp-lemma}
+  Given a function
+  \[
+    \mathrm{lift}_{\Gamma', \Delta'} :
+    \Gamma' \env\U \Delta' \to \forallb{\V\,\Delta' \dotto \W\,\Gamma'}
+  \]
+  we can compose environments $\rho : \Gamma \env\U \Delta$ and
+  $\sigma : \Delta \env\V \Theta$ into an environment
+  $\rho \gg \sigma : \Gamma \env\W \Theta$.
 \end{lemma}
 \begin{proof}
-  Let $\rho$ be as in the statement, and let $\sigma : \Delta \env\V \Theta$.
-  For the environment we are constructing, let
-  $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
-  $\grP = \grQ\plr{\rho.\gr\Psi} =
-  \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
-  For the action on variables, we are given $\grPprime = \grRprime\gr\Psi$ with
-  $\grRprime\theta \sqni A$.
-  We can immediately apply the action of $\sigma$, giving us a value of type
-  $\V\,\plr{\grRprime\plr{\sigma.\gr\Psi}}\,A$.
-  We note that
-  $\grPprime = \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
-  apply $\mathrm{lift}_\rho$ to get the desired value.
+  Let $\Gamma = \grP\gamma$, $\Delta = \grQ\delta$, and $\Theta = \grR\theta$.
+  Take $\gr\Psi$ to be the composition $\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi}$,
+  noting that
+  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
+  (\grR\plr{\sigma.\gr\Psi})\plr{\rho.\gr\Psi} = \grR\gr\Psi$
+  thanks to the inequalities yielded by $\sigma$ and $\rho$.
+  When acting on a variable, we are given $\grPprime \leq \grRprime\gr\Psi$ and
+  a variable $v : \grRprime\theta \sqni A$, and want a value of type
+  $\W\,\grPprime\gamma\,A$.
+  Let $\grQprime \coloneqq \grRprime\plr{\sigma.\gr\Psi}$, with inequality
+  $\grQprime \leq \grRprime\plr{\sigma.\gr\Psi}$ giving us a value
+  $\sigma(v) : \V\,\grQprime\delta\,A$.
+  We wish to apply $\mathrm{lift}$ to $\sigma(v)$ with
+  $\Gamma' \coloneqq \grPprime\gamma$ and $\Delta' \coloneqq \grQprime\delta$ to
+  complete the construction of the $\W$-value.
+  To do this, we need an environment of type
+  $\grPprime\gamma \env\U \grQprime\delta$, which we can get from $\rho$ using
+  \cref{thm:env-resize}, noting that
+  $\grPprime \leq \grRprime\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi} =
+  \grQprime\plr{\rho.\gr\Psi}$.
 \end{proof}
 
-\begin{corollary}[Composition of environments]
-  Given a function
-  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
-  \forall \grPprime, \grQprime.~\grPprime = \grQprime\plr{\rho.\gr\Psi} \to
-  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
-  compose environments of types $\Gamma \env\U \Delta$ and
-  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
-\end{corollary}
-
-\todo{What do these choices mean?}
 \begin{example}
   We can derive the following instances of environment composition.
-  \begin{itemize}
-    \item If $\U = \V = \W = {\sqni}$, then $\mathrm{lift}$ is given by the
-      action of the renaming $\rho$ on variables.
-      This allows us to derive composition of renamings.
-    \item More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
-      the action of the environment $\rho$.
-      This means that renamings post-compose with any other sort of environment.
-    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
-      syntactic traversal.
+  \begin{description}
+    \item[Composition of renamings.]
+      If $\U = \V = \W = {\sqni}$, then we can take
+      $\mathrm{lift}\,\rho\,x \coloneqq \rho(x)$.
+      This allows us to compose two renamings into a single renaming.
+    \item[Post-composition with a renaming.]
+      More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
+      the action of $\rho$ as $\mathrm{lift}$.
+      This means that we can do a composition
+      $\Gamma \env\U \Delta \env\sqni \Theta$ to get an environment of type
+      $\Gamma \env\U \Theta$.
+    \item[Composition of substitutions.]
+      If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
+      syntactic traversal (in the sense of \cref{sec:syntactic-kits} and
+      \cref{sec:lrsub}).
       For example, if $\U = {\sqni}$, we need the action of renaming on terms
@@
       substitutions compose.
-    \item More generally, if $\V = {\vdash}$ and we have a semantics from
+    \item[Composing semantics with substitution.]
+      More generally, if $\V = {\vdash}$ and we have a semantics (in the sense of
+      \cref{sec:gen-sem} and \cref{sec:framework}) from
       $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
       of terms.
-  \end{itemize}
+  \end{description}
 \end{example}

### 2022-12-12 tex/lnd.tex
[ edit ] new draft of chapters 5 and 6

@@
 parts of a rule, means that we keep repeating the context.
-Explicit contexts go against the style of natural deduction, which is based
+Explicit contexts go against the style we established in \cref{sec:simple},
+which is based
 around being parametric in the context, so that substitution is agnostic to the
@@
 I introduce the \emph{bunched connectives} for premises.
-These are inspired by bunched logic~\cite{oHP99}, and will not only be used for
-stating the syntax, but will reappear in the semantics.
+These are inspired by bunched logic~\citep{oHP99}, and will not only be used for
+stating the syntax, but will be used as an abstraction of common patterns in the
+development of the metatheory.
 The idea is to generalise the space between premises from Gentzen's natural
-deduction so as to allow for any linear combination of usage annotations.
+deduction to allow for any linear combination of usage annotations.
 Among other things, this generalisation will allow us to distinguish between
@@
 either \emph{sharing} or \emph{separating} conjunction.
-Similar connectives, but with different interpretations, could be used to
-define other linear-like type theories, like DILL, but here I will focus on the
-usage annotation style.
-The interpretations I use are defined in \cref{fig:bunched}.
+%Similar connectives, but with different interpretations, could be used to
+%define other linear-like type theories, like DILL, but here I will focus on the
+%usage annotation style.
+These connectives are defined in \cref{fig:bunched} in Agda notation.
 
 \begin{figure}
-  \begin{align*}
-    \dot1\,\grR &\coloneqq 1 \\
-    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
-    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
-    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
-    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
-                       \times T\,\grP \times U\,\grQ \\
-    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
-                       \times T\,\grP
-  \end{align*}
+  %\begin{align*}
+  %  \dot1\,\grR &\coloneqq 1 \\
+  %  (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
+  %  (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
+  %  I^*\,\grR &\coloneqq \grR \leq \gr0 \\
+  %  (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
+  %                     \times T\,\grP \times U\,\grQ \\
+  %  (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
+  %                     \times T\,\grP
+  %\end{align*}
+  Parameters:
+  \ExecuteMetaData[\Bunchedtex]{BunchedParams}
+
+  Connectives:
+  \vspace{-1em}
+  \begin{multicols}{2}
+    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseUnit}
+    \noindent\ExecuteMetaData[\Bunchedtex]{forallb}
+    \columnbreak
+    \AgdaNoSpaceAroundCode{}
+    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseConjunction}
+    \vspace{-0.25em}
+    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseImplication}
+    \AgdaSpaceAroundCode{}
+  \end{multicols}
+  \vspace{-2em}
+  \ExecuteMetaData[\Bunchedtex]{BunchedUnit}
+  \ExecuteMetaData[\Bunchedtex]{BunchedConjunction}
+  \ExecuteMetaData[\Bunchedtex]{BunchedImplication}
+  \ExecuteMetaData[\Bunchedtex]{BunchedScaling}
   \caption{The bunched connectives}
@@
 \end{figure}
+
+The bunched connectives are parametrised over two sets and three relations.
+For syntax, the set \AgdaBound{A} will be \AgdaRecord{Ctx}, the type of
+contexts, and \AgdaBound{R} will be \AgdaField{Ann}, the type of usage
+annotations (scalars).
+For the relations, the notation is meant to be suggestive, with
+\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\leq$0} typically stating that all
+of the annotations in \AgdaBound{$\Gamma$} are less than or equal to $0$;
+\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\leq$[}\AgdaSpace{}%
+\AgdaBound{$\Delta$}\AgdaSpace{}\AgdaBound{+}\AgdaSpace{}\AgdaBound{$\Theta$}%
+\AgdaSpace{}\AgdaBound{]}
+typically stating that \AgdaBound{$\Gamma$}, \AgdaBound{$\Delta$}, and
+\AgdaBound{$\Theta$} all agree on their types but the usage context of
+\AgdaBound{$\Gamma$} is less than or equal to the sum of the usage contexts of
+\AgdaBound{$\Delta$} and \AgdaBound{$\Theta$}; and
+\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaBound{$\leq$[}\AgdaSpace{}%
+\AgdaBound{r}\AgdaSpace{}\AgdaBound{*$_l$}\AgdaSpace{}\AgdaBound{$\Delta$}%
+\AgdaSpace{}\AgdaBound{]}
+typically stating that \AgdaBound{$\Gamma$} and \AgdaBound{$\Delta$} agree on
+their types but have the evident scaling relationship with \AgdaBound{r} on
+their usage annotations.
+I use the same symbols for the connectives both in Agda code and in otherwise
+standard mathematical/logical notation.
 
@@
 sharing a context is encoded by $\dottimes$.
-When we interpret a rule as a constructor of an inductive definition, $\dotto$
+When we interpret a typing rule as a constructor of an inductive definition,
+$\dotto$
 interprets the horizontal line, reflecting the fact that the usage annotations
-we start off with in the premises are those of the conclusion.
+we start off with in the premises are those of the conclusion, corresponding to
+a general principle of resource conservation.
 The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
@@
 
-The rest of the bunched connectives are for linear decompositions of the usage
-annotations.
+The rest of the bunched connectives --- $I^*$, $\sep$, $\cdot$, and $\wand$ ---
+involve linear decompositions of the usage annotations.
 The three basic left semimodule operators --- zero, addition, and left-scaling
@@
 \cref{fig:lr-bunched}.
-%I have not included a variable rule because, like with intuitionistic
-%natural deduction,
+The bunched connectives also yield a reasonably consise definition of the Agda
+data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.
 
-These bunched connectives should not be confused for the linear connectives
-they appear in the introduction rules of.
+While we have seen a connection between the bunched connectives and the
+connectives of $\name$, the two should not be confused.
+In particular, the bunched connectives obey different laws to what we would
+expect from linear logic.
 For example, it would make sense to define a bunched connective $\dotplus$,
@@
 We then have maps both ways between $T \dottimes (U \dotplus V)$ and
-$(T \dottimes U) \dotplus (T \dottimes V)$, reminiscent of bunched logic,
+$(T \dottimes U) \dotplus (T \dottimes V)$, reminiscent of the distributivity of
+additive connectives in bunched logic,
 whereas linear logic only has a map from $(A \with B) \oplus (A \with C)$ to
 $A \with (B \oplus C)$, and not a map the other way.
-%For example, linearly we have
-%$\oc{\gr r}(A \with B) \simeq \oc{\gr r} A \otimes \oc{\gr r} B$, while the
-%expressions $\gr r \cdot (A \dottimes B)$ and $\gr r \cdot A \sep \gr r \cdot B$
-%are generally incomparable.
-We will also see later\todo{Forward reference} that, while
-$(\dottimes, \dotto)$ gives a Cartesian \todo{wrong}
-closed structure, $\sep$ is monoidally closed, with an internal hom $\wand$.
-Indeed, the Cartesian closed structure is sufficient to give us the interaction
-between $\dotplus$ and $\dottimes$, by the fact that $A \dottimes {-}$ is a
-left adjoint, and therefore preserves colimits.
 Looking at the interpretations, the connection with bunched logic makes sense.
-Instead of the partial commutative monoid (representing heaps) found in
+Instead of the partial commutative monoid (often representing heaps) found in
 standard semantics of bunched logic, we have a left semimodule of usage
@@
 various subterms.
-\todo{Bunched type system to Agda}
-\todo{Split chapter here. Play around with other systems using bunched
-  connectives. Lists/recursion, annotated arrow, Granule elim, sharing elim.}
+
+A discrepancy with bunched logic is that $\dotto$ is not part of the Cartesian
+closure of context-indexed families.
+In fact, it fails to be a functor because it does not respect the ordering of
+usage contexts.
+This fact has little impact on the development of this thesis, because I mainly
+use $\dotto$ at the top level, directly in conjunction with $\forallb{-}$.
+The combination describes the ``external hom'', i.e.\ the morphisms in the
+category of context-indexed families, with the internal hom being more like the
+Kripke function space, which I do not use.
+
+%\todo{Split chapter here. Play around with other systems using bunched
+%  connectives. Lists/recursion, annotated arrow, Granule elim, sharing elim.}
 
@@
 \end{figure}
+
+\begin{figure}
+  \ExecuteMetaData[\SpecificSyntaxtex]{Ty}
+  \ExecuteMetaData[\SpecificSyntaxtex]{Bind}
+  \ExecuteMetaData[\SpecificSyntaxtex]{lr}
+  \caption{\name{} stated using bunched connectives in Agda}
+  \label{fig:lr-bunched-Agda}
+\end{figure}

### 2022-12-12 tex/lr.tex
[ edit ] new draft of chapters 5 and 6

@@
 In this section, I introduce the syntax of the type theory \name{}, which makes
-use of posemiring usage annotations.
+use of posemiring usage annotations to express the usage restrictions found in
+DILL and other calculi.
 I use this syntax to write some example programs, which will motivate the
 denotational semantics explored in \cref{sec:wrel}.
-For the rest of this thesis, \name{} will serve as both a prototypical
+For the rest of this thesis, particularly
+\cref{sec:semirings,sec:ren-sub-lr,sec:wrel},
+\name{} will serve as both a prototypical
 usage-constrained syntax and a target of semantic analyses.
 
-The calculus \name{} is similar in spirit to intuitionistic linear logic (ILL).
+The calculus \name{} is similar in spirit to intuitionistic linear logic (ILL),
+which we saw in \cref{sec:linearity}.
 The types of \name{}, listed in \cref{fig:lr-types}, are almost identical
@@
 variable in the context.
-In this position, an assumption of type $\oc_{\gr r} A$ acts like an assumption
-of type $A$ that is to be used according to $\gr r$ rather than the standard
-$\gr1$.
 
@@
 
-I will not cover the operational semantics or equational theory of \name{} in
+I will not cover any operational semantics or equational theory of \name{} in
 this thesis.
@@
 certain usage vectors $\grP$ and $\grQ$.
-To a first approximation, $\leq$ can simply be read as $=$.
-The point of using $\leq$ rather than $=$ is to allow for \emph{subusaging},
-i.e., subsumption of usage annotations.
+We saw subusaging in the introduction to this chapter in the specific case of
+$\Ann$ being formed from the poset $\{\gr0 > \gr\omega < \gr1\}$.
+This allowed variables annotated $\gr\omega$ (``unrestricted'') to be both
+weakened/discarded (because $\gr\omega \leq \gr0$) and derelicted/used
+(because $\gr\omega \leq \gr1$).
+Subsumption of usage annotations is essential to nearly all interesting choices
+of $\Ann$.
+However, in the toy example of exact usage counting using the set $\mathbb N$ of
+annotations, we set the order to be just equality as a matter of simplicity.
+
 For usage annotations $\gr r$ and $\gr s$, the inequality $\gr r \leq \gr s$
@@
 A mnemonic is that $\gr r$ is less specific than $\gr s$.
-The principle is reflected by the admissible subusaging rule.
+The principle is reflected by the admissible subusaging rule, where the order
+has been lifted from annotations to usage contexts.
+The subusaging rule is a simple corollary of renaming, as given in
+\cref{sec:ren-sub-lr}.
 
@@
 \]
-
-Subusaging is essential to nearly all interesting choices of $\Ann$.
-For example, we can capture intuitionistic linear logic by choosing $\Ann$ to
-be $\{\gr0 > \gr\omega < \gr1\}$.
-This allows variables annotated $\gr\omega$ (``unrestricted'') to be both
-weakened/discarded (because $\gr\omega \leq \gr0$) and derelicted/used
-(because $\gr\omega \leq \gr1$).
 
@@
 
-To get a feeling for \name{}, I will temporarily fix
-$\Ann \coloneqq (\mathbb N, =, 0, +, 1, \times)$.
-That is, the usual semiring of natural numbers with ordering given by equality.
+\begin{example}
+  Let $\Ann \coloneqq (\mathbb N, =, 0, +, 1, \times)$, that is, specialise to
+  the posemiring made of
+  the usual semiring of natural numbers with ordering given by equality.
   Under this discipline, the usage constraints enforce a form of exact usage
   counting.
-
-\begin{example}
-  The following judgements hold for the natural number semiring.
+  The following judgements then hold.
   Derivations are left as an exercise to the reader.
@@
 
-{\color{red}TODO: cite rntz}
 \begin{example}\label{def:monotonicity-posemiring}
@@
   $\{\gr{\wn\wn}, \gr{\uparrow\uparrow}, \gr{\downarrow\downarrow},
-  \gr{\sim\sim}\}$, with $0 \coloneqq \gr{\wn\wn}$,
-  $1 \coloneqq \gr{\uparrow\uparrow}$, and the following operations:
+  \gr{\sim\sim}\}$.
+  The idea is that each symbol represents the possible \emph{variance} of an
+  input (free variable) with respect to some partial ordering on a semantic
+  domain of elements.
+  $\gr{\uparrow\uparrow}$ represents covariance (if that input goes up, the
+  output goes up), $\gr{\downarrow\downarrow}$ represents contravariance
+  (if that input goes \emph{down}, the output goes up), $\gr{\sim\sim}$ gives no
+  guarantees (if that input remains constant, the output (trivially) goes up),
+  and $\gr{\wn\wn}$ says that that input is irrelevant (whatever changes are
+  made to that input, the output (trivially) goes up).
+
+  I take $0 \coloneqq \gr{\wn\wn}$, $1 \coloneqq \gr{\uparrow\uparrow}$,
+  and define the following operations:
 
@@
 
-  The idea is that each symbol represents the possible \emph{variance} of an
-  input (free variable) with respect to some partial ordering on a semantic
-  domain of elements.
-  $\gr{\uparrow\uparrow}$ represents covariance (if that input goes up, the
-  output goes up), $\gr{\downarrow\downarrow}$ represents contravariance
-  (if that input goes \emph{down}, the output goes up), $\gr{\sim\sim}$ gives no
-  guarantees (if that input remains constant, the output (trivially) goes up),
-  and $\gr{\wn\wn}$ says that that input is irrelevant (whatever changes are
-  made to that input, the output (trivially) goes up).
-
   Addition represents an intersection of guarantees.
@@
   As such, $\oc\gr{\downarrow\downarrow}A$ represents a contravariant $A$.
+  The flipping (involutive) behaviour of $\gr{\downarrow\downarrow}$ lets us
+  notice that $x$ is covariant in
+  a term like $-(-x)$, where $-$ is a constant of type
+  $\oc\gr{\downarrow\downarrow}\mathbb Z \multimap \mathbb Z$.
+
+  A similar, but distinct, collection of modalities for monotonicity is given by
+  \citet{Arntzenius19}.
 \end{example}

### 2022-12-12 tex/lrkits.tex
[ edit ] new draft of chapters 5 and 6

@@
-In an effort to stick to the syntactic kits and traversals approach of
+In an effort to reuse the syntactic kits and traversals approach of
 \cref{sec:syntactic-kits}, I will derive the types of simultaneous renaming
@@
 To get a type of environments suitable for the usage-aware setting, I first
-analyse intuitionistic environments, distilling the slick functional definition
+analyse intuitionistic environments (as introduced in \cref{sec:syntactic-kits}
+definition \AgdaFunction{Env}), distilling the easy-to-use functional definition
 (\cref{def:simple-env}) into a more basic recursive definition
@@
   \begin{itemize}
-    \item There is one environment $\alr{} : \Gamma \env\V_R {\cdot}$.
+    \item There is an environment $\alr{} : \Gamma \env\V_R {\cdot}$.
     \item For $\rho_l : \Gamma \env\V_R \Delta_l$ and
@@
 
-To account for usage, we must replace the simple copying of $\Gamma$ by
-copying of just the types $\gamma$ and \emph{sharing up} of the usage
+To account for usage, we must replace the simple repetition of $\Gamma$ by
+repetition of just the types $\gamma$ and \emph{redistribution} of the usage
 annotations $\grP$.
@@
 \begin{example}
-  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
-  addition and multiplication.
-  There is an intuitionistic recursive environment
+  Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$, with the equality order
+  chosen to avoid any concerns around subsumption of annotations.
+  Then, there is an intuitionistic recursive environment as follows, where
+  $y\,z$ is the application of $y$ to $z$.
   \[
@@
   $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
+
+  When there are multiple environments in question and $\rho$ is such an
+  environment, I use the notation $\rho.\gr\Psi$ to refer to $\gr\Psi$.
+  For example, $\grP \leq \grQ\plr{\rho.\gr\Psi}$.
+  For the action on variables, I write $\rho(x)$, where
+  $x : \grQprime\delta \sqni A$.
+  The expression ``$\rho(x)$'' alone is ambiguous because of the slack in the
+  usage context $\grPprime$ of the resulting value.
+  Therefore, I will always make sure $\grPprime$ and $\grQprime$ clear when
+  using this notation.
 \end{definition}
@@
 \begin{lemma}
-  We can use an environment $\rho : \Gamma \env\V \Delta$ to get from a
+  We can use an environment $\rho : \Gamma \env\V \Delta$ to map a
   usage-checked variable $x : \Delta \sqni A$ to a value of type
@@
   Set $\grPprime \coloneqq \grP$ and $\grQprime \coloneqq \grQ$, then
-  $\grPprime \leq \grQprime\gr\Psi$ by the constraint in $\rho$, and we can
-  use the lookup function to map $x$ to a value.
+  $\grP \leq \grQ\gr\Psi$ by the constraint in $\rho$, so we can take
+  the $\V$-value $\rho(x)$.
 \end{proof}

### 2022-12-12 tex/lrsub.tex
[ build ] get thesis into building state
[ edit ] new draft of chapters 5 and 6

@@
-\def\LRKits{../agda/latex/LRKits.tex}
+\def\LRKits{../agda/processed-latex/LRKits.tex}
 
-In this section, I show that, using the notion of \emph{environment} derived in
+I can now show that, using the notion of \emph{environment} derived in
 \cref{sec:lrkits}, we can replicate the Agda proofs from
-\cref{sec:syntactic-kits} in the usage-aware setting.
+\cref{sec:syntactic-kits} in the usage-aware setting of $\name$.
 From \cref{sec:lenv}, we know that environments are preserved under all
@@
 
-There are a few notational changes necessary in the Agda code.
+There are a few notational changes necessary in the Agda code, compared to the
+typeset mathematics above.
 Usage vectors, elsewhere called $\grP$, $\grQ$, and $\grR$ are rendered as
 \AgdaBound{P}, \AgdaBound{Q}, and \AgdaBound{R}, respectively.
+Usage contexts and typing contexts are tied together with the
+\AgdaInductiveConstructor{ctx} constructor, rather than simple juxtaposition.
 Environments, elsewhere notated $\Gamma \env\V \Delta$, are rendered as
@@
 We start with a slightly modified definition of \AgdaRecord{Kit}.
-We saw in \cref{thm:lr-bind} that in the usage-annotated context, we replace
-general weakening of $\V$-values by weakening by $\gr0$-use variables.
+We saw in \cref{thm:lr-bind} that in the usage-annotated context, we restrict
+weakening of $\V$-values to just $\gr0$-use variables.
 Meanwhile, the function $\mathrm{vr}$, also seen in \cref{thm:lr-bind}, maps
@@
 I state weakening in a slightly different way than previously, so as to help
-unification against a known result type.
+unification against a known result type (avoiding the problem described by
+\citet{McBride12} as \emph{green slime}).
 The type \AgdaFunction{Weakening}\AgdaSpace{}\AgdaBound{$\V$} can be read as
@@
 
-Given \AgdaFunction{bindEnv}, along with \AgdaFunction{env-+} and
-\AgdaFunction{env-*}, standing for \cref{thm:lr-env-add} and
-\cref{thm:lr-env-scale}, respectively, we can reproduce the syntactic traversal
-\AgdaFunction{trav}.
-With all these lemmas in place, we can see how writing \AgdaFunction{trav}
-becomes routine --- when processing a rule, we work our way up through the
+Given \AgdaFunction{bindEnv} (\cref{thm:lr-bind}), \AgdaFunction{env-+}
+(\cref{thm:lr-env-add}), and \AgdaFunction{env-*} (\cref{thm:lr-env-scale}),
+we can reproduce the syntactic traversal \AgdaFunction{trav}.
+With all these lemmas in place, writing \AgdaFunction{trav}
+becomes routine.
+When processing a rule, we work our way up through the
 premise connectives, applying \AgdaFunction{env-*} wherever we see a
@@
 \AgdaFunction{Bind}.
-We when use whatever environments (with names beginning with
+We then use whatever environments (with names beginning with
 \AgdaBound{$\rho$}) and whatever usage vector splitting facts (with names
@@
 \AgdaFunction{ren} to yield a term weakening function.
-However, a usage-aware environment is a function packed together with some
-extra data.
+However, a usage-aware environment is a function packed together with usage
+distribution data.
 As such, we must make an environment version of \AgdaFunction{$\swarrow^v$}.
@@
 This is what we need to instantiate \AgdaFunction{trav} for substitution.
+As a reminder, I also give the type of \AgdaFunction{sub} in rule form.
 
 \ExecuteMetaData[\LRKits]{sub}
+\[
+  \ebrule{%
+    \hypo{\Gamma \env\vdash \Delta}
+    \hypo{\Delta \vdash B}
+    \infer2[sub]{\Gamma \vdash B}
+  }
+\]
 
@@
 \[
-  \begin{prooftree}
+  \ebrule{%
     \hypo{\grR \leq \grQ + \gr r\grP}
@@
     \infer3[singleSub]{\grR\gamma \vdash B}
-  \end{prooftree}
+  }
 \]

### 2022-12-12 tex/rec.tex
[ edit ] new draft of chapters 5 and 6

@@
+The bunched connectives give us a means by which to quickly design and
+experiment with new type systems.
+We will see in \cref{sec:framework} that all such syntaxes are well behaved to
+the extend that they all support the appropriate notion of substitution, which
+again expediates experimentation.
+In this section, I give some example syntactic features that can be covered in
+the bunched connective paradigm.
+
+\subsection{Alternative object-language connectives}
+
+I presented $\name$ as an essentially arbitrary collection of rules, with no
+broader characterisation than that it intuitively seems to capture some notion
+of ``usage'' and that it supports the semantics I give in \cref{sec:wrel}.
+Therefore, in this subsection I give some examples of variant syntaxes others,
+with different use cases, may be interested in.
+In particular, I give a variant of function types where usage annotations go
+``on the arrow'', and a variant of tensor products with a stronger elimination
+rule.
+
+We can produce an annotated function arrow connective by combining parts of the
+rules for the unannotated function arrow $\multimap$ with parts of the rules for
+the modal operator $\oc$.
+
+\begin{mathpar}
+  \ebrule[comb]{%
+    \hypo{\gr rA \vdash B}
+    \infer1[$\gr r\multimap$-I]{\vdash \gr rA \multimap B}
+  }
+  \and
+  \ebrule[comb]{%
+    \hypo{\vdash \gr rA \multimap B}
+    \hypo{\sep}
+    \hypo{\gr r \cdot \plr{\vdash A}}
+    \infer3[$\gr r\multimap$-E]{\vdash B}
+  }
+\end{mathpar}
+
+We can also give an alternative elimination rule for $\otimes$-products inspired
+by Granule~\citep{Granule18,HMWO21}.
+This modified rule allows us to, for example, have a variable
+$\gr rx : A \otimes B$ for any $\gr r$, and pattern-match it into
+$\gr ry : A, \gr rz : B$.
+With the standard $\name$ rule from \cref{fig:lr}, the newly bound variables are
+always given annotation $\gr 1$, and we would only be able to do the match in
+the first place if $\gr r$ were coercible to (i.e.\ less than or equal to)
+$\gr1$.
+The alternative rule below is not compatible with linear logic, because it
+allows us to derive
+$\oc\gr r\plr{A \otimes B} \multimap \oc\gr rA \otimes \oc\gr r B$
+parametric in $\gr r$, $A$, and $B$.
+
+\begin{mathpar}
+  \ebrule[comb]{%
+    \hypo{\gr r \cdot \plr{\vdash A \otimes B}}
+    \hypo{\sep}
+    \hypo{\gr rA, \gr rB \vdash C}
+    \infer3[$\otimes$-E$'$]{\vdash C}
+  }
+\end{mathpar}
+
+\subsection{Recursion}\label{sec:rec}
+
 Based on an intuitive understanding of ``usage'', recursion introduces a new
 phenomenon relative to the forms of programs we have seen so far:
-terms can be used an unbounded number of times.
+Terms can be used an unbounded number of times.
 For example, notice the following reduction in Agda.
 
-\missingfigure{\texttt{foldr \_+\_ 0 (1 :: 2 :: 3 :: []) = 1 + (2 + (3 + 0))}}
+\ExecuteMetaData[\SnippetsTwotex]{sum-reduction}
 
@@
 
-I propose the following additions to \name{} to support a broad class of
+The following additions to \name{} support a broad class of
 inductive types.
 I define strictly positive functors syntactically, with the only notable
-restriction being not being allowed to use the type variable $X$ in the domain
-of a function type.
+restrictions being not being allowed to use the type variable $X$ in the domain
+of a function type and within a $\oc$-type.
 I then add least fixed points of such strictly positive functors to the syntax
@@
 \begin{align*}
-  U &\Coloneqq A \multimap (-) \mid \oc\gr r(-) \\
+  U &\Coloneqq A \multimap (-) \\
   {\odot} &\Coloneqq {\otimes} \mid {\oplus} \mid {\with} \\
@@
         \grQ \leq \gr0 \\
-        \grQ \leq \grQ + \grQ \\
-        \forall \gr r.~\grQ \leq \gr r\grQ
+        \grQ \leq \grQ + \grQ
       \end{matrix*}%
@@
   For lists, we can derive the following introduction and elimination rules
-  (with usage constraints omitted for brevity when obvious).
+  (with usage constraints in the application of \TirName{$\mu$-E} in the
+  \TirName{foldr} rule, relating $\grR$ to $\grP$ and $\grQ$ and restricting
+  $\grQ$, omitted to save space).
 
   \begin{align*}
+    \text{nil}\left[
     \begin{prooftree}
       \hypo{\grR \leq \gr0}
-      \infer1[$I$-I]{I}
+      \infer1[$I$-I]{\grR\gamma \vdash I}
       \infer1[$\oplus$-I$_0$]%
@@
     \end{prooftree}
+    \right.
     &&
+    \text{cons}\left[
     \begin{prooftree}
@@
     \end{prooftree}
+    \right.
   \end{align*}
   \begin{displaymath}
+    \text{foldr}\left[
     \begin{prooftree}
@@
     \end{prooftree}
+    \right.
   \end{displaymath}
@@
 $\grQ$, and $\grR$ into the result of some premise combinators.
-To do this, I introduce a new combinator $\Box^{0{+}{\times}}$ defined below,
+To do this, I introduce a new combinator $\Box^{0{+}}$ defined below,
 along with the resulting implicit-context typing rules.
 
-\begin{align*}
-  \plr{\Box^{0{+}{\times}}\,T}\grR \coloneqq
-  \plr{\grR \leq \gr0} \times \plr{\grR \leq \grR + \grR} \times
-  \plr{\forall \gr r.~\grR \leq \gr r\grR} \times T\,\grR
-\end{align*}
+\ExecuteMetaData[\Bunchedtex]{Box}
 
@@
     \hypo{\sep}
-    \hypo{\Box^{0{+}{\times}}\plr{\gr1F[C] \vdash C}}
+    \hypo{\Box^{0{+}}\plr{\gr1F[C] \vdash C}}
     \infer3[$\mu$-E]{\vdash C}
@@
       \hypo{\sep}
-      \hypo{\Box^{0{+}{\times}}
+      \hypo{\Box^{0{+}}
         \plr{\vdash C\hskip0.75em\dottimes\hskip0.75em\gr1A, \gr1C \vdash C}}
@@
 \end{example}
+
+As suggested by the notation, $\Box^{0{+}}$ may not be particularly canonical in
+terms of what operations the usage context is stable under.
+For example, we could also ask for $\forall \gr r.~\grQ \leq \gr r\grQ$, i.e.\
+that the usage context is stable under any scaling.
+This would allow us to include $\oc\gr r(-)$ in our grammar of strictly positive
+functors.
+In certain semirings, this could give us strange types, like, taking the
+monotonicity semiring, the type
+$\mu X.~I \oplus (\iota \otimes \oc\gr{\downarrow\downarrow} X)$,
+which is the type of lists over the base type $\iota$ where alternating elements
+are treated contravariantly and covariantly.
+This may motivate us to restrict strictly positive functors to allow only a
+subset of usage annotations.
+Separately, in some contexts we may only need to know that $\grQ \leq \gr0$.
+For example, in the linearity semiring, being less than or equal to $\gr0$
+happens to imply closure under addition, because the only annotations less than
+or equal to $\gr0$ are $\gr0$ and $\gr\omega$.
+Also, in semirings for relevant systems, all annotations are stable under
+addition (up to $\leq$ in the appropriate direction), so we need only check for
+being less than or equal to $\gr0$.

### 2022-12-12 tex/semirings.tex
[ edit ] new draft of chapters 5 and 6

@@
 
-The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
+The methods described in \cref{sec:simple} for the simply typed
+$\lambda$-calculus
 make crucial use of \emph{weakening} --- the fact that if we have
 $\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
-We use this property to update environments as they go under binders.
+We use this property to update environments as we take them under binders.
 However, as we saw in \cref{sec:linearity}, there are interesting calculi in
@@
 
+The eventual target of all these $\gr0$-annotated variables is the variable
+rule, which I transcribe as follows.
+The $\gr1$ shows us that we can use the variable thus annotated, while the
+$\gr0$s let us discard all of the other variables in $\gamma$.
+
+\[
+  \ebrule{%
+    \infer0{A \vdash A}
+  }
+  \quad\rightsquigarrow\quad
+  \ebrule{%
+    \infer0{\gr0\gamma, \gr1A \vdash A}
+  }
+\]
+
 The use of $\gr0$ gives us the property that variables never go out of
@@
 
-If we follow the DILL style of variable management, there are not just the two
-states \emph{used} ($\gr1$) and \emph{not used} ($\gr0$), but also
+If we follow the DILL style of variable management explained in \cref{sec:dill},
+there are not just the two
+states \emph{to be used} ($\gr1$) and \emph{not to be used} ($\gr0$), but also
 \emph{usable unrestrictedly}.
@@
 calligraphic $\grP$, $\grQ$, and $\grR$).
-It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
-rule on the right below now accepts $\gr0$-annotated variables in its
+A beneficial side-effect of the fact that $\gr0 + \gr0 = \gr0$ is that the
+rule on the right below is actually more general, and accepts $\gr0$-annotated
+variables in its
 conclusion, which is essential for weakening to be admissible.
@@
 
-Some other transcriptions are as follows.
+Some other transcriptions from DILL to the usage annotation style are as
+follows.
 I unify the variable rules (the one for linear variables and the one for
@@
 weakened away like $\gr0$-annotated variables.
-All together, this means that at the unified variable rule, the variable being
+This ordering information is shown in the diagram
+\begin{tikzpicture}[baseline]
+  \node(omega) at (0,0) {$\gr\omega$};
+  \node(0) [above left of=omega] {$\gr0$};
+  \node(1) [above right of=omega] {$\gr1$};
+
+  \draw (omega) -- (0);
+  \draw (omega) -- (1);
+\end{tikzpicture}.
+All together, this means that at the (only) variable rule, the variable being
 used must have annotation less than or equal to $\gr1$, and every other variable
@@
 $\gr\omega$-many premises.
-This explanation recovers the behaviour that only $\gr\omega$- and
+This explains why only $\gr\omega$- and
 $\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
-and also justifies the designation of multiplication (vector scaling) as the
+and also justifies the choice below of multiplication (vector scaling) as the
 algebraic operation controlling the $\oc$-modality.
@@
 behaved.
-\todo{Refer back to preliminaries for definition of posemiring}
+%\todo{Refer back to preliminaries for definition of posemiring}
 For concreteness, I collect together the definition of the
@@
 \begin{example}\label{def:lin-semiring}
-  The \emph{$\{\gr0, \gr1, \gr\omega\}$ semiring}, also known as the
-  \emph{linearity semiring}, has the operations given as follows, with
+  The \emph{$\{\gr0, \gr1, \gr\omega\}$ posemiring}, also known as the
+  \emph{linearity posemiring}, has the operations given as follows, with
   $0 \coloneqq \gr0$ and $1 \coloneqq \gr1$:
@@
 simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
-The idea of \name{} is to augment simply typed $\lambda$-calculus with
+The idea of \name{} is to augment the simply typed $\lambda$-calculus with
 annotations on free variables, and give enough types to manipulate these
 annotations.
-I use \cref{sec:lnd} to introduce some notation that allows us to restate the
-typing rules of \name{} so as to not mention contexts explicitly.
-The first problem I tackle with this new syntax is proving admissibility of
-substitution.
-I follow the methodology from \cref{sec:kits}, which
+Given this new calculus $\name$, the first goal is to apply the techniques of
+\cref{sec:simple} to it, yielding a simultaneous substitution operation.
+To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
+the typing rules of \name{} to not mention contexts explicitly, as was the style
+in \cref{sec:simple}.
+This new notation --- the \emph{bunched connectives} --- is versatile at
+defining simply typed usage-aware syntaxes, and I give further non-$\name$
+examples in \cref{sec:rec}.
+The application of the techniques of \cref{sec:simple} is left to
+\cref{sec:ren-sub-lr}.
 
@@
 \input{lnd.tex}
+\section{Additions to and variations of \name{}}\label{sec:variant}
+\input{rec.tex}
+\section{Addendum: (lack of) partiality}\label{sec:part}
+\input{part.tex}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\chapter{Renaming and substitution of $\name$}\label{sec:ren-sub-lr}
+
+In \cref{sec:semirings}, I defined my calculus of interest $\name$.
+In this chapter, I develop its syntactic metatheory up to and including
+substitution.
+I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
+significant changes to the underlying notion of \emph{environment} before doing
+so.
+I give and informally motivate these changes to environments in
+\cref{sec:lrkits}, and prove some properties of the new definition in
+\cref{sec:lenv}.
+Finally, I apply these new environments to the syntax of $\name$ in
+\cref{sec:lrsub} to derive renaming and substitution operators.
+
 \section{What are linear renaming and substitution?}\label{sec:lrkits}
@@
 \input{lrsub.tex}
-\section{Adding recursion to \name{}}\label{sec:rec}
-\input{rec.tex}
-\section{Addendum: (lack of) partiality}\label{sec:part}
-\input{part.tex}

### 2022-12-24 agda/LRKits.lagda.tex
[ edit ] another draft of chapters 5 and 6

@@
   open import Relation.Binary.PropositionalEquality as ≡ using (_≡_)
-  open import Relation.Nary
+  open import Relation.Nary.Extra using (_⊆_)
   open import Relation.Unary.Bunched
@@
   open import Generic.Linear.Syntax Ty Ann
-  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
+  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring hiding (◇)
   open import Generic.Linear.Variable Ty rawPoSemiring hiding (_↙ᵛ_; _↘ᵛ_)
@@
   data _⊢_ : Ctx → Ty → Set where
-    var : ∀[ _∋_ ⇒ _⊢_ ]
-    !-I : ∀ {r A} → ∀[ r ·ᶜ (_⊢ A) ⇒ _⊢ `! r A ]
-    !-E : ∀ {r A C} → ∀[ (_⊢ `! r A) ✴ᶜ Bind [ r · A ]ᶜ (_⊢ C) ⇒ _⊢ C ]
+    var : _∋_ ⊆ _⊢_
+    !-I : ∀ {r A} → r · (_⊢ A) ⊆ _⊢ `! r A
+    !-E : ∀ {r A C} → (_⊢ `! r A) ✴ Bind [ r ∙ A ]ᶜ (_⊢ C) ⊆ _⊢ C
 \end{code}
@@
       let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
-      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ Qₗ +* Qᵣ ] →
+      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ (_≤[ Qₗ +* Qᵣ ]) →
       (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
@@
     ∀ {s t P} {γ : Vector Ty (s <+> t)} → P ∘ ↘ ≤* 0* →
-    ∀[ 𝓥 (ctx (P ∘ ↙) (γ ∘ ↙)) ⇒ 𝓥 (ctx P γ) ]
+    𝓥 (ctx (P ∘ ↙) (γ ∘ ↙)) ⊆ 𝓥 (ctx P γ)
 
@@
       ↙ᵏ : Weakening 𝓥
-      vr : ∀[ _∋_ ⇒ 𝓥 ]
-      tm : ∀[ 𝓥 ⇒ _⊢_ ]
+      vr : _∋_ ⊆ 𝓥
+      tm : 𝓥 ⊆ _⊢_
 \end{code}
@@
     bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
-    bindEnv ρ .Ψ = [ [ ρ .Ψ │ 0ᴿ ]ᴿ ─ [ 0ᴿ │ 1ᴿ ]ᴿ ]ᴿ
+\end{code}
+%</bindEnv>
+\begin{code}
+    bindEnv ρ .Ψ = [  [ ρ .Ψ  │ 0ᴿ ]ᴿ
+                              ─
+                      [ 0ᴿ    │ 1ᴿ ]ᴿ ]ᴿ
     bindEnv ρ .fit-here = _↘,_,↙_
@@
 \end{code}
-%</bindEnv>
 
@@
     trav ρ (var x) = tm (ρ .lookup (ρ .fit-here) x)
-    trav ρ (!-I (⟨ sp* ⟩· M)) =
+    trav ρ (!-I (⟨ sp* ⟩·ᶜ M)) =
       let sp*′ , ρ′ = env-* (ρ , sp*) in
-      !-I (⟨ sp*′ ⟩· (trav ρ′ M))
-    trav ρ (!-E (M ✴⟨ sp+ ⟩ N)) =
+      !-I (⟨ sp*′ ⟩·ᶜ (trav ρ′ M))
+    trav ρ (!-E (M ✴ᶜ⟨ sp+ ⟩ N)) =
       let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
-      !-E (trav ρₗ M ✴⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
+      !-E (trav ρₗ M ✴ᶜ⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
 \end{code}
@@
 
+  ren : ∀ {Γ Δ A} → [ _∋_ ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
   ren = trav ∋-kit
@@
 
+  sub : ∀ {Γ Δ A} → [ _⊢_ ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
   sub = trav ⊢-kit
@@
     sub[-] : ∀ {s P Q R γ r A B} → R ≤[ Q +*_] ◇ _≤[ r *ₗ P ] →
-      ctx {s} P γ ⊢ A → ctx Q γ ++ᶜ [ r · A ]ᶜ ⊢ B → ctx R γ ⊢ B
+      ctx {s} P γ ⊢ A → ctx Q γ ++ᶜ [ r ∙ A ]ᶜ ⊢ B → ctx R γ ⊢ B
     sub[-] {Q = Q} {R} {γ} {r} {A} (sp+ , sp*) t = sub σ
       where
-      σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r · A ]ᶜ
-      σ = ++ᵉ (id^Env ✴⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩· t))
+      σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r ∙ A ]ᶜ
+      σ = ++ᵉ (id^Env ✴ᶜ⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩·ᶜ t))
 \end{code}

### 2022-12-24 tex/examples.tex
[ writing ] content sections to full draft
[ edit ] another draft of chapters 5 and 6

@@
 
-\subsection{Renaming and substitution}\label{sec:kits}
+\subsection{Renaming and substitution}\label{sec:kit-to-sem}
 
@@
 
+Standard denotational semantics falls out as a somewhat trivial case of
+\AgdaRecord{Semantics}.
+A lot of work is done in the generic traversal \AgdaFunction{semantics} to
+maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
+family.
+
+{\color{red}Old stuff below\ldots}
+
 To justify the name \emph{semantics}, we give an example traversal that is a
@@
 
-\todo{Define big intersection.}
 When the poset of worlds forms a (relational) commutative monoid, such

### 2022-12-24 tex/lenv.tex
[ edit ] another draft of chapters 5 and 6

@@
   $\grPprime \leq \grP \to
-  \forallb{\V\,\grP\gamma \dotto \V\,\grPprime\gamma}$).
+  \V\,\grP\gamma \rightarrowtriangle \V\,\grPprime\gamma$).
   \begin{itemize}
-    \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
-    \item $\forallb{\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
-      \dotlr \plr{{-} \env\V \Delta_l, \Delta_r}}$
-    \item
-      $\forallb{\gr r \cdot \plr{\V\,(-)\,A} \dotlr \plr{{-} \env\V \gr rA}}$
+    \item $I^{\sep} \leftrightarrowtriangle \plr{{-} \env\V {\cdot}}$
+    \item $\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
+      \leftrightarrowtriangle \plr{{-} \env\V \Delta_l, \Delta_r}$
+    \item $\gr r \cdot \plr{\V\,(-)\,A}
+      \leftrightarrowtriangle \plr{{-} \env\V \gr rA}$
   \end{itemize}
@@
   \begin{description}
-    \item[$I(\to)$]
+    \item[$I^{\sep}(\rightarrowtriangle)$]
       Let $\gr\Psi$ be the unique linear map out of the zero space.
@@
       There are no variables to act upon.
-    \item[$I(\gets)$]
+    \item[$I^{\sep}(\leftarrowtriangle)$]
       $\grQ\gr\Psi$ is an empty sum, so if $\grP \leq \grQ\gr\Psi$ then
       $\grP \leq \gr0$.
-    \item[$\sep(\to)$]
+    \item[$\sep(\rightarrowtriangle)$]
       Let the given environments be $\rho_l : \grPl\gamma \env\V \grQl\delta$
@@
       and we can act on the variable using $\rho_l$.
-    \item[$\sep(\gets)$]
+    \item[$\sep(\leftarrowtriangle)$]
       Let the unnamed context be $\Gamma$, also written $\grP\gamma$.
@@
       We can therefore act using the original environment.
-    \item[$\cdot(\to)$]
+    \item[$\cdot(\rightarrowtriangle)$]
       Let $\grP$ and $\grPprime$ be such that $\grP \leq \gr r\grPprime$ and let
@@
       $\V\,\grPprime\gamma\,A$, which we can take to be $v$.
-    \item[$\cdot(\gets)$]
+    \item[$\cdot(\leftarrowtriangle)$]
       Let us have an environment of type $\grP\gamma \env\V \gr rA$.
@@
   ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
-  \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
-  $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
+  \V\,\Gamma \rightarrowtriangle \V\,\plr{\Gamma, \grR\theta}$ and
+  $\mathrm{vr} : {\sqni} \rightarrowtriangle \V$, we can turn an environment of
   type $\Gamma \env\V \Delta$ into an environment of type
@@
 Notice that, given $F,G : \blr{\mathrm{Ty}, \Set}$, a morphism from $F$ to $G$
-is a function of type $\forallb{F \dotto G}$ (with naturality being trivial).
+is a function of type $F \rightarrowtriangle G$ (with naturality being trivial).
 Therefore, we expect a relative monad, given as a Kleisli triple, to have a unit
-$\eta_\Gamma : \forallb{\Gamma \ni \plr{-} \dotto \Gamma \vdash \plr{-}}$
+$\eta_\Gamma : \Gamma \ni \plr{-} \rightarrowtriangle \Gamma \vdash \plr{-}$
 given by the variable rule, and a Kleisli extension operator
 ${^*}_{\Gamma,\Delta} :
-\forallb{\Gamma \ni \plr{-} \dotto \Delta \vdash \plr{-}} \to
-\forallb{\Gamma \vdash \plr{-} \dotto \Delta \vdash \plr{-}}$
+\plr{\Gamma \ni \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}} \to
+\plr{\Gamma \vdash \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}}$
 given by substitution.
@@
 However, in the usage-aware case, substitution needs not just a mapping of
-variables $f : \forallb{\Gamma \sqni \plr{-} \dotto \Delta \vdash \plr{-}}$,
+variables $f : \Gamma \sqni \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}$,
 but rather an environment $\rho : \Delta \env\vdash \Gamma$, as we have already
@@
 
-\Cref{thm:env-comp-lemma} below amounts to deriving a modified notion of Kleisli
+\Cref{thm:env-comp} below amounts to deriving a modified notion of Kleisli
 composition from a modified Kleisli extension.
-Additionally, \cref{thm:env-id-lemma} is required to turn a monadic unit into an
+Additionally, \cref{thm:env-id} is required to turn a monadic unit into an
 identity environment.
@@
 
-\begin{lemma}[Identity environment]\label{thm:env-id-lemma}
+\begin{lemma}[Identity environment]\label{thm:env-id}
   Given a function
   \[
-    \mathrm{vr}_{\Gamma'} : \forallb{\Gamma' \sqni \plr{-} \dotto \V\,\Gamma'}
+    \mathrm{vr}_{\Gamma'} :
+    \Gamma' \sqni \plr{-} \rightarrowtriangle \V\,\Gamma'
   \]
@@
 
-\begin{lemma}[Composition of environments]\label{thm:env-comp-lemma}
+\begin{lemma}[Composition of environments]\label{thm:env-comp}
   Given a function
@@
     \mathrm{lift}_{\Gamma', \Delta'} :
-    \Gamma' \env\U \Delta' \to \forallb{\V\,\Delta' \dotto \W\,\Gamma'}
+    \Gamma' \env\U \Delta' \to \V\,\Delta' \rightarrowtriangle \W\,\Gamma'
   \]

### 2022-12-24 tex/lnd.tex
[ edit ] another draft of chapters 5 and 6

@@
-%This should stay fairly general, so it could be compiled to a DILL-style
-%calculus as well as the one with usage annotations.
-%Our framework compiles a ND system down to a sequent calculus.
-%It isolates the building blocks of typing rules ($\vdash$, $\wedge$, $*$,
-%$\Box$, $\cdot$) so that we can do more generic programming (e.g., usage
-%checker).
-
 The typing rules of $\name$ presented in \cref{fig:lr} contain a lot of detail
@@
     \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseUnit}
-    \noindent\ExecuteMetaData[\Bunchedtex]{forallb}
+    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseConjunction}
     \columnbreak
-    \AgdaNoSpaceAroundCode{}
-    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseConjunction}
-    \vspace{-0.25em}
-    \noindent\ExecuteMetaData[\Bunchedtex]{PointwiseImplication}
-    \AgdaSpaceAroundCode{}
+    %\AgdaNoSpaceAroundCode{}
+    \noindent\ExecuteMetaData[\Bunchedtex]{Entails}
+    %\AgdaSpaceAroundCode{}
   \end{multicols}
@@
 
-The first three connectives are those we've already seen for intuitionistic
-systems --- $\dot1$, $\dottimes$, and $\dotto$.
+The first two connectives are those we've already seen for intuitionistic
+systems --- $\dot1$ and $\dottimes$.
 The absence of premises is encoded by $\dot1$, while the space between premises
 sharing a context is encoded by $\dottimes$.
+As for implication, I temporarily avoid giving a $\dotto$ connectives, instead
+fusing it together with $\forallb{-}$ to produce the \emph{set} of
+context-preserving functions $T \rightarrowtriangle U$.
 When we interpret a typing rule as a constructor of an inductive definition,
-$\dotto$
+$\rightarrowtriangle$
 interprets the horizontal line, reflecting the fact that the usage annotations
@@
 
+\subsection{$\name$ stated using bunched connectives}\label{sec:lr-bunched}
 The full system \name{} is stated in terms of bunched connectives in
@@
 data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.
-
-While we have seen a connection between the bunched connectives and the
-connectives of $\name$, the two should not be confused.
-In particular, the bunched connectives obey different laws to what we would
-expect from linear logic.
-For example, it would make sense to define a bunched connective $\dotplus$,
-defined analogously to $\dottimes$.
-This $\dotplus$ could be used to rephrase the introduction rules for $\oplus$.
-We then have maps both ways between $T \dottimes (U \dotplus V)$ and
-$(T \dottimes U) \dotplus (T \dottimes V)$, reminiscent of the distributivity of
-additive connectives in bunched logic,
-whereas linear logic only has a map from $(A \with B) \oplus (A \with C)$ to
-$A \with (B \oplus C)$, and not a map the other way.
-Looking at the interpretations, the connection with bunched logic makes sense.
-Instead of the partial commutative monoid (often representing heaps) found in
-standard semantics of bunched logic, we have a left semimodule of usage
-contexts, which we are similarly interested in splitting and sharing between
-various subterms.
-
-A discrepancy with bunched logic is that $\dotto$ is not part of the Cartesian
-closure of context-indexed families.
-In fact, it fails to be a functor because it does not respect the ordering of
-usage contexts.
-This fact has little impact on the development of this thesis, because I mainly
-use $\dotto$ at the top level, directly in conjunction with $\forallb{-}$.
-The combination describes the ``external hom'', i.e.\ the morphisms in the
-category of context-indexed families, with the internal hom being more like the
-Kripke function space, which I do not use.
-
-%\todo{Split chapter here. Play around with other systems using bunched
-%  connectives. Lists/recursion, annotated arrow, Granule elim, sharing elim.}
 
@@
 \end{figure}
+
+\subsection{Connection with bunched logic}\label{sec:bunched-logic}
+While we have seen a connection between the bunched connectives and the
+connectives of $\name$, the two should not be confused.
+In particular, the bunched connectives obey different laws to what we would
+expect from linear logic.
+For example, it would make sense to define a bunched connective $\dotplus$,
+defined analogously to $\dottimes$.
+This $\dotplus$ could be used to rephrase the introduction rules for $\oplus$.
+We then have maps both ways between $T \dottimes (U \dotplus V)$ and
+$(T \dottimes U) \dotplus (T \dottimes V)$, reminiscent of the distributivity of
+additive connectives in bunched logic,
+whereas linear logic only has a map from $(A \with B) \oplus (A \with C)$ to
+$A \with (B \oplus C)$, and not a map the other way.
+Looking at the interpretations, the connection with bunched logic makes sense.
+Instead of the partial commutative monoid (often representing heaps) found in
+standard semantics of bunched logic, we have a left semimodule of usage
+contexts, which we are similarly interested in splitting and sharing between
+various subterms.
+
+From bunched logic, we would expect the Cartesian product $\dottimes$ to have an
+internal hom.
+In the intuitionistic case, $\dotto$ filled this role.
+However, with usage contexts, it makes sense for open types to be presheaves
+over the partial order of contexts under pointwise $\leq$ of usage annotations.
+The family $T \dotto U$ does not satisfy the functoriality condition because of
+the contravariance in the domain $T$.
+Instead, as found in models of bunched logic, we would want a Kripke function
+space, like
+$\lambda\Gamma.~\forall \Gamma' \leq \Gamma.~T\,\Gamma' \to U\,\Gamma'$.
+However, I do not make use of such a connective.
+
+The separating conjunction $\sep$ can be seen as a decategorified version of Day
+convolution~\citep{Day70}.
+It also resembles the use of ternary frames \todo{cite} in semantics of
+substructural logics.
+
+\subsection{Operations on bunched connectives}\label{sec:bunched-op}
+To manipulate terms and other open types defined using bunched connectives, we
+need the zero, addition, and multiplication relations to satisfy some laws.
+For example, to achieve a symmetry map $T \sep U \rightarrowtriangle U \sep T$,
+we need addition to satisfy the commutativity law
+$\forall x,y,z : A.~x \leq y + z \to x \leq z + y$.
+
+For all uses of bunched connectives in this thesis, the carrier set $A$ will
+form a partial order --- for example, with contexts, the order is given by the
+pointwise order on the usage vectors.
+We then consider the category whose objects are posets and whose morphisms are
+relations $R : A \rel B$ satisfying the contravariant-covariant law
+$\forall x,x',y,y'.~x' \leq x \to y \leq y' \to xRy \to x'Ry'$.
+This category can be given the usual monoidal product of relations, which is
+the pointwise product of posets on objects.
+Then, we will always expect zero and addition to together form a cocommutative
+comonoid in this category.
+With this structure, we can get the following equivalences and functions.
+
+\begin{mathpar}
+  I^* \sep T \leftrightarrowtriangle T \and
+  T \sep I^* \leftrightarrowtriangle T \and
+  (T \sep U) \sep V \leftrightarrowtriangle T \sep (U \sep V) \and
+  T \sep U \leftrightarrowtriangle U \sep T \and
+  (T \wand U) \sep T \rightarrowtriangle U \and
+  I^* \wand T \leftrightarrowtriangle T \and
+  (T \sep U) \wand V \leftrightarrowtriangle T \wand (U \wand V)
+\end{mathpar}
+
+I do not use algebraic properties of multiplication in conjunction with
+manipulation of bunched connectives in this thesis, but we could expect
+scalar multiplication
+to add a comodule structure over cosemiring $R$ to the cocommutative comonoid
+given by zero and addition.

### 2022-12-24 tex/lr.tex
[ edit ] another draft of chapters 5 and 6

@@
 
-  This posemiring can be used for sensitivity analysis, where we want to bound
-  the effect of a perturbation on inputs in terms of some semantic notion of
-  distance between values.
-  An annotation $\gr r$ says that if that input is perturbed by at most $r$,
-  then the output will change by at most $1$.
-  An $\gr\infty$-annotated variable gives the very strong guarantee that any
-  change in the input will make a minimal change to the output, while a
-  $\gr0$-annotated variable provides no guarantee at all.
+  The intended semantics for these annotations is to interpret types as metric
+  spaces and terms as Lipschitz-continuous maps.
+  That is to say, each type $A$ comes with a notion of distance $d_A$, and a map
+  $f : A \to B$ satisfies
+  \[
+    \exists\gr r \in \mathbb R^+.\forall x,y.~%
+    d_B(f(x), f(y)) \leq \gr rd_A(x, y).
+  \]
+  I internalise this constant $\gr r$ so that derivations of
+  $\gr rA \vdash B$ are interpreted as $\gr r$-Lipschitz-continuous functions.
+  If $\gr r = \gr\infty$, then the Lipschitz condition is trivially satisfied,
+  meaning that $\gr\infty$-annotated variables can be used without constraint.
+  If $\gr r = \gr0$, then $f$ has to be constant, and the corresponding
+  syntactic constraint is that a $\gr0$-annotated variable cannot be used.
+
   Addition forbids general contraction, which would otherwise allow arbitrary
@@
   weakening, so the resulting sensitivity calculus has an affine flavour.
+
+  Such a system has been applied to sensitivity analysis, a component of
+  differential privacy, by \citet{reed10distance}.
 \end{example}

### 2022-12-24 tex/lrsub.tex
[ edit ] another draft of chapters 5 and 6

@@
 
-Given a \AgdaRecord{Kit}, \cref{thm:lr-bind} looks like the following.
-The \AgdaField{lookup} clauses still contain essentially the same structure as
-in the intuitionistic case: discriminating on whether the variable is old or
-new, using the given environment \AgdaBound{$\rho$} and weakening on the old
-variables, and using \AgdaField{vr} to repackage new variables.
-I will not explain any of the algebraic manipulations here; see
-\cref{thm:lr-bind}.
+Given a \AgdaRecord{Kit}\AgdaSpace{}\AgdaBound{$\V$},
+\cref{thm:lr-bind} gives a function with the following type.
 
@@
 Finally, the simultaneous substitution \AgdaFunction{sub} specialises to
-single substitution \AgdaFunction{sub[-]}.
-Single substitution is stated as an admissible rule below.
-To substitute in for $\gr r$-many $A$ in the second term, we need to derive
-one $A$ with usages $\grP$, and then assert that the result can handle the
-usages of the original term $\grQ$, plus $\gr r$-many copies of $\grP$.
+single substitution.
 
-\[
+\begin{corollary}[Single substitution]\label{thm:single-sub}
+  The following equivalent rules are admissible.
+  \begin{mathpar}
     \ebrule{%
-    \hypo{\grR \leq \grQ + \gr r\grP}
+      \hypo{\grR \leq \gr r\grP + \grQ}
       \hypo{\grP\gamma \vdash A}
       \hypo{\grQ\gamma, \gr rA \vdash B}
-    \infer3[singleSub]{\grR\gamma \vdash B}
+      \infer3{\grR\gamma \vdash B}
     }
-\]
-
-The proof strategy for producing the substitution \AgdaFunction{$\sigma$} is
-to proceed structurally on the codomain context $\grQ\gamma, \gr rA$ using
-\cref{thm:construct-env}, applying the identity substitution
-\AgdaFunction{id\^{}Env} on the $\gamma$ half, and dropping the term
-\AgdaBound{t} in place of the variable we are substituting for.
-
-\ExecuteMetaData[\LRKits]{subSingle}
+    \and
+    \ebrule[comb]{%
+      \hypo{\gr r \cdot \plr{{} \vdash A}}
+      \hypo{\sep}
+      \hypo{\gr rA \vdash B}
+      \infer3{{} \vdash B}
+    }
+  \end{mathpar}
+\end{corollary}
+\begin{proof}
+  It is enough to construct a substitution of type
+  $\grR\gamma \env\vdash \grQ\gamma, \gr rA$.
+  To do this, we use \cref{thm:construct-env} cases $\sep(\rightarrowtriangle)$
+  and $\cdot(\rightarrowtriangle)$ on inequalities
+  $\grR \leq \grQ + \gr r\grP$ and $\gr r\grP \leq \gr r\grP$ respectively to
+  leave us needing a substitution of type $\grQ\gamma \env\vdash \grQ\gamma$ and
+  a term of type $\grP\gamma \vdash A$.
+  For the substitution, we give the identity substitution (\cref{thm:env-id}),
+  and we have the term as a hypothesis.
+\end{proof}

### 2022-12-24 tex/rec.tex
[ edit ] another draft of chapters 5 and 6

@@
 
-\subsection{Alternative object-language connectives}
+\subsection{Alternative object-language connectives}\label{sec:alt}
 
@@
 
-We can also give an alternative elimination rule for $\otimes$-products inspired
-by Granule~\citep{Granule18,HMWO21}.
+We can also give an alternative elimination rule for $\otimes$-products, as
+found in Granule~\citep{Granule18,HMWO21} and the work of
+\citet{AbelBernardy2020} and \citet{reed10distance}.
 This modified rule allows us to, for example, have a variable
@@
   \begin{align*}
-    \text{nil}\left[
+    \TirName{nil}\left[
     \begin{prooftree}
@@
     &&
-    \text{cons}\left[
+    \TirName{cons}\left[
     \begin{prooftree}
@@
   \begin{displaymath}
-    \text{foldr}\left[
+    \TirName{foldr}\left[
     \begin{prooftree}
@@
 
+Among the inductive types given by this $\mu$ operator are also
+\emph{$\with$-lists} (\emph{with-lists}).
+
+\begin{example}
+  Let us define \emph{$\with$-lists} as follows.
+  \[
+    \mathrm{WithList}_A \coloneqq \mu X.~\top \oplus \plr{A \with X}
+  \]
+  Unlike ordinary lists (which we may call \emph{$\otimes$-lists}),
+  $\with$-lists support a \emph{lookup} operation with the type below, where
+  $\mathbb N \coloneqq \mu X.~I \oplus X$.
+  \begin{align*}
+    \mathit{lookup} :
+    \mathrm{WithList}_A \multimap \mathbb N \multimap \top \oplus A
+  \end{align*}
+  The $\name$ term implementing $\mathit{lookup}$ in terms of iterators
+  (\TirName{$\mu$-E}) is complex and tedious, so I do not show it here.
+  Instead, I provide the following clausal specification, where
+  $\mathrm{nil}\plr{-}$ and $\mathrm{cons}\plr{-}$ are the constructors of
+  $\with$-lists, $\mathrm{zero}\plr{}$ and $\mathrm{succ}\plr{}$ are the
+  constructors of $\mathbb N$, $\mathrm{inl}\plr{-}$ and $\mathrm{inr}\plr{-}$
+  are the constructors of $\oplus$, $\langle\rangle$ constructs the
+  inhabitant of $\top$, and postfix $.\pi_l$ and $.\pi_r$ are the projections
+  out of a $\with$-pair.
+  %\begin{align*}
+  %  \mathit{lookup}\,\mathrm{nil}\plr{t}\,i
+  %  &\coloneqq \mathrm{inl}\plr{\langle\rangle} \\
+  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{zero}\plr{}
+  %  &\coloneqq \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
+  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{succ}\plr{i}
+  %  &\coloneqq \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i
+  %\end{align*}
+
+  %\noindent{}\hspace{-0.75ex}
+  \[
+    \begin{pmboxed}
+      \defaultcolumn{l}
+      \> \mathit{lookup} \> \mathrm{nil}\plr{t} \> i
+      \> \coloneqq \> \mathrm{inl}\plr{\langle\rangle} \\
+      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
+      \> \mathrm{zero}\plr{} \> \coloneqq
+      \> \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
+      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
+      \> \mathrm{succ}\plr{i} \> \coloneqq
+      \> \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i \\
+    \end{pmboxed}
+  \]
+  %\begin{verbatim}
+  %  lookup := \xs. wfoldr(xs;
+  %    List I -o T + A;
+  %    t. \i. foldr(i; T -o T + A; \t'. inl(t'); u, r. let () := u in r) t;
+  %    xr. \i. let (i', r) :=
+  %      foldr(i;
+  %        A & (T + A) -o List I * (T + A);
+  %        \xr'. ([], prl(xr'));
+  %        u, ir. \xr'.
+  %          let () := u in
+  %          let (i', r) := ir in
+  %          (() :: i', prr(xr') prl(ir)))
+  %      xr in
+  %      foldr(i'; T + A; r; u, r'. let () := u in r'))
+  %\end{verbatim}
+  %\begin{align*}
+  %  \mathit{lookup} &:
+  %  \mathrm{WithList}_A \multimap \mathrm{List}_I \multimap \top \oplus A
+  %  \\
+  %                  &\coloneqq \lambda\mathit{xs}.~\mathrm{rec}\plr{xs; %
+  %                    \mathrm{List}_I \multimap \top \oplus A; %
+  %                    t.~\lambda i.~\mathrm{inl}\plr{%
+  %                    \mathrm{foldr}\plr{i, \top \multimap \top; \lambda t'.~t'; %
+  %                    u, r.~\textrm{let }{*} \coloneqq u\textrm{ in }r}\,t}; %
+  %                    \mathit{xr}.~\lambda i.~%
+  %                    \mathrm{foldr}\plr{i; %
+  %                    A \with \plr{\top \oplus A} \multimap \top \oplus A; %
+  %                    \lambda
+  %                    }\,\mathit{xr}}
+  %\end{align*}
+\end{example}
+
 As suggested by the notation, $\Box^{0{+}}$ may not be particularly canonical in
@@
 being less than or equal to $\gr0$.
+
+A similar modality, $\Box$, appears in the work of \citet{BB18} when they deal
+with \emph{persistent predicates} in bunched logic.
+They similarly aim to pick out those objects not depending on any external
+resources.

### 2022-12-24 tex/semirings.tex
[ edit ] another draft of chapters 5 and 6

@@
 In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
-appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
-simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
+appeared in my previous work~\citep{WA21}, and can be seen as a
+simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
 The idea of \name{} is to augment the simply typed $\lambda$-calculus with
@@
 examples in \cref{sec:rec}.
-The application of the techniques of \cref{sec:simple} is left to
-\cref{sec:ren-sub-lr}.
+Finally, I justify connections to linear logic and modal logic in
+\cref{sec:translation}, where I translate $\name$ terms to and from
+DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
+%The application of the techniques of \cref{sec:simple} is left to
+%\cref{sec:ren-sub-lr}.
 
@@
 \input{rec.tex}
+\section{Translation to and from existing systems}\label{sec:translation}
+\input{trans.tex}
 \section{Addendum: (lack of) partiality}\label{sec:part}
@@
 In \cref{sec:semirings}, I defined my calculus of interest $\name$.
-In this chapter, I develop its syntactic metatheory up to and including
-substitution.
+In this chapter, I develop the necessary syntactic metatheory for
+specifying and implementing the substitution operation.
 I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
@@
 \input{lrsub.tex}
+\section{Conclusion}\label{sec:ren-sub-lr-conc}
+
+In the preceding two chapters, I have developed a discipline for specifying
+the syntax of linear and modal type systems, and furthermore developing the
+syntactic metatheory of those type systems.
+All of these are based on semirings, and the linear algebra arising from
+considering a usage context full of semiring elements as a vector.
+
+These developments can be seen in retrospect as a generalisation of the methods
+explained in \cref{sec:simple}.
+In terms of premise connectives in the syntactic rules, we have generalised from
+just $\{\dot1, \dottimes\}$ to
+$\{\dot1, \dottimes, I^*, \sep, \gr r\cdot{}, \Box^{0{+}}\}$, maintaining our
+ability to keep the context implicit.
+Similarly to how rule premises can require separation of usage annotations, our
+new environments can require such a separation between their entries thanks to
+the linear map they now contain.
+I have generalised the key property of a kit from arbitrary weakening to
+weakening by $\gr0$-annotated variables, and using that have produced a
+substitution operation based on the same principles as that from
+\cref{sec:kits}.
+
+Having generalised all of the components --- namely the contexts, the syntax,
+and the notion of environment --- the type of the substitution operation looks
+the same as it did for intuitionistic ST$\lambda$C\@.
+Being able to maintain this uniformity is a key step towards generalising the
+rest of \cref{sec:simple} (i.e., \cref{sec:gen-sem,sec:gen-syn}), as I do in
+\cref{sec:framework}.
+
+A similar substitution lemma appears in the PhD thesis of
+\citet[p.\ 138]{petricek-thesis} under the name \emph{multi-nary substitution}.
+In my notation, \citeauthor{petricek-thesis}'s substitution rule looks like the
+following, up to permutation of the contexts containing $\Gamma$.
+Note that if $\Delta = \grQ\delta$, then $\gr r\Delta$ denotes the context
+$\plr{\gr r\grQ}\delta$.
+This rule is essentially an iterated version of the standard linear single
+substitution principle, and is used by \citeauthor{petricek-thesis} as a
+strengthened induction hypothesis required to derive single substitution.
+
+\[
+  \ebrule{%
+    \hypo{\Delta_1 \vdash A_1}
+    \hypo{\cdots}
+    \hypo{\Delta_n \vdash A_n}
+    \hypo{\Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n \vdash B}
+    \infer4{\Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n \vdash B}
+  }
+\]
+
+We can derive \citeauthor{petricek-thesis}-style multi-nary substitution as a
+corollary of my simultaneous substitution, using reasoning similar to that of
+\cref{thm:single-sub}.
+
+\begin{corollary}\label{thm:petricek-sub}
+  \Citeauthor{petricek-thesis}'s multi-nary substitution, as stated above, is
+  admissible in $\name$.
+\end{corollary}
+\begin{proof}
+  It is enough to provide a substitution of type
+  \[
+    \Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n
+    \env\vdash \Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n.
+  \]
+  To do this, we use \cref{thm:construct-env} repeatedly, leaving us needing a
+  substitution of type
+  $\Gamma, \gr0\Delta_1, \ldots, \gr0\Delta_n \env\vdash \Gamma$ and terms of
+  types
+  \begin{align*}
+    \gr0\gamma, \Delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \gr0\delta_n
+    \vdash A_1 \\
+    &\vdots \\
+    \gr0\gamma, \gr0\delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \Delta_n
+    \vdash A_n.
+  \end{align*}
+  The identity substitution and weakening by $\gr0$-annotated variables is
+  enough to make these requirements line up with the given hypotheses.
+\end{proof}
+
+My substitution principle is stronger than \citeauthor{petricek-thesis}'s.
+Where \citeauthor{petricek-thesis} requires that distinct variables be
+available for each hypothesis, I allow for separation of uses via addition of
+contexts.
+Below is a prototypical example.
+
+\begin{example}
+  Let $\Ann \coloneqq \plr{\mathbb N, =, 0, +, 1, \times}$, the exact
+  usage-counting posemiring.
+  Then, we can construct a substitution $\rho : \gr2A \env\vdash \gr1A, \gr1A$,
+  yielding a transformation of terms of the following form:
+  \[
+    \ebrule{%
+      \hypo{\gr1A, \gr1A \vdash B}
+      \infer1{\gr2A \vdash B}
+    }.
+  \]
+  To construct $\rho$, we use \cref{thm:construct-env} case
+  $\sep(\rightarrowtriangle)$, using the fact that $\gr2 \leq \gr1 + \gr1$.
+  From there, two identity substitutions suffice.
+  The action of $\rho$ on terms is to merge the two variables into one.
+  Note that a renaming, rather than a substitution, would also suffice.
+\end{example}
+
+Most notably, my (single) substitution principle more naturally fits the
+requirement we would have for the reduct of the $\beta$-rule for functions in
+$\name$, whereas \citeauthor{petricek-thesis}'s substitution principle would
+need some additional transformation for it to fit properly.
+This comes from the fact that the $\name$ function application rule introduces
+an algebraic ($+$) separation between its premises, whereas
+\citeauthor{petricek-thesis}'s substitution principle separates premises only
+via concatenation.

### 2022-12-24 tex/trans.tex
[ edit ] another draft of chapters 5 and 6

\newcommand\instDILL{\gr{01\omega}}
\newcommand\instPD{\gr{01\Box}}

\newcommand\unused{\gr0}
\newcommand\true{\gr1}
\newcommand\valid{\gr\Box}

A motivating reason to consider the system $\name$ is that
instances of it correspond to previously studied systems.
In this section, I present translations from \name{} to Dual Intuitionistic
Linear Logic~\citep{Barber1996} and the modal system of \citet{judgmental},
and vice versa.
These translations are not mechanised, as part of the reason for developing
\name{} was to avoid mechanising these systems directly.
We cannot prove that the translations form an equivalence, because I have not
written down an equational theory for \name{}, but I expect this to be easy
enough to do.

\subsection{Dual Intuitionistic Linear Logic}\label{sec:dill}

Dual Intuitionistic Linear Logic is a particular formulation of intuitionistic
linear logic introduced by \citet{Barber1996}.
Its key feature, which simplifies the metatheory of linear logic, is the use of
separate contexts for linear and intuitionistic free variables.
Here I show that DILL is a fragment of the instantiation of \name{} at the
linearity semiring $\{\gr0, \gr1, \gr\omega\}$.

The types of DILL are the same as the types of \name, except for the
restriction of $\oc\gr{r}$ to just $\oc\gr\omega$.
I will write the latter simply as $\oc$ when it appears in DILL\@.
I add sums and with-products to the calculus of \cite{Barber1996}, with the
obvious rules (stated fully in \cref{fig:dill}).
These additive type formers present no additional difficulty to the translation.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=Int-Ax]{ }
    {\gamma, A; \cdot \vdash A}
    \and
    \inferrule*[right=Lin-Ax]{ }
    {\gamma; A \vdash A}
    \and
    \inferrule*[right=$I$-I]{ }
    {\gamma; \cdot \vdash I}
    \and
    \inferrule*[right=$I$-E]
    {\gamma; \delta_1 \vdash I \\ \gamma; \delta_2 \vdash A}
    {\gamma; \delta_1, \delta_2 \vdash A}
    \and
    \inferrule*[right=$\otimes$-I]
    {\gamma; \delta_1 \vdash A \\ \gamma, \delta_2 \vdash B}
    {\gamma; \delta_1, \delta_2 \vdash A \otimes B}
    \and
    \inferrule*[right=$\otimes$-E]
    {\gamma; \delta_1 \vdash A \otimes B \\ \gamma; \delta_2, A, B \vdash C}
    {\gamma; \delta_1, \delta_2 \vdash C}
    \and
    \inferrule*[right=$\multimap$-I]
    {\gamma; \delta, A \vdash B}
    {\gamma; \delta \vdash A \multimap B}
    \and
    \inferrule*[right=$\multimap$-E]
    {\gamma; \delta_1 \vdash A \multimap B \\ \gamma; \delta_2 \vdash A}
    {\gamma; \delta_1, \delta_2 \vdash B}
    \and
    \inferrule*[right=$\oc$-I]
    {\gamma; \cdot \vdash A}
    {\gamma; \cdot \vdash \oc A}
    \and
    \inferrule*[right=$\oc$-E]
    {\gamma; \delta_1 \vdash \oc A \\ \gamma, A; \delta_2 \vdash B}
    {\gamma; \delta_1, \delta_2 \vdash B}
    \and
    \inferrule*[right=$\top$-I]{ }
    {\gamma; \delta \vdash \top}
    \and
    \inferrule*[right=$\with$-I]
    {\gamma; \delta \vdash A \\ \gamma, \delta \vdash B}
    {\gamma; \delta \vdash A \with B}
    \and
    \inferrule*[right=$\with$-E$_i$]
    {\gamma; \delta \vdash A_0 \with A_1}
    {\gamma; \delta \vdash A_i}
    \and
    \inferrule*[right=$0$-E]
    {\gamma; \delta_1 \vdash 0}
    {\gamma; \delta_1, \delta_2 \vdash A}
    \and
    \inferrule*[right=$\oplus$-I$_i$]
    {\gamma; \delta \vdash A_i}
    {\gamma; \delta \vdash A_0 \oplus A_1}
    \and
    \inferrule*[right=$\oplus$-E]
    {
      \gamma; \delta_1 \vdash A \oplus B \\
      \gamma; \delta_2, A \vdash C \\
      \gamma; \delta_2, B \vdash C
    }
    {\gamma; \delta_1, \delta_2 \vdash C}
  \end{mathpar}
  \label{fig:dill}
  \caption{The rules of DILL, extended with additive connectives}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \begin{eqns}
      \mathrm{DILL} &\hookrightarrow& \name_\instDILL \\
      Y &\mapsto& \iota_Y \\
      I &\mapsto& I \\
      A \otimes B &\mapsto& A \otimes B \\
      A \multimap B &\mapsto& A \multimap B \\
      \oc A &\mapsto& \oc{\gr\omega}{A} \\
      0 &\mapsto& 0 \\
      A \oplus B &\mapsto& A \oplus B \\
      \top &\mapsto& \top \\
      A \with B &\mapsto& A \with B
    \end{eqns}
    \and
    \begin{eqns}
      \mathrm{PD} &\hookrightarrow& \name_\instPD \\
      Y &\mapsto& \iota_Y \\
      \top &\mapsto& I \\
      A \wedge B &\mapsto& A \with B \\
      A \supset B &\mapsto& A \multimap B \\
      \Box A &\mapsto& \oc{\valid}{A} \\
      \bot &\mapsto& 0 \\
      A \vee B &\mapsto& A \oplus B
    \end{eqns}
  \end{mathpar}
  \label{fig:dill-pd}
  \caption{Embedding of DILL and PD types into \name}
\end{figure}

\begin{proposition}[DILL $\to$ \name]
  Given a DILL derivation of $\gamma; \delta \vdash A$, we can produce a
  $\name_{\instDILL}$ derivation of
  $\gr\omega\gamma, \gr1\delta \vdash A$.
\end{proposition}
\begin{proof}
  By induction on the derivation.
  We have $\gr\omega \leq \gr0$, which allows us to discard
  intuitionistic variables at the var rules, and both
  $\gr1 \leq \gr1$ and $\gr\omega \leq \gr1$, which allow
  us to use both linear and intuitionistic variables.

  Weakening is used when splitting linear variables between two premises.
  For example, \TirName{$\otimes$-I} in DILL is as follows.
  \[
    \inferrule*[right=$\otimes$-I]
    {\gamma; \delta_t \vdash t : A \\ \gamma; \delta_u \vdash u : B}
    {\gamma; \delta_t, \delta_u \vdash t \otimes u : A \otimes B}
  \]
  From this, our new derivation is as follows.
  \[
    \inferrule*[right=$\otimes$-I]
    {
      \inferrule*[right=Weak]
      {\mathit{ih}_t \\\\
        \gr\omega\gamma, \gr1\delta_t
        \vdash M_t : A}
      {\gr\omega\gamma, \gr1\delta_t,
        \gr0\delta_u
        \vdash M_t : A}
      \\
      \inferrule*[right=Weak]
      {\mathit{ih}_u \\\\
        \gr\omega\gamma, \gr1\delta_u
        \vdash M_u : A}
      {\gr\omega\gamma, \gr0\delta_t,
        \gr1\delta_u
        \vdash M_u : A}
    }
    {\gr\omega\gamma, \gr1\delta_t,
      \gr1\delta_u
      \vdash \plr{M_t, M_u} : A \otimes B}
  \]
\end{proof}

When translating from \name{} to DILL, we first coerce the \name{} derivation
to be in a form easily amenable to translation into DILL\@.
An example of a \name{} derivation with no direct translation into DILL is the
following.
In DILL terms, the intuitionistic variable of the conclusion becomes a linear
variable in the premises.
Such a move is admissible in DILL, but does not come naturally.

\[
  \inferrule*[right=$\otimes$-I]
  {
    \inferrule*[right=var]{ }
    {\gr1A : x \vdash A}
    \\
    \inferrule*[right=var]{ }
    {\gr1A : x \vdash A}
    \\
    \gr\omega \leq \gr1 + \gr1
  }
  {\gr\omega A : x \vdash A \otimes A}
\]

To avoid such situations, and therefore manipulations on DILL derivations, I
show that all $\name_{\instDILL}$ derivations can be made in \emph{bottom-up}
style.
In bottom-up style, the algebraic facts we make use of are dictated by making
most general choices based on the conclusions of rules.
Bottom-up style corresponds to a (non-deterministic) form of
\emph{usage checking}, and the following lemma can be understood as saying
that that form of usage checking is sufficiently general.

\begin{definition}
  A derivation is said to be \emph{$\instDILL$-bottom-up} if only the following
  facts about addition and multiplication are used, and all proofs of
  inequalities not at leaves are by reflexivity (i.e, not using the facts that
  $\gr\omega \leq \gr0$ and $\gr\omega \leq \gr1$).

  \makebox[\textwidth][s]{
    \begin{tabular}{c|ccc}
      $+$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
      $\gr0$ & $\gr0$ & $\gr1$ & - \\
      $\gr1$ & $\gr1$ & - & - \\
      $\gr\omega$ & - & - & $\gr\omega$ \\
    \end{tabular}
    \begin{tabular}{c|ccc}
      $*$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
      $\gr0$ & - & - & $\gr0$ \\
      $\gr1$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
      $\gr\omega$ & $\gr0$ & - & $\gr\omega$ \\
    \end{tabular}
  }
\end{definition}

Bottom-up style enforces that whenever we split a context into two (for
example, in the rule \TirName{$\otimes$-I}) all unused variables in the
conclusion stay unused in the premises, intuitionistic variables stay
intuitionistic, and linear variables go either left or right.
Multiplication is only used in the rule \TirName{$\oc\gr{r}$-I}, at which point
both the result and left argument are available.
Here, the bottom-up style enforces that linear variables never appear in the
premise of \TirName{$\oc{\gr\omega}$-I}.

\begin{lemma}
  Every $\name_{\instDILL}$ derivation can be translated into a bottom-up
  $\name_{\instDILL}$ derivation.
\end{lemma}
\begin{proof}
  By induction on the shape of the derivation.
  When we come across a non-bottom-up use of addition, it must be that the
  corresponding variable in the conclusion has annotation $\gr\omega$.
  By subusaging, we can give this variable annotation $\gr\omega$ in
  the premises too, before translating the subderivations to bottom-up
  style.
  A similar argument applies to uses of multiplication, remembering that both
  the left argument and result are fixed.
\end{proof}

\begin{proposition}[\name{} $\to$ DILL]
  Given a $\name_{\instDILL}$ derivation of
  $\gr\omega\gamma, \gr1\delta,
  \gr0\theta \vdash A$ which contains only types expressible
  in DILL, we can produce a DILL derivation of $\gamma; \delta \vdash A$.
\end{proposition}
\begin{proof}
  By induction on the derivation having been translated to bottom-up form.

  In the case of \TirName{var}, all of the unused variables have annotation
  greater than $\gr0$, i.e., $\gr0$ or $\gr\omega$.
  Those annotated $\gr0$ are absent from the DILL derivation, and those
  annotated $\gr\omega$ are in the intuitionistic context.
  The used variable is annotated either $\gr1$ or $\gr\omega$.
  In the first case, we use \TirName{Lin-Ax}, and in the second case,
  \TirName{Int-Ax}.

  All binding of variables in \name{} maps directly onto DILL\@.

  Because we translated to bottom-up form, additions, as seen in, for example,
  the \TirName{$\otimes$-I} rule, can be handled straightforwardly.
  Any intuitionistic variables in the conclusion correspond to intuitionistic
  variables in both premises.
  Any linear variables in the conclusion correspond to a linear variable in
  exactly one of the premises, and is absent in the other premise.

  The only remaining rule is \TirName{$\oc\gr{r}$-I}, of which we only cover
  \TirName{$\oc{\gr\omega}$-I} (the other two targeting types not found
  in DILL).
  In this case, we know that every variable in the conclusion is annotated
  either $\gr0$ or $\gr\omega$, and every variable in the premise is
  annotated the same way.
  This corresponds exactly to the restrictions of DILL's \TirName{$\oc$-I}.
\end{proof}

\subsection{Pfenning Davies}\label{sec:pd}

The translation to and from the modal system of Pfenning and Davies
\cite{judgmental} (henceforth \emph{PD}) is similar to the translation to and
from DILL\@.
I present my variant of PD, again adding some common connectives, in
\cref{fig:pd}
The main difference is the algebra at which \name{} is instantiated.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=hyp]{ }
    {\gamma; \delta, A\;\mathit{true} \vdash A\;\mathit{true}}
    \and
    \inferrule*[right=hyp*]{ }
    {\gamma, A\;\mathit{valid}; \delta \vdash A\;\mathit{true}}
    \and
    \inferrule*[right=$\supset$I]
    {\gamma; \delta, A\;\mathit{true} \vdash B\;\mathit{true}}
    {\gamma; \delta \vdash A \supset B\;\mathit{true}}
    \and
    \inferrule*[right=$\supset$E]
    {
      \gamma; \delta \vdash A \supset B\;\mathit{true} \\
      \gamma; \delta \vdash A\;\mathit{true}
    }
    {\gamma; \delta \vdash B\;\mathit{true}}
    \and
    \inferrule*[right=$\Box$I]
    {\gamma; \cdot \vdash A\;\mathit{true}}
    {\gamma; \delta \vdash \Box A\;\mathit{true}}
    \and
    \inferrule*[right=$\Box$-E]
    {
      \gamma; \delta \vdash \Box A\;\mathit{true} \\
      \gamma, A\;\mathit{valid}; \delta \vdash B\;\mathit{true}
    }
    {\gamma; \delta \vdash B\;\mathit{true}}
    \and
    \inferrule*[right=$\top$-I]{ }
    {\gamma; \delta \vdash \top\;\mathit{true}}
    \and
    \inferrule*[right=$\wedge$-I]
    {
      \gamma; \delta \vdash A\;\mathit{true} \\
      \gamma, \delta \vdash B\;\mathit{true}
    }
    {\gamma; \delta \vdash A \wedge B\;\mathit{true}}
    \and
    \inferrule*[right=$\wedge$-E$_i$]
    {\gamma; \delta \vdash A_0 \wedge A_1\;\mathit{true}}
    {\gamma; \delta \vdash A_i\;\mathit{true}}
    \and
    \inferrule*[right=$\bot$-E]
    {\gamma; \delta \vdash \bot\;\mathit{true}}
    {\gamma; \delta \vdash A\;\mathit{true}}
    \and
    \inferrule*[right=$\vee$-I$_i$]
    {\gamma; \delta \vdash A_i\;\mathit{true}}
    {\gamma; \delta \vdash A_0 \vee A_1\;\mathit{true}}
    \and
    \inferrule*[right=$\vee$-E]
    {
      \gamma; \delta \vdash A \vee B\;\mathit{true} \\
      \gamma; \delta, A \vdash C\;\mathit{true} \\
      \gamma; \delta, B \vdash C\;\mathit{true}
    }
    {\gamma; \delta \vdash C\;\mathit{true}}
  \end{mathpar}
  \label{fig:pd}
  \caption{The rules of PD, extended with several standard connectives}
\end{figure}

\begin{definition}
  Let $\instPD$ denote the following semiring on the partially ordered set
  $\{\valid \triangleleft \true \triangleleft \unused\}$.
  \begin{itemize}
    \item $0 := \unused$.
    \item $+$ is the meet ($\wedge$) according to the subusaging order.
    \item $1 := \true$.
    \item
      \begin{tabular}{c|ccc}
        $*$ & $\unused$ & $\true$ & $\valid$ \\ \hline
        $\unused$ & $\unused$ & $\unused$ & $\unused$ \\
        $\true$ & $\unused$ & $\true$ & $\valid$ \\
        $\valid$ & $\unused$ & $\valid$ & $\valid$ \\
      \end{tabular}
  \end{itemize}
\end{definition}

The $\unused$ annotation plays only a formal role in this example.
Meanwhile, $\true$ and $\valid$ correspond to the judgement forms
$\mathit{true}$ and $\mathit{valid}$ from PD\@.
Addition being the meet makes it idempotent.
Furthermore, it gives us that $\true + \valid = \valid$ --- if somewhere we
require an assumption to be true, and elsewhere require it to be valid, then
ultimately it must be valid (from which we can deduce that it is true).
Multiplication is designed to make $\oc{\valid}$ act like PD's $\Box$.
In particular, $\valid * \valid = \valid$ says that the valid assumptions are
available before and after \TirName{$\oc{\valid}$-I}, whereas
$\valid * \true = \valid$ says that valid assumptions in the conclusion can be
weakened to true assumptions in the premise.
The latter fact does not appear in PD, and will be excluded from
\emph{bottom-up} derivations.

To keep my notation consistent with that of DILL, I swap the roles of
$\gamma$ and $\delta$ in PD compared to what they were in the original paper.
Thus, my PD judgements are of the form $\gamma; \delta \vdash A~\mathit{true}$,
where $\gamma$ contains valid assumptions and $\delta$ contains true
assumptions.

\begin{proposition}[PD $\to$ \name]
  Given a PD derivation of $\gamma; \delta \vdash t : A~\mathit{true}$, we can
  produce a $\name_{\instPD}$ derivation of
  $\valid\gamma, \true\delta \vdash A$.
\end{proposition}
\begin{proof}
  By induction on the PD derivation.
  Most PD rules have direct $\name$ counterparts, noting that variables of any
  annotation can be discarded and duplicated because we have both
  $\gr r \leq \gr 0$ and
  $\gr r \leq \gr r + \gr r$ for all
  $\gr r$.

  Care must be taken with the \TirName{$\Box$I} rule.
  We have, from the induction hypothesis, a $\name$ derivation of
  $\gr\Box\gamma \vdash A$.
  By \TirName{$\oc\valid$-I}, we have
  $\gr\Box\gamma \vdash \oc\valid A$.
  To get the desired conclusion, we must use \TirName{Weak} to get
  $\gr\Box\gamma, \gr\unused\delta \vdash \oc\valid A$, and
  then \TirName{Subuse} on the variables we just introduced (noting that
  $\true \leq \unused$) to get
  $\gr\Box\gamma, \gr\true\delta \vdash \oc\valid A$.
\end{proof}

For translating from $\name_{\instPD}$ to PD, I introduce a similar notion of
\emph{bottom-up} derivations as I did for DILL\@.
Every $\name_{\instPD}$ derivation can be translated into bottom-up style, and
then be directly translated into PD\@.

\begin{definition}
  A derivation is said to be \emph{$\instPD$-bottom-up} if only the following
  facts about addition and multiplication are used, and all proofs of
  inequalities not at leaves are by reflexivity.

  \makebox[\textwidth][s]{
    \begin{tabular}{c|ccc}
      $+$ & $\unused$ & $\true$ & $\valid$ \\ \hline
      $\unused$ & $\unused$ & - & - \\
      $\true$ & - & $\true$ & - \\
      $\valid$ & - & - & $\valid$ \\
    \end{tabular}
    \begin{tabular}{c|ccc}
      $*$ & $\unused$ & $\true$ & $\valid$ \\ \hline
      $\unused$ & - & - & $\unused$ \\
      $\true$ & $\unused$ & $\true$ & $\valid$ \\
      $\valid$ & $\unused$ & - & $\valid$ \\
    \end{tabular}
  }
\end{definition}

\begin{lemma}
  Every $\name_{\instPD}$ derivation can be translated into a bottom-up
  $\name_{\instPD}$ derivation.
\end{lemma}
\begin{proof}
  By induction on the shape of the derivation.
  Given that addition is a meet, it is clear that any non-bottom-up uses of
  addition come from one of the arguments being greater than the result.
  Therefore, it is safe to make this argument smaller in the corresponding
  premise (via subusaging), before translating that subderivation.
  For multiplication, again, there is always a lesser value of the right
  argument that will take us from a non-bottom-up fact to a bottom-up fact with
  the same left argument and result.
\end{proof}

\begin{proposition}[\name{} $\to$ PD]
  Given a $\name_{\instPD}$ derivation of
  $\valid\gamma, \true\delta, \unused\theta
  \vdash M : A$ which does not contain types using $\oc{\unused}{}$ or
  $\oc{\true}$, we can produce a PD derivation of
  $\gamma; \delta \vdash A~\mathit{true}$.
\end{proposition}
\begin{proof}
  We translate away tensor products and tensor units using
  \cref{thm:top-meet}, and translate the resulting derivation to bottom-up
  form.
  The proof proceeds by induction on the resulting derivation in the obvious
  way.
\end{proof}

As mentioned in \cref{sec:alt}, the system of \citet{AbelBernardy2020}
is unable to embed PD in this way, as it would prove
$\Box(A \wedge B) \to \Box A \wedge \Box B$, where PD and $\name$ do not.
In fact, this example shows that, even when weakening and contraction are
admissible, with- and tensor-products are distinct in their system in the
presence of modalities.

### 2023-01-10 tex/framework.tex
[ writing ] renew generic syntax chapter

@@
-\chapter{A framework for usage-restricted calculi}\label{sec:framework}
+\chapter{Generic usage-annotated syntax}\label{sec:framework}
 
@@
 by \citet{AACMM21} and apply the principles we discovered in
-\cref{sec:semirings} to yield a framework allowing for semiring-based usage
-restrictions on variables.
+\cref{sec:semirings} to yield a framework for defining calculi with
+semiring-based usage restrictions on variables.
 Syntactically, I claim that this framework ranges over all finitiary
@@
 \cref{sec:other-syntaxes}.
-I also derive analogues to some of the semantic results of \citet{AACMM21},
-strengthening them to take advantage of usage restrictions.
+%I also derive analogues to some of the semantic results of \citet{AACMM21},
+%strengthening them to take advantage of usage restrictions.
 
@@
 precise about the various levels of domain-specific languages which appear.
-I assume that the reader is familiar with the bunched connectives introduced in
-\cref{fig:bunched} and the usage-aware environments of \cref{def:lr-env}.
-
-\section{Syntax}
 
@@
 
-\section{Semantics}
+\chapter{Generic usage-aware semantics}\label{sec:semantics}
 

### 2023-01-10 tex/intro.tex
[ more ] thesis writing
[ writing ] renew generic syntax chapter

--- 2022-01-25 tex/esop22/intro.tex
@@
-%\begin{itemize}
-%  \item Review standard linear algebra constructions in the posemiring setting.
-%  \item Give a general definition of simple type systems with usage annotations
-%    based on connectives from bunched logic and constructions from linear
-%    algebra.
-%  \item Give some novel examples of such systems.
-%  \item Derive admissibility of renaming and substitution for all such systems.
-%  \item Give a usage checker generic in syntax.
-%\end{itemize}
+\chapter{Introduction}
+
+FIXME: this chapter is directly from the ESOP22 paper.
 
@@
 We build upon both the general structural framework of
-\citet{AACMM21} and the substructural techniques of \citet{WA20}.
+\citet{AACMM21} and the substructural techniques of \citet{WA21}.
 The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
@@
 Finally, we conclude and discuss future work in \cref{sec:conc}
+
+The work presented in this paper has been mechanised in Agda,
+with the code available for building upon~\cite{generic-lr}.

### 2023-01-10 tex/semantics.tex
[ edit ] some corrections and notes after meeting
[ writing ] renew generic syntax chapter

@@
 going under variable binders (\cref{sec:kripke}).
+\todo{Introduction for the chapter}
 
@@
 
-\subsection{A layer of syntax is functorial}\label{sec:functorial}
-
-A basic property of the universe of syntaxes we described in \cref{sec:syntax}
-is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
-Its type says that to map a function \AgdaBound{f}
-over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
-domain and codomain usage contexts, and \AgdaBound{f} should be usable
-wherever the domain and codomain usage contexts are similarly related by
-\AgdaBound{F}.
-
-\ExecuteMetaData[\Maptex]{map-s-type}
-
-This generality is needed because usage contexts change between
-a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
-\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
-\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
-variables newly bound in that subterm).
-Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
-This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.
-
-The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
-works through the structure \AgdaBound{ps} of premises of the rule applied,
-acting on each subterm it finds.
-Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
-\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
-function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
---- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
-and \AgdaBound{Q$'$}.
-In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
-$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
-collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
-Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
-$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
-$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
-The cases for \AgdaInductiveConstructor{`$\cdot$} and
-\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
-of linearity.
-In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
-\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
-systems, do not involve any changes in the usage contexts.
-
-\ExecuteMetaData[\Maptex]{map-p}
-
-\subsection{The Kripke function space}\label{sec:kripke}
+\section{The Kripke function space}\label{sec:kripke}
 
@@
 
-\subsection{Semantic traversal}\label{sec:traversal}
+\section{Semantic traversal}\label{sec:traversal}
 

### 2023-01-10 tex/syntax.tex
[ writing ] renew generic syntax chapter

@@
-The aim of this section is to produce a domain-specific language of
+The aim of this chapter is to produce a domain-specific language of
 \emph{syntax descriptions}.
@@
 
-\subsection{Descriptions of Systems}
+\section{Descriptions of Systems}
 
@@
 A \AgdaRecord{Rule} is a pair of some \AgdaDatatype{Premises} and a
-conclusion. We use an infix arrow as a suggestive notation for rules.
+conclusion.
+I suggestively use a quoted version of the ``universal entailment'' arrow
+$\rightarrowtriangle$, the unquoted version of which interprets the horizontal
+line in a traditionally presented typing rule.
 
@@
   \ExecuteMetaData[\PaperExamplestex]{Ty}
-  \ExecuteMetaData[\PaperExamplestex]{Side}
+  \ExecuteMetaData[\Handtex]{Hand}
 \end{minipage}
@@
 
-\subsection{Terms of a \AgdaRecord{System}}\label{sec:terms}
+\section{Terms of a \AgdaRecord{System}}\label{sec:terms}
 
@@
 
-\subsection{Other syntaxes and syntactic forms}\label{sec:other-syntaxes}
+\section{More example syntaxes}\label{sec:other-syntaxes}
 
-\paragraph{The system $\mu\tilde\mu$.}
-We can encode a usage-annotated version of System $L$/the
-$\mu\tilde\mu$-calculus~\cite{CH00} --- a syntax for classical logic --- in
+With the range of representable syntaxes now formalised, we can explore encoding
+techniques for syntaxes more exotic than ST$\lambda$C and $\name$.
+As well as the variations presented in \cref{sec:variant}, we can represent a
+usage-annotated $\mu\tilde\mu$-calculus and a Linear/non Linear system.
+
+\subsection{The system $\mu\tilde\mu$}
+I encode a usage-annotated version of System $L$/the
+$\mu\tilde\mu$-calculus~\citep{CH00} --- a syntax for classical logic --- in
 such a way that contexts capture the undistinguished parts of the sequent.
@@
 Though the $\mu\tilde\mu$-calculus is originally described as a sequent
-calculus~\cite{CH00}, we use the techniques of
-\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it as a natural
-deduction system, thus giving a notion of \emph{variable} to the system.
+calculus~\citep{CH00}, I use the techniques of
+\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it using hypothetical
+judgements, thus giving a notion of \emph{variable} to the system.
 
@@
 \AgdaDatatype{Ty} not being exposed directly to the generic framework.
-For now, we just consider multiplicative disjunction $\parr$ (\emph{par}) and
+For now, I just consider multiplicative disjunction $\parr$ (\emph{par}) and
 negation/duality, beside an uninterpreted base type.
@@
 
-\paragraph{Duplicability}
-There is one more bunched combinator we have experimented with adding to the
-framework:
+\subsection{Duplicability and L/nL}\label{sec:dup-lnl}
+In \cref{sec:rec}, I introduced a bunched connective $\Box^{0{+}}$, used to
+ensure that such a premise is derived in a context that allows it to be
+duplicated and discarded.
+Syntax for this connective can be added to \AgdaDatatype{Premises} and worked
+through the various parts of the framework, though I do not show these parts in
+the text of this thesis for the sake of simplicity and brevity.
+Additionally, variations such as $\Box^0$ (requiring only that the context is
+${} \leq \gr0$) and $\Box^{0{+}{*}}$ (requiring, in addition to
+$\Box^{0{+}}$, closure under multiplication by any scalar) can be added to the
+framework independently of each other and of $\Box^{0{+}}$.
 
-\[
-  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
-  \times \plr{\grRprime \leq \gr0}
-  \times \plr{\grRprime \leq \grRprime + \grRprime}
-  \times T\,\grRprime
-\]
+As well as the use case of inductive types, as shown in \cref{sec:rec}, I use
+the $\Box^{0{+}{*}}$-modality to encode the restriction to intuitionistic
+variables in the intuitionistic judgement forms of the L/nL
+calculus~\citep{Benton94}.
+I will explain this in detail in due course.
 
-The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
-of it, can be both discarded and duplicated indefinitely, and in the
-refinement we have a $T$.
-We use this combinator to introduce subterms that are used an unknown number of
-times, for example the continuations of the eliminator of an inductive type,
-or other fixed points.
-We can also use it in linear/non-linear style systems~\cite{Benton94} to make
-sure linear variables are not available in the intuitionistic fragment.
+\todo{Maybe partially merge with the later L/nL section.}
+Linear/non-Linear Logic has two sorts of types: linear types $A, B, C$ and
+intuitionistic types $X, Y, Z$.
+I encode these two sorts in the Agda types
+\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{lin} and
+\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{int}, respectively.
+Using an indexed type lets me take the total space of \AgdaDatatype{Ty}, named
+\AgdaFunction{$\Sigma$Ty}, as the evident dependent pair type.
+\AgdaFunction{$\Sigma$Ty} becomes the type of types seen by the framework.
+The types themselves are, on the linear side,
+a tensor unit \AgdaInductiveConstructor{tI},
+a tensor product \AgdaInductiveConstructor{\_t$\otimes$\_}, and
+a linear function space \AgdaInductiveConstructor{\_t$\multimap$\_};
+on the intuitionistic side,
+a unit \AgdaInductiveConstructor{t1},
+a product \AgdaInductiveConstructor{\_t$\times$\_}, and
+an intuitionistic function space \AgdaInductiveConstructor{\_t$\to$\_};
+and adjoint type formers \AgdaInductiveConstructor{tF} and
+\AgdaInductiveConstructor{tG}.
+I also throw in a linear base type \AgdaInductiveConstructor{$\iota$}.
 
-Adding the $\Box$ combinator is the only thing we have found that requires our
-linear maps be functional rather than merely relational.
+\ExecuteMetaData[\LnLtex]{Frag}
+\ExecuteMetaData[\LnLtex]{Ty}
+%\ExecuteMetaData[\LnLtex]{SigTy}
+
+As in the other examples, I define a data type of rule labels
+\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
+The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
+annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
+systems.
+
+We want to be able to embed the L/nL judgement forms
+$\Theta; \Gamma \vdash_{\mathcal L} A$ and $\Theta \vdash_{\mathcal C} X$, where
+$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
+variables.
+Following allowed usage patterns, intuitionistic variables will all have usage
+annotation $\gr\omega$, while linear variables will each be annotated
+either $\gr1$ or $\gr0$.
+These assignments of usage annotations are enforced wherever the rules bind new
+variables.
+By usage subsumption, intuitionistic variables may also end up with annotation
+$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
+$\gr\omega$ annotation, not causing a soundness problem.
+The key invariant is that linear variables never get annotation $\gr\omega$.
+
+The framework already tells us what variables and contexts we can
+have --- namely that we can make any assumption in any context.
+This means that we can propose ill formed sequents like
+$\gr1A \vdash X$.
+To make sure that this is not a problem, we want to maintain the invariant that
+any \emph{derivation} of a well formed sequent contains only well formed
+sequents.
+I enforce this locally using the $\Box^{0{+}{*}}$ bunched modality:
+Every rule targeting an intuitionistic judgement has its premises wrapped in a
+$\Box^{0{+}{*}}$ to restrict the usage annotations to $\gr0$ and $\gr\omega$.
+Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
+judgements but have intuitionistic premises, similarly get a $\Box^{0{+}{*}}$.
+Note that having the annotation $\gr0$ means that well formed intuitionistic
+sequents can contain linear variables annotated $\gr0$, but these variables are
+not usable.
+
+The rules are presented in Agda code below, and also using bunched inference
+rule notation in \cref{fig:LnL-bunched}
+
+\ExecuteMetaData[\LnLtex]{quoteLnL}
+\ExecuteMetaData[\LnLtex]{LnL}
+
+\begin{figure}
+  \begin{mathpar}
+    \ebrule[comb]{%
+      \hypo{I^{\sep}}
+      \infer1[$I$i]{\vdashL I}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdashL I}
+      \hypo{\sep}
+      \hypo{\vdashL C}
+      \infer3[$I$e]{\vdashL C}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdashL A}
+      \hypo{\sep}
+      \hypo{\vdashL B}
+      \infer3[$\otimes$i]{\vdashL A \otimes B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdashL A \otimes B}
+      \hypo{\sep}
+      \hypo{\gr1A, \gr1B \vdashL C}
+      \infer3[$\otimes$e]{\vdashL C}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\gr1A \vdashL B}
+      \infer1[$\multimap$i]{\vdashL A \multimap B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdashL A \multimap B}
+      \hypo{\sep}
+      \hypo{\vdashL A}
+      \infer3[$\multimap$e]{\vdashL B}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Boxzpt\plr{\vdashC X}}
+      \infer1[$F$i]{\vdashL FX}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\vdashL FX}
+      \hypo{\sep}
+      \hypo{\gr\omega X \vdashL C}
+      \infer3[$F$e]{\vdashL C}
+    }
+    \and
+    \ebrule[comb]{%
+      \hypo{\Boxzpt\dot1}
+      \infer1[$1$i]{\vdashC 1}
+    }
+    \and
+    \text{(no \TirName{$1$e})}
+    \and
+    \ebrule[comb]{%
+      \hypo{\Boxzpt(\vdashC X}
+      \hypo{\dottimes}
+      \hypo{\vdashC Y)}
+      \infer3[$\times$i]{\vdashC X \times Y}
+    }
+  \end{mathpar}
+  \caption{Linear/non-Linear Logic in bunched notation}\label{fig:LnL-bunched}
+\end{figure}
+
+Later,\todo{When?} I will give translations between this encoding of L/nL and
+$\name$, providing some assurance that the correct logic has been encoded.
+
+%There is one more bunched combinator we have experimented with adding to the
+%framework:
+%
+%\[
+%  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
+%  \times \plr{\grRprime \leq \gr0}
+%  \times \plr{\grRprime \leq \grRprime + \grRprime}
+%  \times T\,\grRprime
+%\]
+%
+%The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
+%of it, can be both discarded and duplicated indefinitely, and in the
+%refinement we have a $T$.
+%We use this combinator to introduce subterms that are used an unknown number of
+%times, for example the continuations of the eliminator of an inductive type,
+%or other fixed points.
+%We can also use it in linear/non-linear style systems~\cite{Benton94} to make
+%sure linear variables are not available in the intuitionistic fragment.
+%
+%Adding the $\Box$ combinator is the only thing we have found that requires our
+%linear maps be functional rather than merely relational.
+
+\section{A layer of syntax is functorial}\label{sec:functorial}
+
+A basic property of the universe of syntaxes
+is that every syntax supports a functorial action on subterms, realised by a
+function \AgdaFunction{map-s}.
+Its type says that to map a function \AgdaBound{f}
+over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
+domain and codomain usage contexts, and \AgdaBound{f} should be usable
+wherever the domain and codomain usage contexts are similarly related by
+\AgdaBound{$\Psi$}.
+The notation
+\AgdaBound{$\Psi$}\AgdaSpace{}\AgdaSymbol{.}\AgdaField{rel}\AgdaSpace{}%
+\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}
+means the same as $\grQ \leq \grP\gr\Psi$ from \cref{sec:ren-sub-lr}.
+The new notation comes about because I state \AgdaBound{$\Psi$} as a total and
+deterministic linear relation, rather than a linear map, for reasons related to
+mechanisation beyond the scope of this thesis.
+
+\ExecuteMetaData[\Maptex]{map-s-type}
+
+This generality is needed because usage contexts change between
+a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
+\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
+\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
+variables newly bound in that subterm).
+Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
+This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.
+
+The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
+works through the structure \AgdaBound{ps} of premises of the rule applied,
+acting on each subterm it finds.
+Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
+\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
+function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
+--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
+and \AgdaBound{Q$'$}.
+In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
+$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
+collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
+Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
+$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
+$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
+The cases for \AgdaInductiveConstructor{`$\cdot$} and
+\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
+of linearity.
+In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
+\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
+systems, do not involve any changes in the usage contexts.
+
+The linearity of relation \AgdaBound{$\Psi$} is given by fields
+\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
+the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
+\todo{Refer to preliminaries}
+
+\ExecuteMetaData[\Maptex]{map-p}
+
+I have also extended \AgdaFunction{map-p} to handle the various
+$\Box$-modalities described in \cref{sec:dup-lnl}.
+The Agda code for this extension is not particularly readable, so I do not
+include it in this document.
+However, this extension is notable as the only part of the framework requiring
+that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
+deterministic).

### 2023-02-01 agda/LRKits.lagda.tex
[ writing ] more

@@
     env-+ (ρ , sp) =
-      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (sp , ρ .fit-here) in
+      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (ρ .fit-here , sp) in
       record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereₗ }
@@
     env-* (ρ , sp) =
-      let fit-here′ , sp′ = ρ .Ψ .rel-*ₘ (sp , ρ .fit-here) in
+      let fit-here′ , sp′ = ρ .Ψ .rel-*ₘ (ρ .fit-here , sp) in
       sp′ , record { [_]_⇒ᵉ_ ρ; fit-here = fit-here′ }
@@
       (≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br))
-      (ρ .lookup (ρ .Ψ .rel-≤ₘ ≤*-refl (+ₘ-identityʳ→ (sp+l , dl)) ul)
+      (ρ .lookup (ρ .Ψ .rel-≤ₘ (+ₘ-identityʳ→ (sp+l , dl)) ≤*-refl ul)
         (lvar i q bl))
@@
       (≤*-trans (+ₘ-identityʳ→ (sp+l , dl))
-        (0*→≤* (ρ .Ψ .rel-0ₘ (≤*→0* bl , ul)))
+        (0*→≤* (ρ .Ψ .rel-0ₘ (ul , ≤*→0* bl)))
       ++ₙ

### 2023-02-01 tex/conc.tex
[ more ] thesis writing
[ writing ] renew generic syntax chapter
[ writing ] more

--- 2021-10-14 tex/esop22/conc.tex
@@
+\chapter{Conclusions}\label{sec:conc}
+
+FIXME: this chapter is directly from the ESOP22 paper.
+
 We have presented a framework for doing metatheory for a class of substructural
@@
 Our account of substructurality is based on the linear algebraic
-principles described by \citet{WA20}.
+principles described by \citet{WA21}.
 However, these details only really affect the definition of environment,
@@
 finer substructural systems.
+
+Is this a reusable library?
+
+Fitch-style systems (can't do).
+
+\section{Addendum: (lack of) partiality}\label{sec:part}
+\input{part.tex}

### 2023-02-01 tex/petricek.tex
[ writing ] more

--- 2022-11-18 tex/semirings.tex
@@
-\chapter{Usage restriction via semirings}\label{sec:semirings}
-
-The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
-make crucial use of \emph{weakening} --- the fact that if we have
-$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
-We use this property to update environments as they go under binders.
-However, as we saw in \cref{sec:linearity}, there are interesting calculi in
-which general weakening does not hold.
-As such, one of the aims of this chapter will be to find a form of weakening
-applicable to variables of any type, while essentially retaining linearity
-(as opposed to affineness).
-
-As explained by \citet{McBride16}, the first insight is to, instead of
-removing variables from the context of certain subterms, add an annotation to
-free variables saying whether or not they are to be used.
-I use an annotation $\gr0$ on variables that are not to be used, and an
-annotation $\gr1$ on variables that are to be used.
-This convention lets us transcribe the usual $\otimes$-introduction rule
-(below left) as a rule with usage annotations (below right).
-In the notation on the right, I let $\Gamma = \grP\gamma$ and
-$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
-\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
-A usage context is a list of usage annotations, so
-$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
-$\gamma = A_1, \ldots, A_m$.
-When combined, the usage context and the typing context will be of the same
-length.
-Explicit contexts will usually be written with usage annotations and types
-interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
-I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.
+A similar substitution lemma to the one presented in this chapter appears in the
+PhD thesis of \citet[p.\ 138]{petricek-thesis} under the name
+\emph{multi-nary substitution}.
+In my notation, \citeauthor{petricek-thesis}'s substitution rule looks like the
+following, up to permutation of the contexts containing $\Gamma$.
+Note that if $\Delta = \grQ\delta$, then $\gr r\Delta$ denotes the context
+$\plr{\gr r\grQ}\delta$.
+This rule is essentially an iterated version of the standard linear single
+substitution principle, and is used by \citeauthor{petricek-thesis} as a
+strengthened induction hypothesis required to derive single substitution.
 
@@
   \ebrule{%
-    \hypo{\Gamma \vdash A}
-    \hypo{\Delta \vdash B}
-    \infer2{\Gamma, \Delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
+    \hypo{\Delta_1 \vdash A_1}
+    \hypo{\cdots}
+    \hypo{\Delta_n \vdash A_n}
+    \hypo{\Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n \vdash B}
+    \infer4{\Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n \vdash B}
   }
@@
 
-The use of $\gr0$ gives us the property that variables never go out of
-scope in subterms; rather, we lose the ability to use certain variables, but
-retain the ability to refer to them metatheoretically.
-Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
-$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
-variables from $\delta$.
-I prove the admissibility of weakening for terms will come in \cref{sec:lrsub}.
+We can derive \citeauthor{petricek-thesis}-style multi-nary substitution as a
+corollary of my simultaneous substitution, using reasoning similar to that of
+\cref{thm:single-sub}.
 
-If we follow the DILL style of variable management, there are not just the two
-states \emph{used} ($\gr1$) and \emph{not used} ($\gr0$), but also
-\emph{usable unrestrictedly}.
-If we assign unrestricted (or \emph{intuitionistic}) variables an annotation
-$\gr\omega$, we can make the following transcription of the DILL
-$\otimes$-introduction rule.
+\begin{corollary}\label{thm:petricek-sub}
+  \Citeauthor{petricek-thesis}'s multi-nary substitution, as stated above, is
+  admissible in $\name$.
+\end{corollary}
+\begin{proof}
+  It is enough to provide a substitution of type
+  \[
+    \Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n
+    \env\vdash \Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n.
+  \]
+  To do this, we use \cref{thm:construct-env} repeatedly, leaving us needing a
+  substitution of type
+  $\Gamma, \gr0\Delta_1, \ldots, \gr0\Delta_n \env\vdash \Gamma$ and terms of
+  types
+  \begin{align*}
+    \gr0\gamma, \Delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \gr0\delta_n
+    \vdash A_1 \\
+    &\vdots \\
+    \gr0\gamma, \gr0\delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \Delta_n
+    \vdash A_n.
+  \end{align*}
+  The identity substitution and weakening by $\gr0$-annotated variables is
+  enough to make these requirements line up with the given hypotheses.
+\end{proof}
 
+My substitution principle is stronger than \citeauthor{petricek-thesis}'s.
+Where \citeauthor{petricek-thesis} requires that distinct variables be
+available for each hypothesis, I allow for separation of uses via addition of
+contexts.
+Below is a prototypical example.
+
+\begin{example}
+  Let $\Ann \coloneqq \plr{\mathbb N, =, 0, +, 1, \times}$, the exact
+  usage-counting posemiring.
+  Then, we can construct a substitution $\rho : \gr2A \env\vdash \gr1A, \gr1A$,
+  yielding a transformation of terms of the following form:
   \[
     \ebrule{%
-    \hypo{\Theta; \Gamma \vdash A}
-    \hypo{\Theta; \Delta \vdash B}
-    \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  }
+      \hypo{\gr1A, \gr1A \vdash B}
+      \infer1{\gr2A \vdash B}
+    }.
   \]
-
-To conceptualise the criteria on the usage annotations involved in this rule,
-I introduce an additive structure over usage annotations.
-The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
-$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
-Addition lifts pointwise to vectors of usage annotations (the green capital
-calligraphic $\grP$, $\grQ$, and $\grR$).
-It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
-rule on the right below now accepts $\gr0$-annotated variables in its
-conclusion, which is essential for weakening to be admissible.
-
-\[
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR = \grP + \grQ}
-    \hypo{\grP\gamma \vdash A}
-    \hypo{\grQ\gamma \vdash B}
-    \infer3{\grR\gamma \vdash A \otimes B}
-  }
-\]
-
-Some other transcriptions are as follows.
-I unify the variable rules (the one for linear variables and the one for
-intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
-annotations.
-We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
-demand of a linear variable by dereliction.
-We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
-weakened away like $\gr0$-annotated variables.
-All together, this means that at the unified variable rule, the variable being
-used must have annotation less than or equal to $\gr1$, and every other variable
-must have annotation less than or equal to $\gr0$.
-I write this requirement as $\grR \leq \langle x \rvert$, where
-$\langle x \rvert$ is the \emph{basis vector} at position $x$.
-
-\[
-  \ebrule{%
-    \infer0{\Theta; A \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \bra x}
-    \hypo{\gamma_x = A}
-    \infer2{\grR\gamma \vdash A}
-  }
-\]
-
-\[
-  \ebrule{%
-    \infer0{\Theta, A; {\cdot} \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \bra x}
-    \hypo{\gamma_x = A}
-    \infer2{\grR\gamma \vdash A}
-  }
-\]
-
-The final interesting rule form to cover is found in DILL's
-$\oc$-introduction rule.
-DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
-$\otimes$-introduction, though with the same premise each time rather than
-$\gr\omega$-many premises.
-This explanation recovers the behaviour that only $\gr\omega$- and
-$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
-and also justifies the designation of multiplication (vector scaling) as the
-algebraic operation controlling the $\oc$-modality.
-
-\[
-  \ebrule{%
-    \hypo{\Theta; {\cdot} \vdash A}
-    \infer1{\Theta; {\cdot} \vdash \oc A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr0\delta \vdash A}
-    \infer1{\gr\omega\theta, \gr0\delta \vdash \oc A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \gr\omega\grP}
-    \hypo{\grP\gamma \vdash A}
-    \infer2{\grR\gamma \vdash \oc_{\gr\omega} A}
-  }
-\]
-
-In summary, the structure we have required of the set of usage annotations is
-that they have addition (for $\otimes$-introduction and similar rules),
-multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
-$0$ (for a variable not being used), and an ordering (allowing for subsumption
-of usage restrictions).
-Together, these form a \emph{partially ordered semiring} (posemiring), the laws
-of which are both supported by examples and necessary for the syntax to be well
-behaved.
-\todo{Refer back to preliminaries for definition of posemiring}
-For concreteness, I collect together the definition of the
-$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.
-
-\begin{example}\label{def:lin-semiring}
-  The \emph{$\{\gr0, \gr1, \gr\omega\}$ semiring}, also known as the
-  \emph{linearity semiring}, has the operations given as follows, with
-  $0 \coloneqq \gr0$ and $1 \coloneqq \gr1$:
-
-  \makebox[\textwidth][s]{
-    \begin{tabular}{c|ccc}
-      $+$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
-      $\gr0$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
-      $\gr1$ & $\gr1$ & $\gr\omega$ & $\gr\omega$ \\
-      $\gr\omega$ & $\gr\omega$ & $\gr\omega$ & $\gr\omega$ \\
-    \end{tabular}
-    \begin{tabular}{c|ccc}
-      $*$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
-      $\gr0$ & $\gr0$ & $\gr0$ & $\gr0$ \\
-      $\gr1$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
-      $\gr\omega$ & $\gr0$ & $\gr\omega$ & $\gr\omega$ \\
-    \end{tabular}
-    \begin{tikzpicture}[baseline]
-      \node(omega) at (0,0) {$\gr\omega$};
-      \node(0) [above left of=omega] {$\gr0$};
-      \node(1) [above right of=omega] {$\gr1$};
-
-      \draw (omega) -- (0);
-      \draw (omega) -- (1);
-    \end{tikzpicture}
-  }
+  To construct $\rho$, we use \cref{thm:construct-env} case
+  $\sep(\rightarrowtriangle)$, using the fact that $\gr2 \leq \gr1 + \gr1$.
+  From there, two identity substitutions suffice.
+  The action of $\rho$ on terms is to merge the two variables into one.
+  Note that a renaming, rather than a substitution, would also suffice.
 \end{example}
 
-%In \cref{sec:simple}, we saw that the logical rules of simply typed
-%$\lambda$-calculus can be described in terms of three basic premise combinators:
-%$\dot1$, standing for no premises; $\dottimes$, allowing for multiple premises
-%in the same context; and $\Theta \vdash A$, requiring a subterm of type $A$
-%having bound the extra variables from context $\Theta$.
-%However, we remember from \cref{sec:linearity} that in a substructural setting,
-%we do not always want to copy assumptions for use in all subterms.
-%This motivates me to introduce the additional premise combinators $I^*$, $\sep$,
-%and $\cdot$ in \cref{sec:lnd}, allowing for the modes of usage exhibited in
-%the introduction rules for $I$, $\otimes$, and $\oc$, respectively.
-
-The rest of this chapter proceeds as follows.
-In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
-appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
-simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
-The idea of \name{} is to augment simply typed $\lambda$-calculus with
-annotations on free variables, and give enough types to manipulate these
-annotations.
-I use \cref{sec:lnd} to introduce some notation that allows us to restate the
-typing rules of \name{} so as to not mention contexts explicitly.
-The first problem I tackle with this new syntax is proving admissibility of
-substitution.
-I follow the methodology from \cref{sec:kits}, which
-
-\section{A usage-annotated calculus $\name$}\label{sec:lr}
-\input{lr.tex}
-\section{Bunched connectives}\label{sec:lnd}
-\input{lnd.tex}
-\section{What are linear renaming and substitution?}\label{sec:lrkits}
-\input{lrkits.tex}
-\section{Properties of linear environments}\label{sec:lenv}
-\input{lenv.tex}
-\section{Substitution is admissible in \name{}}\label{sec:lrsub}
-\input{lrsub.tex}
-\section{Adding recursion to \name{}}\label{sec:rec}
-\input{rec.tex}
-\section{Addendum: (lack of) partiality}\label{sec:part}
-\input{part.tex}
+Most notably, my (single) substitution principle more naturally fits the
+requirement we would have for the reduct of the $\beta$-rule for functions in
+$\name$, whereas \citeauthor{petricek-thesis}'s substitution principle would
+need some additional transformation for it to fit properly.
+This comes from the fact that the $\name$ function application rule introduces
+an algebraic ($+$) separation between its premises, whereas
+\citeauthor{petricek-thesis}'s substitution principle separates premises only
+via concatenation.

### 2023-02-01 tex/rec.tex
[ writing ] more

@@
 
-\subsection{Recursion}\label{sec:rec}
+\subsection{Adding inductive types and recursion}\label{sec:rec}
 
@@
 
-A similar modality, $\Box$, appears in the work of \citet{BB18} when they deal
-with \emph{persistent predicates} in bunched logic.
-They similarly aim to pick out those objects not depending on any external
+Similar modalities named $\Box$ appear in the work of \citet{BB18} when they
+deal with \emph{persistent predicates} in bunched logic, and in the work of
+\citet{CK20} in a capability-aware calculus.
+Both of these papers, in common with this thesis, aim to pick out the
+\emph{pure} or \emph{safe} objects, i.e.\ those not depending on any external
 resources.

### 2023-02-01 tex/rensub.tex
[ writing ] more

--- 2022-11-18 tex/semirings.tex
@@
-\chapter{Usage restriction via semirings}\label{sec:semirings}
+\chapter{Renaming and substitution for $\name$}\label{sec:ren-sub-lr}
 
-The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
-make crucial use of \emph{weakening} --- the fact that if we have
-$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
-We use this property to update environments as they go under binders.
-However, as we saw in \cref{sec:linearity}, there are interesting calculi in
-which general weakening does not hold.
-As such, one of the aims of this chapter will be to find a form of weakening
-applicable to variables of any type, while essentially retaining linearity
-(as opposed to affineness).
+In \cref{sec:semirings}, I defined my calculus of interest $\name$.
+In this chapter, I develop the necessary syntactic metatheory for
+specifying and implementing the substitution operation.
+I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
+significant changes to the underlying notion of \emph{environment} before doing
+so.
+I give and informally motivate these changes to environments in
+\cref{sec:lrkits}, and prove some properties of the new definition in
+\cref{sec:lenv}.
+Finally, I apply these new environments to the syntax of $\name$ in
+\cref{sec:lrsub} to derive renaming and substitution operators.
 
-As explained by \citet{McBride16}, the first insight is to, instead of
-removing variables from the context of certain subterms, add an annotation to
-free variables saying whether or not they are to be used.
-I use an annotation $\gr0$ on variables that are not to be used, and an
-annotation $\gr1$ on variables that are to be used.
-This convention lets us transcribe the usual $\otimes$-introduction rule
-(below left) as a rule with usage annotations (below right).
-In the notation on the right, I let $\Gamma = \grP\gamma$ and
-$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
-\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
-A usage context is a list of usage annotations, so
-$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
-$\gamma = A_1, \ldots, A_m$.
-When combined, the usage context and the typing context will be of the same
-length.
-Explicit contexts will usually be written with usage annotations and types
-interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
-I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.
-
-\[
-  \ebrule{%
-    \hypo{\Gamma \vdash A}
-    \hypo{\Delta \vdash B}
-    \infer2{\Gamma, \Delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
-  }
-\]
-
-The use of $\gr0$ gives us the property that variables never go out of
-scope in subterms; rather, we lose the ability to use certain variables, but
-retain the ability to refer to them metatheoretically.
-Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
-$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
-variables from $\delta$.
-I prove the admissibility of weakening for terms will come in \cref{sec:lrsub}.
-
-If we follow the DILL style of variable management, there are not just the two
-states \emph{used} ($\gr1$) and \emph{not used} ($\gr0$), but also
-\emph{usable unrestrictedly}.
-If we assign unrestricted (or \emph{intuitionistic}) variables an annotation
-$\gr\omega$, we can make the following transcription of the DILL
-$\otimes$-introduction rule.
-
-\[
-  \ebrule{%
-    \hypo{\Theta; \Gamma \vdash A}
-    \hypo{\Theta; \Delta \vdash B}
-    \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  }
-\]
-
-To conceptualise the criteria on the usage annotations involved in this rule,
-I introduce an additive structure over usage annotations.
-The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
-$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
-Addition lifts pointwise to vectors of usage annotations (the green capital
-calligraphic $\grP$, $\grQ$, and $\grR$).
-It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
-rule on the right below now accepts $\gr0$-annotated variables in its
-conclusion, which is essential for weakening to be admissible.
-
-\[
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
-    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
-    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR = \grP + \grQ}
-    \hypo{\grP\gamma \vdash A}
-    \hypo{\grQ\gamma \vdash B}
-    \infer3{\grR\gamma \vdash A \otimes B}
-  }
-\]
-
-Some other transcriptions are as follows.
-I unify the variable rules (the one for linear variables and the one for
-intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
-annotations.
-We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
-demand of a linear variable by dereliction.
-We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
-weakened away like $\gr0$-annotated variables.
-All together, this means that at the unified variable rule, the variable being
-used must have annotation less than or equal to $\gr1$, and every other variable
-must have annotation less than or equal to $\gr0$.
-I write this requirement as $\grR \leq \langle x \rvert$, where
-$\langle x \rvert$ is the \emph{basis vector} at position $x$.
-
-\[
-  \ebrule{%
-    \infer0{\Theta; A \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \bra x}
-    \hypo{\gamma_x = A}
-    \infer2{\grR\gamma \vdash A}
-  }
-\]
-
-\[
-  \ebrule{%
-    \infer0{\Theta, A; {\cdot} \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \bra x}
-    \hypo{\gamma_x = A}
-    \infer2{\grR\gamma \vdash A}
-  }
-\]
-
-The final interesting rule form to cover is found in DILL's
-$\oc$-introduction rule.
-DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
-$\otimes$-introduction, though with the same premise each time rather than
-$\gr\omega$-many premises.
-This explanation recovers the behaviour that only $\gr\omega$- and
-$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
-and also justifies the designation of multiplication (vector scaling) as the
-algebraic operation controlling the $\oc$-modality.
-
-\[
-  \ebrule{%
-    \hypo{\Theta; {\cdot} \vdash A}
-    \infer1{\Theta; {\cdot} \vdash \oc A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\gr\omega\theta, \gr0\delta \vdash A}
-    \infer1{\gr\omega\theta, \gr0\delta \vdash \oc A}
-  }
-  \quad\rightsquigarrow\quad
-  \ebrule{%
-    \hypo{\grR \leq \gr\omega\grP}
-    \hypo{\grP\gamma \vdash A}
-    \infer2{\grR\gamma \vdash \oc_{\gr\omega} A}
-  }
-\]
-
-In summary, the structure we have required of the set of usage annotations is
-that they have addition (for $\otimes$-introduction and similar rules),
-multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
-$0$ (for a variable not being used), and an ordering (allowing for subsumption
-of usage restrictions).
-Together, these form a \emph{partially ordered semiring} (posemiring), the laws
-of which are both supported by examples and necessary for the syntax to be well
-behaved.
-\todo{Refer back to preliminaries for definition of posemiring}
-For concreteness, I collect together the definition of the
-$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.
-
-\begin{example}\label{def:lin-semiring}
-  The \emph{$\{\gr0, \gr1, \gr\omega\}$ semiring}, also known as the
-  \emph{linearity semiring}, has the operations given as follows, with
-  $0 \coloneqq \gr0$ and $1 \coloneqq \gr1$:
-
-  \makebox[\textwidth][s]{
-    \begin{tabular}{c|ccc}
-      $+$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
-      $\gr0$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
-      $\gr1$ & $\gr1$ & $\gr\omega$ & $\gr\omega$ \\
-      $\gr\omega$ & $\gr\omega$ & $\gr\omega$ & $\gr\omega$ \\
-    \end{tabular}
-    \begin{tabular}{c|ccc}
-      $*$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
-      $\gr0$ & $\gr0$ & $\gr0$ & $\gr0$ \\
-      $\gr1$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
-      $\gr\omega$ & $\gr0$ & $\gr\omega$ & $\gr\omega$ \\
-    \end{tabular}
-    \begin{tikzpicture}[baseline]
-      \node(omega) at (0,0) {$\gr\omega$};
-      \node(0) [above left of=omega] {$\gr0$};
-      \node(1) [above right of=omega] {$\gr1$};
-
-      \draw (omega) -- (0);
-      \draw (omega) -- (1);
-    \end{tikzpicture}
-  }
-\end{example}
-
-%In \cref{sec:simple}, we saw that the logical rules of simply typed
-%$\lambda$-calculus can be described in terms of three basic premise combinators:
-%$\dot1$, standing for no premises; $\dottimes$, allowing for multiple premises
-%in the same context; and $\Theta \vdash A$, requiring a subterm of type $A$
-%having bound the extra variables from context $\Theta$.
-%However, we remember from \cref{sec:linearity} that in a substructural setting,
-%we do not always want to copy assumptions for use in all subterms.
-%This motivates me to introduce the additional premise combinators $I^*$, $\sep$,
-%and $\cdot$ in \cref{sec:lnd}, allowing for the modes of usage exhibited in
-%the introduction rules for $I$, $\otimes$, and $\oc$, respectively.
-
-The rest of this chapter proceeds as follows.
-In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
-appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
-simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
-The idea of \name{} is to augment simply typed $\lambda$-calculus with
-annotations on free variables, and give enough types to manipulate these
-annotations.
-I use \cref{sec:lnd} to introduce some notation that allows us to restate the
-typing rules of \name{} so as to not mention contexts explicitly.
-The first problem I tackle with this new syntax is proving admissibility of
-substitution.
-I follow the methodology from \cref{sec:kits}, which
-
-\section{A usage-annotated calculus $\name$}\label{sec:lr}
-\input{lr.tex}
-\section{Bunched connectives}\label{sec:lnd}
-\input{lnd.tex}
 \section{What are linear renaming and substitution?}\label{sec:lrkits}
@@
 \input{lrsub.tex}
-\section{Adding recursion to \name{}}\label{sec:rec}
-\input{rec.tex}
-\section{Addendum: (lack of) partiality}\label{sec:part}
-\input{part.tex}
+\section{Comparison with Petricek's substitution lemma}\label{sec:petricek}
+\input{petricek.tex}
+\section{Conclusion}\label{sec:ren-sub-lr-conc}
+
+In this and the preceding chapter, I have developed a discipline for specifying
+the syntax of linear and modal type systems, and furthermore developing the
+syntactic metatheory of those type systems.
+All of these are based on semirings, and the linear algebra arising from
+considering a usage context full of semiring elements as a vector.
+
+These developments can be seen in retrospect as a generalisation of the methods
+explained in \cref{sec:simple}.
+In terms of premise connectives in the syntactic rules, we have generalised from
+just $\{\dot1, \dottimes\}$ to
+$\{\dot1, \dottimes, I^*, \sep, \gr r\cdot{}, \Box^{0{+}}\}$, maintaining our
+ability to keep the context implicit.
+Similarly to how rule premises can require separation of usage annotations, our
+new environments can require such a separation between their entries thanks to
+the linear map they now contain.
+I have generalised the key property of a kit from arbitrary weakening to
+weakening by $\gr0$-annotated variables, and using that have produced a
+substitution operation based on the same principles as that from
+\cref{sec:kits}.
+
+Having generalised all of the components --- namely the contexts, the syntax,
+and the notion of environment --- the type of the substitution operation looks
+the same as it did for intuitionistic ST$\lambda$C\@.
+Being able to maintain this uniformity is a key step towards generalising the
+rest of \cref{sec:simple} (i.e., \cref{sec:gen-sem,sec:gen-syn}), as I do in
+\cref{sec:framework}.
+
+\begin{itemize}
+  \item Everything presented in this chapter is perfect.
+  \item But equivalence of substitutions is hard to define/work with.
+  \item The process of this development gives a template for defining
+    simultaneous substitution for other substructural calculi.
+\end{itemize}
+
+I have made few unforced design decisions in this chapter, so there are few
+aspects I can see future work improving upon.
+However, future work may want to extend the work of this chapter, in which case
+there are some unanswered questions.
+Principle among these, in my mind, is dealing with equivalence/equality of
+environments.
+Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
+if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
+$\rho\,x = \sigma\,x$.
+This follows from what we expect of equality of functions (function
+extensionality).
+Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
+dividing up the usage annotations of $\Gamma$, and then a function producing
+$\V$-values whose usage annotations come from that division.
+Equality of $\Sigma$-types is tricky --- we need to equate the first components,
+then rewrite the types of the second components by this equation before equating
+them.
+In practice, the equations rewriting other other equations build up so much that
+I have given up on a first effort to give a treatment of the equational theory
+of substitutions.
+Note that recursively defined environments (\cref{def:lr-rec-env}) are also
+$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
+definition does not clearly help.

### 2023-02-01 tex/semantics.tex
[ writing ] more

--- 2022-05-27 tex/semantics.tex
@@
-Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
-\AgdaFunction{semantics} we define in this section assigns a
-$\C$-value in context $\Gamma$ to every term in context $\Delta$,
-where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
-semantic interpretation of terms ($\V$ being the semantic
-interpretation of variables). Before we can define
-\AgdaFunction{semantics}, we need to treat recursion through rules'
-premises (\cref{sec:functorial}) and extension of environments when
-going under variable binders (\cref{sec:kripke}).
+Having fixed a universe of syntaxes, in which we can build terms, the next thing
+to do is to write recursive functions on terms.
+With terms being given by an inductive \AgdaSymbol{data} type definition, they
+already come with a recursion and an induction principle.
+However, these principles do not handle variable-binding, which we have seen
+with the fact that we had to write the \AgdaFunction{bind} helper
+function for renaming and substitution in \cref{sec:lrsub}.
+
+In this chapter, the central construct is a function \AgdaFunction{semantics}
+which, for a $\V$-environment $\rho : \Gamma \env\V \Delta$,
+maps a term $M : \Delta \vdash A$ to some semantic value in the type
+$\C\,\Gamma\,A$.
+This is a direct adaptation of the \AgdaFunction{semantics} function of
+\cref{sec:gen-sem}, which has the same kind of action on intuitionistic terms,
+given similar operations on $\V$ and $\C$ as what we had earlier.
+The \AgdaFunction{semantics} function recurses on the term $M$, updating $\rho$
+whenever new variables are bound.
+In our usage-aware case, $\rho$ is also updated whenever we come across linear
+combinations induced by premise combinators $I^{\sep}$, $\sep$, and
+$\gr r \cdot {}$.
+
+This chapter is structured as follows.
+I start by giving a quick introduction to linear relations --- a generalisation
+of linear maps --- with reference to their use in mechanised algebraic
+reasoning, in \cref{sec:lin-rel}.
+Using linear relations, I give a functorial \emph{map} operation to a single
+layer of syntax in \cref{sec:functorial}.
+I then adapt the \AgdaFunction{Kripke} function space to the usage-aware
+setting in \cref{sec:kripke}.
+Then I apply the \AgdaFunction{Kripke} function space, along with much of the
+machinery I have introduced in previous chapters, to give the
+\AgdaFunction{semantics} function in \cref{sec:traversal}.
+Finally, I give some example uses of \AgdaFunction{semantics} in
+\cref{sec:example-semantics}.
+\todo{What are the examples?}
 
@@
 
-\subsection{A layer of syntax is functorial}\label{sec:functorial}
+\section{Linear relations in Agda}\label{sec:lin-rel}
 
-A basic property of the universe of syntaxes we described in \cref{sec:syntax}
-is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
+In \cref{sec:lrkits}, I defined \emph{usage-annotated environments}
+(\cref{def:lr-env}).
+One component of a usage-annotated environment is a linear map $\gr\Psi$ which,
+when applied to the target usage vector, gives a vector compatible with the
+source usage vector.
+
+When it comes to mechanisation, I prefer to replace an assertion
+``$\grP \leq \grQ\gr\Psi$'', involving a linear map $\gr\Psi$, by an assertion
+``$\grP\gr\Psi\grQ$'', where $\gr\Psi$ is now a linear \emph{relation} said to
+relate $\grP$ and $\grQ$.
+I define linear relations as follows, where the reader may wish to check that
+a linear map gives rise to a linear relation via the expression
+$\grP \leq \grQ\gr\Psi$.
+
+\begin{definition}\label{def:linear-relation}
+  Given a posemiring $\Ann$ and modules $\mathscr M$ and $\mathscr N$ over
+  $\Ann$, a \emph{linear relation} between $\mathscr M$ and $\mathscr N$ is
+  a relation $\gr\Psi$ between the underlying sets of $\mathscr M$ and
+  $\mathscr N$ such that the following properties hold of all
+  $\grP, \grPprime, \grPl, \grPr \in \mathscr M$ and all
+  $\grQ, \grQprime, \grQl, \grQr \in \mathscr N$.
+  \begin{align*}
+    \grPprime \leq \grP \land \grP\gr\Psi\grQ \land \grQ \leq \grQprime
+    &\implies \grPprime\gr\Psi\grQprime
+    \\
+    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \gr0}
+    &\implies \grP \leq \gr0
+    \\
+    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \grQl + \grQr}
+    &\implies \plr{\exists\grPl,\grPr.~\grP \leq \grPl + \grPr
+      \land \grPl\gr\Psi\grQl \land \grPr\gr\Psi\grQr}
+    \\
+    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \gr r\grQprime}
+    &\implies \plr{\exists\grPprime.~\grP \leq \gr r\grPprime
+      \land \grPprime\gr\Psi\grQprime}
+  \end{align*}
+  I write $\mathscr M \rel \mathscr N$ as the type of linear relations between
+  $\mathscr M$ and $\mathscr N$.
+\end{definition}
+
+Relations have several advantages over functions when doing mechanised algebra
+in type theory.
+For one, what are compound expressions in functional style --- for example
+$x \leq f(y) + g(z)$ --- become collections of simple relationships in
+relational style --- for example $\exists v,w.~vfy \land wgz \land
+\mathrm{Add}\,x\,v\,w$.
+The advantage of this is that we have immediate access to all of the expressions
+and subexpressions, and the proofs of the relationships between them.
+This means that there is no need for congruence or monotonicity lemmas, and
+correspondingly no need to explicitly describe the syntactic context in which
+some algebraic manipulation is being applied and we rely less on the unifier.
+Another advantage is that one can design relations so that pattern-matching
+suggestively decomposes complex relationships.
+For example, given $F : \mathscr M \rel \mathscr M'$ and
+$G : \mathscr N \rel \mathscr N'$, we can define a relation
+$F \oplus G : \mathscr M \oplus \mathscr N \rel \mathscr M' \oplus \mathscr N'$
+pointwise, so that a proof of $(x, x')(F \oplus G)(y, y')$ is a proof of $xFy$
+together with a proof of $x'Gy'$.
+Pattern-matching on such a proof immediately gives us these constituent parts,
+whereas proofs of the corresponding statement involving functions would require
+using a lemma to get the parts.
+There is a dual advantage when producing one of these proofs, where we can
+introduce the canonical constructor (for pairs, in this example) rather than
+having to find the appropriate lemma.
+
+Relations also have several disadvantages, though I have found that for my use
+case, these are outweighed by the advantages.
+For example, automated algebraic solvers are better developed for function-based
+algebraic expressions, and sometimes the fact that functions satisfy unitality
+and associativity up to decidable judgemental equality means that some proofs
+can be avoided.
+The handling of compound expressions can also be a disadvantage in that it
+necessitates lots of new variable names and obscures goal and context displays.
+Finally, in predicative systems such as Agda, relations typically live in a
+larger universe than the corresponding functions.
+In practice, this means quantifying over an extra level variable for each
+relation involved in general lemmas.
+
+There are more relations than there are functions, so statements involving
+relations are more general than the corresponding statements involving
+functions.
+However, one part of the development requires functions rather than relations,
+so I impose functionality on relations after the fact.
+The appropriate notion of functional relation I use is slightly different to the
+standard one, in that I take account of the order on the codomain, and thus ask
+for the \emph{largest} solution rather than the \emph{unique} solution.
+
+\begin{definition}\label{def:functional-linear-relation}
+  A linear relation $\gr\Psi$ between $\mathscr M$ and $\mathscr N$ is
+  \emph{(right-to-left) functional} if, for every $\grQ \in \mathscr N$, there
+  exists universally a $\grP \in \mathscr M$ such that $\grP\gr\Psi\grQ$.
+  Universality means that, for all $\grPprime$ such that $\grPprime\gr\Psi\grQ$,
+  we have $\grPprime \leq \grP$ (i.e.\ $\grP$ is the largest solution).
+\end{definition}
+
+In Agda code, $\gr\Psi$ becomes \AgdaBound{$\Psi$} and the fact that $\gr\Psi$
+relates $\grP$ and $\grQ$ (in this section written $\grP\gr\Psi\grQ$) is
+rendered as
+\PsiDot{rel}\AgdaSpace{}\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}.
+That $\gr\Psi$ respects the orders on its arguments is given by
+\PsiDot{rel-$\leq_m$}, and the various linearity properties are given by
+\PsiDot{rel-0$_m$}, \PsiDot{rel-+$_m$}, and \PsiDot{rel-*$_m$}.
+
+\section{A layer of syntax is functorial}\label{sec:functorial}
+
+A basic property of the universe of syntaxes
+is that every syntax supports a functorial action on subterms, realised by a
+function \AgdaFunction{map-s}.
 Its type says that to map a function \AgdaBound{f}
-over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
+over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
 domain and codomain usage contexts, and \AgdaBound{f} should be usable
 wherever the domain and codomain usage contexts are similarly related by
-\AgdaBound{F}.
+\AgdaBound{$\Psi$}.
 
@@
 collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
-Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
+Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
 $\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
@@
 
+The linearity of relation \AgdaBound{$\Psi$} is given by fields
+\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
+the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
+\todo{Refer to preliminaries}
+
 \ExecuteMetaData[\Maptex]{map-p}
 
-\subsection{The Kripke function space}\label{sec:kripke}
+I have also extended \AgdaFunction{map-p} to handle the various
+$\Box$-modalities described in \cref{sec:dup-lnl}.
+The Agda code for this extension is not particularly readable, so I do not
+include it in this document.
+However, this extension is notable as the only part of the framework requiring
+that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
+deterministic).
 
-At this point we introduce a minor generalisation to
-\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam}:
+\section{The Kripke function space}\label{sec:kripke}
+
+At this point I introduce a minor generalisation to
+\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam} (as defined in
+\cref{sec:terms}):
 \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
@@
 replacing the textual occurrence of \AgdaBound{Ty} by the parameter
-\AgdaBound{I}.
+\AgdaBound{I}, though there is still reference to the ambient \AgdaBound{Ty}
+via \AgdaRecord{Ctx}.
+The main value I am interested in \AgdaBound{I} taking, other than
+\AgdaBound{Ty}, is \AgdaRecord{Ctx} --- for example, the type family of
+$\V$-environments, for a given $\V$, is a
+\AgdaRecord{Ctx}\AgdaSpace{}\AgdaFunction{---OpenFam}%
+\AgdaSpace{}\AgdaSymbol{\_}.
+However, I only use this generalisation much later.
+\todo{When?}
 
-The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
-of function space that describes a $\C$ value parametrised by
-$\Delta$-many additional $\V$s (all correctly typed and usage
-annotated). It is used to describe how to go under binders in a
-Higher-Order Abstract Syntax style---to go under a binder we must
-provide semantic interpretations for all the additional variables:
+\ExecuteMetaData[\Syntaxtex]{dashOpenFam}
+
+The definition
+\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}%
+\AgdaSpace{}\AgdaBound{$\Delta$} is a kind
+of function space that describes a \AgdaBound{$\C$}-value parametrised by
+\AgdaBound{$\Delta$}-many additional \AgdaBound{$\V$}-values (all correctly
+typed and usage-annotated).
+It is used to describe how to go under binders in a
+Higher-Order Abstract Syntax style: To go under a binder we must
+provide semantic interpretations for all the additional variables.
 
@@
 the presence of additional variables introduced by weakening.
+\todo{Is wand defined?}
 

### 2023-02-01 tex/semirings.tex
[ writing ] renew generic syntax chapter
[ writing ] more

--- 2022-12-12 tex/semirings.tex
@@
 \input{rec.tex}
-\section{Addendum: (lack of) partiality}\label{sec:part}
-\input{part.tex}
+\section{Translation to and from existing systems}\label{sec:translation}
+\input{trans.tex}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}\label{sec:semirings-conc}
 
-\chapter{Renaming and substitution of $\name$}\label{sec:ren-sub-lr}
+In this chapter, I have presented the calculus $\name$, with particular focus on
+the posemiring usage annotations.
+Posemirings are sufficient for many use cases, as shown by the examples in this
+chapter.
+However, there are many more substructural disciplines in the literature which
+cannot be expressed using posemirings.
 
-In \cref{sec:semirings}, I defined my calculus of interest $\name$.
-In this chapter, I develop its syntactic metatheory up to and including
-substitution.
-I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
-significant changes to the underlying notion of \emph{environment} before doing
-so.
-I give and informally motivate these changes to environments in
-\cref{sec:lrkits}, and prove some properties of the new definition in
-\cref{sec:lenv}.
-Finally, I apply these new environments to the syntax of $\name$ in
-\cref{sec:lrsub} to derive renaming and substitution operators.
+An important point to note is that $\name$ cannot be instantiated to become a
+bunched logic, in the sense of \citet{oHP99}.
+This is despite the use of the bunched connectives in the \emph{definition} of
+$\name$, and despite the formally tree-shaped contexts.
+However, the variable rule of $\name$ essentially treats the variables in the
+context independently, except for the individual checking of usage annotations
+on each variable.
+Therefore, we cannot talk about more interesting spatial relationships between
+variables, as required by bunched logic.
+This property also precludes us from capturing calculi without exchange
+(non-commutative logics), such as Lambek calculus~\citep{Lambek58}.
+Fitch-style systems\todo{cite} probably also are precluded similarly.
 
-\section{What are linear renaming and substitution?}\label{sec:lrkits}
-\input{lrkits.tex}
-\section{Properties of linear environments}\label{sec:lenv}
-\input{lenv.tex}
-\section{Substitution is admissible in \name{}}\label{sec:lrsub}
-\input{lrsub.tex}
+The work I present here has some methodological similarities with the earlier
+work of \citet{LicataSR17}.
+In that framework, one can provide a very precise \emph{mode theory}, expressing
+which structural rules are available, and from it get a sequent calculus with
+cut-elimination.
+They encode many systems, including a non-associative logic (where context are
+trees obeying no structural rules) and a bunched logic, and the framework is
+probably expressive enough to encode all posemiring-based usage disciplines.
+However, aside from the structural rules, the sequent calculus is fixed ---
+there are two connectives: $\mathsf F$ (internalising the left of the sequent,
+comparable to my
+$\oc_{\gr{r_1}}\plr{-} \otimes \cdots \otimes \oc_{\gr{r_n}}\plr{-}$) and
+$\mathsf U$ (internalising the whole sequent, comparable to my
+$\oc_{\gr{r_1}}\plr{-} \multimap \cdots \multimap \oc_{\gr{r_n}}\plr{-}
+\multimap \plr{-}$), which have left and right rules, and then the only other
+rule is the variable rule.
+This restriction is what allows the cut elimination theorem to be even
+plausible, but means that other connectives, like additives and inductive types,
+and more exotic syntaxes, like those I will present in
+\cref{sec:other-syntaxes}, would have to be developed from scratch.
+
+The distinction between sharing and separating conjunction of premises naturally
+falls out of the posemiring approach to usage restrictions.
+However, a similar distinction also appears in other approaches to linearity,
+and perhaps in other substructural systems.
+Indeed, in the paper that inspired the bunched connectives~\citep{RPKV20},
+linearity is enforced in a similar style as in Yalla~\citep{laurent18}, using
+lists which can be split.
+It would be interesting to see future work on substructural systems work out the
+appropriate premise connectives and define their systems from there, rather than
+directly manipulating contexts.
+It may also be possible to see future work profitably abstracting over the
+posemiring annotations, requiring only that contexts form something like a
+commutative $\mathrm{Rel}$-monoid supporting variable-binding and
+variable access.

### 2023-02-01 tex/syntax.tex
[ writing ] more

@@
 $\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
-terms, coterms, and commands.
+terms (traditional notation: $\Gamma \vdash A \mid \Delta$), coterms
+($\Gamma \mid A \vdash \Delta$), and commands ($\Gamma \vdash \Delta$).
 Read logically, terms and coterms are seen to, respectively, prove and refute
@@
 \emph{par}.
+I give a traditional presentation of this fragment of the
+$\mu\tilde\mu$-calculus in \cref{fig:mmt-orig}, with my encoding of the rules
+below.
 
 \ExecuteMetaData[\MuMuTildetex]{MMT}
+
+\begin{figure}
+  \begin{mathpar}
+    \ebrule{%
+      \hypo{\grP \leq \grPl + \grPr}
+      \hypo{\grQ \leq \grQl + \grQr}
+      \hypo{\grPl\gamma \vdash A \mid \grQl\delta}
+      \hypo{\grPr\gamma \mid A \vdash \grQr\delta}
+      \infer4[Cut$_A$]{\grP\gamma \vdash \grQ\delta}
+    } \and
+    \ebrule{%
+      \hypo{\Gamma \vdash \gr1A, \Delta}
+      \infer1[$\mu$]{\Gamma \vdash A \mid \Delta}
+    } \and
+    \ebrule{%
+      \hypo{\Gamma, \gr1A \vdash \Delta}
+      \infer1[$\tilde\mu$]{\Gamma \mid A \vdash \Delta}
+    } \and
+    \ebrule{%
+      \hypo{\Gamma \mid A \vdash \Delta}
+      \infer1[$\lambda$]{\Gamma \vdash A^\bot \mid \Delta}
+    } \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A \mid \Delta}
+      \infer1[$\tilde\lambda$]{\Gamma \mid A^\bot \vdash \Delta}
+    } \and
+    \ebrule{%
+      \hypo{\grP \leq \gr r\grPl + \gr s\grPr}
+      \hypo{\grQ \leq \gr r\grQl + \gr s\grQr}
+      \hypo{\grPl\gamma \vdash A \mid \grQl\delta}
+      \hypo{\grPr\gamma \vdash B \mid \grQr\delta}
+      \infer4[$\alr{-,-}$]%
+      {\grP\gamma \vdash \gr rA \parr \gr sB \mid \grQ\delta}
+    } \and
+    \ebrule{%
+      \hypo{\Gamma, \gr rA, \gr sB \vdash \Delta}
+      \infer1[$\mu\alr{-,-}$]{\Gamma \mid \gr rA \parr \gr sB \vdash \Delta}
+    }
+  \end{mathpar}
+  \caption{A fragment of a usage-annotated $\mu\tilde\mu$-calculus presented in
+    traditional sequent notation}
+  \label{fig:mmt-orig}
+\end{figure}
 
@@
 systems.
+For reference, I give a standard presentation of the rules of L/nL in
+\cref{fig:LnL-orig}
 
@@
 \begin{figure}
+  \begin{align*}
+    \mathcal J \Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
+  \end{align*}
+  \begin{mathpar}
+    \ebrule{%
+      \infer0[$\mathcal L$-var]{\Theta; A \vdashL A}
+    } \and
+    \ebrule{%
+      \infer0[$\mathcal C$-var]{\Theta, X \vdashC X}
+    } \and
+    \ebrule{%
+      \infer0[$I$i]{\Theta; {\cdot} \vdashL I}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma \vdashL I}
+      \hypo{\Theta; \Delta \vdashL C}
+      \infer2[$I$e]{\Theta; \Gamma, \Delta \vdashL C}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma \vdashL A}
+      \hypo{\Theta; \Delta \vdashL B}
+      \infer2[$\otimes$i]{\Theta; \Gamma, \Delta \vdashL A \otimes B}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma \vdashL A \otimes B}
+      \hypo{\Theta; \Delta, A, B \vdashL C}
+      \infer2[$\otimes$e]{\Theta; \Gamma, \Delta \vdashL C}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma, A \vdashL B}
+      \infer1[$\multimap$i]{\Theta; \Gamma \vdashL A \multimap B}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma \vdashL A \multimap B}
+      \hypo{\Theta; \Delta \vdashL A}
+      \infer2[$\multimap$e]{\Theta; \Gamma, \Delta \vdashL B}
+    } \and
+    \ebrule{%
+      \hypo{\Theta \vdashC X}
+      \infer1[$F$i]{\Theta; {\cdot} \vdashL FX}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; \Gamma \vdashL FX}
+      \hypo{\Theta, X; \Delta \vdashL C}
+      \infer2[$F$e]{\Theta; \Gamma, \Delta \vdashL C}
+    } \and
+    \ebrule{%
+      \infer0[$1$i]{\Theta \vdashC 1}
+    } \and
+    \text{(no \TirName{$1$e})} \and
+    \ebrule{%
+      \hypo{\Theta \vdashC X}
+      \hypo{\Theta \vdashC Y}
+      \infer2[$\times$i]{\Theta \vdashC X \times Y}
+    } \and
+    \ebrule{%
+      \hypo{\Theta \vdashC X_0 \times X_1}
+      \infer1[$\times$e$_i$]{\Theta \vdashC X_i}
+    } \and
+    \ebrule{%
+      \hypo{\Theta, X \vdashC Y}
+      \infer1[$\to$i]{\Theta \vdashC X \to Y}
+    } \and
+    \ebrule{%
+      \hypo{\Theta \vdashC X \to Y}
+      \hypo{\Theta \vdashC X}
+      \infer2[$\to$e]{\Theta \vdashC Y}
+    } \and
+    \ebrule{%
+      \hypo{\Theta; {\cdot} \vdashL A}
+      \infer1[$G$i]{\Theta \vdashC GA}
+    } \and
+    \ebrule{%
+      \hypo{\Theta \vdashC GA}
+      \infer1[$G$e]{\Theta; {\cdot} \vdashL A}
+    }
+  \end{mathpar}
+  \caption{Linear/non-Linear Logic in traditional sequent notation}
+  \label{fig:LnL-orig}
+\end{figure}
+
+\begin{figure}
   \begin{mathpar}
@@
       \infer1[$I$i]{\vdashL I}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer3[$I$e]{\vdashL C}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer3[$\otimes$i]{\vdashL A \otimes B}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer3[$\otimes$e]{\vdashL C}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer1[$\multimap$i]{\vdashL A \multimap B}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer3[$\multimap$e]{\vdashL B}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
-      \hypo{\Boxzpt\plr{\vdashC X}}
+      \hypo{\Box\plr{\vdashC X}}
       \infer1[$F$i]{\vdashL FX}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
@@
       \infer3[$F$e]{\vdashL C}
-    }
-    \and
+    } \and
     \ebrule[comb]{%
-      \hypo{\Boxzpt\dot1}
+      \hypo{\Box\dot1}
       \infer1[$1$i]{\vdashC 1}
-    }
-    \and
-    \text{(no \TirName{$1$e})}
-    \and
+    } \and
+    \text{(no \TirName{$1$e})} \and
     \ebrule[comb]{%
-      \hypo{\Boxzpt(\vdashC X}
+      \hypo{\Box(\vdashC X}
       \hypo{\dottimes}
@@
       \infer3[$\times$i]{\vdashC X \times Y}
+    } \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdashC X_0 \times X_1}}
+      \infer1[$\times$e$_i$]{\vdashC X_i}
+    } \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\gr\omega X \vdashC Y}}
+      \infer1[$\to$i]{\vdashC X \to Y}
+    } \and
+    \ebrule[comb]{%
+      \hypo{\Box(\vdashC X \to Y}
+      \hypo{\dottimes}
+      \hypo{\vdashC X)}
+      \infer3[$\to$e]{\vdashC Y}
+    } \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdashL A}}
+      \infer1[$G$i]{\vdashC GA}
+    } \and
+    \ebrule[comb]{%
+      \hypo{\Box\plr{\vdashC GA}}
+      \infer1[$G$e]{\vdashL A}
     }
   \end{mathpar}
-  \caption{Linear/non-Linear Logic in bunched notation}\label{fig:LnL-bunched}
+  \caption{Linear/non-Linear Logic in bunched notation using
+    $\{\gr0, \gr1, \gr\omega\}$ usage annotations and with $\Boxzpt$ abbreviated
+    to $\Box$}
+  \label{fig:LnL-bunched}
 \end{figure}
@@
 
-\section{A layer of syntax is functorial}\label{sec:functorial}
+\section{Conclusion}
 
-A basic property of the universe of syntaxes
-is that every syntax supports a functorial action on subterms, realised by a
-function \AgdaFunction{map-s}.
-Its type says that to map a function \AgdaBound{f}
-over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
-domain and codomain usage contexts, and \AgdaBound{f} should be usable
-wherever the domain and codomain usage contexts are similarly related by
-\AgdaBound{$\Psi$}.
-The notation
-\AgdaBound{$\Psi$}\AgdaSpace{}\AgdaSymbol{.}\AgdaField{rel}\AgdaSpace{}%
-\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}
-means the same as $\grQ \leq \grP\gr\Psi$ from \cref{sec:ren-sub-lr}.
-The new notation comes about because I state \AgdaBound{$\Psi$} as a total and
-deterministic linear relation, rather than a linear map, for reasons related to
-mechanisation beyond the scope of this thesis.
+I have shown via examples that the syntax descriptions defined in this chapter
+can capture a significant range of simply typed usage-aware calculi.
+Together with the examples of posemirings given in \cref{sec:semirings}, the
+whole framework is capable of encoding many specific substructural calculi.
+However, the language of descriptions has several restrictions, with more and
+less clear practical consequences.
+Many of these restrictions are shared with the syntax descriptions of
+\citet{AACMM21}, but some pertain to usage annotations.
 
-\ExecuteMetaData[\Maptex]{map-s-type}
+One clear limitation is that the forms of judgements do not allow enough
+interdependency to capture dependently typed calculi.
+In particular, for any given calculus, there is one fixed set of types, with no
+indexing over contexts.
+This allows contexts to have a flat list-like structure rather than a telescope,
+and means that the type of a term does not depend on that context, but clearly
+restricts the expressiveness of the calculi which can be encoded.
+As well as dependently typed calculi, this also prevents the encoding of the
+well formed types of System $F_\omega$.
 
-This generality is needed because usage contexts change between
-a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
-\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
-\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
-variables newly bound in that subterm).
-Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
-This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.
+Less clear than describability of System $F_\omega$ is describability of System
+$F$.
+Though I have not presented the details, I believe it should be possible to
+encode calculi with polymorphism, such as System $F$ and the usage-annotated
+system found in \citet{AbelBernardy2020}, via a two-stage process.
+First, we make a syntax description for the language of types, and then use
+the well typed terms arising from that description as the set of types when
+making the description for the System $F$ terms.
+Such an encoding would yield all of the benefits of the framework presented in
+this thesis separately to the language of types and the language of terms, but
+how these two languages interact is unclear.
+For example, in \cref{sec:semantics}, I provide a syntax-generic substitution
+operation.
+This operation would yield substitution for type variables in types and
+substitution for term variables in terms, but substitution for type variables in
+terms would have to be provided separately, and would likely be more
+complicated.
+It would be interesting to see future work in this direction, possibly inspired
+by how Autosubst~\citep{Autosubst15} handles the same issue.
 
-The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
-works through the structure \AgdaBound{ps} of premises of the rule applied,
-acting on each subterm it finds.
-Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
-\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
-function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
---- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
-and \AgdaBound{Q$'$}.
-In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
-$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
-collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
-Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
-$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
-$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
-The cases for \AgdaInductiveConstructor{`$\cdot$} and
-\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
-of linearity.
-In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
-\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
-systems, do not involve any changes in the usage contexts.
+As for restrictions peculiar to the usage-aware setting, we of course have the
+general limitations of posemiring annotations listed in
+\cref{sec:semirings-conc}, which I will not repeat here.
+Additionally, there are things we could wish to do with posemiring-based usage
+annotations which we do not have access to with the descriptions given in this
+chapter.
+For example, an issue one may have with the encoding of Linear/non-Linear Logic
+given in \cref{sec:dup-lnl} is that linear variables and intuitionistic
+variables, despite having morally distinct usage disciplines, necessarily share
+the $\{\gr0, \gr1, \gr\omega\}$ posemiring of usage annotations.
+If one wanted to generalise this construction to different pairs of usage
+disciplines, one may well find that they do not combine adequately into a single
+posemiring.
+A possible solution would be to have a range of kinds, and for each kind to have
+its own set of types and its own posemiring --- for example, a \emph{linear}
+kind and an \emph{intuitionistic} kind, using $\{\gr0, \gr1, \gr\omega\}$ and
+the 1-element posemiring, respectively.
+However, the descriptions given in this thesis only allow for one posemiring.
 
-The linearity of relation \AgdaBound{$\Psi$} is given by fields
-\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
-the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
-\todo{Refer to preliminaries}
-
-\ExecuteMetaData[\Maptex]{map-p}
-
-I have also extended \AgdaFunction{map-p} to handle the various
-$\Box$-modalities described in \cref{sec:dup-lnl}.
-The Agda code for this extension is not particularly readable, so I do not
-include it in this document.
-However, this extension is notable as the only part of the framework requiring
-that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
-deterministic).
+Finally, while there is a simple and well motivated core of premise combinators
+--- $\dot1$, $\dottimes$, $I^{\sep}$, $\sep$, and $\gr r \cdot \plr{-}$ --- the
+various $\Box$-modalities feel like ad hoc additions.
+While I have shown that $\Boxzp$ and $\Boxzpt$ have several practical uses, it
+is unclear whether there should be more modalities.
+Furthermore, I have not made an effort to isolate the properties that make these
+modalities well behaved.
+We will see in \cref{sec:functorial} that any such modality will need to be
+functorial in the appropriate sense, but other examples of generic programming,
+such as the usage elaborator which will be detailed in
+\cref{sec:usage-elaborator}, may need further properties.

### 2023-02-20 tex/gen-sem.tex
[ build ] get thesis into building state
[ writing ] lots of background, before feedback on later chapters
[ writing ] new draft of mechanisation chapter

@@
-\def\SimpleSem{../agda/latex/SimpleSem.tex}
+\def\SimpleSem{../agda/processed-latex/SimpleSem.tex}
 
@@
 One of the basic connectives in this setting is the \emph{pointwise} arrow
-\AgdaFunction{\_$\Rightarrow$\_}, which acts in
+\AgdaFunction{\_$\dotto$\_}, which acts in
 \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
@@
 the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
-\AgdaFunction{\_$\V\vdash$\_} and \AgdaFunction{\_$\C\vdash$\_}).
+\AgdaFunction{\_$\V\vDash$\_} and \AgdaFunction{\_$\C\vDash$\_}).
 The principal use of these aliases is to fill the right argument with a type
@@
 \ExecuteMetaData[\SimpleSem]{kit}
+
+With \AgdaRecord{Semantics} fixed, I also give the corresponding implementation
+of \AgdaFunction{sem}.
+Like \AgdaFunction{trav} from \cref{sec:syntactic-kits}, we need a
+\AgdaFunction{bindEnv} function, but I have updated this to deal with renamings
+and the \AgdaFunction{$\Box$}-operator, and also to fit better with concepts
+important for generic \emph{syntax} (as in \cref{sec:gen-syn}).
+\AgdaFunction{bindEnv} now takes the old environment \AgdaBound{$\rho$}%
+\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaFunction{[}\AgdaSpace{}%
+\AgdaBound{$\V$}\AgdaSpace{}\AgdaFunction{]}\AgdaSpace{}\AgdaBound{$\Gamma$}%
+\AgdaSpace{}\AgdaFunction{$\Rightarrow^e$}\AgdaSpace{}\AgdaBound{$\Delta$}, a
+renaming \AgdaBound{r}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaFunction{Ren}\AgdaSpace{}\AgdaBound{$\Gamma^{+}$}\AgdaSpace{}%
+\AgdaBound{$\Gamma$}, and an environment \AgdaBound{$\sigma$} in the renamed
+context \AgdaBound{$\Gamma^{+}$} targeting the context extension
+\AgdaBound{$\Delta$r}.
+In the \AgdaInductiveConstructor{lam} case of \AgdaFunction{sem}, the
+\AgdaField{$\llbracket$lam$\rrbracket$} field gives us exactly the renaming to
+pass to \AgdaFunction{bindEnv}, while the extension environment
+\AgdaBound{$\sigma$} is made as a singleton environment containing just the
+\AgdaBound{A}-value \AgdaBound{v}.
+
+\ExecuteMetaData[\SimpleSem]{sem}
+\ExecuteMetaData[\SimpleSem]{bindEnv}

### 2023-02-20 tex/gen-syn.tex
[ writing ] new draft of mechanisation chapter

@@
 In our generic version of \AgdaRecord{Semantics}, we keep the
-\AgdaField{ren\textasciicaron$\V$} and \AgdaField{$\llbracket$var$\rrbracket$}
+\AgdaField{ren\textasciicircum$\V$} and \AgdaField{$\llbracket$var$\rrbracket$}
 fields as before, and replace \AgdaField{$\llbracket$app$\rrbracket$} and
@@
 for uniformity and simplicity of presentation.
+
+To generate the expressions involving ellipses, I give an interpretation of the
+formal rule descriptions.
+The interpretation is parametrised on some
+\ExecuteMetaData[\SimpleSyntaxtex]{WithScope}, where, in
+\AgdaBound{,}\AgdaSpace{}\AgdaBound{$\Delta$}\AgdaSpace{}%
+\AgdaBound{$\llbracket$}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
+\AgdaBound{$\vdash$}\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}%
+\AgdaBound{$\rrbracket$}, the context \AgdaBound{$\Delta$} stands for the newly
+bound variables of a premise, \AgdaBound{$\Gamma$} is the context as it was
+below the rule's horizontal line, and \AgdaBound{A} is the type of the premise.
+In the \AgdaInductiveConstructor{con} constructor for terms, the parameter is
+$\Gamma, \Delta \vdash A$, and in the \AgdaField{$\llbracket$con$\rrbracket$}
+field for semantics, the parameter is
+$\Box\plr{{} \env\V \Delta \dotto {} \sdtstile{}\C A}\,\Gamma$.
+
+A single premise with newly bound variables is interpreted by shuffling the
+parts into the right place, while multiple premises are interpreted as pointwise
+products of the individual premises (giving the ellipses above).
+
+\ExecuteMetaData[\SimpleSyntaxtex]{semp}
+
+A rule, with all its parameters instantiated, targets a specific type
+\AgdaBound{A$'$}, which we check to match the desired type \AgdaBound{A}.
+Finally, a whole \AgdaRecord{System} comprises a set \AgdaBound{L} of rule
+labels, and \AgdaBound{rs}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaBound{L}%
+\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaRecord{Rule}.
+The interpretation of these data is to pick a rule label \AgdaBound{l}, and then
+take the interpretation of the rule \AgdaBound{rs}\AgdaSpace{}\AgdaBound{l}.
+For the sake of defining terms as a least fixed point, it is important to note
+that the interpretation of a syntax description is strictly positive in the
+parameter \AgdaBound{,\_$\llbracket$\_$\vdash$\_$\rrbracket$}.
+
+\ExecuteMetaData[\SimpleSyntaxtex]{semr}
+\ExecuteMetaData[\SimpleSyntaxtex]{sems}
+
+The interpretation of a system description as a single layer of syntax is
+functorial, supporting the \AgdaFunction{map-s} function when the parameter
+\AgdaBound{,\_$\llbracket$\_$\vdash$\_$\rrbracket$} is given as an extra
+argument named \AgdaBound{X} or \AgdaBound{Y} (which are both fixed as
+parameters of \AgdaFunction{map-s}, together with \AgdaBound{$\Gamma$} and
+\AgdaBound{$\Delta$}).
+
+\ExecuteMetaData[\SimpleSyntaxtex]{map-s-type}
+
+The implementation of \AgdaFunction{map-s} is straightforward, so I do not list
+it here.
+We preserve
+the shape of the syntactic layer, applying the function to each \AgdaBound{X}
+we find (wherever the description contains \AgdaInductiveConstructor{$\langle$}%
+\AgdaSpace{}\AgdaBound{$\Delta$}\AgdaSpace{}%
+\AgdaInductiveConstructor{`$\vdash$}\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}%
+\AgdaInductiveConstructor{$\rangle$}).
+
+This \AgdaFunction{map-s} will be used in the generic syntax version of
+\AgdaFunction{sem} to recursively apply \AgdaFunction{sem} to all subterms.
+However, a major distinction between generic syntax and the specific syntax of
+previous sections is that the subterms found by \AgdaFunction{map-s} are not
+recognised by Agda's termination checker as \emph{structurally smaller} than
+the original term.
+Therefore, a na\"{i}vely written \AgdaFunction{sem} will fail Agda's termination
+check.
+
+To make \AgdaFunction{sem} pass the termination check, we have three major
+options:
+\begin{enumerate}
+  \item Assert \AgdaFunction{sem} to be terminating, bypassing the termination
+    check.
+  \item Use Agda's \emph{sized types} to remember that the subterms are smaller.
+  \item Avoid sized types, and index terms over some user-defined type (for
+    example, natural numbers or ordinal notations) which is structurally smaller
+    at subterms.
+\end{enumerate}
+
+Each of these approaches has drawbacks.
+Approach 1 is clearly unsafe, in the sense that the fundamental lemma
+\AgdaFunction{sem} is not being completely checked for type-correctness.
+Approach 2 is also unsafe, because Agda's sized type implementation is known to
+make the system inconsistent~\citep{AgdaIssue1201}.
+Meanwhile, approach 3 is safe, but entails a lot of manually extracted and
+supplied extra arguments, which I think would distract from the presentation and
+make the resulting code harder to use.
+I choose to follow \citet{AACMM21} in using sized types, justified by the idea
+that Agda may eventually have a sound implementation of sized types, at which
+point I would want my code to be as easy to update for that new version of Agda
+as possible.
+
+Using sized types, my type family of \AgdaRecord{System}-generic terms is as
+below.
+\AgdaFunction{Scope} is a name for the transformation of an
+\AgdaFunction{OpenFam} into a \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaFunction{OpenFam} which appends the extra context to the
+existing context before applying the original \AgdaFunction{OpenFam}
+(in this case, producing something like $\Gamma, \Delta \vdash A$ from
+$\vdash$).
+The Agda builtin \AgdaPostulate{$\uparrow$} produces a bigger size from an
+existing size \AgdaBound{sz}, giving us here that the size of a term is 1 bigger
+than the size of all of its immediate subterms.
+Agda's elaborator and termination checker have special support for sizes, so we
+do not have to worry much about them from this point on.
+
+\ExecuteMetaData[\SimpleTermtex]{Tm}
+
+Corresponding to the generic \AgdaInductiveConstructor{`con} constructor for
+terms, we have a generic field \AgdaField{$\llbracket$con$\rrbracket$} in the
+updated \AgdaRecord{Semantics} record.
+In place of where one might expect
+\AgdaFunction{Scope}\AgdaSpace{}\AgdaBound{$\C$}, we instead have
+\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$},
+with \AgdaFunction{Kripke} defined below.
+The form of \AgdaFunction{Kripke} follows from the shape we saw in the type of
+the \AgdaField{$\llbracket$lam$\rrbracket$} field we saw in \cref{sec:gen-sem},
+where I use an environment targeting \AgdaBound{$\Delta$} as a way to say
+``a value for each type in \AgdaBound{$\Delta$}''.
+
+\ExecuteMetaData[\SimpleSemanticstex]{Kripke}
+\ExecuteMetaData[\SimpleSemanticstex]{Semantics}
+
+Finally, we get a generic semantic traversal as follows.
+The function \AgdaFunction{bindEnv} is unchanged from \cref{sec:gen-sem}, as it
+never mentions the syntax.
+The type of the traversal \AgdaFunction{sem} is also basically unchanged --- we
+just need to account for arbitrary term sizes (\AgdaBound{sz}), which will get
+smaller when recursing on subterms.
+I have chosen, as did \citet{AACMM21}, to define \AgdaFunction{sem} mutually
+with a function \AgdaFunction{body}, which is like a counterpart to
+\AgdaFunction{sem} dealing with newly bound variables.
+Note that the mutual recursion is not essential --- for example,
+\AgdaFunction{body} could simply be inlined.
+The \AgdaInductiveConstructor{`var} case of \AgdaFunction{sem} is as before.
+The \AgdaInductiveConstructor{`con} case, if viewed appropriately, is a direct
+generalisation of the \AgdaInductiveConstructor{lam} case from earlier.
+We recursively apply \AgdaFunction{sem} to all immediate subterms contained in
+\AgdaBound{M} (as found by \AgdaFunction{map-s}), with an environment updated to
+reflect the newly bound variables in each premise of the rule that was applied.
+
+\ExecuteMetaData[\SimpleSemanticstex]{sem}

### 2023-02-20 tex/mech-related.tex
[ writing ] new draft of mechanisation chapter

@@
 
-\subsection{Fiore, Plotkin, Turi}
+\subsection{Fiore et al.}
 
-TODO: cite \citet{FPT99} and \citet{FS22}.
+\todo{This section is currently a random selection of paragraphs.}
+Marcelo Fiore and various collaborators have a long line of work aiming for a
+categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
+A recent, particularly relevant paper from this line of work is \citet{FS22},
+which mechanises some of this work to obtain a framework similar in scope to
+the work of \citet{AACMM21}.
+
+Perhaps the main methodological difference between the works by
+\citet{AACMM21} and \citet{FS22} is that \citet{AACMM21} start from typed
+syntaxes and then give them semantics, whereas \citet{FS22} start by
+characterising the category of models of a theory, and then later showing that
+syntaxes form the initial model.
+The latter approach means that \citeauthor{FS22} can do a significant amount of
+theory not mentioning syntax in a compositional and structured style.
+In contrast, the former approach is more lightweight when one wants to write
+programs without a categorical specification.
+
+I chose to base the work of this thesis on that of \citet{AACMM21} for largely
+circumstantial reasons.
+I only became aware of the work of \citet{FS22} when it was published, by which
+time I had already completed the bulk of the work presented in this thesis.
+Also, one of the authors of the former paper is my PhD supervisor, and the other
+authors too were working nearby, so it was easy to discuss their work quickly
+and informally.
+
+In terms of the underlying theory, both works extend multi-sorted universal
+algebra\todo{cite} with variable-binding.
+However, the two extensions are subtly different.
+Universal algebra already has a notion of variable, which supports renaming and
+substitution, and which allows a given term to be evaluated when the free
+variables of that term are assigned semantic values.
+\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
+variables in terms.
+On the other hand, \citeauthor{FS22} recast the existing variables as
+\emph{metavariables}, and introduce a new notion of (bindable) variables
+separately into the syntax.
+The resulting metavariables can then stand for arbitrary \emph{open} terms, and
+thus each one remembers its context and requires a substitution whenever it is
+used.
+
+%Compared to the work of \citet{AACMM21}, which is what I chose to guide this
+%chapter and the rest of this thesis, \citet{FS22} can be understood as providing
+%many extra features.
+%I happened not to be interested in these specific features, so have not
+%implemented them, but future development of the work I present in this thesis
+%could benefit from such features.
+%The extra features include an externally compiled language of syntax and theory
+%descriptions, internal support for metavariables, and a treatment of the
+%category of models (of which terms are the initial model and what I call
+%\AgdaFunction{sem} forms the universal morphism from terms to any other model).
+
+\Citeauthor{FS22} use metavariables to form descriptions of relations over
+terms, like an equational theory over the simply typed $\lambda$-calculus.
+Terms have, as well as their context of variables, a context of metavariables,
+and terms can contain a metavariable wherever they could contain a subterm.
+There is then an operation of metavariable substitution, which substitutes terms
+in the place of metavariables.
+Metavariable substitution is used to instantiate the rules of described
+relations/theories.
+In contrast, \citeauthor{AACMM21} make do with the variables of the metalanguage
+(i.e.\ Agda variables).
+\todo{Congruence once and for all.}
+
+Technologically, \citeauthor{FS22} provide an external domain-specific language
+for syntax descriptions.
+From such a description, a Python program generates some boilerplate Agda code
+providing the types, the algebraic signature, the well typed terms, and a proof
+that the terms are the initial model.
+Using code generation like this resembles parts of the Autosubst plugin, as
+opposed to the purely internal, generic programming-based solution given by
+\citeauthor{AACMM21} and in this thesis.
+
+\Citet[p.\ 19]{FS22} avoid sized types by defining their equivalent of the
+functorial mapping \AgdaFunction{map-s} specialised to \AgdaFunction{sem}, with
+these two functions being mutually recursive.
+This is a standard solution, where code reuse is traded away in favour of
+satisfying the termination checker and avoiding sized types.

### 2023-02-20 tex/terms.tex
[ edit ] some corrections and notes after meeting
[ writing ] new draft of mechanisation chapter

@@
 The advantages of the functional vector representation will not become clear
-until later chapters\todo{forward reference}, where I make use of the ease of
-look-up and $\eta$-law of functions.
+until later chapters --- particularly the example in
+\cref{sec:usage-elaborator}, where I make use of the ease of look-up and
+$\eta$-law of functions.
 However, I claim for now that there is little to no disadvantage in the
@@
 \Term{}
+
+\todo{Example terms}

### 2023-05-02 tex/linear.tex
[ edit ] new draft of chapters 5 and 6
[ writing ] mostly C3 related work

--- 2021-12-03 tex/linear.tex
@@
-\chapter{Linearity}\label{sec:linearity}
+\chapter{Linearity and modality}\label{sec:linearity}
+
+\begin{itemize}
+  \item Motivation via prior work:
+    \begin{itemize}
+      \item Rust
+      \item Session types
+      \item ATS
+      \item Complexity \citep{Hofmann03}
+      \item Modal\ldots
+    \end{itemize}
+  \item Can \citet{AACMM21} capture linearity/modality?
+    It's unclear based on standard presentations.
+    Let's find one that looks good (DILL/PD), and see whether it works.
+  \item In-depth introduction of DILL and PD
+  \item It's close to working, but doesn't.
+  \item Mention \citet{BBdePH93}, problems with substitution.
+\end{itemize}
+
+The techniques of \cref{sec:simple} were developed to handle idealised core
+calculi, like the simply typed $\lambda$-calculus and its many variants.
+Such core calculi are fairly close to some real intermediate languages,
+\todo{cite} though many intermediate languages also have polymorphism in their
+type systems, which I have not discussed, but has been considered by much prior
+work\todo{cite}.
+
+%Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
+%Its distinguishing feature is that hypotheses, by default, must be used exactly
+%once.
+%I devote a chapter to introducing linear logic because it is better understood
+%than any individual or general semiring-annotated calculi, and is thus an
+%important point of comparison.
+%Linear logic will be an instance of the semiring-annotated systems we study
+%later.
 
@@
 
-\section{DILL}
+\section{DILL}\label{sec:dill}
 \input{dill.tex}

### 2023-05-02 tex/lrkits.tex
[ edit ] another draft of chapters 5 and 6
[ writing ] mostly C3 related work

@@
 
-\begin{example}
+\begin{example}\label{ex:rec-sub}
   Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$, with the equality order
   chosen to avoid any concerns around subsumption of annotations.
-  Then, there is an intuitionistic recursive environment as follows, where
-  $y\,z$ is the application of $y$ to $z$.
+  Then, there is an intuitionistic recursive environment (substitution) as
+  follows, where $y\,z$ is the application of $y$ to $z$.
   \[
     \alr{\alr{z},\alr{y\,z}} :
-    \plr{x : A, y : B \to C, z : B} \env\vdash_R \plr{B, C}.
+    \plr{x : A, y : B \to C, z : B} \env\vdash_R \plr{B, C}
   \]
@@
 
-From the example, we can see that the important usage vectors are the initial
-one $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix}$ and the usage vectors
-at which terms are derived: $\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$
+\begin{example}\label{ex:rec-ren}
+  Take $\Ann = \plr{\mathbb N, =, 0, +, 1, \times}$.
+  Then, there is an intuitionistic recursive environment (renaming) as follows,
+  where $\alr{-,-,-}$ abbreviates $\alr{-,\alr{-,-}}$ (matching a similar
+  abbreviation in the notation of contexts).
+  \[
+    \alr{\alr{c},\alr{a},\alr{a}} :
+    \plr{a : A, b : B, c : C, d : D} \env\sqni_R \plr{C, A, A}.
+  \]
+  There is also a usage-aware recursive environment
+  \[
+    \alr{\alr{c},\alr{a},\alr{a}} :
+    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D}
+    \env\sqni_R \plr{\gr1C, \gr2A, \gr4A}.
+  \]
+  Intuitively, this choice of usage annotations works because the $\gr6$ $A$s on
+  the left can be divided into the $\gr2 + \gr4$ $A$s on the right.
+  Similarly, the $\gr0$ $B$s and $\gr0$ $D$s on the left can be discarded to
+  yield none on the right.
+  Note that these divisions are directed, so we cannot merge variables or
+  introduce new variables on the right.
+\end{example}
+
+From \cref{ex:rec-sub}, we can see that the important usage vectors are the
+initial one $\begin{pmatrix} \gr0 & \gr2 & \gr3 \end{pmatrix}$ and the usage
+vectors at which terms are derived:
+$\begin{pmatrix} \gr0 & \gr0 & \gr1 \end{pmatrix}$
 and $\begin{pmatrix} \gr0 & \gr1 & \gr1 \end{pmatrix}$.
@@
   $\grQprime\delta \sqni A$ to $\V\,\grPprime\gamma\,A$.
-
+\end{definition}
+\begin{notation}
   When there are multiple environments in question and $\rho$ is such an
@@
   using this notation.
-\end{definition}
+\end{notation}
 

### 2023-05-02 tex/mech-related.tex
[ writing ] mostly C3 related work

@@
 
-\subsection{Fiore et al.}
+\subsection{Second order abstract syntax}\label{sec:fiore}
 
-\todo{This section is currently a random selection of paragraphs.}
 Marcelo Fiore and various collaborators have a long line of work aiming for a
 categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
-A recent, particularly relevant paper from this line of work is \citet{FS22},
-which mechanises some of this work to obtain a framework similar in scope to
-the work of \citet{AACMM21}.
+A recent, particularly relevant paper from this line of work is that of
+\citet{FS22}, which mechanises some of this work to obtain a framework similar
+in scope to the work of \citet{AACMM21}.
+However, there are several differences between the resulting mechanised
+frameworks, of methodological, mathematical, and technical nature.
 
-Perhaps the main methodological difference between the works by
-\citet{AACMM21} and \citet{FS22} is that \citet{AACMM21} start from typed
-syntaxes and then give them semantics, whereas \citet{FS22} start by
-characterising the category of models of a theory, and then later showing that
-syntaxes form the initial model.
-The latter approach means that \citeauthor{FS22} can do a significant amount of
-theory not mentioning syntax in a compositional and structured style.
-In contrast, the former approach is more lightweight when one wants to write
-programs without a categorical specification.
+Though \citet{AACMM21} did not state their results in categorical terms, it is
+still useful to infer what the category-theoretic statement would have been, and
+compare it to the statement actually given by \citet{FPT99} and \citet{FS22}.
+When we do this, we see that while \citeauthor{FPT99} deal with the category of
+contexts under renaming, and presheaves on that category.
+This means that every model must be shown to respect renaming before touching
+the framework of \citeauthor{FS22}.
+Meanwhile, \citeauthor{AACMM21} make use of the interplay between the discrete
+category of contexts and the category of contexts under renaming.
+For example, the models (\AgdaRecord{Semantics}) of \citet{AACMM21} require only
+that the family of semantic values $\V$ be shown to respect renaming, while the
+fact that the result family $\C$ respects renaming follows as a corollary of the
+traversal function \AgdaFunction{trav}.
+\Citeauthor{FS22} make no distinction between $\V$ and $\C$, essentially only
+having $\C$, but requiring it to respect renaming before getting the morphism
+to that model from the initial model (the syntax).
+In particular, this interplay allows us to derive renaming for terms in the way
+we saw in \cref{sec:gen-syn} --- by making $\V$ the family of variables, which
+trivially respects renaming.
+\todo{Every model is an environment model?}
 
-I chose to base the work of this thesis on that of \citet{AACMM21} for largely
-circumstantial reasons.
-I only became aware of the work of \citet{FS22} when it was published, by which
-time I had already completed the bulk of the work presented in this thesis.
-Also, one of the authors of the former paper is my PhD supervisor, and the other
-authors too were working nearby, so it was easy to discuss their work quickly
-and informally.
+Perhaps the relative complexity of the categorical account of the work of
+\citet{AACMM21} is why the authors decided not to state it in such terms.
+However, it is also likely that \citeauthor{AACMM21} developed their work in
+quite a different style to how \citeauthor{FPT99} did, despite arriving at
+similar theories.
+The work of \citet{AACMM21} is designed first and foremost to facilitate
+programming language mechanisations, and thus pays a lot of attention to syntax
+and traversals of it, making sure that the results compute well in Agda.
+On the other hand, \citeauthor{FPT99} started from theoretical investigations
+about the category of models of theories with binding, and only applied their
+work to the mechanisation of programming languages in the much later work of
+\citet{FS22}.
 
@@
 The resulting metavariables can then stand for arbitrary \emph{open} terms, and
-thus each one remembers its context and requires a substitution whenever it is
-used.
+thus each one remembers its context and requires an explicit substitution
+whenever it is used.
 
@@
 (i.e.\ Agda variables).
-\todo{Congruence once and for all.}
+
+More concrete work by \citet{MMS18}, building on the approach to semantics of
+\citet{ACMM17} and \cref{sec:gen-sem}, also deals with syntactic contexts into
+which terms can be substituted (as part of developing notions of contextual
+equivalence).
+However, this work makes use of several different notions of syntactic context,
+rather than just the one given naturally by metavariables in the framework of
+\citeauthor{FS22}.
+This suggests that more research is needed about the various roles of
+metavariables before any particular approach is standardised.
 
@@
 satisfying the termination checker and avoiding sized types.
+
+Despite all of these differences,
+I chose to base the work of this thesis on that of \citet{AACMM21} for largely
+circumstantial reasons.
+I only became aware of the work of \citet{FS22} when it was published, by which
+time I had already completed the bulk of the work presented in this thesis.
+Also, one of the authors of the former paper is my PhD supervisor, and the other
+authors too were working nearby, so it was easy to discuss their work quickly
+and informally.
+I think it is clear what it would mean to adapt the later work of this thesis to
+a framework in the style of \citeauthor{FS22} rather than \citeauthor{AACMM21},
+but it may not be obvious how usage restrictions (for ordinary variables) and
+metavariables should interact.
+
+Notably for this thesis, the work of \citet{FPT99} has been extended to
+substructural syntaxes by \citet{TP06}.
+This work gives a mechanism for turning a description of contexts and their
+structural rules (expressed as a pseudo-monad on the 2-category of categories)
+into a framework for defining substructural syntaxes, and more generally a
+category of algebras of which the syntax is the initial object.
+As examples, they give the untyped $\lambda$-calculus, an untyped multiplicative
+linear logic, and a bunched logic.
+These examples show a broad range of substructural disciplines they support ---
+comparable to the work of \citet{LicataSR17} (which I discuss in
+\cref{sec:semirings-conc}), and more than I discuss in this thesis.
+However, they also show two of the limitations of their approach.
+Firstly, this work provides no way to track types, though it should be possible
+to incorporate types at the expense of complicating the categorical
+constructions they use.
+Secondly, it appears to be impossible to encode the syntactic forms used for the
+additive connectives (i.e., the Cartesian product and coproduct) of linear
+logic.
+Essentially, subterms can only be combined into a larger term in the same ways
+as how contexts can be appended together.
+For example, in a bunched logic, contexts can be combined through both sharing
+conjunction (the Cartesian product) and separating conjunction
+(a monoidal product).
+Correspondingly, the syntax descriptions of \citeauthor{TP06} allow for the
+syntax of sharing pairs and separating pairs.
+However, in the case of linear logic, contexts can only be combined via a
+monoidal product, so we only get separating pairs (tensor-products) and not
+sharing pairs (with-products).
+\todo{Maybe this should go in the linearity chapter.}
+
+\subsection{Substitution-based semantics}
+
+A feature shared by the frameworks of \citet{AACMM21} and \citet{FPT99} (and
+the rest of the work described in \cref{sec:fiore}) is that they are both often
+concerned about how renamings act upon semantics/models.
+The basic currency of both systems is presheaves on the category of contexts
+under renaming, and when the user wants to produce a semantics, they must show
+that the desired semantics forms such a presheaf --- amounting to showing that
+the relevant type family respects renamings (in the direction that may introduce
+new, unused variables to the context).
+
+In contrast, the work of \citet{HHLM22} is based around the category of contexts
+under \emph{substitution} or, more precisely, the morphisms given by semantic
+environments of whatever semantics is being defined.
+
+\begin{itemize}
+  \item Hard to bootstrap.
+  \item All terms have the same, countably infinite context.
+\end{itemize}
+
+\subsection{Nominal techniques}
+
+There have been essentially two approaches when it comes to incorporating
+nominal techniques into proof assistants.
+The first is to develop a new foundational theory and develop a new proof
+assistant on top of it.
+The second is to take an existing proof assistant and provide a library,
+possibly based on unsafe features.
+
+The best-developed nominal foundational theory at the time of writing is
+\emph{Fraenkel-Mostowski (FM) set theory}, as introduced by
+\citet{Gabbay-thesis} and \citet{GP02}.
+\citeauthor{Gabbay-thesis}'s thesis presents FM set theory, and then uses it as
+the basis for a proof assistant Isabelle/FM and programming language FreshML\@.
+FM is a variant of ZF set theory containing an infinite set $\mathbb A$ of
+\emph{atoms} or \emph{names}, and the \emph{equivariance axiom}, stating that
+every FM proposition is invariant under consistent permutation of $\mathbb A$.
+In this setting, one can define the quantifier
+$\reflectbox{$\mathsf N$}a.~\phi$,
+read as introducing a \emph{new} or \emph{fresh} name $a$ to be used in the
+proposition $\phi$.
+As a material set theory, FM is unusual in that it refutes the axiom of choice.
+In terms of applications to proof assistants, this means that FM is incompatible
+with much existing Isabelle-based work, including anything using the Hilbert
+$\varepsilon$-operator, which is used liberally in Isabelle/HOL\@.
+For example, if the syntax of a programming language is formalised in FM, then
+there is no formal connection to the kind of foundations in which interesting
+denotational semantics have likely been formalised.
+
+Related to nominal set theories is the recent work of \citet{Pitts23} about
+theories of \emph{locally nameless sets}.
+This behaves quite similarly to nominal theories, in that free variables have
+names, and renaming is given by primitive operations.
+In fact, \citet{Pitts23} shows that every locally nameless set is a nominal set.
+
+The other approach --- implementing a nominal library within an existing proof
+assistant --- is exemplified by the Nominal Isabelle library of \citet{Urban08}.
+This work is based around having a countably infinite discrete type
+\texttt{name} (which could be a type of natural numbers or strings, for example)
+which is treated abstractly, and then defining liftings of permutations on
+\texttt{name} to permutations on any other types involving \texttt{name}s.
+Because these liftings are defined explicitly within Isabelle/HOL, the Nominal
+Isabelle setup amounts to an explicit model of a nominal logic within
+Isabelle/HOL.
+This setup resembles what \citet{AACMM21} did within Agda, and similarly what I
+do later in this thesis --- creating a library within an existing proof
+assistant.
+The main downside, compared to working internally to a nominal set theory, is
+that the external constructions of Nominal Isabelle are more complicated, and
+involve keeping track of more properties that do not come as part of the theory
+the proof assistant understands.
+
+There has also been an effort to replicate the Nominal Isabelle work in Coq by
+\citet{ABW06}.
+However, this work appears to have been abandoned before publication, and the
+approach is unclear from the work-in-progress paper.
+Nominal libraries have also been made in general-purpose programming languages,
+like in Haskell with the nom package~\citep{Gabbay20}.
+
+From a broader perspective, I claim that nominal techniques solve a subtly
+different problem to that solved by the likes of Autosubst, the work of Fiore,
+and the techniques discussed in detail in this chapter.
+The names found in nominal techniques are more general than (names of)
+variables in formal systems.
+This means that nominal techniques have been applicable to problems outside
+standard programming language theory\todo{Find applications}.
+However, in certain ways, this is arguably not a good fit for type systems.
+In a basic nominal formalisation of a syntax with binding, there is no real
+distinction between the context of a term and the collection of that term's free
+variables.
+This, for example, leaves no natural place to put the types of variables, except
+as either a partial function from the set of names to types, or from the
+computed set of free variables to types.
+The discrepancy becomes even more apparent in substructural systems, as we see
+later, where we want tight control over the context, as opposed to the scope.
+
+\subsection{Others}
+
+\begin{itemize}
+  \item JFDI --- \citet{BHKM12}
+\end{itemize}

### 2023-05-02 tex/rensub.tex
[ writing ] mostly C3 related work

@@
 
-\begin{itemize}
-  \item Everything presented in this chapter is perfect.
-  \item But equivalence of substitutions is hard to define/work with.
-  \item The process of this development gives a template for defining
-    simultaneous substitution for other substructural calculi.
-\end{itemize}
-
 I have made few unforced design decisions in this chapter, so there are few
-aspects I can see future work improving upon.
+aspects I can see future work improving upon.\todo{Which?}
 However, future work may want to extend the work of this chapter, in which case
 there are some unanswered questions.
-Principle among these, in my mind, is dealing with equivalence/equality of
+Principal among these, in my mind, is dealing with equivalence/equality of
 environments.
+We want to talk about equality of environments for two related purposes.
+The most immediate is that we want to develop the equational theory of renaming
+and substitution --- for example, when we use \cref{thm:env-comp} to compose
+substitutions, we expect that composition to be associative and unital (with
+respect to \cref{thm:env-id}).
+These equations of substitutions should yield equations on terms in which such
+substitutions have been applied.
+Slightly more abstractly, I would like to develop a theory of
+\emph{quantitative multicategories}, in which multimorphisms have in their
+domain a list of objects paired with usage annotations.
+I would hope for $\name$ types and terms to give an example quantitative
+multicategory, analogously to how Lambek calculus gives an example of an
+ordinary multicategory and the simply typed $\lambda$-calculus gives an example
+of a Cartesian multicategory.
+
 Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
@@
 definition does not clearly help.
+
+I hope that people working on substructural type systems in the future can take
+inspiration from the process laid out in \cref{sec:lrkits} when working out the
+appropriate notion of environment for their discipline.
+Particularly, \cref{def:lr-rec-env} (the recursive definition) should serve as a
+specification, if not the actual implementation, when coming across a new
+substructural discipline.
+As for the progression to \cref{def:lr-env}, this appears to arise from the fact
+that the quantitative usage information is a refinement of the intuitionistic de
+Bruijn index-based syntax.
+%Being just a refinement means that the action of an environment on a variable
+%is largely as it was in the intuitionistic case,

### 2023-05-02 tex/simple.tex
[ writing ] lots of background, before feedback on later chapters
[ edit ] new draft of chapters 5 and 6
[ edit ] another draft of chapters 5 and 6
[ writing ] mostly C3 related work

@@
 substituted in to share a context.
+\todo{Maybe delete multicategory remark.}
 
-\section{Term representation}
+This chapter almost entirely reviews prior work.
+In \cref{sec:terms}, I use a presentation of the dependently typed encoding of
+the simply typed $\lambda$-calculus from \citet{AR99} to set notational
+conventions.
+\Cref{sec:kits} presents the unpublished work of \citet{McBride05},
+successively deriving simultaneous renaming and simultaneous substitution for
+the terms defined in \cref{sec:terms}.
+The rest of the chapter generalises the shared core of renaming and substitution
+in two dimensions: in \cref{sec:gen-sem} following \citet{ACMM17} to cover
+semantic traversals, and in \cref{sec:gen-syn} following \citet{AACMM21} to
+cover a whole range of simply typed syntaxes with binding, rather than just
+a specific syntax.
+Finally, I review some related work in \cref{sec:mech-related}.
+
+\section{Term representation}\label{sec:terms}
 \input{terms.tex}
@@
 \input{kits.tex}
-\section{Generic semantics}
+\section{Generic semantics}\label{sec:gen-sem}
 \input{gen-sem.tex}
-\section{Generic syntax}
-\section{Natural deduction vs sequent calculus}
-\input{ndsc.tex}
+\section{Generic syntax}\label{sec:gen-syn}
+\input{gen-syn.tex}
+%\section{Natural deduction vs sequent calculus}
+%\input{ndsc.tex}
+\section{Related work}\label{sec:mech-related}
+\input{mech-related.tex}

### 2023-05-15 agda/PD.lagda.tex
[ writing ] about PD modal logic

\begin{code}
{-# OPTIONS --safe --without-K --postfix-projections #-}

module PD where

  data Ty : Set where
    ι : Ty
    □ : Ty → Ty

  data Jud : Set where
    true valid : Ty → Jud

  data True : Jud → Set where
    true : ∀ {A} → True (true A)
  data Valid : Jud → Set where
    valid : ∀ {A} → Valid (valid A)

  data TypeMatch (A : Ty) : Jud → Set where
    true : TypeMatch A (true A)
    valid : TypeMatch A (valid A)

  infix 50 _++_

  data Tree (X : Set) : Set where
    [_] : X → Tree X
    [] : Tree X
    _++_ : Tree X → Tree X → Tree X

  data All {X} (T : X → Set) : Tree X → Set where
    [_] : ∀ {x} → T x → All T [ x ]
    [] : All T []
    _++_ : ∀ {xt yt} → All T xt → All T yt → All T (xt ++ yt)

  data Any {X} (T : X → Set) : Tree X → Set where
    here : ∀ {x} → T x → Any T [ x ]
    ↙ : ∀ {xt yt} → Any T xt → Any T (xt ++ yt)
    ↘ : ∀ {xt yt} → Any T yt → Any T (xt ++ yt)

  map-Any : ∀ {X} {T U : X → Set} →
    (∀ {x} → T x → U x) → (∀ {xt} → Any T xt → Any U xt)
  map-Any f (here x) = here (f x)
  map-Any f (↙ i) = ↙ (map-Any f i)
  map-Any f (↘ i) = ↘ (map-Any f i)

  open import Data.Empty

  Any-⊥ : ∀ {X xt} → Any {X = X} (λ _ → ⊥) xt → ⊥
  Any-⊥ (↙ i) = Any-⊥ i
  Any-⊥ (↘ i) = Any-⊥ i

  open import Data.Product as ×
  open import Data.Sum as ⊎
  open import Function.Base using (id; _∘_; _|>_; case_of_)
  open import Relation.Binary.PropositionalEquality as ≡ using (_≡_; refl)

  lookup-All : ∀ {X} {T : X → Set} {xt x} → All T xt → Any (x ≡_) xt → T x
  lookup-All [ t ] (here refl) = t
  lookup-All (ttl ++ ttr) (↙ i) = lookup-All ttl i
  lookup-All (ttl ++ ttr) (↘ i) = lookup-All ttr i

  Ctx : Set
  Ctx = Tree Jud

  infix 40 _∋_ _⊢_

  _∋_ : Ctx → Ty → Set
  Γ ∋ A = Any (TypeMatch A) Γ

  view-∋-true-valid : ∀ {Γ A} → Γ ∋ A → Any (true A ≡_) Γ ⊎ Any (valid A ≡_) Γ
  view-∋-true-valid (here true) = inj₁ (here refl)
  view-∋-true-valid (here valid) = inj₂ (here refl)
  view-∋-true-valid (↙ v) = ⊎.map ↙ ↙ (view-∋-true-valid v)
  view-∋-true-valid (↘ v) = ⊎.map ↘ ↘ (view-∋-true-valid v)

  keep-valid : Ctx → Tree Ty
  keep-valid [ true A ] = []
  keep-valid [ valid A ] = [ A ]
  keep-valid [] = []
  keep-valid (Γ ++ Δ) = keep-valid Γ ++ keep-valid Δ

  all-valid : Tree Ty → Ctx
  all-valid [ A ] = [ valid A ]
  all-valid [] = []
  all-valid (As ++ Bs) = all-valid As ++ all-valid Bs

  filter-valid : Ctx → Ctx
  filter-valid = all-valid ∘ keep-valid

  pres-valid : ∀ {Γ A} →
    Any (valid A ≡_) Γ → Any (valid A ≡_) (filter-valid Γ)
  pres-valid (here refl) = here refl
  pres-valid (↙ i) = ↙ (pres-valid i)
  pres-valid (↘ i) = ↘ (pres-valid i)

  data _⊢_ (Γ : Ctx) : Ty → Set where
    var : ∀ {A} → Γ ∋ A → Γ ⊢ A
    □i : ∀ {A} → filter-valid Γ ⊢ A → Γ ⊢ □ A
    □e : ∀ {A B} → Γ ⊢ □ A → Γ ++ [ valid A ] ⊢ B → Γ ⊢ B

  record Env (T : Ctx → Ty → Set) (Γ Δ : Ctx) : Set where
    field
      true-part : ∀ {A} → Any (true A ≡_) Δ → T Γ A
      valid-part : ∀ {A} → Any (valid A ≡_) Δ → T (filter-valid Γ) A
  open Env public

  Ren = Env _∋_
  Sub = Env _⊢_

  view-Env-++ : ∀ {T Γ Δl Δr} → Env T Γ (Δl ++ Δr) → Env T Γ Δl × Env T Γ Δr
  view-Env-++ ρ .proj₁ .true-part = ρ .true-part ∘ ↙
  view-Env-++ ρ .proj₁ .valid-part = ρ .valid-part ∘ ↙
  view-Env-++ ρ .proj₂ .true-part = ρ .true-part ∘ ↘
  view-Env-++ ρ .proj₂ .valid-part = ρ .valid-part ∘ ↘

  filter-valid-Valid : ∀ {Γ} → All Valid (filter-valid Γ)
  filter-valid-Valid {Γ = [ true A ]} = []
  filter-valid-Valid {Γ = [ valid A ]} = [ valid ]
  filter-valid-Valid {Γ = []} = []
  filter-valid-Valid {Γ = Γ ++ Δ} = filter-valid-Valid ++ filter-valid-Valid

  -- Valid-in-filter-valid :
  --   ∀ {Γ A} → Any (valid A ≡_) Γ → Any (valid A ≡_) (filter-valid Γ)
  -- Valid-in-filter-valid i = {!!}

  filter-valid-⊆ : ∀ {T Γ} → Any T (filter-valid Γ) → Any T Γ
  filter-valid-⊆ {Γ = [ valid A ]} i = i
  filter-valid-⊆ {Γ = Γ ++ Δ} (↙ i) = ↙ (filter-valid-⊆ i)
  filter-valid-⊆ {Γ = Γ ++ Δ} (↘ i) = ↘ (filter-valid-⊆ i)

  filter-valid-idem : ∀ {Γ} → filter-valid (filter-valid Γ) ≡ filter-valid Γ
  filter-valid-idem {Γ = [ true A ]} = refl
  filter-valid-idem {Γ = [ valid A ]} = refl
  filter-valid-idem {Γ = []} = refl
  filter-valid-idem {Γ = Γ ++ Δ} =
    ≡.cong₂ _++_ filter-valid-idem filter-valid-idem

  filter-valid-Ren : ∀ {Γ} → Ren Γ (filter-valid Γ)
  filter-valid-Ren .true-part i = case (lookup-All filter-valid-Valid i) of λ ()
  filter-valid-Ren .valid-part i = map-Any (λ { refl → valid }) i

  module _ {T : Ctx → Ty → Set}
    (ren : ∀ {Γ Δ A} → Ren Γ Δ → T Δ A → T Γ A)
    where

    lookup : ∀ {Γ Δ A} → Env T Γ Δ → Δ ∋ A → T Γ A
    lookup ρ v = view-∋-true-valid v |>
      [ ρ .true-part , ren filter-valid-Ren ∘ ρ .valid-part ]′

  module BindEnv {T : Ctx → Ty → Set}
    (vr : ∀ {Γ A} → Any (TypeMatch A) Γ → T Γ A)
    (wk : ∀ {Γ A Θ} → T Γ A → T (Γ ++ Θ) A)
    where

    vt : ∀ {Γ A} → Any (true A ≡_) Γ → T Γ A
    vt = vr ∘ map-Any (λ { refl → true })
    vv : ∀ {Γ A} → Any (valid A ≡_) Γ → T Γ A
    vv = vr ∘ map-Any (λ { refl → valid })

    bindEnv : ∀ {Γ Δ Θ} → Env T Γ Δ → Env T (Γ ++ Θ) (Δ ++ Θ)
    bindEnv ρ .true-part (↙ i) = wk (ρ .true-part i)
    bindEnv ρ .true-part (↘ i) = vt (↘ i)
    bindEnv ρ .valid-part (↙ i) = wk (ρ .valid-part i)
    bindEnv ρ .valid-part (↘ i) = vv (↘ (pres-valid i))

  module _ {T : Ctx → Ty → Set} where

    Env-filter-valid : ∀ {Γ Δ} →
      Env T Γ Δ → Env T (filter-valid Γ) (filter-valid Δ)
    Env-filter-valid ρ .true-part i =
      case (lookup-All filter-valid-Valid i) of λ ()
    Env-filter-valid ρ .valid-part i =
      let t = ρ .valid-part (filter-valid-⊆ i) in
      ≡.subst (λ ∙ → T ∙ _) (≡.sym filter-valid-idem) t

  module Trav {T : Ctx → Ty → Set}
    (vr : ∀ {Γ A} → Any (TypeMatch A) Γ → T Γ A)
    (ren : ∀ {Γ Δ A} → Ren Γ Δ → T Δ A → T Γ A)
    (tm : ∀ {Γ A} → T Γ A → Γ ⊢ A)
    where

    wk : ∀ {Γ A Θ} → T Γ A → T (Γ ++ Θ) A
    wk = ren λ where
      .true-part i → ↙ (map-Any (λ { refl → true }) i)
      .valid-part i → ↙ (map-Any (λ { refl → valid }) (pres-valid i))

    open BindEnv vr wk

    trav : ∀ {Γ Δ A} → Env T Γ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (lookup ren ρ x)
    trav ρ (□i M) = □i (trav (Env-filter-valid ρ) M)
    trav ρ (□e M N) = □e (trav ρ M) (trav (bindEnv ρ) N)
\end{code}

\begin{code}
  ren^∋ : ∀ {Γ Δ A} → Env _∋_ Γ Δ → Δ ∋ A → Γ ∋ A
  ren^∋ ρ v = view-∋-true-valid v |>
    [ ρ .true-part , filter-valid-⊆ ∘ ρ .valid-part ]′

  ren^⊢ : ∀ {Γ Δ A} → Env _∋_ Γ Δ → Δ ⊢ A → Γ ⊢ A
  ren^⊢ = Trav.trav id ren^∋ var

  sub^⊢ : ∀ {Γ Δ A} → Env _⊢_ Γ Δ → Δ ⊢ A → Γ ⊢ A
  sub^⊢ = Trav.trav var ren^⊢ id
\end{code}

### 2023-05-15 tex/linear.tex
[ writing ] about PD modal logic

@@
 type systems, which I have not discussed, but has been considered by much prior
-work\todo{cite}.
+work~\citep{POPLmark}.
 
-%Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
-%Its distinguishing feature is that hypotheses, by default, must be used exactly
-%once.
-%I devote a chapter to introducing linear logic because it is better understood
-%than any individual or general semiring-annotated calculi, and is thus an
-%important point of comparison.
-%Linear logic will be an instance of the semiring-annotated systems we study
-%later.
+Another potential feature --- required for many desired semantics but only
+relatively recently appearing in programming languages --- that is not handled
+by the techniques of \cref{sec:simple} is restrictions on how variables can be
+used.
+In the simple type systems covered in \cref{sec:simple}, the only restriction on
+the use of a variable is that it is used as an expression of the appropriate
+type.
+However, we could also consider having somehow ``guarded'' subexpressions in
+which some variables bound outside cannot occur, or systems in which the use of
+a variable in one part of an expression precludes its use in another part of
+that expression (the variable having been ``used up'').
+I will introduce and motivate some such systems in this chapter, and argue that
+they are impossible to capture directly in the framework presented in
+\cref{sec:gen-syn}.
+
+Languages and systems implementing features inspired by linear logic include
+Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
+session types~\citep{HLVCCDMPRTVTZ16}.
+Additionally, forms of linearity have been used in theoretical work, for example
+to bound computational complexity~\citep{GSS92,Hofmann03} or maintain
+differentiability~\citep{ER03,Ehrhard18}.
+As for modality, \citet{IMO20} show that common data-flow analyses for
+imperative languages can be recast as modal type systems, though they have
+generally been developed in an ad hoc fashion.
+Such a recasting may be useful to understand the metatheoretic properties of
+programs passing such static analyses.
+More user-facing implementations of modal type systems include the system for
+stack allocation in OCaml~\citep{DW22} and the approach to erasure found in
+Agda and Idris\todo{cite}.
+
+In this chapter\ldots\todo{Chapter map}
+
+\section{Intuitionistic S4 modal logic}\label{sec:modal}
+\input{modal.tex}
 

### 2023-05-15 tex/modal.tex
[ writing ] about PD modal logic

Modal logics, and in particular their modalities, are traditionally presented in
an axiomatic style.
For example, the common modal logic S4 may be presented by taking a presentation
of classical logic, adding a unary $\Box$ (``box'') operator to the formulas,
keeping the logical rules as-are, and adding all of the axioms and rules
listed in \cref{fig:S4-axioms}.
Note that, unlike elsewhere in this thesis, the empty context in the
necessitation rule \TirName{N} is a true restriction.
While the axioms hold in any context, thanks to weakening, necessitation only
holds in the empty context.
We may apply weakening to the conclusion of \TirName{N} to embed such a
derivation into a larger derivation, but we may not use hypotheses from the
outside in such a subderivation.

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\vdash A}
      \infer1[N]{\vdash \Box A}
    }
    \and
    \ebrule{%
      \infer0[K]{\vdash \Box\plr{A \to B} \to \plr{\Box A \to \Box B}}
    }
    \and
    \ebrule{%
      \infer0[T]{\vdash \Box A \to A}
    }
    \and
    \ebrule{%
      \infer0[4]{\vdash \Box A \to \Box\plr{\Box A}}
    }
  \end{mathpar}
  \caption{The axioms and rules required in a traditional presentation of S4}
  \label{fig:S4-axioms}
\end{figure}

In this thesis, I am working intuitionistically, so I start with a base of
intuitionistic logic, and study intuitionistic S4 (IS4).

The traditional presentation is convenient for talking about semantics, with
different choices of axioms corresponding fairly directly to natural
restrictions on models.
However, the syntax is quite poorly behaved, at least for programming language
applications.
As mentioned earlier, the necessitation rule makes requirements of its context,
which makes it incompatible with the methods discussed in \cref{sec:simple}.
Essentially, it requires necessitation to be a special case in renaming,
substitution, and all the other traversals.
Additionally, axioms require us to modify the notion of canonical form, and redo
any proofs that rely on all closed terms of function type being
$\lambda$-abstractions.

The methodology I use to produce a nicer syntax is that of \citet{judgmental}:
capturing the desired modality in the judgemental structure of the syntax.
To start, note the following.

\begin{proposition}
  In NJ + \TirName{N} + \TirName{K} + \TirName{T}, the following principle is
  admissible.
  \[
    \ebrule{%
      \hypo{\Box B_1, \ldots, \Box B_n \vdash A}
      \infer1[NKT]{\Box B_1, \ldots, \Box B_n \vdash \Box A}
    }
  \]
\end{proposition}
\begin{proof}
  By induction on $n$.
  When $n = 0$, \TirName{NKT} becomes just the necessitation rule \TirName{N}.
  For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
  $\Box\Gamma$ abbrevates
  $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{T$'$} and
  \TirName{K$'$} abbrevate the obvious combinations of
  \TirName{T} and \TirName{K}, respectively, and \TirName{$\to$-E}.
  I silently apply weakening thoughout the derivation for the sake of concision.
  \[
    \begin{prooftree}
      \hypo{\Box\Gamma, \Box B \vdash A}
      \infer1[$\to$-I]{\Box\Gamma \vdash \Box B \to A}
      \infer1[IH]{\Box\Gamma \vdash \Box\plr{\Box B \to A}}
      \infer0[Var]{\Box B \vdash \Box B}
      \infer1[T$'$]{\Box B \vdash \Box\Box B}
      \infer2[K$'$]{\Box\Gamma, \Box B \vdash \Box A}
    \end{prooftree}
  \]
\end{proof}

However, we do not want to add such a rule to our system because it is not
stable under substitution.
For example, consider a substitution
$\sigma : \Box A \wedge \Box B \env\vdash \Box A, \Box B$.
The domain context contains a single formula headed by $\wedge$, not $\Box$, so
when we apply $\sigma$ to a derivation ending with an application of
\TirName{NKT} deriving $\Box A, \Box B \vdash \Box C$, it becomes unclear how to
conclude $\Box A \wedge \Box B \vdash \Box C$, as \TirName{NKT} does not apply.

Instead of constraining the types of all of the items in the context, the
solution given by \citet{judgmental} is to diversify the shapes of contexts, and
constrain those shapes when giving our necessitation rule.
In this case, we are interested in hypotheses being $\Box$-like, so we introduce
a judgement form $A~\mathit{valid}$, contrasting with the judgement we had
previously written just $A$, but will now write $A~\mathit{true}$.
We construct the system so that $A~\mathit{valid}$ is equivalent to
$\Box A~\mathit{true}$.
%Note that, in S4, $\Box\Box A$ is logically equivalent to $\Box A$
Because the choice of $\mathit{valid}$ or $\mathit{true}$ is part of the
structure of the context, rather than being part of the type, we are able to
define substitution between contexts in a way that treats valid and true entries
differently --- restricting substitution so that it is compatible with an
adapted version of \TirName{NKT}.

Let us take the natural deduction presentation of the simply typed
$\lambda$-calculus given in \cref{sec:terms}.
Where that system has sequents of the form $A_1, \ldots, A_n \vdash B$, we
instead now write
$A_1~\mathit{true}, \ldots, A_n~\mathit{true} \vdash B~\mathit{true}$.
We may also have hypothetical occurrences of $A~\mathit{valid}$, but all logical
rules will target judgements of the form $A~\mathit{true}$.
Hypotheses of the form $A~\mathit{valid}$ arise from the elimination rule for
the $\Box$ operator, \TirName{$\Box$-E}.
The only special thing about valid hypotheses is that, once they are bound, they
are preserved by the \TirName{$\Box$-I} rule, which is our recasting of
\TirName{NKT} as the introduction rule for $\Box$.
Otherwise, like true hypotheses, an $A~\mathit{valid}$ hypothesis can be used to
obtain $A~\mathit{true}$, via the \TirName{vVar} rule, as justified by axiom
\TirName{4}.

The three new rules are listed in \cref{fig:PD}.
Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
leaving only valid hypotheses.
Because I am working with a de Bruijn index-style presentation, weakening needs
to be admissible.
Therefore, the \TirName{$\Box$-I} rule needs to remove its true hypotheses,
rather than only applying in a context in which there are no true hypotheses.

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma_v \vdash A~\mathit{true}}
      \infer1[$\Box$-I]{\Gamma \vdash \Box A~\mathit{true}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash \Box A~\mathit{true}}
      \hypo{\Gamma, A~\mathit{valid} \vdash B~\mathit{true}}
      \infer2[$\Box$-E]{\Gamma \vdash B~\mathit{true}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \ni \plr{A~\mathit{valid}}}
      \infer1[vVar]{\Gamma \vdash A~\mathit{true}}
    }
  \end{mathpar}
  \caption{The new rules of the \citeauthor{judgmental} presentation of IS4.}
  \label{fig:PD}
\end{figure}

Whereas \citet{judgmental} prove single substitution for both true and valid
hypotheses (as well as \emph{possible} hypotheses, and for possible judgements,
in their system dealing with a possibility modality $\Diamond$) separately, I
build on the simultaneous substitution procedure given in
\cref{sec:syntactic-kits} to give a unified simultaneous substitution procedure.
Recall that the basic definition was that of \emph{$\vDash$-environment},
defined in the intuitionistic case by
$\Gamma \env\vDash \Delta \coloneqq \Pi A.~\Delta \ni A \to \Gamma \vDash A$.
Essentially the same formula deals with true hypotheses, but for valid
hypotheses, we need to think of something new.
We want to be able to instantiate $\vDash$ to $\vdash$ so as to derive
substitution, but $\Gamma \vdash A~\mathit{valid}$ is not a valid sequent in our
system.
Instead, I consider $\Gamma \vdash \Box A~\mathit{true}$, and in particular the
canonical way of deriving such a sequent: the \TirName{$\Box$-I} rule.
\TirName{$\Box$-I} says that we can get
$\Gamma \vdash \Box A~\mathit{true}$ from $\Gamma_v \vdash A~\mathit{true}$.
I take the form of latter sequent to replace $\Gamma \vDash A~\mathit{valid}$,
giving the following definition.

\begin{definition}\label{def:PD-env}
  A \emph{(modal) $\vDash$-environment} from $\Gamma$ to $\Delta$ is given as
  follows.
  \[
    \Gamma \env\vDash \Delta \coloneqq
    \Pi A.~\plr{\Delta \ni A~\mathit{true} \to \Gamma \vDash A}
    \times \plr{\Delta \ni A~\mathit{valid} \to \Gamma_v \vDash A}
  \]
\end{definition}

With this definition, I reproduce the kit-based machinery from
\cref{sec:syntactic-kits}.
However, I make crucial use of the \emph{stability under renaming} property ---
rather than \emph{stability under context extension} --- to deal with not just
context extensions, but also discarding of true hypotheses by the
\TirName{$\Box$-I} rule.
To state this property, I firstly need to be precise about what a renaming is.

\begin{definition}\label{def:PD-var}
  The set of variables in context $\Gamma$ of type $A$, notated $\Gamma \ni A$,
  is the disjoint union of the sets $\Gamma \ni A~\mathit{true}$ and
  $\Gamma \ni A~\mathit{valid}$ --- respectively, the true variables and the
  valid variables in $\Gamma$ of type $A$.

  The set of \emph{renamings} from $\Gamma$ to $\Delta$ is
  $\Gamma \env\ni \Delta$.
\end{definition}

\Cref{def:PD-var} allows us to state a unified variable rule: For each
$x : \Gamma \ni A$, we have a corresponding term
$x : \Gamma \vdash A~\mathit{true}$.
However, with the new definition of both environments and variables, we must
check that they interact in the desired way.

\begin{lemma}[lookup]\label{thm:PD-lookup}
  If $\vDash$ respects renaming (i.e., we have a function
  $\mathrm{ren}^\vDash :
  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$
  for each $\Gamma$, $\Delta$, and $A$), then
  from an environment $\rho : \Gamma \env\vDash \Delta$ and a variable
  $x : \Delta \ni A$, we get a value $\rho(x) : \Gamma \vDash A$.
\end{lemma}
\begin{proof}
  I proceed by cases on whether $x$ is a true or valid variable.
  In the true case, the result is straightforward.
  In the valid case, \cref{def:PD-env} gives us a value in $\Gamma_v \vDash A$.
  We need a renaming of type $\Gamma \env\vDash \Gamma_v$ to get the desired
  value in $\Gamma \vDash A$.
  The renaming is the one that, conceptually, discards the true hypotheses.
\end{proof}

We then need to prove that environments are preserved by everything we do to
contexts in a derivation.
One thing we do, like in the simply typed case, is to bind new variables, and
this is handled by \cref{thm:PD-bindEnv}.
The other thing, which is new for modal logic, is to discard the true
hypotheses, as in the \TirName{$\Box$-I} rule, which is handled by
\cref{thm:PD-vEnv}.

\begin{lemma}[bindEnv]\label{thm:PD-bindEnv}
  If we have $\mathrm{vr} : \Gamma \ni A \to \Gamma \vDash A$ for all
  $\Gamma$ and $A$, and
  $\mathrm{ren}^\vDash :
  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$
  for all $\Gamma$, $\Delta$, and $A$,
  then from an environment $\rho : \Gamma \env\vDash \Delta$
  we can get an environment of type $\Gamma, \Theta \env\vDash \Delta, \Theta$
  for all $\Gamma$, $\Delta$, and $\Theta$.
\end{lemma}
\begin{proof}
  I consider separately the part of the environment we are constructing that
  deals with true hypotheses and the part that deals with valid hypotheses.
  These cases correspond to the two factors in the expression in
  \cref{def:PD-env}.

  The case for true hypotheses is essentially the same as in
  \cref{sec:syntactic-kits}.
  We take cases on whether the $x : \Delta, \Theta \ni A~\mathit{true}$ is in
  $\Delta$ or $\Theta$.
  In the $\Delta$ case, we apply $\rho$ and then rename via
  $\mathrm{ren}^\vDash$ using the obvious renaming of type
  $\Gamma, \Theta \env\ni \Gamma$.
  In the $\Theta$ case, we embed our $z : \Theta \ni A~\mathit{true}$ as
  $\searrow z : \Gamma, \Theta \ni A~\mathit{true}$, and use $\mathrm{vr}$ to
  get the required value.

  The case for valid hypotheses is similar.
  We take the same cases, with the $\Delta$ case differing only in which part of
  $\rho$ we need to use (the valid part, rather than the true part).
  In the $\Theta$ case, we have $z : \Theta \ni A~\mathit{valid}$.
  The $_v$ operation keeps all valid hypotheses, so we have
  $z_v : \Theta_v \ni A~\mathit{valid}$.
  Using the fact that $(\Gamma, \Theta)_v = \Gamma_v, \Theta_v$, we have
  $\searrow z_v : (\Gamma, \Theta)_v \ni A~\mathit{valid}$, and we use
  $\mathrm{vr}$ to get the required value.
\end{proof}

\begin{lemma}\label{thm:PD-vEnv}
  From an environment $\rho : \Gamma \env\vDash \Delta$, we can get an
  environment $\rho_v : \Gamma_v \env\vDash \Delta_v$.
\end{lemma}
\begin{proof}
  Suppose $x : \Delta_v \ni A$.
  Because $\Delta_v$ contains only and all of the valid hypotheses from
  $\Delta$, we actually have $x' : \Delta \ni A~\mathit{valid}$.
  Applying the second part of $\rho$ on $x'$ gives us a value in
  $\Gamma_v \vDash A$.
  We actually wanted a value in $\plr{\Gamma_v}_v \vDash A$, but $_v$ is
  idempotent, so we already have this.
\end{proof}

Then, we use all of the previous features to prove the key theorem.

\begin{theorem}[trav]\label{thm:PD-trav}
  For any notion of value $\vDash$ which is stable under renaming, admits
  variables (via some $\mathrm{vr}$ as in \cref{thm:PD-bindEnv}), and
  embeds into terms (via some
  $\mathrm{tm} : \Gamma \vDash A \to \Gamma \vdash A~\mathit{true}$ for all
  $\Gamma$ and $A$),
  an environment $\rho : \Gamma \env\vDash \Delta$ and a term
  $M : \Delta \vdash A~\mathit{true}$ yield a term
  $M[\rho] : \Gamma \vdash A~\mathit{true}$.
\end{theorem}
\begin{proof}
  I proceed by induction on the term $M$.
  Here I consider only variables and the two rules governing the $\Box$
  connective.
  It is easy to adapt this procedure to any of the types from simply typed
  $\lambda$-calculus.

  % var
  I consider variables being given by the unified variable rule proposed in the
  paragraph before \cref{thm:PD-lookup}.
  Given this, from a variable $x : \Delta \ni A$, \cref{thm:PD-lookup} gives us
  a value $\rho(x) : \Gamma \vDash A$, which $\mathrm{tm}$ turns into a term of
  the desired form.

  % \Box-I
  If the term was made from the \TirName{$\Box$-I} rule, we use that rule to
  produce the resulting term, and need to get from a term
  $M : \Delta_v \vdash A~\mathit{true}$ to a term in
  $\Gamma_v \vdash A~\mathit{true}$.
  It is enough to use the induction hypothesis, updating the environment using
  \cref{thm:PD-vEnv}.

  % \Box-E
  If the term was made from the \TirName{$\Box$-E} rule, we again use the same
  rule to construct the output, but have to get from a term
  $M : \Delta \vdash \Box A~\mathit{true}$ to a term in
  $\Gamma \vdash \Box A~\mathit{true}$, and from a term
  $N : \Delta, A~\mathit{valid} \vdash B~\mathit{true}$ to a term in
  $\Gamma, A~\mathit{valid} \vdash B~\mathit{true}$.
  The former follows from a straightforward use of the induction hypothesis,
  while the latter needs \cref{thm:PD-bindEnv} before applying the induction
  hypothesis.
\end{proof}

To get our desired corollaries of simultaneous renaming and substitution, we
first have to show that variables respect renaming.
In the pure intuitionistic case of \cref{sec:simple}, this was trivial because
renamings were precisely functions between sets of variables.
In the modal case, we have to be careful about the distinction between true and
valid hypotheses, and how the context is restricted in the valid case.

\begin{lemma}[variables respect renaming]\label{thm:PD-ren-var}
  Given a renaming $\rho : \Gamma \env\ni \Delta$ and a variable
  $x : \Delta \ni A$, we get a variable in $\Gamma \ni A$.
\end{lemma}
\begin{proof}
  Note that we cannot use \cref{thm:PD-lookup} with $\vDash$ instantiated to
  $\ni$ because it would circularly require that variables respect renaming.
  However, the procedure in this case is similar to that of
  \cref{thm:PD-lookup}.

  We take cases on whether $x$ is true or valid, and the result in the true case
  comes immediately by applying the true part of $\rho$.
  For $x : \Delta \ni A~\mathit{valid}$, $\rho$ gives us a variable in
  $\Gamma_v \ni A$.
  This variable must be valid ($\Gamma_v \ni A~\mathit{valid}$), and from there
  it is easy to get a variable in the larger context $\Gamma$.
\end{proof}

\begin{corollary}[renaming]\label{thm:PD-ren}
  Given a renaming $\rho : \Gamma \env\ni \Delta$ and a term
  $M : \Delta \vdash A~\mathit{true}$, we get a term in
  $\Gamma \vdash A~\mathit{true}$.
\end{corollary}
\begin{proof}
  We use \cref{thm:PD-trav}, with $\vDash$ being $\ni$, $\mathrm{vr}$ being the
  identity function, $\mathrm{tm}$ being the unified variable rule, and renaming
  of $\ni$ being given by \cref{thm:PD-ren-var}.
\end{proof}

\begin{corollary}[substitution]\label{thm:PD-sub}
  Given a substitution $\rho : \Gamma \env\vdash \Delta$ and a term
  $M : \Delta \vdash A~\mathit{true}$, we get a term in
  $\Gamma \vdash A~\mathit{true}$.
\end{corollary}
\begin{proof}
  We use \cref{thm:PD-trav}, with $\vDash$ being $- \vdash -~\mathit{true}$,
  $\mathrm{vr}$ being the unified variable rule, $\mathrm{tm}$ being the
  identity function, and renaming being given by \cref{thm:PD-ren}.
\end{proof}

With these basic syntactic lemmas proved in a manner largely following the
proofs of \cref{sec:simple}, it is plausible that this approach could be
extended to handle generic semantics and generic syntax, following
\cref{sec:gen-sem,sec:gen-syn}, respectively.
However, I hold off from developing this until phrasing it more generally in
\cref{sec:framework}.

### 2023-05-15 tex/rensub.tex
[ writing ] about PD modal logic

@@
 
-I have made few unforced design decisions in this chapter, so there are few
-aspects I can see future work improving upon.\todo{Which?}
-However, future work may want to extend the work of this chapter, in which case
+Future work may want to extend the work of this chapter, in which case
 there are some unanswered questions.

### 2023-06-22 tex/trans.tex
[ writing ] renew generic syntax chapter
[ edit ] inferrule -> ebproof

@@
 
-\subsection{Dual Intuitionistic Linear Logic}\label{sec:dill}
+\subsection{Dual Intuitionistic Linear Logic}\label{sec:trans-dill}
 
@@
 I add sums and with-products to the calculus of \cite{Barber1996}, with the
-obvious rules (stated fully in \cref{fig:dill}).
+obvious rules (stated fully in \cref{fig:dill2}).
 These additive type formers present no additional difficulty to the translation.
@@
   \begin{mathpar}
-    \inferrule*[right=Int-Ax]{ }
-    {\gamma, A; \cdot \vdash A}
+    \ebrule{\infer0[Int-Ax]{\gamma, A; {\cdot} \vdash A}} \and
+    \ebrule{\infer0[Lin-Ax]{\gamma; A \vdash A}} \and
+    \ebrule{\infer0[$I$-I]{\gamma; {\cdot} \vdash I}} \and
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash I}
+      \hypo{\gamma; \delta_2 \vdash A}
+      \infer2[$I$-E]{\gamma; \delta_1, \delta_2 \vdash A}
+    }
     \and
-    \inferrule*[right=Lin-Ax]{ }
-    {\gamma; A \vdash A}
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash A}
+      \hypo{\gamma, \delta_2 \vdash B}
+      \infer2[$\otimes$-I]{\gamma; \delta_1, \delta_2 \vdash A \otimes B}
+    }
     \and
-    \inferrule*[right=$I$-I]{ }
-    {\gamma; \cdot \vdash I}
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash A \otimes B}
+      \hypo{\gamma; \delta_2, A, B \vdash C}
+      \infer2[$\otimes$-E]{\gamma; \delta_1, \delta_2 \vdash C}
+    }
     \and
-    \inferrule*[right=$I$-E]
-    {\gamma; \delta_1 \vdash I \\ \gamma; \delta_2 \vdash A}
-    {\gamma; \delta_1, \delta_2 \vdash A}
+    \ebrule{%
+      \hypo{\gamma; \delta, A \vdash B}
+      \infer1[$\multimap$-I]{\gamma; \delta \vdash A \multimap B}
+    }
     \and
-    \inferrule*[right=$\otimes$-I]
-    {\gamma; \delta_1 \vdash A \\ \gamma, \delta_2 \vdash B}
-    {\gamma; \delta_1, \delta_2 \vdash A \otimes B}
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash A \multimap B}
+      \hypo{\gamma; \delta_2 \vdash A}
+      \infer2[$\multimap$-E]{\gamma; \delta_1, \delta_2 \vdash B}
+    }
     \and
-    \inferrule*[right=$\otimes$-E]
-    {\gamma; \delta_1 \vdash A \otimes B \\ \gamma; \delta_2, A, B \vdash C}
-    {\gamma; \delta_1, \delta_2 \vdash C}
+    \ebrule{%
+      \hypo{\gamma; \cdot \vdash A}
+      \infer1[$\oc$-I]{\gamma; \cdot \vdash \oc A}
+    }
     \and
-    \inferrule*[right=$\multimap$-I]
-    {\gamma; \delta, A \vdash B}
-    {\gamma; \delta \vdash A \multimap B}
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash \oc A}
+      \hypo{\gamma, A; \delta_2 \vdash B}
+      \infer2[$\oc$-E]{\gamma; \delta_1, \delta_2 \vdash B}
+    }
     \and
-    \inferrule*[right=$\multimap$-E]
-    {\gamma; \delta_1 \vdash A \multimap B \\ \gamma; \delta_2 \vdash A}
-    {\gamma; \delta_1, \delta_2 \vdash B}
+    \ebrule{%
+      \infer0[$\top$-I]{\gamma; \delta \vdash \top}
+    }
     \and
-    \inferrule*[right=$\oc$-I]
-    {\gamma; \cdot \vdash A}
-    {\gamma; \cdot \vdash \oc A}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A}
+      \hypo{\gamma, \delta \vdash B}
+      \infer2[$\with$-I]{\gamma; \delta \vdash A \with B}
+    }
     \and
-    \inferrule*[right=$\oc$-E]
-    {\gamma; \delta_1 \vdash \oc A \\ \gamma, A; \delta_2 \vdash B}
-    {\gamma; \delta_1, \delta_2 \vdash B}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A_0 \with A_1}
+      \infer1[$\with$-E$_i$]{\gamma; \delta \vdash A_i}
+    }
     \and
-    \inferrule*[right=$\top$-I]{ }
-    {\gamma; \delta \vdash \top}
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash 0}
+      \infer1[$0$-E]{\gamma; \delta_1, \delta_2 \vdash A}
+    }
     \and
-    \inferrule*[right=$\with$-I]
-    {\gamma; \delta \vdash A \\ \gamma, \delta \vdash B}
-    {\gamma; \delta \vdash A \with B}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A_i}
+      \infer1[$\oplus$-I$_i$]{\gamma; \delta \vdash A_0 \oplus A_1}
+    }
     \and
-    \inferrule*[right=$\with$-E$_i$]
-    {\gamma; \delta \vdash A_0 \with A_1}
-    {\gamma; \delta \vdash A_i}
-    \and
-    \inferrule*[right=$0$-E]
-    {\gamma; \delta_1 \vdash 0}
-    {\gamma; \delta_1, \delta_2 \vdash A}
-    \and
-    \inferrule*[right=$\oplus$-I$_i$]
-    {\gamma; \delta \vdash A_i}
-    {\gamma; \delta \vdash A_0 \oplus A_1}
-    \and
-    \inferrule*[right=$\oplus$-E]
-    {
-      \gamma; \delta_1 \vdash A \oplus B \\
-      \gamma; \delta_2, A \vdash C \\
-      \gamma; \delta_2, B \vdash C
+    \ebrule{%
+      \hypo{\gamma; \delta_1 \vdash A \oplus B}
+      \hypo{\gamma; \delta_2, A \vdash C}
+      \hypo{\gamma; \delta_2, B \vdash C}
+      \infer3[$\oplus$-E]{\gamma; \delta_1, \delta_2 \vdash C}
     }
-    {\gamma; \delta_1, \delta_2 \vdash C}
   \end{mathpar}
-  \label{fig:dill}
+  \label{fig:dill2}
   \caption{The rules of DILL, extended with additive connectives}
@@
   \[
-    \inferrule*[right=$\otimes$-I]
-    {\gamma; \delta_t \vdash t : A \\ \gamma; \delta_u \vdash u : B}
+    \ebrule{%
+      \hypo{\gamma; \delta_t \vdash t : A}
+      \hypo{\gamma; \delta_u \vdash u : B}
+      \infer2[$\otimes$-I]%
       {\gamma; \delta_t, \delta_u \vdash t \otimes u : A \otimes B}
+    }
   \]
@@
   \[
-    \inferrule*[right=$\otimes$-I]
-    {
-      \inferrule*[right=Weak]
-      {\mathit{ih}_t \\\\
-        \gr\omega\gamma, \gr1\delta_t
-        \vdash M_t : A}
-      {\gr\omega\gamma, \gr1\delta_t,
-        \gr0\delta_u
-        \vdash M_t : A}
-      \\
-      \inferrule*[right=Weak]
-      {\mathit{ih}_u \\\\
-        \gr\omega\gamma, \gr1\delta_u
-        \vdash M_u : A}
-      {\gr\omega\gamma, \gr0\delta_t,
-        \gr1\delta_u
-        \vdash M_u : A}
+    \ebrule{%
+      \hypo{\mathit{ih}_t}
+      \infer[no rule]1{\gr\omega\gamma, \gr1\delta_t \vdash M_t : A}
+      \infer1[Weak]%
+      {\gr\omega\gamma, \gr1\delta_t, \gr0\delta_u \vdash M_t : A}
+      \hypo{\mathit{ih}_u}
+      \infer[no rule]1{\gr\omega\gamma, \gr1\delta_u \vdash M_u : A}
+      \infer1[Weak]%
+      {\gr\omega\gamma, \gr0\delta_t, \gr1\delta_u \vdash M_u : A}
+      \infer2[$\otimes$-I]%
+      {\gr\omega\gamma, \gr1\delta_t, \gr1\delta_u
+        \vdash \plr{M_t, M_u} : A \otimes B}
     }
-    {\gr\omega\gamma, \gr1\delta_t,
-      \gr1\delta_u
-      \vdash \plr{M_t, M_u} : A \otimes B}
   \]
@@
 \[
-  \inferrule*[right=$\otimes$-I]
-  {
-    \inferrule*[right=var]{ }
-    {\gr1A : x \vdash A}
-    \\
-    \inferrule*[right=var]{ }
-    {\gr1A : x \vdash A}
-    \\
-    \gr\omega \leq \gr1 + \gr1
+  \ebrule{%
+    \infer0[Var]{\gr1A : x \vdash A}
+    \infer0[Var]{\gr1A : x \vdash A}
+    \hypo{\gr\omega \leq \gr1 + \gr1}
+    \infer3[$\otimes$-I]{\gr\omega A : x \vdash A \otimes A}
   }
-  {\gr\omega A : x \vdash A \otimes A}
 \]
@@
 
-\begin{definition}
+\begin{definition}\label{def:DILL-bottom-up}
   A derivation is said to be \emph{$\instDILL$-bottom-up} if only the following
@@
 
-\subsection{Pfenning Davies}\label{sec:pd}
+\subsection{Pfenning Davies}\label{sec:trans-pd}
 
@@
   \begin{mathpar}
-    \inferrule*[right=hyp]{ }
-    {\gamma; \delta, A\;\mathit{true} \vdash A\;\mathit{true}}
+    \ebrule{%
+      \infer0[hyp]{\gamma; \delta, A\;\mathit{true} \vdash A\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=hyp*]{ }
-    {\gamma, A\;\mathit{valid}; \delta \vdash A\;\mathit{true}}
+    \ebrule{%
+      \infer0[hyp*]{\gamma, A\;\mathit{valid}; \delta \vdash A\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\supset$I]
-    {\gamma; \delta, A\;\mathit{true} \vdash B\;\mathit{true}}
-    {\gamma; \delta \vdash A \supset B\;\mathit{true}}
+    \ebrule{%
+      \hypo{\gamma; \delta, A\;\mathit{true} \vdash B\;\mathit{true}}
+      \infer1[$\supset$I]{\gamma; \delta \vdash A \supset B\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\supset$E]
-    {
-      \gamma; \delta \vdash A \supset B\;\mathit{true} \\
-      \gamma; \delta \vdash A\;\mathit{true}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A \supset B\;\mathit{true}}
+      \hypo{\gamma; \delta \vdash A\;\mathit{true}}
+      \infer2[$\supset$E]{\gamma; \delta \vdash B\;\mathit{true}}
     }
-    {\gamma; \delta \vdash B\;\mathit{true}}
     \and
-    \inferrule*[right=$\Box$I]
-    {\gamma; \cdot \vdash A\;\mathit{true}}
-    {\gamma; \delta \vdash \Box A\;\mathit{true}}
+    \ebrule{%
+      \hypo{\gamma; \cdot \vdash A\;\mathit{true}}
+      \infer1[$\Box$I]{\gamma; \delta \vdash \Box A\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\Box$-E]
-    {
-      \gamma; \delta \vdash \Box A\;\mathit{true} \\
-      \gamma, A\;\mathit{valid}; \delta \vdash B\;\mathit{true}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash \Box A\;\mathit{true}}
+      \hypo{\gamma, A\;\mathit{valid}; \delta \vdash B\;\mathit{true}}
+      \infer2[$\Box$-E]{\gamma; \delta \vdash B\;\mathit{true}}
     }
-    {\gamma; \delta \vdash B\;\mathit{true}}
     \and
-    \inferrule*[right=$\top$-I]{ }
-    {\gamma; \delta \vdash \top\;\mathit{true}}
+    \ebrule{%
+      \infer0[$\top$-I]{\gamma; \delta \vdash \top\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\wedge$-I]
-    {
-      \gamma; \delta \vdash A\;\mathit{true} \\
-      \gamma, \delta \vdash B\;\mathit{true}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A\;\mathit{true}}
+      \hypo{\gamma, \delta \vdash B\;\mathit{true}}
+      \infer2[$\wedge$-I]{\gamma; \delta \vdash A \wedge B\;\mathit{true}}
     }
-    {\gamma; \delta \vdash A \wedge B\;\mathit{true}}
     \and
-    \inferrule*[right=$\wedge$-E$_i$]
-    {\gamma; \delta \vdash A_0 \wedge A_1\;\mathit{true}}
-    {\gamma; \delta \vdash A_i\;\mathit{true}}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A_0 \wedge A_1\;\mathit{true}}
+      \infer1[$\wedge$-E$_i$]{\gamma; \delta \vdash A_i\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\bot$-E]
-    {\gamma; \delta \vdash \bot\;\mathit{true}}
-    {\gamma; \delta \vdash A\;\mathit{true}}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash \bot\;\mathit{true}}
+      \infer1[$\bot$-E]{\gamma; \delta \vdash A\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\vee$-I$_i$]
-    {\gamma; \delta \vdash A_i\;\mathit{true}}
-    {\gamma; \delta \vdash A_0 \vee A_1\;\mathit{true}}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A_i\;\mathit{true}}
+      \infer1[$\vee$-I$_i$]{\gamma; \delta \vdash A_0 \vee A_1\;\mathit{true}}
+    }
     \and
-    \inferrule*[right=$\vee$-E]
-    {
-      \gamma; \delta \vdash A \vee B\;\mathit{true} \\
-      \gamma; \delta, A \vdash C\;\mathit{true} \\
-      \gamma; \delta, B \vdash C\;\mathit{true}
+    \ebrule{%
+      \hypo{\gamma; \delta \vdash A \vee B\;\mathit{true}}
+      \hypo{\gamma; \delta, A \vdash C\;\mathit{true}}
+      \hypo{\gamma; \delta, B \vdash C\;\mathit{true}}
+      \infer3[$\vee$-E]{\gamma; \delta \vdash C\;\mathit{true}}
     }
-    {\gamma; \delta \vdash C\;\mathit{true}}
   \end{mathpar}

### 2023-06-22 tex/dill.tex
[ more ] thesis writing
[ writing ] redo linearity and modality chapter

--- 2021-12-03 tex/dill.tex
@@
-Dual intuitionistic linear logic is a syntax for intuitionistic linear logic
-introduced by \citep{Barber1997}.
-Its key feature is sequents with \emph{dual contexts} --- splitting assumptions
-into \emph{linear} assumptions and \emph{intuitionistic} assumptions.
-The idea of the intuitionistic assumptions is that they will behave like the
-variables of simply typed $\lambda$-calculus, in contrast to the linear
-assumptions, which behave as in the linear calculus we saw in \autoref{sec:ill}.
+Dual Intuitionistic Linear Logic (DILL) is a syntax for intuitionistic linear
+logic introduced by \citet{Barber1996}.
+Its key feature is splitting assumptions into \emph{linear} assumptions and
+\emph{intuitionistic} assumptions --- sometimes called the \emph{dual context}
+approach.
+Intuitionistic assumptions behave like the variables of simply typed
+$\lambda$-calculus.
+In contrast, the linear assumptions behave as in the linear calculus we saw in
+\cref{sec:ill}.
 For example, an intuitionistic assumption of $A$ in an instance of the
 $\otimes$-introduction rule is automatically copied to both premises.
-In contrast, an assumption of $\oc A$ in the purely linear calculus must first
+This contrasts with an assumption of $\oc A$ in the purely linear calculus,
+which must first
 be contracted into two assumptions, with one going to each premise.
+Compared to the modal system we saw in \cref{sec:modal}, linear assumptions
+correspond to true assumptions, and intuitionistic assumptions correspond to
+valid assumptions.
 
-The high-level purpose of intuitionistic assumptions is to universally
-characterise the connective $\oc$.
-That is to say, the existence of intuitionistic assumptions, differentiated from
-linear assumptions, gives rise to $\oc$ in the same way that the existence of
-multiple (linear) assumptions gives rise to $\otimes$.
-We informally express the latter as ``$\otimes$ is the internalisation of
-context concatenation (comma)'', and we can similarly express the former as
-``$\oc$ is the internalisation of intuitionistic assumptions''.
+The new feature when dealing with linear logic, compared to modal logic, is that
+linear and intuitionistic assumptions satisfy different structural rules.
+In our de Bruijn index style, where every multi-premise rule does the maximal
+contraction and weakening occurs in the leafwardmost possible position,
+this means that all of the rules have to manage the substructurality of linear
+assumptions.
+To help manage linear assumptions, I introduce three pieces of notation.
 
-%\begin{align*}
-%  \begin{proof}
-%    \hypo{\Gamma_0 \vdash \oc A}
-%    \hypo{\Gamma_1, \mathbf{int}~A \vdash C}
-%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
-%  \end{proof}
-%  &
-%  \begin{proof}
-%    \hypo{\Gamma_0 \vdash A \otimes B}
-%    \hypo{\Gamma_1, \mathbf{lin}~A, \mathbf{lin}~B \vdash C}
-%    \infer2{\Gamma_0 \cupdot \Gamma_1 \vdash C}
-%  \end{proof}
-%\end{align*}
+\begin{definition}\label{def:DILL-empty}
+  I write $\Gamma~\mathit{int}$ to state that context $\Gamma$ contains only
+  intuitionistic assumptions, i.e., no linear assumptions.
+\end{definition}
+
+\begin{definition}\label{def:DILL-split}
+  I write $\Gamma + \Delta$ to stand for a context combining $\Gamma$ and
+  $\Delta$.
+  Specifically, the operation is only defined when $\Gamma$ and $\Delta$ contain
+  the same intuitionistic assumptions, and in that case the result contains the
+  same intuitionistic assumptions.
+  The linear assumptions of the result are the disjoint union of the linear
+  assumptions of each of $\Gamma$ and $\Delta$.
+\end{definition}
+
+\begin{definition}\label{def:DILL-var}
+  I write $\Gamma \ni A$, like in \cref{sec:modal}, to stand for the set of
+  variables of type $A$ we can get from $\Gamma$.
+  Specifically, if $\Gamma$ contains an assumption $x : A~\mathit{int}$, then
+  $x$ gives rise to an inhabitant of $\Gamma \ni A$ if and only if
+  $\Gamma~\mathit{int}$; and if $\Gamma$ contains an assumption
+  $x : A~\mathit{lin}$, then $x$ gives rise to an inhabitant of $\Gamma \ni A$
+  if and only if $x$ is the only linear assumption in $\Gamma$.
+  These restrictions essentially encode the fact that we do not allow linear
+  variables to be discarded via the use of another variable.
+\end{definition}
+
+\Citet{Barber1996} uses the notational convenience of dividing linear and
+intuitionistic assumptions into two separate contexts, hence the name of the
+\emph{dual context} approach.
+I choose not to do this so as to draw out the connection with the approach of
+\citet{judgmental}, as well as the approach I will take in later chapters.
+I also think that it is more instructive to treat the context as a single
+unified thing, so that we can clearly see what operations on the context have
+to be preserved by environments (for use in renaming, substitution, and other
+traversals).
+
+I list the complete rules of DILL in \cref{fig:dill}.
+Given \cref{def:DILL-empty,def:DILL-split,def:DILL-var}, the rules for
+multiplicative and additive connectives look not too dissimilar to those in
+\cref{fig:mall}.
+However, their treatment of intuitionistic assumptions (contraction and
+weakening automatically, as in non-substructural calculi) is new.
+Meanwhile, the rules for the $\oc$-modality are the same as those for
+the $\Box$-modality given in \cref{fig:PD}, except for the implicit structural
+rules applying to linear assumptions.
+For example, where the \TirName{$\Box$-I} rule discards true assumptions, the
+\TirName{$\oc$-I} rule cannot discard linear assumptions, but instead requires
+that there are no linear assumptions in the context when the rule is applied, so
+that no linear assumptions can be used in the subproof.
+
+\begin{figure}
+  \begin{align*}
+    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
+              \mid 0 \mid A \oplus B \mid \top \mid A \with B \mid \oc A \\
+    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A~\mathit{lin}
+                             \mid \Gamma, A~\mathit{int} \\
+    \mathcal S &\Coloneqq \Gamma \vdash A~\mathit{lin}
+  \end{align*}
+  \begin{mathpar}
+    \ebrule{%
+      \hypo{\Gamma \ni A}
+      \infer1[Var]{\Gamma \vdash A~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash I~\mathit{lin}}
+      \hypo{\Delta \vdash C~\mathit{lin}}
+      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma~\mathit{int}}
+      \infer1[$I$-I]{\Gamma \vdash I~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A \otimes B~\mathit{lin}}
+      \hypo{\Delta, A~\mathit{lin}, B~\mathit{lin} \vdash C~\mathit{lin}}
+      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A~\mathit{int}}
+      \hypo{\Delta \vdash B~\mathit{int}}
+      \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A \multimap B~\mathit{int}}
+      \hypo{\Delta \vdash A~\mathit{int}}
+      \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{int}}
+      \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash 0~\mathit{lin}}
+      \infer1[$0$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
+    }
+    \and
+    \text{(no \TirName{$0$-I})}
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A \oplus B~\mathit{lin}}
+      \hypo{\Delta, A~\mathit{lin} \vdash C~\mathit{lin}}
+      \hypo{\Delta, B~\mathit{lin} \vdash C~\mathit{lin}}
+      \infer3[$\oplus$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A_i~\mathit{lin}}
+      \infer1[$\oplus$-I$_i$]{\Gamma \vdash A_0 \oplus A_1~\mathit{lin}}
+    }
+    \and
+    \text{(no \TirName{$\top$-E})}
+    \and
+    \ebrule{%
+      \infer0[$\top$-I]{\Gamma \vdash \top~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A_0 \with A_1~\mathit{lin}}
+      \infer1[$\with$-E$_i$]{\Gamma \vdash A_i~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A~\mathit{lin}}
+      \hypo{\Gamma \vdash B~\mathit{lin}}
+      \infer2[$\with$-I]{\Gamma \vdash A \with B~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash \oc A~\mathit{lin}}
+      \hypo{\Delta, A~\mathit{int} \vdash \oc C~\mathit{lin}}
+      \infer2[$\oc$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash A~\mathit{lin}}
+      \hypo{\Gamma~\mathit{int}}
+      \infer2[$\oc$-I]{\Gamma \vdash \oc A~\mathit{lin}}
+    }
+  \end{mathpar}
+  \caption{Dual Intuitionistic Linear Logic}
+  \label{fig:dill}
+\end{figure}
+
+I will not give a direct substitution procedure for DILL like the one I gave for
+the modal system of \cref{sec:modal}.
+Suffice to say, where environments for the modal calculus had to be preserved by
+binding of variables and pruning of all true variables, environments for DILL
+have to be preserved by similar plus the operations of
+\cref{def:DILL-empty,def:DILL-split}.
+These definitions themselves are somewhat technical, and I do not believe that
+delving into the further technicalities of how they interact with environments
+will provide useful enough intuitions to justify including here.
+However, I will revisit the idea of environments for linear calculi in a more
+general semiring-annotated setting in \cref{sec:ren-sub-lr}.
+
+Finally, notice that the rules given in \cref{fig:dill} characterise $\oc$ up to
+logical equivalence.
+The main feature ensuring this property is that each logical rule contains
+exactly one occurrence of a logical connective, meaning that each logical rule
+is only defining that connective in that place.
+For the full characterisation result, we also require local soundness and
+completeness of $\oc$, as discussed by \citet{judgmental} in the modal setting.

### 2023-06-22 tex/ill.tex
[ writing ] redo linearity and modality chapter

@@
-Linear logic is a good starting point for our investigations into usage
-restrictions because it is well known, well understood, and exhibits most of the
-difficulties we will come across.
-Linear logic's lack of general weakening and contraction will be a typical
-feature of the calculi we will see, and the exponential modality $\oc$ (bang)
-will be the prototype for a range of modalities I will consider.
+Where modal logics can be seen as additions to an underlying classical or
+intuitionistic logic, linear logic is a much more radical change.
+Formally, many modal logics, including (I)S4, are conservative extensions of
+the underlying classical or intuitionistic logic, meaning that statements not
+mentioning the modal operators $\Box$ and $\Diamond$ are provable in modal logic
+if and only if they are provable in the underlying non-modal logic.
+In linear logic, however, we a priori severely restrict provability of basic
+formulae, and then use modalities to recover the strength of classical or
+intuitionistic logic.
+
+In this thesis, I consider intuitionistic linear logic (ILL).
+ILL can be understood as a logic of resources.
+Whereas in classical logic we read a proposition $A$ to implicitly mean ``$A$ is
+true'', we may read a proposition $A$ of linear logic to implicitly mean ``I
+have an $A$''.
+A sequent $A \vdash B$ can then be understood as saying ``if I have an $A$, I
+can give it up so as to have a $B$'', and the corresponding implication
+$A \multimap B$ can be understood as saying ``I have a \emph{mechanism} for
+having a $B$ if I have and am willing to give up an $A$''.
+\todo{Camels example}
+In this section, I will carefully introduce intuitionistic linear logic so as to
+make the intuitive readings more precise.
+
+%Linear logic is a good starting point for our investigations into usage
+%restrictions because it is well known, well understood, and exhibits most of the
+%difficulties we will come across.
+%Linear logic's lack of general weakening and contraction will be a typical
+%feature of the calculi we will see, and the exponential modality $\oc$ (bang)
+%will be the prototype for a range of modalities I will consider.
 
@@
 
-While classical linear logic is perhaps better understood, for the sake of this
-thesis I focus on intuitionistic linear logic.
-The syntactic difference between the two is analogous to the difference between
-classical and intuitionistic logic: The latter restricts the former by only
-allowing one formula on the right of a sequent.
-Classical linear logic is not irrelevant to this thesis, and appears as an
-instance of the framework presented in \cref{sec:framework}.
-However, the connection of this work to intuitionistic linear logic is more
-direct, and reflects the basis of existing semiring-annotated calculi on
-(intuitionistic) typed $\lambda$-calculi.
-
-Intuitionistic linear logic is often motivated by its sensitivity to resources.
-The fact that, in intuitionistic (and classical) logic, \emph{any} hypothesis
-can be discarded or duplicated means that propositions cannot be used to model
-things that may be ``possessed'' or ``spent'', despite hypotheses often being
-described as things we ``have''.
-In contrast, linear logic allows propositions to behave like physical objects,
-being moved around but not being created or destroyed (a priori).
-
+\todo{Incorporate into chapter introduction.}
 Programming applications of linearity include stateful protocols
@@
 
-%In LJ, two distinct, but equiderivable, canonical definitions of $\wedge$ can
-%be given.
-%
-%In the first approach, $\wedge^-$ is conceived to be a
-%\emph{categorial product}.
-%Such a product is constructed by providing two maps in from the same antecedents
-%$\Gamma$.
-%It is used by projecting out one of the sides.
-%
-%\begin{mathpar}
-%  \inferrule*[right=$\wedge^-$-r]
-%  {\Gamma \vdash A \\ \Gamma \vdash B}
-%  {\Gamma \vdash A \wedge^- B}
-%
-%  \and
-%
-%  \inferrule*[right=$\wedge^-$-l$_0$]
-%  {\Gamma, A \vdash C}
-%  {\Gamma, A \wedge^- B \vdash C}
-%
-%  \and
-%
-%  \inferrule*[right=$\wedge^-$-l$_1$]
-%  {\Gamma, B \vdash C}
-%  {\Gamma, A \wedge^- B \vdash C}
-%\end{mathpar}
-%
-%In the second approach, $\wedge^+$ is conceived to be a \emph{tensor product}.
-%Intuitively, $\wedge^+$ internalises the comma of context concatenation.
-%Such a product is used by giving access to both sides simultaneously.
-%It is constructed by constructing both sides and combining the antecedents
-%required by each side.
-%
-%\begin{mathpar}
-%  \inferrule*[right=$\wedge^+$-r]
-%  {\Gamma \vdash A \\ \Delta \vdash B}
-%  {\Gamma, \Delta \vdash A \wedge^+ B}
-%
-%  \and
-%
-%  \inferrule*[right=$\wedge^+$-l]
-%  {\Gamma, A, B \vdash C}
-%  {\Gamma, A \wedge^+ B \vdash C}
-%\end{mathpar}
-%
-%When we prove the logical equivalence of these two formulations, we notice that
-%the structural rules of weakening and contraction are essential.
-%When we remove weakening and contraction, $\wedge^-$ and $\wedge^+$ become
-%logically distinct connectives, which we notate $\&$ and $\otimes$,
-%respectively.
-%
-%\begin{mathpar}
-%  \inferrule*[right=$\wedge^-$-r]
-%  {%
-%    \inferrule*[right=Weak$^*$,fraction={\cdot\cdots\cdot}]
-%    {\Gamma \vdash A}
-%    {\Gamma, \Delta \vdash A}
-%    \\
-%    \inferrule*[Right=Weak$^*$,fraction={\cdot\cdots\cdot}]
-%    {\Delta \vdash B}
-%    {\Gamma, \Delta \vdash B}
-%  }
-%  {\Gamma, \Delta \vdash A \wedge^- B}
-%
-%  \and
-%
-%  \inferrule*[right=Contr]
-%  {%
-%    \inferrule*[Right=$\wedge^-$-l$_0$]
-%    {%
-%      \inferrule*[Right=$\wedge^-$-l$_1$]
-%      {\Gamma, A, B \vdash C}
-%      {\Gamma, A, A \wedge^- B \vdash C}
-%    }
-%    {\Gamma, A \wedge^- B, A \wedge^- B \vdash C}
-%  }
-%  {\Gamma, A \wedge^- B \vdash C}
-%\end{mathpar}
-%
-%\begin{mathpar}
-%  \inferrule*[right=Contr$^*$,fraction={\cdot\cdots\cdot}]
-%  {%
-%    \mprset{defaultfraction}
-%    \inferrule*[Right=$\wedge^+$-r]
-%    {\Gamma \vdash A \\ \Gamma \vdash B}
-%    {\Gamma, \Gamma \vdash A \wedge^+ B}
-%  }
-%  {\Gamma \vdash A \wedge^+ B}
-%
-%  \and
-%
-%  \inferrule*[right=$\wedge^+$-l]
-%  {%
-%    \inferrule*[Right=Weak]
-%    {\Gamma, A \vdash C}
-%    {\Gamma, A, B \vdash C}
-%  }
-%  {\Gamma, A \wedge^+ B \vdash C}
-%
-%  \and
-%
-%  \inferrule*[right=$\wedge^+$-l]
-%  {%
-%    \inferrule*[Right=Weak]
-%    {\Gamma, B \vdash C}
-%    {\Gamma, A, B \vdash C}
-%  }
-%  {\Gamma, A \wedge^+ B \vdash C}
-%\end{mathpar}
-
 \subsection{The multiplicative-additive fragment}
@@
 where all hypotheses are linear (must be used exactly once).
-We will extend MALL with the \emph{exponential} modality in
+I will extend MALL with the \emph{exponential} modality in
 \cref{sec:bang-modality}.
-MALL is unable to embed intuitionistic or classical logic (collectively
-known as \emph{structural logics}), as MALL is unable to reflect any of the
-discarding or duplication that can be done in proofs in structural logics.
+MALL is unable to embed intuitionistic or classical logic, as MALL is unable to
+reflect any of the discarding or duplication that can be done in proofs using
+weakening or contraction.
 
@@
 However, without the presence of weakening and contraction, we have to be more
-careful about the rules we state, so as not to accidentally introduce weakening
-and contraction admissibly.
+careful about the rules we state, so as not to accidentally admit weakening
+and contraction.
 The lack of these structural rules also allows us to observe a new phenomenon:
 the distinction (at the level of provability) between \emph{additive} and
-\emph{multiplicative} formulations of existing connectives (in particular, in
-the intuitionistic case, the conjunction connective).
+\emph{multiplicative} formulations of existing connectives (in particular, the
+conjunction connective).
 
@@
 gives us the with-product ($\with$).
-These products are equivalent up to provability in structural logics but
-distinct in linear logic.
+These products are equivalent up to provability in logics with weakening and
+contraction, but distinct in linear logic.
 
@@
 
+\begin{figure}
+  \begin{align*}
+    A, B, C &\Coloneqq \dots \mid \oc A
+  \end{align*}
   \begin{mathpar}
@@
   \end{mathpar}
+  \caption{The sequent calculus rules for the $\oc$-modality}
+  \label{fig:bang-seq}
+\end{figure}
 
-In the intuitionistic linear logic sequent calculus ILL, $\oc A$ is defined
+\Cref{fig:bang-seq} shows the rules we can add to MALL to get the full sequent
+calculus for intuitionistic linear logic (ILL).
+In ILL, $\oc A$ is defined
 to be a proposition whose occurrences as antecedents can be deleted
@@
 
-While this definition of $\oc$ works, in the sense that it gives the intended
-class of models and cut elimination is maintained, it has some disadvantages.
-Firstly, while the multiplicative and additive connectives are all characterised
-by universal properties, $\oc$ is not.
+The \TirName{Dereliction} rule is comparable to the \TirName{T} axiom we saw in
+\cref{sec:modal}.
+The \TirName{Promotion} rule is exactly like the \TirName{Promotion} rule we saw
+in \cref{sec:modal}, and for the same reasons we want to avoid having such a
+rule in a natural deduction system for intuitionistic linear logic.
+Note that \TirName{Promotion} is not problematic in the sequent calculus, at
+least insofar as it maintains \TirName{Cut}-elimination.
+
+Additional to the problems with \TirName{Promotion} in a natural deduction
+system, defining the $\oc$-modality as in \cref{fig:bang-seq} has the odd
+feature that, unlike all of the other connectives of ILL, $\oc$ is not
+characterised by a universal property.
 This can be seen by the fact that taking the rules for $\oc$ and replacing each
@@
 $\oc X$.
-Finally, while $\oc$ is supposed to be a positive connective, it sometimes
-behaves like a negative connective.
-For example, for a positive connective like $\otimes$, the normal form proof
-of the identity sequent $P \otimes Q \vdash P \otimes Q$ starts (from the
-bottom) with a left rule and then, with the left in a more amenable form,
-applies the right rule.
-In contrast, the normal form proof of $\oc P \vdash \oc P$ starts with the
-right rule, as it needs to have everything on the left be of the form $\oc X$.
+This lack of characterisation also holds of the rule and axioms we first gave to
+the $\Box$-modality in \cref{fig:S4-axioms}, and essentially any presentation
+that does not encorporate the modality into the judgemental structure of the
+calculus.
 
+A noteworthy solution to the substitution problem of \TirName{Promotion} is
+given by \citet{BBdePH93}.
+The solution they give is compatible with the form of sequents introduced above,
+but works purely in terms of right-rules like standard natural deduction
+calculi.
+I give their rules for the $\oc$-modality in \cref{fig:BBdePH}.
+The \TirName{Weakening} and \TirName{Contraction} rules are reminiscent of
+pattern-matching elimination principles, allowing us to form an inhabitant of
+the type of interest ($\oc A$) and then continue producing the originally
+desired result ($B$) in an updated context.
+However, somewhat unusually, we can choose which of the two ``pattern-matching''
+principles to use, depending on whether we want to delete or duplicate the
+value of type $\oc A$.
+Another elimination-like rule is \TirName{Dereliction}, which straightforwardly
+lets us derive $A$ from $\oc A$ similarly to the corresponding sequent calculus
+rule from \cref{fig:bang-seq}, but on the right.
+
+\begin{figure}
   \begin{mathpar}
     \ebrule{%
-    \infer0[Id]{P \vdash P}
-    \infer0[Id]{Q \vdash Q}
-    \infer2[$\otimes$-r]{P, Q \vdash P \otimes Q}
-    \infer1[$\otimes$-l]{P \otimes Q \vdash P \otimes Q}
+      \hypo{\Delta_1 \vdash \oc A_1}
+      \hypo{\cdots}
+      \hypo{\Delta_n \vdash \oc A_n}
+      \hypo{\oc A_1, \ldots, \oc A_n \vdash B}
+      \infer4[Promotion]{\Delta_1, \ldots, \Delta_n \vdash \oc B}
     }
-
     \and
-
     \ebrule{%
-    \infer0[Id]{P \vdash P}
-    \infer1[Der]{\oc P \vdash P}
-    \infer1[Pro]{\oc P \vdash \oc P}
+      \hypo{\Gamma \vdash \oc A}
+      \infer1[Dereliction]{\Gamma \vdash A}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash \oc A}
+      \hypo{\Delta \vdash B}
+      \infer2[Weakening]{\Gamma, \Delta \vdash B}
+    }
+    \and
+    \ebrule{%
+      \hypo{\Gamma \vdash \oc A}
+      \hypo{\Delta, \oc A, \oc A \vdash B}
+      \infer2[Contraction]{\Gamma, \Delta \vdash B}
     }
   \end{mathpar}
+  \caption{The \citet{BBdePH93} rules for the $\oc$-modality}
+  \label{fig:BBdePH}
+\end{figure}
+
+The \citeauthor{BBdePH93} \TirName{Promotion} rule is more formally complex, but
+can be understood as follows.
+We notice that the problem with the \TirName{Promotion} rule of
+\TirName{fig:bang-seq} when added to a natural deduction system is that it
+restricts the types of assumptions in the concluding sequent.
+In other words, the sequent calculus \TirName{Promotion} rule is acting like
+both a right-rule and, problematically, a left-rule, to some degree.
+Acting on the context in a natural deduction system risks making substitution
+inadmissible, as happens in this case.
+Instead, \citeauthor{BBdePH93} paraphrase this restriction on the context as the
+construction of a \emph{new} context $\oc A_1, \ldots, \oc A_n$ for the primary
+subderivation.
+Heuristically, this new \TirName{Promotion} rule admits substitution because any
+substitution will pass into the relevant premises deriving $\oc A_1$ to
+$\oc A_n$.
+More abstractly, we can notice that the first $n$ premises essentially form an
+explicit substitution from $\Delta_1, \ldots, \Delta_n$ to
+$\oc A_1, \ldots, \oc A_n$, so any further substitution applied
+to a \TirName{Promotion}-headed term is precomposed onto this explicit
+substitution~\citep{ACCL91}.
+
+The \citeauthor{BBdePH93} \TirName{Promotion} rule is a clever solution to the
+substitution problem of (intuitionistic) linear logic, but the resulting system
+comes with its own problems.
+If we were to write programs in a language directly implementing the
+\citeauthor{BBdePH93} calculus, then the \TirName{Promotion} rule would be a
+pain point because it makes us rebind all of the $oc$-typed variables we need to
+new variables.
+Additionally, we have to be explicit everywhere about weakening and contraction,
+potentially making working with $\oc$-typed variables much more fiddly than the
+corresponding variables would be in a non-substructural programming language.
+These syntactic annoyances may be worked around in a realistic implementation by
+some new elaboration procedures, but we would prefer not to rely on such
+elaboration if there is an acceptable core calculus which more closely matches
+the programs we want to write.
+Therefore, I investigate such a system, Dual Intuitionistic Linear Logic, in the
+following subsection.
+
+\subsection{Dual Intuitionistic Linear Logic}\label{sec:dill}
+\input{dill.tex}

### 2023-06-22 tex/linear.tex
[ writing ] redo linearity and modality chapter

@@
 \chapter{Linearity and modality}\label{sec:linearity}
-
-\begin{itemize}
-  \item Motivation via prior work:
-    \begin{itemize}
-      \item Rust
-      \item Session types
-      \item ATS
-      \item Complexity \citep{Hofmann03}
-      \item Modal\ldots
-    \end{itemize}
-  \item Can \citet{AACMM21} capture linearity/modality?
-    It's unclear based on standard presentations.
-    Let's find one that looks good (DILL/PD), and see whether it works.
-  \item In-depth introduction of DILL and PD
-  \item It's close to working, but doesn't.
-  \item Mention \citet{BBdePH93}, problems with substitution.
-\end{itemize}
 
@@
 Additionally, forms of linearity have been used in theoretical work, for example
-to bound computational complexity~\citep{GSS92,Hofmann03} or maintain
-differentiability~\citep{ER03,Ehrhard18}.
+to bound computational complexity~\citep{GSS92,Hofmann03} or to write programs
+which support incremental updates to computations~\citep{ER03,Ehrhard18}.
 As for modality, \citet{IMO20} show that common data-flow analyses for
@@
 More user-facing implementations of modal type systems include the system for
-stack allocation in OCaml~\citep{DW22} and the approach to erasure found in
-Agda and Idris\todo{cite}.
+stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
+erasure found in Agda and Idris~\citep{Atkey18}.
 
-In this chapter\ldots\todo{Chapter map}
+In this chapter, I look at two standard type-theoretic features that motivate
+considering calculi going beyond simple type theories as deliniated in
+\cref{sec:simple} or equivalently by \citet{AACMM21}.
+The first feature is a $\Box$-modality --- specifically, the $\Box$-modality of
+intuitionistic S4, which I discuss in \cref{sec:modal}.
+The second feature is linearity, which I discuss in \cref{sec:ill}.
+In particular, I discuss possible syntaxes for each, with a mind to
+being able to apply the techniques described in \cref{sec:simple}.
+Then, I\ldots
 
@@
 
-\section{Intuitionistic linear logic}\label{sec:ill}
+\section{Intuitionistic Linear Logic}\label{sec:ill}
 \input{ill.tex}
 
-\section{DILL}\label{sec:dill}
-\input{dill.tex}
-
-\section{Mechanisation survey}
+\section{Mechanisations and systemitisations of substructural logics}
+\label{sec:linmech}
 \input{linmech.tex}

### 2023-06-22 tex/linmech.tex
[ writing ] more
[ writing ] redo linearity and modality chapter

@@
 In this section, I give an overview of techniques which have been used in
-previous work to encode linear logic in proof assistants.
-To my knowledge, there have been no other significant attempts to mechanise
-semiring-annotated logics.
+previous work to mechanise linear logic in proof assistants.
+Na\"{i}ve approaches often struggle to represent concatenation of contexts in a
+way which is amenable to the way dependent type theory-based proof assistants
+work.
+Problems even arise when working rigorously on paper when trying to avoid an
+explicit exchange rule, such as how \cref{def:DILL-split} is not a precise
+definition of a binary operator on lists.
 
@@
 
-Typing with leftovers~\cite{allais17} is a technique developed to capture a
-straightforward approach to linearity checking.
-The idea is to, at any point, consider an input context, a term, and an output
-context.
-The input context contains all of the variables in scope, while the output
+Typing with leftovers, introduced by \citet{allais17}, is a technique developed
+to specify an algorithm for linear type checking as a declarative type system.
+The idea is to consider an input context, a term, and an output context, where
+the input context contains all of the variables in scope, and the output
 context is the same minus any variables used by the term.
-Checking of adjacent terms, for example in a $\otimes$-introduction, is done by
+Type-checking of adjacent subterms of, for example, an application of
+the $\otimes$-introduction rule, is done by
 threading the context through from the output of the first term to the input of
@@
 are bound, and are expected to be absent in the output context of that term.
-\Cref{fig:twl,fig:twl-mult} give rules in this style for the multiplicative
-fragment of intuitionistic linear logic.
+
+\Cref{fig:twl,fig:twl-mult} give rules in the typing-with-leftovers style for
+the multiplicative fragment of intuitionistic linear logic.
+Where \citeauthor{allais17} marks \emph{fresh} and \emph{stale} variables, I use
+the notation I will use starting in \cref{sec:semirings}, labelling such
+variables with $\gr 1$ and $\gr 0$, respectively.
+Intuitively, the number describes how many more times that variable is to be
+used.
 
@@
 However, it is less clear how to handle nullary additives --- $\top$ and $0$ ---
-as we would have 0 (rather than 2) candidates for the output context.
+as we would have 0 (rather than 2) potential candidates for the output context.
 At some level, this problem is unavoidable in a system modelling linearity
 checking because any checking strategy will expose the ambiguity in sequents
-like $\gr1x : A \vdash (\{\}, \{\}) : \top \otimes \top$ of whether the variable
-$x$ was used in the left half or the right half.
-As such, we may expect uses of $\top$-introduction (and similarly
-$0$-elimination) to be annotated with which variables they use, in which case we
-get a viable typing with leftovers rule.
+like $\gr1x : A \vdash (\langle\rangle, \langle\rangle) : \top \otimes \top$ of
+whether the variable
+$x$ was consumed in the left half or the right half.
+Such an example is also considered in related work on proof seach for linear
+logics, such as the work of \citet[p.\ 11]{WH94} and \citet[p.\ 150]{CHP00}.
+It is not immediately clear whether the different solutions proposed by these
+papers will apply to \citeauthor{allais17}' and my settings, given that they
+both act on a set of formulas restricted to facilitate proof search.
+The solutions also add significant, seemingly somewhat ad hoc, structure to the
+syntax of sequents, with no semantic justification (rather being justified by
+making their respective proof search algorithms efficient).
+%As such, we may expect uses of $\top$-introduction (and similarly
+%$0$-elimination) to be annotated with which variables they consume, in which case we
+%get a viable typing with leftovers rule.
 
@@
 $\oc$.
-We could perhaps, inspired by DILL, distinguish between linear variables and
-intuitionistic variables.
+The solution given by both \citet{WH94} and \citet{CHP00} is, as in DILL, to
+distinguish between linear variables and intuitionistic variables.
 This gives rules like those of \cref{fig:twl-exp}.
@@
 $\gr\omega$) must be intuitionistic in the output context.
-In the $\oc$-introduction rule, there are several choices we could make, but I
-have chosen to keep all the linear variables in scope but used so as to match
-the general style of variables staying in scope with the $\gr0$ annotation..
+%In the $\oc$-introduction rule, there are several choices we could make, but I
+%have chosen to keep all the linear variables in scope but used so as to match
+%the general style of variables staying in scope with the $\gr0$ annotation.
 
@@
 
-The unusal form of sequents also somewhat obscures the denotational semantics of
-a typing with leftovers system.
+The unusual form of sequents also somewhat obscures any attempt to interpret the
+terms of a typing-with-leftovers system.
 Though the $\boxtimes$ notation suggests a semantics into symmetric monoidal
@@
 the output context.
+
+Another piece of related work using a typing-with-leftovers style is that of
+\citet{polakow15}.
+There, \citeauthor{polakow15} encodes a linear embedded domain-specific language
+inside Haskell using typeclass constraints.
 
@@
 with an explicit exchange rule being used to move such formulas into place.
-\Citet{laurent18} explains why explicit permutations are necessary in such a
-system, and why using multisets instead of lists would conflate distinct proofs
-when there are repeated formulas in a sequent.
-The system I will present in the next chapter avoids permutations essentially by
-being a refinemet of the system presented in the previous chapter, which has an
-ordered context and specific references (de Bruijn indices) into it.
+\Citet{laurent18} points out that using multisets, as do some less formal
+presentations of linear logic, is insufficient at distinguishing certain proofs
+involving repeated assumptions or conclusions.
+For example, we expect there to be two distinct derivations of
+$A \otimes A \vdash A \otimes A$ (up to the appropriate equational theory):
+the one that keeps the pair in the same order and the one that flips the order.
+But if the \TirName{$\otimes$-L} rule unpacks the formula $A \otimes A$ into the
+\emph{multiset} $\{A, A\}$, then we forget the order of the input pair.
 
@@
 
-\begin{itemize}
-  \item Sequent calculus
-  \item Specialised to linear logic
-\end{itemize}
+The relevance of Yalla to the work in this thesis is limited by the fact that
+Yalla is based entirely on sequent calculi, whereas I am considering only
+natural deduction calculi.
+
+\subsection{Co-de-Bruijn syntax}
+
+\Citet{McBride18} presents a mechanisation of the simply typed
+$\lambda$-calculus in what he calls \emph{co-de-Bruijn} style.
+He notes that syntax based on de Bruijn indices, as presented in
+\cref{sec:simple}, finds a canonical way to place contractions and weakenings
+by eagerly placing contractions wherever they could be needed (i.e., whenever a
+rule has multiple premises) and leaving weakening as late as possible (i.e., at
+the variable rule and at rules with no premises).
+Co-de-Bruijn syntax, by contrast, finds a canonical way to place contractions
+and weakenings by doing the reverse: weakening as early as possible (i.e., as
+soon as the variable is bound) and contracting only where necessary.
+
+Such a scheme straightforwardly adapts to multiplicative linear logic by
+modifying the data structures presented by \citet{McBride18} to disallow all
+contraction and weakening.
+With the additive rules, however, we get cases where variables appear multiple
+times syntactically in a term but are still considered linear by the type
+system, the simplest example being
+$x : A \vdash \langle x, x\rangle : A \with A$.
+Such rules are perhaps a new kind compared to what \citeauthor{McBride18}
+considers, but it seems likely that just copying the same context to all the
+premises would not be too hard to accommodate.
+Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
+intuitionistic variables handled using the regular co-de-Bruijn machinery from
+the paper.
+In summary, the co-de-Bruijn approach is promising for capturing linearity, but
+has not been thoroughly investigated.
+
+\subsection{Fitch-style modalities}
+
+An alternative to the approach of \citet{judgmental} to adapt modal logics (and
+particularly IS4) to natural deduction is using Fitch-style calculi.
+Fitch-style calculi, as codified and studied by \citet{Borghuis-thesis}, are
+distinguished by allowing for contexts containing \emph{locks}, written $\Lock$,
+with the variable rule being restricted so that only variables not behind locks
+are immediately accessible.
+
+Below I give the main rules of Fitch-style IK (intuitionistic logic K).
+Other normal modal logics are obtained by strengthening the $\Box$-elimination
+rule to remove varying numbers of locks.
+For example, we can add axiom \TirName{T} by allowing for 0 or 1 locks to be
+removed, axiom \TirName{4} by allowing 1 or more locks to be removed, or both
+axioms together by allowing any number of locks to be removed from the
+right-hand end of the context (together with any variables to the right of a
+removed lock).
+The \TirName{$\Box$-I} rule stays the same, and forms part of an adjunction of
+the form ``$\Lock \dashv \Box$''.
+
+\begin{mathpar}
+  \ebrule{%
+    \hypo{\Lock \notin \Gamma'}
+    \infer1[Var]{\Gamma, A, \Gamma' \vdash A}
+  }
+  \and
+  \ebrule{%
+    \hypo{\Gamma, \Lock \vdash A}
+    \infer1[$\Box$-I]{\Gamma \vdash \Box A}
+  }
+  \and
+  \ebrule{%
+    \hypo{\Lock \notin \Gamma'}
+    \hypo{\Gamma \vdash \Box A}
+    \infer2[$\Box$-E$_K$]{\Gamma, \Lock, \Gamma' \vdash A}
+  }
+\end{mathpar}
+
+A syntactic advantage of Fitch-style calculi over the calculus introduced by
+\citet{judgmental} is that Fitch-style calculi support a projection-style
+eliminator for $\Box$, which makes it easier to use than the pattern-matching
+eliminator of \citeauthor{judgmental}.
+A disadvantage is that $\Lock$ cannot be understood as a kind of hypothetical
+judgement like the rest of the context, so many of the heuristics we relied on
+in \cref{sec:simple} and \cref{sec:modal} fail.
+In fact, the addition of locks represents a large change to the judgemental
+structure of the calculus, apparently requiring a complete overhaul of the basic
+metatheory.
+
+\Citet{VRC22} have completed a significant mechanisation of the metatheory of
+a Fitch-style calculus in Agda.
+This work shows that, despite the change in the structure of the metatheory,
+Fitch-style calculi are amenable to mechanised proofs.
+
+I am not aware of any work discussing linear Fitch-style calculi.
+
+\subsection{Systematisations of substructural logics}
+
+Several pieces of prior work have aimed to give general accounts of a range of
+substructural calculi, in a similar vein to existing accounts of aspects of
+non-substructural calculi.
+I review some of these, particularly as a way to give a comparison to the
+adaptation of the methods of \cref{sec:simple} that I spend the rest of this
+thesis on.
+
+\paragraph{Linear Logical Framework}
+In \cref{sec:lf}, I discussed logical frameworks based on the
+$\lambda^\Pi$-calculus, and their use in the mechanisation of non-substructural
+programming languages.
+\Citet{CP02} extend $\lambda^\Pi$ to a calculus
+$\lambda^{\Pi\multimap\with\top}$, and use that to create a logical framework
+supporting linear logics: the Linear Logical Framework (LLF).
+The $\Pi$ type former still forms intuitionistic weak dependent function spaces,
+while the new $\multimap$ forms linear weak non-dependent function spaces.
+They handle the distinction between intuitionistic and linear variables thus
+introduced in the same style as DILL, with the argument of an intuitionistic
+application having the same intuitionistic restriction as we saw in DILL's
+\TirName{$\oc$-I} rule.
+
+The additive connectives $\with$ and $\top$ provide a way to state rules whose
+premises respectively share and arbitrarily consume linear variables, like in
+the rules for additive connectives in linear logic.
+I will revisit this method of stating typing rules in terms of \emph{sharing}
+and \emph{separating} (as given by right-nested sequences of $\multimap$s)
+conjunctions of premises in \cref{sec:lnd}, though in \cref{sec:bunched-logic}
+I argue that there is a closer connection to bunched logics than linear logics.
+
+The main focus of the original paper is to represent mutation in an ML-like
+language via state updates mediated by $\multimap$, though they also mention
+having mechanised some metatheory of linear calculi.
+Some example programs are currently available at
+\url{https://www.cs.cmu.edu/~iliano/projects/LLF/index.html}.
+
+\paragraph{Encoding linearity in LF}
+\Citet{crary10} gives a method of encoding linearity constraints in a
+conventional, non-substructural, logical framework.
+He implements this approach in the LF-based proof assistant Twelf~\citep{Twelf}.
+He uses a predicate \texttt{linear : (term -> term) -> type}, where
+\texttt{term} is a type of preterms, and thus \texttt{term -> term} is (thanks
+to the weak function space of LF) the type of preterms with one free variable.
+The predicate \texttt{linear} then says that that free variable is used linearly
+in its term, which is defined inductively on the structure of preterms.
+The \texttt{linear} predicate is used by the typing relation wherever variables
+are bound.
+The development handles all of intuitionistic linear logic, with the
+$\oc$-modality treated with a DILL-style distinction between linear and
+intuitionistic variables.
+Intuitionistic variables are implemented simply by not checking for linearity in
+the \TirName{$\oc$-E} rule.
+\Citet[\S 4]{crary10} also shows how to adapt this technique to a PD-style
+presentation of IS4.
+
+As an example, let us look at the typing and linearity rules for the binary
+tensor product.
+Typing is given by a relation \texttt{of : term -> tp -> type}, where
+\texttt{tp} is the type of object-level types.
+Each syntactic form has a typing rule and potentially several linearity rules,
+understood disjunctively as logic programming clauses.
+The rules for the introduction form are listed below.
+The typing relation is just like it would be for pairs in the simply typed
+$\lambda$-calculus: $(M, N) : A \otimes B$ if $M : A$ and $N : B$.
+The linearity rules are symmetric, so I will just consider
+\texttt{linear/tpair1}.
+It says that $x \vdash (M[x], N)$ is linear if $x \vdash M[x]$ is linear.
+This rule is subtle in that not applying \texttt{x} to \texttt{N} implies that
+\texttt{x} is fresh (and therefore unused) in \texttt{N}.
+Where $\otimes$-pairs have two linearity rules, the $I$-unit, and also the
+introduction form for $\oc$, have no linearity rules, meaning that no linear
+variables can be used in or by them.
+
+\begin{verbatim}
+of/tpair : of (tpair M N) (tensor A B) <- of M A <- of N B.
+linear/tpair1 : linear ([x] tpair (M x) N) <- linear ([x] M x).
+linear/tpair2 : linear ([x] tpair M (N x)) <- linear ([x] N x).
+\end{verbatim}
+
+Meanwhile, the rules for the eliminator are somewhat more involved, thanks to
+the bound variables.
+First, the typing rule shows how \texttt{of} relies on \texttt{linear}, checking
+each bound variable for linearity.
+Because we have two bound variables, we need to check that the term \texttt{N}
+is linear in both.
+We do this by checking that, for all \texttt{y}, \texttt{N} is linear in
+\texttt{x}, and that for all \texttt{x}, \texttt{N} is linear in \texttt{y}.
+The linearity rules have the same choice and careful management of free
+variables as they did for the introduction form.
+In addition, in the rule \texttt{linear/lett2}, the bound variables in
+\texttt{N} have to be universally quantified while we check for linearity in the
+free variable \texttt{z}.
+
+\begin{verbatim}
+of/lett : of (lett M ([x] [y] N x y)) C
+  <- of M (tensor A B) <- ({x} of x A -> {y} of y B -> of (N x y) C)
+  <- ({y} linear ([x] N x y)) <- ({x} linear ([y] N x y))
+linear/lett1 : linear ([z] lett (M z) ([x] [y] N x y))
+  <- linear ([z] M z)
+linear/lett2 : linear ([z] lett M ([x] [y] N z x y))
+  <- ({x} {y} linear ([z] N z x y))
+\end{verbatim}
+
+\Citet{crary10} goes on to extend this encoding with intuitionistic dependent
+$\Pi$-types, producing an object theory comparable to the
+$\lambda^{\Pi\multimap\with\top}$ metatheory developed by \citet{CP02}.
+If one wants to mix linearity and dependency following the methodology of
+\citet{Atkey18}, then it is crucial that linear variables are still free in
+subterms from which they have been discarded.
+At first sight, it appears that \citeauthor{crary10}'s encoding violates this
+because of its use of ``does not appear free'' to mean ``is not used'' in many
+linearity rules.
+However, one could imagine introducing an \texttt{unused} predicate similar to
+\texttt{linear} in order to handle unused free variables, at the cost of a few
+extra rules and a somewhat heavier encoding (scaling with the \emph{size} of the
+term, rather than the depth).
+Indeed, one could imagine parametrising the \texttt{linear} predicate so as to
+encode the semiring-annotated systems I discuss in \cref{sec:semirings}.
+
+The approach of \citet{crary10} removes the objection to the work of
+\citet{CP02} that each new substructural discipline would need a new proof
+assistant by encoding linearity in a standard intuitionistic logical framework.
+However, the encoding makes linear variables second-class compared to
+intuitionistic variables.
+While intuitionistic variables are just there thanks to the metatheory, linear
+variables must essentially be explicitly quantified.
+
+\paragraph{Licata-Shulman-Riley}
+\Citet{LicataSR17} describe a framework for specifying and working with a wide
+range of substructural logics.
+I discuss exactly what this range is in \cref{sec:semirings-conc}, in relation
+to the calculi I describe in the rest of this thesis.
+For now, it suffices to say that this framework is specified in enough detail
+that it should be possible to mechanise it directly, but I am not aware of
+anyone having done so.
+\Citet{Restall1999} describes a similar approach.
+
+\paragraph{Tanaka-Power}
+The work of \citet{FPT99}, which I discussed in \cref{sec:fiore}, has been
+extended to substructural syntaxes by \citet{TP06}.
+This work gives a mechanism for turning a description of contexts and their
+structural rules (expressed as a pseudo-monad on the 2-category of categories)
+into a framework for defining substructural syntaxes, and more generally a
+category of algebras of which the syntax is the initial object.
+As examples, they give the untyped $\lambda$-calculus, an untyped multiplicative
+linear logic, and a bunched logic.
+These examples show a broad range of substructural disciplines they support ---
+comparable to the work of \citet{LicataSR17} (which I discuss in
+\cref{sec:semirings-conc}), and more than I discuss in this thesis.
+However, they also show two of the limitations of their approach.
+Firstly, this work provides no way to track types, though it should be possible
+to incorporate types at the expense of complicating the categorical
+constructions they use.
+Secondly, it appears to be impossible to encode the syntactic forms used for the
+additive connectives (i.e., the Cartesian product and coproduct) of linear
+logic.
+Essentially, subterms can only be combined into a larger term in the same ways
+as how contexts can be appended together.
+For example, in a bunched logic, contexts can be combined through both sharing
+conjunction (the Cartesian product) and separating conjunction
+(a monoidal product).
+Correspondingly, the syntax descriptions of \citeauthor{TP06} allow for the
+syntax of sharing pairs and separating pairs.
+However, in the case of linear logic, contexts can only be combined via a
+monoidal product, so we only get separating pairs (tensor-products) and not
+sharing pairs (with-products).
+
+\paragraph{Granule}
+\citet{BrunelGMZ14}, \citet{GhicaS14}, \ldots

### 2023-06-22 tex/modal.tex
[ writing ] redo linearity and modality chapter

@@
-Modal logics, and in particular their modalities, are traditionally presented in
-an axiomatic style.
+Modal logics, and in particular their modalities, are usually presented in
+philosophical and mathematical logic in an axiomatic style.
 For example, the common modal logic S4 may be presented by taking a presentation
@@
 Note that, unlike elsewhere in this thesis, the empty context in the
-necessitation rule \TirName{N} is a true restriction.
-While the axioms hold in any context, thanks to weakening, necessitation only
-holds in the empty context.
+necessitation rule \TirName{N} is a proper restriction:
+While the axioms hold in any context, thanks to weakening, and ordinary logical
+rules hold in any context, the necessitation rule only holds in the empty
+context.
 We may apply weakening to the conclusion of \TirName{N} to embed such a
@@
 
-In this thesis, I am working intuitionistically, so I start with a base of
-intuitionistic logic, and study intuitionistic S4 (IS4).
+In this thesis, I am working with intuitionistic logics, so I start with a base
+of intuitionistic logic, and study intuitionistic S4 (IS4).
+The axiomatic presentation can be introduced on top of various presentations of
+intuitionistic/classical logic --- including natural deduction, sequent
+calculi, Hilbert systems, and abstractly taking the set of
+intuitionistic/classical tautologies.
+For concreteness and rigour, I will assume a natural deduction system NJ
+allowing for explicit weakening.
 
-The traditional presentation is convenient for talking about semantics, with
+The axiomatic presentation is convenient for talking about semantics, with
 different choices of axioms corresponding fairly directly to natural
@@
 which makes it incompatible with the methods discussed in \cref{sec:simple}.
-Essentially, it requires necessitation to be a special case in renaming,
+Essentially, we are required to make necessitation a special case in renaming,
 substitution, and all the other traversals.
-Additionally, axioms require us to modify the notion of canonical form, and redo
+Additionally, axioms require us to modify the notion of canonical
+form~\citep[p.\ 79]{Prawitz65}, and redo
 any proofs that rely on all closed terms of function type being
@@
 capturing the desired modality in the judgemental structure of the syntax.
-To start, note the following.
+To start, note the following two facts about a ``promotion'' rule similar to one
+mentioned by \citet{BBdPH93}.
 
-\begin{proposition}
-  In NJ + \TirName{N} + \TirName{K} + \TirName{T}, the following principle is
+\begin{proposition}\label{thm:NKT4-P}
+  In NJ + \TirName{N} + \TirName{K} + \TirName{4}, the following principle is
   admissible.
@@
       \hypo{\Box B_1, \ldots, \Box B_n \vdash A}
-      \infer1[NKT]{\Box B_1, \ldots, \Box B_n \vdash \Box A}
+      \infer1[Promotion]{\Box B_1, \ldots, \Box B_n \vdash \Box A}
     }
@@
   By induction on $n$.
-  When $n = 0$, \TirName{NKT} becomes just the necessitation rule \TirName{N}.
+  When $n = 0$, \TirName{Promotion} becomes just the necessitation rule \TirName{N}.
   For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
   $\Box\Gamma$ abbrevates
-  $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{T$'$} and
+  $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{4$'$} and
   \TirName{K$'$} abbrevate the obvious combinations of
-  \TirName{T} and \TirName{K}, respectively, and \TirName{$\to$-E}.
+  \TirName{4} and \TirName{K}, respectively, and \TirName{$\to$-E}.
   I silently apply weakening thoughout the derivation for the sake of concision.
@@
       \infer0[Var]{\Box B \vdash \Box B}
-      \infer1[T$'$]{\Box B \vdash \Box\Box B}
+      \infer1[4$'$]{\Box B \vdash \Box\Box B}
       \infer2[K$'$]{\Box\Gamma, \Box B \vdash \Box A}
@@
 
+\begin{proposition}\label{thm:PT-NKT4}
+  In NJ + \TirName{Promotion} + \TirName{T}, we can derive the original axioms
+  and rule of S4: \TirName{N}, \TirName{K}, \TirName{T}, and \TirName{4}
+\end{proposition}
+\begin{proof}
+  The rule \TirName{N} and axiom \TirName{T} are immediate.
+  For axioms \TirName{K} and \TirName{4}, we have the following derivations,
+  where \TirName{T$'$} abbrevates the obvious combination of \TirName{T} and
+  \TirName{$\to$-E}.
+  \begin{mathpar}
+    \ebrule{%
+      \infer0[T$'$]{\Box\plr{A \to B} \vdash A \to B}
+      \infer0[T$'$]{\Box A \vdash A}
+      \infer2[$\to$-E]{\Box\plr{A \to B}, \Box A \vdash B}
+      \infer1[Promotion]{\Box\plr{A \to B}, \Box A \vdash \Box B}
+      \infer1[$\to$-I]{\Box\plr{A \to B} \vdash \Box A \to \Box B}
+      \infer1[$\to$-I]{\vdash \Box\plr{A \to B} \to \Box A \to \Box B}
+    }
+    \and
+    \ebrule{%
+      \infer0[Var]{\Box A \vdash \Box A}
+      \infer1[Promotion]{\Box A \vdash \Box\Box A}
+      \infer1[$\to$-I]{\vdash \Box A \to \Box\Box A}
+    }
+  \end{mathpar}
+\end{proof}
+
+\Cref{thm:NKT4-P,thm:PT-NKT4} suggest that a \TirName{Promotion}-like rule to
+introduce the $\Box$ operator and a \TirName{T}-like rule to eliminate the
+$\Box$ operator may be a possible approach to capture S4 in a natural deduction
+style.
 However, we do not want to add such a rule to our system because it is not
 stable under substitution.
-For example, consider a substitution
-$\sigma : \Box A \wedge \Box B \env\vdash \Box A, \Box B$.
-The domain context contains a single formula headed by $\wedge$, not $\Box$, so
-when we apply $\sigma$ to a derivation ending with an application of
-\TirName{NKT} deriving $\Box A, \Box B \vdash \Box C$, it becomes unclear how to
-conclude $\Box A \wedge \Box B \vdash \Box C$, as \TirName{NKT} does not apply.
+For example, let
+$\sigma : \Box A \wedge \Box B \env\vdash \Box A, \Box B$ be the substitution
+whose two components are given by left projection and right projection,
+respectively.
+The substitution principle says that we should be able to turn any derivation of
+$\Box A, \Box B \vdash \Box C$ into a derivation of
+$\Box A \wedge \Box B \vdash \Box C$.
+However, if we concluded $\Box A, \Box B \vdash \Box C$ by \TirName{Promotion},
+then it becomes unclear how to conclude $\Box A \wedge \Box B \vdash \Box C$.
+In particular, \TirName{Promotion} does not apply because the formula
+$\Box A \wedge \Box B$ is headed by $\wedge$, rather than $\Box$.
 
@@
 differently --- restricting substitution so that it is compatible with an
-adapted version of \TirName{NKT}.
+adapted version of \TirName{Promotion}.
 
@@
 are preserved by the \TirName{$\Box$-I} rule, which is our recasting of
-\TirName{NKT} as the introduction rule for $\Box$.
+\TirName{Promotion} as the introduction rule for $\Box$.
 Otherwise, like true hypotheses, an $A~\mathit{valid}$ hypothesis can be used to

### 2023-06-22 tex/prelim.tex
[ writing ] move semiring definition out of prelims

@@
 \section{Algebra}
-
-\begin{definition}
-  A \emph{semiring} is a monoid in the multicategory of commutative monoids and
-  multilinear maps.
-  Unpacked, this means that we have a set $\mathscr R$ together with elements
-  $0$ and $1$ and binary operators $+$ and $\cdot$ (with $\cdot$ usually written
-  as juxtaposition) such that the following hold for all
-  $x, y, z \in \mathscr R$.
-  \begin{itemize}
-    \item $0 + x = x$; $x + 0 = x$; $(x + y) + z = x + (y + z)$; $x + y = y + x$
-    \item $1x = x$; $x1 = x$; $(xy)z = x(yz)$
-    \item $0x = 0$; $x0 = 0$; $(x + y)z = xz + yz$; $x(y + z) = xy + xz$
-  \end{itemize}
-\end{definition}

### 2023-06-22 tex/semirings.tex
[ writing ] mostly C3 related work
[ writing ] move semiring definition out of prelims

@@
 behaved.
-%\todo{Refer back to preliminaries for definition of posemiring}
+
+\begin{definition}
+  A \emph{semiring} is a monoid in the multicategory of commutative monoids and
+  multilinear maps.
+  Unpacked, this means that we have a set $\mathscr R$ together with elements
+  $0$ and $1$ and binary operators $+$ and $\cdot$ (with $\cdot$ usually written
+  as juxtaposition) such that the following hold for all
+  $x, y, z \in \mathscr R$.
+  \begin{itemize}
+    \item $0 + x = x$; $x + 0 = x$; $(x + y) + z = x + (y + z)$; $x + y = y + x$
+    \item $1x = x$; $x1 = x$; $(xy)z = x(yz)$
+    \item $0x = 0$; $x0 = 0$; $(x + y)z = xz + yz$; $x(y + z) = xy + xz$
+  \end{itemize}
+\end{definition}
+\begin{definition}
+  A \emph{posemiring} is a semiring in the category of partially ordered sets
+  (posets).
+  Unpacked, this means that we have a semiring (in the category of sets)
+  $(\mathscr R, 0, +, 1, \cdot)$ such that $\mathscr R$ has a partial order
+  (written $\leq$) and addition and multiplication are monotonic with respect
+  to $\leq$ in both arguments.
+\end{definition}
+
 For concreteness, I collect together the definition of the
@@
 \input{rec.tex}
-\section{Translation to and from existing systems}\label{sec:translation}
+\section{Representing existing linear and modal logics}\label{sec:translation}
 \input{trans.tex}

### 2023-06-22 tex/lr.tex
[ add ] note on subusaging order

@@
 
+Note that my subusaging order is reversed relative to the similar ordering found
+in previous work --- particularly \citet{Granule18} and related work from those
+authors.
+I find the order I use preferable because it matches the standard for
+simultaneous renaming and substitution (we will see later that a
+$\leq$-relationship between contexts gives rise to a renaming), and all of these
+make rules like \TirName{Subuse} interpretable as generalised notions of
+composition between orderings/renamings/substitutions and terms.
+
 \paragraph{Tensor- and with-products}

### 2023-06-22 tex/syntax.tex
[ add ] graph example

@@
 usage-annotated $\mu\tilde\mu$-calculus and a Linear/non Linear system.
+
+\subsection{An encoding of graphs}
+
+As a non-logical example of a syntax that can be encoded using linear syntax
+descriptions, I consider a language of directed acyclic hypergraphs, which can
+be used as string diagrams for symmetric monoidal categories.
+I want to represent hypergraphs like the following, made up of operators
+$\{
+\tikz{\node (n) [circle,draw,minimum size=2mm,inner sep=0] {};%
+  \draw[->] (n) to (0,-3mm)},
+\tikz{\node (n) [circle,draw,minimum size=2mm,inner sep=0] {};%
+  \draw[->] (2.12mm,2.12mm) to (n); \draw[->] (-2.12mm,2.12mm) to (n);%
+  \draw[->] (n) to (0,-3mm)},
+\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
+  \draw[->] (0,3mm) to (n)},
+\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
+  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
+  \draw[->] (0,3mm) to (n)}
+\}$
+and wires between them with no fan-in or fan-out.
+I have given the wires names (a, b, w, x, y, and z), which I use in a textual
+representation of the graph on the right.
+The textual representation is in A-normal form, where the two nullary operators
+have names beginning with \texttt{eta}, the two binary operators have names
+beginning with \texttt{mu}, and names having the suffix \texttt{white} or
+\texttt{black} based on the corresponding operator's colour.
+
+\noindent
+\begin{minipage}[t]{0.5\textwidth}
+  \begin{tikzpicture}[baseline]
+    \node (i0) {};
+    \node[right=2cm of i0] (i1) {};
+    \node[below=1cm of i0] (mub) [circle,fill] {};
+    \node[below right=1cm of mub] (muw) [circle,draw] {};
+    \node[below left=1cm of muw] (etaw) [circle,draw] {};
+    \node[below right=1cm of etaw] (etab) [circle,fill] {};
+    \node[below left=1cm of etab] (o0) {};
+    \node[below right=1cm of etab] (o1) {};
+    \draw[->] (i0) to node[auto] {a} (mub);
+    \draw[->] (i1) to[out=-90,in=45] node[auto] {b} (muw);
+    \draw[->] (mub) to node[auto] {x} (muw);
+    \draw[->] (muw) to node[above right] {y} (etab);
+    \draw[->] (etaw) to node[auto] {z} (o0);
+    \draw[->] (mub) to[out=-135,in=90] node[left,near start] {w} (o1);
+  \end{tikzpicture}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+  \vspace{1.5em}
+\begin{verbatim}
+a, b |-
+let w, x := mu_black(a) in
+let y := mu_white(x, b) in
+let z := eta_white() in
+let := eta_black(y) in
+z, w
+\end{verbatim}
+\end{minipage}
+
+In the textual representation, I treat inputs a and b as free variables, while
+outputs z and w are listed on the final line.
+The rest of the expression is a series of \texttt{let}-expressions binding zero
+or more variables to the results of an operator applied to other variables.
+All of the variables are used linearly, avoiding any fan-out.
+
+To turn this representation into a type system, I do the following.
+First, I fix the usage annotation semiring as the
+$\{\gr0, \gr1, \gr\omega\}$-semiring to achieve linearity.
+Then, I set \AgdaBound{Ty} to be the following type.
+All of the variables get type \AgdaInductiveConstructor{wire}, with a context
+full of \AgdaInductiveConstructor{wire}-type variables corresponding to the
+inputs of the graph.
+Meanwhile, the whole expression has type
+\AgdaInductiveConstructor{bundle}\AgdaSpace{}\AgdaBound{s}, where \AgdaBound{s}
+gives the shape of the outputs.
+
+\ExecuteMetaData[\Graphtex]{Sort}
+
+The type system itself is as follows.
+Each operator gets its own syntactic construct, with essentially the same naming
+conventions as for the textual representation above.
+Each of these constructs is an entire \texttt{let}-expression.
+There is also an \AgdaInductiveConstructor{`end} construct, allowing us to
+finish and list the outputs.
+
+To see how the encodings of the operations work, let us look at the
+\AgdaInductiveConstructor{`$\upmu\bullet$}-construct
+($\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
+  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
+  \draw[->] (0,3mm) to (n)}$).
+It has two premises conjoined by $\sep$ (in fact, all premises in this language
+are separated, with no use of sharing conjunction).
+The first premise expects a simple value, corresponding to the input of the
+operator.
+The second premise is more complex, and represents the body/continuation of the
+\texttt{let}-expression.
+It binds two variables, corresponding to the two outputs, which can then be used
+in later \texttt{let}-expressions or the outputs.
+The bound variables all have type \AgdaInductiveConstructor{wire} and usage
+annotation $\gr1$ (written \AgdaInductiveConstructor{u1} in Agda) to make them
+behave linearly.
+
+\ExecuteMetaData[\Graphtex]{qGraphL}
+\ExecuteMetaData[\Graphtex]{GraphL}
+
+The required premises of the \AgdaInductiveConstructor{`end}-rule are calculated
+by \AgdaFunction{end-premises}.
+This follows the tree structure of the shape, requiring a
+\AgdaInductiveConstructor{wire}-term in an unextended context for each leaf.
+
+\ExecuteMetaData[\Graphtex]{end-premises}
+
+A graph is a term of the graph language.
+Specifically, a graph with inputs of shape \AgdaBound{s} and outputs of shape
+\AgdaBound{t} has the following type --- a term in a context of shape
+\AgdaBound{s}, all of whose entries are wire variables with annotation $\gr1$,
+whose result type is \AgdaInductiveConstructor{bundle}\AgdaSpace{}\AgdaBound{t}.
+
+\ExecuteMetaData[\Graphtex]{Graph}
+
+Finally, I give an example term \AgdaFunction{myGraph}.
+As I said at the end of \cref{sec:terms}, writing a term out in full is tedious,
+so I instead choose to use the machinery I discuss in detail in
+\cref{sec:usage-elaborator}.
+The \AgdaFunction{elab-unique} tool elaborates a well typed term into a well
+typed and usage-correct term as long as it can infer assignments of usage
+annotations that satisfy the constraints.
+I add the prefix \texttt{u} to signify the unannotated (just well typed) terms.
+All of the checks required by the elaboration procedure are done at Agda's type
+checking time, so we know that the program listed below can be elaborated by
+virtue of the whole Agda definition being type-checked.
+
+The \AgdaFunction{uvar\#} tool is like the \AgdaFunction{var\#} tool I used
+earlier --- allowing us to refer to variables by counting from the lefthand end
+of the context.
+I add these numbers to the picture of the graph for convenience.
+Variables never go out of scope, despite being used, so these numbers do not
+change with any bindings.
+
+\noindent
+\begin{minipage}[t]{0.33\textwidth}
+  \begin{tikzpicture}[baseline]
+    \node (i0) {};
+    \node[right=2cm of i0] (i1) {};
+    \node[below=1cm of i0] (mub) [circle,fill] {};
+    \node[below right=1cm of mub] (muw) [circle,draw] {};
+    \node[below left=1cm of muw] (etaw) [circle,draw] {};
+    \node[below right=1cm of etaw] (etab) [circle,fill] {};
+    \node[below left=1cm of etab] (o0) {};
+    \node[below right=1cm of etab] (o1) {};
+    \draw[->] (i0) to node[auto] {0} (mub);
+    \draw[->] (i1) to[out=-90,in=45] node[auto] {1} (muw);
+    \draw[->] (mub) to node[auto] {3} (muw);
+    \draw[->] (muw) to node[above right] {4} (etab);
+    \draw[->] (etaw) to node[auto] {5} (o0);
+    \draw[->] (mub) to[out=-135,in=90] node[left,near start] {2} (o1);
+  \end{tikzpicture}
+\end{minipage}
+\begin{minipage}[t]{0.67\textwidth}
+  \AgdaNoSpaceAroundCode{}
+  \ExecuteMetaData[\Graphtex]{myGraph}
+  \AgdaSpaceAroundCode{}
+\end{minipage}
 

### 2023-06-22 tex/examples.tex
[ writing ] expand and reorganise examples

@@
-In this section, we provide three example uses of semantic traversals: generic
-renaming and substitution, a usage elaborator, and a denotational semantics.
+In this chapter, I provide three example uses of semantic traversals: generic
+renaming and substitution\todo{Moved}, a usage elaborator, and a denotational semantics.
 The reader is also encouraged to see the far greater range of examples in the
@@
 
-A result we will use throughout this section is \emph{reification}.
-When we have an index-preserving mapping from usage-checked variables to
-$\V$-environments, we can construct environments of the form
-$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
-This lets us write the \AgdaFunction{reify} function, which  simplifies our
-obligations in giving a \AgdaRecord{Semantics} by coercing
-\AgdaFunction{Kripke} functions into just
-\AgdaBound{$\C$}-values in an extended context.
-
-%To show this, we instantiate the Kripke function with the renaming
-%$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
-%pass it an argument environment
-%$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
-%extended part.
-%Post-composition of a renaming onto an environment comes from
-%\cref{thm:env-postren}.
-
-%\ExecuteMetaData[\Syntactictex]{reify}
-
-\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
-  If $\V$ is an open family such that there is a function
-  $v : \forallb{{\sqni} \dotto \V}$, we get a function of type
-  $\forallb{\mathrm{Kripke}\,\V\,\C \dotto \mathrm{Scope}\,\C}$ for any $\C$.
-\end{lemma}
-\begin{proof}
-  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
-  That is, $b$ is a Kripke function yielding $\C$-computations
-  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
-  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
-  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
-  $\Gamma, \gr0\delta$.
-  Then we give the $\wand$-function an argument in context
-  $\gr0\gamma, \Delta$, noting that
-  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
-  as we wanted from the result.
-  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
-  We produce this via \cref{thm:env-postren} from an environment
-  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
-  and a renaming which is the complement to that used on $\Box^r$.
-\end{proof}
-
-All of the $\V$s used in examples in this paper support identity environments.
-However, \citet[p.~27]{AACMM21} give some important examples that do not
-support identity environments, and thus cannot use \AgdaFunction{reify}
-(\cref{thm:reify}).
-The feature that causes the lack of support for identity environments is that
-a semantics can make use of the fact that only variables of particular kinds
-are bound by the syntax.
-In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
-binds variables that synthesise their type, as opposed to those whose type is
-checked.
-The semantics of type-checking and elaboration rely on variables synthesising
-their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
-Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
-only binds such synthesising variables.
-Similar phenomena would appear in, say, a call-by-value language where
-variables are values (not computations), or a polarised language where
-variables always have a polarity matching their type.
-
-\subsection{Renaming and substitution}\label{sec:kit-to-sem}
-
-In an unpublished note, \citet{McBride05} gives a parametrised traversal
-yielding homomorphisms of syntax, the canonical examples of which are
-simultaneous renaming and simultaneous substitution.
-The parameters are collected in the record \AgdaRecord{Kit}.
-We make a minor change to the original presentation, where instead of our
-\AgdaField{ren\textasciicircum{}$\V$} field, \citeauthor{McBride05} has the
-field \AgdaField{wk} allowing only context extensions.
-As for the other two fields, \AgdaField{vr} allows us to map variables to
-$\V$-values, so as to put newly bound variables in environments; and
-\AgdaField{tm} allows us to extract terms from $\V$-values, as required when
-we use the environment to handle a free variable.
-
-\ExecuteMetaData[\Syntactictex]{Kit}
-
-Where \citeauthor{McBride05} gave the traversal explicitly, we go via our
-generic semantic traversal.
-The first two fields of \AgdaRecord{Semantics} derive directly from fields of
-\AgdaRecord{Kit}.
-Meanwhile, to handle term constructors, we first \AgdaFunction{reify} to get a
-collection of traversed subterms, and then use \AgdaInductiveConstructor{`con}
-to assemble these subterms into a similarly shaped syntactic form as we started
-with.
-The \AgdaField{vr} field is used implicitly in \AgdaFunction{reify}, as it is
-used to show that $\V$-identity environments exist.
-
-\ExecuteMetaData[\Syntactictex]{kit-to-sem}
-
-The action of a syntactic traversal on logical rules is basically fixed: we
-preserve the logical rule and extend the environment with any newly bound
-variables according to \AgdaField{vr}.
-Meanwhile, the action on variables is relatively unconstrained: we look up the
-variable in the environment to get a $\V$-value, then transform that $\V$-value
-into a term using \AgdaField{tm}.
-
-The idea of simultaneous renaming is that variables replace variables, whereas
-with simultaneous substitution, terms replace variables.
-This translates to environments for renaming containing $\sqni$-values
-(variables), and environments for substitution containing $\vdash$-values
-(terms).
-
-%To implement renaming and substitution for terms, we now just implement
-%syntactic kits for variables and terms, respectively.
-
-\ExecuteMetaData[\Syntactictex]{Ren-Kit}
-
-Notice that \AgdaFunction{ren\textasciicircum$\vdash$}, witnessing the fact
-that terms are renameable, is a corollary of \AgdaFunction{Ren-Kit}.
-
-\ExecuteMetaData[\Syntactictex]{Sub-Kit}
-
-\subsection{A usage elaborator}\label{sec:usage-elaborator}
+\section{A usage elaborator}\label{sec:usage-elaborator}
 
@@
 with the size of terms.
-An additional factor, which we can't see on paper, is that type checking time
+An additional factor, which we can't see on paper but is nonetheless
+significant, is that type checking time
 for these proofs soon becomes prohibitive to interactive development.
 
-Our aim in this subsection is to automate usage constraint proofs, making terms
+Our aim in this section is to automate usage constraint proofs, making terms
 both easier to write and more performant to check.
@@
 
-\subsection{A denotational semantics}
+\section{A denotational semantics}\label{sec:den-sem}
 
@@
 \end{example}
+
+\section{Translating between $\name$ and L/nL}\label{sec:lnl}
+\input{lnl.tex}

### 2023-06-22 tex/lnl.tex
[ writing ] expand and reorganise examples

--- 2021-06-08 tex/lnl.tex
@@
-We can express Benton's linear/non-linear logic~\cite{Benton94} in the
-framework.
-There are two apparent discrepancies between L/nL and what we have seen so far
-to be expressible in the framework: the existence of multiple judgement modes
-and the restriction on the kinds of assumptions based on the mode.
-We will start by addressing these discrepancies, before presenting the encoding
-in full and proving the resulting system logically equivalent to
-$\lambda\gr{\mathcal R}$.
+In \cref{sec:dup-lnl}, I gave what I claimed to be an encoding of
+Linear/non-Linear logic~\citep{Benton94} as a syntax description.
+In this section, I rigorously state and prove the correspondence between
+\citeauthor{Benton94}'s definition of L/nL and my encoding of it.
+Then, I give translations from this encoding to my encoding of $\name$, and
+vice versa, using two generic semantic traversals.
+These results together should give us confidence that the encoding of L/nL is
+correct up to logical equivalence.
 
-\subsection{Encoding L/nL}
+\subsection{Encoding L/nL}\label{sec:encoding-LnL}
 
-In L/nL, we have a \emph{linear} and an \emph{intuitionistic} mode, with
-separate types (respectively $A$ and $X$) and typing judgements (respectively
-$\vdash_{\mathcal L}$ and $\vdash_{\mathcal C}$) connected by modalities $F$ and
-$G$.
-The mode of the conclusion type is the same as the mode of the judgement, so
-only a distinction in the conclusion mode is necessary.
-To achieve this distinction in the types, we have an indexed type family
-$\mathrm{Ty} : \mathrm{Frag} \to \mathrm{Set}$, and set the framework types to
-be the type $(f : \mathrm{Frag}) \times \mathrm{Ty}\,f$.
-% \AgdaDatatype{Ty}\AgdaSymbol{:}\AgdaDatatype{Frag}\AgdaSymbol{$\to$}
-% \AgdaDatatype{Set}
+I will present translations between the systems given by
+\cref{fig:LnL-orig,fig:LnL-bunched}.
 
-In L/nL, while linear judgements can have both linear and intuitionistic
-assumptions, intuitionistic judgements can only have intuitionistic assumptions.
-We encode this invariant on intuitionistic judgements using the $\Box$
-premise combinator on all intuitionistic conclusions and wherever
-intuitionistic premises occur for linear conclusions.
+I use $S$ to range over both linear and intuitionistic variables.
+In this section, I use the notations $\Gamma \vdash A$ and $\Gamma \vdash X$
+without subscripts on the turnstile to refer to the encoded version of the
+calculus.
+This notation keeps the encoded calculus distinct from the reference L/nL
+calculus I am translating from and to.
+
+The main difference between the original L/nL calculus and the encoded version
+is that the encoded version contains some extra ``junk'', not corresponding to
+anything in the original L/nL calculus.
+This junk includes all of the wrinkles we saw when translating to and from DILL
+in \cref{sec:trans-dill} --- where in the semiring-annotated system, variables
+annotated $\gr\omega$ (corresponding to intuitionistic variables) can slip into
+having annotation $\gr1$ or $\gr0$ whenever there are any algebraic
+manipulations of the context.
+In linear/non-linear logic, this slipping causes extra problems because
+intuitionistic variables are supposed to be of a distinct sort to other
+(i.e., linear) variables.
+Additionally, we have no means in the framework to correlate types with usage
+annotations, so we must deal with free variables carefully to ensure the
+required correlation between linear and intuitionistic types and annotations.
+
+With the above remarks in mind, I take it as clear how to translate the original
+L/nL calculus into the encoded version, and just state the type of the
+translation in \cref{thm:lnl-to-enc} without including a proof.
+In contrast, I spend most of this subsection on the reverse translation, which
+I provide a proof sketch of in \cref{thm:enc-to-lnl}.
+
+\begin{proposition}\label{thm:lnl-to-enc}
+  We can construct the following translations.
+  \begin{align}
+    (\Theta \vdashC X) &\to (\gr\omega\Theta \vdash X) \\
+    (\Theta; \Gamma \vdashL A) &\to (\gr\omega\Theta, \gr1\Gamma \vdash A)
+  \end{align}
+\end{proposition}
+
+The key property needed to sensibly do the reverse translation is
+\emph{linear well-formedness}, as given by \cref{def:lin-well-formed}.
+Linear well-formedness says that variables of linear type have linear usage
+annotations.
+It does not say anything about intuitionistic types and usage annotations for
+two reasons.
+First, talking about $\gr\omega$ is not sufficiently stable.
+As we work up a derivation, the ``slip'' described earlier says that usage
+annotations will tend to get larger, i.e.\ more precise.
+Therefore, it makes more sense to make conditions of being greater than or equal
+to some collection of usage annotations.
+Second, it turns out to be unnecessary to add any conditions on variables with
+intuitionistic type, because we can just treat them as if they were annotated
+$\gr\omega$.
+We can forget the specificness of annotations $\gr0$ and $\gr1$ when not needed,
+because whatever a specifically annotated variable can do can be done by an
+$\gr\omega$-annotated variable.
+
+\begin{definition}\label{def:lin-well-formed}
+  A semiring-annotated context for L/nL is \emph{linearly well formed} when all
+  linear variables are annotated either $\gr0$ or $\gr1$.
+\end{definition}
+
+\begin{lemma}\label{thm:lwf}
+  If $\Gamma$ is linearly well formed and $M : \plr{\Gamma \vdash S}$, then the
+  context of every subterm of $M$ is linearly well formed.
+\end{lemma}
+\begin{proof}
+  This lemma follows by inspection of the syntax description
+  (\cref{fig:LnL-bunched}).
+  In the subterms, the linear variables in $\Gamma$ are only changed by binding
+  new variables (all instances of which maintain linear well formedness) and by
+  existing variables being shared out or coerced (which never produces
+  $\gr\omega$ annotations from $\gr0$ or $\gr1$).
+\end{proof}
+
+Linear well-formedness is the only condition needed for the translation given by
+\cref{thm:enc-to-lnl}.
+We can translate a derivation with any such context, with no further
+specification of its shape.
+In particular, the shape is not calculated from an original L/nL context.
+
+\begin{proposition}\label{thm:enc-to-lnl}
+  Let $\Gamma_{\mathcal C}$ be the list of variables of intuitionistic type in
+  $\Gamma$.
+  Let $\Gamma_{\mathcal L}$ be the list of variables of linear type in $\Gamma$
+  with usage annotation $\gr1$.
+  Then, whenever $\Gamma$ is linearly well formed, we can construct the
+  following translations.
+  \begin{align}
+    (\Gamma \vdash X) &\to \plr{\Gamma_{\mathcal C} \vdashC X} \\
+    (\Gamma \vdash A) &\to
+      \plr{\Gamma_{\mathcal C}; \Gamma_{\mathcal L} \vdashL A}
+  \end{align}
+\end{proposition}
+\begin{proof}
+  We proceed by mutual induction on the derivations.
+
+  First, I consider variables.
+  Suppose we have $\Gamma \sqni S$.
+  If $S$ is a linear type $A$, then $\Gamma$ contains one variable of type $A$
+  annotated $\gr1$, while all of the other linear variables are annotated $\gr0$
+  (by linear well formedness, we have no linear variables annotated
+  $\gr\omega$).
+  Therefore, $\Gamma_{\mathcal L} = A$, and \TirName{$\mathcal L$-var} applies.
+  Otherwise, if $S$ is an intuitionistic type $X$, then
+  \TirName{$\mathcal C$-var} applies to yield $\Gamma_{\mathcal C} \vdashC X$.
+
+  For the logical rules, linear well formedness means that all variables of
+  linear type act linearly.
+  Additionally, every L/nL rule requiring there to be no linear variables in
+  scope is guarded by $\Boxzpt$ or $I^*$ in the syntax description, excluding
+  linear variables.
+  Given these behaviours, the two calculi correspond closely, and it is a matter
+  of inspection (and use of \cref{thm:lwf} when using the induction hypothesis)
+  to complete the proof.
+\end{proof}
 
@@
 
-The translations largely follow Benton's originals.
+\Citet[\S 3.3]{Benton94} gives syntactic translations back and forth between
+Linear/non-Linear logic and the presentation of intuitionistic linear logic
+given by \citet{BBdePH93}.
+In this section, I give analogous translations between my encodings of L/nL and
+$\name$ as instances of the generic traversal \AgdaFunction{semantics}.
+More precisely, I instantiate $\name$ to the $\{\gr0 > \gr\omega < \gr1\}$
+posemiring and restrict it to the fragment containing connectives $I$,
+$\otimes$, $\multimap$, and $\oc\gr\omega$, matching the fragment of L/nL
+presented by \citeauthor{Benton94} and in this section.
+Notably, this fragment of $\name$ excludes $\oc\gr0$ and $\oc\gr1$.
+I write $\oc\gr\omega$ as just $\oc$, as in traditional linear logic.
+
+Under the above restrictions and conventions, \citeauthor{Benton94}'s
+translations between the types of ILL and L/nL can be used verbatim,
+and are reproduced in \cref{fig:lnl-lr-types}.
+Notably, the image of each ILL type under the $\plr{-}^o$-translation falls in
+the \emph{linear} types of L/nL.
+In the other direction (the $\plr{-}^*$-translation), we make extensive use of
+the $\oc$-type former to translate the intuitionistic types of L/nL.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{.49\linewidth}
+    \centering
+    \begin{align*}
+      &(-)^* : \mathrm{Ty}_\lnl \to \mathrm{Ty}_{\name} \\
+      &\begin{aligned}
+        I^* &= I \\
+        \plr{A \otimes B}^* &= A^* \otimes B^* \\
+        \plr{A \multimap B}^* &= A^* \multimap B^* \\
+        \plr{FX}^* &= \oc X^* \\
+        1^* &= I \\
+        \plr{X \times Y}^* &= \oc X^* \otimes \oc Y^* \\
+        \plr{X \to Y}^* &= \oc X^* \multimap Y^* \\
+        \plr{GA}^* &= A^*
+      \end{aligned}
+    \end{align*}
+    \begin{align*}
+      &(-)^* : \mathrm{Ctx}_\lnl \to \mathrm{Ctx}_{\name} \\
+      &\plr{\plr{\grR\gamma}^*}_i = \grR_i\gamma_i^*
+    \end{align*}
+  \end{subfigure}
+  \begin{subfigure}{.49\linewidth}
+    \centering
+    \begin{align*}
+      &(-)^\circ : \mathrm{Ty}_{\name} \to \mathrm{Ty}_{\lin} \\
+      &\begin{aligned}
+        I^\circ &= I \\
+        \plr{A \otimes B}^\circ &= A^\circ \otimes B^\circ \\
+        \plr{A \multimap B}^\circ &= A^\circ \multimap B^\circ \\
+        \plr{\oc A}^\circ &= GFA^\circ
+      \end{aligned}
+    \end{align*}
+    \begin{align*}
+      &(-)^\circ : \oiw \times \mathrm{Ty}_{\name} \to \Sigma_f~\mathrm{Ty}_f \\
+      &\begin{aligned}
+        \plr{\gr0A}^\circ &= \plr{\lin, A^\circ} \\
+        \plr{\gr1A}^\circ &= \plr{\lin, A^\circ} \\
+        \plr{\gr\omega A}^\circ &= \plr{\intu, GA^\circ}
+      \end{aligned}
+    \end{align*}
+    \begin{align*}
+      &(-)^\circ : \mathrm{Ctx}_{\name} \to \mathrm{Ctx}_\lnl \\
+      &\plr{\plr{\grR\gamma}^\circ}_i = \grR_i\plr{\grR_i\gamma_i}^\circ
+    \end{align*}
+  \end{subfigure}
+  \caption{Translation of types between L/nL and $\name$}
+  \label{fig:lnl-lr-types}
+\end{figure}
+
+I extend $\plr{-}^*$ to contexts pointwise on the type context.
+Note that this differs from \citeauthor{Benton94}'s translation of contexts in
+that the intuitionistic variables of $\lnl$ are translated using usage
+annotation $\gr\omega$, rather than type former $\oc$.
+A translation from $\lnl$ to DILL would probably similarly use DILL's
+intuitionistic variables rather than $\oc$.
+
+For $\plr{-}^\circ$, I use an extra step to avoid producing contexts
+that are not linearly well formed per \cref{def:lin-well-formed}.
+Specifically, wherever there is an annotation $\gr\omega$ in a $\name$ context,
+the corresponding type is wrapped in a $G$ to make it intuitionistic.
+For example, $\plr{\gr0A, \gr1B, \gr\omega C}^\circ =
+\plr{\gr0A^\circ, \gr1B^\circ, \gr\omega GC^\circ}$.
+
+In Agda code, I define the $\plr{-}^\circ$ operator on
+$\oiw \times \mathrm{Ty}_{\name}$ (the one that introduces a $G$ for each
+$\gr\omega$) via a \emph{view} \AgdaDatatype{LIView}~\citep{MM04}.
+I define usage annotations $\gr0$ and $\gr1$ (\AgdaInductiveConstructor{u0} and
+\AgdaInductiveConstructor{u1} in Agda) to be \emph{linear}, with only
+$\gr\omega$ (\AgdaInductiveConstructor{u$\upomega$}) being
+\emph{intuitionistic}.
+\AgdaDatatype{LIView} is a view because of the existence of
+\ExecuteMetaData[\LinIntViewtex]{liview-type}, and well behaved in the sense
+that any two views of the same usage annotation are equal (witnessed by
+\AgdaFunction{liview-prop}, not shown here).
+The translation from $\name$ to $\lnl$ takes cases between linear and non-linear
+annotations at many points, so having these cases expressed as a view avoids
+duplication of arguments between the cases for $\gr0$ and $\gr1$.
+
+\ExecuteMetaData[\LinIntViewtex]{Linear}
+\ExecuteMetaData[\LinIntViewtex]{LIView}
+
+\begin{theorem}\label{thm:lnl-to-lr}
+  Let $S$ be an $\lnl$ type, either linear or intuitionistic.
+  Then, we can translate any L/nL term to a $\name$ term as follows.
+  \begin{align}
+    (\Gamma \vdash_\lnl S) &\to (\Gamma^* \vdash_{\name} S^*)
+  \end{align}
+\end{theorem}
+\begin{proof}
+  The proof is mostly straightforward, largely following Benton's translation.
+  Similarly to the denotational semantics of \cref{sec:den-sem}, we may use a
+  \AgdaRecord{Semantics} with $\V$ set to $\sqni_{\lnl}$, and then set
+  $\C\,\Gamma\,S \coloneqq \Gamma^* \vdash_{\name} S^*$.
+  Whenever we have induction hypotheses of \AgdaFunction{Kripke} type, we use
+  the \AgdaFunction{reify} function for $\name$ to get $\name$ terms.
+  Therefore, we are essentially just doing a proof by induction on the structure
+  of the input term.
+
+  Translating the \TirName{$F$i} rule into \TirName{$\oc$i} relies
+  on the equivalence between duplicability (as expressed by the $\Box$ premise
+  connective) and having been scaled by $\gr\omega$ (as expressed by the
+  $\gr\omega \cdot {}$ premise connective).
+  This holds of the $\oiw$ semiring, but not of general semirings (and is not
+  even stateable for general semirings because of the mention of $\gr\omega$).
+  The same reasoning is needed when translating the introduction rules for
+  intuitionistic connectives, because they always have $\Box$ed premises and
+  are translated using $\oc$.
+
+  As an example of translating an intuitionistic elimination rule, let us
+  consider the \TirName{$\times$e$_0$} rule.
+  I reproduce it here with explicit contexts.
+  \[
+    \ebrule{%
+      \hypo{\Gamma' \leq \Gamma}
+      \hypo{\Gamma\text{ duplicable}}
+      \hypo{\Gamma \vdash_{\lnl} X \times Y}
+      \infer3[$\times$e$_0$]{\Gamma' \vdash_{\lnl} X}
+    }
+  \]
+
+  Recall that $\plr{X \times Y}^* = \oc X^* \otimes \oc Y^*$.
+  This means that we must pattern-match the hypothesis to get variables
+  $\gr\omega X^*, \gr\omega Y^*$ so as to be able to use the $X^*$ for the
+  conclusion and discard the $Y^*$.
+  The formal derivation is as follows.
+  We are able to copy $\Gamma^*$ between all of these subterms because its usage
+  annotations are the same as those of $\Gamma$, which is duplicable by the
+  assumption in \TirName{$\times$e$_0$}.
+  The distinction between $\Gamma'$ and $\Gamma$ is minor.
+  When we apply the \TirName{$\otimes$e} rule, we use the fact that
+  $\Gamma' \leq \Gamma + \Gamma$, by the fact that $\Gamma' \leq \Gamma$ and
+  $\Gamma \leq \Gamma + \Gamma$.
+  From then on, we need only think about $\Gamma$, which behaves well because it
+  is duplicable.
+
+  \begin{align*}
+    &\ebrule{%
+      \hypo{\text{IH}}
+      \infer[no rule]1{\Gamma^* \vdash_{\name} \oc X^* \otimes \oc Y^*}
+      \infer0[Var]{\Gamma^*, \gr1\oc X^*, \gr0\oc Y^* \vdash_{\name} \oc X^*}
+      \hypo{\nabla}
+      \infer2[$\oc$e]{\Gamma^*, \gr1\oc X^*, \gr1\oc Y^* \vdash_{\name} X^*}
+      \infer2[$\otimes$e]{\Gamma^* \vdash_{\name} X^*}
+    }
+    \\
+    &\text{where }\nabla \coloneqq
+    \\
+    &\ebrule{%
+      \infer0[Var]{\Gamma^*, \gr0\oc X^*, \gr1\oc Y^*, \gr\omega X^* \vdash_{\name} \oc Y^*}
+      \infer0[Var]{\Gamma^*, \gr0\oc X^*, \gr0\oc Y^*, \gr\omega X^*, \gr\omega Y^* \vdash_{\name} X^*}
+      \infer2[$\oc$e]{\Gamma'^*, \gr0\oc X^*, \gr1\oc Y^*, \gr\omega X^* \vdash_{\name} X^*}
+    }
+  \end{align*}
+
+  In the Agda proof, renaming is required to perform lots of minor functions
+  that we would ignore on paper.
+  For example, the equation $\plr{\Gamma, \Delta}^* = \Gamma^*, \Delta^*$ ---
+  required when induction hypotheses have newly bound variables --- is not true
+  definitionally.
+  Furthermore, because of the functional representation I use for contexts, it
+  is not even provable without function extensionality.
+  Therefore, renaming is the simplest way to get the required coercion.
+  Such renamings perhaps could be avoided in most cases if contexts were
+  represented as non-functional lists.
+\end{proof}
+
+\begin{theorem}\label{thm:lr-to-lnl}
+  We can translate from $\name$ to the linear fragment of $\lnl$.
+  \begin{align}
+    (\Gamma \vdash_{\name} A) \to (\Gamma^\circ \vdash_\lnl A^\circ)
+  \end{align}
+\end{theorem}
+\begin{proof}
+  We use the same simple induction scheme as in \cref{thm:lnl-to-lr}, but with
+  the places of $\name$ and $\lnl$ switched.
+  However, some of the rule translations are more complex, mainly caused by the
+  complexity of the $\plr{-}^\circ$ translation on contexts.
+
+  For variables, we must consider separately the cases where the variable being
+  used is annotated $\gr1$ and $\gr\omega$.
+  The case for a variable $\gr1A$ is straightforward, and translates directly
+  to an $\lnl$ variable $\gr1A^\circ$.
+  A variable $\gr\omega A$, however, is translated to $\gr\omega GA^\circ$, so
+  we must eliminate the $G$ in order to get the conclusion $A^\circ$.
+  The \TirName{$G$e} rule requires its context to be duplicable, which is true
+  by the fact that all of the unused variables are annotated either $\gr0$ or
+  $\gr\omega$ (because they are all less than or equal to $\gr0$), and the
+  variable being used is annotated $\gr\omega$.
+
+  Most logical rules are handled very similarly to each other, so I will just
+  translate the \TirName{$\otimes$i} rule as an example.
+  I reproduce it here with explicit contexts (split into their usage and typing
+  contexts).
+
+  \[
+    \ebrule{%
+      \hypo{\grR \leq \grP + \grQ}
+      \hypo{\grP\gamma \vdash_{\name} A}
+      \hypo{\grQ\gamma \vdash_{\name} B}
+      \infer3[$\otimes$i]{\grR\gamma \vdash_{\name} A \otimes B}
+    }
+  \]
+
+  We cannot do a na\"{i}ve translation to the corresponding application of the
+  \TirName{$\otimes$i} rule of $\lnl$ because $\plr{\grR\gamma}^\circ$,
+  $\plr{\grP\gamma}^\circ$, and $\plr{\grQ\gamma}^\circ$ may all have different
+  typing contexts.
+  For example, consider the following instance.
+  There are two problems.
+  First, our induction hypotheses give us contexts
+  containing $C^\circ$, where our conclusion wants a context containing
+  $GC^\circ$.
+  Second, $\gr1GC^\circ$ is not linearly well formed, because an intuitionistic
+  type is given a linear annotation.
+  It is therefore difficult to work with such a sequent.
+
+  \[
+    \ebrule{%
+      \hypo{\gr1 C \vdash_{\name} A}
+      \hypo{\gr1 C \vdash_{\name} B}
+      \infer2[$\otimes$i]{\gr\omega C \vdash_{\name} A \otimes B}
+    }
+    \quad\rightsquigarrow\quad
+    \ebrule{%
+      \hypo{\gr1C^\circ \vdash_{\lnl} A^\circ}
+      \ellipsis{?}{\gr1 GC^\circ \vdash_{\lnl} A^\circ}
+      \hypo{\gr1C^\circ \vdash_{\lnl} B^\circ}
+      \ellipsis{?}{\gr1 GC^\circ \vdash_{\lnl} B^\circ}
+      \infer2[$\otimes$i]{\gr\omega GC^\circ \vdash_{\lnl} A^\circ \otimes B^\circ}
+    }
+  \]
+
+  To fix these issues, firstly I overwrite the context-splitting given by the
+  input term to conform to the bottom-up style of \cref{def:DILL-bottom-up}.
+  This means precisely that wherever $\gr\omega$ appears in the context of the
+  conclusion, it will also appear in the context of all the hypotheses.
+  This gives the following partial derivation.
+
+  \[
+    \ebrule{%
+      \hypo{\gr1C^\circ \vdash_{\lnl} A^\circ}
+      \ellipsis{?}{\gr\omega GC^\circ \vdash_{\lnl} A^\circ}
+      \hypo{\gr1C^\circ \vdash_{\lnl} B^\circ}
+      \ellipsis{?}{\gr\omega GC^\circ \vdash_{\lnl} B^\circ}
+      \infer2[$\otimes$i]{\gr\omega GC^\circ \vdash_{\lnl} A^\circ \otimes B^\circ}
+    }
+  \]
+
+  Then, I fix the discrepancy in types using substitution.
+  In the running example, the substitution needed for both subterms is of the
+  type $\gr\omega GC^\circ \env{\vdash_{\lnl}} \gr1C^\circ$, which amounts to a
+  term of type $\gr\omega GC^\circ \vdash_{\lnl} C^\circ$, as follows.
+  Note that \TirName{$G$e} is only applicable thanks to changing the usage
+  annotation from $\gr1$ to $\gr\omega$ in the previous step.
+
+  \[
+    \ebrule{%
+      \infer0[Var]{\gr\omega GC^\circ \vdash_{\lnl} GC^\circ}
+      \infer1[$G$e]{\gr\omega GC^\circ \vdash_{\lnl} C^\circ}
+    }
+  \]
+
+  More generally, we may have to produce substitutions of type
+  $$\gr0A^\circ, \gr1B^\circ, \gr\omega GC^\circ, \gr\omega GD^\circ
+  \env{\vdash_{\lnl}}
+  \gr0A^\circ, \gr1B^\circ, \gr1C^\circ, \gr\omega GD^\circ.$$
+  These can be produced pointwise, from substitutions of types
+  $\gr0A^\circ \env{\vdash_{\lnl}} \gr0A^\circ$ and
+  $\gr1B^\circ \env{\vdash_{\lnl}} \gr1B^\circ$ and
+  $\gr\omega GC^\circ \env{\vdash_{\lnl}} \gr1C^\circ$ and
+  $\gr\omega GD^\circ \env{\vdash_{\lnl}} \gr\omega GD^\circ$.
+  We have just seen how to produce the third of these, and the rest are
+  identity substitutions.
+
+  The two sui generis rules to translate are the rules for the $\oc$-modality,
+  with the $\oc$ of intuitionistic linear logic becoming the composite $FG$ in
+  linear/non-linear logic.
+  The way of handling \TirName{$\oc$i} is similar to the way of handling rules
+  like \TirName{$\otimes$i}, but involving multiplication/scaling by $\gr\omega$
+  rather than addition.
+  We have a similar difficult instance, shown below, where an $\gr\omega$ gets
+  specialised to a $\gr1$ via the algebraic operation.
+
+  \[
+    \ebrule{%
+      \hypo{\gr\omega \leq \gr\omega \cdot \gr1}
+      \hypo{\gr1 C \vdash_{\name} A}
+      \infer2[$\oc$i]{\gr\omega C \vdash_{\name} \oc A}
+    }
+  \]
+
+  Again, the solution is to fix up the operation to keep the more general
+  $\gr\omega$, allowing us to apply \TirName{$F$i} and \TirName{$G$i}, and
+  the same substitution as before possibly including an application of
+  \TirName{$G$e} as necessary.
+
+  Finally, the translation of \TirName{$\oc$e} is simple, but worth checking.
+  I reproduce the rule below with explicit contexts.
+
+  \[
+    \ebrule{%
+      \hypo{\grR \leq \grP + \grQ}
+      \hypo{\grP\gamma \vdash_{\name} \oc A}
+      \hypo{\grQ\gamma, \gr\omega A \vdash_{\name} B}
+      \infer3[$\oc$e]{\grR\gamma \vdash_{\name} B}
+    }
+  \]
+
+  The main thing to note is that the translation of the context of the
+  right-hand premise, $\grQ\gamma, \gr\omega A$, is
+  $\plr{\grQ\gamma}^\circ, \gr\omega GA^\circ$ --- i.e., the translation of
+  contexts gives us a $G$ thanks to the $\gr\omega$ usage annotation.
+  Therefore, we do not have to eliminate the $G$, because the right-hand subterm
+  is expected to do it for us.
+  Indeed, we have seen in previous cases many uses of \TirName{$G$e} already.
+
+  I translate the \TirName{$\oc$e} rule as follows.
+  As in the \TirName{$\otimes$i} case, I pick $\grPprime$ and $\grQprime$ to fit
+  bottom-up form, and use the same substitutions as in that case to mend the
+  terms arriving from the induction hypotheses.
+  Then, just \TirName{$F$e} suffices.
+
+  \[
+    \ebrule{%
+      \infer0{\grR \leq \grPprime + \grQprime}
+      \hypo{}
+      \ellipsis{}{\plr{\grPprime\gamma}^\circ \vdash_{\lnl} FGA^\circ}
+      \hypo{}
+      \ellipsis{}{\plr{\grQprime\gamma}^\circ, \gr\omega GA^\circ \vdash_{\lnl} B^\circ}
+      \infer3[$F$e]{\plr{\grR\gamma}^\circ \vdash_{\lnl} B^\circ}
+    }
+  \]
+
+  %In the Agda version of the translation, manipulations are complicated by the
+  %fact that the $\plr{-}^\circ$ translation for contexts does not distribute
+\end{proof}

### 2023-06-22 tex/semantics.tex
[ writing ] expand and reorganise examples

@@
 semantics to terms. For open families $\V$ and $\C$, interpreting
-variables and terms respectively, we assume that $\V$ is renameable,
+variables and terms respectively, we assume that $\V$ is renameable
+(i.e., that $\sdtstile{}\V A \rightarrowtriangle \Box\plr{\sdtstile{}\V A}$ for
+all $A$),
 that $\V$ is embeddable in $\C$, and that we have an algebra for a
@@
 % at \AgdaBound{$\Theta$}.
+
+\AgdaFunction{semantics} is the fundamental lemma of the framework.
+With it proven, I move onto corollaries and specific applications.
+
+\section{Reifying the Kripke function space}
+
+A result I will use throughout the rest of this thesis is \emph{reification}.
+When we have an index-preserving mapping from usage-checked variables to
+$\V$-environments, we can construct environments of the form
+$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
+This lets us write the \AgdaFunction{reify} function, which  simplifies our
+obligations in giving a \AgdaRecord{Semantics} by coercing
+\AgdaFunction{Kripke} functions into just
+\AgdaBound{$\C$}-values in an extended context.
+
+\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
+  If $\V$ is an open family such that there is a function
+  $v : {\sqni} \rightarrowtriangle \V$, we get a function of type
+  $\mathrm{Kripke}\,\V\,\C \rightarrowtriangle \mathrm{Scope}\,\C$ for any $\C$.
+\end{lemma}
+\begin{proof}
+  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
+  That is, $b$ is a Kripke function yielding $\C$-computations
+  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
+  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
+  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
+  $\Gamma, \gr0\delta$.
+  Then we give the $\wand$-function an argument in context
+  $\gr0\gamma, \Delta$, noting that
+  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
+  as we wanted from the result.
+  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
+  We produce this via \cref{thm:env-postren} from an environment
+  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
+  and a renaming which is the complement to that used on $\Box^r$.
+\end{proof}
+
+All of the $\V$s used in examples in this paper support identity environments.
+However, \citet[p.~27]{AACMM21} give some important examples that do not
+support identity environments, and thus cannot use \AgdaFunction{reify}
+(\cref{thm:reify}).
+The feature that causes the lack of support for identity environments is that
+a semantics can make use of the fact that only variables of particular kinds
+are bound by the syntax.
+In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
+binds variables that synthesise their type, as opposed to those whose type is
+checked.
+The semantics of type-checking and elaboration rely on variables synthesising
+their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
+Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
+only binds such synthesising variables.
+Similar phenomena would appear in, say, a call-by-value language where
+variables are values (not computations), or a polarised language where
+variables always have a polarity matching their type.
+
+\section{Renaming and substitution}\label{sec:kit-to-sem}
+
+The final completely syntax-generic result I present is simultaneous
+substitution.
+I derive this as I did in the simply typed case in \cref{sec:gen-sem}:
+I first show that a syntactic kit can be turned into a semantics, and then by
+instantiating the notion of kit for, in turn, renaming and substitution, the
+general semantic traversal gives the result we want.
+
+The notion of \AgdaRecord{Kit} is essentially the same as in the simply typed
+case, once we allow for changes to the basic definitions of variables, terms,
+and environments (in particular, renamings).
+
+\ExecuteMetaData[\Syntactictex]{Kit}
+
+The first two fields of \AgdaRecord{Semantics} derive directly from fields of
+\AgdaRecord{Kit}.
+Meanwhile, to handle term constructors, we first \AgdaFunction{reify} to get a
+collection of traversed subterms, and then use \AgdaInductiveConstructor{`con}
+to assemble these subterms into a similarly shaped syntactic form as we started
+with.
+The \AgdaField{vr} field is used implicitly in \AgdaFunction{reify}, as it is
+used to show that $\V$-identity environments exist.
+
+\ExecuteMetaData[\Syntactictex]{kit-to-sem}
+
+The action of a syntactic traversal on logical rules is basically fixed: we
+preserve the logical rule and extend the environment with any newly bound
+variables according to \AgdaField{vr}.
+Meanwhile, the action on variables is relatively unconstrained: we look up the
+variable in the environment to get a $\V$-value, then transform that $\V$-value
+into a term using \AgdaField{tm}.
+
+The idea of simultaneous renaming is that variables replace variables, whereas
+with simultaneous substitution, terms replace variables.
+This translates to environments for renaming containing $\sqni$-values
+(variables), and environments for substitution containing $\vdash$-values
+(terms).
+
+%To implement renaming and substitution for terms, we now just implement
+%syntactic kits for variables and terms, respectively.
+
+\ExecuteMetaData[\Syntactictex]{Ren-Kit}
+
+Notice that \AgdaFunction{ren\textasciicircum$\vdash$}, witnessing the fact
+that terms are renameable, is a corollary of \AgdaFunction{Ren-Kit}.
+
+\ExecuteMetaData[\Syntactictex]{Sub-Kit}

### 2023-06-22 tex/syntax.tex
[ writing ] expand and reorganise examples

@@
 
-\todo{Maybe partially merge with the later L/nL section.}
 Linear/non-Linear Logic has two sorts of types: linear types $A, B, C$ and
@@
 \AgdaInductiveConstructor{tG}.
-I also throw in a linear base type \AgdaInductiveConstructor{$\iota$}.
+I also include a linear base type \AgdaInductiveConstructor{$\iota$}.
 
@@
 The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
-annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
+annotations $\{\gr0, \gr1, \gr\omega\}$, as I have used before for linear
 systems.
 For reference, I give a standard presentation of the rules of L/nL in
-\cref{fig:LnL-orig}
-
-We want to be able to embed the L/nL judgement forms
-$\Theta; \Gamma \vdash_{\mathcal L} A$ and $\Theta \vdash_{\mathcal C} X$, where
-$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
-variables.
-Following allowed usage patterns, intuitionistic variables will all have usage
-annotation $\gr\omega$, while linear variables will each be annotated
-either $\gr1$ or $\gr0$.
-These assignments of usage annotations are enforced wherever the rules bind new
-variables.
-By usage subsumption, intuitionistic variables may also end up with annotation
-$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
-$\gr\omega$ annotation, not causing a soundness problem.
-The key invariant is that linear variables never get annotation $\gr\omega$.
-
-The framework already tells us what variables and contexts we can
-have --- namely that we can make any assumption in any context.
-This means that we can propose ill formed sequents like
-$\gr1A \vdash X$.
-To make sure that this is not a problem, we want to maintain the invariant that
-any \emph{derivation} of a well formed sequent contains only well formed
-sequents.
-I enforce this locally using the $\Box^{0{+}{*}}$ bunched modality:
-Every rule targeting an intuitionistic judgement has its premises wrapped in a
-$\Box^{0{+}{*}}$ to restrict the usage annotations to $\gr0$ and $\gr\omega$.
-Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
-judgements but have intuitionistic premises, similarly get a $\Box^{0{+}{*}}$.
-Note that having the annotation $\gr0$ means that well formed intuitionistic
-sequents can contain linear variables annotated $\gr0$, but these variables are
-not usable.
-
-The rules are presented in Agda code below, and also using bunched inference
-rule notation in \cref{fig:LnL-bunched}
-
-\ExecuteMetaData[\LnLtex]{quoteLnL}
-\ExecuteMetaData[\LnLtex]{LnL}
+\cref{fig:LnL-orig}.
 
@@
   \begin{align*}
-    \mathcal J \Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
+    A, B, C &\Coloneqq I \mid A \otimes B \mid A \multimap B \mid FX \\
+    X, Y, Z &\Coloneqq 1 \mid X \times Y \mid X \to Y \mid GA \\
+    \mathcal J &\Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
   \end{align*}
@@
 \end{figure}
+
+We want to be able to embed the L/nL judgement forms
+$\Theta; \Gamma \vdashL A$ and $\Theta \vdashC X$, where
+$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
+variables.
+I take the subscript on $\vdashL$ and $\vdashC$ to be a redundant annotation of
+whether the conclusion type is \AgdaInductiveConstructor{lin} or
+\AgdaInductiveConstructor{int}, so it does not appear in the Agda code, but I
+retain it in all non-mechanised presentations.
+
+I will generally ensure that all intuitionistic variables will all have usage
+annotation $\gr\omega$, while linear variables will each be annotated
+either $\gr1$ or $\gr0$.
+These assignments of usage annotations are enforced wherever the rules bind new
+variables.
+By usage subsumption, intuitionistic variables may also end up with annotation
+$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
+$\gr\omega$ annotation, not causing a soundness problem.
+The key invariant is that linear variables never get annotation $\gr\omega$.
+
+The key difference in form between linear and intuitionistic sequents is that
+intuitionistic sequents cannot contain linear variables.
+However, syntax descriptions give us no ability to directly state such a
+restriction on the kinds of variables.
+Therefore, any encoded syntax will allow us to propose ill formed sequents like
+$\gr1A \vdashC X$, where a linear variable is to be used for an intuitionistic
+conclusion.
+To make sure that this is not a problem I use the $\Boxzpt$ bunched modality to
+guard many of the rules, making sure that variables with usage annotation $\gr1$
+cannot occur in \emph{provable} intuitionistic sequents.
+Specifically, every rule targeting an intuitionistic judgement has its premises
+wrapped in a $\Boxzpt$.
+Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
+judgements but have intuitionistic premises, similarly get a $\Boxzpt$.
+Note that having the annotation $\gr0$ means that well formed intuitionistic
+sequents can contain linear variables annotated $\gr0$, but these variables are
+not usable.
+
+For ease of \emph{producing} terms, we may prefer a ``garbage in; garbage out''
+style with a minimal number of uses of $\Boxzpt$, which could be achieved in two
+different ways:
+\begin{itemize}
+  \item
+    Ensure that no sequent of the form $\Gamma, \gr1A \vdashC X$ is derivable.
+    This leaves boxes only on rules \TirName{$1$-I} and \TirName{$G$-I} --- the
+    two logical rules where an intuitionistic conclusion is derived without any
+    intuitionistic premises.
+  \item
+    Ensure that any rule targeting a well formed conclusion has only well formed
+    premises.
+    This leaves boxes only on rules \TirName{$F$-I} and \TirName{$G$-E} --- the
+    two rules where a linear conclusion is derived from intuitionistic premises.
+\end{itemize}
+It would be interesting to show the equivalence of these two approaches with the
+one I have taken in this thesis, but that is left to future work.
+
+The rules are presented in Agda code below, and also using bunched inference
+rule notation in \cref{fig:LnL-bunched}.
 
@@
 
-Later,\todo{When?} I will give translations between this encoding of L/nL and
-$\name$, providing some assurance that the correct logic has been encoded.
+\ExecuteMetaData[\LnLtex]{quoteLnL}
+\ExecuteMetaData[\LnLtex]{LnL}
+
+In \cref{sec:lnl}, I will give translations between the original L/nL system,
+this encoding of L/nL, and $\name$, providing some assurance that the correct
+logic has been encoded.
 

### 2023-06-22 tex/mech-related.tex
[ add ] a bit more related work

@@
 
-Notably for this thesis, the work of \citet{FPT99} has been extended to
-substructural syntaxes by \citet{TP06}.
-This work gives a mechanism for turning a description of contexts and their
-structural rules (expressed as a pseudo-monad on the 2-category of categories)
-into a framework for defining substructural syntaxes, and more generally a
-category of algebras of which the syntax is the initial object.
-As examples, they give the untyped $\lambda$-calculus, an untyped multiplicative
-linear logic, and a bunched logic.
-These examples show a broad range of substructural disciplines they support ---
-comparable to the work of \citet{LicataSR17} (which I discuss in
-\cref{sec:semirings-conc}), and more than I discuss in this thesis.
-However, they also show two of the limitations of their approach.
-Firstly, this work provides no way to track types, though it should be possible
-to incorporate types at the expense of complicating the categorical
-constructions they use.
-Secondly, it appears to be impossible to encode the syntactic forms used for the
-additive connectives (i.e., the Cartesian product and coproduct) of linear
-logic.
-Essentially, subterms can only be combined into a larger term in the same ways
-as how contexts can be appended together.
-For example, in a bunched logic, contexts can be combined through both sharing
-conjunction (the Cartesian product) and separating conjunction
-(a monoidal product).
-Correspondingly, the syntax descriptions of \citeauthor{TP06} allow for the
-syntax of sharing pairs and separating pairs.
-However, in the case of linear logic, contexts can only be combined via a
-monoidal product, so we only get separating pairs (tensor-products) and not
-sharing pairs (with-products).
-\todo{Maybe this should go in the linearity chapter.}
-
 \subsection{Substitution-based semantics}
@@
 
-\subsection{Others}
+\subsection{Logical frameworks}\label{sec:lf}
 
-\begin{itemize}
-  \item JFDI --- \citet{BHKM12}
-\end{itemize}
+Logical frameworks based on higher order abstract syntax are another approach
+which has been used to mechanise the metatheory of logics and programming
+languages.
+Primary among logical frameworks in recent study is the Edinburgh Logical
+Framework~\citep{HHP93}, also known as \emph{LF}.
+The underlying theory of LF is a dependent type theory $\lambda^\Pi$ featuring
+\emph{weak} function spaces.
+Weak function spaces differs from the usual strong function spaces in that
+functions in a weak function space cannot inspect their arguments --- only
+place them whole in their result.
+Therefore, variables of the logical framework can be used as if they were the
+variables of the object language.
+
+Below I give two example constructors of a type family $\mathrm{tm}$.
+These declarations are similar to the rule descriptions I gave in
+\cref{fig:app-lam} for their lack of explicit contexts.
+In this example, $\Pi$ is the type former for dependent weak function spaces,
+with $\to$ being the non-dependent specialisation of $\Pi$.
+The quoted arrow $\qto$ is a type former of the object language, as in the
+previous examples in this chapter.
+
+\begin{align*}
+  \mathrm{lam} &: \Pi A,B.~\plr{\mathrm{tm}\,A \to \mathrm{tm}\,B} \to
+                 \mathrm{tm}\,\plr{A \qto B} \\
+  \mathrm{app} &: \Pi A,B.~\mathrm{tm}\,\plr{A \qto B} \to
+                 \mathrm{tm}\,A \to \mathrm{tm}\,B
+\end{align*}
+
+Like several other approaches I have described in this chapter, logical
+frameworks aim to manage variables, variable binding, and contexts in a generic
+way, giving the user only the choice of what logical rules they want to add to a
+basic calculus.
+In fact, logical frameworks go further than any other approach I have described,
+never giving the user reified access to the context of object language terms,
+and giving immediate access to (single) substitution simply as application of
+weak functions.
+LF also natively supports more complex calculi than the simply typed
+$\lambda$-calculus --- for example allowing us to implement and reason about
+System F.
+
+The main drawback of logical frameworks is that making one requires making a new
+proof assistant, typically with no compatibility with existing proof assistants
+and their libraries of useful mathematical definitions and proofs.
+Particularly, if you want to consider calculi not (conveniently) expressible in
+$\lambda^\Pi$ (as I do, with substructural calculi), then you need to make a new
+logical framework which is probably also incompatible with existing logical
+frameworks.
+As such, no single logical framework acts as a convenient and natural foundation
+upon which to build a broad range of mathematics including substructural logics
+and their metatheory.

### 2023-06-22 tex/conc.tex
[ writing ] start on introduction and conclusion

@@
 \chapter{Conclusions}\label{sec:conc}
+
+In this thesis, I have developed a foundation for semiring-annotated calculi in
+natural deduction style.
 
@@
 
-Is this a reusable library?
-
-Fitch-style systems (can't do).
+\begin{itemize}
+  \item Novel simultaneous substitution for linear type theories
+  \item Generalise \citet{AACMM21} work to substructural logics
+  \item Is this a reusable library? Yes, examples
+  \item Fitch-style systems (can't do).
+  \item Methodology for metatheory of new calculi
+  \item Semirings have been a reasonably robust basis.
+  \item Can the code be directly generalised?
+    Abstract away the algebraic structure of contexts.
+\end{itemize}
 

### 2023-06-22 tex/intro.tex
[ writing ] start on introduction and conclusion

@@
 
-FIXME: this chapter is directly from the ESOP22 paper.
+FIXME: the start of this chapter is directly from the ESOP22 paper.
 
@@
 
-This paper proceeds as follows.
-In \cref{sec:algebra}, we review and fix conventions pertaining to partially
-ordered semirings and vectors over them.
-In \cref{sec:bunched-rules}, we introduce an informal meta-syntax allowing us
-to state substructural typing rules succinctly and without explicit reference
-to contexts.
-In \cref{sec:syntax}, we mechanise that meta-syntax, giving a type of
-\emph{descriptions} of type systems, and interpreting those descriptions as
-types of intrinsically typed terms.
-In \cref{sec:env}, we discuss usage-aware environments: a generalisation of
-the structures used in simultaneous renaming and substitution proofs.
-We use environments in \cref{sec:semantics} to state an alternative
-elimination principle for terms, and give examples of such eliminations in
-\cref{sec:examples}.
-The examples are syntax-generic renaming and substitution, a specific
-denotational semantics, and a syntax-generic usage elaborator.
-Finally, we conclude and discuss future work in \cref{sec:conc}
+\section{Naming and notation conventions}
 
-The work presented in this paper has been mechanised in Agda,
-with the code available for building upon~\cite{generic-lr}.
+I assume familiarity with the Curry-Howard correspondence throughout this
+thesis.
+I make no distinction between logics and type theories, and use terminology from
+each interchangeably.
+Each following bullet point lists a collection of synonyms.
+
+\begin{itemize}
+  \item assumption, hypothesis, variable
+  \item derivation, proof, term
+  \item proposition, formula, type
+  \item connective, type former
+\end{itemize}
+
+I carry out mechanised constructions and proofs in the proof assistant and
+programming language Agda~\citep{Agda}.
+Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
+similarly present non-mechanised constructions and proofs assuming a foundation
+given by dependent type theory, in a style inspired by the HoTT
+Book~\citep{hottbook}.
+
+\section{Outline of the thesis}
+
+This thesis proceeds as follows.
+The next two chapters, \cref{sec:simple,sec:linearity}, are introductory in
+nature, and cover two largely independent strands of prior work.
+In \cref{sec:simple}, I introduce existing methods of representing and reasoning
+about type systems in proof assistants based on dependent type theory.
+I start from well established representations of well scoped and well typed
+terms, and develop these towards a recent approach to environment-based
+semantics given by \citet{AACMM21}.
+In \cref{sec:linearity}, I discuss the challenges faced when one extends a
+treatment of a simple type system, such as that given in \cref{sec:simple}, to
+modal and linear type systems.
+We see that modal and linear type systems apparently violate some of the nice
+properties of the simply typed $\lambda$-calculus we required in
+\cref{sec:simple}.
+I present a solution for intuitionistic S4 modal logic, but leave a solution for
+linear logic to the following chapters.
+
+In the following two chapters, \cref{sec:semirings,sec:ren-sub-lr}, I present a
+calculus $\name$ parametrised by a partially ordered semiring of \emph{usage
+annotations}.
+In \cref{sec:semirings}, I define the calculus, give some possible extensions,
+and show that it subsumes intuitionistic S4 modal logic and Intuitionistic
+Linear Logic.
+In \cref{sec:ren-sub-lr}, I show that $\name$ enjoys generalised versions of the
+nice properties required in \cref{sec:simple}, and I proceed to give novel
+definitions of simultaneous substitutions and their action on $\name$ terms.
+These two chapters are adapted from the work of \citet{WA21}.
+
+The remaining three main chapters,
+\cref{sec:framework,sec:semantics,sec:example-semantics}, adapt the syntactic
+and semantic framework of \citet{AACMM21}, as presented at the end of
+\cref{sec:simple}, to semiring-annotated calculi.
+\Cref{sec:framework,sec:semantics} generalise the work on $\name$ presented in
+\cref{sec:semirings,sec:ren-sub-lr}, respectively.
+\Cref{sec:framework} shows how to formally describe the syntax of an arbitrary
+semiring-annotated calculus, following the constructions used in
+\cref{sec:semirings}.
+\Cref{sec:semantics} then provides the generic environment-based semantic
+traversal on such syntaxes, providing renaming and substitution as per
+\cref{sec:ren-sub-lr} for all syntaxes as special cases of the generic
+traversal.
+\Cref{sec:examples} then gives further example uses of the generic traversal.
+
+Finally, I conclude with \cref{sec:conc}, which discusses the achievements of
+this thesis and openings for future work.

### 2023-07-12 tex/conc.tex
[ writing ] various up to WRel example

@@
 
-In this thesis, I have developed a foundation for semiring-annotated calculi in
-natural deduction style.
+In this thesis, I have developed a foundation for semiring-annotated calculi
+presented in natural deduction style.
+I have given a consolidated account of the semiring-annotated calculus $\name$,
+including its relations to existing linear and modal calculi.
+As part of this, I have adapted what I call \emph{bunched connectives} from
+\cite{RPKV20} as a way to state the typing rules of the calculus as well as to
+work with the metatheory.
+The distinction between sharing and separating conjunction given by the bunched
+connectives corresponds well to the notions of \emph{additive} and
+\emph{multiplicative} connectives in the object language, respectively.
+Following this, I have given a novel linear algebra-based definition of
+\emph{environment} for semiring-annotated calculi, together with a motivation
+which may serve as a basis for the corresponding definition for other
+substructural systems.
+The adequacy of this definition of environment is shown by my implementation of
+simultaneous renaming and substitution, as well as other operations on
+environments, like composition of renamings and substitutions.
 
-FIXME: this chapter is directly from the ESOP22 paper.
+With the details of $\name$ worked out, I then moved on to adapting the work of
+\citet{AACMM21} so as to make it able to capture semiring-based usage
+restrictions, as found in $\name$.
+The syntax descriptions of the resulting system are based on the bunched
+connectives, and are shown to be expressive enough to encode calculi of a
+variety of forms.
+In adapting the semantics, I am forced to be precise about \emph{sharing} and
+\emph{separating} bunched connectives, but largely add these as a refinement of
+the work of \citet{AACMM21}.
+I provide the renaming and substitution operations for all expressible calculi.
+I also provide more specialised examples of semantic traversals:
+a usage elaborator, an NbE algorithm, a denotational semantics, and translations
+between different calculi.
+The usage elaborator gives an unexpected example of generic programming, which
+one could not write without syntax descriptions.
+% Other examples...
+Together, these examples show the applicability and versatility of the framework
+I have developed.
+
+FIXME: the next two paragraphs are directly from the ESOP22 paper.
 
@@
   \item Is this a reusable library? Yes, examples
-  \item Fitch-style systems (can't do).
   \item Methodology for metatheory of new calculi
@@
 
-\section{Addendum: (lack of) partiality}\label{sec:part}
+\section{Future work}
+
+\begin{itemize}
+  \item Fitch-style systems (can't do).
+\end{itemize}
+
+\subsection{Further work}
+
+\paragraph{Equality}
+Perhaps the most fundamental missing piece from the metatheoretic account of
+semiring-annotated calculi I have given in this thesis is equations between
+terms.
+Reasoning about equality between terms and environments is a problem I have
+tried to solve, but I have not arrived at a satisfactory solution in the time
+available to me.
+
+I believe that the basic difficulty of giving an account for equality in a
+linear setting is the proliferation of $\Sigma$-types.
+For example, describing equality between two applications of
+\TirName{$\with$-I} is immediate: $\Gamma \vdash (M, N) = (M', N') : A \with B$
+if and only if $\Gamma \vdash M = M' : A$ and $\Gamma \vdash N = N' : B$.
+However, to do the same with \TirName{$\otimes$-I} requires us to be careful
+about the contexts of the subterms.
+The two applications of \TirName{$\otimes$-I} may a priori split the context in
+different ways, and should only be equated when those splittings are equal
+(in the appropriate sense).
+If the splittings are equal, then the contexts of the subterms will line up, and
+only then can the subterms themselves be compared for equality.
+These multiple stages come about because $\plr{T * U}\,\Gamma$ is a
+$\Sigma$-type, and equality of elements of $\Sigma$-types always follows this
+pattern.
+Such reasoning becomes even more complex for environments, which are equivalent
+to iterated $*$-families.
+Additionally, it is unclear what effect subsumption of contexts (like subusaging
+in this thesis or explicit structural rules in other calculi) should have on
+equality, particularly when the subsumption commutes with parts of the subterms.
+
+Previous work on linear logic has preferred to deal with equations only on
+untyped terms. \todo{Check this.}
+
+\paragraph{Polymorphism}
+An important feature of most contemporary statically typed programming languages
+is polymorphism.
+In particular, parametric polymorphism over types can be used to significantly
+improve code reuse, and is well undestood theoretically via System $F$ and its
+variants.
+I have not considered polymorphism in this thesis, and neither did
+\citet{AACMM21} in their paper, so whether it can be supported in the framework
+presented earlier is an open question.
+However, \citet{Autosubst15} did apply their related system to System $F_{<:}$,
+so it may be possible to support polymorphic systems using their techniques.
+
+A separate but related question concerns polymorphism over usage annotations.
+The status of polymorphism over usage annotations is less well established both
+in practice and in theory.
+\Citet{Granule18} present an implementation allowing for polymorphism of usage
+annotations, and even polymorphism over semirings, but provide no more than
+example programs to justify the feature.
+This thesis provides no advance on understanding annotation polymorphism, unless
+it can be encoded into a semiring to fit the framework.
+
+%\subsection{Further generalisations}
+
+\paragraph{Structure of contexts}
+As I have presented it, the work of \citet{AACMM21} has two axes in which it is
+generic: the syntax, which can be controlled through syntax descriptions to
+produce a wide range of calculi and features; and the semantics, where we can
+produce a wide range of maps out of terms with the help of environments.
+To this, the work of this thesis has added a third axis of genericity: the usage
+discipline of variables, as described by a partially ordered semiring.
+
+Starting at least with the bunched connectives in \cref{sec:lnd}, if not earlier
+when talking about usage contexts forming modules over the semiring of
+annotations, I have made productive use of abstractions over the basic usage
+annotations throughout this thesis.
+These abstractions suggest a next step of completely abstracting away much of
+the representation of contexts and their individual entries.
+One may imagine that it is possible to develop a framework in which the required
+operations and properties of contexts are axiomatised, similar to how usage
+annotations are axiomatised to form a semiring in this thesis, and to how
+categories-with-families models are defined~\citep{Dybjer95,CCD19}.
+Instances of such a framework would include the work of \citet{RPKV20}, which
+uses a very similar bunched connective abstraction over a very different
+representation of contexts, based on relational interleaving of lists.
+
+The use of semirings is motivated in this thesis and elsewhere largely because
+they are general enough to cover a wide range of examples.
+However, I cannot claim to have a derivation from first principles of why we
+should choose partially ordered semirings over any of a range of similar
+algebraic structures.
+Additionally, some of the specific constructions done in this thesis fit
+somewhat unnaturally with the semiring-based approach.
+For example, when translating semiring-annotated systems to traditional systems,
+I tended to need to make a \emph{bottom-up} assumption
+(\cref{def:DILL-bottom-up,def:PD-bottom-up}) so as to avoid some ``junk'' facts
+given by the semiring.
+Meanwhile, the usage elaborator of \cref{sec:usage-elaborator} eschews the
+``forward'' computation of semiring operations in favour of non-deterministic
+backwards computation, e.g., from a sum to the collection of possible summands.
+Possibly consciously working more abstractly, as described in the previous
+paragraph, would make a more natural structure appear.
+
+If we are to retain an annotation-based approach to usage restrictions, then a
+possible feature request that falls out of the encoding of linear/non-linear
+logic is to have some sort of kinding system by which different kinds of types
+are annotated using different sets of annotations.
+In the L/nL example, we would want linear types to be annotated with $\gr0$ and
+$\gr1$, and intuitionistic types to be annotated with $\gr\omega$ (as the sole
+element of a trivial instance of an algebraic structure), with no crossover
+between the two kinds.
+Algebraic means to handle such mixed-kind usage vectors may be inspired by the
+work of \citet{Hart95,MF21} on dimensional analysis in linear algebra.
+
+\subsubsection{Partiality}\label{sec:part}
 \input{part.tex}

### 2023-07-12 tex/examples.tex
[ writing ] various up to WRel example

@@
 
+\section{Normalisation by evaluation}\label{sec:nbe}
+
+To give an example of a common operation on programming languages, I present a
+normalisation by evaluation (NbE) algorithm~\citep{BS91} for a fragment of
+$\name$.
+The algorithm contains not only a structural induction on a term using an
+environment, but also functions which create terms.
+It therefore gives an example of various parts of the framework.
+
+Though it is easy to define a syntax of normal and neutral forms, I do not do so
+here.
+Converting between terms and normal/neutral forms introduces significant
+overheads, and the syntax of normal/neutral forms is somewhat complicated by the
+invariant that all variables are of neutral type.
+It would be worthwhile future work to work out this refinement, thereby showing
+that the normaliser presented in this section actually puts terms into normal
+form.
+Another limitation is that I do no reasoning about equivalence of terms, so
+there is no proof that the resultant term is equivalent to the original.
+
+For the sake of simplicity and conciseness, I consider just the fragment
+containing a unary function type, but whose argument carries a usage annotation.
+The partially ordered semiring of usage annotations is arbitrary throughout.
+I also include an arbitrary set of base types \AgdaBound{BaseTy} via constructor
+\AgdaInductiveConstructor{$\iota$}.
+
+\ExecuteMetaData[\NbESimpletex]{Ty}
+
+With only function types, the only term formers are application and
+$\lambda$-abstraction.
+These are similar to the corresponding term formers in $\name$, but application
+scales its argument in accordance with the usage annotation of the function
+type, and $\lambda$-abstraction binds a variable with the given usage
+annotation.
+For convenience, I define an alias \AgdaFunction{Term} for the open family of
+terms in this type system of unspecified (\AgdaPostulate{$\infty$}) size.
+
+\ExecuteMetaData[\NbESimpletex]{system}
+
+The NbE model of types is given by \AgdaFunction{\_$\vDash$\_}.
+It is defined by induction on the type.
+The base type is interpreted by terms of base type, while
+the function type is interpreted by the Kripke function space (essentially a
+special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$, though reusing
+the \AgdaFunction{Kripke} definition causes difficulty for the termination
+checker).
+The interpretation of the base type additionally requires a \AgdaRecord{Lift}
+of its universe level, because \AgdaFunction{$\Box^r$} produces a large type
+(a type in \AgdaPrimitive{Set$_1$}).
+The largeness of \AgdaFunction{$\Box^r$} comes from the largeness of
+environments, and hence renamings, because of the relationally specified linear
+map.
+
+\ExecuteMetaData[\NbESimpletex]{vDash}
+
+\AgdaFunction{\_$\vDash$\_} is a renameable family, as can be seen by cases on
+the type.
+It is renameable at the base type because terms are renameable, and it is
+renameable at function type because any open type formed using
+\AgdaFunction{$\Box^r$} is renameable by composition of renamings.
+I refer to this proof of renameability as
+\AgdaFunction{ren\textasciicircum$\vDash$}.
+
+Now that I am working with a specific syntax, I introduce the function
+\AgdaFunction{reify[]}.
+This is a variant of the \AgdaFunction{reify} function given in \cref{sec:reify}
+usable only for rules that bind no new variables --- i.e., the parameter
+\AgdaBound{$\Delta$} to \AgdaFunction{Kripke} is the empty context.
+
+\begin{lemma}[\AgdaFunction{reify[]}]
+  We have a function of type
+  $\mathrm{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any open
+  families $\V$ and $\C$.
+\end{lemma}
+\begin{proof}
+  Let $b : \mathrm{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
+  We give $b$ the identity renaming (corresponding to no extension of the
+  context), yielding
+  $b\,1 : \plr{\plr{-} \env\V {\cdot} \wand \plr{-} \sdtstile{}\C A}\,\Gamma$.
+  By \cref{thm:construct-env}, we have that $\plr{-} \env\V {\cdot}$ is
+  equivalent to $I^*$, and by monoidal closure,
+  $\plr{I^* \wand \plr{-} \sdtstile{}\C A}$ is equivalent to
+  $\plr{-} \sdtstile{}\C A$, so the result follows.
+\end{proof}
+
+The first part of the NbE algorithm I give is the evaluator.
+The evaluator interprets terms into the model via an environment of values from
+the model, making it a clear instance of the generic semantic traversal.
+Specifically, I fix both \AgdaBound{$\V$} and \AgdaBound{$\C$} to
+\AgdaFunction{\_$\vDash$\_} (allowing the conversion from semantic values to
+semantic terms to be given by the identity function \AgdaFunction{id}), and
+provide the other required data as below.
+The case for $\lambda$-abstraction is almost trivial, given that its job is to
+convert the Kripke function \AgdaBound{M} coming from the generic traversal into
+a value in the model of function type (i.e.\ a Kripke function).
+The main thing we need to do is to wrap the argument to the function into a
+singleton environment using \AgdaFunction{[-]$^e$}.
+In the case for function application, we use \AgdaFunction{reify[]} to get rid of
+the extraneous Kripke functions coming from the generic traversal.
+Having done so for both subterms, we apply the value corresponding to
+\AgdaBound{M} (which is a Kripke function) to the value corresponding to
+\AgdaBound{N} in an unextended context (\AgdaFunction{1$^r$}), with usage
+information divided up as in the original application.
+
+\ExecuteMetaData[\NbESimpletex]{evalSem}
+
+The function \AgdaFunction{eval} is given by the appropriate instantiation of
+the generic traversal \AgdaFunction{semantics}.
+Note that we have not yet shown that \AgdaFunction{\_$\vDash$\_} supports
+identity environments (which requires the reify and reflect functions), so we
+cannot yet evaluate arbitrary open terms.
+
+\ExecuteMetaData[\NbESimpletex]{eval}
+
+The final pieces of NbE are the \emph{reify} and \emph{reflect} functions.
+I give these the names \AgdaFunction{nbeReify} and \AgdaFunction{nbeReflect}
+respectively to disambiguate from the \AgdaFunction{reify} and
+\AgdaFunction{reify[]} functions of the framework.
+\AgdaFunction{nbeReify} and \AgdaFunction{nbeReflect} are defined mutually by
+recursion on the type, as is standard in NbE.
+Therefore, neither uses the generic semantic traversal, though both make use of
+some of the general components I have developed in this thesis.
+
+At base type, values in the model are terms (modulo \AgdaRecord{Lift}), so the
+conversion between these and terms is trivial.
+At function type, \AgdaFunction{nbeReify} is given a Kripke function
+\AgdaBound{v} and aims to turn it into a normal form of function type.
+It does so in the canonical way --- introducing a $\lambda$-abstraction --- and
+then aims to produce a term in an extended context.
+Because of the context extension, we apply \AgdaBound{v} using the renaming
+\AgdaFunction{$\swarrow^r$}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaFunction{++$^c$}\AgdaSpace{}%
+\AgdaFunction{[}\AgdaSpace{}\AgdaField{0\#}\AgdaSpace{}\AgdaFunction{$\bullet$}%
+\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}\AgdaFunction{]$^c$}
+to get it ``back'' to the original context.
+The splitting \AgdaBound{sp+} adds together $\Gamma, \gr0A$ and
+$\gr0\Gamma, \gr rA$ to get $\Gamma, \gr rA$, while \AgdaBound{sp*} scales down
+$\gr0\Gamma, \gr rA$ to $\gr0\Gamma, \gr1A$.
+Semantic value \AgdaBound{v} then wants a semantic value of type $A$, which it
+gets from reflecting the newly bound variable into the model.
+Finally, \AgdaBound{v} produces a semantic value, which a recursive call to
+\AgdaFunction{nbeReify} at the smaller type \AgdaBound{B} converts into a term
+(in fact, a normal form).
+Meanwhile, \AgdaFunction{nbeReflect} at function type is given a syntactic
+function (in fact, a neutral form) and aims to turn it into a Kripke function.
+To produce a Kripke function, we consider a renaming \AgdaBound{$\rho$} into
+\AgdaBound{$\Gamma$}, splittings of the resultant context, and a semantic value
+\AgdaBound{N} of type \AgdaBound{A}.
+To produce the value of type \AgdaBound{B}, we recursively call
+\AgdaFunction{nbeReflect} at smaller type \AgdaBound{B}, but with a larger term
+formed by reifying \AgdaBound{N} (again of smaller type) and applying it to the
+term \AgdaBound{M}.
+To make the application work, we must rename both \AgdaBound{M} and
+\AgdaBound{N} to ignore the (empty) context extensions introduced by the
+application rule, and we must also rename \AgdaBound{M} by \AgdaBound{$\rho$} to
+put \AgdaBound{M} into the extended context the reflection is being done at.
+
+\ExecuteMetaData[\NbESimpletex]{reify-reflect}
+
+Finally, we compose \AgdaFunction{eval}, \AgdaFunction{nbeReify}, and
+\AgdaFunction{nbeReflect} to get the normaliser \AgdaFunction{nbe}.
+The definition of \AgdaFunction{nbe} uses \AgdaFunction{eval} to interpret the
+term into the model, and then uses \AgdaFunction{nbeReify} to read the semantic
+value back as a term (a normal form).
+However, as noted earlier, we need to give an environment of type
+$\Gamma \env\vDash \Gamma$ to \AgdaFunction{eval} in order to interpret an open
+term in its own context.
+I provide this environment via \cref{thm:env-id}, which says that if we have a
+mapping from variables in $\Gamma$ to values in $\Gamma$, we can get the desired
+identity environment.
+The required map is given in
+\AgdaFunction{identityEnv\textasciicircum{}$\vDash$}
+by turning the variable into a term using \AgdaInductiveConstructor{`var} and
+then applying \AgdaFunction{nbeReflect}.
+The definition \AgdaFunction{identityEnv\textasciicircum{}$\vDash$} is marked as
+an \AgdaKeyword{instance} so that it is picked up by
+\AgdaFunction{id\textasciicircum{}Env}.
+
+\ExecuteMetaData[\NbESimpletex]{nbe}
+
+The main limitation of this example is that the results are not specified as
+being in normal form by their type.
+Ideally, we would have a syntax of normal and neutral forms for
+\AgdaFunction{nbe} and supporting functions to target.
+We can produce such a syntax --- similarly to how \citet{AACMM21} produce a
+syntax for bidirectional type checking --- but this syntax is difficult to work
+with because of the crucial restriction on the allowable kind of variables ---
+namely that variables can only appear as neutral forms, and not normal forms.
+I believe that this difficulty could be worked around via carefully placed
+assertions that various contexts are well formed --- and in particular that the
+extended context introduced by a \AgdaFunction{$\Box^r$} is well formed.
+
+Also, I have not given an account of many of the connectives of $\name$.
+Of particular interest would be the types with pattern-matching eliminators ---
+$I$, $\otimes$, $0$, $\oplus$, and $\oc$.
+I believe these could be handled by ordinary techniques for NbE for sums ---
+probably not doing $\eta$-expansion for these types.
+
 \section{A denotational semantics}\label{sec:den-sem}
@@
 family.
+For a simple denotational semantics, we set $\V$ to be variables
 
-{\color{red}Old stuff below\ldots}
-
-To justify the name \emph{semantics}, we give an example traversal that is a
-denotational semantics in the usual sense.
-The semantics we take is a refinement of that of \citet{AbelBernardy2020},
+As an example denotational semantics for a semiring-annotated calculus, I take
+a semantics in world-indexed relations.
+This is a refinement of the semantics given by \citet{AbelBernardy2020},
 which gives a way to extract parametricity theorems from substructurally typed
@@
 
-To abbreviate this section, we use a simplified syntax compared to \name{}.
-We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
-type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
-\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.
+For the sake of brevity, I use the same term language as in \cref{sec:nbe}, with
+a set of base types and function types with a usage-annotated domain type.
 
-\ExecuteMetaData[\WReltex]{Ty}
+As a running example, I take the usage annotations to be the 4-element
+variance posemiring (\cref{def:monotonicity-posemiring}).
+I establish the property that all terms are monotonic in their free variables.
+This monotonicity can be covariant or contravariant (or neither or both)
+depending on the annotation of each free variable (respectively,
+\AgdaInductiveConstructor{$\uparrow\uparrow$},
+\AgdaInductiveConstructor{$\downarrow\downarrow$},
+\AgdaInductiveConstructor{$\wn\wn$}, and
+\AgdaInductiveConstructor{\textasciitilde\textasciitilde}).
+This provides an additional example to those of \citet{AbelBernardy2020}.
 
-In the term syntax, $\lambda$-abstraction now binds a variable with annotation
-\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
-(both in accordance with the function type they are acting on).
-
-\ExecuteMetaData[\WReltex]{AnnArr}
-
-As a running example, we take the usage annotations to be the 4-element
-variance posemiring (\cref{thm:variance}).
-We establish the property that all terms are monotonic in their free variables.
-This monotonicity can be covariant or contravariant (or neither or both)
-depending on the annotation of each free variable.
-This provides an additional example to those of \citeauthor{AbelBernardy2020}.
-
-We will take semantics of this system into
-\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
+In the semantics of this section, types are interpreted as
+\emph{world-indexed relations}~\citep{AbelBernardy2020,context-constrained}.
 A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
-\AgdaBound{W} is a set over which
-we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
-property with respect to the order on \AgdaBound{W}.
+\AgdaBound{W} is a set \AgdaField{set} over which
+we have a \AgdaBound{W}-indexed binary relation \AgdaField{ren} satisfying a
+presheaf-like property \AgdaBound{subres} with respect to the order on
+\AgdaBound{W}.
+I declare the record type \AgdaRecord{WRel} to have
+\AgdaKeyword{no-eta-equality} so that the usual $\eta$-law of records does not
+apply to it.
+The lack of an $\eta$-law allows the connectives on \AgdaRecord{WRel}
+(e.g.\ \AgdaFunction{\_$\otimes^R$\_} and \AgdaFunction{\_$\multimap^R$\_},
+defined below) to be definitionally injective, helping the unification in Agda's
+elaborator.
 
@@
 Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
-and \AgdaBound{S} consist of a mapping between the underlying sets such that, at
-each fixed world \AgdaBound{w}, the mapping preserves relatedness from
-\AgdaBound{R} to \AgdaBound{S}.
+and \AgdaBound{S} consist of a mapping \AgdaField{set$\Rightarrow$} between the
+underlying sets such that, at each world \AgdaBound{w}, the mapping preserves
+relatedness from \AgdaBound{R} to \AgdaBound{S}.
+In the type of the field \AgdaField{rel$\Rightarrow$}, the world index is
+handled implicitly via \AgdaFunction{\_$\rightarrowtriangle$\_}.
 
@@
 \AgdaRecord{$\wand$}, now over worlds rather than contexts.
+Their definitions are listed below.
+
+\ExecuteMetaData[\WReltex]{IR}
+\ExecuteMetaData[\WReltex]{tensorR}
+\ExecuteMetaData[\WReltex]{lollyR}
 
 The final piece of semantics we need is a \emph{bang} operator.
-We allow the
+I allow the
 semantic \emph{bang} to be an arbitrary annotation-indexed functor on
@@
 \ExecuteMetaData[\WReltex]{Bang}
+
+The properties required of the semantic \emph{bang} operator are slightly
+weaker than those given by \citet{AbelBernardy2020}.
+Specifically, the morphisms given by \AgdaField{!$^R$-1}, \AgdaField{!$^R$-I},
+and \AgdaField{!$^R$-$\otimes$} only go one way, rather than being
+bi-implications as they are for \citet{AbelBernardy2020}.
+\citeauthor{AbelBernardy2020} need the stronger laws because of their strong
+eliminator for tensor products, discussed in \cref{sec:alt}.
+Also, I do not use the \emph{non-idempotent intersection} operators (which
+handle worlds in the same way as \AgdaFunction{I$^R$} and
+\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, but keep the
+underlying set the same, rather than using products of the underlying sets),
+and thus do not have axioms for them.
+I avoid the need for non-idempotent intersections by giving the semantics all at
+once, rather than first giving a $\mathrm{Set}$-semantics and then giving a
+$\mathrm{WRel}$-semantics on top of it.
 
@@
   predicate on worlds \AgdaFunction{$\dot1$}.
+  With the worlds being trivial, so too is the property \AgdaField{subres}.
 
@@
 
-The semantics of a type is given by \AgdaFunction{$\llbracket$\_$\rrbracket$},
-which maps into world-indexed relations.
-The function type is interpreted using \AgdaFunction{$\multimap^R$} and
+Let us assume that we have a mapping
+\AgdaBound{$\upiota\llbracket$\_$\rrbracket$} from base types to world-indexed
+relations.
+We extend this interpretation to all types via
+\AgdaFunction{$\llbracket$\_$\rrbracket$}, shown below, with
+the function type being interpreted using \AgdaFunction{$\multimap^R$} and
 \AgdaField{!$^R$}.
 Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
-\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}.
+\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}, and \AgdaField{!$^R$}
+for singleton contexts.
 Terms are interpreted as morphisms by the open family
@@
 
+\ExecuteMetaData[\WReltex]{sem}
+\ExecuteMetaData[\WReltex]{sem-vdash}
 \ExecuteMetaData[\WReltex]{lookupR-type}
 
-Now we give a \AgdaRecord{Semantics}.
+The laws \AgdaField{!$^R$-0}, \AgdaField{!$^R$-+}, and \AgdaField{!$^R$-*} lift
+from singleton contexts to all contexts via lemmas \AgdaFunction{ctx-0},
+\AgdaFunction{ctx-+}, and \AgdaFunction{ctx-*}, with the latter saying that we
+have morphisms of type
+\mbox{$\sem{\plr{\gr r\grP}\gamma} \to \oc^R\gr r\sem{\grP\gamma}$}.
+We can see each of these three lemmas as ways to turn the algebraic structure of
+contexts into structured world-indexed relations --- allowing us to use general
+facts about \AgdaFunction{I$^R$}, \AgdaFunction{\_$\otimes^R$\_}, and
+\AgdaField{!$^R$}, respectively.
+
+Now I give a \AgdaRecord{Semantics}.
 The choice of \AgdaBound{$\V$} as
@@
 environment entries by \AgdaFunction{lookup$^R$}.
+
 Meanwhile, for the logical rules, we ignore environments by using
-\AgdaFunction{reify} to just deal with morphisms in an extended context.
-As such, $\lambda$-abstractions are easy to interpret, while applications
-require some massaging to remove the extension by an empty context, followed by
-some plumbing to split the interpretation of the context according to the usage
-constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
-the interpretation of the function \AgdaBound{m$'$}.
+\AgdaFunction{reify} and \AgdaFunction{reify[]} to just deal with morphisms in
+an extended context.
+The crucial structure of world-indexed relations is given by lemmas
+\AgdaFunction{curry$^R$} and \AgdaFunction{uncurry$^R$}, which translate between
+\mbox{\AgdaRecord{WRelMor}\AgdaSpace{}\AgdaSymbol(\AgdaBound{R}\AgdaSpace{}%
+\AgdaFunction{$\otimes^R$}\AgdaSpace{}\AgdaBound{S}\AgdaSymbol)%
+\AgdaSpace{}\AgdaBound{T}} and
+\mbox{\AgdaRecord{WRelMor}\AgdaSpace{}\AgdaBound{R}\AgdaSpace{}\AgdaSymbol(%
+\AgdaBound{S}\AgdaSpace{}\AgdaFunction{$\multimap^R$}\AgdaSpace{}\AgdaBound{T}%
+\AgdaSymbol)}.
+Also, \AgdaFunction{map-$\otimes^R$} implements the tensor
+product of world-indexed relation morphisms.
+With these lemmas, the interpretations of $\lambda$-abstraction and application
+are straightforward.
+For $\lambda$-abstraction, we have
+\mbox{$\sem{\Gamma, \gr rA} = \sem\Gamma \otimes^R \oc^R\gr r\sem{A}$},
+so we just need to use \AgdaFunction{curry$^R$} to get the desired semantic
+function in the original context.
+For application, we are producing the composition pictured in
+\cref{fig:Wrel-app}.
+In this picture, I ignore the possible subusaging between $\grP + \gr r\grQ$ and
+the original context for the sake of notational suggestiveness.
 
 \ExecuteMetaData[\WReltex]{Wrel}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node(0) {$\sem{\plr{\grP + \gr r\grQ}\gamma}$};
+    \node(1) [right=of 0] {$\otimes$};
+    \node(2) [above=1em of 1] {$\sem{\grP\gamma}$};
+    \node(3) [below=1em of 1] {$\sem{\gr r\grQ\gamma}$};
+    \node(4) [right=of 3] {$\oc\gr r\sem{\grQ\gamma}$};
+    \node(5) [right=of 4] {$\oc\gr r\sem{A}$};
+    \node(6) [above=1em of 5] {$\otimes$};
+    \node(7) [above=1em of 6] {$\sem{\grP\gamma}$};
+    \node(8) [right=of 6] {$\sem{B}$};
+
+    \draw[->] (0) to node[above] {\AgdaFunction{ctx-+}} (1);
+    \draw[->] (2) to node[above] {\AgdaFunction{id$^R$}} (7);
+    \draw[->] (3) to node[below] {\AgdaFunction{ctx-*}} (4);
+    \draw[->] (4) to node[below] {\AgdaBound{n}} (5);
+    \draw[->] (6) to node[above] {\AgdaBound{m}} (8);
+  \end{tikzpicture}
+  \caption{Interpretation of application in the world-indexed relation
+    semantics}
+  \label{fig:Wrel-app}
+\end{figure}
 

### 2023-07-12 tex/gen-sem.tex
[ writing ] various up to WRel example

@@
 other than terms, and act in an interesting way on term constructors.
-Doing a straight generalisation of the type of \AgdaFunction{trav}, this
-suggests that we want a function with the following type, where
+\Citet{ACMM17} show how to do this generalisation of \AgdaFunction{trav} to a
+function \AgdaFunction{sem} with the following type, where
 \AgdaBound{$\C$} is the type family we are targeting.
+In this section, I discuss the assumptions needed to implement such a function.
 
@@
 However, we want to treat binding constructs specially, particularly because
-there are semanticses with no notion of binding.
+there are semantics with no notion of binding.
 We instead provide a function from values to computations that works in any

### 2023-07-12 tex/gen-syn.tex
[ writing ] various up to WRel example

@@
 
-I will present a scheme such that \cref{fig:app-lam} is interpreted as the type
+I will present a scheme based on the work of \citet{AACMM21} such that
+\cref{fig:app-lam} is interpreted as the type
 system we studied in the previous sections (simply typed $\lambda$-caluclus with

### 2023-07-12 tex/intro.tex
[ writing ] various up to WRel example

@@
 \chapter{Introduction}
+
+This thesis advances the frontier of programming languages work that can be done
+in a proof assistant.
+I will elaborate on both of these aspects in the following paragraphs.
+
+The main aim of programming language research is to find common patterns in
+computer programs and represent these patterns via programming language
+features.
+As a basic example, consider functions in, say, a low level language like C, in
+contrast to what these functions are compiled to in assembly language/machine
+code.
+In order to support function calls, the C runtime system manages a
+\emph{call stack}.
+A function call is compiled to a sequence of instructions which store the return
+address and the values of the arguments on the stack, move the stack pointer,
+and jump to the compiled code of the function.
+A return from a function then places the return value into the appropriate
+register, moves the stack pointer back, and jumps back to the return address.
+Advances in programming language techniques are evaluated by the usefulness of
+behaviours captured and goodness of the mathematical properties of the
+abstractions produced.
+
+Within programming language research, type theory is a methodology for producing
+abstractions.
+A type system lets us enforce in our language complex invariants which maintain
+abstraction boundaries while also ideally being machine-checkable.
+For example, we give functions function type to abstract away their
+implementation via call stacks and jumps and so on.
+
+In this thesis, the invariants of interest revolve around restricting the usage
+of variables.
+I introduce this topic thoroughly in \cref{sec:linear}, but in brief\ldots
+
+The work of this thesis relies upon type theory in two distinct ways.
+Firstly, the main objects of study are programming languages with interesting
+type systems.
+Secondly, type theory provides the basis of the proof assistant Agda I use to
+implement the aforementioned programming languages and operations upon them.
+
+As for the other topic of this thesis,
+a \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
+a piece of software that allows for the encoding of mathematical definitions,
+theorems, constructions, and proofs, and furthermore check that such encoded
+proofs are correct and that such encoded constructions are well formed.
+To truly by interactive, i.e.\ to actually assist, a proof assistant will
+usually have a user interface which can read partial proofs, display
+information about what more proof needs to be given, and provide actions that
+will help complete the proof.
+
+Proof assistants have seen increasing use in programming language research in
+recent years.
+The most obvious reason why working in a proof assistant is seen as beneficial
+is that it ensures correctness.
+If the proof assistant accurately implements a suitable mathematical foundation,
+then any theorem proved in the proof assistant is guaranteed to be a true
+theorem of that foundation.
+These guarantees of correctness are particularly important when working with
+combinatorially complex mathematical objects, proofs about which often require
+the consideration of a large number of cases.
+Programming language syntaxes are often such complex objects, motivating the use
+of proof assistants when studying programming languages.
+
+A second reason to use proof assistants is for the assistance they provide when
+exploring a mathematical theory.
+When we make a new definition, we may want to test how it works in a special
+case, or what constructions it allows us to perform.
+In a proof assistant, the assistance tools give us immediate feedback as to what
+moves are and aren't allowed.
+For example, if we define a complex type system, a proof assistant will let us
+interactively build typing derivations, making clear any side conditions and
+types of subderivations as we go.
+Also, as I do later in this thesis, a proof assistant allows us to build a very
+general theory, and practically use that theory directly in more specific cases
+without losing rigour.
+
+Thirdly, analogously to how a strong static type system can give us more
+confidence when refactoring a program, the constant checking of proofs in a
+proof assistant gives us the confidence to change definitions and lemmas knowing
+that we will be guided towards the parts of our theory that need to be
+correspondingly changed.
+This can help if we are developing a new programming language with a changing
+specification.
+
+Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
+this thesis --- double as programming languages themselves.
+This means that we can write progams and prove properties of them using the same
+tool.
+Also, many theorems proven in such a proof assistant have computational content.
+For example, if we prove a normalisation theorem for a programming language,
+this will typically yield a (verified) normalisation algorithm for it, which we
+can really run on a computer.
+As such, a development in a proof assistant can provide a reference
+implementation of a programming language, or even --- as with Idris
+2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
+actual implementation of a programming language.
 
@@
 
-\section{Naming and notation conventions}
-
-I assume familiarity with the Curry-Howard correspondence throughout this
-thesis.
-I make no distinction between logics and type theories, and use terminology from
-each interchangeably.
-Each following bullet point lists a collection of synonyms.
-
-\begin{itemize}
-  \item assumption, hypothesis, variable
-  \item derivation, proof, term
-  \item proposition, formula, type
-  \item connective, type former
-\end{itemize}
-
-I carry out mechanised constructions and proofs in the proof assistant and
-programming language Agda~\citep{Agda}.
-Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
-similarly present non-mechanised constructions and proofs assuming a foundation
-given by dependent type theory, in a style inspired by the HoTT
-Book~\citep{hottbook}.
-
 \section{Outline of the thesis}
@@
 traversal.
-\Cref{sec:examples} then gives further example uses of the generic traversal.
+\Cref{sec:example-semantics} then gives further example uses of the generic
+traversal.
 
@@
 this thesis and openings for future work.
+
+\section{Naming and notation conventions}
+
+I assume familiarity with the Curry-Howard correspondence throughout this
+thesis.
+I make no distinction between logics and type theories, and use terminology from
+each interchangeably.
+Each following bullet point lists a collection of synonyms.
+
+\begin{itemize}
+  \item assumption, hypothesis, variable
+  \item derivation, proof, term
+  \item proposition, formula, type
+  \item connective, type former
+\end{itemize}
+
+I carry out mechanised constructions and proofs in the proof assistant and
+programming language Agda~\citep{Agda}.
+Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
+similarly present non-mechanised constructions and proofs assuming a foundation
+given by dependent type theory, in a style inspired by the HoTT
+Book~\citep{hottbook}.
+
+\section{Agda primer}
+
+I use the proof assistant and programming language Agda throughout this thesis,
+with Agda code being used particularly in \cref{sec:simple} and the later
+chapters.
+As such, it is important for the reader to be able to read basic Agda syntax in
+order to benefit from the parts of the exposition that reside in code listings.
+The syntax of Agda is broadly similar to that of Haskell, and relatively close
+to that of Standard ML, OCaml, and Coq's Gallina sublanguage.
+I will assume that the reader is able to read basic Haskell code, and spend most
+time explaining differences thereof.
+
+\subsection{Lexing and parsing}
+
+Agda is extremely liberal in its set of allowed names.
+There is just a single lexical class (unlike in Haskell, where, for example,
+constructors start with a capital letter and definitions start with a lowercase
+letter), and names can be any string of Unicode characters except whitespace and
+special characters \verb|.;{}()@"|, apart from those strings reserved as
+keywords or literals.
+Therefore, we can introduce names like
+\AgdaFunction{0x-+-$\uplambda\rightarrow$}
+to stand for any kind of identifiable thing.
+With such free-form names, ample spacing is required between identifiers.
+For example, while \AgdaNumber{0}\AgdaSpace{}\AgdaDatatype{$\leq$}\AgdaSpace{}%
+\AgdaNumber{1} is a possible expression containing three
+identifiers, \AgdaInductiveConstructor{0$\leq$1} is a single
+valid identifier.
+Only the special characters may appear next to names without being separated by
+whitespace.
+
+A character with unique behaviour in Agda's syntax is the underscore
+(\AgdaSymbol{\_}).
+Within a name, an underscore signifies that the name will function as a mixfix
+operator, allowing for an argument in the position of the underscore.
+For example, the full name of the \AgdaDatatype{$\leq$} operator used in the
+previous paragraph is \AgdaDatatype{\_$\leq$\_}, signifying that it can take an
+argument to its left and its right.
+We can also introduce closed operators, like \AgdaDatatype{[\_]}, which can take
+an argument between the square brackets (e.g.\ \AgdaDatatype{[}\AgdaSpace{}%
+\AgdaNumber{1}\AgdaSpace{}\AgdaDatatype{]}, with spaces still being important).
+Mixfix operators can be partially applied by leaving underscores in the name in
+the application.
+For example, \AgdaDatatype{\_$\leq$}\AgdaSpace{}\AgdaNumber{1} could be the
+predicate asserting that a number is less than or equal to 1.
+
+On its own, an underscore has a completely different meaning, which can depend
+on context.
+In patterns, an underscore has the same meaning as it has in Haskell and ML ---
+it holds the place of a pattern variable, but does not name that variable.
+In expressions, an underscore stands for an unspecified subterm which will be
+solved by unification.
+The solving of unspecified terms is canonical and respects $\beta\eta$-equality,
+unlike in Coq.
+
+Spacing is important particularly important when dealing with underscores.
+For example,
+\AgdaDatatype{\_$\leq$}\AgdaSpace{}\AgdaSymbol{\_} (with a space after the
+\AgdaDatatype{$\leq$} but not before) standing for the predicate asserting that
+a number is less than or equal to some unspecified number.
+
+Like Haskell, Agda's syntax is indentation-sensitive.
+The distinctions conveyed by indentation are largely obvious or intuitive to
+human readers (for example, allowing for line-continuation or delineating nested
+modules), so I will not discuss them explicitly here.
+
+\subsection{Functions, $\Pi$-types}\label{sec:pi-types}
+
+Simple function types take the form
+\AgdaArgument{A}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{B},
+coinciding with Haskell's syntax.
+Also as in Haskell and ML, the function arrow nests to the right.
+However, Agda has a termination checker ensuring that all defineable functions
+are total, so many Haskell functions do not have a corresponding Agda function.
+
+The key feature distinguishing Agda from Haskell is the presence of arbitrary
+dependent types, including dependent function types ($\Pi$-types).
+The basic syntax for $\Pi$-types is
+\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{A}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaArgument{B},
+where variable \AgdaBound{x} can occur free in expression \AgdaArgument{B}.
+However, there are several syntactic conveniences I use throughout the code
+listings.
+For one, iterated $\Pi$-types can be abbreviated so that
+\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{A}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaBound{y}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{B}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaArgument{C}
+is written just
+\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{A}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaBound{y}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{B}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaArgument{C},
+omitting the first arrow.
+For another, prefixing an arrow with the \AgdaSymbol{$\forall$} symbol allows us
+to omit domain types.
+For example,
+\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaArgument{B}
+is equivalent to
+\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\_}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaArgument{B}.
+Notice that this is a very different type to
+\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{B},
+which is a non-dependent function type equivalent to
+\AgdaSymbol{(}\AgdaSymbol{\_}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaArgument{B}.
+When writing
+\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaArgument{B},
+we assume that the occurrence of \AgdaBound{x} in \AgdaArgument{B} tells us what
+type \AgdaBound{x} should have (i.e.\ there is enough information to solve the
+underscore in
+\AgdaSymbol{(}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\_}\AgdaSymbol{)}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaArgument{B}).
+
+Just like in Haskell, functions in Agda can be introduced via
+\AgdaSymbol{$\uplambda$}-abstractions and clausal definitions, and are applied
+by juxtaposition.
+Agda also includes \emph{extended} \AgdaSymbol{$\uplambda$}-abstractions,
+introduced via equivalent syntaxes
+\AgdaSymbol{$\uplambda$}\AgdaSpace{}\AgdaKeyword{where}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{M} and
+\AgdaSymbol{$\uplambda$}\AgdaSpace{}\AgdaSymbol{\{}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{M}%
+\AgdaSpace{}\AgdaSymbol{\}},
+which allow for pattern-matching on the variable \AgdaBound{x} (or all of the
+variables, if there are multiple variables).
+
+Agda allows for arbitrary function arguments to be marked as \emph{implicit} by
+replacing the round brackets in the type by curly braces.
+For example, if we have
+\AgdaBound{f}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaArgument{A}\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
+\AgdaArgument{B},
+then the argument to \AgdaBound{f} is implicit.
+Being implicit means that an occurrence of \AgdaBound{f} is treated as if it has
+been applied to an underscore, giving the expression \AgdaBound{f} the type
+$\AgdaArgument{B}[\AgdaSymbol{\_}/\AgdaBound{x}]$ (i.e.\ \AgdaArgument{B} with
+\AgdaSymbol{\_} substituted in for \AgdaBound{x}; the substitution syntax is not
+part of Agda syntax).
+An implicit argument can also be given explicitly in two ways.
+The first of a series of implicit arguments can be given by surrounding the
+argument in curly braces, and any other implicit arguments in the series can be
+given by including the name of the argument.
+For example,
+\AgdaBound{f}\AgdaSpace{}\AgdaSymbol{\{}\AgdaSymbol{\_}\AgdaSymbol{\}},
+\AgdaBound{f}\AgdaSpace{}\AgdaSymbol{\{}\AgdaArgument{x}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}\AgdaSymbol{\_}\AgdaSymbol{\}}, and just
+\AgdaBound{f} are all equivalent expressions, and the underscore can be filled
+in in either of the first two expressions to provide an actual value for the
+implicit argument.
+Implicit arguments are usually left out of \AgdaSymbol{$\uplambda$}-abstractions
+and clausal definitions, but can be bound to names and pattern-matched on using
+the same syntax as in expressions.
+
+There are a few other places in the syntax using single curly braces, all of
+which have meanings related to implicit arguments.
+I also make a small amount of use of double curly braces
+(\AgdaSymbol{\{\{} and \AgdaSymbol{\}\}}), which denote arguments which are to
+be solved by instance resolution.
+Instance resolution is very similar to Haskell's typeclass resolution ---
+finding non-canonical solutions based on the instances in scope.
+
+Agda uses $\Pi$-types where in Haskell we would use polymorphism.
+For example, we can define an identity function as below.
+The definition relies on quantifying over terms of type \AgdaPrimitive{Set},
+i.e.\ (small) types.
+This definition also gives an example of defining a function with an implicit
+argument (\AgdaBound{X}), which can typically be inferred from either the
+kargument type or the return type, so can be omitted.
+
+\ExecuteMetaData[\SnippetsTwotex]{id0}
+
+An unfortunate feature of the definition \AgdaFunction{id$_0$} is that we cannot
+apply it to the expression \AgdaPrimitive{Set}, because \AgdaPrimitive{Set}
+contains only small types, and itself is a large type.
+We can work around these size issues using \emph{universe level polymorphism},
+as in the following definition.
+
+\ExecuteMetaData[\SnippetsTwotex]{id}
+
+Universe levels start at \AgdaPrimitive{0$\ell$}, with \AgdaPrimitive{Set} being
+an alias for \AgdaPrimitive{Set}\AgdaSpace{}\AgdaPrimitive{0$\ell$} (and also
+\AgdaPrimitive{Set$_0$}).
+Larger levels can be produced with the successor operator \AgdaPrimitive{suc},
+and we can take the least upper bound of two levels using the operator
+\AgdaPrimitive{\_$\sqcup$\_}.
+
+\subsection{Data types}
+
+Agda's \AgdaKeyword{data}-declarations are similar in scope to Haskell's, with
+the addition of indexing by terms of arbitrary type.
+\AgdaKeyword{data}-declarations give us indexed inductive sum-of-product types.
+
+All \AgdaKeyword{data}-declarations use GADT syntax.
+The body of a declaration comprises a list of constructor names paired with
+their types.
+Where two constructors have the same type, they may be written on the same line
+with their names separated by whitespace, as I do with the two constructors of
+\AgdaDatatype{Bool} below.
+\AgdaDatatype{Bool} has two constructors --- \AgdaInductiveConstructor{true} and
+\AgdaInductiveConstructor{false} --- both of which have type
+\AgdaDatatype{Bool}.
+\AgdaDatatype{$\mathbb N$} also has two constructors, where
+\AgdaInductiveConstructor{zero} has type \AgdaDatatype{$\mathbb N$} and
+\AgdaInductiveConstructor{suc} is inductive, with type
+\AgdaDatatype{$\mathbb N$}\AgdaSpace{}\AgdaKeyword{$\to$}\AgdaSpace{}%
+\AgdaDatatype{$\mathbb N$}.
+
+\noindent
+\begin{minipage}[t]{0.5\textwidth}
+  \ExecuteMetaData[\SnippetsThreetex]{Bool}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+  \ExecuteMetaData[\SnippetsThreetex]{Nat}
+\end{minipage}
+
+\AgdaDatatype{Bool} and \AgdaDatatype{$\mathbb N$} are both types, and indeed
+small types, as we can see by the fact that they are annotated to have type
+\AgdaPrimitive{Set}.
+We can also use \AgdaKeyword{data}-declarations to define type \emph{families}
+in various ways.
+The simplest is to add \emph{parameters}, as in the type family
+\AgdaDatatype{List} below.
+Parameters always appear to the left of the colon of the first line of the
+\AgdaKeyword{data}-declaration, and are constant throughout the
+\AgdaKeyword{data}-declaration.
+Variables to the left of the colon can appear in the body of the
+\AgdaKeyword{data}-declaration without further quantification.
+
+\ExecuteMetaData[\SnippetsThreetex]{List}
+
+Slightly more flexible than parameters are \emph{Protestant indices}%
+\footnote{The terminology of Protestant/Catholic indices comes orally from Conor
+McBride. I am not aware of a written reference.}.
+Protestant indices also appear to the left of the colon, and also must appear
+unmodified in the \emph{return type} of all of the constructors.
+However, they may take different values in inductive appearances of the type
+family in the argument types of constructors.
+Protestant indices give a generalisation of polymorphic
+recursion to indices of arbitrary type~\citep{Mycroft84,Henglein93}.
+
+I give two examples of type families with Protestant indices.
+The first, \AgdaDatatype{NestedList} is standard from the polymorphic recursion
+literature.
+It is worth noting at this point that Agda permits overloading of constructors,
+which are disambiguated by the type family they are being used to construct.
+This overloading allows \AgdaDatatype{List} and \AgdaDatatype{NestedList} to
+have constructors with the same names without confusion.
+The second example, \AgdaDatatype{ScopedTerm} is a data structure representing
+well scoped untyped $\lambda$-calculus terms.
+The Protestant index \AgdaBound{s} describes the number of variables in scope,
+which increases by 1 when we introduce a $\lambda$-abstraction.
+I will introduce \AgdaDatatype{Fin}, a type family with a specified natural
+number of inhabitants, in the next set of examples.
+As a syntactic note, in the type of the \AgdaInductiveConstructor{app}
+constructor, I use the two variable names \AgdaBound{M} and \AgdaBound{N}
+separated by whitespace to name two arguments with the same type.
+
+\ExecuteMetaData[\SnippetsThreetex]{NestedList}
+\ExecuteMetaData[\SnippetsThreetex]{ScopedTerm}
+
+The most general way to make a type family is to introduce a
+\emph{Catholic index}.
+The types of Catholic indices are specified to the right of the colon, and can
+be instantiated arbitrarily throughout the \AgdaKeyword{data}-declaration.
+Catholic indices are not in scope for the body of the
+\AgdaKeyword{data}-declaration, so the values filling them may need to be
+quantified over in each constructor.
+When this quantification is over a large type, like \AgdaPrimitive{Set}, the
+type family being defined will itself need to be large, e.g.\ inhabiting
+\AgdaPrimitive{Set$_1$}.
+This is a major reason for not defining types like \AgdaDatatype{List} and
+\AgdaDatatype{NestedList} using Catholic indices.
+
+I give two examples of type families with Catholic indices.
+The first is the \AgdaDatatype{Fin} family, as used in \AgdaDatatype{ScopedTerm}
+above.
+By inspection of the return types of the constructors, there is no way to
+produce a canonical inhabitant of
+\AgdaDatatype{Fin}\AgdaSpace{}\AgdaInductiveConstructor{zero}.
+For
+\AgdaDatatype{Fin}\AgdaSpace{}\AgdaSymbol(\AgdaInductiveConstructor{suc}%
+\AgdaSpace{}\AgdaBound{n}\AgdaSymbol),
+we can potentially use either of the constructors.
+Either we use \AgdaInductiveConstructor{zero} to get a canonical inhabitant, or
+if we can make a number with a smaller bound (i.e.\ an inhabitant of
+\AgdaDatatype{Fin}\AgdaSpace{}\AgdaBound{n}), we can use
+\AgdaInductiveConstructor{suc} to produce a larger number.
+
+\ExecuteMetaData[\SnippetsThreetex]{Fin}
+
+The second example of a type family with Catholic indices is more general in
+nature.
+Below I define \emph{propositional equality}, written
+\AgdaDatatype{\_$\equiv$\_}.
+It has two parameters and one Catholic index (though the standard library
+version of propositional equality I use throughout this thesis has an extra
+level parameter for the sake of universe level polymorphism).
+The constructor \AgdaInductiveConstructor{refl} constructs an inhabitant of
+\AgdaArgument{M}\AgdaSpace{}\AgdaDatatype{\_$\equiv$\_}\AgdaSpace{}%
+\AgdaArgument{N} only when \AgdaArgument{N} is definitionally equal to
+\AgdaArgument{M} (because terms are considered ``the same'' to the type checker
+exactly when they are definitionally equal).
+Notice that \AgdaInductiveConstructor{refl} does not quantify over \AgdaBound{x}
+because \AgdaBound{x} is already in scope as a parameter.
+
+\ExecuteMetaData[\SnippetsThreetex]{Eq}
+
+It is through type families like \AgdaDatatype{\_$\equiv$\_} that we can state
+and prove mathematical theorems in Agda.
+In the following subsection, I show how to use such indexed type families.
+
+\subsection{Clausal definitions}
+
+Clausal definitions of functions in Agda look very similar to their equivalents
+in Haskell.
+However, definitions in Agda regularly make use of
+\emph{dependent pattern matching}, which is our primary way of using indexed
+data types.
+Recursive definitions are also conservatively checked for termination.
+
+I will explain the salient aspects of clausal function definitions via two
+examples.
+The first, unimaginatively named \AgdaFunction{lemma}, shows a simple case where
+pattern matching modifies the context through unification of Catholic indices.
+The second, named \AgdaFunction{elim-Fin-zero}, gives an example of proper
+dependent pattern matching.
+
+In the following definition \AgdaFunction{lemma}, we want to chase equations in
+order to prove that \AgdaBound{x} is propositionally equal to \AgdaBound{z}.
+We start with the following incomplete definition, where the expression
+\AgdaHole{\{ \}0} marks an interation point, or \emph{hole}, in the program, to
+which we can apply interactive commands to complete the program.
+
+\begin{code}
+\>[0]\AgdaFunction{lemma}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{∀}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{A}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaPrimitive{Set}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{x}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaBound{A}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{z}\<%
+\\
+\>[0]\AgdaFunction{lemma}\AgdaSpace{}%
+\AgdaBound{p}\AgdaSpace{}%
+\AgdaBound{q}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaHole{\{ \}0}\<%
+\end{code}
+
+As a first step, I choose to match on the variable
+\AgdaBound{p}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{y}.
+The only applicable pattern is \AgdaInductiveConstructor{refl}.
+Doing this match has the effect of unifying \AgdaBound{y} --- which is taking
+the position of the Catholic index of \AgdaDatatype{\_$\equiv$\_} --- with
+\AgdaBound{x} --- which is the value of the index specified in the type of
+\AgdaInductiveConstructor{refl}.
+Local variables act as unification variables, so the unification succeeds with
+most general unifier $[x \coloneqq x, y \coloneqq x]$.
+Therefore, the type of \AgdaBound{q} becomes
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{x}.
+
+\begin{code}
+\>[0]\AgdaFunction{lemma}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{∀}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{A}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaPrimitive{Set}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{x}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaBound{A}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{y}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaBound{x}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{z}\<%
+\\
+\>[0]\AgdaFunction{lemma}\AgdaSpace{}%
+\AgdaInductiveConstructor{refl}\AgdaSpace{}%
+\AgdaBound{q}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaHole{\{ \}0}\<%
+\end{code}
+
+The next step is to match on \AgdaBound{q}.
+This similarly unifies \AgdaBound{z} and \AgdaBound{x}, making the conclusion
+type
+\AgdaBound{z}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{≡}}\AgdaSpace{}%
+\AgdaBound{z}.
+Finally, this conclusion type is in the image of the
+\AgdaInductiveConstructor{refl} constructor, so we may fill the hole with
+\AgdaInductiveConstructor{refl}.
+
+\ExecuteMetaData[\SnippetsThreetex]{lemma}
+
+Full \emph{dependent} pattern matching, as described by \citet{MM04}, is when
+the unification of indices described above takes account of constructors.
+In particular, the constructors of a data type satisfy the ``no confusion''
+property --- constructors are injective and mutually disjoint.
+Where we encounter disjoint constructors during unification, we may dismiss the
+corresponding case as impossible.
+Consider the following example (\AgdaFunction{elim-Fin-zero}).
+We start with an argument
+\AgdaBound{i}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaDatatype{Fin}\AgdaSpace{}%
+\AgdaInductiveConstructor{zero}, and consider which constructors could possibly
+construct such a value.
+However, as noted earlier, both constructors of \AgdaDatatype{Fin} target
+successor values of the index, from which \AgdaInductiveConstructor{zero} is
+disjoint.
+Therefore, both cases are impossible.
+The notation when all cases are impossible is to place empty round brackets
+\AgdaSymbol{()} in the place of the impossible argument, and to not provide a
+righthand side to the clause.
+
+\ExecuteMetaData[\SnippetsThreetex]{elim-Fin-zero}
+
+As an example of the injectivity of constructors, the obvious example is to
+internalise the proof of injectivity for a given constructor, as I do in
+\AgdaFunction{suc-injective}.
+We start with an argument
+\AgdaBound{p}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}%
+\AgdaInductiveConstructor{suc}\AgdaSpace{}\AgdaBound{m}%
+\AgdaSpace{}\AgdaOperator{\AgdaDatatype{$\equiv$}}\AgdaSpace{}%
+\AgdaInductiveConstructor{suc}\AgdaSpace{}\AgdaBound{n}
+and match on it.
+This time, we do have a possible pattern --- \AgdaInductiveConstructor{refl} ---
+but working out how to change the context relies on unifying
+\AgdaInductiveConstructor{suc}\AgdaSpace{}\AgdaBound{m} with
+\AgdaInductiveConstructor{suc}\AgdaSpace{}\AgdaBound{n}.
+We are justified in doing this, with most general unifier
+$[m \coloneqq m, n \coloneqq m]$, because \AgdaInductiveConstructor{suc} is
+injective (with respect to propositional equality).
+If the checker for dependent pattern matching did not know that
+\AgdaInductiveConstructor{suc} was injective --- for example, if it were instead
+a defined function --- then the unification would fail.
+This leads to the intuition that constructors and variables are well behaved
+with respect to dependent pattern matching, while other expressions are not.
+
+\ExecuteMetaData[\SnippetsThreetex]{suc-injective}
+
+Ordinarily, each clause of a definition gives rise to a \emph{definitional}
+equation between its lefthand side and righthand side.
+In intensional type theory, as implemented by Agda, definitional and
+propositional equality are contrasted to each other.
+Definitional equality corresponds to a decidable fragment of the natural
+equational theory of the type theory.
+As such, definitional equality is an entirely metatheoretic notion, and we can
+neither assume nor prove directly definitional equations within the language.
+Definitional equality is sometimes also called \emph{judgemental equality},
+because it forms a judgement which plays a part in the rules of the type theory.
+As well as from the clauses of definitions, we also get definitional equations
+from $\beta$-reductions of $\lambda$-abstractions and $\eta$-laws of functions
+and records.
+Because the type checker treats definitionally equal terms equivalently, we are
+able to refactor up to definitional equality without changing any downstream
+code.
+
+On the other hand, propositional equality is a notion internal to the language,
+as we have seen by defining propositional equality (\AgdaDatatype{\_$\equiv$\_})
+and proving things about it (\AgdaFunction{lemma}).
+Propositional equality is sometimes known as \emph{typal equality} or
+\emph{mathematical equality}.
+The latter name comes from the fact that propositional equality is the closest
+notion to what mathematicians usually call \emph{equality}, because, for
+example, it allows us to prove things like
+\AgdaBound{m}\AgdaSpace{}\AgdaFunction{+}\AgdaSpace{}\AgdaBound{n}%
+\AgdaSpace{}\AgdaDatatype{$\equiv$}\AgdaSpace{}%
+\AgdaBound{n}\AgdaSpace{}\AgdaFunction{+}\AgdaSpace{}\AgdaBound{m} for all
+natural numbers \AgdaBound{m} and \AgdaBound{n}.
+Propositional equality satisfies Leibniz' law, meaning that an inhabitant of a
+type \AgdaBound{A} can be coerced into an inhabitant of any type propositionally
+equal to \AgdaBound{A}.
+However, this cast requires marking in the code, so is less convenient to use
+than definitional equality.
+
+Definitional equality between two terms implies their propositional equality,
+because exactly when two terms are definitionally equal, the type checker is
+happy to accept \AgdaInductiveConstructor{refl} as a proof.
+This relationship between the two is simple, but can still be deceptive.
+For example, consider the notion of injectivity with respect to definitional
+and propositional equality.
+A function $f$ is injective (with respect to some notion of equality $\approx$)
+when, for all $x$ and $y$, we have $f\,x \approx f\,y \to x \approx y$.
+Because $\approx$ appears both covariantly and contravariantly in this
+definition, we have implications in neither direction between definitional
+injectivity and propositional injectivity.
+Indeed, we can find examples of all four possibilities:
+constructors are injective in both senses; type formers, like \AgdaDatatype{Fin}
+and \AgdaDatatype{List}, are definitionally injective but not propositionally
+injective;
+\ExecuteMetaData[\SnippetsThreetex]{double}\ can be proven to be
+propositionally injective, but is not definitionally injective because
+\AgdaFunction{\_+\_} is not injective;
+and nearly everything else is not injective in either sense.
+
+Because the notions of definitional and propositional injectivity are
+incomparable, so too are the corresponding unification procedures.
+Propositional unification (using only the injectivity of constructors) is used
+during dependent pattern matching, while solving of implicit arguments and
+underscores in expressions is done by definitional unification.
+
+\subsection{Records, $\Sigma$-types}
+
+While Agda provides built-in basic $\Pi$-types, with special syntax described in
+\cref{sec:pi-types}, it does not do the same for $\Sigma$-types.
+Instead, the default way to get the functionality of $\Sigma$-types is to
+declare record types, similarly to how we get sums via
+\AgdaKeyword{data}-declarations.
+However, the standard library does provide $\Sigma$-types, via record types,
+using the following declaration.
+
+\ExecuteMetaData[\SnippetsThreetex]{Sigma}
+
+As does the standard library, I will begin to use universe level polymorphism in
+these example definitions.
+Here, \AgdaBound{a} and \AgdaBound{b} are the levels of the two projections.
+The level of the record type must be at least the level of the type of each
+field, and in this case, the smallest such level is
+\AgdaBound{a}\AgdaSpace{}\AgdaOperator{\AgdaPrimitive{$\sqcup$}}\AgdaSpace{}%
+\AgdaBound{b}.
+As for the main points of interest in this \AgdaKeyword{record}-declaration, it
+contains two fields.
+The first, \AgdaField{proj$_1$}, simply has type \AgdaBound{A}.
+The second, \AgdaField{proj$_2$}, then has a type dependent on the value of the
+first field.
+Additionally, we give this record type a named constructor
+\AgdaInductiveConstructor{\_,\_}.
+Any record type can also be constructed using the more verbose syntax
+\begin{code}[inline]%
+\>[0]\AgdaKeyword{record}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaSpace{}%
+\AgdaField{proj$_1$}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaHole{\{ \}1}\AgdaSpace{}%
+\AgdaSymbol{;}\AgdaSpace{}%
+\AgdaField{proj$_2$}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaHole{\{ \}2}\AgdaSpace{}%
+\AgdaSymbol{\}}\<%
+\end{code}.
+
+The standard library provides various notations for
+\AgdaRecord{$\Upsigma$}-types, useful in various situations.
+In this thesis, I use \ExecuteMetaData[\SnippetsFourtex]{Sigma-syntax}\ and
+\ExecuteMetaData[\SnippetsFourtex]{exists}\ as equivalent notations for
+\AgdaRecord{$\Upsigma$}\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}\AgdaBound{B}.
+Indeed, the $\eta$-contracted form can be used with \AgdaFunction{$\exists$}, as
+in \AgdaFunction{$\exists$}\AgdaSpace{}\AgdaBound{B} (\AgdaSymbol{\textbackslash}
+is an alternative notation for \AgdaSymbol{$\uplambda$}, as in Haskell).
+\AgdaRecord{$\Upsigma$} also specialises to non-dependent products, as given by
+the infix operator \AgdaFunction{\_$\times$\_}.
+This is achieved by setting the parameter \AgdaBound{B} to be a constant type
+family.
+The resulting operator \AgdaFunction{\_$\times$\_}, as well as the non-dependent
+function type, behave better than their dependent counterparts with respect to
+unification because they allow us to remain in the first order fragment of
+higher order unification.
+
+There are two main ways of using the fields of a record.
+The first is to put the projections into scope using
+\AgdaKeyword{open}\AgdaSpace{}\AgdaRecord{$\Upsigma$}, and then to use the field
+names to project out of arbitrary terms of \AgdaRecord{$\Upsigma$}-type.
+This is what I will always do when using the \AgdaRecord{$\Upsigma$}-type
+family.
+Within this paradigm, there are two further notational choices.
+Either, we can use the field names as functions, so that
+\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaField{proj$_1$}%
+\AgdaSpace{}\AgdaBound{z}\AgdaSpace{}\AgdaOperator{\AgdaInductiveConstructor,}%
+\AgdaSpace{}\AgdaField{proj$_2$}\AgdaSpace{}\AgdaBound{z},
+or we can use postfix projections via the space-dot notation, as in
+\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaBound{z}%
+\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_1$}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor,}\AgdaSpace{}\AgdaBound{z}%
+\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_2$}.
+I tend to prefer the latter, also using it occasionally in ordinary mathematical
+notation (without the space).
+Both notations can also be used on the lefthand side of a clausal definition as
+\emph{copatterns}.
+Copatterns let us think of records as being function-like, with the fields of a
+record type being the possible arguments we can pass to such a function.
+
+The second way of using the fields of a record requires a motivating example.
+Consider the below definition of the type of semigroups at universe level
+\AgdaBound{$\ell$}.
+A semigroup has a carrier set, a binary operation on that set, and an
+associativity law for that binary operation.
+
+\ExecuteMetaData[\SnippetsFourtex]{Semigroup}
+
+In order to use the fields of \AgdaRecord{Semigroup} in the intended way, we
+do not open them into global scope.
+Doing so would mean that, for example, \AgdaField{\_$\bullet$\_} would take
+three arguments: the semigroup and its two intended arguments.
+Instead, we get to the point where we have a semigroup \AgdaBound{G} in scope
+and use
+\AgdaKeyword{open}\AgdaSpace{}\AgdaRecord{Semigroup}\AgdaSpace{}\AgdaBound{G}
+to put into scope the components \emph{of \AgdaBound{G}}.
+Then, the name \AgdaField{Carrier} in scope will refer to the carrier set of
+\AgdaBound{G}, the name \AgdaField{\_$\bullet$\_} will refer to the binary
+operator (which really takes two arguments), et cetera.
+Doing this gives the impression of working ``inside'' \AgdaBound{G}, which is
+the way I typically work with algebraic sets with structure.
+
+By $\eta$-equality, two inhabitants of a record type are definitionally equal
+exactly when they agree definitionally on all fields.
+This often makes record types much more convenient to work with than the
+corresponding single-constructor data types, which do not enjoy any $\eta$-laws.
+Notably, all inhabitants of the record type \AgdaRecord{$\top$} with no fields
+are definitionally equal.
+
+Along with \AgdaRecord{$\Upsigma$} and \AgdaRecord{$\top$}, there are two more
+general-purpose record types I need to cover which take advantage of two special
+features of \AgdaKeyword{record}-declarations (and also
+\AgdaKeyword{data}-delcarations, but I use \AgdaKeyword{record}-declarations for
+the convenience reason given in the previous paragraph).
+The first feature is that the universe level of a record type has a lower bound
+(the level of each field) but no upper bound.
+Therefore, we can introduce the following declaration \AgdaRecord{Lift}, which
+takes a type \AgdaBound{A} at level \AgdaBound{a} and produces an equivalent
+type at a potentially higher level
+\AgdaBound{a}\AgdaSpace{}\AgdaOperator{\AgdaPrimitive{$\sqcup$}}\AgdaSpace{}%
+\AgdaBound{$\ell$}.
+This type former is useful in situations which require a type at a specific
+level, such as when constructing a type using a function.
+
+\ExecuteMetaData[\SnippetsFourtex]{Lift}
+
+The other interesting property we get from \AgdaKeyword{record}-declarations is
+that the resulting type family is definitionally injective in its parameters.
+Therefore, record types behave well in the form of unification that solves
+implicit arguments.
+We can use this property to take any type family \AgdaBound{F} and produce an
+equivalent family \AgdaRecord{Wrap}\AgdaSpace{}\AgdaBound{F} which is
+definitionally injective.
+
+\ExecuteMetaData[\SnippetsFourtex]{Wrap}
+
+As an example, if we have a variable
+\AgdaBound{f}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaRecord{Wrap}%
+\AgdaSpace{}\AgdaFunction{F}\AgdaSpace{}\AgdaBound{y}
+and pass it to a function with a type of the form
+\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol\{\AgdaBound{x}\AgdaSymbol\}%
+\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaRecord{Wrap}\AgdaSpace{}%
+\AgdaFunction{F}\AgdaSpace{}\AgdaBound{x}\AgdaSpace{}\AgdaSymbol{$\to$}%
+\AgdaSpace{}\AgdaSymbol{\_},
+Agda will successfully unify the type of \AgdaBound{f} with the expected type of
+the argument, setting $[x \coloneqq y]$.
+However, without the \AgdaRecord{Wrap}, we would need to unify
+\AgdaFunction{F}\AgdaSpace{}\AgdaBound{y} with
+\AgdaFunction{F}\AgdaSpace{}\AgdaBound{x}, which would fail if \AgdaFunction{F}
+were not injective, because there may be multiple acceptable values of
+\AgdaBound{x} up to definitional equality.
+
+The version of \AgdaRecord{Wrap} found in Agda's standard library is
+significantly more complicated to allow for type families with arbitrarily many
+arguments in a convenient syntax, using the $n$-ary functions of
+\citet{Allais19}.
+The version in the standard library is the one I use in this thesis.
+In fact, both versions of the \AgdaRecord{Wrap} type family are the first pieces
+of novel work to be presented in this thesis.
+
+\subsection{Colours}
+
+I use the ``Conor colours'' option for Agda syntax highlighting.
+This set of colours is inspired by Conor McBride's syntax highlighting for
+Epigram 2.
+The colour given to a name is determined by the type of declaration that name
+is bound to.
+The main colours are \AgdaDatatype{blue} for types and type families,
+\AgdaField{red} for constructors of data types and fields of records,
+\AgdaFunction{green} for definitions which may unfold/compute, and
+\AgdaBound{purple} for local variables.
+
+Separately, I use \gr{green} in many places for usage annotations in traditional
+typeset mathematical notation.
+This usage of green contrasts only with ordinary black text.

### 2023-07-12 tex/kits.tex
[ writing ] various up to WRel example

@@
 
-For this syntax, a renaming from \AgdaBound{$\Gamma$} to \AgdaBound{$\Delta$}
+A renaming from \AgdaBound{$\Gamma$} to \AgdaBound{$\Delta$}
 is a map from variables in \AgdaBound{$\Delta$} to variables in
-\AgdaBound{$\Gamma$}.
-Substitutions instead map into terms in \AgdaBound{$\Gamma$}.
+\AgdaBound{$\Gamma$}, represented in Agda as follows.
 
 \Ren{}
-\Sub{}
 
-We define \AgdaFunction{ren} as the action of a renaming on terms.
+The action of a renaming \AgdaBound{$\rho$} on terms is given by
+\AgdaFunction{ren}\AgdaSpace{}\AgdaBound{$\rho$}, with \AgdaFunction{ren}
+defined below.
 The idea of simultaneous renaming is to preserve the structure of the term, but
@@
 
+Meanwhile, a substitution from \AgdaBound{$\Gamma$} to \AgdaBound{$\Delta$} is
+an inhabitant of \AgdaFunction{Sub}\AgdaSpace{}\AgdaBound{$\Gamma$}\AgdaSpace{}%
+\AgdaBound{$\Delta$}, as defined below.
+This definition is identical to the definition of \AgdaFunction{Ren}, except
+that it gives us \emph{terms} in \AgdaBound{$\Gamma$} rather than variables.
+
+\Sub{}
+
 The \AgdaFunction{sub} function below gives the action of a substitution.
@@
 
-The statements of simultaneous renaming and simultaneous substitution are
+As observed by \citet{McBride05,BHKM12},
+the statements of simultaneous renaming and simultaneous substitution are
 very similar, with substitution being the generalisation that allows
-replacement by terms rather than just variables.
+replacement of variables by terms rather than just other variables.
+Following \citet{McBride05},
 I will introduce a type family \AgdaFunction{Env} of \emph{environments}, and
@@
 \AgdaFunction{bindSub}, we notice only three key
-differences:~\cite{McBride05,BHKM12}
+differences:
 

### 2023-07-12 tex/terms.tex
[ writing ] various up to WRel example

@@
 nullary \& binary nodes, and serves as a generalised ``length'' of the context.
+The tree shape makes concatenation definitionally
+injective, so that in cases where multiple new variables are bound in a subterm
+(for example, $\otimes$-elimination), Agda's unification-based solving will
+be more able to infer which variables have just been bound.
+Within a given \AgdaBound{t}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}%
+\AgdaDatatype{LTree}, we can define the positions of \AgdaBound{t} using
+\AgdaDatatype{Ptr}.
+A \emph{pointer} (\AgdaDatatype{Ptr}) into a tree picks out a leaf
+(\AgdaInductiveConstructor{[-]}) following a path of lefts
+(\AgdaInductiveConstructor{$\swarrow$}) and rights
+(\AgdaInductiveConstructor{$\searrow$}) at any binary nodes encountered.
+
+\ExecuteMetaData[\LTreetex]{LTree}
+\ExecuteMetaData[\LTreetex]{Ptr}
+
 The contents of the context --- the types --- are then stored in the functional
@@
 until later chapters --- particularly the example in
-\cref{sec:usage-elaborator}, where I make use of the ease of look-up and
+\cref{sec:usage-elaborator}, where I make use of the ease of look-up and the
 $\eta$-law of functions.
@@
 use renaming.
-As for the tree shape, this makes context concatenation definitionally
-injective, so that in cases where multiple new variables are bound in a subterm
-(for example, $\otimes$-elimination), Agda's unification-based solving will
-be more able to infer which variables have just been bound.
 
-\ExecuteMetaData[\LTreetex]{LTree}
+\ExecuteMetaData[\Vectortex]{Vector}
 \Ctx{}
@@
 with a proof \AgdaField{tyq} that this type is equal to \AgdaBound{A}.
+
+\Var{}
+
 Variables embed into terms via the \AgdaInductiveConstructor{var} constructor of
@@
 
-\Var{}
 \Term{}
 
-\todo{Example terms}
+Using this encoding, the Church numeral for 2 appears as follows.
+In standard notation, this would be
+$\lambda f.~\lambda x.~f\,(f\,x)$.
+To refer to $f$ in the main body of the expression, we skip one binder (using
+\AgdaInductiveConstructor{$\swarrow$}) and pick the next one
+(using \AgdaInductiveConstructor{$\searrow$}) and pick its only bound variable
+(using \AgdaInductiveConstructor{here}).
+To refer to $x$, we do not skip its binder, instead picking it and its only
+bound variable.
+
+\ExecuteMetaData[../agda/processed-latex/SimpleKits.tex]{two}

### 2023-07-12 tex/trans.tex
[ writing ] various up to WRel example

@@
 
-\begin{definition}
+\begin{definition}\label{def:PD-bottom-up}
   A derivation is said to be \emph{$\instPD$-bottom-up} if only the following

### 2023-07-16 tex/agda-primer.tex
[ writing ] address all remaining todos

--- 2023-07-12 tex/intro.tex
@@
-\chapter{Introduction}
-
-This thesis advances the frontier of programming languages work that can be done
-in a proof assistant.
-I will elaborate on both of these aspects in the following paragraphs.
-
-The main aim of programming language research is to find common patterns in
-computer programs and represent these patterns via programming language
-features.
-As a basic example, consider functions in, say, a low level language like C, in
-contrast to what these functions are compiled to in assembly language/machine
-code.
-In order to support function calls, the C runtime system manages a
-\emph{call stack}.
-A function call is compiled to a sequence of instructions which store the return
-address and the values of the arguments on the stack, move the stack pointer,
-and jump to the compiled code of the function.
-A return from a function then places the return value into the appropriate
-register, moves the stack pointer back, and jumps back to the return address.
-Advances in programming language techniques are evaluated by the usefulness of
-behaviours captured and goodness of the mathematical properties of the
-abstractions produced.
-
-Within programming language research, type theory is a methodology for producing
-abstractions.
-A type system lets us enforce in our language complex invariants which maintain
-abstraction boundaries while also ideally being machine-checkable.
-For example, we give functions function type to abstract away their
-implementation via call stacks and jumps and so on.
-
-In this thesis, the invariants of interest revolve around restricting the usage
-of variables.
-I introduce this topic thoroughly in \cref{sec:linear}, but in brief\ldots
-
-The work of this thesis relies upon type theory in two distinct ways.
-Firstly, the main objects of study are programming languages with interesting
-type systems.
-Secondly, type theory provides the basis of the proof assistant Agda I use to
-implement the aforementioned programming languages and operations upon them.
-
-As for the other topic of this thesis,
-a \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
-a piece of software that allows for the encoding of mathematical definitions,
-theorems, constructions, and proofs, and furthermore check that such encoded
-proofs are correct and that such encoded constructions are well formed.
-To truly by interactive, i.e.\ to actually assist, a proof assistant will
-usually have a user interface which can read partial proofs, display
-information about what more proof needs to be given, and provide actions that
-will help complete the proof.
-
-Proof assistants have seen increasing use in programming language research in
-recent years.
-The most obvious reason why working in a proof assistant is seen as beneficial
-is that it ensures correctness.
-If the proof assistant accurately implements a suitable mathematical foundation,
-then any theorem proved in the proof assistant is guaranteed to be a true
-theorem of that foundation.
-These guarantees of correctness are particularly important when working with
-combinatorially complex mathematical objects, proofs about which often require
-the consideration of a large number of cases.
-Programming language syntaxes are often such complex objects, motivating the use
-of proof assistants when studying programming languages.
-
-A second reason to use proof assistants is for the assistance they provide when
-exploring a mathematical theory.
-When we make a new definition, we may want to test how it works in a special
-case, or what constructions it allows us to perform.
-In a proof assistant, the assistance tools give us immediate feedback as to what
-moves are and aren't allowed.
-For example, if we define a complex type system, a proof assistant will let us
-interactively build typing derivations, making clear any side conditions and
-types of subderivations as we go.
-Also, as I do later in this thesis, a proof assistant allows us to build a very
-general theory, and practically use that theory directly in more specific cases
-without losing rigour.
-
-Thirdly, analogously to how a strong static type system can give us more
-confidence when refactoring a program, the constant checking of proofs in a
-proof assistant gives us the confidence to change definitions and lemmas knowing
-that we will be guided towards the parts of our theory that need to be
-correspondingly changed.
-This can help if we are developing a new programming language with a changing
-specification.
-
-Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
-this thesis --- double as programming languages themselves.
-This means that we can write progams and prove properties of them using the same
-tool.
-Also, many theorems proven in such a proof assistant have computational content.
-For example, if we prove a normalisation theorem for a programming language,
-this will typically yield a (verified) normalisation algorithm for it, which we
-can really run on a computer.
-As such, a development in a proof assistant can provide a reference
-implementation of a programming language, or even --- as with Idris
-2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
-actual implementation of a programming language.
-
-FIXME: the start of this chapter is directly from the ESOP22 paper.
-
-In this paper, we treat the metatheory of a class of substructural type
-systems related to linear logic~\cite{girard87linear}.
-This class is variously known as
-\emph{coeffectful}~\cite{PetricekOM14,Granule18},
-\emph{quantitative}~\cite{BrunelGMZ14,Atkey18}, or
-\emph{resource-aware}~\cite{GhicaS14},
-or is given no particular name~\cite{reed10distance,abadi99core},
-and generalises bounded linear logic to track variable usage with semiring
-annotations.
-In all of these systems, we have some ambient semiring $\Ann$, and in the
-judgements of the type system, variables are annotated by elements of $\Ann$
-describing \emph{how} that variable can be used.
-The additive structure of $\Ann$ gives the ability to count, or otherwise
-accumulate, usages of variables in multiple subterms.
-The multiplicative structure gives rise to a form of modality, for example
-allowing multiple or unlimited reuse, or movement between security levels, in
-the type system.
-
-The aspect of such systems we tackle here is their basic metatheory and
-mechanisation thereof.
-
-We build upon both the general structural framework of
-\citet{AACMM21} and the substructural techniques of \citet{WA21}.
-The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
-propositional natural deduction systems based on intrinsically typed syntax and
-de Bruijn indices, we aim to replicate for linear-like systems based on
-semiring usage annotations.
-By picking a trivial semiring, our work can subsume that of
-\citeauthor{AACMM21}, except for the many pieces of machinery we have not yet
-ported to this new framework.
-
-Our work complements that of \citet{Granule18} on the Granule programming
-language.
-Where Granule focuses on writing programs \emph{in} the language and running
-them, we focus on metatheoretic reasoning \emph{about} type systems.
-%As such, whereas Granule has a convenient syntax, performant type-checker,
-%interpreter
-
-Our work is similar in scope to that of \citet{LicataSR17}, though we work in
-a natural deduction style rather than a sequent calculus style.
-Where \citeauthor{LicataSR17} are much more agnostic in terms of
-substructurality --- allowing for non-commutative and bunched logics ---
-we are much more agnostic in terms of syntax.
-The system of \citeauthor{LicataSR17} is essentially a single calculus,
-supporting ``product'' ($\mathrm F$) types and ``function'' ($\mathrm U$)
-types, parametrised on a \emph{mode theory} describing its structural rules.
-For this system, they derive the strong result of cut elimination.
-Meanwhile, we leave syntax design to the user, and consequently can only
-guarantee substitution (which we can only get because of our commitment to
-natural deduction).
-
-\section{Outline of the thesis}
-
-This thesis proceeds as follows.
-The next two chapters, \cref{sec:simple,sec:linearity}, are introductory in
-nature, and cover two largely independent strands of prior work.
-In \cref{sec:simple}, I introduce existing methods of representing and reasoning
-about type systems in proof assistants based on dependent type theory.
-I start from well established representations of well scoped and well typed
-terms, and develop these towards a recent approach to environment-based
-semantics given by \citet{AACMM21}.
-In \cref{sec:linearity}, I discuss the challenges faced when one extends a
-treatment of a simple type system, such as that given in \cref{sec:simple}, to
-modal and linear type systems.
-We see that modal and linear type systems apparently violate some of the nice
-properties of the simply typed $\lambda$-calculus we required in
-\cref{sec:simple}.
-I present a solution for intuitionistic S4 modal logic, but leave a solution for
-linear logic to the following chapters.
-
-In the following two chapters, \cref{sec:semirings,sec:ren-sub-lr}, I present a
-calculus $\name$ parametrised by a partially ordered semiring of \emph{usage
-annotations}.
-In \cref{sec:semirings}, I define the calculus, give some possible extensions,
-and show that it subsumes intuitionistic S4 modal logic and Intuitionistic
-Linear Logic.
-In \cref{sec:ren-sub-lr}, I show that $\name$ enjoys generalised versions of the
-nice properties required in \cref{sec:simple}, and I proceed to give novel
-definitions of simultaneous substitutions and their action on $\name$ terms.
-These two chapters are adapted from the work of \citet{WA21}.
-
-The remaining three main chapters,
-\cref{sec:framework,sec:semantics,sec:example-semantics}, adapt the syntactic
-and semantic framework of \citet{AACMM21}, as presented at the end of
-\cref{sec:simple}, to semiring-annotated calculi.
-\Cref{sec:framework,sec:semantics} generalise the work on $\name$ presented in
-\cref{sec:semirings,sec:ren-sub-lr}, respectively.
-\Cref{sec:framework} shows how to formally describe the syntax of an arbitrary
-semiring-annotated calculus, following the constructions used in
-\cref{sec:semirings}.
-\Cref{sec:semantics} then provides the generic environment-based semantic
-traversal on such syntaxes, providing renaming and substitution as per
-\cref{sec:ren-sub-lr} for all syntaxes as special cases of the generic
-traversal.
-\Cref{sec:example-semantics} then gives further example uses of the generic
-traversal.
-
-Finally, I conclude with \cref{sec:conc}, which discusses the achievements of
-this thesis and openings for future work.
-
-\section{Naming and notation conventions}
-
-I assume familiarity with the Curry-Howard correspondence throughout this
-thesis.
-I make no distinction between logics and type theories, and use terminology from
-each interchangeably.
-Each following bullet point lists a collection of synonyms.
-
-\begin{itemize}
-  \item assumption, hypothesis, variable
-  \item derivation, proof, term
-  \item proposition, formula, type
-  \item connective, type former
-\end{itemize}
-
-I carry out mechanised constructions and proofs in the proof assistant and
-programming language Agda~\citep{Agda}.
-Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
-similarly present non-mechanised constructions and proofs assuming a foundation
-given by dependent type theory, in a style inspired by the HoTT
-Book~\citep{hottbook}.
-
-\section{Agda primer}
+\section{Agda primer}\label{sec:agda-primer}
 
 I use the proof assistant and programming language Agda throughout this thesis,
-with Agda code being used particularly in \cref{sec:simple} and the later
-chapters.
+with Agda code being used particularly in this chapter and
+\cref{sec:framework,sec:semantics,sec:example-semantics}.
 As such, it is important for the reader to be able to read basic Agda syntax in

### 2023-07-16 tex/conc.tex
[ writing ] address all remaining todos

@@
 environments, like composition of renamings and substitutions.
+The definition of environments, and hence simultaneous renaming/substitution, is
+novel, but acts like the appropriate generalisation of environments for simply
+typed theories.
 
@@
 one could not write without syntax descriptions.
-% Other examples...
-Together, these examples show the applicability and versatility of the framework
+Meanwhile, the NbE algorithm gives further justification that I have the correct
+notion of environment for semiring-annotated type systems.
+Together, the examples show the applicability and versatility of the framework
 I have developed.
-
-FIXME: the next two paragraphs are directly from the ESOP22 paper.
-
-We have presented a framework for doing metatheory for a class of substructural
-type systems in Agda.
-The framework gives us renaming, substitution, and a usage elaborator for new
-syntaxes for free, which we hope can facilitate prototyping and the
-mechanisation of more interesting semantic results.
-Beside the mechanised framework itself, we believe its methodology --- the use
-of bunched premise combinators --- can guide and simplify the development of
-(potentially unmechanised) substructural type systems.
-
-Our account of substructurality is based on the linear algebraic
-principles described by \citet{WA21}.
-However, these details only really affect the definition of environment,
-in which the use of linear maps is motivated by them being the standard notion
-of morphism between vectors.
-We could imagine that a similar notion of morphism is found for the kind of
-annotations found in \citet{LicataSR17}, allowing a framework to consider
-finer substructural systems.
-
-\begin{itemize}
-  \item Novel simultaneous substitution for linear type theories
-  \item Generalise \citet{AACMM21} work to substructural logics
-  \item Is this a reusable library? Yes, examples
-  \item Methodology for metatheory of new calculi
-  \item Semirings have been a reasonably robust basis.
-  \item Can the code be directly generalised?
-    Abstract away the algebraic structure of contexts.
-\end{itemize}
 
@@
 
-\begin{itemize}
-  \item Fitch-style systems (can't do).
-\end{itemize}
-
-\subsection{Further work}
+To conclude, I discuss various possible future directions of the work started in
+this thesis.
 
@@
 presented earlier is an open question.
-However, \citet{Autosubst15} did apply their related system to System $F_{<:}$,
-so it may be possible to support polymorphic systems using their techniques.
+However, both \citet{Autosubst15} and \citet{Autosubst18} have applied similar
+work to polymorphic calculi via special support, suggesting that it would be
+possible to modify the framework of this thesis in a similar way.
 
@@
 
-\subsubsection{Partiality}\label{sec:part}
+\paragraph{Partiality}\label{sec:part}
 \input{part.tex}
+
+\paragraph{A separate variable sort}
+\Citet{AACMM21} mention as a limitation of their work the fact that variables
+can have unrestricted type.
+This limitation carries over to the work of this thesis.
+In many languages --- for example, the $\mu\tilde\mu$-calculus of
+\cref{sec:mu-mu-tilde} and the normal/neutral forms of a $\lambda$-calculus ---
+we would like to be able to restrict the syntax so as to disallow variables of
+certain kinds.
+Such a restriction would make it easier to construct traversals over these
+syntaxes, which I have not attempted in this thesis.
+
+In private communication, Allais has suggested a simple solution to this problem
+by which the type of variable types and the type of term types are distinct, and
+related by an arbitrary relation $R$.
+Each use of the \AgdaInductiveConstructor{`var} constructor then requires a
+proof that $R$ relates the type of the variable picked to the type of the
+conclusion desired.
+For example, in the syntax of normal/neutral forms, we let the variable types be
+just the types, while the term types are types tagged with whether the term is
+normal or neutral.
+The relation $R$ relates equal types where the term type is marked neutral.
+Allais also suggests that such a scheme could be used to bind patterns in the
+context, with the relatedness proof serving as a path through the pattern to a
+variable of the desired type.
+It remains non-trivial to work out the appropriate notion of renaming, and more
+generally the appropriate notion of environment, in this setting, which I leave
+to future work.
+
+\paragraph{Logical frameworks}
+I discussed in \cref{sec:lin-sys} two previous approaches to representing linear
+languages in logical frameworks.
+The work of this thesis should help in adapting the first approach --- making a
+new logical framework with direct support for linearity --- to a broader range
+of substructural calculi.
+In particular, the bunched premise connectives give candidates for the required
+type formers in such a logical framework.
+On the other hand, the second approach --- to use an existing non-linear logical
+framework but include a \texttt{linear} predicate --- should be easy to adapt to
+semiring-annotated systems by replacing the \texttt{linear} predicate by a
+relation between semiring elements and \texttt{term -> term} weak functions.
+
+\paragraph{Fitch-style systems}
+I mentioned in \cref{sec:fitch} an alternative way to present modal logics,
+namely Fitch-style natural deduction systems.
+These appear to be largely irreconcilable with the work of this thesis due to
+the fact that the $\Box$-elimination rule is very sensitive to the context.
+A general treatment of Fitch-style syntaxes, comparable to the treatment of
+semiring-indexed syntaxes given in this thesis, would probably have to assume a
+$\Box$ connective whose behaviour was determined as part of the structural rules
+of the calculus.

### 2023-07-16 tex/examples.tex
[ writing ] address all remaining todos

@@
-In this chapter, I provide three example uses of semantic traversals: generic
-renaming and substitution\todo{Moved}, a usage elaborator, and a denotational semantics.
+In this chapter, I provide four example uses of semantic traversals:
+a usage elaborator, a normalisation by evaluation algorithm, a denotational
+semantics, and translations between calculi.
 The reader is also encouraged to see the far greater range of examples in the
 work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
-Renaming and substitution are essential results, while the latter two examples
-focus on usage annotations.
+The usage elaborator (\cref{sec:usage-elaborator}) gives an example of a program
+that is generic in both syntax and usage annotations, as well as being an
+essential tool for practical use of the framework.
+The normalisation by evaluation algorithm (\cref{sec:nbe}) is for a specific
+syntax, but exercises parts of the generic semantic traversal --- particularly,
+its handling of environments --- to good effect.
+The denotational semantics in terms of world-indexed relations
+(\cref{sec:den-sem}) presents an interesting semantics for semiring-annotated
+type systems, but with minimal use of operations provided by the framework.
+Finally, the translations between $\name$ and L/nL (\cref{sec:lnl}) both justify
+the encoding of L/nL given in \cref{sec:dup-lnl} and give examples of syntactic
+work benefitting from the generic renaming and substitution operations.
 
@@
 of variable binding.
-For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-comb}, the right
+For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
 premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
@@
 We can now use the elaborator to automatically infer the usage
-annotations for the example at the end of \cref{sec:terms}. This
-allows us to write:
+annotations for the example at the end of \cref{sec:terms-of-system}.
+This allows us to write:
 \ExecuteMetaData[\PaperExamplestex]{cojoin}

### 2023-07-16 tex/gen-syn.tex
[ writing ] address all remaining todos

@@
 
-To make \AgdaFunction{sem} pass the termination check, we have three major
+To make \AgdaFunction{sem} pass the termination check, we have four major
 options:
@@
     at subterms.
+  \item Inline a new, instantiated version of \AgdaFunction{map-s} wherever it
+    is used.
 \end{enumerate}
@@
 make the resulting code harder to use.
+Finally, approach 4 is safe, but limits code reuse (both of the function
+\AgdaFunction{map-s} itself and any lemmas we may prove about it).
 I choose to follow \citet{AACMM21} in using sized types, justified by the idea
@@
 as possible.
+\Citet{FS22} use approach 4, and in fact have only one use of (their equivalent
+of) \AgdaFunction{map-s} in their development.
 

### 2023-07-16 tex/ill.tex
[ writing ] address all remaining todos

@@
 having a $B$ if I have and am willing to give up an $A$''.
-\todo{Camels example}
+
+The resource interpretation of intuitionistic linear logic is often explained by
+analogy with a vending machine.
+Suppose we have a vending machine which takes pound coins and sells bottles of
+water for \pounds{}1 and chocolate bars for \pounds{}2.
+We can represent the latter two facts by the judgements
+$\textrm{\pounds}1 \vdash \mathrm{water}$ and
+$\textrm{\pounds}1, \textrm{\pounds}1 \vdash \mathrm{bar}$, respectively.
+Notice that having two \pounds{}1 coins is very different from having just one
+\pounds{}1 coin.
+We can represent having compound objects using $\otimes$-products
+(tensor-products), so that having a \pounds{}1 coin together with a chocolate
+bar is represented as $\textrm{\pounds}1\otimes\mathrm{bar}$, and the mechanism
+for getting a chocolate bar from the vending machine is described by
+$\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}$.
+The entire protocol of one transaction with the vending machine is represented
+as
+$\plr{\textrm{\pounds}1 \multimap \mathrm{water}} \with
+\plr{\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}}$,
+where the $\with$-product (with-product) represents giving the user a choice of
+which of the two conjuncts to have.
+Finally, that we can do multiple transactions with the vending machine is
+represented as
+$\oc\plr{\plr{\textrm{\pounds}1 \multimap \mathrm{water}} \with
+\plr{\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}}}
+$, where the $\oc$-modality (read ``bang'' or ``of course'') represents having
+an arbitrary number of copies of its argument.
+
+Such reasoning is useful in computer science because it allows us to distinguish
+what we had before some event with what we got after it.
+In both intuitionistic and classical logic, if we had a \pounds{}1 coin and a
+way to turn a \pounds{}1 coin into a bottle of water, we'd be able to have both
+the \pounds{}1 coin and the bottle of water at once.
+Instead, linear logic acknowledges that the \pounds{}1 coin was spent, so at one
+time we had just the \pounds{}1 coin, and at a later time we had just the bottle
+of water.
+In real applications, such reasoning is useful in, for example, dealing with
+mutable state~\citep{MK19} --- in which, when we mutate the state, we forget the
+old value and have only the new value --- and producing a session type system~%
+\citep{Wadler12} --- in which the protocol progresses as messages are sent.
+
 In this section, I will carefully introduce intuitionistic linear logic so as to
 make the intuitive readings more precise.
-
-%Linear logic is a good starting point for our investigations into usage
-%restrictions because it is well known, well understood, and exhibits most of the
-%difficulties we will come across.
-%Linear logic's lack of general weakening and contraction will be a typical
-%feature of the calculi we will see, and the exponential modality $\oc$ (bang)
-%will be the prototype for a range of modalities I will consider.
-
-\subsection{Motivation of linearity}
-
-\todo{Incorporate into chapter introduction.}
-Programming applications of linearity include stateful protocols
-\citep{Wadler12}, mutually exclusive capabilities (TODO: cite),
-and mutation (TODO: cite).
-In each of these cases, the act of using up a hypothesis allows us to divide
-time into ``before'' and ``after'', with lack of duplication avoiding confusion
-over which half we are in, and lack of deletion allowing us to know whether the
-corresponding action actually happened.
-For example, when modelling mutation, a variable refers to a single state of a
-mutable value.
-The \texttt{write} operation uses up such a variable, making the old state
-inaccessible, and returns a new linear value to represent the new state of the
-mutable value.
-Such a protocol naturally also supports a \emph{freezing} operation, by which we
-relinquish the ability to mutate the value in return for an immutable reference
-to the value produced.
 

### 2023-07-16 tex/intro.tex
[ writing ] address all remaining todos

--- 2023-06-22 tex/intro.tex
@@
 
-FIXME: the start of this chapter is directly from the ESOP22 paper.
+This thesis advances the frontier of programming languages work that can be done
+in a proof assistant.
+I will elaborate on both of these aspects in the following paragraphs.
 
-In this paper, we treat the metatheory of a class of substructural type
-systems related to linear logic~\cite{girard87linear}.
-This class is variously known as
-\emph{coeffectful}~\cite{PetricekOM14,Granule18},
-\emph{quantitative}~\cite{BrunelGMZ14,Atkey18}, or
-\emph{resource-aware}~\cite{GhicaS14},
-or is given no particular name~\cite{reed10distance,abadi99core},
-and generalises bounded linear logic to track variable usage with semiring
-annotations.
-In all of these systems, we have some ambient semiring $\Ann$, and in the
-judgements of the type system, variables are annotated by elements of $\Ann$
-describing \emph{how} that variable can be used.
-The additive structure of $\Ann$ gives the ability to count, or otherwise
-accumulate, usages of variables in multiple subterms.
-The multiplicative structure gives rise to a form of modality, for example
-allowing multiple or unlimited reuse, or movement between security levels, in
-the type system.
+The main aim of programming language research is to find common patterns in
+computer programs and represent these patterns via programming language
+features.
+As a basic example, consider functions in, say, a low level language like C, in
+contrast to what these functions are compiled to in assembly language/machine
+code.
+In order to support function calls, the C runtime system manages a
+\emph{call stack}.
+A function call is compiled to a sequence of instructions which store the return
+address and the values of the arguments on the stack, move the stack pointer,
+and jump to the compiled code of the function.
+A return from a function then places the return value into the appropriate
+register, moves the stack pointer back, and jumps back to the return address.
+Advances in programming language techniques are evaluated by the usefulness of
+behaviours captured and goodness of the mathematical properties of the
+abstractions produced.
 
-The aspect of such systems we tackle here is their basic metatheory and
-mechanisation thereof.
+Within programming language research, type theory is a methodology for producing
+abstractions via classifying program behaviour.
+The simplest and most common type systems classify programs by what values they
+may return.
+For example, the body of a C function with return type \texttt{char} is a
+program that, if it returns a value, will return a character.
+If that function has a parameter of type \texttt{int}, then we can compose it
+with a function with return type \texttt{int} to build up a larger program.
+We expect type systems to stop us from running programs with arguments of the
+wrong types.
+This holds of both static and dynamic type systems --- with a static type
+system, the compiler will refuse to compile our code if we pass a \texttt{char}
+value to a function expecting an \texttt{int}, while with a dynamic type system,
+our program will do a check before running the function to make sure that we
+have passed it an \texttt{int}.
+The abstraction produced in such simple type systems is the idea that a function
+will take arguments in accordance with its parameter types and produce a result
+in accordance with its result type, and furthermore, as readers, we do not have
+to inspect the function's implementation to see that these properties are true.
 
-We build upon both the general structural framework of
-\citet{AACMM21} and the substructural techniques of \citet{WA21}.
-The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
-propositional natural deduction systems based on intrinsically typed syntax and
-de Bruijn indices, we aim to replicate for linear-like systems based on
-semiring usage annotations.
-By picking a trivial semiring, our work can subsume that of
-\citeauthor{AACMM21}, except for the many pieces of machinery we have not yet
-ported to this new framework.
+A more recent trend in type theory is to use types to describe other parts of a
+program's behaviour than what range of values it may return.
+For example, Java's \emph{checked exceptions} can be seen as part of the static
+type system, in which programs are classified by which exceptions they may
+throw.
+More generally, \emph{effect systems} classify programs based on all of the
+effects they may have --- such as reading input and writing to files, and also
+internally defined effects, such as non-determinism and using an accumulator.
+In a type system which tracks effects (an \emph{effect system}), programs by
+default are pure (i.e.\ have no effects), with effects being opt-in.
+Pure programs generally enjoy good properties, making them easy to reason about
+and easy for an optimising compiler to optimise.
 
-Our work complements that of \citet{Granule18} on the Granule programming
-language.
-Where Granule focuses on writing programs \emph{in} the language and running
-them, we focus on metatheoretic reasoning \emph{about} type systems.
-%As such, whereas Granule has a convenient syntax, performant type-checker,
-%interpreter
+In this thesis, I consider the dual of effect systems ---
+\emph{coeffect systems} --- as introduced by \citet{POM14}.
+In a coeffect system, we are interested not in what extra behaviour a program
+may exhibit (as with effects), but rather what extra abilities the context of a
+program may provide.
+Analogously to the case of effect systems, we typically restrict our
+coeffect-free programs to be ``more pure'' than usual.
+A standard example is to restrict to \emph{linear} programs, in which each
+variable in the context is used exactly once.
+The ability to duplicate and discard variables is then seen as a coeffect, which
+can be tracked by a coeffect system.
+Restricting to linear programs may seem like an arbitrary restriction at first,
+but the expectation of linearity arises naturally in applications such as
+file-handling, session-typed communication, and approaches to mutable memory.
+I introduce linearity and its applications more fully in \cref{sec:linearity}.
 
-Our work is similar in scope to that of \citet{LicataSR17}, though we work in
-a natural deduction style rather than a sequent calculus style.
-Where \citeauthor{LicataSR17} are much more agnostic in terms of
-substructurality --- allowing for non-commutative and bunched logics ---
-we are much more agnostic in terms of syntax.
-The system of \citeauthor{LicataSR17} is essentially a single calculus,
-supporting ``product'' ($\mathrm F$) types and ``function'' ($\mathrm U$)
-types, parametrised on a \emph{mode theory} describing its structural rules.
-For this system, they derive the strong result of cut elimination.
-Meanwhile, we leave syntax design to the user, and consequently can only
-guarantee substitution (which we can only get because of our commitment to
-natural deduction).
+%A type system lets us enforce in our language complex invariants which maintain
+%abstraction boundaries while also ideally being machine-checkable.
+%For example, we give functions function type to abstract away their
+%implementation via call stacks and jumps and so on.
 
-\section{Naming and notation conventions}
+%In this thesis, the invariants of interest revolve around restricting the usage
+%of variables.
+%I introduce this topic thoroughly in \cref{sec:linearity}, but in brief\ldots
 
-I assume familiarity with the Curry-Howard correspondence throughout this
-thesis.
-I make no distinction between logics and type theories, and use terminology from
-each interchangeably.
-Each following bullet point lists a collection of synonyms.
+The work of this thesis relies upon type theory in two distinct ways.
+Firstly, as I have introduced above, the main objects of study in this thesis
+are programming languages with interesting type systems.
+Secondly, however, type theory provides the basis of the proof assistant Agda I
+use to implement the aforementioned programming languages and operations upon
+them.
+I will now introduce the idea of proof assistants.
 
-\begin{itemize}
-  \item assumption, hypothesis, variable
-  \item derivation, proof, term
-  \item proposition, formula, type
-  \item connective, type former
-\end{itemize}
+A \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
+a piece of software that allows for the encoding of mathematical definitions,
+theorems, constructions, and proofs, and furthermore check that such encoded
+proofs are correct and that such encoded constructions are well formed.
+To truly by interactive, i.e.\ to actually assist, a proof assistant will
+usually have a user interface which can read partial proofs, display
+information about what more proof needs to be given, and provide actions that
+will help complete the proof.
 
-I carry out mechanised constructions and proofs in the proof assistant and
-programming language Agda~\citep{Agda}.
-Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
-similarly present non-mechanised constructions and proofs assuming a foundation
-given by dependent type theory, in a style inspired by the HoTT
-Book~\citep{hottbook}.
+Proof assistants have seen increasing use in programming language research in
+recent years.
+The most obvious reason why working in a proof assistant is seen as beneficial
+is that it ensures correctness.
+If the proof assistant accurately implements a suitable mathematical foundation,
+then any theorem proved in the proof assistant is guaranteed to be a true
+theorem of that foundation.
+These guarantees of correctness are particularly important when working with
+combinatorially complex mathematical objects, proofs about which often require
+the consideration of a large number of cases.
+Programming language syntaxes are often such complex objects, motivating the use
+of proof assistants when studying programming languages.
+
+A second reason to use proof assistants is for the assistance they provide when
+exploring a mathematical theory.
+When we make a new definition, we may want to test how it works in a special
+case, or what constructions it allows us to perform.
+In a proof assistant, the assistance tools give us immediate feedback as to what
+moves are and aren't allowed.
+For example, if we define a complex type system, a proof assistant will let us
+interactively build typing derivations, making clear any side conditions and
+types of subderivations as we go.
+Also, as I do later in this thesis, a proof assistant allows us to build a very
+general theory, and practically use that theory directly in more specific cases
+without losing rigour.
+
+Thirdly, analogously to how a strong static type system can give us more
+confidence when refactoring a program, the constant checking of proofs in a
+proof assistant gives us the confidence to change definitions and lemmas knowing
+that we will be guided towards the parts of our theory that need to be
+correspondingly changed.
+This can help if we are developing a new programming language with a changing
+specification.
+
+Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
+this thesis --- double as programming languages themselves.
+This means that we can write progams and prove properties of them using the same
+tool.
+Also, many theorems proven in such a proof assistant have computational content.
+For example, if we prove a normalisation theorem for a programming language,
+this will typically yield a (verified) normalisation algorithm for it, which we
+can really run on a computer.
+As such, a development in a proof assistant can provide a reference
+implementation of a programming language, or even --- as with Idris
+2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
+actual implementation of a programming language.
 
@@
 this thesis and openings for future work.
+
+\section{Naming and notation conventions}
+
+I assume familiarity with the Curry-Howard correspondence throughout this
+thesis.
+I make no distinction between logics and type theories, and use terminology from
+each interchangeably.
+Each following bullet point lists a collection of synonyms.
+
+\begin{itemize}
+  \item assumption, hypothesis, variable
+  \item derivation, proof, term
+  \item proposition, formula, type
+  \item connective, type former
+\end{itemize}
+
+I carry out mechanised constructions and proofs in the proof assistant and
+programming language Agda~\citep{Agda}.
+Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
+similarly present non-mechanised constructions and proofs assuming a foundation
+given by dependent type theory, in a style inspired by the HoTT
+Book~\citep{hottbook}.
+I give a fuller introduction to Agda in \cref{sec:agda-primer}.
+
+This thesis is written in \colour{}, but should be readable without.
+Agda code has syntax highlighting, and various pieces of notation related to
+usage annotations are coloured \gr{green} for emphasis.

### 2023-07-16 tex/lenv.tex
[ writing ] address all remaining todos

@@
 
-\begin{example}
-  We can derive the following instances of environment composition.
-  \begin{description}
-    \item[Composition of renamings.]
-      If $\U = \V = \W = {\sqni}$, then we can take
-      $\mathrm{lift}\,\rho\,x \coloneqq \rho(x)$.
-      This allows us to compose two renamings into a single renaming.
-    \item[Post-composition with a renaming.]
-      More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
-      the action of $\rho$ as $\mathrm{lift}$.
-      This means that we can do a composition
-      $\Gamma \env\U \Delta \env\sqni \Theta$ to get an environment of type
-      $\Gamma \env\U \Theta$.
-    \item[Composition of substitutions.]
-      If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
-      syntactic traversal (in the sense of \cref{sec:syntactic-kits} and
-      \cref{sec:lrsub}).
-      For example, if $\U = {\sqni}$, we need the action of renaming on terms
-      to show that a renaming followed by a substitution composes to a
-      substitution.
-      If $\U = {\vdash}$, then the action of substitution on terms gives us that
-      substitutions compose.
-    \item[Composing semantics with substitution.]
-      More generally, if $\V = {\vdash}$ and we have a semantics (in the sense of
-      \cref{sec:gen-sem} and \cref{sec:framework}) from
-      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
-      of terms.
-  \end{description}
-\end{example}
+We can derive the following corollaries as instances of environment composition.
+
+\begin{corollary}[Composition of renamings]\label{thm:ren-comp}
+  Given renamings $\rho : \Gamma \env\sqni \Delta$ and
+  $\sigma : \Delta \env\sqni \Theta$, we can form their composite
+  $\rho; \sigma : \Gamma \env\sqni \Theta$.
+\end{corollary}
+\begin{proof}
+  Take $\U = \V = \W = {\sqni}$ in \cref{thm:env-comp}.
+  Then let $\mathrm{lift}\,\rho\,x \coloneqq \rho(x)$.
+\end{proof}
+
+\begin{corollary}[Post-composition with a renaming]\label{thm:ren-env-comp}
+  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
+  $\sigma : \Delta \env\sqni \Theta$, we can form their composite
+  $\rho; \sigma : \Gamma \env\U \Theta$.
+\end{corollary}
+\begin{proof}
+  As in \cref{thm:ren-comp}.
+\end{proof}
+
+\begin{corollary}[Pointwise renaming of an environment]\label{thm:env-ren}
+  If $\sdtstile{}\V$ respects renaming, then so does $\env\V$ (on the left).
+\end{corollary}
+\begin{proof}
+  Suppose we have $\rho : \Gamma \env\sqni \Delta$ and
+  $\sigma : \Delta \env\V \Theta$.
+  We want to compose these via \cref{thm:env-comp} with $U = {\sqni}$ and
+  $\V = \W$.
+  The function $\mathrm{lift}$ is given exactly by the fact that $\V$ respects
+  renaming.
+\end{proof}
+
+\begin{corollary}[Composition of substitutions]\label{thm:sub-comp}
+  Given substitutions $\rho : \Gamma \env\vdash \Delta$ and
+  $\sigma : \Delta \env\vdash \Theta$, we can form their composite
+  $\rho; \sigma : \Gamma \env\vdash \Theta$.
+\end{corollary}
+\begin{proof}
+  Take $\U = \V = \W = {\vdash}$ in \cref{thm:env-comp}.
+  Then, $\mathrm{lift}$ is given by the action of a substitution on a term
+  (see \AgdaFunction{sub} in the following section).
+\end{proof}
+
+\begin{corollary}[Composing semantics with substitution]
+  If we have a semantics (in the sense of \cref{sec:gen-sem} and
+  \cref{sec:traversal}) from $\U$ to $\W$, then from an environment
+  $\rho : \Gamma \env\U \Delta$ and a substitution
+  $\sigma : \Delta \env\vdash \Theta$, we can form the composite
+  $\rho; \sigma : \Gamma \env\W \Theta$.
+\end{corollary}
 

### 2023-07-16 tex/linear.tex
[ writing ] various up to WRel example
[ writing ] address all remaining todos

@@
 calculi, like the simply typed $\lambda$-calculus and its many variants.
-Such core calculi are fairly close to some real intermediate languages,
-\todo{cite} though many intermediate languages also have polymorphism in their
-type systems, which I have not discussed, but has been considered by much prior
-work~\citep{POPLmark}.
-
-Another potential feature --- required for many desired semantics but only
-relatively recently appearing in programming languages --- that is not handled
-by the techniques of \cref{sec:simple} is restrictions on how variables can be
-used.
+Such core calculi can be seen as simplified models of real programming languages
+and/or their intermediate representations.
+A potential programming language feature --- required for many desired semantics
+but only relatively recently appearing in programming languages --- that is not
+handled by the techniques of \cref{sec:simple} is restrictions on how variables
+can be used.
 In the simple type systems covered in \cref{sec:simple}, the only restriction on
@@
 As for modality, \citet{IMO20} show that common data-flow analyses for
-imperative languages can be recast as modal type systems, though they have
-generally been developed in an ad hoc fashion.
+imperative languages can be recast as modal type systems, though such analyses
+have generally been developed in an ad hoc fashion.
 Such a recasting may be useful to understand the metatheoretic properties of
@@
 being able to apply the techniques described in \cref{sec:simple}.
-Then, I\ldots
+After having introduced these topics, I finish this chapter with a survey of
+prior work on representing such type theories in proof assistants in
+\cref{sec:linmech}.
 
@@
 
-\section{Mechanisations and systemitisations of substructural logics}
+\section{Mechanisations and systematisations of substructural logics}
 \label{sec:linmech}

### 2023-07-16 tex/linmech.tex
[ writing ] address all remaining todos

@@
 $\Gamma \vdash M : A$ corresponds to many different typing with leftovers
-sequents:
+sequents, including:
 \begin{itemize}
@@
   \item $\gr1\Gamma, \gr0x : B \vdash M : A \boxtimes \gr0\Gamma, \gr0x : B$
-  \item \&c.
 \end{itemize}
@@
 depending on what terms surround a given subterm.
+In particular, many different variations can appear for the same term in
+different subterms of the same derivation, even when the scope of the two
+occurrences is the same.
 This means that if we want to implement substitution, which involves putting a
@@
 modifying the data structures presented by \citet{McBride18} to disallow all
-contraction and weakening.
+contraction and weakening, as presented by \citet{RPKV20}.
 With the additive rules, however, we get cases where variables appear multiple
@@
 
-\subsection{Fitch-style modalities}
+The other relevant contribution from \citet{RPKV20} is to recast the
+context-splitting relations via connectives inspired by bunched
+logic~\citet{oHP99}.
+I adapt this method in \cref{sec:lnd}, but with a different notion of
+context-splitting.
+
+\subsection{Fitch-style modalities}\label{sec:fitch}
 
@@
 
-\subsection{Systematisations of substructural logics}
+\subsection{Systematisations of substructural logics}\label{sec:lin-sys}
 
@@
 
-\paragraph{Granule}
-\citet{BrunelGMZ14}, \citet{GhicaS14}, \ldots
+\paragraph{Semiring-annotated systems}
+There have been several appearances in the literature of calculi in which
+variables are annotated with some algebraic usage information which is reflected
+in the types of the calculus.
+Such calculi appear in the work of \citet{abadi99core} and
+\citet{reed10distance}, where annotations are used to control information flow
+and sensitivity to perturbations, respectively.
+Following these disparate calculi, the work of \citet{BrunelGMZ14},
+\citet{GhicaS14}, and \citet{POM14} sought to unify these calculi using
+(variations on) partially ordered semirings as the source of annotations.
+Each of these papers also provided further examples of calculi with semiring
+usage annotations.
+I discuss calculi based on partially ordered semiring annotations further in the
+following chapter, as it provides the basis for the rest of this thesis.

### 2023-07-16 tex/lr.tex
[ writing ] address all remaining todos

@@
 DILL and other calculi.
-I use this syntax to write some example programs, which will motivate the
-denotational semantics explored in \cref{sec:wrel}.
 For the rest of this thesis, particularly
-\cref{sec:semirings,sec:ren-sub-lr,sec:wrel},
+\cref{sec:semirings,sec:ren-sub-lr},
 \name{} will serve as both a prototypical
@@
 this thesis.
-I will discuss a denotational semantics in \cref{sec:wrel}.
+I will discuss a denotational semantics inspired by that of
+\citet{AbelBernardy2020} in \cref{sec:den-sem}.
 

### 2023-07-16 tex/mech-related.tex
[ writing ] address all remaining todos

@@
 \citet{Autosubst18} propose some modifications to Autosubst which, as far as I
-can tell, have not yet been incorporated, but are presented in mechanised form
-for the paper.
-Some of these modifications aim to bring Autosubst into line with
-\citeauthor{AACMM21}'s work, in particular taking semantic traversals as a
-theoretical basis.
-TODO: more.
+can tell, have not yet been incorporated, although all of the case studies in
+the paper are mechanised in Coq.
+In the paper, they adapt and extend the work of \citeauthor{ACMM17} on generic
+semantic traversals to cover a variant of system F.
+They use the term \emph{multivariate traversal} for the generalisation of
+heterogeneous substitution to semantic traversals.
+It appears that this work could be followed through to produce syntax
+descriptions covering polymorphic calculi, which would provide a route for this
+work to be incorporated into the Autosubst library.
 
@@
 trivially respects renaming.
-\todo{Every model is an environment model?}
+%\todo{Every model is an environment model?}
 
@@
 In terms of the underlying theory, both works extend multi-sorted universal
-algebra\todo{cite} with variable-binding.
+algebra%\todo{cite}
+with variable-binding.
 However, the two extensions are subtly different.
@@
 Isabelle setup amounts to an explicit model of a nominal logic within
-Isabelle/HOL.
+Isabelle/HOL\@.
 This setup resembles what \citet{AACMM21} did within Agda, and similarly what I
@@
 This means that nominal techniques have been applicable to problems outside
-standard programming language theory\todo{Find applications}.
+standard programming language theory, such as in the representation of graphs
+and in topology~\citep{GG17}.
 However, in certain ways, this is arguably not a good fit for type systems.

### 2023-07-16 tex/part.tex
[ writing ] address all remaining todos

@@
 However, we find that the most natural notion of \emph{partial semiring} is
-degenerate, in that all partial semirings are actually (total) semirings.
+degenerate, in the sense that all partial semirings are actually (total)
+semirings.
 
@@
 A partial commutative monoid is exactly a commutative monoid object in the
-category of sets and partial functions with the usual monoidal product
-\todo{Define this monoidal product somewhere.}.
+category of sets and partial functions with the usual monoidal product given
+by pairing of objects and morphisms (like the Cartesian product in $\Set$).
 However, semirings need a Cartesian category in order to state the interaction
@@
 form bimonoids in $\mathcal C$ together with the cocommutative comonoid.
+
+The distributivity laws of semirings are stated below.
+I include these to show that the cocommutative comonoids of a monoidal category
+give enough structure to state these laws.
+The other laws --- that all morphisms respect $\eta$ and $\delta$, that addition
+forms a commutative monoid, and that multiplication forms a monoid --- are
+standard in symmetric monoidal category theory.
 
@@
     (0,-1) node(res) {}
-    (0,1) node[circle,draw](del) {}
+    (0,1) node(del) {$\eta$}
     ;
@@
       (0,3) node(z) {}
-      (0,2) node[circle,draw](dup) {}
+      (0,2) node(dup) {$\delta$}
       (-1,1) node(x*) {*}
@@
       \draw (y) to[out=270,in=135] (y*);
-      \draw (dup) to[out=180,in=45] (x*);
+      \draw (dup) to[out=-150,in=45] (x*);
       \draw (dup) -- (y*);
@@
       (0,3) node(z) {}
-      (0,2) node[circle,draw](dup) {}
+      (0,2) node(dup) {$\delta$}
       (1,1) node(x*) {*}
@@
       \draw (y) to[out=270,in=45] (y*);
-      \draw (dup) to[out=0,in=135] (x*);
+      \draw (dup) to[out=-30,in=135] (x*);
       \draw (dup) -- (y*);
@@
 
-\begin{lemma}
-  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, there is
+It is well known that all commutative comonoids in $(\Set, \times)$, and indeed
+any Cartesian monoidal category, are trivial, in the sense that every object of
+$\Set$ gives rise to exactly one commutative comonoid.
+We find in the following two lemmas that this property also holds of
+$\plr{\Set_{\mathrm{part}}, \otimes}$.
+
+\begin{lemma}\label{thm:ccomon-exists}
+  For each object $X$ in $\plr{\Set_{\mathrm{part}}, {\otimes}}$, there is
   a cocommutative comonoid over $X$.
@@
   being defined for all $x$.
-  Checking that these satisfy the laws is routine.
+  Checking that these satisfy the cocomutative comonoid laws is routine.
   Alternatively, we can see that both $\eta$ and $\delta$, being total, are
@@
 
-\begin{lemma}
-  For each object $X$ in $(\mathrm{Set}_{\mathrm{part}}, {\otimes})$, any
-  comonoid over $X$ is the one described in the previous lemma.
+\begin{lemma}\label{thm:ccomon-unique}
+  For each object $X$ in $\plr{\Set_{\mathrm{part}}, {\otimes}}$, any
+  comonoid over $X$ is the one described in \cref{thm:ccomon-exists}.
 \end{lemma}
@@
   some $y$ such that $\delta(x) = (y, x)$ and $\eta(y) = ()$.
-  Symmetrically, from the right unit law, there is some $z$ such that
+  Symmetrically, from the right unit law, we get some $z$ such that
   $\delta(x) = (x, z)$ and $\eta(z) = ()$.
-  But because $\delta$, as a partial function, is deterministic, we have
+  But because $\delta$, being a partial function, is deterministic, we have
   $(y, x) = (x, z)$, giving us that $y = z = x$, and $\delta(x) = (x, x)$.
-  Moreover, $\eta(x) = \eta(y) = ()$.
+  Moreover, because the chosen $y$ is equal to $x$, we have for all $x$ that
+  $\eta(x) = ()$.
 \end{proof}
+
+That a morphism $f$ respects the $\eta$ given in \cref{thm:ccomon-exists} is
+equivalent to saying that $f$ is total.
+Therefore, all possible semiring operators in
+$\mathrm{CComon}\plr{\Set_{\mathrm{part}}, \otimes}$ are total, meaning that
+there is a corresponding semiring in $\plr{\Set, \times}$.
+
+The above reasoning shows that semirings in the category of sets and partial
+functions are not worth studying.
+If we want partiality, there appear to be two options.
+The first option is to give up on multiplication.
+We could imagine replacing the binary multiplication operator by a set of
+unary modalities satisfying fewer laws.
+In particular, I make little use of addition on the left of a multiplication,
+and multiplying by $\gr0$ on the left (as done by $\oc\gr0$) is unwanted in some
+cases (such as when encoding DILL and PD, as in \cref{sec:translation}).
+With unary modalities, we could expect all of the required laws to be
+expressible in a symmetric monoidal category.
+The second option is to use a different notion of partiality.
+The notion of partiality given by the category of sets and partial functions is
+``strict'', in that composing with an everywhere-undefined function yields an
+everywhere-undefined function.
+With a non-strict notion of partial function, we may be able to have interesting
+partial semirings.

### 2023-07-16 tex/semantics.tex
[ writing ] various up to WRel example
[ writing ] address all remaining todos

@@
 \AgdaFunction{semantics} function in \cref{sec:traversal}.
-Finally, I give some example uses of \AgdaFunction{semantics} in
-\cref{sec:example-semantics}.
-\todo{What are the examples?}
+Finally, I give the syntax-generic simultaneous renaming and substitution
+operations in \cref{sec:kit-to-sem}.
 
@@
 the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
-\todo{Refer to preliminaries}
 
@@
 \AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam} (as defined in
-\cref{sec:terms}):
+\cref{sec:terms-of-system}):
 \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
@@
 \AgdaSpace{}\AgdaSymbol{\_}.
-However, I only use this generalisation much later.
-\todo{When?}
+I use this generality in the type of \AgdaFunction{extend} in the next section.
 
@@
 the presence of additional variables introduced by weakening.
-\todo{Is wand defined?}
 
@@
 
-\section{Reifying the Kripke function space}
+\section{Reifying the Kripke function space}\label{sec:reify}
 

### 2023-07-16 tex/semirings.tex
[ writing ] address all remaining todos

@@
 
-As explained by \citet{McBride16}, the first insight is to, instead of
+This chapter proceeds as follows.
+In \cref{sec:semiring-motivation}, I give an intuitive introduction to semiring
+annotations on variables, based on replicating features of DILL (introduced in
+\cref{sec:dill}).
+Then, \cref{sec:lr} formalises the ideas of \cref{sec:semiring-motivation} into
+a calculus $\name$.
+This calculus has
+appeared in my previous work~\citep{WA21}, and can be seen as a
+simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
+%The idea of \name{} is to augment the simply typed $\lambda$-calculus with
+%annotations on free variables, and give enough types to manipulate these
+%annotations.
+Given this new calculus $\name$, the first goal is to apply the techniques of
+\cref{sec:simple} to it, yielding a simultaneous substitution operation.
+To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
+the typing rules of \name{} to not mention contexts explicitly, as was the style
+in \cref{sec:simple}.
+This new notation --- the \emph{bunched connectives} --- is versatile at
+defining simply typed usage-aware syntaxes, and I give further non-$\name$
+examples in \cref{sec:variant}.
+Finally, I justify connections to linear logic and modal logic in
+\cref{sec:translation}, where I translate $\name$ terms to and from
+DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
+
+\section{Motivation for semiring annotations}\label{sec:semiring-motivation}
+
+The question of defining calculi which do not semantically admit weakening and
+contraction but also do not rely on variables going out of scope is directly
+addressed by \citet{McBride16}.
+The first technique suggested is to, instead of
 removing variables from the context of certain subterms, add an annotation to
@@
 
-The rest of this chapter proceeds as follows.
-In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
-appeared in my previous work~\citep{WA21}, and can be seen as a
-simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
-The idea of \name{} is to augment the simply typed $\lambda$-calculus with
-annotations on free variables, and give enough types to manipulate these
-annotations.
-Given this new calculus $\name$, the first goal is to apply the techniques of
-\cref{sec:simple} to it, yielding a simultaneous substitution operation.
-To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
-the typing rules of \name{} to not mention contexts explicitly, as was the style
-in \cref{sec:simple}.
-This new notation --- the \emph{bunched connectives} --- is versatile at
-defining simply typed usage-aware syntaxes, and I give further non-$\name$
-examples in \cref{sec:rec}.
-Finally, I justify connections to linear logic and modal logic in
-\cref{sec:translation}, where I translate $\name$ terms to and from
-DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
-%The application of the techniques of \cref{sec:simple} is left to
-%\cref{sec:ren-sub-lr}.
-
 \section{A usage-annotated calculus $\name$}\label{sec:lr}
@@
 (non-commutative logics), such as Lambek calculus~\citep{Lambek58}.
-Fitch-style systems\todo{cite} probably also are precluded similarly.
+Fitch-style systems~\citep{Borghuis-thesis} probably also are precluded
+similarly.
 

### 2023-07-16 tex/simple.tex
[ writing ] address all remaining todos

@@
 each variable in the context by a term in another context.
-This simultaneous substitution operation fits the form of composition in a
-multicategory.
-Specifically, in the case of a simple intuitionistic type system, this is the
-composition of a Cartesian multicategory, which allows all of the terms being
-substituted in to share a context.
-\todo{Maybe delete multicategory remark.}
+%This simultaneous substitution operation fits the form of composition in a
+%multicategory, giving us a connection to categorical logic~\citep{}.
+%Specifically, in the case of a simple intuitionistic type system, this is the
+%composition of a Cartesian multicategory, which allows all of the terms being
+%substituted in to share a context.
 
 This chapter almost entirely reviews prior work.
-In \cref{sec:terms}, I use a presentation of the dependently typed encoding of
-the simply typed $\lambda$-calculus from \citet{AR99} to set notational
-conventions.
+First, in \cref{sec:agda-primer}, I give an introduction to Agda, the proof
+assistant I work in throughout this thesis.
+Then, in \cref{sec:terms}, I use a presentation of the dependently typed
+encoding of the simply typed $\lambda$-calculus from \citet{AR99} to set
+notational conventions.
 \Cref{sec:kits} presents the unpublished work of \citet{McBride05},
@@
 
+\input{agda-primer.tex}
 \section{Term representation}\label{sec:terms}

### 2023-07-16 tex/syntax.tex
[ writing ] address all remaining todos

@@
 
-\section{Terms of a \AgdaRecord{System}}\label{sec:terms}
+\section{Terms of a \AgdaRecord{System}}\label{sec:terms-of-system}
 
@@
 
-To specify the meaning of descriptions, we assume some \AgdaBound{X} : \AgdaFunction{ExtOpenFam},
+To specify the meaning of descriptions, we assume some
+\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
+\AgdaSpace{}\AgdaSymbol{\_},
 % \ExecuteMetaData[\Interpretationtex]{X},
@@
 The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
-are the interpretation into bunched combinators.
-The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
-they have been lifted from predicates on usage vectors to predicates on
-contexts, with the type component of the context shared throughout.
-Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
-already polymorphic (not relying on anything specific about usage vectors), so
-do not need a \AgdaFunction{$^c$} variant.
+are given by the bunched connectives, as listed below.
 
@@
 
-The most obvious way to make such an \AgdaBound{X} is to use some existing
+The most obvious way to make an
+\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
+\AgdaSpace{}\AgdaSymbol{\_}
+is to use some existing
 \AgdaFunction{OpenFam} on an extended context.
-We defined \AgdaFunction{Scope} to do this: take the new variables
+I define \AgdaFunction{Scope} to do this: take the new variables
 \AgdaBound{$\Delta$}, concatenate them onto the existing context
@@
 
-We use \AgdaFunction{Scope} to deal with new variables in syntax.
+I use \AgdaFunction{Scope} to deal with new variables in syntax.
 Terms resemble the free monad over a layer-of-syntax functor, though
@@
 variable or a use of a logical rule together with terms for each of
-the required subterms. The \AgdaFunction{Size} argument is a use of
+the required subterms.
+The \AgdaFunction{Size} argument \AgdaBound{sz} is a use of
 Agda's sized types to record that subterms are smaller than the
@@
 
-This definition uses \AgdaFunction{$\dotto$}, which, analogously to
-\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
-space.
-We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
-case two (the context and the type).
-The notation
-\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
-stands for
-\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
-\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
-\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
-\AgdaSpace{}%
-\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
-where \AgdaBound{T} is a type family with $n$ many indices.
+%This definition uses \AgdaFunction{$\dotto$}, which, analogously to
+%\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
+%space.
+%We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
+%case two (the context and the type).
+%The notation
+%\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
+%stands for
+%\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
+%\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
+%\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
+%\AgdaSpace{}%
+%\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
+%where \AgdaBound{T} is a type family with $n$ many indices.
 
@@
 Finally, I give an example term \AgdaFunction{myGraph}.
-As I said at the end of \cref{sec:terms}, writing a term out in full is tedious,
-so I instead choose to use the machinery I discuss in detail in
+As I said at the end of \cref{sec:terms-of-system}, writing a term out in full
+is tedious, so I instead choose to use the machinery I discuss in detail in
 \cref{sec:usage-elaborator}.
@@
 
-\subsection{The system $\mu\tilde\mu$}
+\subsection{The system $\mu\tilde\mu$}\label{sec:mu-mu-tilde}
 I encode a usage-annotated version of System $L$/the

### 2023-07-25 tex/agda-primer.tex
[ writing ] full draft

@@
 Also as in Haskell and ML, the function arrow nests to the right.
-However, Agda has a termination checker ensuring that all defineable functions
+However, Agda has a termination checker ensuring that all definable functions
 are total, so many Haskell functions do not have a corresponding Agda function.
@@
 argument (\AgdaBound{X}), which can typically be inferred from either the
-kargument type or the return type, so can be omitted.
+argument type or the return type, so can be omitted.
 
@@
 We start with the following incomplete definition, where the expression
-\AgdaHole{\{ \}0} marks an interation point, or \emph{hole}, in the program, to
+\AgdaHole{\{ \}0} marks an interaction point, or \emph{hole}, in the program, to
 which we can apply interactive commands to complete the program.
@@
 \AgdaSymbol{()} in the place of the impossible argument, and to not provide a
-righthand side to the clause.
+right-hand side to the clause.
 
@@
 Ordinarily, each clause of a definition gives rise to a \emph{definitional}
-equation between its lefthand side and righthand side.
+equation between its left-hand side and right-hand side.
 In intensional type theory, as implemented by Agda, definitional and
@@
 notation (without the space).
-Both notations can also be used on the lefthand side of a clausal definition as
+Both notations can also be used on the left-hand side of a clausal definition as
 \emph{copatterns}.
@@
 features of \AgdaKeyword{record}-declarations (and also
-\AgdaKeyword{data}-delcarations, but I use \AgdaKeyword{record}-declarations for
+\AgdaKeyword{data}-declarations, but I use \AgdaKeyword{record}-declarations for
 the convenience reason given in the previous paragraph).

### 2023-07-25 tex/conc.tex
[ writing ] full draft

@@
 In particular, parametric polymorphism over types can be used to significantly
-improve code reuse, and is well undestood theoretically via System $F$ and its
+improve code reuse, and is well understood theoretically via System $F$ and its
 variants.

### 2023-07-25 tex/examples.tex
[ writing ] full draft

@@
 the encoding of L/nL given in \cref{sec:dup-lnl} and give examples of syntactic
-work benefitting from the generic renaming and substitution operations.
+work benefiting from the generic renaming and substitution operations.
 
@@
 from the ambient posemiring.
-The scaling combinator also makes direct reference to the posemiring.
+The scaling connective also makes direct reference to the posemiring.
 To produce a simplified syntax description, where usage constraints are

### 2023-07-25 tex/framework.tex
[ writing ] expand and reorganise examples
[ writing ] full draft

@@
 semiring-based usage restrictions on variables.
-Syntactically, I claim that this framework ranges over all finitiary
+Syntactically, I claim that this framework ranges over all finitary
 variable-based simply typed semiring-annotated calculi, with justification by
@@
 
-\section{Example semantics}\label{sec:example-semantics}
+\chapter{Applications}\label{sec:example-semantics}
 

### 2023-07-25 tex/gen-sem.tex
[ writing ] full draft

@@
 The principal use of these aliases is to fill the right argument with a type
-(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
+(occurring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
 a context given through the point-free machinery.

### 2023-07-25 tex/gen-syn.tex
[ writing ] full draft

@@
 \cref{fig:app-lam} is interpreted as the type
-system we studied in the previous sections (simply typed $\lambda$-caluclus with
+system we studied in the previous sections (simply typed $\lambda$-calculus with
 a base type and function types).

### 2023-07-25 tex/ill.tex
[ writing ] full draft

@@
 proof of $B$ transitively using $\Delta$ and $\Gamma$.
-The exchange rule Exch says that use is invariant under permutation.
+The exchange rule \TirName{Exch} says that use is invariant under permutation.
 
@@
 the $\Box$-modality in \cref{fig:S4-axioms}, and essentially any presentation
-that does not encorporate the modality into the judgemental structure of the
+that does not incorporate the modality into the judgemental structure of the
 calculus.
@@
 \citeauthor{BBdePH93} calculus, then the \TirName{Promotion} rule would be a
-pain point because it makes us rebind all of the $oc$-typed variables we need to
+pain point because it makes us rebind all of the $\oc$-typed variables we need to
 new variables.

### 2023-07-25 tex/intro.tex
[ writing ] full draft

@@
 
-Within programming language research, type theory is a methodology for producing
-abstractions via classifying program behaviour.
+Within programming language research, type theory is a methodology for enforcing
+abstractions by classifying program behaviour.
 The simplest and most common type systems classify programs by what values they
@@
 are programming languages with interesting type systems.
-Secondly, however, type theory provides the basis of the proof assistant Agda I
+Secondly, type theory provides the basis of the proof assistant Agda I
 use to implement the aforementioned programming languages and operations upon
@@
 this thesis --- double as programming languages themselves.
-This means that we can write progams and prove properties of them using the same
+This means that we can write programs and prove properties of them using the same
 tool.
@@
 
-I assume familiarity with the Curry-Howard correspondence throughout this
-thesis.
+I assume familiarity with the Curry-Howard correspondence~\citep{Howard80}
+throughout this thesis.
 I make no distinction between logics and type theories, and use terminology from
@@
   \item assumption, hypothesis, variable
-  \item derivation, proof, term
   \item proposition, formula, type
   \item connective, type former
+  \item derivation, proof, term
+  \item derivable (formula), inhabited (type)
 \end{itemize}

### 2023-07-25 tex/kits.tex
[ writing ] full draft

@@
 We see formally that $A \to B, A$ derives $B, C \to C$ by deriving each element
-of the latter from the former --- hence the first two premises of the Subst
+of the latter from the former --- hence the first two premises of the \TirName{Subst}
 rule below, deriving $B$ and $C \to C$ from $A \to B, A$.

### 2023-07-25 tex/linmech.tex
[ writing ] full draft

@@
 $x$ was consumed in the left half or the right half.
-Such an example is also considered in related work on proof seach for linear
+Such an example is also considered in related work on proof search for linear
 logics, such as the work of \citet[p.\ 11]{WH94} and \citet[p.\ 150]{CHP00}.

### 2023-07-25 tex/lnd.tex
[ add ] Restall reference
[ writing ] full draft

@@
 \cref{fig:lr-bunched}.
-The bunched connectives also yield a reasonably consise definition of the Agda
+The bunched connectives also yield a reasonably concise definition of the Agda
 data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.
@@
 convolution~\citep{Day70}.
-It also resembles the use of ternary frames \todo{cite} in semantics of
-substructural logics.
+It also resembles the use of ternary frames in semantics of non-distributive
+logics~\citep[chapter 12]{Restall99}.
 

### 2023-07-25 tex/lr.tex
[ writing ] full draft

@@
   For example, if a variable is used covariantly in one subterm and
-  contravariantly in another, we can only make the trivial guaratee represented
+  contravariantly in another, we can only make the trivial guarantee represented
   by $\gr{\sim\sim}$.

### 2023-07-25 tex/modal.tex
[ writing ] full draft

@@
   For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
-  $\Box\Gamma$ abbrevates
+  $\Box\Gamma$ abbreviates
   $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{4$'$} and
-  \TirName{K$'$} abbrevate the obvious combinations of
+  \TirName{K$'$} abbreviate the obvious combinations of
   \TirName{4} and \TirName{K}, respectively, and \TirName{$\to$-E}.
-  I silently apply weakening thoughout the derivation for the sake of concision.
+  I silently apply weakening throughout the derivation for the sake of concision.
   \[
@@
   For axioms \TirName{K} and \TirName{4}, we have the following derivations,
-  where \TirName{T$'$} abbrevates the obvious combination of \TirName{T} and
+  where \TirName{T$'$} abbreviates the obvious combination of \TirName{T} and
   \TirName{$\to$-E}.

### 2023-07-25 tex/rec.tex
[ writing ] full draft

@@
 the extend that they all support the appropriate notion of substitution, which
-again expediates experimentation.
+again expedites experimentation.
 In this section, I give some example syntactic features that can be covered in
@@
 broader characterisation than that it intuitively seems to capture some notion
-of ``usage'' and that it supports the semantics I give in \cref{sec:wrel}.
+of ``usage'' and that it supports the semantics I give in \cref{sec:den-sem}.
 Therefore, in this subsection I give some examples of variant syntaxes others,
@@
 Following \cref{sec:lnd}, I want to turn the ad hoc constraints on $\grP$,
-$\grQ$, and $\grR$ into the result of some premise combinators.
-To do this, I introduce a new combinator $\Box^{0{+}}$ defined below,
+$\grQ$, and $\grR$ into the result of some premise connectives.
+To do this, I introduce a new connective $\Box^{0{+}}$ defined below,
 along with the resulting implicit-context typing rules.

### 2023-07-25 tex/semantics.tex
[ writing ] full draft

@@
 In our usage-aware case, $\rho$ is also updated whenever we come across linear
-combinations induced by premise combinators $I^{\sep}$, $\sep$, and
+combinations induced by premise connectives $I^{\sep}$, $\sep$, and
 $\gr r \cdot {}$.
@@
 \AgdaFunction{semantics} function in \cref{sec:traversal}.
-Finally, I give the syntax-generic simultaneous renaming and substitution
-operations in \cref{sec:kit-to-sem}.
+Finally, I show that the Kripke function space simplifies under certain
+conditions in \cref{sec:reify}, and I use that case to give the syntax-generic
+simultaneous renaming and substitution operations in \cref{sec:kit-to-sem}.
 
@@
 is a device that turns any type family into an equivalent type family
-that is judgementally injective in its indices, which helps with
+that is definitionally injective in its indices, which helps with
 Agda's type inference.
@@
   The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
-  We produce this via \cref{thm:env-postren} from an environment
+  We produce this via \cref{thm:ren-env-comp} from an environment
   $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
@@
 \ExecuteMetaData[\Syntactictex]{Sub-Kit}
+
+\section{Conclusion}
+
+In this chapter, I have completed my generalisation of the core of the work of
+\citet{AACMM21}.
+The key decision to be made in this generalisation process was that the Kripke
+function spaces would be based on the $\wand$ bunched connective.
+That this change works justifies the idea of a deep connection between linear
+syntaxes and bunched-logic-style handling of contexts.
+
+The main pay-off we have seen so far from this and the previous chapter is that
+all syntaxes we can describe respect substitution.
+However, in the following chapter, I give more specific worked examples using
+the generic semantic traversal and operations derived from it.

### 2023-07-25 tex/syntax.tex
[ writing ] full draft

@@
 The \AgdaDatatype{Premise} data type describes premises of rules,
-using the bunched combinators from \cref{fig:bunched}. A single
+using the bunched connectives from \cref{fig:bunched}. A single
 premise is introduced by the
@@
 inferring the usage contexts from the proofs because addition and
-multiplication are no longer (judgementally) injective.
+multiplication are no longer (definitionally) injective.
 The function \AgdaFunction{var\#} is a way of turning a statically known de
@@
 The \AgdaFunction{uvar\#} tool is like the \AgdaFunction{var\#} tool I used
-earlier --- allowing us to refer to variables by counting from the lefthand end
+earlier --- allowing us to refer to variables by counting from the left-hand end
 of the context.
@@
 
-Finally, while there is a simple and well motivated core of premise combinators
+Finally, while there is a simple and well motivated core of premise connectives
 --- $\dot1$, $\dottimes$, $I^{\sep}$, $\sep$, and $\gr r \cdot \plr{-}$ --- the

### 2023-07-25 tex/terms.tex
[ writing ] full draft

@@
-We could mechanise Gentzen's definition of a natural deduction system directly,
-but this definition is quite complicated.
+As part of the Curry-Howard correspondence, terms of a $\lambda$-calculus
+correspond to derivations in a natural deduction system.
+We could mechanise Gentzen's original definition of a natural deduction system
+directly, but this definition is quite complicated.
 In particular, if we want to give derivations an inductive definition, the use

### 2023-07-25 tex/trans.tex
[ writing ] full draft

@@
 I will write the latter simply as $\oc$ when it appears in DILL\@.
-I add sums and with-products to the calculus of \cite{Barber1996}, with the
+I add sums and with-products to the calculus of \citet{Barber1996}, with the
 obvious rules (stated fully in \cref{fig:dill2}).
@@
 
-\subsection{Pfenning Davies}\label{sec:trans-pd}
+\subsection{Pfenning-Davies}\label{sec:trans-pd}
 
-The translation to and from the modal system of Pfenning and Davies
-\cite{judgmental} (henceforth \emph{PD}) is similar to the translation to and
-from DILL\@.
+The translation to and from the modal system of \citet{judgmental}
+(henceforth \emph{PD}) is similar to the translation to and from DILL\@.
 I present my variant of PD, again adding some common connectives, in
@@
 
+Additionally, PD has no direct equivalents to the tensor unit and tensor
+products.
+Therefore, I note that if the semiring of usage annotations satisfies some
+simple but strong criteria (as $\instPD$ does), then we can use the Cartesian
+unit and products in their stead.
+
+\begin{lemma}\label{thm:top-meet}
+  If $\gr1 \leq \gr0$ and $\gr1 \leq \gr1 + \gr1$, then $I$ is interderivable
+  with $\top$, and for all $A$ and $B$, $A \otimes B$ is interderivable with
+  $A \with B$.
+\end{lemma}
+\begin{proof}
+  The following derivations suffice.
+  I omit standard inequalities to emphasise those which use this lemma's
+  assumptions.
+  \begin{mathpar}
+    \ebrule{%
+      \infer0[$\top$-I]{\gr1I \vdash \top}
+    }
+    \and
+    \ebrule{%
+      \infer0{\plr{\gr1} \leq \plr{\gr0}}
+      \infer1[$I$-I]{\gr1\top \vdash I}
+    }
+    \and
+    \ebrule{%
+      %\infer0{\plr{\gr1} \leq \plr{\gr1 + \gr0}}
+      \infer0[Var]{\gr1\plr{A \otimes B} \vdash A \otimes B}
+      \infer0{\plr{\gr0, \gr1, \gr1} \leq \plr{\gr0, \gr1, \gr0}}
+      \infer1[Var]{\gr0\plr{A \otimes B}, \gr1A, \gr1B \vdash A}
+      \infer0{\plr{\gr0, \gr1, \gr1} \leq \plr{\gr0, \gr0, \gr1}}
+      \infer1[Var]{\gr0\plr{A \otimes B}, \gr1A, \gr1B \vdash B}
+      \infer2[$\with$-I]{\gr0\plr{A \otimes B}, \gr1A, \gr1B \vdash A \with B}
+      \infer2[$\otimes$-E]{\gr1\plr{A \otimes B} \vdash A \with B}
+    }
+    \and
+    \ebrule{%
+      \infer0{\plr{\gr1} \leq \plr{\gr1} + \plr{\gr1}}
+      \infer0[Var]{\gr1\plr{A \with B} \vdash A \with B}
+      \infer1[$\with$-E$_0$]{\gr1\plr{A \with B} \vdash A}
+      \infer0[Var]{\gr1\plr{A \with B} \vdash A \with B}
+      \infer1[$\with$-E$_1$]{\gr1\plr{A \with B} \vdash B}
+      \infer3[$\otimes$-I]{\gr1\plr{A \with B} \vdash A \otimes B}
+    }
+  \end{mathpar}
+\end{proof}
+
+\begin{remark}
+  By multiplying through on both sides, $\gr1 \leq \gr0$ is equivalent to
+  $\forall\gr x.~\gr x \leq \gr0$, and $\gr1 \leq \gr1 + \gr1$ is equivalent to
+  $\forall\gr x.~\gr x \leq \gr x + \gr x$.
+\end{remark}
+
+Finally, I give the translation itself.
+
 \begin{proposition}[\name{} $\to$ PD]

### 2023-10-29 tex/ack.tex
[ writing ] ready for submission

My first thanks go to my supervisor, Bob Atkey.
The combination of his own knowledge and his understanding of the lacks in my
knowledge has let me have a positive and fulfilling experience as a PhD student.
The wider Mathematically Structured Programming group at Strathclyde has also
taught me much, well beyond what appears in this thesis.
The group has also often been the source of fun and excitement in research,
which I see as crucial to progress.

Special thanks go to Guillaume Allais, Jan de Muijnck-Hughes, and James McKinna,
all of whom have read all or part of this thesis in draft form and whose
comments have lead to great improvements in the final document.
Similar thanks go to the anonymous reviewers of the papers that have lead to and
formed part of this thesis.

The programming language research community as a whole has helped me get to
where I am today.
Interactions at conferences and other meetings have brought me plenty of new
perspectives.
Additionally, starting even before my PhD studies, the formerly Twitter, now
Fediverse, community of researchers and practitioners have been crucial in
getting me interested in programming languages and various more specific topics
thereof.

I thank my family, particularly for our time together in the COVID-19
lockdown/work-from-home period.
The pandemic caused difficulties for us all, but a positive approach got us
through it and let us take advantage of what we could.
More recently, I thank Ayaka, who has listened to and supported me over the last
year of write-up.
Above all, she has been very patient, and with my thesis-writing now ending, I
hope that we can live our lives more fully together.

I have made many friends outside academia during my studies while living in
Glasgow and, latterly, Edinburgh.
Particularly, those of the Strathclyde Board Games Society and Japanese language
learning groups have given me much needed fun times in evenings and weekends, as
well as deeper friendships.
I owe much of my good mood and motivation to carry on to these people.

The work described in this thesis would not have happened without the Agda
programming language/proof assistant and both the contents and style developed
in its standard library.
I thank all contributors to these tools, as well as everyone who has offered
help on how to use them.
Similarly, the preparation of this thesis and related papers was much aided by
\TeX, \LaTeX, and the various packages needed to practically produce beautiful
mathematical documents.

This document is adapted from the template by Jethro Browell
(\url{https://www.overleaf.com/latex/templates/thesis-template-for-university-of-strathclyde/nfnrnmjqyxqg}),
which was licensed under CC BY 4.0.

### 2023-10-29 tex/agda-primer.tex
[ writing ] ready for submission

@@
 order to benefit from the parts of the exposition that reside in code listings.
-The syntax of Agda is broadly similar to that of Haskell, and relatively close
-to that of Standard ML, OCaml, and Coq's Gallina sublanguage.
+The syntax of Agda is broadly similar to that of Haskell~\citep{Haskell}, and
+relatively close to that of Standard ML, OCaml, and Coq version 8's Gallina
+sublanguage~\citep{SML,OCaml,Coq}.
 I will assume that the reader is able to read basic Haskell code, and spend most
@@
 
-\subsection{Lexing and parsing}
+\subsection{Lexical structure}
 
@@
 In expressions, an underscore stands for an unspecified subterm which will be
-solved by unification.
+solved by unification~\citep{AP11,Miller92}.
 The solving of unspecified terms is canonical and respects $\beta\eta$-equality,
@@
 contains only small types, and itself is a large type.
-We can work around these size issues using \emph{universe level polymorphism},
+We can work around these size issues using
+\emph{universe level polymorphism}~\citep{BCDE22},
 as in the following definition.
@@
 Slightly more flexible than parameters are \emph{Protestant indices}%
-\footnote{The terminology of Protestant/Catholic indices comes orally from Conor
-McBride. I am not aware of a written reference.}.
+\footnote{The terminology of Protestant/Catholic indices is due to Peter
+  Hancock. The mnemonic is that Catholics believe in transubstantiation, which
+  is seen as analogous to the instantiation of Catholic indices with expressions
+  that occurs during dependent pattern matching.}.
 Protestant indices also appear to the left of the colon, and also must appear
@@
 The constructor \AgdaInductiveConstructor{refl} constructs an inhabitant of
-\AgdaArgument{M}\AgdaSpace{}\AgdaDatatype{\_$\equiv$\_}\AgdaSpace{}%
+\AgdaArgument{M}\AgdaSpace{}\AgdaDatatype{$\equiv$}\AgdaSpace{}%
 \AgdaArgument{N} only when \AgdaArgument{N} is definitionally equal to

### 2023-10-29 tex/dill.tex
[ writing ] ready for submission

@@
 linear and intuitionistic assumptions satisfy different structural rules.
-In our de Bruijn index style, where every multi-premise rule does the maximal
+In our De Bruijn index style, where every multi-premise rule does the maximal
 contraction and weakening occurs in the leafwardmost possible position,
@@
     \ebrule{%
-      \hypo{\Gamma \vdash A~\mathit{int}}
-      \hypo{\Delta \vdash B~\mathit{int}}
+      \hypo{\Gamma \vdash A~\mathit{lin}}
+      \hypo{\Delta \vdash B~\mathit{lin}}
       \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
@@
     \ebrule{%
-      \hypo{\Gamma \vdash A \multimap B~\mathit{int}}
-      \hypo{\Delta \vdash A~\mathit{int}}
+      \hypo{\Gamma \vdash A \multimap B~\mathit{lin}}
+      \hypo{\Delta \vdash A~\mathit{lin}}
       \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
@@
     \ebrule{%
-      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{int}}
+      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{lin}}
       \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}

### 2023-10-29 tex/examples.tex
[ writing ] ready for submission

@@
 convenience for converting statically known natural numbers,
-representing de Bruijn \emph{levels}, into variable terms.
+representing De Bruijn \emph{levels}, into variable terms.
 
@@
 It therefore gives an example of various parts of the framework.
+
+Given the linear theme of $\name$, one may consider using the size-based
+normalisation proof given by \citet[p.\ 71]{girard87linear} and
+\citet{Krishnaswami13}.
+However, this fails for two reasons for $\name$, both of which also apply to the
+fragment I consider here defined by
+$A, B \Coloneqq \iota \mid \oc\gr rA \multimap B$.
+Firstly, we can use the $\oc$ connective in the linear instantiation of $\name$
+to introduce contractable variables.
+The size maintenance/reduction principles do not hold in the presence of
+contraction, because substitution for a contracted variable can make the term
+grow arbitrarily larger.
+Secondly, there are instantiations other than the linear
+$\{\gr0, \gr1, \gr\omega\}$ semiring, many of which allow for non-linear
+behaviour even without the $\oc$-connective.
+For example, with the trivial semiring, we recover an intuitionistic calculus
+with no usage restrictions.
+We therefore have to use a normalisation proof which also works for standard
+Simply Typed $\lambda$-Calculus.
 
@@
 given that a standard denotational semantics would not use intermediate
-environments in the same sense as renaming and substitution, but it allows us to
-reuse the standard facts that variables support renaming and identity
-environments.
-With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
-environment entries by \AgdaFunction{lookup$^R$}.
+environments in the same sense as renaming, substitution, and evaluation.
+The only requirements are that the choice of \AgdaBound{$\V$} respects renaming
+and yields inhabitants of the target family
+\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
+Picking \AgdaRecord{\_$\sqni$\_} means that we can use the already proven lemma
+\AgdaFunction{ren\textasciicircum$\sqni$} for the former requirement; and we use
+\AgdaFunction{lookup$^R$} --- which we would have to provide somewhere anyway
+--- to yield semantic objects.
 
@@
 \input{lnl.tex}
+
+\section{Conclusion}
+
+In this chapter, I have provided a range of example syntaxes and operations on
+them, defined with the help of the generic semantic traversal.
+Where \citet{AACMM21} give examples which focus on using the semantic
+environment effectively, my examples focus more on problems we can only work on
+using usage restrictions.
+These examples tie together the two strands of this thesis --- representation of
+syntax and semantics in Agda and usage restrictions via semiring annotations.
+
+Though all of the applications have been successfully completed, there are some
+points in them where details of the framework have forced us into an unnatural
+or suboptimal approach, relative to what we could achieve with a
+reimplementation of each specific calculus from first principles.
+Specifically, in the NbE example of \cref{sec:nbe}, I was significantly hampered
+by the inability to cleanly exclude variables from the syntax of normal forms,
+and in the encoding of a linear/non-linear calculus in \cref{sec:lnl}, we were
+forced to consider variables with linear type but intuitionistic usage
+annotation and vice versa.
+I will discuss these problems further, with suggestions of solutions, in
+\cref{sec:conc}.

### 2023-10-29 tex/framework.tex
[ writing ] ready for submission

@@
 precise about the various levels of domain-specific languages which appear.
+Most of the code and text of this chapter, \cref{sec:semantics}, and the
+examples in \cref{sec:usage-elaborator,sec:den-sem} is adapted from
+\citet{WA22}.
+The updated code for all of the remainder of this thesis is available at
+\url{https://github.com/laMudri/generic-lr/releases/tag/thesis-submission}, and
+can be cited via \doi{10.5281/zenodo.10050850}.
 

### 2023-10-29 tex/intro.tex
[ writing ] ready for submission

@@
 
-This thesis advances the frontier of programming languages work that can be done
-in a proof assistant.
-I will elaborate on both of these aspects in the following paragraphs.
+Programming language design has a history stretching back almost to the
+inception of electronic computing.
+Programming language \emph{theory} has arguably an even longer history, with
+parts of mathematical logic developed in the early 20th century forming the
+vocabulary of programming language theory as we know it today.
+Programming language theory studies mathematical models of programming languages
+to help us better understand programs in existing languages, and to inform the
+design of new language features, new languages, and new programming paradigms.
 
-The main aim of programming language research is to find common patterns in
-computer programs and represent these patterns via programming language
-features.
-As a basic example, consider functions in, say, a low level language like C, in
-contrast to what these functions are compiled to in assembly language/machine
-code.
-In order to support function calls, the C runtime system manages a
-\emph{call stack}.
-A function call is compiled to a sequence of instructions which store the return
-address and the values of the arguments on the stack, move the stack pointer,
-and jump to the compiled code of the function.
-A return from a function then places the return value into the appropriate
-register, moves the stack pointer back, and jumps back to the return address.
-Advances in programming language techniques are evaluated by the usefulness of
-behaviours captured and goodness of the mathematical properties of the
-abstractions produced.
+%The field of programming language theory expands on at least two axes.
+%On one axis, we become ever more able to give multiple mathematical
+%interpretations of realistic programming language features.
+%On another, the field generates new ideas with a theoretical basis which turn
+%out to give rise to useful programming language features and paradigms.
+%Most work pushes along both of these axes, including the work covered in this
+%thesis, though most of the work covered in this thesis is phrased and directed
+%in a way that emphasises the theoretical properties over the specifics of
+%applications.
+%In phrasing things this way, I hope to impart clear intuitions about the scope
+%of this work, given its level of generality.
 
-Within programming language research, type theory is a methodology for enforcing
-abstractions by classifying program behaviour.
+%This thesis advances the frontier of programming languages work that can be done
+%in a proof assistant.
+%I will elaborate on both of these aspects in the following paragraphs.
+
+%The main aim of programming language research is to find common patterns in
+%computer programs and represent these patterns via programming language
+%features.
+%As a basic example, consider functions in, say, a low level language like C, in
+%contrast to what these functions are compiled to in assembly language/machine
+%code.
+%In order to support function calls, the C runtime system manages a
+%\emph{call stack}.
+%A function call is compiled to a sequence of instructions which store the return
+%address and the values of the arguments on the stack, move the stack pointer,
+%and jump to the compiled code of the function.
+%A return from a function then places the return value into the appropriate
+%register, moves the stack pointer back, and jumps back to the return address.
+%Advances in programming language techniques are evaluated by the usefulness of
+%behaviours captured and goodness of the mathematical properties of the
+%abstractions produced.
+
+Within programming language research, type theory is a methodology for
+classifying program behaviour.
 The simplest and most common type systems classify programs by what values they
 may return.
+For example, in many --- perhaps most --- widely used programming languages, the
+compiler will keep track of the types of expressions and their subexpressions,
+and give an error or warning to the programmer whenever there is a mismatch
+between the expected type and the actual type.
 For example, the body of a C function with return type \texttt{char} is a
@@
 throw.
-More generally, \emph{effect systems} classify programs based on all of the
-effects they may have --- such as reading input and writing to files, and also
+More generally, \emph{effect systems}~\citep{LG88} classify programs based on
+all of the effects they may have ---
+such as reading input and writing to files, and also
 internally defined effects, such as non-determinism and using an accumulator.
@@
 proofs are correct and that such encoded constructions are well formed.
-To truly by interactive, i.e.\ to actually assist, a proof assistant will
+To truly be interactive, i.e.\ to actually assist, a proof assistant will
 usually have a user interface which can read partial proofs, display
@@
 usage annotations are coloured \gr{green} for emphasis.
+As a test, the word \emph{colour} in the first sentence of this paragraph should
+appear with the first two letters red, the next two green, and the final two
+blue, with the shades of red, green, and blue used for Agda code.
+Additionally, the word \emph{green} in the second sentence is written in the
+shade of green used for usage annotations.

### 2023-10-29 tex/kits.tex
[ writing ] ready for submission

@@
 substitution could fall foul of variable capture.
-In our approach, based on de Bruijn indices, the difficulty is that an index $i$
+In our approach, based on De Bruijn indices, the difficulty is that an index $i$
 outside a binder of $n$ variables corresponds to an index $n + i$ inside the

### 2023-10-29 tex/lenv.tex
[ writing ] full draft
[ writing ] ready for submission

@@
 environment is.
-However, I find them difficult to work with, as it is often easier to do
-linear algebraic proofs separately from the rest of an environment.
-For example, for identity and composition of environments, as we are about to
-see, our working definition (\cref{def:lr-env})
-is easier to use because we can rely on the closure of linear maps under
-identity and composition.
-Concretely, an inductive construction of, for example, an identity environment
-would involve constructing an environment of type
+However, such a definition appears to require creative induction hypotheses in
+the proving of simple lemmas, in contrast to the more direct proofs I achieve
+below using \cref{def:lr-env}.
+To take a concrete example, consider how we may construct an ``identity''
+environment of type $\Gamma \env\V \Gamma$, as in \cref{thm:env-id} below.
+If we try to directly proceed by induction on $\Gamma$, we get to the case where
+we are aiming to construct an environment of type
 $\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
@@
 $\gr0\gamma, \grQ\delta \env\V \grQ\delta$.
-These are not identity environments, so we would have to strengthen the
-induction hypothesis.
-Similar strengthening of the induction hypothesis would be required for
-composition.
+These are not identity environments, and thus do not come from the hypotheses of
+a simple induction.
+In contrast, using \cref{def:lr-env}, in \cref{thm:env-id} we are able to use
+the standard fact that there are identity linear maps, and on top of such a map
+worry only about the value assigned to each variable.
 
@@
 
-I put together the preceeding pieces to give a syntactic traversal operation over
+I put together the preceding pieces to give a syntactic traversal operation over
 $\name$ in the following section.

### 2023-10-29 tex/linear.tex
[ writing ] full draft
[ writing ] ready for submission

@@
 Languages and systems implementing features inspired by linear logic include
-Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
+Rust~\citep{MK14,Rust}, Granule~\citep{Granule18}, ATS~\citep{Xi04,ZX05}, an
+experimental Haskell extension~\citep{BBNPS17}, and various implementations of
 session types~\citep{HLVCCDMPRTVTZ16}.
@@
 stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
-erasure found in Agda and Idris~\citep{Atkey18}.
+erasure found in Agda and Idris~\citep{Atkey18,Brady21}.
 
 In this chapter, I look at two standard type-theoretic features that motivate
-considering calculi going beyond simple type theories as deliniated in
+considering calculi going beyond simple type theories as delineated in
 \cref{sec:simple} or equivalently by \citet{AACMM21}.

### 2023-10-29 tex/linmech.tex
[ writing ] ready for submission

@@
 variable.
+An algebraic presentation of this fragment is given by \citet{ZD21}, in terms of
+a partial functional cancellative addition operation.
 For exponentials, though, and particularly the $\oc$-introduction rule, what I
@@
 
-\subsection{Co-de-Bruijn syntax}
+\subsection{Co-De-Bruijn syntax}
 
 \Citet{McBride18} presents a mechanisation of the simply typed
-$\lambda$-calculus in what he calls \emph{co-de-Bruijn} style.
-He notes that syntax based on de Bruijn indices, as presented in
+$\lambda$-calculus in what he calls \emph{co-De-Bruijn} style.
+He notes that syntax based on De Bruijn indices, as presented in
 \cref{sec:simple}, finds a canonical way to place contractions and weakenings
@@
 the variable rule and at rules with no premises).
-Co-de-Bruijn syntax, by contrast, finds a canonical way to place contractions
+Co-De-Bruijn syntax, by contrast, finds a canonical way to place contractions
 and weakenings by doing the reverse: weakening as early as possible (i.e., as
@@
 Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
-intuitionistic variables handled using the regular co-de-Bruijn machinery from
+intuitionistic variables handled using the regular co-De-Bruijn machinery from
 the paper.
-In summary, the co-de-Bruijn approach is promising for capturing linearity, but
+In summary, the co-De-Bruijn approach is promising for capturing linearity, but
 has not been thoroughly investigated.
@@
 context-splitting relations via connectives inspired by bunched
-logic~\citet{oHP99}.
+logic~\citep{oHP99}.
 I adapt this method in \cref{sec:lnd}, but with a different notion of

### 2023-10-29 tex/mech-related.tex
[ writing ] ready for submission

@@
 Autosubst is based on similar ideas to those \citeauthor{AACMM21} use:
-de Bruijn-indexed terms with a distinguished variable rule and notion of
+De Bruijn-indexed terms with a distinguished variable rule and notion of
 binding, acted upon by simultaneous renaming and substitution.
@@
 environments of whatever semantics is being defined.
+The focus on substitution makes this work simpler than renaming-based
+frameworks, because we get to avoid talking about renaming, whereas we always
+ultimately need to talk about substitution.
+However, what is a gain for internal simplicity is a loss for usability.
+For the syntax, starting from substitution means that the framework provides no
+help in \emph{proving} the admissibility of substitution, because the syntax is
+not considered a model \emph{until} we can show that it admits substitution.
+Similarly, for semantic models, we have to prove that each model we consider
+respects semantic substitution, which is a stronger requirement than
+respecting renaming.
 
-\begin{itemize}
-  \item Hard to bootstrap.
-  \item All terms have the same, countably infinite context.
-\end{itemize}
+Additionally, \citet{HHLM22} gain further simplicity of definitions by not
+keeping track of contexts.
+For example, untyped substitutions on a set (as opposed to scope-indexed family)
+$X$ are functions from $\mathbb N$ to $X$.
+Similarly, the simply typed families introduced in section 6 of the
+aforementioned paper are indexed on their type, but not their context.
+Ignoring scopes/contexts lets one talk about monads (as in
+\emph{De Bruijn monads}) rather than a more complex notion like relative
+monads~\citep{ACU15}.
+However, it also means losing potentially useful and important information.
 

### 2023-10-29 tex/modal.tex
[ writing ] ready for submission

@@
 To start, note the following two facts about a ``promotion'' rule similar to one
-mentioned by \citet{BBdPH93}.
+mentioned by \citet{BBdePH93}.
 
 \begin{proposition}\label{thm:NKT4-P}
-  In NJ + \TirName{N} + \TirName{K} + \TirName{4}, the following principle is
-  admissible.
+  When $\Gamma = B_1, \ldots, B_n$, let $\Box\Gamma$ abbreviate
+  $\Box B_1, \ldots, \Box B_n$.
+  Then, in NJ + \TirName{N} + \TirName{K} + \TirName{4}, the following principle
+  is admissible.
   \[
     \ebrule{%
-      \hypo{\Box B_1, \ldots, \Box B_n \vdash A}
-      \infer1[Promotion]{\Box B_1, \ldots, \Box B_n \vdash \Box A}
+      \hypo{\Box\Gamma \vdash A}
+      \infer1[Promotion]{\Box\Gamma \vdash \Box A}
     }
@@
   For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
-  $\Box\Gamma$ abbreviates
-  $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{4$'$} and
-  \TirName{K$'$} abbreviate the obvious combinations of
-  \TirName{4} and \TirName{K}, respectively, and \TirName{$\to$-E}.
+  \TirName{4$'$} and \TirName{K$'$} abbreviate the obvious combinations of
+  \TirName{4} and \TirName{K}, respectively, with \TirName{$\to$-E}.
   I silently apply weakening throughout the derivation for the sake of concision.
@@
 leaving only valid hypotheses.
-Because I am working with a de Bruijn index-style presentation, weakening needs
+Because I am working with a De Bruijn index-style presentation, weakening needs
 to be admissible.

### 2023-10-29 tex/rensub.tex
[ writing ] ready for submission

@@
 them.
-In practice, the equations rewriting other other equations build up so much that
+In practice, the equations rewriting other equations build up so much that
 I have given up on a first effort to give a treatment of the equational theory
@@
 As for the progression to \cref{def:lr-env}, this appears to arise from the fact
-that the quantitative usage information is a refinement of the intuitionistic de
+that the quantitative usage information is a refinement of the intuitionistic De
 Bruijn index-based syntax.

### 2023-10-29 tex/semirings.tex
[ writing ] ready for submission

@@
 DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
+
+This chapter and the following chapter re-present and expand the work of
+\citet{WA21}.
+For the thesis version, I have dropped mention of \emph{skew} semirings, which
+allows the algebraic components to be more robust and better abstracted.
+In particular, in \cref{sec:ren-sub-lr}, I talk about linear maps rather than
+matrices, and define environments in terms of usage-checked variables rather
+than raw well typed variables.
 

### 2023-10-29 tex/syntax.tex
[ writing ] ready for submission

@@
 multiplication are no longer (definitionally) injective.
-The function \AgdaFunction{var\#} is a way of turning a statically known de
+The function \AgdaFunction{var\#} is a way of turning a statically known De
 Bruijn level and a usage proof into an application of \AgdaInductiveConstructor{`var}.
+In the type, \AgdaPrimitive{$\infty$} is the ``infinite'' size, which all sizes
+are less than.
+Effectively, writing \AgdaPrimitive{$\infty$} allows us to ignore sizes when we
+are not doing recursion involving sizes.
 
@@
 %or other fixed points.
-%We can also use it in linear/non-linear style systems~\cite{Benton94} to make
+%We can also use it in linear/non-linear style systems~\citep{Benton94} to make
 %sure linear variables are not available in the intuitionistic fragment.

### 2023-10-29 tex/terms.tex
[ writing ] ready for submission

@@
-As part of the Curry-Howard correspondence, terms of a $\lambda$-calculus
-correspond to derivations in a natural deduction system.
+The main use I have for Agda in this thesis is to represent and reason about
+programming languages.
+In particular, I am interested in representing \emph{core languages} or
+\emph{core calculi}.
+Given a standard multi-pass compiler, the core language can be identified as the
+first stage after which there can be no compile-time errors.
+In other words, (some) terms of the core language arise from taking the source
+code for a term, parsing it, scope-checking it, type-checking it, and doing
+whatever other static checks are done to it.
+Core language terms can also arise via optimisations on other core language
+terms, or could not actually appear in the compilation of any real program, but
+nonetheless be potentially handled by the optimiser and backend.
+The core language can also be thought of as a representation of the
+\emph{meaningful} part of a programming language, excluding (meaningless)
+erroneous programs.
+
+In most compilers, the core language is represented as an annotated abstract
+syntax tree.
+By construction, the abstract syntax tree representation makes syntactically
+ill formed programs unrepresentable.
+When representing a core language in a dependently typed language, including
+Agda, we can take this idea further to make ill scoped, ill typed, and otherwise
+meaningless programs unrepresentable.
+Thus, we can truly \emph{define} a core language, and moreover reason
+mathematically about it.
+
+As is standard in the study of functional programming languages, I will take the
+core languages I consider to be variants of the simply typed
+$\lambda$-calculus~\citep{Church40,Barendregt93}.
+As part of the Curry-Howard correspondence~\citep{Howard80}, terms of a typed
+$\lambda$-calculus correspond to derivations in a natural deduction system, and
+therefore we can take inspiration in terms of definitions and methodology from
+logic when we come to mechanise core calculi in Agda.
 We could mechanise Gentzen's original definition of a natural deduction system
@@
 $\to$-introduction.
-\cref{fig:explicit-contexts} gives an example of the same derivation written
+\Cref{fig:explicit-contexts} gives an example of the same derivation written
 in Gentzen's style and in the explicit context style.
@@
 equality, and also that we never have to reason about freshness of names.
-Agda does not have support for nominal techniques~\cite{GP02}, which may have
+Agda does not have support for nominal techniques~\citep{GP02}, which may have
 made names a better option.
 
-Most mechanisations choose contexts to be an inductive list of types.
+Most mechanisations choose contexts to be an inductive list of object-language
+types.
 However, I instead choose a functional, tree-shaped representation, as shown
 with the type \AgdaRecord{Ctx}.
-The type \AgdaDatatype{LTree} is the inductive type generated by leaves and
-nullary \& binary nodes, and serves as a generalised ``length'' of the context.
+The type \AgdaDatatype{LTree} is the inductive type generated by
+leaves (\AgdaInductiveConstructor{[-]}) and
+nullary (\AgdaInductiveConstructor{$\upvarepsilon$}) \&
+binary (\AgdaInductiveConstructor{\_<+>\_}) nodes, and serves as a generalised
+``length'' of the context.
 The tree shape makes concatenation definitionally
@@
 
-The contents of the context --- the types --- are then stored in the functional
-vector \AgdaField{ty-ctx}, which is a mapping from leaves in \AgdaField{shape}
-to object language types \AgdaDatatype{Ty}.
+I use \AgdaDatatype{Ptr} to form a type family of \AgdaDatatype{LTree}-indexed
+vectors \AgdaFunction{Vector}.
+These vectors serve as the data structure containing the types, and in later
+chapters usage information, of contexts.
 The advantages of the functional vector representation will not become clear
@@
 \ExecuteMetaData[\Vectortex]{Vector}
+
+The basic operations for building vectors from parts are \AgdaFunction{[\_]} to
+create a singleton vector, \AgdaFunction{[]} to create an empty vector, and
+\AgdaFunction{\_++\_} to append two vectors.
+
+The shape of the context is usually not worth indexing over in term
+representation, so I hide this index using the record type \AgdaRecord{Ctx}.
+I also instantiate the element type of these vectors to \AgdaDatatype{Ty}, the
+type of object-language types.
+
 \Ctx{}
+
+The three operators for building vectors lift to operations on \AgdaRecord{Ctx}
+by suffixing \AgdaFunction{$^c$} --- giving us \AgdaFunction{[\_]$^c$},
+\AgdaFunction{[]$^c$}, and \AgdaFunction{\_++$^c$\_}.
 

### 2024-01-20 tex/ack.tex
[ update ] acknowledgements

@@
 formed part of this thesis.
+More recently, I also thank my examiners Kathrin Stark and Conor McBride,
+whose keen and careful reading have lead to yet more improvements to this final
+version of my thesis.
 
@@
 year of write-up.
-Above all, she has been very patient, and with my thesis-writing now ending, I
-hope that we can live our lives more fully together.
+Alas, the writing process can take a toll on relationships, so I wish her all
+the best.
 

### 2024-03-05 tex/agda-primer.tex
[ correction ] first pass over chapter 2

@@
 
-All \AgdaKeyword{data}-declarations use GADT syntax.
+All \AgdaKeyword{data}-declarations use \emph{generalised algebraic data type}
+(GADT) syntax~\citep{CH03,SP08,XCC03}.
 The body of a declaration comprises a list of constructor names paired with
@@
 The first, \AgdaDatatype{NestedList} is standard from the polymorphic recursion
-literature.
+literature~\citep{BM98}.
 It is worth noting at this point that Agda permits overloading of constructors,
@@
 This set of colours is inspired by Conor McBride's syntax highlighting for
-Epigram 2.
+Epigram.
 The colour given to a name is determined by the type of declaration that name
@@
 \AgdaBound{purple} for local variables.
+The actual rendering of Agda code into \LaTeX{} is done by a backend of the Agda
+compiler invoked via the command \texttt{agda --latex}.
 

### 2024-03-05 tex/gen-sem.tex
[ correction ] first pass over chapter 2

@@
 
+\begin{figure}
+  \begin{minipage}[t]{0.333\textwidth}
+    \ExecuteMetaData[\SimpleSem]{dotone}
+  \end{minipage}
+  \begin{minipage}[t]{0.666\textwidth}
+    \ExecuteMetaData[\SimpleSem]{dottimes}
+  \end{minipage}
+  \begin{minipage}[t]{0.549\textwidth}
+    \ExecuteMetaData[\SimpleSem]{dotto}
+  \end{minipage}
+  \begin{minipage}[t]{0.450\textwidth}
+    \ExecuteMetaData[\SimpleSem]{allb}
+  \end{minipage}
+  \begin{minipage}[t]{0.450\textwidth}
+    \ExecuteMetaData[\SimpleSem]{Circle}
+  \end{minipage}
+  \begin{minipage}[t]{0.549\textwidth}
+    \ExecuteMetaData[\SimpleSem]{Box}
+  \end{minipage}
+  \caption{Simple connectives on \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}%
+    \AgdaSpace{}\AgdaPrimitive{Set}}
+  \label{fig:simple-connectives}
+\end{figure}
+
 With the aim of abstracting away from explicit contexts, bringing us closer to
@@
 \AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
-One of the basic connectives in this setting is the \emph{pointwise} arrow
-\AgdaFunction{\_$\dotto$\_}, which acts in
+The definitions of the main connectives I use while working in
 \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
-like the non-dependent arrow does in \AgdaPrimitive{Set}.
-Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
+are listed in \cref{fig:simple-connectives}.
+In brief, \AgdaFunction{$\dot1$}, \AgdaFunction{\_$\dottimes$\_}, and
+\AgdaFunction{\_$\dotto$\_} are pointwise liftings of the singleton set, the
+Cartesian product of sets, and function sets from \AgdaPrimitive{Set} to
+\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}.
+Another component is the \AgdaFunction{$\forall[\_]$} notation, which
 embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
@@
 context.
+I use all of these connectives in both Agda code and typeset mathematical
+expressions.
+In the latter case, they will appear black rather than green.
+
 To facilitate working in this point-free setting, I give infix versions of
@@
 a context given through the point-free machinery.
-
-\ExecuteMetaData[\SimpleSem]{Circle}
 
@@
 same context.
+I give the definition of each of the fields of \AgdaFunction{RenSem} using
+copattern matching.
+The definition below defines, for example, that taking the
+\AgdaField{$\llbracket$var$\rrbracket$} projection of \AgdaFunction{RenSem} will
+yield \AgdaInductiveConstructor{var}, and that taking the
+\AgdaField{$\llbracket$var$\rrbracket$} projection and then passing an argument
+\AgdaBound{b} will yield \AgdaInductiveConstructor{lam}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaBound{b}\AgdaSpace{}\AgdaSymbol{(}%
+\AgdaFunction{$\searrow^v$}\AgdaSpace{}\AgdaFunction{here$^v$}\AgdaSymbol{)}%
+\AgdaSymbol{)}.
 
@@
 The interpretation of a context is the interpretation for each of its types.
-We still have environments storing variables, which delays the interpretation of
+
+\ExecuteMetaData[\SimpleSem]{semTy}
+\ExecuteMetaData[\SimpleSem]{semCtx}
+\ExecuteMetaData[\SimpleSem]{semvdash}
+
+Moving on to the interpretation of terms, we still have environments storing
+variables, which delays the interpretation of
 variables to the \AgdaField{$\llbracket$var$\rrbracket$} case and allows newly
@@
 \ExecuteMetaData[\SimpleKits]{GD}\ for any \AgdaBound{$\Delta$}).
-
-\ExecuteMetaData[\SimpleSem]{Box}
+Refer back to \cref{fig:simple-connectives} for the formal definition.
 
@@
 \ExecuteMetaData[\SimpleSem]{bindEnv}
+
+Because a lot of type formers have type
+\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaDatatype{Ty}%
+\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
+(for example, \AgdaRecord{\_$\ni$\_}, \AgdaDatatype{\_$\vdash$\_},
+\AgdaBound{$\V$}, and \AgdaBound{$\C$}),
+I introduce the abbreviation \AgdaFunction{OpenFam} for this type, and use the
+abbreviation from now on.
+The intuition behind the name is that such a type former contains ``open''
+things --- i.e., things with free variables, analogous to open terms --- and
+form a family of sets indexed over object language types.
+The following definition comes from the code of \cref{sec:gen-syn}, where the
+specific datatype \AgdaDatatype{Ty} has been replaced by a module parameter
+\AgdaBound{Ty}, hence the change in colour.
+
+\ExecuteMetaData[\SimpleSyntaxtex]{OpenFam}

### 2024-03-05 tex/gen-syn.tex
[ writing ] ready for submission
[ correction ] first pass over chapter 2

@@
 
+The form of syntactic rule descriptions is given in \cref{fig:simple-syntax}.
+A rule comprises a conclusion type and a collection of premises.
+Premises are defined inductively, with a single premise comprising a context
+extension $\Delta$ (a context, representing the newly bound variables in that
+subterm) and a required type $A$ (the type of the subterm).
+Individual premises can be combined in a nullary-binary tree fashion.
+Given that conjunction of premises is interpreted as a Cartesian product, which
+is associative, the tree structure is unnecessary.
+Indeed, such a tree structure does not appear in the work of \citet{AACMM21},
+who prefer a list-like representation.
+However, a tree structure will become necessary for the extensions to
+descriptions I provide in \cref{sec:linearity} and \cref{sec:framework}.
+
 \begin{figure}
   \begin{align*}
-    \text{Premises} && \mathit{ps}, \mathit{qs} &\Coloneqq
-      \Delta \vdash A \mid {} \mid \mathit{ps} \quad \mathit{qs}
+    \text{Premises} && \mathit{ps}, \mathit{qs} \Coloneqq{}& \Delta \vdash A
+    &&\text{\color{gray}(premise with new variables $\Delta$ and type $A$)} \\
+                    &&\mid{}&
+    &&\text{\color{gray}(no premises)} \\
+                    &&\mid{}& \mathit{ps} \quad \mathit{qs}
+    &&\text{\color{gray}(conjunction of premises)} \\
     \\
-    \text{Rule} && r, s &\Coloneqq {\mathit{ps} \over \vdash A}
+    \text{Rule} && r, s \Coloneqq{}& {\mathit{ps} \over \vdash A}
+    &&\text{\color{gray}(rule with premises $\mathit{ps}$ concluding $A$)}
   \end{align*}
@@
 
+What I wrote in \cref{fig:simple-syntax} as $\Delta \vdash A$, in Agda I write
+as \ExecuteMetaData[\SimpleSyntaxtex]{single-premise}.
+I write the empty tree of premises as \AgdaInductiveConstructor{`$\dot1$}, and
+conjoin trees of premises using \AgdaInductiveConstructor{\_`$\dottimes$\_}.
 A single premise with newly bound variables is interpreted by shuffling the
@@
 
+\ExecuteMetaData[\SimpleSyntaxtex]{Premises}
 \ExecuteMetaData[\SimpleSyntaxtex]{semp}
 
+I transcribe the horizontal bar in the syntax of rules given in
+\cref{fig:simple-syntax} into Agda via the infix constructor
+\AgdaInductiveConstructor{\_`$\rightarrowtriangle$\_}.
+Given that this is the only form a rule can take, I define \AgdaRecord{Rule} as
+a \AgdaKeyword{record} type rather than a \AgdaKeyword{data} type, so as to get
+projections automatically.
 A rule, with all its parameters instantiated, targets a specific type
 \AgdaBound{A$'$}, which we check to match the desired type \AgdaBound{A}.
-Finally, a whole \AgdaRecord{System} comprises a set \AgdaBound{L} of rule
-labels, and \AgdaBound{rs}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}\AgdaBound{L}%
-\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaRecord{Rule}.
+
+\ExecuteMetaData[\SimpleSyntaxtex]{Rule}
+\ExecuteMetaData[\SimpleSyntaxtex]{semr}
+
+Finally, a whole \AgdaRecord{System} comprises a set \AgdaField{Label} of rule
+labels, and \AgdaField{rules}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaField{Label}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaRecord{Rule}.
 The interpretation of these data is to pick a rule label \AgdaBound{l}, and then
@@
 
-\ExecuteMetaData[\SimpleSyntaxtex]{semr}
+\ExecuteMetaData[\SimpleSyntaxtex]{System}
 \ExecuteMetaData[\SimpleSyntaxtex]{sems}
@@
     check.
-  \item Use Agda's \emph{sized types} to remember that the subterms are smaller.
+  \item Use Agda's \emph{sized types}~\citep{Abel10} to remember that the
+    subterms are smaller.
   \item Avoid sized types, and index terms over some user-defined type (for

### 2024-03-05 tex/kits.tex
[ correction ] first pass over chapter 2

@@
 While substitution has this extra role in a lot of the syntaxes with binding we
-care about, variable-binding also significantly complicates the substitution
+care about, variable binding also significantly complicates the substitution
 operation.
@@
 of derivability restricted to structural rules.
+Renamings and substitutions are so named because of the operations
+\emph{renaming} and \emph{substitution}, which take a renaming (respectively, a
+substitution) and produce a map from terms in $\Delta$ to terms in $\Gamma$.
+The conspicuous contravariance in the previous definitions --- by which
+renamings/substitutions from $\Gamma$ to $\Delta$ are associated with maps from
+things in $\Delta$ to things in $\Gamma$ --- arises from the fact that terms
+(and, as a special case, variables) are naturally interpreted denotationally as
+maps from the interpretation of the context to the interpretation of the type of
+the term.
+Because the context is in the domain position, we expect context maps to act
+contravariantly on terms.
+Additionally, with the order I use, we have that the renamings from $\Gamma$ to
+a singleton context $A$ correspond to variables in context $\Gamma$ of type $A$,
+and similarly substitutions from $\Gamma$ to $A$ are just terms in context
+$\Gamma$ of type $A$, which is the way round we expect sets of variables and
+terms to be specified.
 
-In the derivation below, we assume the existence of a derivation of
+Following the above conventions, we expect the following admissible rules for
+simultaneous renaming and substitution.
+In this subsection, I use floor brackets $\lfloor$ and $\rfloor$ to group
+together those hypotheses that form the simultaneous renaming/substitution.
+The floor brackets are only a visual guide, and do not change the meaning
+of the rules.
+In these rules, $\Delta_i$ denotes the $i$th entry in context $\Delta$.
+
+\begin{mathpar}
+  \ebrule{%
+    \hypo{\Gamma \ni \Delta_1}
+    \delims{\left\lfloor}{\right.}
+    \hypo{\cdots}
+    \hypo{\Gamma \ni \Delta_n}
+    \delims{\left.}{\right\rfloor}
+    \hypo{\Delta \vdash A}
+    \infer4[Rename]{\Gamma \vdash A}
+  }
+  \and
+  \ebrule{%
+    \hypo{\Gamma \vdash \Delta_1}
+    \delims{\left\lfloor}{\right.}
+    \hypo{\cdots}
+    \hypo{\Gamma \vdash \Delta_n}
+    \delims{\left.}{\right\rfloor}
+    \hypo{\Delta \vdash A}
+    \infer4[Subst]{\Gamma \vdash A}
+  }
+\end{mathpar}
+
+As an example of applying \TirName{Subst}, consider the example below.
+We assume the existence of a derivation of
 $B, C \to C \vdash C$, and by the admissibility of substitution we thus have a
@@
   &\begin{prooftree}
+    % Vertical spacing
+    \infer0{F}
+    \infer1{F}
+    \rewrite{}
     \hypo{\Pi}
     \infer[no rule]1{A \to B, A \vdash B}
+    \overlay
+    \delims{\left\lfloor}{\right.}
     \infer0[Var]{A \to B, A, C \vdash C}
     \infer1[$\to$-I]{A \to B, A \vdash C \to C}
+    \delims{\left.}{\right\rfloor}
     \hypo{B, C \to C \vdash C}

### 2024-03-05 tex/mech-related.tex
[ correction ] first pass over chapter 2

@@
 user, and having substitution be just one instance.
-\citet{ACMM17} and \citet{AACMM21} provide many examples of traversals over
+\Citet{ACMM17} and \citet{AACMM21} provide many examples of traversals over
 syntax using the same generic environment management as used by substitution.
@@
 \citet{ACCL91}.
+This difference in exposure of the internals can be traced back to the differing
+implementation approaches of the two tools.
+Autosubst is implemented via metaprogramming in Coq's Ltac sublanguage, whereas
+\citeauthor{AACMM21}'s tool is implemented via generic programming within the
+main term language of Agda.
 
@@
 
-\citet{Autosubst18} propose some modifications to Autosubst which, as far as I
-can tell, have not yet been incorporated, although all of the case studies in
-the paper are mechanised in Coq.
-In the paper, they adapt and extend the work of \citeauthor{ACMM17} on generic
-semantic traversals to cover a variant of system F.
-They use the term \emph{multivariate traversal} for the generalisation of
-heterogeneous substitution to semantic traversals.
-It appears that this work could be followed through to produce syntax
-descriptions covering polymorphic calculi, which would provide a route for this
-work to be incorporated into the Autosubst library.
+\Citet{Autosubst19} describe a new version of the tool --- \emph{Autosubst 2}.
+Autosubst 2 continues the Autosubst implementation style as metaprogramming over
+Coq, but this time implements the metaprogramming as an external Haskell tool
+rather than a collection of Ltac tactics.
+Autosubst 2 moves closer to the work of \citeauthor{AACMM21} by producing types
+of well scoped terms, with scope-respecting operations over them.
+It also adopts the semantic traversals of \citet{ACMM17}, as described by
+\citet{Autosubst18}.
+Autosubst 2 surpasses both the original Autosubst and the work of
+\citet{AACMM21} in its dedicated support for mutually inductive syntax
+definitions, which would have to be encoded for use in the other two tools.
+\Citet{Autosubst19} give a running example of call-by-value System F,
+demonstrating both the handling of polymorphic calculi and mutually inductively
+defined syntax --- with expressions and values being mutually defined.
+It would be an interesting experiment to port this example to
+\citeauthor{AACMM21}'s framework, to check that these features can indeed be
+encoded.
+It is also conceivable that the syntax descriptions of \citeauthor{AACMM21}
+could be extended with explicit support for multiple mutually defined term
+sorts.
 
@@
 Marcelo Fiore and various collaborators have a long line of work aiming for a
-categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
+categorical account of variable binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
 A recent, particularly relevant paper from this line of work is that of
@@
 In terms of the underlying theory, both works extend multi-sorted universal
-algebra%\todo{cite}
-with variable-binding.
+algebra with variable binding.
 However, the two extensions are subtly different.

### 2024-03-05 tex/simple.tex
[ writing ] ready for submission
[ correction ] first pass over chapter 2

@@
 In this chapter, I review and justify the family of approaches usually used to
-represent simple type systems inside dependently typed proof assistants.
-These approaches were first presented by \citet{AR99}, who showed a way of
-representing well \emph{scoped} terms in a language with polymorphic recursion,
-and extended the representation to well \emph{typed} terms in a language with
-dependent types.
+represent simple type systems inside dependently typed proof assistants:
+intrinsically well scoped and well typed representations.
+The line of work leading to intrinsically well typed representations of syntax
+starts with \citet{BH94}, who defined, in Standard ML, a type of
+$\lambda$-calculus terms indexed on the set/type of variables they contain, and
+gave some operations on them.
+Later, \citet{AR99} (working from a suggestion by Hook) and \citet{BP99DeBruijn}
+used Haskell's support for nested
+datatypes and polymorphic recursion over them to define types of terms indexed
+only over \emph{free} variables, with a $\lambda$-abstraction having one fewer
+free variable than its immediate subterm.
+\Citet{AR99} also showed how to extend this well \emph{scoped} representation of
+terms to a well \emph{typed} representation, given a metalanguage with dependent
+types.
 The representation of terms relies on indexing on both a context --- giving the
@@
 
-This chapter almost entirely reviews prior work.
+This chapter reviews prior work, except for a mildly novel connecting stage at
+the start of \cref{sec:gen-sem} introducing the
+\AgdaFunction{$\bigcirc$}-modality.
 First, in \cref{sec:agda-primer}, I give an introduction to Agda, the proof
@@
 notational conventions.
-\Cref{sec:kits} presents the unpublished work of \citet{McBride05},
+\Cref{sec:kits} presents the work of \citet{BHKM12},
 successively deriving simultaneous renaming and simultaneous substitution for
-the terms defined in \cref{sec:terms}.
+the terms defined in \cref{sec:terms} from a single traversal of the syntax.
 The rest of the chapter generalises the shared core of renaming and substitution
@@
 Finally, I review some related work in \cref{sec:mech-related}.
+
+The Agda code displayed in this chapter is available for further study at
+\url{https://github.com/laMudri/thesis/tree/master/agda}, except for the code
+in \cref{sec:gen-syn}, which is available at
+\url{https://github.com/laMudri/generic-lr/tree/thesis/src/Generic/Simple}.
 

### 2024-03-05 tex/dill.tex
[ correction ] first pass over chapter 3

@@
 weakening automatically, as in non-substructural calculi) is new.
-Meanwhile, the rules for the $\oc$-modality are the same as those for
-the $\Box$-modality given in \cref{fig:PD}, except for the implicit structural
+Meanwhile, the rules for the $\oc$-modality are the same as \TirName{$\Box$-I}
+and \TirName{$\Box$-E} from \cref{sec:modal}, except for the implicit structural
 rules applying to linear assumptions.
@@
     }
-    \and
+    \\
     \ebrule{%
@@
     \text{(no \TirName{$0$-I})}
-    \and
+    \\
     \ebrule{%
@@
     }
-    \and
+    \\
     \text{(no \TirName{$\top$-E})}
@@
 
-Finally, notice that the rules given in \cref{fig:dill} characterise $\oc$ up to
+Notice that the rules given in \cref{fig:dill} characterise $\oc$ up to
 logical equivalence.
@@
 completeness of $\oc$, as discussed by \citet{judgmental} in the modal setting.
+
+As a concluding note on DILL, and as motivation for \cref{sec:semirings}, I
+consider the extension of DILL to type dependency.
+Such an extension has been studied by \citet{Vakar14}, and a similar extension
+to linear/non-linear logic~\citep{Benton94} has been studied by \citet{KPB15}.
+While these systems are well behaved on their own terms, both conspicuously omit
+dependency on linear values.
+This limitation means that one is not able to reason inside the language about
+linear values --- for example, we cannot state equations between linear values.
+The limitation arises from the fact that, in DILL and most other traditional
+presentations of linear logic, linear variables go out of scope in any subterms
+in which they are not used.
+As noted by \citet{McBride16}, appearance in types should not be considered
+\emph{use}.
+According to DILL, unused linear variables are not in scope, and thus linear
+variables are not in scope for types, and thus cannot appear in types.
+\Citet{McBride16} develops a system, refined by \citet{Atkey18}, in which unused
+linear variables remain in scope, and thus can appear in types.
+It is this system, based on semiring annotations on variables, which is
+most closely related to the system I present in \cref{sec:semirings}.

### 2024-03-05 tex/modal.tex
[ correction ] first pass over chapter 3

@@
   In NJ + \TirName{Promotion} + \TirName{T}, we can derive the original axioms
-  and rule of S4: \TirName{N}, \TirName{K}, \TirName{T}, and \TirName{4}
+  and rule of S4: \TirName{N}, \TirName{K}, \TirName{T}, and \TirName{4}.
 \end{proposition}
@@
     \ebrule{%
-      \infer0[T$'$]{\Box\plr{A \to B} \vdash A \to B}
-      \infer0[T$'$]{\Box A \vdash A}
+      \infer0[Var]{\Box\plr{A \to B} \vdash \Box\plr{A \to B}}
+      \infer1[T$'$]{\Box\plr{A \to B} \vdash A \to B}
+      \infer0[Var]{\Box A \vdash \Box A}
+      \infer1[T$'$]{\Box A \vdash A}
       \infer2[$\to$-E]{\Box\plr{A \to B}, \Box A \vdash B}
@@
 style.
-However, we do not want to add such a rule to our system because it is not
-stable under substitution.
+However, we do not want to add \TirName{Promotion} to our system because it is
+not stable under substitution.
 For example, let
@@
 $\Box A \wedge \Box B \vdash \Box C$.
-However, if we concluded $\Box A, \Box B \vdash \Box C$ by \TirName{Promotion},
-then it becomes unclear how to conclude $\Box A \wedge \Box B \vdash \Box C$.
+Let $A \coloneqq B \to C$.
+Then, we have the following derivation of $\Box A, \Box B \vdash \Box C$, where
+\TirName{T$'$} is the same abbreviation as it was in the proof of
+\cref{thm:PT-NKT4}.
+
+\begin{displaymath}
+  \begin{prooftree}
+    \infer0[Var]{\Box\plr{B \to C}, \Box B \vdash \Box\plr{B \to C}}
+    \infer1[T$'$]{\Box\plr{B \to C}, \Box B \vdash B \to C}
+    \infer0[Var]{\Box\plr{B \to C}, \Box B \vdash \Box B}
+    \infer1[T$'$]{\Box\plr{B \to C}, \Box B \vdash B}
+    \infer2[$\to$-E]{\Box\plr{B \to C}, \Box B \vdash C}
+    \infer1[Promotion]{\Box\plr{B \to C}, \Box B \vdash \Box C}
+  \end{prooftree}
+\end{displaymath}
+
+However, it is unclear how to derive
+$\Box\plr{B \to C} \wedge \Box B \vdash \Box C$ in NJ + \TirName{Promotion} +
+\TirName{T}.
 In particular, \TirName{Promotion} does not apply because the formula
-$\Box A \wedge \Box B$ is headed by $\wedge$, rather than $\Box$.
+$\Box\plr{B \to C} \wedge \Box B$ is headed by $\wedge$, rather than $\Box$.
 
@@
 
-The three new rules are listed in \cref{fig:PD}.
+The three new rules are listed below.
 Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
@@
 
-\begin{figure}
+\vspace{-2.5em}
 \begin{mathpar}
@@
 \end{mathpar}
-  \caption{The new rules of the \citeauthor{judgmental} presentation of IS4.}
-  \label{fig:PD}
-\end{figure}
 
@@
   If $\vDash$ respects renaming (i.e., we have a function
-  $\mathrm{ren}^\vDash :
-  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$
+  \mbox{$\mathrm{ren}^\vDash :
+  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$}
   for each $\Gamma$, $\Delta$, and $A$), then

### 2024-03-05 tex/lnd.tex
[ writing ] ready for submission
[ correction ] first pass over chapter 4

@@
 
-The first two connectives are those we've already seen for intuitionistic
+The first two connectives are those we have already seen for intuitionistic
 systems --- $\dot1$ and $\dottimes$.
-The absence of premises is encoded by $\dot1$, while the space between premises
-sharing a context is encoded by $\dottimes$.
-As for implication, I temporarily avoid giving a $\dotto$ connectives, instead
+The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
+rules for $\top$ and $\with$, respectively.
+The $\dot1$ connective encodes the absence of premises, while $\dottimes$
+encodes the space between premises sharing a context.
+As for implication, I temporarily avoid giving a $\dotto$ connective, instead
 fusing it together with $\forallb{-}$ to produce the \emph{set} of
@@
 a general principle of resource conservation.
-The prototypical rules that use $\dot1$ and $\dottimes$ are the introduction
-rules for $\top$ and $\with$, respectively.
 
@@
 data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.
+
+The one other change I have made between \cref{fig:lr} and \cref{fig:lr-bunched}
+is to abbreviate the conjunction of $\gamma \ni x : A$ and
+$\grP \leq \langle x \rvert$ as $\grP\gamma \sqni A$, and then omit the common
+context between the hypothesis and conclusion of the \TirName{Var} rule to give
+${} \sqni A$.
+Just as I prefer to work with a full context $\Gamma$ rather than its components
+$\gamma$ and $\grP$, I prefer to work with the usage-checked variables in
+$\Gamma \sqni A$, rather than ever considering non-usage-checked variables in
+$\gamma \ni A$ and then taking usage-correctness as an extra property.
 
@@
 It also resembles the use of ternary frames in semantics of non-distributive
-logics~\citep[chapter 12]{Restall99}.
+logics~\citep[chapter 12]{Restall1999}.
 
@@
 Then, we will always expect zero and addition to together form a cocommutative
-comonoid in this category.
+comonoid in this category --- i.e., we have a poset $M$ and order-respecting
+relations $0 : M \rel 1$ and ${+} : M \rel M \times M$ such that:
+
+\begin{itemize}
+  \item $\forall a,b.~\plr{\exists x.~{+}(a; x, b) \wedge 0(x; {\cdot})}
+    \iff a \leq b$
+  \item $\forall a,b.~\plr{\exists x.~{+}(a; b, x) \wedge 0(x; {\cdot})}
+    \iff a \leq b$
+  \item $\forall a,b,c,d.~\plr{\exists x.~{+}(a; x, d) \wedge {+}(x; b, c)}
+    \iff \plr{\exists y.~{+}(a; b, y) \wedge {+}(y; c, d)}$
+  \item $\forall a,b,c.~{+}(a; b, c) \iff {+}(a; c, b)$
+\end{itemize}
+
+For example, when dealing with contexts, we let $M \coloneqq \Ann^n$, with
+$0(\grR; {\cdot}) \coloneqq \plr{\grR \leq \gr0}$ and
+${+}(\grR; \grP, \grQ) \coloneqq \plr{\grR \leq \grP + \grQ}$, and where
+the required properties follow from the commutative monoid properties of vectors
+over a commutative monoid (the additive monoid of the semiring $\Ann$).
 With this structure, we can get the following equivalences and functions.
@@
 scalar multiplication
-to add a comodule structure over cosemiring $R$ to the cocommutative comonoid
+to add a comodule structure over the cocommutative comonoid
 given by zero and addition.
+I would then expect equivalences $\gr r \cdot I^* \leftrightarrowtriangle I^*$
+and $\gr r \cdot \plr{T \sep U} \leftrightarrowtriangle
+\gr r \cdot T \sep \gr r \cdot U$, which both at least hold in the case of
+vectors over a semiring.

### 2024-03-05 tex/lr.tex
[ correction ] first pass over chapter 4

@@
 I will not cover any operational semantics or equational theory of \name{} in
-this thesis.
+this thesis, though the substitution principle and notion of environment I
+develop in \cref{sec:ren-sub-lr} make it easy to define any desired operational
+semantics.
 I will discuss a denotational semantics inspired by that of
@@
 composition between orderings/renamings/substitutions and terms.
+
+One could imagine a system in which all $\leq$-constraints in \cref{fig:lr} are
+replaced by equality constraints, and the \TirName{Subuse} rule is turned into a
+non-syntax-directed defining rule of the system.
+Indeed, a similar system is presented by \citet{Granule18} among others.
+I prefer \TirName{Subuse} to be admissible in this thesis because it means that
+all algebra, including inequalities, happens in the same place.
+Having all of the algebra happening in the same place sets us up for
+\cref{sec:lnd}, in which I pick out some of the common patterns in the rules of
+\cref{fig:lr} and partially abstract away from the algebraic details.
+Using the abstractions of \cref{sec:lnd}, there is no natural place for a
+\TirName{Subuse} rule, though one could imagine adapting the operators there to
+introduce tightening of usage constraints.
+In terms of \cref{sec:framework}, where I consider $\name$-like syntaxes in
+general, one would expect to add a \TirName{Subuse} rule, similarly to the
+\TirName{Var} rule present in all syntaxes, if subusaging were not already
+admissible.
 

### 2024-03-05 tex/rec.tex
[ correction ] first pass over chapter 4

@@
 parametric in $\gr r$, $A$, and $B$.
+In particular, with the linearity semiring of \cref{def:lin-semiring}, we get
+$\oc\gr\omega\plr{A \otimes B} \multimap \oc\gr\omega A \otimes \oc\gr\omega B$,
+corresponding to the ordinary linear logic formula
+$\oc\plr{A \otimes B} \multimap \oc A \otimes \oc B$, which is not provable.
 
@@
 I define strictly positive functors syntactically, with the only notable
-restrictions being not being allowed to use the type variable $X$ in the domain
-of a function type and within a $\oc$-type.
+restrictions relative to the syntax of types with a type variable $X$
+being not being allowed to use $X$ in the domain of a function type or within a
+$\oc$-type.
 I then add least fixed points of such strictly positive functors to the syntax
@@
 \end{example}
+
+One could also consider supporting mutually inductive types, but I do not in
+this thesis for the sake of simplicity.
 
@@
   \end{align*}
+
+  Specially for lists, we can note that the \TirName{nil} constructor appears
+  exactly once (at the end of a series of \TirName{cons}es) in any value.
+  This suggests that we can handle \TirName{nil} linearly when doing recursion
+  over lists --- i.e., we do not need to handle the \TirName{nil} constructor in
+  a usage context $\grQ$ satisfying $\grQ \leq \gr0$ and
+  $\grQ \leq \grQ + \grQ$.
+
+  The key in treating \TirName{nil} linearly is to use \TirName{foldr} with
+  endofunction return type $C \multimap C$.
+  Such functions let us pass the desired \TirName{nil} value through all of the
+  \TirName{cons}es into its required place.
+  I give this construction rigorously in the derivation below, which describes
+  how to take a list, a \TirName{nil} value $\nabla^n : \grS\gamma \vdash C$,
+  and a \TirName{cons} value $\nabla^c : \grQ\gamma, \gr1A, \gr1C \vdash C$ and
+  produce a result of type $C$.
+  I have omitted side conditions
+  $\grR \leq \grP + \grQ$, $\grQ \leq \gr0$, $\grQ \leq \grQ + \grQ$, and
+  $\grT \leq \grR + \grS$ in order to save space in the figure.
+  I have also omitted subderivation $\mathrm{id}$, constructing the identity
+  function (noting that $\grQ \leq \gr0$, and therefore all of the additional
+  variables can be discarded), and the details of a use of substitution ---
+  labelled \TirName{Subst}.
+  I introduce substitution in its full generality in \cref{sec:ren-sub-lr}, but
+  the intuitive justification for using substitution here is that we apply the
+  variable of type $C \multimap C$ to the variable of type $C$ to produce a
+  value of type $C$.
+  This uses up both the function and the argument, leaving just a single linear
+  value.
+  One could also imagine introducing a let-binding, together with weakening of
+  $\gr0$-annotated variables, to make the shape of the sequent line up with that
+  of $\nabla^c$.
+
+  \begin{displaymath}
+    \begin{prooftree}
+      \hypo{\grP\gamma \vdash \mathrm{List}_A}
+      \hypo{\mathrm{id}}
+      \infer[no rule]1{\grQ\gamma \vdash C \multimap C}
+      \hypo{\nabla^c}
+      \infer[no rule]1{\grQ\gamma, \gr1A, \gr1C \vdash C}
+      \infer1[Subst]{\grQ\gamma, \gr1A, \gr1\plr{C \multimap C}, \gr1C \vdash C}
+      \infer1[$\multimap$-I]{\grQ\gamma, \gr1A, \gr1\plr{C \multimap C} \vdash C \multimap C}
+      \infer3[foldr]{\grR\gamma \vdash C \multimap C}
+      \hypo{\nabla^n}
+      \infer[no rule]1{\grS\gamma \vdash C}
+      \infer2[$\multimap$-E]{\grT\gamma \vdash C}
+    \end{prooftree}
+  \end{displaymath}
+
+  Given an annotation $\gr r$ satisfying $\gr r \leq \gr0$,
+  $\gr r \leq \gr r + \gr r$, and $\gr r \leq 1$ (for example, $\gr\omega$ from
+  \cref{def:lin-semiring}), we can produce a function
+  \[
+    \mathrm{foldr}_{A,C} : \mathrm{List}_A \multimap \oc\gr rC \multimap
+    \oc\gr r\plr{A \multimap C \multimap C} \multimap C
+  \]
+  as a simple wrapping of the \TirName{foldr} rule.
+  We can get a function $\mathrm{foldr}_{A,C}'$ as follows using
+  $\mathrm{foldr}_{A,C \multimap C}$, and similar reasoning as above to
+  treat the \TirName{nil} case linearly.
+  I use square brackets to denote the \TirName{$\oc$-I} rule, a pattern
+  matching let-expression to denote the \TirName{$\oc$-E} rule, and standard
+  $\lambda$-calculus notation for function introduction and elimination.
+  \begin{align*}
+    \mathrm{foldr}_{A,C}' &: \mathrm{List}_A \multimap C \multimap
+    \oc\gr r\plr{A \multimap C \multimap C} \multimap C \\
+    \mathrm{foldr}_{A,C}' &\coloneqq \lambda e.\lambda n.\lambda c'.~%
+                            \textrm{let }[c] \coloneqq c'\textrm{ in }
+                            \mathrm{foldr}_{A,C \multimap C}\,e\,
+                            [\lambda r.~r]\,
+                            [\lambda x.\lambda f.\lambda r.~c\,x\,(f\,r)]\,
+                            n
+  \end{align*}
+\end{example}
+
+Among the inductive types given by this $\mu$ operator are also
+\emph{$\with$-lists} (\emph{with-lists}).
+
+\begin{example}
+  Let us define \emph{$\with$-lists} as follows.
+  \[
+    \mathrm{WithList}_A \coloneqq \mu X.~\top \oplus \plr{A \with X}
+  \]
+  Unlike ordinary lists (which we may call \emph{$\otimes$-lists}),
+  $\with$-lists support a \emph{lookup} operation with the type below, where
+  $\mathbb N \coloneqq \mu X.~I \oplus X$.
+  \begin{align*}
+    \mathit{lookup} :
+    \mathrm{WithList}_A \multimap \mathbb N \multimap \top \oplus A
+  \end{align*}
+  The $\name$ term implementing $\mathit{lookup}$ in terms of iterators
+  (\TirName{$\mu$-E}) is complex and tedious, so I do not show it here.
+  Instead, I provide the following clausal specification, where
+  $\mathrm{nil}\plr{-}$ and $\mathrm{cons}\plr{-}$ are the constructors of
+  $\with$-lists, $\mathrm{zero}\plr{}$ and $\mathrm{succ}\plr{}$ are the
+  constructors of $\mathbb N$, $\mathrm{inl}\plr{-}$ and $\mathrm{inr}\plr{-}$
+  are the constructors of $\oplus$, $\langle\rangle$ constructs the
+  inhabitant of $\top$, and postfix $.\pi_l$ and $.\pi_r$ are the projections
+  out of a $\with$-pair.
+
+  \[
+    \begin{pmboxed}
+      \defaultcolumn{l}
+      \> \mathit{lookup} \> \mathrm{nil}\plr{t} \> i
+      \> \coloneqq \> \mathrm{inl}\plr{\langle\rangle} \\
+      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
+      \> \mathrm{zero}\plr{} \> \coloneqq
+      \> \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
+      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
+      \> \mathrm{succ}\plr{i} \> \coloneqq
+      \> \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i \\
+    \end{pmboxed}
+  \]
 \end{example}
@@
 
+\begin{samepage}
   \ExecuteMetaData[\Bunchedtex]{Box}
+\end{samepage}
 
@@
 
-Among the inductive types given by this $\mu$ operator are also
-\emph{$\with$-lists} (\emph{with-lists}).
-
-\begin{example}
-  Let us define \emph{$\with$-lists} as follows.
-  \[
-    \mathrm{WithList}_A \coloneqq \mu X.~\top \oplus \plr{A \with X}
-  \]
-  Unlike ordinary lists (which we may call \emph{$\otimes$-lists}),
-  $\with$-lists support a \emph{lookup} operation with the type below, where
-  $\mathbb N \coloneqq \mu X.~I \oplus X$.
-  \begin{align*}
-    \mathit{lookup} :
-    \mathrm{WithList}_A \multimap \mathbb N \multimap \top \oplus A
-  \end{align*}
-  The $\name$ term implementing $\mathit{lookup}$ in terms of iterators
-  (\TirName{$\mu$-E}) is complex and tedious, so I do not show it here.
-  Instead, I provide the following clausal specification, where
-  $\mathrm{nil}\plr{-}$ and $\mathrm{cons}\plr{-}$ are the constructors of
-  $\with$-lists, $\mathrm{zero}\plr{}$ and $\mathrm{succ}\plr{}$ are the
-  constructors of $\mathbb N$, $\mathrm{inl}\plr{-}$ and $\mathrm{inr}\plr{-}$
-  are the constructors of $\oplus$, $\langle\rangle$ constructs the
-  inhabitant of $\top$, and postfix $.\pi_l$ and $.\pi_r$ are the projections
-  out of a $\with$-pair.
-  %\begin{align*}
-  %  \mathit{lookup}\,\mathrm{nil}\plr{t}\,i
-  %  &\coloneqq \mathrm{inl}\plr{\langle\rangle} \\
-  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{zero}\plr{}
-  %  &\coloneqq \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
-  %  \mathit{lookup}\,\mathrm{cons}\plr{\mathit{xxs}}\,\mathrm{succ}\plr{i}
-  %  &\coloneqq \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i
-  %\end{align*}
-
-  %\noindent{}\hspace{-0.75ex}
-  \[
-    \begin{pmboxed}
-      \defaultcolumn{l}
-      \> \mathit{lookup} \> \mathrm{nil}\plr{t} \> i
-      \> \coloneqq \> \mathrm{inl}\plr{\langle\rangle} \\
-      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
-      \> \mathrm{zero}\plr{} \> \coloneqq
-      \> \mathrm{inr}\plr{\mathit{xxs}.\pi_l} \\
-      \> \mathit{lookup} \> \mathrm{cons}\plr{\mathit{xxs}}
-      \> \mathrm{succ}\plr{i} \> \coloneqq
-      \> \mathit{lookup}\,\plr{\mathit{xxs}.\pi_r}\,i \\
-    \end{pmboxed}
-  \]
-  %\begin{verbatim}
-  %  lookup := \xs. wfoldr(xs;
-  %    List I -o T + A;
-  %    t. \i. foldr(i; T -o T + A; \t'. inl(t'); u, r. let () := u in r) t;
-  %    xr. \i. let (i', r) :=
-  %      foldr(i;
-  %        A & (T + A) -o List I * (T + A);
-  %        \xr'. ([], prl(xr'));
-  %        u, ir. \xr'.
-  %          let () := u in
-  %          let (i', r) := ir in
-  %          (() :: i', prr(xr') prl(ir)))
-  %      xr in
-  %      foldr(i'; T + A; r; u, r'. let () := u in r'))
-  %\end{verbatim}
-  %\begin{align*}
-  %  \mathit{lookup} &:
-  %  \mathrm{WithList}_A \multimap \mathrm{List}_I \multimap \top \oplus A
-  %  \\
-  %                  &\coloneqq \lambda\mathit{xs}.~\mathrm{rec}\plr{xs; %
-  %                    \mathrm{List}_I \multimap \top \oplus A; %
-  %                    t.~\lambda i.~\mathrm{inl}\plr{%
-  %                    \mathrm{foldr}\plr{i, \top \multimap \top; \lambda t'.~t'; %
-  %                    u, r.~\textrm{let }{*} \coloneqq u\textrm{ in }r}\,t}; %
-  %                    \mathit{xr}.~\lambda i.~%
-  %                    \mathrm{foldr}\plr{i; %
-  %                    A \with \plr{\top \oplus A} \multimap \top \oplus A; %
-  %                    \lambda
-  %                    }\,\mathit{xr}}
-  %\end{align*}
-\end{example}
-
 As suggested by the notation, $\Box^{0{+}}$ may not be particularly canonical in

### 2024-03-05 tex/semirings.tex
[ correction ] first pass over chapter 4

@@
 $\gamma = A_1, \ldots, A_m$.
+Note that, in a single context $\Gamma$, components $\grP$ and $\gamma$ have the
+same length $m$ (and the same shape when represented in Agda).
 When combined, the usage context and the typing context will be of the same
@@
 interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
-I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.
+I use $\gr r\gamma$ to abbreviate $\gr rA_1, \ldots, \gr rA_m$.
 
@@
 $\oc$-introduction rule.
-DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
+DILL's $\oc$-introduction can be thought of as an $\gr\omega$-ary counterpart to
 $\otimes$-introduction, though with the same premise each time rather than
@@
 posemiring annotations, requiring only that contexts form something like a
-commutative $\mathrm{Rel}$-monoid supporting variable-binding and
+commutative $\mathrm{Rel}$-monoid supporting variable binding and
 variable access.

### 2024-03-05 tex/trans.tex
[ correction ] first pass over chapter 4

@@
 These additive type formers present no additional difficulty to the translation.
+As a cosmetic difference, I assume in $\name$ a base type $\iota_Y$ for each
+atomic formula $Y$ in DILL.
 
@@
   \end{mathpar}
+  \caption{The rules of DILL, extended with additive connectives}
   \label{fig:dill2}
-  \caption{The rules of DILL, extended with additive connectives}
 \end{figure}
@@
   \end{mathpar}
+  \caption{Embedding of DILL and PD types into \name}
   \label{fig:dill-pd}
-  \caption{Embedding of DILL and PD types into \name}
 \end{figure}
@@
 \begin{definition}\label{def:DILL-bottom-up}
-  A derivation is said to be \emph{$\instDILL$-bottom-up} if only the following
-  facts about addition and multiplication are used, and all proofs of
+  A derivation is said to be \emph{$\instDILL$-bottom-up} if only the facts
+  about addition and multiplication listed in the tables below are used,
+  and all proofs of
   inequalities not at leaves are by reflexivity (i.e, not using the facts that
   $\gr\omega \leq \gr0$ and $\gr\omega \leq \gr1$).
+  In the tables, each row corresponds to a value for the left argument, each
+  column corresponds to a value for the right argument, and the dash (-)
+  indicates a fact that is not to be used.
 
@@
 I present my variant of PD, again adding some common connectives, in
-\cref{fig:pd}
+\cref{fig:pd}.
 The main difference is the algebra at which \name{} is instantiated.
@@
   \end{mathpar}
+  \caption{The rules of PD, extended with several standard connectives}
   \label{fig:pd}
-  \caption{The rules of PD, extended with several standard connectives}
 \end{figure}
@@
   Let $\instPD$ denote the following semiring on the partially ordered set
-  $\{\valid \triangleleft \true \triangleleft \unused\}$.
+  $\{\valid < \true < \unused\}$.
   \begin{itemize}
@@
 \begin{definition}\label{def:PD-bottom-up}
-  A derivation is said to be \emph{$\instPD$-bottom-up} if only the following
-  facts about addition and multiplication are used, and all proofs of
+  A derivation is said to be \emph{$\instPD$-bottom-up} if only the facts
+  about addition and multiplication listed in the tables below are used,
+  and all proofs of
   inequalities not at leaves are by reflexivity.
+  In the tables, each row corresponds to a value for the left argument, each
+  column corresponds to a value for the right argument, and the dash (-)
+  indicates a fact that is not to be used.
 

### 2024-03-05 tex/syntax.tex
[ correction ] first pass over chapter 6

@@
 
-The \AgdaDatatype{Premise} data type describes premises of rules,
+The \AgdaDatatype{Premises} data type describes premises of rules,
 using the bunched connectives from \cref{fig:bunched}. A single
@@
 constructor names), and for each label a description of the
-corresponding rule. We use $\rhd$ as infix notation for systems to
+corresponding rule. I use $\rhd$ as infix notation for systems to
 associate the label set with the rules.
@@
 
-As an example, we transcribe a fragment of $\name$ (as defined in
-\cref{fig:lr} and \cref{fig:lr-bunched}) into a description.
-We give the set of types of
+As an example, I transcribe a fragment of $\name$ (as defined in
+\cref{fig:lr} and \cref{fig:lr-bunched}) as a description.
+I give the set of types of
 this system as a data type \AgdaDatatype{Ty} (together with a base
-type \AgdaInductiveConstructor{$\iota$}). We assume that there is a
+type \AgdaInductiveConstructor{$\iota$}).
+I assume that there is a
 posemiring \AgdaInductiveConstructor{Ann} in scope for the
@@
 rule, but the labels contain no further information about subterms or
-restrictions on the context. This will be provided when we associate
-labels with \AgdaRecord{Rule}s in a \AgdaRecord{System}.
+restrictions on the context.
+This will be provided by the \AgdaRecord{Rule} associated to each label in a
+\AgdaRecord{System}.
 
@@
 one-to-one, and the bunched premises are the same. A major difference
-is that we do not include a counterpart to the
+is that I do not include a counterpart to the
 \AgdaInductiveConstructor{var} rule in a
 \AgdaRecord{System}. Variables are common to all the systems
-representable in our framework.
+representable in this framework.
 
@@
 The next thing we want to do is to build terms in the described type system.
-The following definitions are useful for talking about types indexed over
-contexts, judgement forms, and judgement forms admitting newly bound variables,
-respectively.
+The following three definitions are useful for talking about types indexed over
+contexts.
+An \emph{open type} (\AgdaFunction{OpenType}) is a type
+indexed over a context.
+These are the things that bunched connectives connect, and thus serve as the
+interpretation of \AgdaDatatype{Premises}.
+
+\ExecuteMetaData[\Syntaxtex]{OpenType}
+
+An \emph{open family} (\AgdaFunction{OpenFam}) is
+indexed on a context and additionally on an object language type.
+Judgement forms, such as the families of variables, terms, and semantics
+thereof, are examples of open families.
 
@@
 
-To specify the meaning of descriptions, we assume some
-\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
-\AgdaSpace{}\AgdaSymbol{\_},
-% \ExecuteMetaData[\Interpretationtex]{X},
+An \emph{extensible open family} (\AgdaFunction{ExtOpenFam})
+is like an open family, but is indexed over an additional context.
+This additional context is intended to represent newly bound variables, which
+may just be added to the existing context, or the extensible open family may
+treat them differently to the existing variables.
+
+\ExecuteMetaData[\Syntaxtex]{ExtOpenFam}
+
+To specify the meaning of descriptions, let us assume some
+extensible open family \AgdaBound{X},
 over which we form one layer of syntax, using the function
@@
 %\ExecuteMetaData[\HeavyItex]{lR-term}
-
-% A layer of syntax supports the following functorial action.
-
-% \ExecuteMetaData[\Maptex]{map-s-type}
 
@@
 framework independently of each other and of $\Box^{0{+}}$.
+I define $\Box^{0{+}{*}}$ as follows.
+
+\[
+  \plr{\Box^{0{+}{*}}T}\,\grP\gamma \coloneqq
+  \plr{\grP \leq \grQ} \times
+  \plr{\grQ \leq \gr0} \times \plr{\grQ \leq \grQ + \grQ} \times
+  \plr{\forall\gr r.~\grQ \leq \gr r\grQ} \times T\,\grQ\gamma
+\]
 
@@
 rule notation in \cref{fig:LnL-bunched}.
+Recall that \AgdaDatatype{Hand} is a type containing two inhabitants ---
+\AgdaInductiveConstructor{ll} and \AgdaInductiveConstructor{rr}.
+I use the notation \AgdaFunction{[} \AgdaBound{X} \AgdaFunction{>}
+\AgdaBound{i} \AgdaFunction{<} \AgdaBound{Y} \AgdaFunction{]}, for
+\AgdaBound{i} \AgdaSymbol: \AgdaDatatype{Hand}, to mean \AgdaBound{X} when
+\AgdaBound{i} is \AgdaInductiveConstructor{ll} and \AgdaBound{Y} when
+\AgdaBound{i} is \AgdaInductiveConstructor{rr}.
 

### 2024-03-06 tex/examples.tex
[ correction ] finish chapter 7

@@
   We have a function of type
-  $\mathrm{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any open
-  families $\V$ and $\C$.
+  $\AgdaFunction{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any
+  open families $\V$ and $\C$.
 \end{lemma}
 \begin{proof}
-  Let $b : \mathrm{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
+  Let $b : \AgdaFunction{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
   We give $b$ the identity renaming (corresponding to no extension of the

### 2024-03-06 tex/semantics.tex
[ correction ] finish chapter 7

@@
 already come with a recursion and an induction principle.
-However, these principles do not handle variable-binding, which we have seen
+However, these principles do not handle variable binding, which we have seen
 with the fact that we had to write the \AgdaFunction{bind} helper
@@
 \AgdaBound{$\Psi$}.
+The things contained in a layer of syntax are inhabitants of extensible open
+families --- either subterms with newly bound variables \AgdaBound{$\Theta$} or
+semantic interpretations thereof.
+Therefore, function \AgdaBound{f} maps between extensible open families
+\AgdaBound{X} and \AgdaBound{Y}, which are given the same newly bound variables
+and result type.
 
@@
 It is used to describe how to go under binders in a
-Higher-Order Abstract Syntax style: To go under a binder we must
+parametric higher-order abstract syntax (PHOAS) style~\citep{Chlipala08}:
+To go under a binder we must
 provide semantic interpretations for all the additional variables.
@@
 
-\AgdaFunction{Wrap}
-is a device that turns any type family into an equivalent type family
-that is definitionally injective in its indices, which helps with
-Agda's type inference.
-It turns the type family into a parametrised
-record with a single field \AgdaField{get} whose type is the type in
-the body of the $\lambda$-abstraction.
+I introduced \AgdaRecord{Wrap} in \cref{sec:records} as a device that turns a
+unary type family into an equivalent type family that is definitionally
+injective in its indices, which helps with Agda's type inference.
+It turns the original type family into a parametrised
+record with a single field \AgdaField{get}, whose type is the type in
+the body of the $\lambda$-abstraction, and a constructor
+\AgdaInductiveConstructor{mk} mapping from the original family to the wrapped
+family.
+The green \AgdaFunction{Wrap} is an extension of the blue \AgdaRecord{Wrap} to
+type families of arbitrary arity, made possible thanks to the work of
+\citet{Allais19}.
 For understanding the meaning of
@@
 variable, at the multiplicity of its annotation, together with the
-resources supporting each of those values. We use the ``magic wand''
+resources supporting each of those values. I use the ``magic wand''
 function space here to enforce the invariant that the freshly bound
 variables have usage annotations that are added to the existing
-variables, not shared with them. The use of the
+variables, not shared with them.
+This use of the magic wand goes hand-in-hand with the understanding of
+environments as separating conjunctions developed in \cref{sec:lrkits}, and both
+are crucial refinements of the previous intuitionistic work.
+The use of the
 \AgdaFunction{$\Box^r$} modality ensures that we can still use it in
@@
   $v : {\sqni} \rightarrowtriangle \V$, we get a function of type
-  $\mathrm{Kripke}\,\V\,\C \rightarrowtriangle \mathrm{Scope}\,\C$ for any $\C$.
+  $\AgdaFunction{Kripke}\,\V\,\C \rightarrowtriangle \AgdaFunction{Scope}\,\C$
+  for any $\C$.
 \end{lemma}
 \begin{proof}
-  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
+  Let $b : \AgdaFunction{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
   That is, $b$ is a Kripke function yielding $\C$-computations

### 2024-03-06 tex/examples.tex
[ correction ] finish chapter 8

@@
 annotations, the type of types does not depend on the type of usage annotations.
-This means that, in our simplified syntax, terms have types from the original
+This means that, in the simplified syntax, terms have types from the original
 system even though variables have trivial usage annotations.
-We define the $\mathbf0$ posemiring as follows, being careful to use the
+I define the $\mathbf0$ posemiring as follows, being careful to use the
 0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
@@
 canonically inferred.
+This inference extends to entire contexts.
+Recall that a context is a record containing a size and two functional vectors.
+This means that, if we give the size and vector of types, the context is
+determined up to definitional equality, because we know that the usage vector
+is a function that always returns the unique usage annotation.
 
@@
 \ExecuteMetaData[\UsageChecktex]{NonDetInverses}
+
+\begin{example}\label{def:01w-inv}
+  The example \AgdaRecord{NonDetInverses} I use in this section is for the
+  linearity ($\instDILL$) semiring.
+  The fields of the record are filled by the following functions.
+  They implement all of the ``bottom-up'' laws, as defined in
+  \cref{def:DILL-bottom-up}.
+
+  To invert both \AgdaInductiveConstructor{0\#} and
+  \AgdaInductiveConstructor{1\#}, we give one result whenever the respective
+  relation (\AgdaBound{x} \AgdaDatatype{$\leq$} \AgdaInductiveConstructor{0\#}
+  and \AgdaBound{x} \AgdaDatatype{$\leq$} \AgdaInductiveConstructor{1\#}) holds.
+  This means giving a result both when \AgdaBound{x} is the annotation itself,
+  and also when \AgdaBound{x} is \AgdaInductiveConstructor{$\omega$\#}.
+  We are forced into this choice by the need to make $\gr\omega$-annotated
+  variables both discardable and usable, without any intervening operations.
+
+  \noindent
+  \begin{minipage}[t]{0.5\textwidth}
+    \ExecuteMetaData[\ZeroOneManytex]{0-inv}
+  \end{minipage}
+  \begin{minipage}[t]{0.5\textwidth}
+    \ExecuteMetaData[\ZeroOneManytex]{1-inv}
+  \end{minipage}
+
+  In inverting addition, we return all possible results for
+  \AgdaInductiveConstructor{0\#} and \AgdaInductiveConstructor{1\#}, but
+  restrict the results for \AgdaInductiveConstructor{$\omega$\#} so as to avoid
+  unnecessarily specialisation of $\gr\omega$-annotated variables into $0$- or
+  $1$-annotated variables.
+  Any variables annotated $\gr\omega$ will stay annotated $\gr\omega$ after
+  applying rules using separating conjunction.
+
+  \begin{samepage}
+    \ExecuteMetaData[\ZeroOneManytex]{plus-inv}
+  \end{samepage}
+
+  Similarly to the case for addition, when inverting multiplication we apply
+  the bottom-up restrictions.
+  We enforce that multiplying by \AgdaInductiveConstructor{1\#} acts as a strict
+  identity function, rather than allowing specialisation of
+  \AgdaInductiveConstructor{$\omega$\#}.
+  For \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{0\#}, we know that
+  \AgdaBound{z} has to be \AgdaInductiveConstructor{0\#} by annihilation.
+  However, this much leaves \AgdaBound{y} completely unspecified, so we choose
+  the most permissive result \AgdaInductiveConstructor{$\omega$\#}.
+  For \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{$\omega$\#},
+  \AgdaInductiveConstructor{0\#}-annotated variables stay
+  \AgdaInductiveConstructor{0\#}-annotated,
+  \AgdaInductiveConstructor{$\omega$\#}-annotated variables stay
+  \AgdaInductiveConstructor{$\omega$\#}-annotated, and
+  \AgdaInductiveConstructor{1\#}-annotated variables are not allowed in the
+  conclusion.
+  The choices for
+  \AgdaBound{x} \AgdaSymbol{=} \AgdaInductiveConstructor{$\omega$\#} correspond
+  to how the $\oc$-introduction rule in DILL works.
+
+  \begin{samepage}
+    \ExecuteMetaData[\ZeroOneManytex]{mult-inv}
+  \end{samepage}
+
+  As argued by \cref{thm:DILL-bottom-up}, the bottom-up restrictions do not
+  constrain the range of meaningfully different terms that can be produced.
+\end{example}
 
@@
 annotations for the example at the end of \cref{sec:terms-of-system}.
-This allows us to write:
+Instantiating the elaborator to the \AgdaRecord{NonDetInverses} given in
+\cref{def:01w-inv} allows us to write:
+
 \ExecuteMetaData[\PaperExamplestex]{cojoin}
-We have instantiated the usage elaborator so that:
-\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
-empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
-$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
-\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
-$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
-gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
-[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
-$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
-consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
-this splitting would introduce more non-determinism but not allow any
-more terms to be typed. As such, the only non-determinism comes when
-we have variables annotated $\gr1$ and need to do an additive split,
-like when we apply the \AgdaInductiveConstructor{!E} rule below. At
-this point, the variable can become either $\gr0$-annotated in the
-left subterm and $\gr1$-annotated on the right, or vice-versa. We will
-find that, because the left subterm wants to use that variable, the
-former choice will be rejected. The function \AgdaFunction{var\#} is a
-convenience for converting statically known natural numbers,
-representing De Bruijn \emph{levels}, into variable terms.
+
+When elaborating this definition, the only non-determinism comes when
+applying the \AgdaInductiveConstructor{!E}-rule with the
+\AgdaInductiveConstructor{1\#}-annotated variable of type
+\AgdaInductiveConstructor{!} \AgdaInductiveConstructor{$\omega$\#} %
+\AgdaBound{A} in scope.
+The non-determinism is resolved in the left subterm of this application, as the
+application of the variable rule forces the annotation to become not
+\AgdaInductiveConstructor{0\#}, hence \AgdaInductiveConstructor{1\#}, on the
+left, and therefore become \AgdaInductiveConstructor{0\#} on the right.
+
+Compare the one-line definition of \AgdaFunction{cojoin-!$\omega$} above with
+the equivalent definition we saw in \cref{sec:terms-of-system} not using the
+usage elaborator, in which we have to repeatedly specify both the usage contexts
+and the (entirely obvious) proofs relating them.
+
+\ExecuteMetaData[\HeavyItex]{cojoin-explicit}
+
+%The function \AgdaFunction{var\#} is a
+%convenience for converting statically known natural numbers,
+%representing De Bruijn \emph{levels}, into variable terms.
 
@@
 the function type is interpreted by the Kripke function space (essentially a
-special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$, though reusing
-the \AgdaFunction{Kripke} definition causes difficulty for the termination
-checker).
+special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$)
+\footnote{
+  Using
+  \AgdaFunction{Kripke} \AgdaFunction{\_$\vDash\_$} \AgdaFunction{\_$\vDash\_$}
+  here causes difficulty for the termination checker because
+  \AgdaFunction{Wrap}, and hence \AgdaFunction{Kripke}, does not reduce, leaving
+  occurrences of \AgdaFunction{\_$\vDash\_$} not applied to smaller (or indeed
+  any) arguments.
+}.
 The interpretation of the base type additionally requires a \AgdaRecord{Lift}
@@
 
+However, we can see by inspection that the algorithm given above is producing
+normal forms and neutral forms in the places we would expect them to be.
+At function type, \AgdaFunction{nbeReify} adds a $\lambda$-abstraction, around
+the result of a recursive call, so if the base type values (terms) it receives
+from the semantics are always normal, \AgdaFunction{nbeReify} will always
+produce normal forms.
+In fact, we expect the base type terms to be neutral forms.
+We can see that they are neutral forms because we only ever pass
+\AgdaFunction{nbeReflect} neutral forms, and the only larger terms it produces
+are applications of the term it is given to normal forms coming from
+\AgdaFunction{nbeReify}.
+The traversal \AgdaFunction{evalSem} does not directly produce any terms; terms
+(of base type) only appear via the environment.
+
 Also, I have not given an account of many of the connectives of $\name$.
@@
 family.
-For a simple denotational semantics, we set $\V$ to be variables
+For a simple denotational semantics, we set $\V$ to be variables --- the minimal
+family of values able to retain enough information to be meaningful in any
+semantics --- and gradually build up an identity renaming ($\sqni$-environment)
+as we go under binders.
 
@@
 This provides an additional example to those of \citet{AbelBernardy2020}.
+The example includes the following type declaration, giving the carrier type of
+the variance posemiring.
+For the operations, refer to \cref{def:monotonicity-posemiring}.
+
+\ExecuteMetaData[\Monotex]{WayUp}
 

### 2024-03-06 tex/syntax.tex
[ correction ] finish chapter 8

@@
 The bound variables all have type \AgdaInductiveConstructor{wire} and usage
-annotation $\gr1$ (written \AgdaInductiveConstructor{u1} in Agda) to make them
+annotation $\gr1$ (written \AgdaInductiveConstructor{1\#} in Agda) to make them
 behave linearly.

### 2024-03-07 tex/lrsub.tex
[ correction ] first pass over chapter 5

@@
 
-I can now show that, using the notion of \emph{environment} derived in
+I now show that, using the notion of \emph{environment} derived in
 \cref{sec:lrkits}, we can replicate the Agda proofs from
@@
 
-We start with a slightly modified definition of \AgdaRecord{Kit}.
-We saw in \cref{thm:lr-bind} that in the usage-annotated context, we restrict
-weakening of $\V$-values to just $\gr0$-use variables.
-Meanwhile, the function $\mathrm{vr}$, also seen in \cref{thm:lr-bind}, maps
-usage-checked variables to $\V$-values, and the function $\mathrm{tm}$, used
-to coerce $V$-values yielded by the environment into terms, stays the same.
-I state weakening in a slightly different way than previously, so as to help
+I start with the definition \AgdaFunction{Weakening}, which says what it means
+for a family of values \AgdaBound{$\V$} to respect one step of weakening on the
+right by $\gr0$-use variables.
+I state weakening in a slightly different way to what appears in the statement
+of \cref{thm:lr-bind}, so as to help
 unification against a known result type (avoiding the problem described by
@@
 to a value in the whole of $\grP\gamma$.
+
+\ExecuteMetaData[\LRKits]{Weakening}
+
+Given this new definition of \AgdaFunction{Weakening}, the record
+\AgdaRecord{Kit} remains largely unchanged relative to what we saw in
+\cref{sec:syntactic-kits}.
+We still want $\V$-values to respect weakening (as used in \cref{thm:lr-bind}),
+and to support maps \AgdaField{vr} from variables (also used in
+\cref{thm:lr-bind}) and \AgdaField{tm} into terms (used when traversal reaches a
+variable, we get a value from the environment, and want to produce a term from
+that value).
+As well as \AgdaFunction{Weakening}, note that \AgdaRecord{\_$\sqni$\_} and, of
+course, \AgdaDatatype{\_$\vdash$\_} have different definitions to the
+corresponding intuitionistic notions, but they still represent morally the same
+parts of the language and its metatheory.
 
@@
 we can reproduce the syntactic traversal \AgdaFunction{trav}.
-With all these lemmas in place, writing \AgdaFunction{trav}
+Similarly to the unchanged high-level definition of \AgdaRecord{Kit}, we are
+aiming for an unchanged traversal principle expressed by the rule below.
+When $\V$ has a \AgdaRecord{Kit} structure, and we have a $\V$-environment from
+$\Gamma$ to $\Delta$, we can transform a term in $\Delta$ to a term in $\Gamma$
+of the same type.
+
+\[
+  \begin{prooftree}
+    \hypo{\text{\AgdaRecord{Kit}\AgdaSpace{}}\V}
+    \hypo{\Gamma \env\V \Delta}
+    \hypo{\Delta \vdash A}
+    \infer3[Trav]{\Gamma \vdash A}
+  \end{prooftree}
+\]
+
+With all the lemmas of \cref{sec:lenv} in place, writing \AgdaFunction{trav}
 becomes routine.

### 2024-03-07 tex/rensub.tex
[ correction ] first pass over chapter 5

@@
 In this chapter, I develop the necessary syntactic metatheory for
-specifying and implementing the substitution operation.
-I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
-significant changes to the underlying notion of \emph{environment} before doing
-so.
-I give and informally motivate these changes to environments in
-\cref{sec:lrkits}, and prove some properties of the new definition in
+specifying and implementing the simultaneous substitution operation.
+My aim in this chapter is to replay the development in \cref{sec:kits}, making
+minimal changes to lemmas and theorems so as to support semiring usage
+annotations.
+The big change I make in this chapter is to the definition of
+\emph{environment}.
+In terms of ideas, I believe that the definition of environment I give in
+\cref{sec:lrkits} is the main contribution of this thesis.
+Its primary novel feature is the use of a linear map --- in the sense found in
+linear algebra --- to distribute usage annotations across the values in the
+environment.
+
+The resulting substitution operation bears some similarity to a multi-variable
+substitution operation given by \citet{petricek-thesis}.
+This is the only other presentation I have found in the literature of an
+operation similar to simultaneous substitution for a linear or linear-like
+calculus.
+I discuss Petricek's substitution lemma in more detail in \cref{sec:petricek}.
+In brief, relative to Petricek's work, I extract the environment from the
+substitution lemma, tidy up the definition, and present usage-annotated
+environments as a reusable tool and object of study in their own right.
+I reuse the resulting definition of environment in the remainder of this thesis,
+taking advantage of the properties and generality of definition established in
+this chapter, particularly in \cref{sec:semantics,sec:example-semantics}.
+
+The body of this chapter starts with a motivation for and a definition of
+environment in the usage-annotated setting in \cref{sec:lrkits}.
+I then give some key properties of and operations on these environments in
 \cref{sec:lenv}.
-Finally, I apply these new environments to the syntax of $\name$ in
-\cref{sec:lrsub} to derive renaming and substitution operators.
+I apply the previous definitions and properties to the problems of simultaneous
+renaming and substitution for $\name$ in \cref{sec:lrsub}, which also connects
+the previous sections' mathematical developments with Agda code.
+I then compare the yielded substitution operation with the one given by Petricek
+in \cref{sec:petricek}, before a concluding discussion in
+\cref{sec:ren-sub-lr-conc}.
 
@@
 \input{petricek.tex}
+
 \section{Conclusion}\label{sec:ren-sub-lr-conc}
